### 2005-05-09 docs/background.tex
moving fs into the proper directory. moving the predicate stuff to the predicate directory. deleting some random stuff

\section{Sorting}

Sorting algorithms are used to arrange items consisting of \n{record}s and
\n{key}s. A key gives order to a collection of data called a record.  A real-life
example is a phone book: the records, phone numbers, are sorted by their keys,
names and addresses.

Sorts are frequently classified as \n{in-place} and \n{out-of-place}. An
out-of-place sort typically uses an extra array of the same size as the array to
be sorted. This may require the results to be copied back at the end. An
in-place sort does not, though it may use a large stack, especially in recursive
algorithms.

A \n{stable} sort is one in which records which have the same key stay in the
same order across the sort. All the elementary sorts are stable, as is
mergesort, heapsort and radixsort. Quicksort is not stable.

Sorting algorithms tend to be written using an \n{Abstract Data Type}, rather
than for a specific data type. This allows efficient reuse of the sorts, usually
as library functions. An example from the C standard library is the \cc{qsort}
function, which performs a sort on an arbitrary data type based on a comparative
function passed to it.

The ADT interface in this project is taken from \cite{Sedgewick02}. An \cc{Item}
is a record, and a function is defined to extract its key.  Comparisons are done
via the \cc{less} function, and swapping is done using the \cc{exch} function.
These are replaced with their actual functions by the C preprocessor. The code
used for ADTs is in Figure \vref{Definitions used by the ADT}. Items in this
project were \cc{unsigned integer}s, and were compared using a simple
\n{less-than} function\footnote{This abstraction broke down in several cases,
such as where a \n{greater-than} function was required, such as in quicksort}.
In this case, a record and key are equivalent, and the word \n{key} is used to
refer to items to be sorted.

\code{Definitions used by the ADT}{adt.h}

There are special cases among sorts, where some sorts perform better on some
types of data. Most data to be sorted generally involves a small key.  This type
of data has low-cost comparisons and low-cost exchanges.  However, if strings
are being compared, then the comparison would have a much higher cost than the
exchange, since comparing keys could involved many comparisons - in the worst
case one on each letter.

Sorts which are not comparison based also exist. These sorts are called counting
based sorts, as they count the number of keys to be put in each segment of an
array before putting them in place. The most important of these is radixsort,
discussed in Chapter \ref{Radixsort}.

Sorts sometimes use a sentinel to avoid exceeding bounds of an array. If an
algorithm spends its time moving a key right based on a comparison with the key
next to it, then it will need to check that it doesn't go over the edge of the
array. This check, which can be expensive, can be replaced with a sentinel. This
is done by placing a maximal key off the right of the array, or by finding the
largest key in the array, and placing it in the final position. This technique
is used in quicksort and insertion sort.

\section{Caches}
Processors run many times faster than main memory. As a result, a processor
which must constantly wait for memory is not used to its full potential. To
overcome this problem, a \n{cache hierarchy} is used. A cache contains a small
subset of main memory in a smaller memory, which is much faster to access than
main memory. Modern computers typically contain many levels of cache, each with
increasing size and \n{latency}\footnote{Latency is the time between a request
being made, and its successful execution.}. For example, it may take twice as
long to access the level 1 cache as it does to access a register. It may take
ten times as long as that to access the level 2 cache, and fifty times as long
to access main memory.

When a cache contains the data being sought, then this is called a cache
\n{hit}. When it does not, it is called a cache \n{miss}, and the next largest
cache in the hierarchy is queried for the data. There are three types of cache
miss: \n{compulsory} or \n{cold-start} misses, \n{conflict} misses and
\n{capacity} misses. A compulsory miss occurs when the data is first fetched
from memory. Since it has not been accessed before, it cannot already be in the
cache. A conflict miss occurs when the addressing scheme used by the cache
causes two separate memory locations to be mapped to the same area of the cache.
This only occurs in direct mapped or set-associative cache, described below.
Finally, a capacity miss occurs when the data sought has been ejected from the
cache due to a lack of space.

Caches range from direct mapped to fully associative. In a direct mapped cache,
every memory address has an exact position in the cache where it can be found.
Since the cache is smaller than main memory, many addresses map to the same position.
Fully associative is the opposite of this: data can be stored anywhere in the
cache. A set-associative cache is a mixture of both: more than one piece of data
can be stored in the same place. A 4-way set-associative cache allows four items
to be stored in the same place; when a fifth item needs to be stored in that
position, one of the items is ejected. The item may be ejected randomly, in a
specific order, or the \n{least recently used} item may be removed.

When data is loaded into the cache, data near it is loaded with it. This set of
data is stored in a \n{cache line} or \n{cache block}. The associativity and
length of the cache line dictate the mapping of memory addresses to cache
storage. Figure \vref{Addressing} shows how a memory address is divided.

\begin{figure}[h]
\centering
\includegraphics{cache_addressing}
\caption{Addressing}
\label{Addressing}
\end{figure}

This cache is a 2MB direct mapped cache.  There is a 32 byte cache line, which
is indexed by the 5 bit \n{offset}. There are $2^{16}$ cache lines, which are
indexed by the middle 16 bits of the address. The most significant 11 bits form
the \n{tag}. A cache line is tagged to check if the data stored at an index is
the same as the data requested. For a fully associative cache, there is no
index, and all the bits except the offset make up the tag. A fully associative
cache has no conflict misses for this reason; similarly, a direct mapped cache
has no capacity misses, since even when full, misses are due to conflicts.

The address used by the cache is not necessarily the address used by a program.
The addresses used within a program 
are called \n{virtual addresses}. Some caches are addressed by virtual address,
some caches by \n{physical address}, and some caches by \n{bus
address}\footnote{A bus address is used to address parts of the computer
connected over a bus, including main memory, multimedia controllers such as
sound and graphics cards, and storage devices such as hard drivers and floppy
disk controllers.}. Several optimisations in later chapters use the address to
offset arrays so that conflicts are not caused. However, because a program
cannot be sure of what type of addressing is used, the actual manipulated
address may not directly relate to the index used to address the cache. This
problem is not encountered in simulations in this project, as SimpleScalar
addresses its caches by virtual address.

Caches work on the principle of \n{locality}. It is expected that after some
data has been used once, it is likely to be used again in the near future. This
is called \n{\label{temporal locality}temporal locality}. Similarly, if some data is being used, it is
likely that data near it will also be used soon, such as two consecutive
elements of an array. This is called \n{spatial locality}. Many of the
algorithms in this project are designed to take advantage of locality. Spatial
locality is exploited by ensuring that once a cache line is loaded, each key in
the cache line is used. Temporal locality is exploited by ensuring that an
algorithm uses data as much as it can as soon as it is loaded, rather than over
a large time scale, during which it may have been ejected from the cache.

\section{Branch Predictors}

A branch is a conditional statement used to control the next instruction a
machine must execute. Typically, these are \cc{if} and \cc{while} statements,
or their derivatives. Branches cause delays in a processor's instruction
pipeline.  The processor must wait until the branch is resolved before it knows
what the next instruction is, and so it cannot fetch that instruction, and must
stall. To avoid this problem, modern processors use branch predictors, which
guess the branch direction and allow the processors to execute subsequent
instructions straight away.

Each branch can be resolved in two ways: \n{taken} or \n{not-taken}. If the branch is
taken, then the program counter moves to the address provided by the branch
instruction. Otherwise, the program counter increments, and takes the next
sequential instruction.

If a branch is mispredicted, then a long delay will occur, though this delay is
the same as if a branch predictor had not been used. During this delay, the
pipeline must be emptied, the instructions already executed must be discarded,
and the next instruction must be fetched.

For this project, I split branches into two types. The first is a \n{flow
control branch}. Flow control refers to controlling the flow of execution of a
program, and branches of this type are loops and other control statements.
The second type is a \n{comparative branch}: these are comparisons between two
pieces of data, such as two keys in an array being sorted. A branch can be both
a flow control and a comparative branch, but these will be referred to as
comparative branches here. This is due to the properties of the branches: a
comparative branch should not be easily predictable; a flow control branch
should be easy to predict based on previous predictions of the same branch.

There are several kinds of branch predictors: \n{static}, \n{semi-static} and
\n{dynamic}. This project is only interested in dynamic predictors, due to
their high levels of accuracy. The other types are none-the-less discussed
below, as an introduction to dynamic predictors. A good discussion on predictor
types is \cite{Uht97}, from which the percentages below are taken. These
predictors were tested by Uht on the SPECint92 benchmarking suite. This is a
standard suite of integer-based programs used to measure the performance of an
architecture against another. Programs in this benchmark include \cc{gcc}, a
major C compiler, \cc{espresso}, which generates Programmable Logic Arrays and
\cc{li}, a LISP interpreter.

\subsection{Static Predictors}
A static predictor makes hardwired predictions, not based on information about
the program itself. Typically the processor will predict that branches are all
taken, or all not-taken, with 60\% and 40\% accuracy, respectively. Another type
of static branch predictor is Backwards-Taken-Forwards-Not-taken. Backward
branches are usually taken, so this increases the prediction accuracy to 65\%.

\subsection{Semi-static Predictors}
Semi-static predictors rely on the compiler to make their predictions for them.
The predictions don't change over the execution of a program, but the
improvement in predicting forward branches, which predominantly execute one way,
increases the accuracy to 75\%.

\subsection{Dynamic Predictors}
Dynamic predictors store a small amount of data about each branch, which
predicts the direction of the next branch. This data changes over the course of
a program's execution. Examples of this type of predictor are \n{1-bit},
\n{bimodal} and \n{two-level adaptive}.

\subsubsection{1-bit}
A 1-bit dynamic predictor uses a single bit to store whether a particular branch was last
taken or not-taken.  If there is a misprediction, then the bit is changed. 1-bit
predictors have the potential to mispredict every time, if the branch is
alternately taken and not taken. In the case of a loop with the sequence
T-T-T-N-T-T-T-N, a 1-bit predictor will mispredict twice per loop, or about half
the time.  Typically, though, it achieves an accuracy of 77 to 79\%.

\subsubsection{Bimodal}
A 2-bit dynamic predictor is called a bimodal predictor. It is a simple 2-bit
counter, each combination representing a state. In states 00 and 01, not-taken
is predicted; in states 10 and 11, taken is predicted. Each taken prediction
increments the count, to a maximum of 11, and each not-taken prediction
decrements the counter, to a minimum of zero.

In the sequence T-T-T-N-T-T-T-N, a 2-bit predictor will only mispredict once.
In general, it will achieve 78 to 89\% accuracy.

\subsubsection{Two-Level Adaptive}
A two-level adaptive predictor uses branch history to predict future branches. A
global branch history register keeps track of the last few branches, and uses
this to index a table of 2-bit predictors, which predict the result of the
branch. Frequently, the program counter is also included, either concatenated or
exclusive-ored with the branch history to form the table index.

This type of predictor can predict the two cases above perfectly.  However, the
table can take time to fill, leading to cold-start misses, much like those of a
cache. Typically, a two-level adaptive predictor will be accurate 93\% of the time.

\section{Big O notation}
Big O notation is used to describe the complexity of an algorithm. It says that
the algorithm will complete in a certain number of steps, in relation to number
of items of data being worked on, without giving specifics of how many
instructions each step will take to execute. It is possible, therefore, that an
algorithm with a lower complexity will complete slower than an algorithm where
each step is executed very quickly, but which has a greater order of complexity.

Big O notation is used to describe how an algorithm scales. Doubling the number
of elements to be sorted by an $O(N)$ algorithm will roughly double the time it
takes to execute, but will quadruple the running time of an $O(N^2)$ algorithm.

An algorithm which is $O(N)$ is said to complete in linear time. An algorithm
which is $O(N^2)$ is said to complete in quadratic time.  $O(NlogN)$ algorithms
may take several orders of magnitude more time than $O(N)$ algorithms, and
typically use the \n{divide-and-conquer} principle, or use a binary tree to
reduce complexity. Quicksort, mergesort and heapsort are all of this type,
whereas elementary sorts, such as insertion sort, selection sort, bubblesort and
shakersort are $O(N^2)$ algorithms. Radixsort is $O(N)$, but can be more
expensive than quicksort, due to the cost of each step.  These are discussed in
later chapters.

An important thing to note is that the time the algorithms take to execute are
proportion to their Order, but different constants of proportionality are used
for different algorithms. An algorithm with $10N$ instructions is $O(N)$, and is
not differentiated from an algorithm with $1000N$ instructions by this
notation.

### 2005-05-09 docs/bugs.tex
moving fs into the proper directory. moving the predicate stuff to the predicate directory. deleting some random stuff

\section{Bugs}
This section lists some of the bugs that exist in the remaining code. In a
project of limited timescale, it is not possible to remove every possible bug
from the code. These bugs are listed here in order that it is possible to
continue on this work with minimal difficutly.

\section{Algorithmic}

\label{bug1}
In multi-quicksort, the results include the sentinel check which should have
been removed before the results were measured. This adds one branch per \cc{N /
THRESHOLD} keys , and the same number of extra instructions. In the same
algorithm, the old number of 1018 is the number of keys in a block of a list.
The effects of this are not significant. Also, the check for the sentinel was
not removed, though the effect of this is minor.

\label{bug2}
In several of the mergesorts, it is possible that an extra merge step is being
taken. This has not been properly analysed, due to time restrictions. It is,
however, obvious from the level 2 miss graph of all the merges, that this
does not affect base mergesort, but it does affect both double mergesorts, and
the tiled mergesort. It is uncertain if multi mergesort is affected. The problem
stems from the end condition of the merge. It should end once the size of the next
merge is equal the size of the array, rather than greater than. However, the
algorithms do not properly merge when this is the case.

\label{bug4}
Double tiled mergesort and double multi-mergesort were optimized to sort in the
level 1 cache first. The number of keys that fit into the cache was fixed at
2048; this number should have been half that, since the sort is out of place.
This probably resulted in thrashing, especially since the arrays were aligned.
Removing this would improve the sorts.

\section{Results}
\label{bug3}
To reduce the costs of creating the array of random numbers, this time was
measured and subtracted from the results. While this works as expected for
instruction count, the effects on cache misses is less clear. Subtracting level
2 cache misses incurred during filling removes compulsory misses from the
equation. For shorter lists, the compulsory misses are completely removed.
For longer lists, those greater than the size of the cache, compulsory misses
come back, since the keys at the start of the array are flushed from the cache
by the time they begin to be sorted. LaMarca's work centers around
reducing conflict misses, however, so this is not necessarily a problem, just
something to be aware of.

### 2005-05-09 docs/code_list.tex
moving fs into the proper directory. moving the predicate stuff to the predicate directory. deleting some random stuff

\section{Simple Script}
Below is the code for the script used to run the SimpleScalar simulations.

\lgrindfile{code/simple_code}
\label{Simple Script}

\section{Sample Simulation Run}
\label{Simple List}
The instructions listed below form a sample simulation run of SimpleScalar. Each
of the commands is run in turn and its results stored by \cc{simple}.

\lgrindfile{code/sample_sim_list}

\section{Sample Simulation}
The text below is an output of a simgle run of \cc{sim-bpred}.
\label{Sample Simulation}

\lgrindfile{code/sample_output}

\section{Convert Script}
The following is the source code to the script used to convert data from the
stored format of the \cc{simple} script to that required by gnuplot.
\label{Convert Script}

\lgrindfile{code/convert}

\section{Sample Gnuplot Script}
\label{Sample Gnuplot Script}
The following script is used to make plot results. The image made by this file
is in Figure \vref{All cycles}.

\lgrindfile{code/cycles_script}

\section{Sort listings}
Included below is a code listing of the sorts used in this project. Note that
the code is not exact. It is edited for brevity and clarity, and may not even
compile. In addition, loops are rolled, edge-cases are removed and variable
declarations are moved or removed. The actual code used is contained on the CD
accompanying this project.
\label{Sort Listings}

Base heapsort is listed in Figure \ref{Simple Base Heapsort}.\\
\textbf{Cache Heapsort:}
\lgrindfile{code/cache_heapsort_code}

\textbf{Base Mergesort:}
\lgrindfile{code/base_mergesort_code}

Tiled Mergesort is very similar to double-tiled mergesort, and so is not
included. Merge and presort functions are also not included, as they are very
similar to those of base mergesort.\\
\textbf{Double Tiled Mergesort:}
\lgrindfile{code/double_tiled_mergesort_code}

\textbf{Double Multi-Mergesort:}
\lgrindfile{code/double_multi_mergesort_code}

\textbf{Base Quicksort:}
\lgrindfile{code/base_quicksort_code}

\textbf{Cache Quicksort:}
\lgrindfile{code/cache_quicksort_code}

\textbf{Multi-Quicksort:}
\lgrindfile{code/multi_quicksort_code}

\textbf{Sequential Multi-Quicksort:}
\lgrindfile{code/multi_quicksort_seq_code}

\textbf{Radixsort:}
\lgrindfile{code/radixsort_code}

### 2005-05-09 docs/conclusions.tex
moving fs into the proper directory. moving the predicate stuff to the predicate directory. deleting some random stuff

This section presents a summary of the results from previous chapters, lists
important results and discoveries, and suggests contributions made to computer
science by this project.

\section{Cache Results}

The test results in this project mostly agreed with LaMarca's results.  The
heapsort graphs had nearly exactly the same shape as LaMarca's, and the ratios
between the base and cache-conscious versions were nearly identical. The
multi-mergesort results were slightly different: instruction count was
identical, but while the cache results of the base and tiled mergesorts were
identical, multi-mergesort results were not the same. However, the basis for
those results - a sharp increase in cache misses followed by a flat-lined graph
- were in both cases.

The quicksort results were also very similar, though the multi-quicksort results
had a slightly different shape to LaMarca's. However, the instruction count
results for each quicksorts agreed with LaMarca's results, as did cache miss
results of the base and cache-conscious versions.

The two sets of radixsort results were slightly similar. The rise and plateau of
the curve to the right of the graphs was the same in both cases, though the left
hand side of the graphs did not agree.

Another interesting observation can be made: the results of the simulations on
the Pentium 4 were not always influenced by the cache misses. Though heapsort
and radixsort began to perform poorly once the data set no longer fit in the
cache, neither quicksort nor mergesort were affected. The reason why radixsort
was affected is obvious - there is no temporal reuse; in heapsort, the full
cache line is not being used, and spatial locality suffers as a result. It is
interesting that the other two sorts were unaffected, even though their cache
miss graphs indicate a significant increase in misses.

Finally, it was shown that in some cases a direct-mapped cache performed better
than a fully-associative one. The reasons for this were not clear.

\section{Branch Prediction Results}

Insertion sort has only one branch misprediction per key. This shows that it is
possible to predict comparative branches correctly, though the instruction cost
of such an algorithm may be high. This is also shown in multi-quicksort: though
the instruction count and memory reference count are both doubled, it is still
faster to use a sequential search than a binary search across small lists.

Selection sort has a similar number of instructions and cache misses as
bubblesort and a larger instruction and cache miss count than shakersort, yet it
still considerably outperforms both of these. This shows the high cost of branch
mispredictions on an algorithm.

Radixsort has almost no branch mispredictions at all. The only mispredictions
are due to its flow control, but it escapes comparative mispredictions, and this
factor allows it to perform as fast as, and in some cases faster than quicksort,
despite its high number of cache misses.

The heapsort results show the disadvantage of using unsorted lists. Even though
there are less branches in the 4-heap version, the number of branch
mispredictions are higher than in a standard 2-heap.

It is also shown that bimodal branch predictors generally out-perform two-level
adaptive predictors when resolving comparative misses. This is shown by the
results of insertion sort, bubblesort, shakersort, heapsort and quicksort. In
the cases where the two-level adaptive predictor does perform as well, the table
required is much larger than that required by the bimodal predictors to get the
same results.

\section{Best Sort}
The results also helped answer the question of which is the best sort. However,
it is important to remember that sorts have different purposes and properties,
so this shall be considered. A comparative chart of the fastest sorts is shown
in Figure \ref{All cycles}. This shows the cycles per key of each of the major
sorts and their derivatives, measured using Pentium 4 performance counters.
Neither the elementary sorts nor heapsorts are on the chart, as their results
are significantly worse than the other results, which makes it difficult to
accurately see those results.

\afterpage{
\thispagestyle{empty}
\clearpage
\enlargethispage{14em}
\vspace*{-10em}
\setcounter{figure}{8}
\begin{figure}[H]
\begin{changemargin}
\includegraphics[angle=90,scale=1.8]{plots/all_cycles}
\vspace*{8em}
\end{changemargin}
\caption{\label{All cycles}Cycles per key of several major sorts and their variations - this was
measured on a Pentium 4 using hardware performance counters.}
\end{figure}
\newpage
}

The best stable sort is radixsort. The best comparison-based stable sort is
double tiled mergesort. The best in-place sort is quicksort. The best in-place
stable sort is, sadly, insertion sort. The best elementary sort is also
insertion sort; as such it should be only used for very small data sets.  The
best sort for very large data sets is radixsort. The best overall sort was
quicksort, though radixsort is very close, and the radixsort considered here was
not optimised. It is expected that an optimised radixsort would beat quicksort
for smaller sets.

\section{Contributions}

This section lists the contributions of this project. Each of these is novel  
and was discovered or developed during the course of this project.

The primary contribution of this project is that it repeats the work of LaMarca
on cache-conscious sorting, and validates many of his claims, while questioning
others. The next most important contribution is having performed a branch
prediction analysis of major sorts - which to my knowledge has not been done
before - and developing a framework by which these results can be repeated and
expanded upon.

Several discoveries were made and analysed: insertion sort has one misprediction
per key; selection sort has $log_2(N/2)$ branch mispredictions per key - an
explanation of this was also provided; tiling mergesort can be done in two
steps, reducing the cache miss count; it is possible to avoid using a
sentinel and still remove the bounds check from a standard 2-heap, simply by
reordering the steps involved; sequential searches are far cheaper than binary
searches across small lists, due to a low branch misprediction rate.

In addition, several ideas were conceived although not implemented, which relate
to sorting algorithms: it should be possible to reduce the conflict misses of
tiled mergesort without padding, by slightly reducing the size of the array
segments sorted in the first sorting phase; it may be possible to reduce the
cost of an insertion sort or selection sort by considering more than one key at
a time; it may be possible to reduce the number of partitions - and hence the
increase in instruction count - of multi-quicksort by choosing from a larger
selection of pivots. Finally, it should be possible to reduce by half the number
of cache misses in radixsort, by counting earlier and by copying back and forth
between two arrays.

### 2005-05-09 docs/report.tex
moving fs into the proper directory. moving the predicate stuff to the predicate directory. deleting some random stuff

\documentclass[dvips]{report}
\usepackage{varioref}
\usepackage{listings}
\usepackage{color}
\usepackage{graphicx}
\usepackage{subfigure}
\usepackage{float}
\usepackage{lgrind}

\usepackage[intlimits]{amsmath}
\usepackage{eufrak}
\usepackage[mathscr]{euscript}
\usepackage{afterpage}

\newenvironment{changemargin}{%
 \begin{list}{}{%
\setlength{\textwidth}{\paperwidth}
\setlength{\textheight}{\paperheight}
%\setlength{\oddsidemargin}{-1in}
%\setlength{\evensidemargin}{-1in}
\setlength{\topmargin}{-1in}
  \setlength{\topsep}{-1in}%
  \setlength{\leftmargin}{-1.5in}%
  \setlength{\rightmargin}{-1.5in}%
  \setlength{\listparindent}{\parindent}%
  \setlength{\itemindent}{\parindent}%
  \setlength{\parsep}{\parskip}%
 }%
\centering%
\item[]%
}{\end{list}}

\newcommand{\cc}[1]{\texttt{#1}}
\newcommand{\n}[1]{\textit{#1}}
\newcommand{\code}[2]{\begin{figure}\lstinputlisting{code/#2}\caption{#1}\label{#1}\end{figure}}

\newcommand{\myscale}{1.2}
\newcommand{\plot}[7]{
% 3 figures first, then 3 page for them
\afterpage{
\setcounter{subfigure}{0}

\thispagestyle{empty}
\clearpage
\enlargethispage{14em}
\vspace*{-8em}
\begin{figure}[H]
\begin{changemargin}
\subfigure[Cycles per key - this was measured on a Pentium 4 using hardware performance counters. #2]
{\label{#1 cycles}\includegraphics[scale=\myscale]{plots/#1_-_cycles.eps}}
\subfigure[Instructions per key - this was simulated using SimpleScalar \cc{sim-cache}. #3]	
{\label{#1 instructions}\includegraphics[scale=\myscale]{plots/#1_-_Instructions.eps}}
\end{changemargin}
\vspace*{7em}
\caption{Simulated Instruction count and empiric cycle count for #1sort}
\label{Simulated Instruction count and empiric cycle count for #1sort}
\end{figure}
\newpage

\thispagestyle{empty}
\clearpage
\enlargethispage{14em}
\vspace*{-8em}
\begin{figure}[H]
\begin{changemargin}
\subfigure[Level 1 cache misses per key - this was simulated using SimpleScalar \cc{sim-cache}, simulating an 8KB data
cache with a 32 byte cache line and separate instruction cache. #4]
{\label{#1 level 1 misses}\includegraphics[scale=\myscale]{plots/#1_-_Level_1_misses.eps}}\\
\subfigure[Level 2 cache misses per key - this was simulated using SimpleScalar \cc{sim-cache}, simulating a 2MB data
and instruction cache and a 32 byte cache line. #5]
{\label{#1 level 2 misses}\includegraphics[scale=\myscale]{plots/#1_-_Level_2_misses.eps}}\\
\end{changemargin}
\vspace*{7em}
\caption{Cache Simulation results for #1sort}
\label{Cache Simulation results for #1sort}
\end{figure}
\newpage

\thispagestyle{empty}
\clearpage
\enlargethispage{14em}
\vspace*{-7em}
\begin{figure}[H]
\begin{changemargin}
\centering
\subfigure[Branches per key - this was simulated using \cc{sim-bpred}. #6]
{\label{#1 branches}\includegraphics[scale=\myscale]{plots/#1_-_Branches.eps}}
\subfigure[Branches misses per key - this was simulated using \cc{sim-bpred}, with bimodal and two-level adaptive
predictors. The simulated two-level adaptive predictor used a 10 bit branch history register which was XOR-ed with the
program counter. #7]
{\label{#1 branch misses}\includegraphics[scale=\myscale]{plots/#1_-_Branch_misses.eps}}
\end{changemargin}
\vspace*{6em}
\caption{Branch Simulation results for #1sort}
\label{Branch Simulation results for #1sort}
\end{figure}
}}

\newcommand{\smallscale}{.74}
\newcommand{\smallplot}[2]{
\setcounter{subfigure}{0}
\thispagestyle{empty}
\clearpage
\enlargethispage{14em}
\vspace*{-5em}
\begin{figure}[H]
\begin{changemargin}
\subfigure[Instructions per key] {\includegraphics[scale=0.74]{plots/#1_-_Instructions.eps}}\\
\subfigure[Level 1 cache misses per key]	{\includegraphics[scale=0.74]{plots/#1_-_Level_1_misses.eps}}
\subfigure[Level 2 cache misses per key]	{\includegraphics[scale=0.74]{plots/#1_-_Level_2_misses.eps}}
\subfigure[Branches per key]				{\includegraphics[scale=0.74]{plots/#1_-_Branches.eps}}
\subfigure[Branch misses per key]			{\includegraphics[scale=0.74]{plots/#1_-_Branch_misses.eps}}
\end{changemargin}
\vspace*{6em}
\caption{Simulation results for #2}
\label{Simulation results for #2}
\end{figure}
\newpage
}

\lstset{language=c,tabsize=3,basicstyle=\small,visiblespaces=false,visibletabs=false,stringspaces=false,labelstep=1,labelstyle=\small,frame=tblr}

\long\def\symbolfootnote[#1]#2{\begingroup\def\thefootnote{\fnsymbol{footnote}}\footnote[#1]{#2}\endgroup}

%\topmargin 0.2cm
%\textwidth 16cm
%\textheight 21cm
%\oddsidemargin 0.0cm
%\evensidemargin 0.0cm

\title{Fast Sorting on Modern Computers}
\author{Paul Biggar\\ B.A. (Mod.) Computer Science \\ Final Year Project April 2004 \\ Supervisor: Dr. David Gregg}
\date{}

\bibliographystyle{these}

\hyphenation{La-Marca}

% ----------------------------------------------------------------------------------
% END OF PREAMBLE
% ----------------------------------------------------------------------------------
\begin{document}

\begin{titlepage}
        \begin{center}
                \vspace*{95mm}
                \LARGE Fast Sorting on Modern Computers\\
                \vspace{1em}
                \large Paul Biggar\\
                \large B.A. (Mod.) Computer Science \\
                \large Final Year Project April 2004\\
                \large Supervisor: Dr. David Gregg\\
        \end{center}
\end{titlepage}

%\maketitle

\chapter*{Declaration}

I hereby declare that this thesis is entirely my own work and that it has not
been submitted as an exercise for a degree at any other university.

\begin{center}
\vspace*{2in}

\underline{\hspace*{3in}} \today

Paul Biggar
\end{center}

\chapter*{Acknowledgements}
I would like to thank my supervisor, Dr. David Gregg, for his help and advice
during this project. I'd also like to thank those who read my first draft, and
my friends and family for their support during the last year, and my entire time
at college. Thank you.
\newpage

\begin{abstract}
This project attempts to create and analyse sorts designed to take advantage of
features of modern processors. The work on cache-conscious sorting by LaMarca
and Ladner is the starting point; this is implemented and analysed using a
hardware simulator. Results relating to a variety of cache and branch predictor
configurations are obtained and analysed for a variety of sorts, beginning with
insertion and selection sorts, bubblesort and shakersort, and continuing to
heapsort, mergesort, quicksort and radixsort. Actual and expected results are
compared together and to the results of LaMarca's research.
\end{abstract}

\tableofcontents
\listoffigures

\setlength{\parskip}{1em}
\setlength{\parindent}{0ex}

\chapter{Introduction}

\input intro.tex

\chapter{Background Information}

\input background.tex

\chapter{Tools and Method}

\input method.tex

\chapter{Elementary Sorts}

\input ordernsquared.tex

\chapter{Heapsort}

\input heap.tex

\chapter{Mergesort}

\input merge.tex

\chapter{Quicksort}

\input quick.tex

\chapter{Radixsort}

\input radix.tex

\chapter{Conclusions}

\input conclusions.tex

%\appendix

%\chapter{Simulation Result Listing}
%\input result_list.tex

%\chapter{Code Listing}
%\input code_list.tex

%\chapter{Bug List}
%\input bugs.tex

%\chapter{SimpleScalar installation}
%\input simplescalar.tex

\bibliography{report} % param is the name of the .bib file
\nocite{*}

\end{document}

### 2005-05-09 base_quicksort.c

@@ -1,0 +1,204 @@
+#include <stdlib.h>
+#include <stdio.h>
+
+#include "base_sorts.h"
+
+#define THRESHHOLD 10
+
+static int* stack = NULL;
+static int stack_index = 0;
+
+static void
+stackinit(unsigned int N)
+{
+
+	int lg2N = 32; /* im hard coding this and i dont care */
+	unsigned int msbN = 1 << (lg2N - 1);
+	while(msbN > N)
+	{
+		lg2N--;
+		msbN = msbN >> 1;
+	}
+
+/*	printf("\nAllocating a stack %d ints big (lg2N == %d)\n", (lg2N + 2) << 2, lg2N); */
+	
+	stack = malloc(((lg2N + 2) << 2) * sizeof(int));
+}
+
+inline static void
+push(int valueA, int valueB)
+{
+	stack[stack_index++] = valueA;
+	stack[stack_index++] = valueB;
+}
+
+inline static int 
+pop()
+{
+	stack_index--;
+	return stack[stack_index];
+}
+
+inline static void
+stackclear()
+{
+	if (stack != NULL) free(stack);
+	stack = NULL;
+}
+
+inline static int
+stackempty()
+{
+	return (stack_index==0);
+}
+
+static struct predictor i_pred;
+static struct predictor j_pred;
+
+static int
+partition(unsigned int a[], int l, int r)
+{
+	unsigned int v;
+	int i, j;
+
+	v = a[r];
+	i = l - 1;
+	j = r;
+
+	for(;;)
+	{
+		while (less(a[++i], v))
+		{
+			branch_taken(&i_pred);
+		}
+		branch_not_taken(&i_pred);
+
+		while (less(v, a[--j]))
+		{
+			branch_taken(&j_pred);
+		}
+		branch_not_taken(&j_pred);
+		
+		if (i >= j)
+			break;
+
+		exch(a[i], a[j]);
+	}
+
+	exch(a[i], a[r]);
+
+	return i;
+}
+
+static void
+insertion_sentinel(unsigned int a[], int N)
+{
+	int i;
+	unsigned min;
+
+	min = 0;
+	for(i = 1; i < N; i++)
+	{
+		if(less(a[i], a[min]))
+		{
+			min = i;
+		}
+	}
+	exch(a[0], a[min]);
+
+}
+
+static void
[…]

### 2005-05-09 base_radixsort.c

@@ -1,0 +1,111 @@
+#include <stdlib.h>
+
+#include "base_sorts.h"
+#include "insertion.c"
+
+#define THRESHHOLD 10
+
+#define bitsword 32
+#define bitsbyte 8
+#define bytesword 4
+#define Radix (1 << bitsbyte)
+
+#define digit(A, B) (((A) >> (bitsword-((B)+1)*bitsbyte)) & (Radix-1))
+
+/* this gets the MSB */
+/*#define digit(A, B) (((A) >> (bitsword-((B)+1))) & 1) */
+
+static void
+quicksortB(unsigned int a[], int l, int r, int w)
+{
+	int i = l;
+	int j = r;
+
+	if (r <= l || w > bitsword) return;
+
+	while(j != i)
+	{
+		while(digit(a[i], w) == 0 && (i<j))
+			i++;
+		while(digit(a[j], w) == 1 && (j>i))
+			j--;
+		exch(a[i], a[j]);
+	}
+	if (digit(a[r], w) == 0) j++;
+	quicksortB(a, l, j-1, w+1);
+	quicksortB(a, j, r, w+1);
+}
+
+
+static unsigned int* radix_aux;
+
+#define bin(A) (l+count[(A)])
+static void
+radixMSD(unsigned int a[], int l, int r, int w)
+{
+/*	printf("\nl:%d, r=%d, W:%d\n", l, r, w); */
+	int i, j;
+	int count[Radix+1];
+
+	if (w > bytesword) return;
+	if (r-l <= THRESHHOLD)
+	{
+		insertion_sentinel(&a[l], r-l);
+		insertion(&a[l], r-l);
+		return;
+	}
+
+	for(j = 0 ; j < Radix; j++) count[j] = 0; /* clear count */
+	
+	for(i = l; i <= r; i++)
+		count[digit(a[i], w) + 1]++; /* count the items with these digits */
+
+	for(j = 1; j < Radix; j++)
+		count[j] += count[j-1]; /* cumulative */
+
+	for(i = l; i <= r; i++)
+		radix_aux[l+count[digit(a[i], w)]++] = a[i]; /* copy into correct places */
+
+	for(i = l; i <= r; i++) 
+		a[i] = radix_aux[i]; /* copy back */
+
+	radixMSD(a,l, bin(0)-1, w+1);
+
+	for(j = 0; j < Radix-1; j++)
+		radixMSD(a, bin(j), bin(j+1)-1, w+1);
+}
+
+static void
[…]

### 2005-05-09 base_sorts.h

@@ -1,0 +1,16 @@
+#ifndef _BASE_SORTS_H_
+#define _BASE_SORTS_H_
+
+#include "utils.h"
+
+void base_heapsort(unsigned int a[], int N);
+void base_quicksort(unsigned int a[], int N);
+void base_radixsort(unsigned int a[], int N);
+void base_mergesort(unsigned int a[], int N);
+
+void knuth_base_mergesort(unsigned int a[], int N);
+void knuth_other_base_mergesort(unsigned int a[], int N);
+
+void do_nothing(unsigned int a[], int N);
+
+#endif

### 2005-05-09 branch_sorts.h

@@ -1,0 +1,8 @@
+#ifndef _BRANCH_SORTS_H_
+#define _BRANCH_SORTS_H_
+
+#include "utils.h"
+
+void multi_quicksort_seq(unsigned int a[], int N);
+
+#endif

### 2005-05-09 cache_radixsort.c

@@ -1,0 +1,70 @@
+#include <stdlib.h>
+
+#include "base_sorts.h"
+#include "insertion.c"
+
+#define THRESHHOLD 10
+
+#define bitsword 32
+#define bitsbyte 8
+#define bytesword 4
+#define Radix (1 << bitsbyte)
+
+#define digit(A, B) (((A) >> (bitsword-((B)+1)*bitsbyte)) & (Radix-1))
+
+	int i, j, w, count[Radix+1]; /* R is the size of a digit, hence the number of bins */
+
+	for(w = bytesword - 1; w >= 0; w--) /* this is from lsb. 3 == lsByte, 0 == msByte */
+	{
+		for(j = 0; j < Radix; j++) count[j] = 0; /* set the counts to 0 */
+		for(i = l; i <= r; i++) /* count each digit */
+		{
+			count[digit(a[i], w) + 1]++;
+		}
+		for(j = 1; j < Radix; j++) /* make the count cumulatative */
+		{
+			count[j] += count[j-1];
+		}
+		for(i = l; i <= r; i++) /* when we look up the array, we now have its position */
+		{
+			radix_aux[count[digit(a[i], w)]++] = a[i];
+		}
+		for(i = l; i <= r; i++) /* copy it back */
+		{
+			a[i] = radix_aux[i];
+		}
+	}
+}
+
+
+void
+cache_radixsort(unsigned int a[], int N)
+{
+	unsigned int radix_aux = malloc(N * sizeof(unsigned int));
+	int r = N-1;
+	int l = 0;
+	int i, j, w, count[Radix+1]; /* R is the size of a digit, hence the number of bins */
+
+	for(w = bytesword - 1; w >= 0; w--) /* this is from lsb. 3 == lsByte, 0 == msByte */
+	{
+		for(j = 0; j < Radix; j++) count[j] = 0; /* set the counts to 0 */
+		for(i = l; i <= r; i++) /* count each digit */
+		{
+			count[digit(a[i], w) + 1]++;
+		}
+		for(j = 1; j < Radix; j++) /* make the count cumulatative */
+		{
+			count[j] += count[j-1];
+		}
+		for(i = l; i <= r; i++) /* when we look up the array, we now have its position */
+		{
+			radix_aux[count[digit(a[i], w)]++] = a[i];
+		}
+		for(i = l; i <= r; i++) /* copy it back */
+		{
+			a[i] = radix_aux[i];
+		}
+	}
+
+	free(radix_aux);
+}

### 2005-05-09 docs/code/base_heapsort.c

@@ -1,0 +1,33 @@
+void fixDown(Item a[], int parent, int N)
+{
+	Item v = a[parent];
+	while(2*parent <= N) /* check for children */
+	{
+		int child = 2*parent;
+
+		/* find larger child */
+		if (less(a[child], a[child+1])) child++; 
+
+		/* stop when not larger than parent */
+		if (!less(v, a[child])) break; 
+	
+		a[parent] = a[child];
+		parent = child;
+	}
+	a[parent] = v;
+}
+
+void base_heapsort(Item a[], int N)
+{
+	/* construct the heap */ 
+	for(int k=N/2; k >= 1; k--) fixDown(&a[-1], k, N);
+
+	/* destroy the heap for the sortdown */
+	while(N>1)
+	{
+		Item temp = a[0]; 
+		a[0] = a[--N]; 
+		fixDown(&a[-1], 1, N); 
+		a[N] = temp; 
+	}
+}

### 2005-05-09 docs/code/base_radixsort.c

@@ -1,0 +1,45 @@
+#define THRESHHOLD 10
+
+#define bitsword 32
+#define bitsbyte 8
+#define bytesword 4
+#define Radix (1 << bitsbyte)
+
+#define digit(A, B) (((A) >> (bitsword-((B)+1)*bitsbyte)) & (Radix-1))
+
+Item * radix_aux;
+
+void
+radixLSD(Item a[], int l, int r)
+{
+	/* R is the size of a digit, hence the number of bins */
+	int count[Radix+1]; 
+
+	/* this is from lsb. 3 == lsByte, 0 == msByte */
+	for(w = bytesword - 1; w >= 0; w--) 
+	{
+		/* set the counts to 0 */
+		for(int j = 0; j < Radix; j++) count[j] = 0; 
+		for(int i = l; i <= r; i++) /* count each digit */
+			count[digit(a[i], w) + 1]++;
+		
+		/* make the count cumulatative */
+		for(int j = 1; j < Radix; j++) 
+			count[j] += count[j-1];
+
+		/* when we look up the array, we now have its position */
+		for(int i = l; i <= r; i++) 
+			radix_aux[count[digit(a[i], w)]++] = a[i];
+		
+		for(int i = l; i <= r; i++) /* copy it back */
+			a[i] = radix_aux[i];
+		
+	}
+}
+
+void base_radixsort(Item a[], int N)
+{
+	radix_aux = malloc(N * sizeof(Item ));
+	radixLSD(a, 0, N-1);
+	free(radix_aux);
+}

### 2005-05-09 docs/code/main.c

@@ -1,0 +1,46 @@
+test_array(const char* description) {
+	if (memcmp(sorted_array, random_array2,
+				RANDOM_SIZE * sizeof(Item)) == 0)
+		printf("%s is sorted\n\n", description);
+	else {
+		printf("Not sorted:\n");
+		print_int_array(random_array2, RANDOM_SIZE,
+				description, -1, -1, -1);
+		printf("\n\nWhat it should be:\n");
+		print_int_array(sorted_array, RANDOM_SIZE,
+				"Sorted array", -1, -1, -1);
+	}
+}
+
+time_sort(void sort(Item*, int), char* description) {
+	memcpy(random_array2, random_array, 
+			RANDOM_SIZE * sizeof(int));
+	start_timer();
+	sort(random_array2, RANDOM_SIZE);
+	stop_timer();
+	print_timer(description);
+	test_array(description);
+}
+
+test_sort(void sort(Item*, int), char* description) {
+	for(int i = 8; i < 256; i++) {
+		int* a = malloc(i * sizeof(Item));
+		int* b = malloc(i * sizeof(Item));
+		fill_random_array(a, i, i);
+		memcpy(b, a, i * sizeof(Item));
+		qsort((void*)a, i, sizeof(Item), compare_ints);
+		sort(b, i);
+		if (memcmp(a, b, i * sizeof(Item)) != 0) {
+			printf("%s is not sorted (size == %d):\n", 
+					description, i);
+			print_int_array(b, i, description, -1, -1, -1);
+			printf("\n\nWhat it should be:\n");
+			print_int_array(a, i, "Sorted array", -1, -1, -1);
+			exit(-1);
+		}
+		free(a); free(b);
+	}
+	printf("Testing of %s complete\n", description);
+}
+test_sort(on2_insertsort, "O(N squared) Insertion");
+time_sort(on2_insertsort, "O(N squared) Insertion");

### 2005-05-09 docs/code/on2_shakersort.c

@@ -1,0 +1,16 @@
+void shakersort(Item a[], int N)
+{
+	int k = N-1;
+	for (int i = 0; i < k; )
+	{
+		// forwards
+		for (int j = k; j > i; j--)
+			if (less(a[j], a[j-1])) exch(a[j], a[j-1]);
+		i++;
+
+		// backwards
+		for (int j = i ; j < k; j++)
+			if (less(a[j+1], a[j])) exch(a[j+1], a[j]);
+		k--;
+	}
+}

### 2005-05-09 docs/code/simple_quicksort.c

@@ -1,0 +1,8 @@
+void quicksort(Item a[], int l, int r)
+{
+	if (r <= l) return;
+	int i = partition(a, l, r);
+	quicksort(a, l, i-1);
+	quicksort(a, i+1, r);
+}
+

### 2005-05-09 knuth_other_base_mergesort.c

@@ -1,0 +1,120 @@
+#include <stdlib.h>
+#include <string.h>
+#include <stdio.h>
+
+void knuth_other_base_mergesort(unsigned int a[], int N)
+{
+	unsigned int* aux = malloc(N * sizeof(unsigned int));
+	int s; /* s keeps track of the target, we can write something to remove this */
+	int i,j,k,l;
+	int d; /* direction counter */
+	
+	int temp;
+
+	int p, q, r; /* p keeps count of the number of items in this merge, q and r are counters for ascending and descending */
+	unsigned int* source;
+	unsigned int* target;
+
+/*S1:*/
+	s = 0; /* we should be able to elimnate s from here */
+	p = 1;
+
+S2:
+	i = 0;
+	j = N-1;
+	k = -1;
+	l = N;
+	d = 1;
+	q = p;
+	r = p;
+
+	if (s == 0)
+	{
+		source = a;
+		target = aux;
+	}
+	else
+	{
+		source = aux;
+		target = a;
+	}
+
+S3:
+	if (source[i] > source[j])
+		goto S8;
+
+/*S4:*/
+	k = k + d;
+	target[k] = source[i];
+
+/*S5:*/
+	i = i+1;
+	q = q-1;
+	if (q > 0)
+		goto S3;
+
+S6:
+	k = k+d;
+	if (k == l)
+		goto S13;
+
+	target[k] = source[j];
+
+/*S7: */
+	j = j-1;
+	r = r-1;
+	if (r>0)
+		goto S6;
+	
+	goto S12;
+
+S8:
+	k = k + d;
+	target[k] = source[j];
+
+/*S9:*/
+	j = j-1;
+	r = r-1;
+	if (r > 0)
+		goto S3;
+
+S10:
+	k = k+d;
+	if (k == l)
+		goto S13;
+
+	target[k] = source[i];
+
+/*S11: */
+	i = i+1;
+	q = q-1;
+	if(q > 0)
+		goto S10;
+
+S12:
+	q = p;
+	r = p;
+	d = -d;
+	temp = k; /* swap */
+	k = l;
+	l = temp;
+	if (j-i<p)
+		goto S10;
+
+	goto S3;
+
+S13:
+	p = p+p;
+	if (p < N)
+	{
+		s = 1-s;
+		goto S2;
+	}
+
+	if (s==0)
+	{
+		memcpy(a, target, N * sizeof(unsigned int)); 
+	}
+
+	free(aux);
+}

### 2005-05-09 main.c

@@ -1,0 +1,259 @@
+/***************************************************************************
+                          main.c  -  description
+                             -------------------
+    begin                : Sat Oct 11 2003
+    copyright            : (C) 2003 by Paul Biggar
+    email                : biggarp@tcd.ie
+ ***************************************************************************/
+
+#include <stdio.h> // fopen, fgets, fclose
+#include <errno.h> // errno
+#include <assert.h> // assert
+#include <string.h> // strerror
+#include <stdlib.h> // atoi, malloc, free
+#include <sys/types.h> // time
+#include <sys/time.h> // time
+
+#include "on2_sorts.h"
+#include "base_sorts.h"
+#include "branch_sorts.h"
+#include "cache_sorts.h"
+#include "visual_sorts.h"
+#include "utils.h"
+
+//#define RANDOM_SIZE (4194304)
+//#define RANDOM_SIZE (534523)
+#define RANDOM_SIZE (262144)
+//#define RANDOM_SIZE (16384)
+//#define RANDOM_SIZE (4096)
+//#define RANDOM_SIZE (128)
+
+#define RUN_VISUAL 1
+#define VISUAL_SIZE (128)
+
+unsigned int sorted_array[RANDOM_SIZE];
+unsigned int random_array[RANDOM_SIZE];
+unsigned int random_array2[RANDOM_SIZE];
+
+FILE* visual_log = NULL;
+
+static struct timeval t1;
+static struct timeval t2;
+
+
+static inline long
+get_milliseconds(struct timeval *tp1, struct timeval *tp2)
+{
+	return ( (tp2->tv_sec - tp1->tv_sec) * 1000000) + (tp2->tv_usec - tp1->tv_usec);
+}
+
+static void inline 
+start_timer()
+{
+	if (gettimeofday(&t1, NULL) != 0)  printf("Bad result: \"%s\"", strerror(errno));
+}
+
+static void inline
+stop_timer()
+{
+	if (gettimeofday(&t2, NULL) != 0) printf("Bad result: \"%s\"", strerror(errno));
+}
+
+static void inline
+print_timer(const char* description)
+{
+	printf("%s: %ld ticks\n", description, get_milliseconds(&t1, &t2));
+}
+
+static void inline
+test_array(const char* description)
[…]

### 2005-05-09 on2_bubblesort.c

@@ -1,0 +1,23 @@
+#include "on2_sorts.h"
+
+void
+on2_bubblesort(unsigned int a[], int N)
+{
+	int i, j;
+	int sorted;
+
+	for (i = 0; i < N-1; i++)
+	{
+		sorted = 1;
+		for (j = N-1; j > i; j--)
+		{
+			if (less(a[j],a[j-1]))
+			{
+				exch(a[j-1],a[j]);
+				sorted = 0;
+			}
+			
+		}
+		if (sorted) return;
+	}
+}

### 2005-05-09 paper/fastsort_perfex/base_heapsort.c

@@ -1,0 +1,146 @@
+#include "branch_sorts.h"
+
+#include <stdio.h>
+
+static void
+fixDown(unsigned int a[], int parent, int N)
+{
+	int child;
+	unsigned int v = a[parent];
+	while(2*parent <= N) /* see if it has any children */
+	{
+		child = 2*parent;
+
+		/* its important that this be less for my improvement */
+		if (less(a[child], a[child+1])) /* see if it has more than 1 child - check which child is larger */
+		{
+			child++;
+		}
+
+		if (!less(v, a[child])) /* when the larger child isnt large enough to be promoted, stop */
+		{
+			break;
+		}
+	
+		/* move down */ 
+		a[parent] = a[child];
+
+		parent = child;
+	}
+	
+	a[parent] = v;
+
+}
+
+/*
+static int
+is_a_heap(unsigned int a[], int l, int r)
+{
+	int N = r-l+1;
+	int k;
+	for(k = 1; k <= N/2; k++)
+	{
+		int k_pos = k - 1;
+		int l_pos = k * 2 - 1;
+		int r_pos = k * 2;
+		int l = l_pos + 1;
+		int r = r_pos + 1;
+
+		if (a[k_pos] < a[l_pos])
+		{
+			printf("FAIL: node %d (a[%d]=%d) is less than left child %d (a[%d]=%d)\n", k, k_pos, a[k_pos], l, l_pos, a[l_pos]);
+			return 0;
+		}
+		if (a[k_pos] < a[r_pos])
+		{
+			printf("FAIL: node %d (a[%d]=%d) is less than right child %d (a[%d]=%d)\n", k, k_pos, a[k_pos], r, r_pos, a[r_pos]);
+			return 0;
+		}
+	}
+	return 1;
+}*/
+
+
+void
+base_heapsort(unsigned int a[], int N)
+{
+	int k;
+
+	int is_even = !(N & 1);
+	if (is_even)
+	{
+		compexch(a[N-1], a[(N-1)/2]);
+		N--;
+	}
+
+	/* construct the heap */ 
+	for(k=N/2; k >= 1; k--)
+	{
+		fixDown(&a[-1], k, N); /* while pretending there are no parents with 1 child*/
+	}
+
+	if (is_even)
+	{
+		exch(a[0], a[N]); /* the final place */
[…]

### 2005-05-09 paper/fastsort_perfex/base_mergesort.c

@@ -1,0 +1,274 @@
+#include <stdlib.h>
+#include <string.h>
+#include <stdio.h>
+
+#include "base_sorts.h"
+
+void
+base_mergesort(unsigned int a[], int N)
+{
+	unsigned int* aux = malloc(N * sizeof(unsigned int)); /* make it twice the size to use the notation */
+
+	int i,j; /* indices for the first array */
+	int k; /* indices for second array */
+	
+	int d; /* d => direction, if (f == 0) keep going */
+
+	int next_count;
+
+	int track; /* this keeps track of i so we know where to merge next */
+	/* an alternative would be to add count/2 and clear the lsb. */
+
+	unsigned int* temp_pointer;
+	unsigned int temp;
+	int min;
+
+	unsigned int* source = a;
+	unsigned int* target = aux;
+
+/*	int remainder = (N & 7); */
+
+/*	print_int_array(a, N, "before pre-sorting", 0, -1, -1); */
+	/* pre-sort */
+	for(i = 0; i <= N-8; i+=8)
+	{
+		j = i+1;
+		if (less(a[i+2], a[j])) j = i+2;
+		if (less(a[i+3], a[j])) j = i+3;
+		compexch(a[i], a[j]);
+
+		j = i+2;
+		if (less(a[i+3], a[j])) j = i+3;
+		compexch(a[i+1], a[j]);
+
+		compexch(a[i+2], a[i+3]);
+
+		j = i+5;
+
+		if (less(a[j], a[i+6])) j = i+6;
+		if (less(a[j], a[i+7])) j = i+7;
+		compexch(a[j], a[i+4]);
+
+		j = i+6;
+		if (less(a[j], a[i+7])) j = i+7;
+		compexch(a[j], a[i+5]);
+
+		compexch(a[i+7], a[i+6]);
+	}
+
+	/* do a selection sort for the next 4, if they exist */
+	temp = i+4;
+	if (N < temp) temp = N;
+	for(; i < temp-1; i++)
+	{
+		min = i;
+		for(j = i+1; j < temp; j++)
+		{
+			if(less(a[j],a[min]))
+			{
+				min = j;
+			}
+		}
+		exch(a[i], a[min]);
+	}
+
+	/* do a reverse sort for the remaining ones, if they exist */
+	for(i = temp; i < N-1; i++)
+	{
+		min = i;
+		for(j = i+1; j < N; j++)
+		{
[…]

### 2005-05-09 paper/fastsort_perfex/base_quicksort.c

@@ -1,0 +1,186 @@
+#include <stdlib.h>
+#include <stdio.h>
+
+#include "base_sorts.h"
+
+#define THRESHHOLD 10
+
+static int* stack = NULL;
+static int stack_index = 0;
+
+static void
+stackinit(unsigned int N)
+{
+
+	int lg2N = 32; /* im hard coding this and i dont care */
+	unsigned int msbN = 1 << (lg2N - 1);
+	while(msbN > N)
+	{
+		lg2N--;
+		msbN = msbN >> 1;
+	}
+
+/*	printf("\nAllocating a stack %d ints big (lg2N == %d)\n", (lg2N + 2) << 2, lg2N); */
+	
+	stack = malloc(((lg2N + 2) << 2) * sizeof(int));
+}
+
+inline static void
+push(int valueA, int valueB)
+{
+	stack[stack_index++] = valueA;
+	stack[stack_index++] = valueB;
+}
+
+inline static int 
+pop()
+{
+	stack_index--;
+	return stack[stack_index];
+}
+
+inline static void
+stackclear()
+{
+	if (stack != NULL) free(stack);
+	stack = NULL;
+}
+
+inline static int
+stackempty()
+{
+	return (stack_index==0);
+}
+
+
+
+static int
+partition(unsigned int a[], int l, int r)
+{
+	unsigned int v;
+	int i, j;
+
+	v = a[r];
+	i = l - 1;
+	j = r;
+
+	for(;;)
+	{
+		while (less(a[++i], v))
+			;
+		while (less(v, a[--j]))
+			;
+		
+
+		if (i >= j)
+			break;
+
+		exch(a[i], a[j]);
+	}
+
+	exch(a[i], a[r]);
+
+	return i;
+}
+
+static void
+insertion_sentinel(unsigned int a[], int N)
+{
+	int i;
+	unsigned min;
+
+	min = 0;
+	for(i = 1; i < N; i++)
+	{
+		if(less(a[i], a[min]))
+		{
+			min = i;
+		}
+	}
+	exch(a[0], a[min]);
+
+}
+
+static void
+insertion(unsigned int a[], int N)
+{
+	int i;
+	unsigned int v;
+	for(i = 2; i < N; i++)
+	{
+		int j = i;
+		v = a[i];
+
+		while(less(v, a[j-1]))
+		{
+			a[j] = a[j-1];
+			j--;
+		}
+		a[j] = v;
+	}
[…]

### 2005-05-09 paper/fastsort_perfex/multi_quicksort_seq.c

@@ -1,0 +1,404 @@
+#include <stdlib.h>
+#include <stdio.h>
+#include <string.h>
+#include <limits.h>
+
+#include "cache_sorts.h"
+
+#define THRESHHOLD 10
+
+static int* stack = NULL;
+static int stack_index = 0;
+
+/* C is the number of keys that fit in the cache */
+#define C IN_PLACE_LIMIT
+#define C_DIV_3 (C / 3)
+static void
+stackinit(unsigned int N)
+{
+
+	int lg2N = 32; /* im hard coding this and i dont care */
+	unsigned int msbN = 1 << (lg2N - 1);
+	while(msbN > N)
+	{
+		lg2N--;
+		msbN = msbN >> 1;
+	}
+
+	if (N > 2 * C) /* make space for all the pivots being pushed */
+	{
+		lg2N +=  N / C_DIV_3;
+	}
+
+/*	printf("\nAllocating a stack %d ints big (lg2N == %d)\n", (lg2N + 2) << 2, lg2N); */
+
+	/* +2 for initial values, *2 cuase we push 2 ints a time */
+	/* we only really need about 3 less, cause of the threshhold, but this doesnt hurt */
+	stack = malloc(((lg2N + 2) * 2) * sizeof(int));
+}
+
+inline static void
+push(int valueA, int valueB)
+{
+	stack[stack_index++] = valueA;
+	stack[stack_index++] = valueB;
+}
+
+inline static int 
+pop()
+{
+	stack_index--;
+	return stack[stack_index];
+}
+
+inline static void
+stackclear()
+{
+	if (stack != NULL) free(stack);
+	stack = NULL;
+}
+
+inline static int
+stackempty()
+{
+	return (stack_index==0);
+}
+
+
+
+static int
+partition(unsigned int a[], int l, int r)
+{
+	unsigned int v;
+	int i, j;
+
+	v = a[r];
+	i = l - 1;
+	j = r;
+
+	for(;;)
+	{
+		while (less(a[++i], v))
+			;
+		while (less(v, a[--j]))
+			;
+		
+
+		if (i >= j)
+			break;
+
+		exch(a[i], a[j]);
+	}
+
+	exch(a[i], a[r]);
+
+	return i;
+}
+
+/*static void
+print_stack()
+{
+	printf("\nThe stack is: ");
+	for(int i = 0; i < stack_index; i++)
+	{
+		printf(" %d, ", stack[i]);
+	}
+}
+*/
+
+/* this is included so that we can replace the < N with <= N
[…]

### 2005-05-09 paper/fastsort_perfex/on2_sorts.h

@@ -1,0 +1,11 @@
+#ifndef _ON2_SORTS_H_
+#define _ON2_SORTS_H_
+
+#include "utils.h"
+
+void on2_insertsort(unsigned int a[], int N);
+void on2_selectsort(unsigned int a[], int N);
+void on2_bubblesort(unsigned int a[], int N);
+void on2_shakersort(unsigned int a[], int N);
+
+#endif

### 2005-05-09 simpleMain.c

@@ -1,0 +1,49 @@
+#include <stdlib.h>
+#include "base_sorts.h"
+#include "cache_sorts.h"
+#include "branch_sorts.h"
+#include "on2_sorts.h"
+
+#ifndef SORT
+#error "You must define the sort function like this: \"gcc -DSORT=base_insertion\""
+#endif
+
+#ifndef RANDOM_SIZE
+#error "You must define the size of the array like this: \"gcc -DRANDOM_SIZE=128\""
+#endif
+
+#ifndef SEED
+#error "You must define the sort function like this: \"gcc -DSEED=0\""
+#endif
+
+
+int main()
+{
+	int i;
+
+	/* if you segfault here, comment out _FASTSORT_STACK */
+	/* this happens because your system stack ownt take a 2 or 4meg stack variable */
+#define _FASTSORT_STACK
+#ifdef _FASTSORT_STACK
+	unsigned int random_array[RANDOM_SIZE];
+#else
+	unsigned int* random_array = malloc(RANDOM_SIZE * sizeof(unsigned int));
+#endif
+
+	srand(SEED);
+	for(i = 0; i < RANDOM_SIZE; i++)
+	{
+		random_array[i] = rand();
+	}
+
+
+	SORT(random_array, RANDOM_SIZE);
+
+#ifndef _FASTSORT_STACK
+	free(random_array);
+#endif
+	return 0;
+
+}
+
+

### 2005-05-09 utils.h

@@ -1,0 +1,63 @@
+#ifndef _UTILS_H_
+#define _UTILS_H_
+
+#include "paper/predictor.h"
+
+#ifndef NULL
+#define NULL 0
+#endif
+
+#define HEAP_SIZE 8
+
+#define ALIGNMENT (HEAP_SIZE * sizeof(unsigned int))
+
+#define LINE_BITS 5
+#define LINE_MASK 0x1F
+/* 32 byte line */ 
+
+#define BLOCK_BITS 16
+#define BLOCK_MASK 0xFFFF
+/* 65536 blocks */
+
+#define BLOCK_AND_LINE_BITS (BLOCK_BITS + LINE_BITS)
+#define BLOCK_AND_LINE_MASK ((BLOCK_MASK << LINE_BITS) | (LINE_MASK))
+
+/* LIMIT isnt the same as 2^(LINE_BITS+BLOCK_BITS)  */
+/* it is: sizeof cache / sizeof(unsigned long) / 2) */
+#define IN_PLACE_LIMIT 524288
+#define IN_PLACE_LIMIT_BITS (19)
+
+#define OUT_OF_PLACE_LIMIT 262144
+#define OUT_OF_PLACE_LIMIT_BITS (18)
+
+static inline unsigned int get_index(unsigned int* address)
+{
+	unsigned int a = (unsigned int)(address);
+	return (a >> LINE_BITS) & BLOCK_MASK;
+}
+
+static inline unsigned int get_offset(unsigned int* address)
+{
+	unsigned int a = (unsigned int)(address);
+	return a & LINE_MASK;
+}
+
+
+#define key(A) (A)
+#define less(A,B) (key(A) < key(B))
+#define exch(A, B) do { unsigned int t = (A); (A) = (B); (B) = t; } while(0)
+#define compexch(A, B) do { if (less((B), (A))) exch((A), (B)); } while(0)
+
+#define OUT(A) do { printf(#A ": 0x%x (%d)\n", A, A); } while(0)
+
+
+int check_sorted(unsigned int a[], int l, int r);
+
+void print_array(void* array, unsigned int count, unsigned int size, const char* description);
+void print_int_array(unsigned int array[], int N, const char* description, int left, int right, int pivot);
+
+void fill_random_array(unsigned int a[], int size, int seed);
+
+int compare_ints(const void* a, const void* b);
+
+#endif

### 2005-05-09 visual_sorts.h

@@ -1,0 +1,15 @@
+#ifndef _VISUAL_SORTS_H_
+#define _VISUAL_SORTS_H_
+
+#include <stdio.h>
+
+int open_visual_log(const char* file_name, const char* flags);
+void close_visual_log();
+
+void visual_base_mergesort(unsigned int a[], int N);
+void visual_tiled_mergesort(unsigned int a[], int N);
+void visual_base_heapsort(unsigned int a[], int l, int r);
+void visual_multi_mergesort(unsigned int a[], int N);
+
+void visual_knuth_base_mergesort(unsigned int a[], int N);
+#endif

### 2005-05-26 docs/demo.tex
make way for the new and improved fs

\documentclass[a4,landscape]{seminar}

\usepackage{semcolor}
\usepackage{graphicx}
\usepackage{color}
\usepackage{pst-all}

\definecolor{borderred}{rgb}{0.27,0.47,0.68}
\definecolor{darkblue}{rgb}{0.27,0.18,0.55}
\newenvironment{stitle}{\LARGE\color{black}}{}

\slideframe[\psset{fillstyle=gradient,gradmidpoint=0.25,gradbegin=white,gradend=gray,linecolor=white}]{scplain} %to show

%\slideframe{none} %to print

\renewcommand{\slidetopmargin}{-.2in}
\addtolength{\slidewidth}{2.4in}
\addtolength{\slideheight}{1.8in}

\newcommand{\plot}[2]{\begin{figure}\centering\includegraphics*[scale=#1]{plots/#2}\end{figure}}
\newcommand{\plotdouble}[3]{\begin{figure}\centering\includegraphics*[scale=#1]{plots/#2}\hspace{5pt}\includegraphics*[scale=#1]{plots/#3}\end{figure}}

\begin{document}
\centering
\sffamily
\input slide1.tex
\input slide2.tex
\input flowchart.tex
\input slide4.tex
\input slide5.tex
\input slide6.tex
\input slide7.tex
\input slide8.tex
\input slide3.tex

\end{document}

### 2005-05-26 docs/heap.tex
make way for the new and improved fs

%\section{Description}

Heapsort is an in-place sort which works in $O(NlogN)$ time in the average and
worst case. It begins by creating a structure known as a \n{heap}, which stores
its largest key at the top, then removing keys from the top and placing them in
their final position in the array.

\subsection{Implicit heaps}

A \n{binary tree} is a linked data structure where a node contains data and links
to two other nodes, which have the same structure. The first node is called the
\n{parent}, while the linked nodes are called \n{children}. One node is called the
\n{root}, and the tree branches out from this, with each level having twice as many
nodes as the previous level.

A heap is a type of binary tree with several important properties. Firstly, the
the data stored by the parent must always be greater than the data stored by its
children. Secondly, in a binary tree there is no requirement that the tree be full.
In a heap, it is important that every node's left child exists before its right
child. With this property, a heap can be stored in an array without any wasted
space, and without any links to other nodes. A heap done in the fashion is
called an \n{implicit heap}.

To take advantage of the heap structure, it must be possible to easily traverse
the heap, going between parent and child, without the presence of links. This
can be done due to the nature of the indices of parents and their children. In
the case of an array indexed between $0$ and $N-1$, then a 
parent whose index is $i$ has two children with indices of
$2i+1$ and $2i+2$. Conversely, the index of a parent whose child has an index
of $j$ is $\lfloor{}(j+1)/2-1\rfloor{}$. It is also possible to index the array between $1$
and $N$, in which case the children have indices of $2i$ and $2i+1$, while the parent
has an index of $\lfloor{}j/2\rfloor{}$. Using this scheme can reduces instruction cost, but
results is slightly more difficult pointer arithmetic, which cannot be performed
in some languages\footnote{In C this can be done by \cc{Item a[] =
\&array[-1];}}.

\subsection{Sorting with heaps}
To sort an array with a heap, it is necessary to first create the heap. This is
known as \n{heapifying}, and can be performed in two ways.

The first method involves \n{sifting}\footnote{Sifting up and down is also
referred to as fixing up and down, and algorithms that do this are called
\n{fix-up} or \n{sift-up}, and \n{fix-down} or \n{sift-down}} keys up the heap.
Assuming that a heap already exists, adding a key and re-heapifying will
maintain the heap property. This can be done by adding the key at the end of
the heap, then moving the key up the tree, swapping it with smaller parents
until it is in place. An unsorted array can be sorted by continually heapifying
larger lists, thereby implicitly adding one key every time to a sorted heap of
initial size one. This technique was invented by Williams, and LaMarca refers to
it as the \n{Repeated-Adds} algorithm.

The second technique is to heapify from the bottom up, by sifting keys down.
Beginning at half way through the array (this is the largest node with children),
a parent is compared to its two children and swapped down if it is smaller. This
continues upwards until the root of the heap is reached. This is the technique
specified by Floyd and LaMarca refers to this as Floyd's method.

The second step is to destroy the heap, leaving a sorted array.

\subsubsection{In-place sorting}
For an in-place sort, a heap is created as specified above. Then the root of the
heap (the largest key) is swapped with the last key in the heap, and the new
root is sifted down the heap which has been shrunk so as not to include the old
root, which is now in its final position. This is repeated for the entire heap,
leaving a fully sorted array.

Floyd suggested an improvement to this, where the new root is sifted straight
down without regard for whether it is larger than its child. Typically, it will
end near the bottom, so sifting up from here should save comparisons. Sedgewick
notes speed improvements only when comparisons are move expensive than swaps,
such as in the case of sorting strings. For integers, this step should not be
done.

\subsubsection{Out-of-place sorting}
Up until this point, the heap property used has been that a parent will be
larger than its child. If this property is change so that the parent is always
smaller than its child, then the result is an out-of-place sort. The smallest
key is removed from the heap, and placed in its final place in the array. The
heap is then re-heapified. LaMarca refers to the steps involved as
\n{Remove-min}.

\section{Base heapsort}
LaMarca recommends to use of the Repeated-Adds algorithm for reducing
cache misses. As such, his base heapsort uses Floyd's method, ostensibly at
the behest of major textbooks, each of whom claim Floyd's method gives better
results. This is certainly backed up by Sedgewick, who recommends this method,
but in initial tests, the out-of-place Repeated-Adds technique had a lower
instruction count. Regardless, the base algorithm used here is Floyd's in-place
sort. Code for a base heapsort is in Figure \ref{Simple Base Heapsort}.

\code{Simple Base Heapsort}{base_heapsort.c}

A difficulty of heapsort is the need to do a bounds check every time. In a heap,
every parent must have either no children, or two children. The only exception
to this is the last parent. It is possible that this has only one child, and it
is necessary to check for this as a result. This occurs every two heapifies, as
one key is removed from the heap every step. Therefore it is not possible to just
pretend the heap is slightly smaller, and sift up the final key later, as it
might be with a staticly sized heap.

LaMarca overcomes the problem by inserting a maximal key at the end of a heap if
there's an even number of nodes in the array. This gives a second child to the
last element of the heap. This is an effective technique for building the heap,
as the number of keys is static. When destroying the heap, however, the
maximal key must be added to every second operation. This can be avoided
relatively simply with no extra instructions. If the last key in the heap is
left in the heap while its being sifted down the heap, it will provide a
sentinel for the sift down operation. The key from the top of the heap is then
put into place at the end. This technique reduces instruction count by
approximately 10\%.

The only other optimisation used is to unroll the loop while destroying the
heap. This has minimal affect, however.

\section{Cache Heapsort}

LaMarca notes that the Repeated-Adds algorithm has a much lower number of cache
misses than Floyd's. The reason for this is simple. Consider the number of keys 
touched when a new key is added to the heap. In the case of Floyd's method, the
keys being heapified are not related to the last keys that were heapified. Reuse
only occurs upon reaching a much higher level of the heap. For large heaps, these
keys may be swapped out of the cache by this time. As a result, Floyd's method
has poor temporal locality.

The Repeated-Adds algorithm, however, displays excellent temporal locality. A
key has a 50\% chance of having same parent, a 75\% of the same grandparent and
so on. The number of instructions, however, increases significantly, and the
sort becomes out-of-place.

The cost of building the heap is completely dominated by the cost of
destroying it. Less than 10\% of heapsort's time is spent building the heap.
Therefore optimisations must be applied when destroying the heap.

LaMarca's first step is to increase spatial locality. When sifting down the
heap, it is necessary to compare the parent with both children. Therefore it is
important that both children reside in the same cache block. If a heap starts at
the start of a cache block, then the sibling of the last child in the cache
block will be in the next cache block.  For a cache block fitting four keys,
this means that half the children will cross cache block boundaries. This can be
avoided by padding the start of the heap. Since the sort is out of place, an
extra cache block can be allocated, and the root of the heap is placed at the end of
the first block. Now, all siblings will reside in the same block.

\begin{figure}[h]
\centering
\subfigure[a normal heap] {\includegraphics[clip=true]{heap_cache_line1}}
\subfigure[a padded heap] {\includegraphics[clip=true]{heap_cache_line2}}
\end{figure}

A second point about heapsort's spatial locality is that when a block is loaded,
only one set of children will be used though two or four sets may be loaded. The
solution is to use a heap higher \n{fanout}, that is, a larger number of
children.  The heap used so far could be described as a 2-heap, since each
parent has two children.  LaMarca proposes using a \n{d-heap}, where $d$ is the
number of keys which fit in a cache line. In this manner, every time a block is
loaded, it is fully used. The number of cache blocks loaded is also decreased,
since the height of the tree is halved by this.

This step changes the indexing formula for the heap. In this method, a parent
$i$ has $d$ children, indexed from $(i*d) + 1$ to $(i*d) + d$. The parent of a
child $j$ is indexed by $\lfloor{}(i-1)/d\rfloor{}$.

The height of a 2-heap is $\lceil{}log_2(N)\rceil{}$, where \textit{N} is the
number of keys in the heap. By changing to a 4-heap, the height is reduced to
$\lceil{}log_4(N)\rceil{} = \lceil{}\tfrac{1}{2}log_2(N)\rceil{}$. The height of
an 8-heap is half again the height of a 4-heap. However, the number of
comparisons needed in each level is doubled, since all the children will be
unsorted. LaMarca claims that as long as the size of the fanout is less than
12, a reduced instruction count will be observed.

In the base heapsort, an improvement was added that removed the need for a
sentinel. Since the 4- and 8-heaps were unsorted, this optimisation had to be
removed. Leaving the key being copied in place was still possible, but it needs
to be replaced by a maximal key before trying to sift the next key down the
heap, lest it were chosen to be sifted up. Placing the sentinel is still cheaper
than the bounds check, however, as the bounds check will need to be performed
$NlogN$ times, compared to $N$ times for the sentinel. The cost of the sentinel
write is more expensive but the extra $logN$ instructions saved will mean that
this increase will be an order of magnitude lower than the bounds check.

\section{Results}
All simulations in this project are performed on 32 byte cache lines. As a
result, 8-heaps are used for heapsort, since 8 4-byte \n{int}s fit the cache line.
LaMarca used 64 bit ints, and while this doesn't affect results, it means that
fewer keys fit into his cache lines. As a result, he used a 4-heap. To try and
emulate his results, extra tests were done, using a 4-heap with a 32 byte
cache line, and using a 4-heap using a 16 byte cache line, so that the entire
cache line is used. \verb!UINT_MAX! was used as the sentinel.

\subsection{Expected Results}
The instruction count of heapsort is $O(Nlog_d(N))$, although a higher $d$ means
that each step will take longer. As the fanout of the heap doubles, then the
height of the heap halves, but each heap step is more expensive. It is expected
though, that increasing $d$ will reduce the instruction count until the point at
which the cost of a heap step doubles, as compared to a heap with a smaller
fanout.

An additional consideration is that the base heapsort has more instructions than
it's out-of-place equivalent. This means that the 4-heap will have a better
instruction count, as well as better cache performance than the base heapsort.
The 8-heap should have the best cache performance of them all, due to fully
using each cache line, taking advantage of the spatial locality of the cache
line. However, the 4-heap should perform just as well with a 16 byte cache line.

It is expected that the number of comparison branches stays the same with
an increased fanout. When the height of the heap halves, the number of
comparisons in each step doubles. However, the proportion of those which are
predictable changes. The key being sifted down the heap is likely to go the
entire way down the tree. Therefore the comparison between the smallest child
and the parent is predictable. However, the comparisons to decide which child is
smallest are not predictable. In a 2-heap, there is one comparison between
children, meaning that half the branches are predictable. An 8-heap has seven
comparisons between children, with the result that only an eighth of the branches
are predictable.

However, the creation of the heap will have fewer misses with a larger fanout,
due to the reduced height. In practice though, the implications of this are
minimal. Most of the time is spent in the destruction of the heap, and this will
have far more branches than the creation phase.

\subsection{Simulation Results}

\plot{heap}{}{}{Cache heapsort (4-heap) was also tested with a 16 byte cache line}{Cache heapsort (4-heap) was also tested with a 16 byte cache line}{}{}

As expected, the 4-heap performs better than the base heapsort. However, the
increased number of comparisons of the 8-heap increases the instruction count to
even higher than base heapsort, despite the reduction in the height of the heap.
This is shown in Figure \ref{heap instructions}.

The level 1 cache results, shown in Figure \ref{heap level 1 misses}, show the
difference between the two types of heap. The slope of the base heapsort result
is much greater than all the others, with the 8-heap having the fewest misses.
This continues into the level 2 cache, where the flattest slope belongs to the
8-heap. The 4-heap with the 32-byte line has fewer misses than the 16-byte
version, which is surprising since the 16-byte version can fit twice as many
blocks in the cache, and uses each block fully.  This may be due to the extra
compulsory misses due to having twice as many blocks to load. The difference
between fully associative and direct mapped is only minor.

In the graph, the base heapsort results flat-line for a step longer than the
other heapsorts' results. This is due to being in-place, since twice as many
keys fit in the cache without a corresponding increase in misses. For the same
reason, the base heapsorts have half as many level 2 cache misses of the later,
out-of-place sorts (See Figure \ref{heap level 2 misses}).

The branch prediction behaves similarly to the expected results, in that the
prediction ratio gets worse as the fanout gets larger. However, the difference
is minor between the 4-heap and 8-heap, and the difference for larger fanout
would be less still. The difference between the 2-heap and 4-heap result is
interesting though, the 2-heap has more branches, but less misses per key.

The difference between different types of predictors is also visible in the
results, shown in Figure \ref{heap branch misses}. The largest table size is
shown for the two-level predictor. The smallest table size is shown for the
bimodal predictor, as all the bimodal table sizes show identical results.  In
the case of the 2- and 8-heaps, the bimodal predictor works better than the
two-level adaptive predictor. The warm-up time is visible in the base heapsort,
though it's amortised for larger data sets. There is no difference between
predictors for the 4-heap.

Figure \ref{heap cycles} shows the actual time taken for heap sort on a Pentium
4 1.8GHz machine, which has an 8-way associative 256KB level 2 cache with a 64 byte
cache line, a 4-way associative 8KB level 1 data cache, and a maximum branch
misprediction delay of 20 cycles.\footnote{This is determined from the Pentium
4's 20 stage pipeline. This number is not exact, and is probably slightly lower,
with the possibility of occasionally rising higher (in the case of an
instruction cache miss, which is exceptionally unlikely for small routines like
sorts).} Its static branch prediction strategy is backward taken and forward
not taken; its dynamic strategy is unspecified: though \cite{Intel248966-010}
mentions a branch history so it is likely a two-level adaptive predictor.

The figure shows that the base heapsort performs best while the data set fits in
the cache. After that point, the 8-heap performs best, and its cycle count increases
with a smaller slope. This shows the importance of LaMarca's cache-conscious
design even eight years after his thesis. It also shows the importance of reduced
branch mispredictions: the base heapsort has more instructions and level 1 cache
misses, but still outperforms the 4-heap, before level 2 cache misses take over.
However, this is hardly conclusive: the lower level 2 cache misses may account
for this.

The similarities to LaMarca's results are startling. The shape of the
instruction and cache miss curves are identical, as are the ratios between the
base and memory-tuned versions in the case of level 1 and 2 cache misses,
instruction count and execution time.

Slight differences exist though. The level 2 cache miss results here indicate
that while still sorting in the cache, the number of misses in the in-place
heapsort is less than that of the out-of-place sorts. LaMarca, however, has the
same amount of misses in both cases. Similarly, the results here indicate that
base heapsort experiences a sharp increase in cache misses. This increase comes
when the set size is twice that of cache mergesort when its sharp increase
occurs. LaMarca's results do not show this.

Finally, though this is likely to be related the previous differences, base
heapsort performs slightly better for small sets than the cache heapsorts do in
the results presented here. LaMarca's results indicate that the cache heapsort
has better results the entire way through the cache. This may relate to the
associativity: LaMarca's tests were performed on a DEC AlphaStation 250 with a
direct mapped cache. The Pentium 4's cache is 8-way associative, and so the
unoptimised algorithm would not suffer as much.

\section{Future Work}

LaMarca later published heapsort code in \cite{LaMarcaHeap96}. The ACM
Journal of Experimental Algorithms requires that source code be published, and
comparing simulated results from there to the sorts created here might have
interesting results.

Comparing the 4-heap with a 16 byte cache line to the 4-heap with a 32 byte
cache line is not as interesting as it would be to compare it to an 8-heap on a
16 byte cache line.

Also, it would be interesting to use performance counters to find out whether
the cache or the branch predictors are responsible for the base heapsort being
faster than the 4-heap, on the Pentium 4.

### 2005-05-26 docs/merge.tex
make way for the new and improved fs

%\section{Description}
\label{Mergesort}
Merging is a technique which takes two sorted lists, and combines them into a
single sorted list. Starting at the smallest key on both lists, it writes the
smaller key to a third list. It then compares the smallest key on one list with
the next smallest on the other, and writes the smaller to the auxiliary list.
The merging continues until both lists are combined into a single sorted list.

Mergesort works by treating an unsorted array of length $N$ as $N$ sorted arrays of
length one. It then merges the first two arrays into one array of length two, and
likewise for all other pairs. These arrays are then merged with their neighbours
into arrays of length four, and so on, until the arrays are completely sorted.

Mergesort is an out-of-place sort, and uses an auxiliary array of the same size
for the merging. Since a lot of steps are required, it is more efficient to
merge back and forth - first using the initial array as the source and the
auxiliary array as the target, then using the auxiliary array as the source and
the initial array as the target - rather than merging to the auxiliary array and
copying back before the next merge step.

The number of merges required is $log_2(N)$, and if this number is even then
the final merge puts the array in the correct place. Otherwise, the result will
need to be copied back into the initial array.

\section{Base Mergesort}

\subsection{Algorithm N}
\label{Algorithm N}

For his base algorithm, LaMarca takes Knuth's algorithm N \cite{Knuth98}.
LaMarca describes this algorithm as functioning as described above; lists of
length one are merged into lists of length two, then four, eight, and so on.
However, algorithm N does not work in this fashion. Instead, it works as a
\n{natural} merge. Rather than each list having a fixed length at a certain
point, lists are merged according to the initial ordering. If it happens that
there exists an ascending list on one side, this is exploited by the algorithm.
While this results in good performance, the merge is not perfectly regular, and
it is difficult or impossible to perform several of the optimisations LaMarca
describes on it.

\subsection{Algorithm S}

Knuth's algorithm S is a \n{straight} merge, and works as described above. It has
15\% lower instruction count than Algorithm N\footnote{This is observed on
random data, which algorithm N is not designed for. On more regular data, or
semi-sorted data, this will probably not occur.}. Algorithm S, like algorithm N,
is a very difficult algorithm to modify. Knuth describes it in assembly, and
making the optimisations that LaMarca describes on it is more difficult than
rewriting the algorithm more simply. The rewritten algorithm, referred to from
here on as `base mergesort', is a simple higher level version of algorithm S,
and performs only slightly more slowly.

A disadvantage of this method is that if an array has slightly more than $2^x$
keys, an entire extra merge step would be required, which would considerably
slow the algorithm. This problem is shared by all other versions of mergesort
considered here, except algorithm N. However, since this project uses exact
powers of two as set sizes, this problem does not affect the results. LaMarca,
however, used set sizes such as 4096000, which at times would suffer from this
type of problem.

\subsection{Base Mergesort}
Three optimisations are described by LaMarca for the base mergesort. Firstly, a
bitonic sort, as described in \cite{Sedgewick02}, removes instructions from the
inner loop. Every second array is created in reverse, so that the larger keys  
are next to each other in the middle. The smallest keys are at the outside,
meaning that the lists are iterated through from the outside in. As a result, it
is not possible to go outside the array, and it is not necessary to consider the
lengths of the arrays, apart from when setting the initial array indices. When
the first array is exhausted, it will point to the last key on the other array,
which will be larger than all other keys in that array. When both are
exhausted, the last key of both arrays will be equal. A check is put in, in this
case, and merging ends if both indices are the same. Figure \vref{Visual Sort}
shows a bitonic merge.

The overhead of preparing arrays to be merged is small. However, for small
lists, the overhead is large enough that it can be more efficient to sort using
a much simpler sort, such as insertion, selection or bubblesort. In Section
\ref{insertion is predictable}, it is shown that insertion sort is the fastest
elementary sort.  Sedgewick's quicksort uses this, and LaMarca recommends the
use of a simple inline pre-sort, which can be made from insertion sort. To make
lists of length four, the smallest of the four keys is found, and exchanged
with the first key, in order to act as a sentinel. The next three keys are
sorted with insertion sort. The next set of four keys is sorted in reverse
using the same method. It is possible to hard code this behaviour, in an attempt
to reduce the instruction count, but in fact the instruction count stays the
same, and the branch prediction suffers as a result.

Finally, the inner loop is \n{unrolled}\footnote{Unrolling a loop means
executing several steps of the loop between iterations, reducing the number of
iterations, and lowering the loop overhead, as well as allowing for better
scheduling of the instructions.}.  There are actually two inner loops, and they
are both unrolled 8 times. The compiler is also set to unroll loops. However,
the observed difference between unrolling manually and compiler unrolling is up
to 10\% of instruction count, so manual unrolling is used.

Overall, these three optimisations reduced the instruction count by 30\%.

\section{Tiled Mergesort}
LaMarca's first problem with mergesort is the placement of the arrays with
respect to each other. If the source and target arrays map to the same cache
block, the writing from one to the other will cause a conflict miss. The next
read will cause another conflict miss, and will continue in this manner. To
solve this problem, the two arrays are positioned so that they map to different
parts of the cache. To do this, an additional amount of memory the same size as
the cache is allocated, which can be expensive for small arrays. The address of
the auxiliary array is then offset such that $index(source) + index(auxiliary) =
C$.

While this step makes sense, whether this is valuable is very much a platform
dependant issue. Since virtual addresses are used, it may be possible that the
addresses used by the cache are not aligned as expected. Even worse, an
operating system with an excellent memory management system may do a similar
step itself. The program's attempt to reduce cache misses could then increase
them. In the event that the \n{Memory Management Unit} in the processor uses
virtual addresses to index its cache, or has a direct mapping between virtual
and physical or bus addresses, then this problem is very unlikely. The
simulations presented in this project address the cache by virtual address, and
so this optimisation is effective in this case.

The next problem considered is that of temporal locality. Once a key is loaded
into the array, it's only used once per merge step. If the data does not fit in
half the cache, then conflict misses will occur, and no temporal reuse will
occur at all. The solution is to apply a technique used in compilers called
\n{tiling}. Using this technique, the number of keys which fit in the cache will
be fully sorted before moving on to the next set. When all the sets have been
sorted in the manner, they are all merged together as usual.

\section{Multi-mergesort}

Multi-mergesort has two phases. The first is the same as tiled mergesort, and
sorts the array into lists sized $C/2$. The second phase attempts to improve
temporal locality by merging these lists together in a single merge step. There
are $k$ lists and this is called a \n{k-way merge}. It is possible to do a
search across each of these keys, maintaining a list of indices to the smallest
unused key from each list. When there is a small number of lists, this isn't a
problem, however, when the lists grow big, the overhead involved in this is
substantial. It is necessary to check each of the lists, and each key searched
would cause a cache miss, since each array segment is aligned, and map to
the same cache block.

To avoid this, a priority queue is used. The smallest key from each list is
added to the queue, and then put into the final target array. This reduces the
problem of the search overhead, but does not reduce the problem of
\n{thrashing}\footnote{Thrashing occurs when two arrays are mapped to the same
cache block, and the constant access of one followed by the other causes every
access to be a miss.}.  To avoid this, an entire cache line is added at a time
to the priority queue. There will still be conflict misses, but there will be no
thrashing since key overwritten in the cache will not be needed again.

LaMarca used a 4-heap for his priority queue, and an 8-heap is used here, for
the same reasons as before. The heap is aligned, and a cache line
is put into the array each time. The smallest key is at the top of the heap,
and this is put into its final position in the array. The heap is then fixed.

LaMarca uses a special tag to mark the last key from each list inserted into
the array. No more details are provided on this, and a sorted list of length $k$
is used here, in its place.  If the key at the top of the heap is the same as
the smallest value in the list, another eight keys are added from the array
segment to which the number belonged. This continues until the array is full
sorted.

Using a sorted list instead of LaMarca's tagging leads to a variety of problems.
The reason the sorted list is used is that no details of the tagging is
provided. The sorted list is $k$ keys long, and only needs to be properly sorted
once. The next time it needs to be sorted is when a new line is added, but in
this case only one key is out of place, and can be put in place with a single
insertion sort run. This is linear at worst, and has excellent branch prediction
properties, as shown earlier.

When a list is fully consumed, the next keys inserted will be from the next
list. This could be fixed with a bounds check, however, since the lists are
bitonic, the next keys inserted are going to be the largest keys from the next
array. These will be added to the heap, and take their place at the bottom,
without harmful effect. A harmful effect is possible, though, when lots of keys 
with the same value are in the array. All these keys would be inserted at a
similar time, and when the first one is taken off the heap, a list would add
eight keys, even though it had just added eight. In this manner, the size of the
heap could be overrun. With random keys in the array, this is unlikely, and so
this is not addressed.

The heap used for the k-way merge should result in a large increase in
instructions, and a significant reduction in level 2 cache misses. The addition
of the heap should not result in an increase in misses, since the heap is small
and will only conflict with the merge arrays for a very short time. The heap is
aligned as described in the previous chapter, and has excellent cache
properties. LaMarca estimates that there should be one miss per cache line
in the first phase, and the same in the second phase, although each miss will
occur in both the source and target arrays. This sums to four misses per cache
block, for a total of one miss for every two keys in the array.

\section{Double Tiled Mergesort \\/ Double Multi-mergesort}

While writing the cache mergesorts, I realized several opportunities for
improvement. The first of these is avoiding the copy back to memory. In the
event of an odd number of merge steps, the auxiliary array will contain the
sorted array, which needs to be copied back. This can be avoided by pre-sorting
to make lists sized two or eight. A large increase of instruction count was realized by
sorting to lists of length two, so a length of eight was used. When an even number
was required, four was used.

The second improvement was to fully sort keys in the level 1 cache, before
continuing to level 2 cache as in the tiled mergesort. The expected improvement
is not as significant as in the case of the level 2 cache, since is the penalty
for level 1 misses is not great.

Finally, the merge is rewritten in a more straightforward and readable manner. This
was necessary due to the increased complexity of debugging the double mergesorts.
During the process, different functions were used to process lists that were
exactly the required size and lists which were not, which needed code to handle
edge-cases. However, the instruction count improvement was in the double digits,
and so these changes were removed in order to have lower instruction cache
misses, as well as to reduce the fill time of the branch predictors and reduce
the complexity and size of the algorithm.

\section{Results}
\subsection{Test Parameters}
The level 1 cache is 8, so 2048 keys can be sorted in the level 1 cache.
The number of keys to be sorted in the level 2 cache is 262144.  When arrays
were pre-sorted, they were pre-sorted to a size four, or eight if the number of passes
was going to be odd. These were reversed for the double multi-mergesort; since
it does a k-way merge into the source array, the sorted cache sized segments
need to be in the auxiliary array. An 8-heap was used for the multi-mergesorts.

\subsection{Expected Performance}

Algorithm N and algorithm S are included in the results for reference, but are
not discussed here.

The instruction count of mergesort is $O(NlogN)$, and does not vary
according to input. Multi-mergesort should have a significantly higher
instruction count, once its threshold is hit, as should its variant, double
multi-mergesort.

The number of memory references is also $O(NlogN)$, and mergesort's cache
performance should be very regular, so long as the data set fits in the cache. Once
the data set exceeds the cache, the number of cache misses should leap. For
multi-mergesort, they should flatten out after this, perhaps with a slight
slope to take into account the conflict misses of the aligned array segments. The
fully associative caches should not have these misses, and the results should be
correspondingly lower. Due to their design, the double mergesorts should have
better performance than their counterparts.

The flow control of mergesort is very regular, but there is no way to predict
the comparisons. As a result the misprediction rate should be $O(NlogN)$ for
all but the multi-mergesorts, which should increase once the threshold is
passed.

\subsection{Simulation Results}
\label{Mergesort results}

\plot{merge}{}{}{Results shown are for a Direct Mapped cache.}{Results shown are
for a Direct Mapped cache.}{}{}

The number of instructions, shown in Figure \ref{merge instructions} is exactly
as predicted, with a 50\% spike at multi-mergesort. Level 2 misses, shown in
\ref{merge level 2 misses} are also as expected, although the multi-mergesort
does not show the same improvement over tiled mergesort as in LaMarca's results.
Double multi-mergesort does show an improvement, but not a great improvement to
double tiled mergesort. It does, however, have a much flatter slope.

The prediction results in Figure \ref{Branch Simulation results for mergesort}
are as expected. Very little variation exists between the types of mergesort. It
is noticeable, though, that the spike in branch misses due to the
multi-mergesorts is considerably lessened by the two level adaptive predictor.
The reason for this is not clear.

Despite the improvement in cache misses due to multi-mergesort, it still cannot
account for the increase in instruction count as a result, and it has a higher
cycle count on the Pentium 4 (see Figure \ref{merge cycles}) as a result. The
other sorts generally perform in $O(NlogN)$ time, with double tiled mergesort
being the fastest. Note that the alignment of the mergesort may not work as
desired due to the size of the cache. This was optimised for a 2MB cache, while
the Pentium 4 cache is only 256KB. This means that the two arrays are aligned to
the same cache block. However, since the cache is 8-way associative, this is not
serious, and should not result in a lot of thrashing.

The results presented here differ from LaMarca's results. In the case of
instruction count, both sets are nearly identical: the graphs are the same shape
and have similar ratios between the base and multi-mergesorts. The cache results
have a similar shape to LaMarca's, with a large increase in misses as soon as
the arrays no longer fit in the cache, and similar ratios between the base and
tiled versions.  But while LaMarca reports a very small increase in level 2
misses once multi-mergesort's data set is larger than the cache, the results
here show nearly the same jump as tiled mergesort. This results in half as many
misses as the base mergesort, while LaMarca shows nearly 10 times fewer misses
for his multi-mergesort.

As a result of this discrepancy, it is not a surprise that LaMarca's results
indicate that multi-mergesort is far faster than base-mergesort, but the same
results are not shown here. The increase in instruction count cripples
multi-mergesort, and tiled mergesort emerges the victor.

This is not to say that LaMarca's results disagree with those here. In fact, it
is noticeable that multi-mergesort's cache misses plateau at the edge of the
graph, while other mergesorts' misses continue to climb. This is in keeping with
both LaMarca's theory and results.

\section{Future work}
Several avenues could be pursued in order to improve the algorithms presented
here. In multi-mergesort, LaMarca spoke of tagging keys as they went into the
cache. Here, a sorted list was used in its place. Implementing and investigating
tagging and comparing of the results of the two techniques would be interesting.

The difference between the actual and expected results of the tiled and
multi-mergesorts is another area to be investigated. While the results point to
similar conclusions about cache misses between the base, tiled and
multi-mergesorts, the relation these cache misses for direct mapped caches is
not expected. Investigating, and possibly fixing this would be interesting.

Algorithm N is a mergesort for more natural layouts of data, such as partially
sorted lists. It can sort these faster than a straight merge, but won't on
random data until it is nearly sorted. Some of the improvements described here
may be able to improve its performance, as they did for the straight merge.

A bug exists in the double tiled and multi-mergesorts, which causes an extra
merge step in some cases. More accurate results could be achieved by solving
this problem. Also, to take advantage of the level 1 cache, the double sorts were
sorted to 2048 keys first, then merged into larger sections. This number should
have been half that. See section \ref{bug4} for more details.

To address one of the problems with multi-mergesort, that of the conflict misses
on the k-way merge, \cite{Xiao00} reveals a padded multi-mergesort with reduced 
conflict misses. It would be interesting to compare these results.

It may also be possible to simply reduce the number of conflict misses of the
k-way merge, simply by reducing the size of the initial sort by the size of several cache
lines. In this way, the smallest keys would map to different cache blocks,
without the need for padding.

### 2005-05-26 docs/ordernsquared.tex
make way for the new and improved fs

Elementary sorts are also referred to as $O(N^2)$ sorts. This is due to the
manner in which they check every key in the array in order to sort a single key;
for every key to be sorted, every other key is checked.  Four of these sorts are
discussed here. They are insertion sort, selection sort, bubblesort and
shakersort.

\section{Selection sort}

Selection sort works in a straightforward manner. It begins by 
searching the unsorted array for the smallest key, then swapping it into the
first position in the array. This key is now the first element of a sorted
array, being built from the left. The remainder of the array is
still unsorted, and it is now searched for its smallest key, which is swapped
into place. This continues until the entire array is sorted. Sample code is in
Figure \vref{Selection sort code}.

\code{Selection sort code}{on2_selectsort.c}
\subsection{Testing}
All elementary sorts are only sorted up to 65536 keys in the simulations, and
262144 keys with the performance counters. Due to the quadratic nature of
these sorts, days and weeks would be required to run tests on the larger arrays.

\subsubsection{Expected Performance}
Sedgewick provides a discussion on the performance of elementary sorts. The
observations on instruction count are his, and come from \cite{Sedgewick02}.

Selection sort uses approximately $N^2/2$ comparisons and exactly $N$ exchanges.
As a result, selection sort can be very useful for sorts involving very large
records, and small keys. In this case, the cost of swapping keys dominates the
cost of comparing them. Conversely, for large keys with small records, the cost
of selection sort is far higher than for other simple sorts.

The performance of selection sort is not affected by input, so the instruction
count varies very little.  Its cache behaviour is also not affected by input,
and selection sort performs very badly from a cache perspective. There are $N$
traversals of the array, leading to bad temporal reuse. If the array doesn't fit
inside the cache, then there will be no temporal reuse.

The level 1 cache should have equally bad performance as a result. Since it is
smaller than the array, the performance of the level 1 cache should simulate the
performance of the level 2 cache.

The branch prediction performance is not expected to be bad. Flow control
predictions are very straightforward, and should result in very few misses.
Comparison predictions, however, are very numerous. Each array traversal has an
average of $N/2$ comparisons, and approximately $log_2(N/2)$ of them should be
taken, while the rest are not-taken. The taken predictions will be misses,
while the rest should be hits, so it is expected that there will be $logN$
misses per key.

This number is not immediately obvious. Consider a single step of the selection
sort. The first key to be sorted is, on average, larger than half the keys and
smaller than the other half, so it will have to travel across half the array to
find a key smaller than it, on average. Once it finds a smaller key, this will
be larger than a quarter of the keys in the array, and so it will need to travel
half of the remaining array, on average, before it finds it. The search will
continue like this until the entire array has been covered. Comparing two keys
and discovering that the new key is not smaller than the current smallest is the
norm; every time a smaller key is found a branch misprediction will occur (if
two or more keys in a row are successively smaller, this may not be the case).
The length of the array to be considered in $N/2$ on average, and an average
total of $Nlog_2(N/2)$ branch misses will occur. This assumes randomly sorted
keys.

\section{Insertion sort}

Insertion sort, as its name suggests, is used to insert a single key into its
correct position in a sorted list. Beginning with a sorted array of length one,
the key to the right of the sorted array is swapped down the array a key at a
time, until it is in its correct position. The entire array is sorted in this
manner.

An important improvement to insertion sort is the removal of a bounds check. In
the case that the key being sorted into the array is smaller than the smallest
key in the array, then it will be swapped off the end of the array, and continue
into the void. To prevent this, a bounds check must be put in place, which will
be checked every comparison, doubling the number of branches and severely
increasing the number of instructions. However this can be avoided by using a
sentinel. This is done by finding the smallest key at the start, and putting it
into place. Then no other key can be smaller, and it will not be possible to
exceed the bounds of the array. Another technique would be to put $0$ as the
smallest key, storing the key currently in that position, then putting it in
place at the end. This would use a different insertion sort, with a bounds
check, going in the opposite direction. The first of these is shown in the
sample code in Figure \vref{Insertion sort code}.

\code{Insertion sort code}{on2_insertsort.c}
\subsection{Testing}
\subsubsection{Expected Performance}
These instruction count observations are again due to Sedgewick.  Insertion sort
performs in linear time on a sorted list, or slightly unsorted list. With a
random list, there are, on average, there are $N^2/4$ comparisons and $N^2/4$
half-exchanges\footnote{A half-exchange is where only one part of the swap is
completed, with the other value stored in a temporary to be put in place later.}.
As a result, the instruction count should be high, though it is expected to be
lower than selection sort.

The cache performance should be similar to selection sort as well. While
selection sort slowly decreases the size of the array to be considered,
insertion sort gradually increases the size. However, each key is not checked
with every iteration. Instead, on average, only half of the array will be
checked, and there will be high temporal and spatial reuse due to the order in
which the keys are accessed. This should result in lower cache misses than
selection sort.

The flow control of insertion sort is determined by an outer loop, which should
be predictable, and an inner loop, which, unlike selection sort, is dependent on
data. As a result, the predictability of the flow control is highly dependant on
the predictability of the comparisons.

\section{Bubblesort}
Bubblesort, as its name suggests, works by bubbling small keys from right of
the array to the left. It begins by taking a key from the end and moving it left
while it is less than any keys it passes. Once it comes across a smaller key, it
changes to this key, and moves it towards the left of the array. In this way, every
iteration the smallest key moves left to its final position, and other small
keys are slowly moved left, so that the unsorted part of the array becomes more
sorted as bubble sort continues.

An improvement made to bubblesort is that it stops as soon as the array is
sorted. Sample bubblesort code is in Figure \vref{Bubblesort code}.

\code{Bubblesort code}{on2_bubblesort.c}
\subsection{Testing}
\subsubsection{Expected Performance}
These instruction count observations are again due to Sedgewick.  Bubblesort has
$N^2/2$ comparisons and exchanges. The number of passes over the array, and size
of each pass is the same as selection sort, and similar instruction count
results are expected.  The extra check to see if the array is sorted adds
$O(NlogN)$ instructions, but this is small in comparison to the number of
instructions saved if the sort ended early.

The cache performance of bubblesort should be the same as that of selection
sort. Each key loaded into the cache is used just once, and will be ejected from
the cache if the data set is large enough.

The branch predictive performance is also expected to be similar. While
selection sort has $O(NlogN)$ misses, each of bubblesort's comparisons has
the potential to be a miss, and not just in the pathological case. However, as
the sort continues and the array becomes more sorted, the number of misses
should decrease. The number of flow control misses should be very low.

\section{Shakersort}
Shakersort is sometimes called back-bubblesort. It behaves like a bubblesort
which moves in both directions rather than just one way. Both ends of the array
become sorted in this manner, with the larger keys slowly moving right and the
smaller keys moving left. Sample code for shakersort is in Figure
\vref{Shakersort code}.

\code{Shakersort code}{on2_shakersort.c}
\subsection{Testing}
\subsubsection{Expected Performance}
The instruction count of shakersort is expected to be exactly the same as
bubblesort. The version of shakersort used here, however, does not incorporate
the improvement to end the sort early if the array is already sorted. This will
probably slightly reduce the instruction count as the sort is unlikely to end
early for random keys. In addition, the worst case for bubblesort is not a
worst case for shakersort and is less likely.

The cache performance of shakersort is expected to be superior to that of
bubblesort. A small amount of temporal reuse occurs at the ends of the array,
as each key is used twice once its loaded into the cache. In addition, when the
array is not more than twice the size of the cache, then the keys in the centre
of the array are not ejected from the array at all.

Finally, the branch prediction results are expected to be the same as in bubble
sort, with one difference.  Keys which must travel the entire way across the
array are taken care of earlier in the sort. If the largest key is in the
leftmost position, it must be moved the entire way right, which causes misses
for every move in bubblesort, since each move is disjoint from the next. In
shakersort, however, each in the correct direction will follow a move in the
same direction, each of which should be predicted. This works for a lot of
keys, which are more likely to move in continuous movements than they are in
bubblesort, where only small keys on the right move continuously. The
performance should improve noticeably as a result.

\section{Simulation Results}

% I'm doing a slightly different result here. Level 2 doesn't interest us, so put
% in 2 prediction results instead - 1 for select/insert, 1 for bubble/shaker
\afterpage{
\label{Elementary sort results}

\thispagestyle{empty}
\clearpage
\enlargethispage{14em}
\vspace*{-8em}
\begin{figure}[H]
\begin{changemargin}
\subfigure[Cycles per key - this was measured on a Pentium 4 using hardware performance counters.]
{\label{elementary cycles}\includegraphics[scale=\myscale]{plots/on2_-_cycles.eps}}
\subfigure[Instructions per key - this was simulated using SimpleScalar \cc{sim-cache}.]	
{\label{elementary instructions}\includegraphics[scale=\myscale]{plots/on2_-_Instructions.eps}}
\end{changemargin}
\vspace*{6em}
\caption{Simulated Instruction count and empiric cycle count for elementary sorts}
\label{Simulated Instruction count and empiric cycle count for elementary sorts}
\end{figure}
\newpage

\thispagestyle{empty}
\clearpage
\enlargethispage{14em}
\vspace*{-8em}
\begin{figure}[H]
\begin{changemargin}
\subfigure[Level 1 cache misses per key - this was simulated using SimpleScalar \cc{sim-cache}, simulating an 8KB data
cache with a 32 byte cache line and separate instruction cache.]
{\label{elementary level 1 misses}\includegraphics[scale=\myscale]{plots/on2_-_Level_1_misses.eps}}
\subfigure[Branches per key - this was simulated using \cc{sim-bpred}.]
{\label{elementary branches}\includegraphics[scale=\myscale]{plots/on2_-_Branches.eps}}
\end{changemargin}
\vspace*{6em}
\caption{Cache and Branch prediction simulation results for elementary sorts}
\label{Cache and Branch prediction simulation results for elementary sorts}
\end{figure}
\newpage

\thispagestyle{empty}
\clearpage
\enlargethispage{14em}
\vspace*{-8em}
\begin{figure}[H]
\begin{changemargin}
\centering
\subfigure[Branches misses per key - this was simulated using \cc{sim-bpred}, with bimodal and two-level adaptive
predictors. The simulated two-level adaptive predictor used a 10 bit branch history register which was XOR-ed with the
program counter.]
{\includegraphics[scale=\myscale]{plots/on2_-_Branch_misses1.eps}}
\subfigure[Branches misses per key - this was simulated using \cc{sim-bpred}, with bimodal and two-level adaptive
predictors. The simulated two-level adaptive predictor used a 10 bit branch history register which was XOR-ed with the
program counter.]
{\includegraphics[scale=\myscale]{plots/on2_-_Branch_misses2.eps}}
\end{changemargin}
\vspace*{6em}
\caption{Branch Simulation results for elementary sorts}
\label{Branch Simulation results for elementary sorts}
\end{figure}
}

Figure \ref{elementary instructions} shows the instruction count of each
elementary sort. Each sort's instruction count is as predicted. Selection sort
has the most instructions, and insertion sort has approximately half that
number, as predicted. Bubblesort has almost the same number of instructions as
selection sort, but shakersort has a significant number less.

The level 2 cache performance of these sorts is not considered, as the set sizes
are never larger than the level 2 cache. Instead, the level 1 caches are used,
and their performance can be used to approximate the level 2 cache in these
cases. These results are shown in Figure \ref{elementary level 1 misses}. Each
set size is larger than the cache, and each sort iterates across the entire
array each time, leading to very poor performance.

As expected, selection sort is the worst. This is due to the fact that it
considers the entire unsorted segment of the array for each key to be put in
place. Bubblesort, with similar behaviour, is only slightly better. Shakersort
has a small amount of temporal reuse (see Section \ref{temporal locality}),
reducing its number of misses.

Insertion sort performs best of the four sorts in relation to cache misses. This
is due to the fact that the keys it considered most recently are used again for
the next key. These keys will be swapped out fairly regularly, however, but not
nearly as often as selection sort, where they are swapped out at nearly every
iteration.

Figure \ref{elementary branches} shows the number of branches per key of these
sorts, while \ref{Branch Simulation results for elementary sorts} shows the
branch misses per key.  Branch prediction performance is not surprising. Each
sort has a similar number of branches, with insertion sort having half as many
as selection and bubblesort, and shakersort being half-way between the two. The
number of misses for bubblesort and shakersort are huge; nearly one in four
comparisons are misses. As the data sets gets large, there are nearly 100,000
branch misses involved in sorting a single key.

Selection sort, meanwhile, has the same number of branches as bubblesort, but
only $O(logN)$ misses per key. The most interesting result, though, is insertion
sort. This has exactly one miss per key. Although this result was not predicted
in advance, on reflection this behaviour is obvious: every iteration involves
moving the key left. The only time it will not move left is when its being put
into place, and this only happens once for every key. This shows that it is
possible to completely eliminate even comparative branch misses. It also
establishes insertion sort as the best of the simple sorts, and it is used later
in quicksort and mergesort as a result of this. \label{insertion is
predictable}

Another interesting result comes from a comparison between shakersort and
selection sort. The number of cycles to sort an array (shown in Figure
\ref{elementary cycles}) is nearly twice as high on shakersort as it is for
selection sort. Shakersort has a significantly lower instruction count and cache
miss count than selection sort, but has several orders of magnitude worse branch
misses. This shows how important it is to consider branch misses in an
algorithm, especially comparative ones. It also establishes that exchange-based
algorithms, such as bubblesort and shakersort, will perform considerably worse
than algorithms which have a correlation between successive branches, as in
insertion and selection sorts.

\section{Future Work}
Simulations from elementary sorts take a lot of time. Running the simulation for
bubblesort with 65536 keys takes nearly a day, and the simulation for 4194304
keys would take nearly 11 years to run. Due to this, it is necessary to
extrapolate results for larger data sets from the results from these smaller
sets. The only place where it's necessary to have larger data sets is in the
case of cache results. The cache behaviour of these sorts should change
dramatically once the keys no longer fit in the cache, as they do in subsequent
chapters with other sorts. The results can be simulated, by using a
smaller cache, but in this case all the data sets are too large. It would be
easy to use data sets smaller than the cache, but in this case many factors
would not have time to be amortised, such as filling the cache and branch
prediction tables.

Each of these sorts could have the number of iterations across the array reduced
by a half or a third by considering two or three keys at a time. The resultant
reduction in cache misses is predictable, and instruction count should reduce.
The new branch prediction results, especially in the case of insertion sort, may
be significantly worse, or may stay the same. The results of this would be
interesting.

Shakersort doesn't have an opt-out-early clause like bubblesort, and as a result
would perform very badly for sorted sets. Adding this enhancement is simple, and
would give better results, which would also be more consistent with bubblesort.

It should be possible to make more cache-conscious versions of these sorts.
Making a bubblesort which performs a bubble across an entire cache line would
be an example. However, these would only make a difference on large data sets,
when better sorts such as quicksort should be used. In addition, the complexity
of doing this means it may be easier to code a simple $O(NlogN)$ sort, which
would perform better for the same effort. Due to this, a cache-conscious
elementary sort is but an academic exercise.

### 2005-05-26 docs/radix.tex
make way for the new and improved fs

\label{Radixsort}
%\section{Description}

All other sorts considered in this project are \n{comparison based} sorts.
They compare two keys at a time, and determine the larger and smaller key,
sorting on that basis. Radixsort is a \n{counting based} sort. It works by
separating all the keys into small parts, and processing these parts without
regard for other keys. For radix of two, these small parts are bits, and for a
radix of 10, these parts would be digits.

The steps to sort by radix are simple. First a radix is chosen, based on the
size of the part to be sorted. For 32 bits integers, a byte is used. This means
the radix is 256, as there are 256 possible byte values. An array of 256
counters is created to count the number of keys with a particular least
significant byte. Then the counters are summed in turn to store the cumulative
count of keys that belong before them in the array. Now each byte can be put in
place by accessing the counter indexed by its least significant byte.

These are written into an auxiliary array, which is written back before the next
iteration using the second least significant byte. Since every key is processed
in order, when the next most significant byte is sorted, the order of the less
significant bytes are maintained, so that once the most significant byte is
sorted, the entire array is.

This type of radixsort is called \n{Least Significant Digit} radixsort.
Other implementations are possible, and several improvements also exist, but as
radixsort is included here for comparative purposes, these are not implemented.
They are, however, discussed in Section \ref{radix future work}.

Note that this method is only suitable for unsigned ints. Because a signed int's
sign is contained in the most significant bit, this sort would place negative
numbers after the positive numbers.

\section{Results}
\subsection{Test parameters}
Keys are to be divided into segments 8 bits long, resulting in four passes of
the array. The radix used is therefore $2^8 = 256$.

\subsection{Expected Performance}
Radixsort is expected to have very good performance, compared to comparison
based sorts. This is due to that fact that radixsort is $O(N)$; for an array
as small as 4096 records, it involves 12 times fewer steps than the $O(NlogN)$ sorts.

The cache performance of radixsort is not expected to be good, however.
Quicksort and mergesort, for example, have very good temporal locality. Keys are
used repeatedly before being swapped out of the cache. In radixsort, however,
every key is used just once each time it is in the cache. When the array is larger than
the size of the cache, every key access will be a miss as a result, and every
new byte being sorted involves four passes over the array. In addition, if the
auxiliary array is pathologically aligned, then the copy back into the array
can result in thrashing.

The number of iterations across the array is constant, however, proportional to
$(bitsPerInt/log_2(Radix))$. For a 32 bit integer, this is four steps and each step
has four passes over the array, for a total of 16 passes over the array.  With
eight  keys per cache line, there should be about two misses per key, when the
array doesn't fit into the array, though this depends on the placing of the
auxiliary array.

The number of branch prediction misses should be exceptionally low. In
comparison based sorts, a branch due to a comparison can not be predicted except
in exceptional circumstances\footnote{See Section \ref{insertion is
predictable}.}. As a result, only flow control branches occur in radixsort, and
they are all predictable.

\subsection{Simulation results}
\plot{radix}{}{}{}{}{}{}

The instruction count of radixsort, shown in Figure \ref{radix instructions} is
very low. It is lower than even quicksort for the smallest set, and it
flat-lines immediately (after amortisation of the loop overhead). The number of
cache misses in Figure \ref{radix level 2 misses} is exactly the same, until the
array no longer fits into the cache. The number of misses then leaps up to 2.5
per
key, which is roughly what was predicted.

The number of branch misses, shown in Figure \ref{radix branch misses} was also
extremely low. Despite 10 branches per key, there was less than .08 of a branch
miss per key. The number of branches itself is quite low, as several of the
sorts considered had nearly that number of misses. However, it is not
unexpected.

With the low number of branch misses, a much better overall performance is
expected than is achieved. When in the cache, it performs only slightly worse
than quicksort, but once the data set grows larger, the gap widens.  However,
the results, shown in Figure \ref{radix cycles}, indicate that radixsort will
never increase beyond 350 cycles per key, meaning that for larger sets it will
perform significantly better than quicksort. The reason for the high number of
cycles is that radixsort was not optimised. Possible optimisations are discussed
below, and should significantly reduce the number of cache misses, the main
cause of radixsorts poor performance.

LaMarca's radixsort results are quite different to those presented here. His
instruction count graph has a logarithmic shape: it begins very high and sharply
falls before flattening out. By contrast, the graph here is almost completely
flat. A similar thing happens in both cache miss graphs: a sharp dip in
LaMarca's results, and a flat-line with a slight slope in these results. Also,
in both cases, once the array no longer fits in the cache, the number of misses
jumps to around 2.5, where they stay.

The cycles per key in both cases are unsurprising based on instruction count and
cache miss results: LaMarca's have a sharp fall before rising slightly once the
data set no longer fits in the cache; the result presented here flat-line before rising
slightly for the same reason. While both results agree for large data sets, they do not
for smaller sets.

\section{Future Work}
\label{radix future work}
Many improvements exist to radixsort, and some are suggested by Sedgewick and
LaMarca. Some of these improvements help to lower cache misses and instruction
count, both of which slow down radixsort significantly. LaMarca suggested
that while the current byte is being put in place, the counting of the next byte
can be done. This increases the memory requirements, but this is not
significant.

The first improvement is to reduce the number of runs across the array per step.
At the moment, there is one step to count, one to copy, and one to write back
(causing potentially two misses per write). LaMarca suggests counting for the next
step in the previous count. In fact, all the counting can be done in a single
pass at the start, instead of four passes at various stages, decreasing instruction
count and cache misses. The downside is the fourfold increase in memory required
for the arrays of counts, but this is small compared to the main and auxiliary
arrays.

The step to copy the array back can also be removed. In mergesort, copying back
to the source array was changed to copying back and forth between the source and 
auxiliary arrays. The same could be potentially done with radixsort, without any
increase in complexity. The number off steps in likely to be even, and is known
at compile time, so the number of times the copy step is done could be reduced
to zero. If the number of steps is odd, which is exceptionally unlikely when
sorting integers, then one copy at the end would be required.

Combining other sorts with radixsort could also be possible. An optimisation to
mergesort was to do a k-way merge at the end, but this dramatically increased
the instruction count. Using a type of radixsort here would be complicated, but
may be possible. Alternatively, sorting inside the cache using radixsort is far
faster than merging. The k-way merge could then be performed at the end. In
practice, though, radixsort is far faster than mergesort, has a lower
instruction count, and less cache misses. While these steps could improve
mergesort, they are more likely to slow radixsort, with no advantage.

Radixsort could also be faster than multi-quicksort's initial step. Replacing
this with a single radix step would split the array into 256 containers, each of
which could be quicksorted. Alternatively, a smaller radix, based on the size of
the array, could be generated at run time. Using LaMarca's analysis,
$C/3$ containers need to be generated. With $N$ as the number of keys in the
array, and $C$ is the number of keys which fit in the level 2 cache, the radix
can be calculated with the formula:
$$2^{\lceil{}log_2(\frac{3N}{CacheSize})\rceil{}}$$
This will only work for random, evenly distributed keys. If all the keys only
vary in their least significant bits, then this will have no effect.

Additional performance improvements could possibly be gained using a different
radix. This radix was chosen simply because Sedgewick used it. Another radix may
lead to better results.

The radixsort presented above is a Least Significant Digit radixsort. A Most
Significant Digit radixsort behaves differently. Once the initial sort is done, each
bin in recursed into, and continually sorted in this manner. This results in a
higher instruction count, and increased stack usage due to the recursion.
However, there should be a large increase in temporal reuse due to this, which
may make up for it.

### 2005-05-26 paper/docs/bugs.tex
make way for the new and improved fs

--- 2005-05-09 docs/bugs.tex
@@
 continue on this work with minimal difficutly.
-
-\section{Algorithmic}
-
-\label{bug1}
-In multi-quicksort, the results include the sentinel check which should have
-been removed before the results were measured. This adds one branch per \cc{N /
-THRESHOLD} keys , and the same number of extra instructions. In the same
-algorithm, the old number of 1018 is the number of keys in a block of a list.
-The effects of this are not significant. Also, the check for the sentinel was
-not removed, though the effect of this is minor.
-
-\label{bug2}
-In several of the mergesorts, it is possible that an extra merge step is being
-taken. This has not been properly analysed, due to time restrictions. It is,
-however, obvious from the level 2 miss graph of all the merges, that this
-does not affect base mergesort, but it does affect both double mergesorts, and
-the tiled mergesort. It is uncertain if multi mergesort is affected. The problem
-stems from the end condition of the merge. It should end once the size of the next
-merge is equal the size of the array, rather than greater than. However, the
-algorithms do not properly merge when this is the case.
-
-\label{bug4}
-Double tiled mergesort and double multi-mergesort were optimized to sort in the
-level 1 cache first. The number of keys that fit into the cache was fixed at
-2048; this number should have been half that, since the sort is out of place.
-This probably resulted in thrashing, especially since the arrays were aligned.
-Removing this would improve the sorts.
 

### 2005-05-26 paper/docs/code/cache_radixsort_code.tex
make way for the new and improved fs

\documentclass[a4paper,twoside]{article}
\usepackage[procnames,noindent]{lgrind}
\usepackage{fancyhdr,a4wide}
\usepackage{german}
\usepackage{makeidx}
\pagestyle{fancy}

\makeindex

\begin{document}
\renewcommand{\footrulewidth}{0.4pt}
\fancyhead[C]{\lgrindhead}
\fancyhead[LO,RE]{\lgrindfilesize~Bytes\\\lgrindmodtime}
\fancyhead[RO,LE]{\bfseries \lgrindfilename\\\lgrindmodday.\lgrindmodmonth.\lgrindmodyear}
\fancyfoot[C]{\bfseries\thepage}
\setlength{\headheight}{24pt}
\begin{lgrind}
\BGfont
\File{cache\_radixsort.c}{2004}{7}{14}{20:54}{1479}
\L{\LB{\K{\#include}_\<\V{stdlib}.\V{h}\>}}
\L{\LB{\K{\#include}_\<\V{string}.\V{h}\>}}
\L{\LB{}}
\L{\LB{\K{\#include}_\S{}\3base\_sorts.h\3\SE{}}}
\L{\LB{}}
\L{\LB{\K{\#define}_\V{THRESHHOLD}_\N{10}}}
\L{\LB{}}
\L{\LB{\K{\#define}_\V{BITSWORD}_\N{32}}}
\L{\LB{\K{\#define}_\V{BITSBYTE}_\N{8}}}
\L{\LB{\K{\#define}_\V{BYTESWORD}_\N{4}}}
\L{\LB{\K{\#define}_\V{R}_\N{256}}}
\L{\LB{}}
\L{\LB{\K{\#define}_\V{digit}(\V{A},_\V{B})_(((\V{A})_\>\>_(\V{BITSWORD}\-((\V{B})+\N{1})*\V{BITSBYTE}))_\&_(\V{R}\-\N{1}))}}
\L{\LB{}}
\L{\LB{\K{void}}}
\index{cache\_radixsort}\Proc{cache\_radixsort}\L{\LB{\V{cache\_radixsort}(\K{unsigned}_\K{int}_\V{a}[\,],_\K{int}_\V{N})}}
\L{\LB{\{}}
\L{\LB{}\Tab{8}{\K{unsigned}_\K{int}*_\V{aux}_=_\V{malloc}(\V{N}_*_\K{sizeof}(\K{unsigned}_\K{int}));}}
\L{\LB{}\Tab{8}{\K{int}_\V{i},_\V{j},_\V{w};_\C{}/*_R_is_the_size_of_a_digit,_hence_the_number_of_bins_*/\CE{}}}
\L{\LB{}\Tab{8}{\K{int}_\V{count0}[\V{R}+\N{1}];}}
\L{\LB{}\Tab{8}{\K{int}_\V{count1}[\V{R}+\N{1}];}}
\L{\LB{}\Tab{8}{\K{int}*_\V{temp};}}
\L{\LB{}}
\L{\LB{}\Tab{8}{\K{int}*_\V{count}_=_\V{count0};}}
\L{\LB{}\Tab{8}{\K{int}*_\V{next\_count}_=_\V{count1};}}
\L{\LB{}}
\L{\LB{}\Tab{8}{\K{unsigned}_\K{int}*_\V{s}_=_\V{a};}}
\L{\LB{}\Tab{8}{\K{unsigned}_\K{int}*_\V{t}_=_\V{aux};}}
\L{\LB{}}
\L{\LB{}\Tab{8}{\C{}/*_count_the_digits_for_the_first_time_*/\CE{}}}
\L{\LB{}\Tab{8}{\K{for}(\V{j}_=_\N{0};_\V{j}_\<_\V{R};_\V{j}++)_\V{count}[\V{j}]_=_\N{0};}}
\L{\LB{}\Tab{8}{\K{for}(\V{i}_=_\N{0};_\V{i}_\<_\V{N};_\V{i}++)_\V{count}[\V{digit}(\V{s}[\V{i}],_\N{3})_+_\N{1}]++;}}
\L{\LB{}\Tab{8}{\K{for}(\V{j}_=_\N{1};_\V{j}_\<_\V{R};_\V{j}++)_\V{count}[\V{j}]_+=_\V{count}[\V{j}\-\N{1}];}}
\L{\LB{}}
\L{\LB{}\Tab{8}{\K{for}(\V{w}_=_\V{BYTESWORD}_\-_\N{1};_\V{w}_\>_\N{0};_\V{w}\-\-)_\C{}/*_this_is_from_lsb._3_==_lsByte,_0_==_msByte_*/\CE{}}}
\L{\LB{}\Tab{8}{\{}}
\L{\LB{}\Tab{16}{\K{for}(\V{j}_=_\N{0};_\V{j}_\<_\V{R};_\V{j}++)_\V{next\_count}[\V{j}]_=_\N{0};_\C{}/*_set_the_next_counts_to_0_*/\CE{}}}
\L{\LB{}}
\L{\LB{}\Tab{16}{\K{for}(\V{i}_=_\N{0};_\V{i}_\<=_\V{N}\-\N{1};_\V{i}++)_\C{}/*_when_we_look_up_the_array,_we_now_have_its_position_*/\CE{}}}
\L{\LB{}\Tab{16}{\{}}
\L{\LB{}\Tab{24}{\V{t}[\V{count}[\V{digit}(\V{s}[\V{i}],_\V{w})]++]_=_\V{s}[\V{i}];}}
\L{\LB{}\Tab{24}{\V{next\_count}[\V{digit}(\V{s}[\V{i}],_\V{w}\-\N{1})_+_\N{1}]++;}}
\L{\LB{}\Tab{16}{\}}}
\L{\LB{}\Tab{8}{}}
\L{\LB{}\Tab{16}{\K{for}(\V{j}_=_\N{1};_\V{j}_\<_\V{R};_\V{j}++)_\V{next\_count}[\V{j}]_+=_\V{next\_count}[\V{j}\-\N{1}];_\C{}/*_make_the_next_counts_cumulative_*/\CE{}}}
\L{\LB{}}
\L{\LB{}\Tab{16}{\C{}/*_swap_the_counts_*/\CE{}}}
\L{\LB{}\Tab{16}{\V{temp}_=_\V{next\_count};}}
\L{\LB{}\Tab{16}{\V{next\_count}_=_\V{count};}}
\L{\LB{}\Tab{16}{\V{count}_=_\V{temp};;}}
\L{\LB{}}
\L{\LB{}\Tab{16}{\C{}/*_swap_the_arrays_*/\CE{}}}
\L{\LB{}\Tab{16}{\V{temp}_=_\V{t};}}
\L{\LB{}\Tab{16}{\V{t}_=_\V{s};}}
\L{\LB{}\Tab{16}{\V{s}_=_\V{temp};}}
\L{\LB{}\Tab{8}{\}}}
\L{\LB{}}
\L{\LB{}\Tab{8}{\C{}/*_finish_it_off_*/\CE{}}}
\L{\LB{}\Tab{8}{\K{for}(\V{i}_=_\N{0};_\V{i}_\<=_\V{N}\-\N{1};_\V{i}++)_\C{}/*_when_we_look_up_the_array,_we_now_have_its_position_*/\CE{}}}
\L{\LB{}\Tab{8}{\{}}
\L{\LB{}\Tab{16}{\V{t}[\V{count}[\V{digit}(\V{s}[\V{i}],_\V{w})]++]_=_\V{s}[\V{i}];}}
\L{\LB{}\Tab{8}{\}}}
\L{\LB{}}
\L{\LB{}\Tab{8}{\V{free}(\V{aux});}}
\L{\LB{\}}}
\end{lgrind}
\printindex
\end{document}

### 2005-05-26 paper/docs/code/sample_sim_list.tex
make way for the new and improved fs

% Remember to use the lgrind style

\Head{}
\File{sample\_sim\_list}{2004}{4}{28}{3:48}{2185}
\L{\LB{\V{sslittle}\-\V{na}\-\V{sstrix}\-\V{gcc}_\-\V{Wall}_\-\V{O3}_\-\V{funroll}\-\V{loops}_\-\V{finline}\-\V{functions}_}}
\L{\LB{}\Tab{8}{\-\V{DRANDOM\_SIZE}=\N{4096}_\-\V{DSEED}=\N{0}_\-\V{DSORT}=\V{base\_heapsort}_}}
\L{\LB{}\Tab{8}{\V{base\_heapsort}.\V{c}_\V{simpleMain}.\V{c}_\-\V{o}_\V{base\_heapsort}}}
\L{\LB{\V{sim}\-\V{cache}_\-\V{cache}:\V{dl1}_\V{dl1}:\N{256}:\N{32}:\N{1}:\V{l}_\-\V{cache}:\V{dl2}_\V{dl2}:\N{131072}:\N{32}:\N{1}:\V{l}_}}
\L{\LB{}\Tab{24}{\-\V{cache}:\V{il1}_\V{il1}:\N{256}:\N{32}:\N{1}:\V{l}_\-\V{cache}:\V{il2}_\V{dl2}_\V{base\_heapsort}}}
\L{\LB{\V{sim}\-\V{cache}_\-\V{cache}:\V{dl1}_\V{dl1}:\N{256}:\N{32}:\N{1}:\V{l}_\-\V{cache}:\V{dl2}_\V{dl2}:\N{65536}:\N{32}:\N{1}:\V{l}_}}
\L{\LB{}\Tab{24}{\-\V{cache}:\V{il1}_\V{il1}:\N{256}:\N{32}:\N{1}:\V{l}_\-\V{cache}:\V{il2}_\V{dl2}_\V{base\_heapsort}}}
\L{\LB{\V{sim}\-\V{cache}_\-\V{cache}:\V{dl1}_\V{dl1}:\N{1}:\N{32}:\N{256}:\V{l}_\-\V{cache}:\V{dl2}_\V{dl2}:\N{1}:\N{32}:\N{131072}:\V{l}_}}
\L{\LB{}\Tab{24}{\-\V{cache}:\V{il1}_\V{il1}:\N{1}:\N{32}:\N{256}:\V{l}_\-\V{cache}:\V{il2}_\V{dl2}_\V{base\_heapsort}}}
\L{\LB{\V{sim}\-\V{cache}_\-\V{cache}:\V{dl1}_\V{dl1}:\N{1}:\N{32}:\N{256}:\V{l}_\-\V{cache}:\V{dl2}_\V{dl2}:\N{1}:\N{32}:\N{65536}:\V{l}_}}
\L{\LB{}\Tab{24}{\-\V{cache}:\V{il1}_\V{il1}:\N{1}:\N{32}:\N{256}:\V{l}_\-\V{cache}:\V{il2}_\V{dl2}_\V{base\_heapsort}}}
\L{\LB{\V{sim}\-\V{bpred}_\-\V{bpred}_\V{bimod}_\-\V{bpred}:\V{bimod}_\N{512}_\V{base\_heapsort}}}
\L{\LB{\V{sim}\-\V{bpred}_\-\V{bpred}_\V{bimod}_\-\V{bpred}:\V{bimod}_\N{1024}_\V{base\_heapsort}}}
\L{\LB{\V{sim}\-\V{bpred}_\-\V{bpred}_\V{bimod}_\-\V{bpred}:\V{bimod}_\N{2048}_\V{base\_heapsort}}}
\L{\LB{\V{sim}\-\V{bpred}_\-\V{bpred}_\N{2$\ell$ev}_\-\V{bpred}:\N{2$\ell$ev}_\N{1}_\N{512}_\N{10}_\N{1}_\V{base\_heapsort}}}
\L{\LB{\V{sim}\-\V{bpred}_\-\V{bpred}_\N{2$\ell$ev}_\-\V{bpred}:\N{2$\ell$ev}_\N{1}_\N{1024}_\N{10}_\N{1}_\V{base\_heapsort}}}
\L{\LB{\V{sim}\-\V{bpred}_\-\V{bpred}_\N{2$\ell$ev}_\-\V{bpred}:\N{2$\ell$ev}_\N{1}_\N{2048}_\N{10}_\N{1}_\V{base\_heapsort}}}
\L{\LB{\V{sim}\-\V{bpred}_\-\V{bpred}_\N{2$\ell$ev}_\-\V{bpred}:\N{2$\ell$ev}_\N{1}_\N{4096}_\N{10}_\N{1}_\V{base\_heapsort}}}
\L{\LB{}}
\L{\LB{}\Tab{56}{.}}
\L{\LB{}\Tab{56}{.}}
\L{\LB{}\Tab{56}{.}}
\L{\LB{}\Tab{56}{.}}
\L{\LB{}\Tab{56}{.}}
\L{\LB{}\Tab{56}{.}}
\L{\LB{}}
\L{\LB{\V{sslittle}\-\V{na}\-\V{sstrix}\-\V{gcc}_\-\V{Wall}_\-\V{O3}_\-\V{funroll}\-\V{loops}_\-\V{finline}\-\V{functions}_}}
\L{\LB{}\Tab{8}{\-\V{DRANDOM\_SIZE}=\N{4194304}_\-\V{DSEED}=\N{0}_\-\V{DSORT}=\V{base\_heapsort}_}}
\L{\LB{}\Tab{8}{\V{base\_heapsort}.\V{c}_\V{simpleMain}.\V{c}_\-\V{o}_\V{base\_heapsort}}}
\L{\LB{\V{sim}\-\V{cache}_\-\V{cache}:\V{dl1}_\V{dl1}:\N{256}:\N{32}:\N{1}:\V{l}_\-\V{cache}:\V{dl2}_\V{dl2}:\N{131072}:\N{32}:\N{1}:\V{l}_}}
\L{\LB{}\Tab{24}{\-\V{cache}:\V{il1}_\V{il1}:\N{256}:\N{32}:\N{1}:\V{l}_\-\V{cache}:\V{il2}_\V{dl2}_\V{base\_heapsort}}}
\L{\LB{\V{sim}\-\V{cache}_\-\V{cache}:\V{dl1}_\V{dl1}:\N{256}:\N{32}:\N{1}:\V{l}_\-\V{cache}:\V{dl2}_\V{dl2}:\N{65536}:\N{32}:\N{1}:\V{l}_}}
\L{\LB{}\Tab{24}{\-\V{cache}:\V{il1}_\V{il1}:\N{256}:\N{32}:\N{1}:\V{l}_\-\V{cache}:\V{il2}_\V{dl2}_\V{base\_heapsort}}}
\L{\LB{\V{sim}\-\V{cache}_\-\V{cache}:\V{dl1}_\V{dl1}:\N{1}:\N{32}:\N{256}:\V{l}_\-\V{cache}:\V{dl2}_\V{dl2}:\N{1}:\N{32}:\N{131072}:\V{l}_}}
\L{\LB{}\Tab{24}{\-\V{cache}:\V{il1}_\V{il1}:\N{1}:\N{32}:\N{256}:\V{l}_\-\V{cache}:\V{il2}_\V{dl2}_\V{base\_heapsort}}}
\L{\LB{\V{sim}\-\V{cache}_\-\V{cache}:\V{dl1}_\V{dl1}:\N{1}:\N{32}:\N{256}:\V{l}_\-\V{cache}:\V{dl2}_\V{dl2}:\N{1}:\N{32}:\N{65536}:\V{l}_}}
\L{\LB{}\Tab{24}{\-\V{cache}:\V{il1}_\V{il1}:\N{1}:\N{32}:\N{256}:\V{l}_\-\V{cache}:\V{il2}_\V{dl2}_\V{base\_heapsort}}}
\L{\LB{\V{sim}\-\V{bpred}_\-\V{bpred}_\V{bimod}_\-\V{bpred}:\V{bimod}_\N{512}_\V{base\_heapsort}}}
\L{\LB{\V{sim}\-\V{bpred}_\-\V{bpred}_\V{bimod}_\-\V{bpred}:\V{bimod}_\N{1024}_\V{base\_heapsort}}}
\L{\LB{\V{sim}\-\V{bpred}_\-\V{bpred}_\V{bimod}_\-\V{bpred}:\V{bimod}_\N{2048}_\V{base\_heapsort}}}
\L{\LB{\V{sim}\-\V{bpred}_\-\V{bpred}_\N{2$\ell$ev}_\-\V{bpred}:\N{2$\ell$ev}_\N{1}_\N{512}_\N{10}_\N{1}_\V{base\_heapsort}}}
\L{\LB{\V{sim}\-\V{bpred}_\-\V{bpred}_\N{2$\ell$ev}_\-\V{bpred}:\N{2$\ell$ev}_\N{1}_\N{1024}_\N{10}_\N{1}_\V{base\_heapsort}}}
\L{\LB{\V{sim}\-\V{bpred}_\-\V{bpred}_\N{2$\ell$ev}_\-\V{bpred}:\N{2$\ell$ev}_\N{1}_\N{2048}_\N{10}_\N{1}_\V{base\_heapsort}}}
\L{\LB{\V{sim}\-\V{bpred}_\-\V{bpred}_\N{2$\ell$ev}_\-\V{bpred}:\N{2$\ell$ev}_\N{1}_\N{4096}_\N{10}_\N{1}_\V{base\_heapsort}}}

### 2005-05-26 paper/docs/heap.tex
make way for the new and improved fs

--- 2005-05-26 docs/heap.tex
@@
 
+\section{Predictor results}
+
+\sixplot{base_heapsort-0.eps}{cache4_heapsort-0.eps}{cache_heapsort-0.eps}{cache_heapsort-1.eps}{}{}{base and cache heapsorts}
+
+The results of the software predictor show the relation of the branches
+in the creation and destruction of the heap. The base heapsort results in
+Figure \ref{base and cache heapsorts-1} show that the time taken to create the
+heap is not significant relative to the time taken to sort from the heap.
+
+The most interesting thing is the relation between the predictivity of the
+comparisons in the destruction of the 2-, 4- and 8-heaps, which are shown in
+Figure \ref{Branch prediction performance for base and cache heapsorts}.
+Overall, the 4-heap is half as high as the 2-heap, and twice as high as the
+8-heap, but has to make more comparisons. However, there are three comparisons
+in the 4-heap and seven comparisons in the 8-heap. In total, there are 50\% more
+comparisons in the 4-heap, and 150\% more in the 8-heap.
+
+These branches become more predictable. The first predicts correctly 50\% of the
+time, while the next predicts correctly 60\% of the time, then 71\%, 77\%, and
+so on in a curve. This curve is shown in Figure TODO.
+%TODO make a pretty graph for this.
+
+Also notable from the results is that branches are more often not-taken then
+they are correctly predicted, implying that a semi-static predictor based on
+hints from the compiler would do a better job at making correct predictions.
+
 \section{Future Work}

### 2005-05-26 paper/docs/intro.tex
make way for the new and improved fs

Most major sorting algorithms in computer science date to the 50s and 60s.
Radixsort was written in 1954, quicksort in 1962, and mergesort has been traced
back to the 1930s. Modern processors meanwhile have features which were not
thought of when these sorts were written: branch predictors, for example, were
not created until the 1970s, and advanced branch prediction techniques were
not used in consumer processors until very recently. It is hardly surprising,
then, that sorting algorithms rarely take these architectural features into
account.

Papers considering sorting performance typically analyse instruction count.
LaMarca and Ladner \cite{LaMarca96} break this trend by discussing the cache
implications of several algorithms, including sorts. This project attempts to
investigate these claims and to test and analyse the high performance sorts
based on them.

LaMarca devotes a chapter of his thesis to sorting. A chapter is spent
discussing implicit heaps, which are later used in heapsort. In
\cite{LaMarca99}, radixsort was added for comparison. Radixsort is added here
for the same purpose. For comparison, elementary sorts - insertion sort,
selection sort, bubblesort and shakersort - are also included here.

In addition to repeating LaMarca's work, this project also discusses hardware
branch prediction techniques and their effects on sorting. When algorithms are
altered to improve their performance, it is important to observe if any change
in the rate of branch prediction misses occurs.

Over the course of the cache-conscious algorithm improvements, branch prediction
results are generated in order to observe the side-effects of the
cache-conscious improvements on the branch predictors, and whether they affect
different branch predictors in different ways.

Several important results are discovered during the course of the project. The
branch prediction rates of insertion sort and selection sort, for example, and
the difference in performance between binary searches and sequential searches due
to branch misses. Several steps to remove instructions from standard algorithms
are devised, such as double tiling mergesort, and removing a sentinel and bounds
check from heapsort.

Chapter 2 provides background information on caches and branch predictors. It
also discusses the theory of sorting, and conventions used in the algorithms in
this project.

Chapter 3 discusses the method and framework used in this project, including the
tools used and the reasons for using them.

Chapter 4 discusses the elementary sorts and their implementations. Results are
included for later comparison.

Chapters 5 through 7 discuss base algorithms and cache-conscious improvements
for heapsort, mergesort and quicksort. Results are presented and explained, and
branch prediction implications of the changes are discussed.

Chapter 8 presents a radixsort implementation and discusses its performance in
the areas of instruction count, cache-consciousness and branch prediction.

Chapter 9 presents conclusions and contributions developed over the course of
this project.

### 2005-05-26 paper/docs/merge.tex
make way for the new and improved fs

--- 2005-05-26 docs/merge.tex
@@
 
-\section{Tiled Mergesort}
+\section{Double Tiled Mergesort}
 LaMarca's first problem with mergesort is the placement of the arrays with
@@
 
-\section{Multi-mergesort}
+It is also possible to avoiding the copy back to memory. In the event of an odd
+number of merge steps, the auxiliary array will contain the sorted array, which
+needs to be copied back. This can be avoided by pre-sorting to make lists sized
+two or eight. A large increase of instruction count was realized by sorting to
+lists of length two, so a length of eight was used. When an even number was
+required, four was used.
 
-Multi-mergesort has two phases. The first is the same as tiled mergesort, and
+The final impriovement was to fully sort keys in the level 1 cache, before
+continuing to level 2 cache as in the tiled mergesort. The expected improvement
+is not as significant as in the case of the level 2 cache, since is the penalty
+for level 1 misses is not great. For this reason, this sort is called Double
+Tiled Mergesort instead of Tiled Mergesort, however, it is referred to as both
+in this report.
+
+\section{Double Multi-mergesort}
+
+Double Multi-mergesort has two phases. The first is the same as tiled mergesort, and
 sorts the array into lists sized $C/2$. The second phase attempts to improve
@@
 
-\section{Double Tiled Mergesort \\/ Double Multi-mergesort}
-
-While writing the cache mergesorts, I realized several opportunities for
-improvement. The first of these is avoiding the copy back to memory. In the
-event of an odd number of merge steps, the auxiliary array will contain the
-sorted array, which needs to be copied back. This can be avoided by pre-sorting
-to make lists sized two or eight. A large increase of instruction count was realized by
-sorting to lists of length two, so a length of eight was used. When an even number
-was required, four was used.
-
-The second improvement was to fully sort keys in the level 1 cache, before
-continuing to level 2 cache as in the tiled mergesort. The expected improvement
-is not as significant as in the case of the level 2 cache, since is the penalty
-for level 1 misses is not great.
-
-Finally, the merge is rewritten in a more straightforward and readable manner. This
-was necessary due to the increased complexity of debugging the double mergesorts.
-During the process, different functions were used to process lists that were
-exactly the required size and lists which were not, which needed code to handle
-edge-cases. However, the instruction count improvement was in the double digits,
-and so these changes were removed in order to have lower instruction cache
-misses, as well as to reduce the fill time of the branch predictors and reduce
-the complexity and size of the algorithm.
-
 \section{Results}
@@
 according to input. Multi-mergesort should have a significantly higher
-instruction count, once its threshold is hit, as should its variant, double
-multi-mergesort.
+instruction count once its threshold is hit.
 
@@
 as predicted, with a 50\% spike at multi-mergesort. Level 2 misses, shown in
-\ref{merge level 2 misses} are also as expected, although the multi-mergesort
-does not show the same improvement over tiled mergesort as in LaMarca's results.
-Double multi-mergesort does show an improvement, but not a great improvement to
-double tiled mergesort. It does, however, have a much flatter slope.
+Figure \ref{merge level 2 misses} are also as expected, although the
+multi-mergesort does not show the same improvement over tiled mergesort as in
+LaMarca's results.
 
@@
 
+\section{Predictor results}
+
+\afterpage{
+\setcounter{subfigure}{0}
+\thispagestyle{empty}
+\clearpage
+\enlargethispage{14em}
+\vspace*{-5em}
+\begin{figure}[H]
+\begin{changemargin}
+\subfigure[]{\includegraphics[scale=0.74]{../counters/algorithm_n-0.eps}}
+\subfigure[]{\includegraphics[scale=0.74]{../counters/algorithm_s-0.eps}}
+\subfigure[]{\includegraphics[scale=0.74]{../counters/base_mergesort-0.eps}}
+\subfigure[]{\includegraphics[scale=0.74]{../counters/double_tiled_mergesort-0.eps}}
+\subfigure[]{\includegraphics[scale=0.74]{../counters/double_multi_mergesort-0.eps}}
+\subfigure[]{\includegraphics[scale=0.74]{../counters/double_multi_mergesort-1.eps}}
+\end{changemargin}
+\vspace*{6em}
+\caption{Branch prediction performance for mergesorts. All graphs use the same scale.}
+\label{Branch prediction performance for mergesorts}
+\end{figure}
+\newpage
+}
+
+%TODO This is a bit to rewrite
+
+figure a shows Algorithm N, knuths natural merge. N3, N5 and N9 indicate parts of the flow chart. 'N3 i==j' is the second branch in N3.
+
+The reuslts show that the seconsary loops are always taken. Taken here means broken out of, hence seldom used. The main action occurs in N3, the main comparator, which has a 50\% prediction ratio.
+
+Algorithm S in graph B shows the same. I never really understood how it worked though.
+
+Figure c shows base mergesort.
+This shows a lower number of branches and a higher precentage of correct predictions.
+The lower number of branches is probably due to the presort.
+there is a higher percentage of correct predictions, but the actual number of correct predictions in the same.
+This implies that doing the presort removes incorrect predictions (or, more to the point, moves incorrect predictions to another part of the algorithm).
+Note too, that both sides are used equally, which doesnt happen later.
+Two thirds of the branches are not-taken, which means loops execute twice before breaking.
+this is the same results that we found in quicksort later.
+Im sure knuth explains this somewhere in the fine print.
+
+Figure d shows double tiled mergesort.
+The total number fo branches is a 10\% increase over the base mergesort.
+Not sure why this is.
+15\% increase in correct predictions so this incrase into predictable range, which should be avoidable (according to what we guess in quicksort).
+
+Figure e and f show the double mulitmergesort results.
+This sorts fully within the cache only, reducing the number of branches relating to this.
+The reason that the previous balance was so muhc in favour of forwards was that the final sort at the end was forwards.
+Here though, the arrays are left bitonically, so there is an even distribution betwwen forwards and backwards.
+There si still the bias towards left rather than right.
+Left comes first in the code, but this doesnt quite explain why this happens.
+However, the number fo misses each way is identical.
+
+The total branches is a decrease of about 20\% on base mergesort.
+The number of mispredictions decreases by less than 10\%.
+Howver, all these branches are made up by the increase in branches due to the heap.
+The number of extra brnahc misses due to this is 26\% of the branches misses in base mergesort.
+In total, the num ber fo branches isa equivalent however.
+So no loss by doing this (except lots of extra instructions, as seen before).
+
+Note that here and in the heaps, comparison 0 does nothing. THe compiler ought to trim it out, except that now it actually does something (increases the branch counter).
+
+finally, in base mergesort, there is an inline presort used.
+For efficiency, an insertion sort is used later.
+However, there are 6.8 misses per key in the inline presort, which is less than the insertion sort does.
+Its not as useful for presorting to 8, however.
+THese results hold forwards and in reverse.
+
+%TODO up to here
+
 \section{Future work}
@@
 
-The difference between the actual and expected results of the tiled and
-multi-mergesorts is another area to be investigated. While the results point to
-similar conclusions about cache misses between the base, tiled and
-multi-mergesorts, the relation these cache misses for direct mapped caches is
-not expected. Investigating, and possibly fixing this would be interesting.
-
 Algorithm N is a mergesort for more natural layouts of data, such as partially
@@
 may be able to improve its performance, as they did for the straight merge.
-
-A bug exists in the double tiled and multi-mergesorts, which causes an extra
-merge step in some cases. More accurate results could be achieved by solving
-this problem. Also, to take advantage of the level 1 cache, the double sorts were
-sorted to 2048 keys first, then merged into larger sections. This number should
-have been half that. See section \ref{bug4} for more details.
 
@@
 It may also be possible to simply reduce the number of conflict misses of the
-k-way merge, simply by reducing the size of the initial sort by the size of several cache
-lines. In this way, the smallest keys would map to different cache blocks,
-without the need for padding.
+k-way merge, simply by reducing the size of the initial sort by the size of
+several cache lines. In this way, the smallest keys would map to different
+cache blocks, without the need for padding.

### 2005-05-26 paper/docs/method.tex
make way for the new and improved fs

This section discusses the method and framework used to complete this project.
From the point of view of a Final Year Project, this is a description of what I
did all year, what I spent my time on, what went right and wrong during the
project, and what I learned. The first section discusses the steps involved in
creating the project, programming the sorts and getting results. The second
section discusses tools that I used or created in the course of the project.

\section{Method}

The following is the initial project description:
\begin{quotation}
Sorting is one of the most important and studied problems in computer science.
Many good algorithms exist which offer various trade-offs in efficiency,
simplicity, and memory use.  One weakness of classical sorting algorithms is
that they are not designed for modern computers.

Cache memories, pipelines, and branch prediction have greatly changed the
relative strengths of different algorithms. For example, comparison based
sorting algorithms (such as Quicksort) can be inefficient on modern machines,
because they involve difficult to predict branches.

The goal of this project is to design, implement and experiment with sorting
algorithms designed to make the best use of modern architectural features.
Several cache-conscious sorting algorithms have been proposed, and the first
step would be to implement these algorithms and compare their performance with
traditional algorithms.  The next step is to consider modifications to the
algorithms to make better use of the processor pipeline and multiple execution
units and reduce branch mispredictions. The project will also involve a detailed
study of the behaviour of the algorithms using a processor simulator or
performance counters on a real machine.

To do this project, you should have a reasonable knowledge of computer
architecture and traditional sorting algorithms. You should also be enthusiastic
about algorithm design, and making things go faster.  All programming is in C or
C++.
\end{quotation}
After discussion with my supervisor, the following project goals were agreed upon:
\begin{itemize}
\item Implement heapsort, mergesort, quicksort, radixsort, and the four
elementary sorts: insertion sort, selection sort, bubblesort and shakersort.
\item Simulate these sorts using the SimpleScalar architecture simulator, using a
variety of branch predictors and cache configurations.
\item Implement cache-conscious sorts from \cite{LaMarca96} and
\cite{LaMarca99}.
\item Investigate and analyse the increase in speed, reduction in cache misses, 
and change in branch prediction accuracy as a result of these changes.
\end{itemize}

Several steps were necessary to prepare for the project. The first of these was
to compile SimpleScalar. A guide to repeating this is included in Section
\ref{SimpleScalar Install}. A framework was required by which I could test sorts
as they were developed. This is discussed in Section \ref{Main}. It was also necessary
to fully understand the technologies to be used in the project. Several texts
were used for this: cache and branch predictors are both in
\cite{HennessyPatterson90}; branch predictors in
\cite{Uht97}, as well as in \cite{McFarling93}
and \cite{Smith81}; sorts are discussed in \cite{Sedgewick02} and
\cite{Knuth98}; cache-conscious sorts are discussed \cite{LaMarca96},
\cite{LaMarca99} and \cite{LaMarcaHeap96}.

There were several steps involved in writing each iteration of the algorithms.
The elementary sorts and radixsort had just one of these iterations; the other sorts had more
than one\footnote{Quicksort, for example, had four: base quicksort, tuned
quicksort, multi-quicksort and sequential multi-quicksort.}. A flowchart of these
steps appears in Figure \vref{Flowchart}.

\begin{figure}
\scalebox{0.57}{\includegraphics*[0pt,165pt][600pt,802pt]{flowchart}}
\caption{Flowchart describing the development of an algorithm}
\label{Flowchart}
\end{figure}

The first step was to write the algorithm. \cite{Sedgewick02} was the primary
source of these, though in several cases LaMarca specified a particular version
which he used. When a step involved in an algorithm wasn't clear, the paper
describing the algorithm was used to clarify the meaning.

Once the algorithm was written, a testing framework was used. This tested for
edge-cases in the algorithms by testing sorts sized from $8$ to $255$ keys, with
different random data sets. Testing was also performed on a very large data set.
This test was timed and the results were considered. The results of this test
were not accurate: the program took up to a minute to run, during which time the
computer was being used by multiple users and processes, leading to context
switches,sharing of resources and scheduling. Despite these inaccuracies, it was
possible to use the results relative to each other, and to determine whether,
for example, memory-tuned quicksort was running faster than base quicksort.

Early in the project, it was decided that exceeding the bounds of an array
during a sort was to be forbidden. No serious sorting library or application
could acceptably work in this manner, and any results obtained while allowing
this were only of academic interest. It isn't possible, from a normal run of a
program, to determine if an arrays bounds are exceeded slightly, though exceeding
the bounds by a large amount will result is a segmentation fault. Because of
this, Valgrind (see Section \ref{Valgrind}), was used to report these errors.

Once it was certain that a sort actually sorted without error, it was necessary
to ensure that this was not a fluke occurrence. Many of the sorts had multiple
passes, and if one pass did not properly sort the array segment it was meant to,
the array might still be sorted by the next pass, though potentially much more
slowly. To test that this did not occur, a set of functions to visually display
arrays and variables at certain points was developed, which created a HTML page
showing the progression of the algorithm. An example of this is in Figure
\vref{Visual Sort}.

\begin{figure}
\scalebox{0.58}{\includegraphics*{visual}}
\caption{A sample visual sort routine, showing a merge from the source array to
the auxiliary array}
\label{Visual Sort}
\end{figure}

This shows two stages of a bitonic mergesort. The top of the image shows the
source array being merged into the target array below. The source array shows
three large segments of the array and four smaller segments. The first two large
segments have been merged into one segment in the target array. The third of
these has been merged into the corresponding target array segment. The first of
the smaller segments has begun being merged with the other small segments. The
rest of the target array is uninitialised.  Below that is a small amount of
debugging information. $i == j$ is a key point in the program and is printed as
debugging information. In the two arrays below this, the merge continues.

While it was not known in advance the desired instruction count of an algorithm,
a certain range was implied by LaMarca's descriptions. The relationship between
one iteration of an algorithm, or between base versions of several algorithms was
known ahead of time. One configuration of the \cc{simple} script (see
Section \ref{Simple Script}) ran quick tests using the \cc{sim-fast} simulator,
giving instruction counts which could be compared to other versions.

If both these tests and the visual display of the sort's behaviour seemed
correct, then the full collection of simulations was run. The results of the
simulation were compared to results of other versions and to the expected
results. If the results were surprising, then it was necessary to explain the
reasons for this. If the cause was a bug, then the bug was fixed and the tests
repeated.

The results were then turned into graphs, from which conclusions about the
performance of the algorithm could be reached, and patterns could be spotted.
More details of this are in Section \ref{Gnuplot}. The next iteration of the
algorithm was then begun, and the process repeated for each of the
algorithm's iterations.

\section{Tools}

\subsection{SimpleScalar}
\cc{SimpleScalar} is an architecture simulator. It works by compiling code into its
native instruction set, PISA, then emulating a processor as it runs the program.
SimpleScalar can be found at \n{http://www.simplescalar.com}. Documentation on how
to use and modify SimpleScalar are \cite{Burger97}, \cite{Austin02},
\cite{Tutorialv4}, \cite{Tutorialv2} and \cite{SimpleScalarUserGuide}.

The installation of SimpleScalar can be difficult. A step by step description
for installing it on an x86 linux machine is in Section \ref{SimpleScalar
Install}.

SimpleScalar works by compiling code into its own instruction set. It provides a
compiler - a port of gcc 2.63 - with related tools such as a linker and assembler.
Several simulators are then provided: \cc{sim-fast} only counts instruction,
quickly; \cc{sim-cache} simulates a cache and counts level 1 and level 2 cache
misses; \cc{sim-bpred} simulates a branch predictor and counts branch hits and
misses. Other simulators are provided, but they were not used in this project.

The output of SimpleScalar is human readable (see Section \ref{Sample
Simulation} for an example) but is not conducive to collecting large amounts of
data. To that end, a script was written to run each required set of simulations. The
script was written in perl, and recorded data from the simulations into a single
file for later retrieval. The source code to the script, \cc{simple}, is in
section \ref{Simple Script}.  A sample list of commands executed by \cc{simple} are
in section \ref{Simple List}. The data file produced was suitable to reading
with a \cc{diff} program.

The SimpleScalar settings were chosen to be similar to LaMarca's simulations.
LaMarca used the \n{Atom} simulator, which was used to test software for the DEC
AlphaStation. It simulated a 2MB cache with a 32 byte cache line. The remainder
of the settings were taken from the DEC Alpha: 8KB level 1 data cache, 8KB level
1 instruction cache and shared instruction and data level 2 cache.

\subsection{Gnuplot}
\label{Gnuplot}
\cc{Gnuplot} is a tool for plotting data files. It provides visualisations of data in
two and three dimensions, and is used to create all of the graphs in this
document. Its homepage is \n{http://www.gnuplot.info/}.

There were several steps involved in using gnuplot for the data. Firstly, the
data was not in the correct format for gnuplot, and had to be converted. It also
needed to be offset by the cost of filling the array with random variables, and
scaled to so that the y-axis could be in the form of \n{plotted data per key},
both to be more readable, and to be consistent with LaMarca's graphs. The source
code to the script to do this is in Figure \ref{Convert Script}. Once the data
was converted, a script to run gnuplot (see Section \ref{Sample Gnuplot Script}
for an example) was required. The number of graphs required made this an
impossible manual task, and so another script had to be written.  Some graphs
still needed to be tweaked by hand, due to difficulties in the script, but the
task was much shortened by this, and the tweaking was mostly accomplished using
macros in the development platform.

Finally, a script to compare the data together was required.  This compared each
sort against each other sort in key metrics, such a branch or cache misses. It
also compared each cache size against other cache sizes for a particular sort,
and branch predictors against other branch predictors for each sort.

\subsection{Main}
\label{Main}

The main program provided the framework for the project. Imaginatively called
\cc{main.c}, this program provided the means for testing and (roughly)
timing each sort. Figure \vref{Key sections of Main.c} contains this code.

\code{Key sections of Main.c}{main.c}

\subsection{Valgrind}
\label{Valgrind}
\cc{Valgrind} is a set of tools which are used for debugging and profiling.  By
default it works as a memory checker, and detects errors such as the use of
uninitialised memory and accessing memory off the end of an array. This use
makes it particularly useful for bounds checking. No configuration was required
for this, and simply running \cc{valgrind a.out} ran the program through
valgrind. The homepage for valgrind is at \n{http://valgrind.kde.org/}.

\subsection{perfex}
\cc{Perfex} is a tool used to access performance counters on a Pentium 4. By default,
it prints out the number of cycles\footnote{The processor automatically
maintains this figure in a 40-bit register called P4\_TSC (Time Stamp Counter),
and perfex prints this value after running the program} which a program takes to
execute. This was used to create the graphs comparing the running time of sorts.
The gnuplot scripts for this were short and easily written.

Perfex was an alternative to using SimpleScalar. By using its hardware
registers, actual real world results could have been measured. Using perfex
instead of SimpleScalar was an option reviewed at the start, but the difficulty
involved in setting up a machine to use it was considerable. Additionally,
because perfex measures results on a real machine, only one type of cache and
branch predictor could have been measured. Since access was
provided late in the project, limited measurements were taken to supplement the
SimpleScalar results.

\subsection{Incidental Tools}
The algorithms in this project were written in C using \cc{vim}, a powerful text
editor. \cc{gcc} was used to compile the framework, when not using SimpleScalar.
The report was written using \LaTeX, and diagrams were written is
\cc{postscript}, or drawn using \cc{OpenOffice.org}, an open-source office
suite. Many files had to be edited in the same way, and this was accomplished
either using \cc{vim} macros, or by writing a short \cc{perl} script.

\section{Implementation of sorts}

The \cc{C} code written for this project was approximately 8000 lines long.
Appendix \ref{Sort Listings} lists many of the key parts of this code.

To show an example of the creation of a sort, mergesort will now be discussed.
Descriptions of some of the terms here is saved for discussion in Chapter
\ref{Mergesort}.

The base mergesort described by LaMarca is Algorithm N. Three improvements
were prescribed by LaMarca: changing the sort to be bitonic to eliminate bounds
checks, unrolling the inner loop, and pre-sorting the array with a simple in-place
sort. He also recommended making the sort pass between two arrays to avoid
unnecessary copying.

Algorithm N is described in \cite{Knuth98}. A flow control diagram and step by
step guide is provided, but very little detail of how the sort actually works is
included. Therefore, the first step was to understand how this sort worked. This
was done by first analysing the flow control of the algorithm. The steps of the
sort are described in assembly, and while they can be written in C, the result
is a very low-level version. I attempted to rewrite the algorithm without
\cc{goto} statements and labels, but instead using standard flow control
statements such as \cc{if}, \cc{while} and \cc{do-while}. Despite the time spent
on this, it was impossible to faithfully reproduce the behaviour without
duplicating code.

As a result, it was necessary to use the original version, and to try to perform
optimisations on it. However, several factors made this difficult or impossible.
LaMarca unrolled the inner loop of mergesort, but it is not clear which is the
inner loop. The two deepest loops were executed very infrequently, and
unrolling them had no effect on the instruction count. Unrolling the other loops
also had no effect, as it was rare for the loop to execute more than five or six
times, and once or twice was more usual.

Pre-sorting the array was also not exceptionally useful. The instruction count
actually increased slightly as a result of this, indicating that algorithm N
does a better job of pre-sorting the array than the inlined pre-sort did.

Algorithm N does not have a bounds check. It treats the entire array
bitonically; that is, rather than have pairs of array beings merged together,
the entire array is slowly merged inwards, with the right hand side being sorted
in descending order and the left hand side being sorted in ascending order.
Since there is no bounds check, there is no bounds check to be eliminated.

As a result of this problems, algorithm S, from the same source, was
investigated. This behaved far more like LaMarca's description: LaMarca
described his base mergesort as sorting into lists of size two, then four and
so on. Algorithm S behaves in this manner, though Algorithm N does not. Fully
investigating algorithm S, which is described in the same manner as algorithm N,
was deemed to be too time-consuming, despite its good performance.

As a result, a mergesort was written partially based on the outline of algorithm
N. The outer loop was very similar, but the number of merges required were
calculated in advance, and the indices of each merge step.  It was very simple
to turn this into a bitonic array; adding pre-sorting reduced the instruction
count, as did unrolling the loop. The reduction in instruction count from the
initial version, which performed the same number of instructions as the same as
algorithm N, and the optimised version, was about 30\%.

A tiled mergesort was then written, based on LaMarca's guidance. Firstly, this
was aligned in memory so that the number misses due to conflicts would be
reduced.  Then it fully sorted segments of the array, each half the size of the
cache, so that temporal locality would be exploited. Finally, all these segments
were merged together in the remaining merge steps.

LaMarca's final step was to create a multi-mergesort. This attempted to reduce
the number of cache misses in the final merge steps of tiled mergesort by
merging all the sorted arrays in one step. This required the use of an 8-heap,
which had been developed as part of heapsort.

Once these sorts were written it was attempted to improve them slightly.
Instead of a tiled mergesort, a double tiled mergesort was written.  This
attempted to completely sort in the level 1 cache, then to completely sort in
the level 2 cache. Double multi-mergesort was a combination of this and of
multi-mergesort. The rewrite was based on the base mergesort, and was written
far more cleanly, being more readable and efficient. These were also found to
reduce the instruction count, as well as the level 2 cache miss count. These
results are discussed in more detail in \ref{Mergesort results}.

Each of these sort variations had to be fully written from the descriptions in
LaMarca's thesis. Only the base algorithm was available from textbooks, and this
needed to be extensively analysed and rewritten. Each step which was written had
to be debugged and simulated. The results were compared with the expected
results and frequently the sorts needed to be edited to reduce the cost of a
feature.

Each of the three $O(NlogN)$ sorts was treated in this fashion. The mergesorts
accounted for over 3000 lines of the 8000 lines of code in this project. Key
portions of this code are shown in Section \ref{Sort Listings}.

\section{Future work}
SimpleScalar ran very slowly, with each simulation run taking at least $2
\frac{1}{2}$ hours. Several steps could be taken to lower this time. Firstly,
compiling SimpleScalar with the \cc{-03} flag, which performs a significant
number of optimisations on the program. This was not done because of the
difficulty of compiling the program in the first place.

\cc{lib-cheetah} is a SimpleScalar library used to simulate multiple
caches in the same simulation. Over half the simulation time was taken by four
cache simulations, and running them concurrently might have significantly
decreased the running time. This was not a significant worry during this project,
but decreasing the running time would be necessary if the either the number of
simulations or the number of sorts was to change dramatically.

An improvement that would have yielded interesting results would be to do simulations
which only consisted of flow control, with no comparisons. This would allow a
division of flow control misses and comparative misses. However, this would not
always be possible; sorts like quicksort and heapsort mix use their data to
control the flow of the program. In addition, bubblesort, shakersort and
selection sort have (analytically) predictable flow control, while radixsort has
no misses at all. Mergesort may benefit from this, however, especially versions
with difficult flow control diagrams, such as \n{algorithm N} and \n{algorithm
S} (see Section \ref{Algorithm N}).

A metric not considered so far is the number of instructions between branches.
This should be an important metric, and investigating it could point to
important properties within an algorithm.

Valgrind comes with a tool called \cc{cachegrind}, which is a cache
profiler, detailing where cache misses occur in a program. Using this on a
sort may indicate areas for potential improvement.

As well as the number of cycles, Pentium 4 performance registers can also provide
most of the same information as SimpleScalar, with different parameters. For
example, the cache line of a Pentium 4 machine is 128 bytes whereas 32 byte
cache lines were used in SimpleScalar.  The addition of real world data would
strengthen the results from this project.

### 2005-05-26 paper/docs/ordernsquared.tex
make way for the new and improved fs

--- 2005-05-26 docs/ordernsquared.tex
@@
 
+\label{selection branches logn}
 The branch prediction performance is not expected to be bad. Flow control
@@
 
+\section{Bubblesort2}
+Bubblesort2 incorporates an improvement to end bubblesort early. The bubblesort
+implementation above ends early if it detects that the array is fully sorted.
+It keeps a binary flag to indicate if there were any exchanges along the way.
+Instead, bubblesort2 keeps track of the last exchange made, and starts the next
+iteration from there. This means that any pockets of sorted keys are skipped
+over, not just if they occur at the end. Bubblesort2 code is in Figure
+\vref{Bubblesort2 code}
+
+\code{Bubblesort2 code}{on2_bubblesort2.c}
+\subsection{Testing}
+\subsubsection{Expected Performance}
+It is expected that the bubblesort2 should perform faster than the original in
+all metrics. Instruction count, were there no conditions that allowed keys to be
+skipped, would be the same, as no extra instructions are added. The flow of the
+algorithm is the same, though there should be fewer iterations, especially at
+the start, when iterations are longer, so cache accesses, and branch misses
+should be reduced along with instruction count.
+
 \section{Shakersort}
@@
 performance should improve noticeably as a result.
+
+\section{Shakersort2}
+Shakersorts incorporates the same improvements over shakersort as bubblesort2
+does over bubblesort, that is, it skips over pockets of sorted keys in both
+directions, as well as ending early. The code for shakersort2 is in Figure
+\vref{Shakersort2 code}.
+
+\code{Shakersort2 code}{on2_shakersort2.c}
+\subsection{Testing}
+\subsubsection{Expected Performance}
+It is expected that the improvements will make shakersort faster in all cases
+than its original form.
 
@@
 number, as predicted. Bubblesort has almost the same number of instructions as
-selection sort, but shakersort has a significant number less.
+selection sort, but shakersort has a significant number less. The improvement to
+bubblesort in bubblesort2 have increased the instruction count, contrary to
+predictions, and similarly in the shakersort2.
 
@@
 has a small amount of temporal reuse (see Section \ref{temporal locality}),
-reducing its number of misses.
+reducing its number of misses. Shakersort2 is placed roughly between insertion
+sort and shakersort, showing that ending early is helpful in its case. This does
+not appear to be the case with bubblesort2 though, which performs identically to
+bubblesort.
 
@@
 Figure \ref{elementary branches} shows the number of branches per key of these
-sorts, while \ref{Branch Simulation results for elementary sorts} shows the
-branch misses per key.  Branch prediction performance is not surprising. Each
-sort has a similar number of branches, with insertion sort having half as many
-as selection and bubblesort, and shakersort being half-way between the two. The
-number of misses for bubblesort and shakersort are huge; nearly one in four
-comparisons are misses. As the data sets gets large, there are nearly 100,000
-branch misses involved in sorting a single key.
+sorts, while Figure \ref{Branch Simulation results for elementary sorts} shows
+the branch misses per key.  Branch prediction performance is not surprising.
+Each sort has a similar number of branches, with insertion sort having half as
+many as selection and bubblesort, and shakersort being half-way between the
+two. The number of misses for bubblesort and shakersort are huge; nearly one in
+four comparisons are misses. As the data sets gets large, there are nearly
+10,000 branch misses involved in sorting a single key. The improved bubblesort
+had exactly the same number of misses and branches. The improved shakersort has
+a lot less branches, but the same number of misses. This is because skipping
+over sorted keys is predictable from the first branch miss.
 
@@
 
-Shakersort doesn't have an opt-out-early clause like bubblesort, and as a result
-would perform very badly for sorted sets. Adding this enhancement is simple, and
-would give better results, which would also be more consistent with bubblesort.
-
 It should be possible to make more cache-conscious versions of these sorts.

### 2005-05-26 paper/docs/quick.tex
make way for the new and improved fs

%\section{Description}

Quicksort is an in-place, unstable sort. It was first described in
\cite{HoarePartition61} and \cite{HoareQuicksort61}, and later in more detail in
\cite{Hoare62}. Sedgewick did his PhD on quicksort, and several important
improvements to quicksort are due to this. They are discussed in
\cite{Sedgewick78} and \cite{Sedgewick02}. \cite{BentleyMcIlroy93} discusses
considerations for a fast C implementation. \cite{SedgewickBentley02} discusses
equal keys and proves that quicksort has the lowest complexity of any comparison
based sort.

Quicksort is an algorithm of the type known as \n{divide and conquer}. It
recursively partitions its array around a pivot, swapping large keys on
the left hand side with smaller keys on the right hand side. The partition
causes the pivot to be put in its final position, with all greater keys to the
right, and lesser keys on the left. The left and right partitions are then
sorted recursively.

The recursive nature of quicksort means that even though quicksort is in-place,
there is a requirement of a stack. In the worst case, this stack can grow to
$O(N)$, if the largest or smallest key is always chosen as the pivot. This
worst case will occur frequently, especially when trying to sort an already
sorted, or almost sorted list.

Figure \vref{Simple quicksort implementation} contains a simple implementation of
the quicksort algorithm.

\code{Simple quicksort implementation}{simple_quicksort.c}

\section{Base Quicksort}

LaMarca uses the optimised quicksort described by Sedgewick for his base
quicksort. This is an iterative quicksort which uses an explicit stack. In the
worst case, this stack must be the same size as the array, and must be
initialised at the start of the algorithm. To reduce the size of the stack,
Sedgewick proposes sorting the smallest partition first. This optimisation
reduces the maximum size of the stack to $O(logN)$. In practice, two \cc{int}s,
the left and the right index, are required for each partition.

Sedgewick's second improvement is the use of an elementary sort for small
subfiles.  Once a partition is smaller than an arbitrary threshold, insertion
sort is used to sort the partition. Sedgewick states that roughly the same
results are obtained for thresholds between 5 and 25. An additional reduction in
instruction count can be had by doing the insertion sort over the entire array
at the end, rather than across each small array as soon as the threshold is
reached. At the end of the partition sorting, LaMarca places a sentinel at the
end of the array and uses an insertion sort, eliminating the bounds check from
the end. In this implementation, the sentinel is placed at the front, and the
smallest of the first \cc{threshold} keys is used as a sentinel, in keeping
with the policy of not exceeding the bounds of the array.

Choosing a pivot which is close to the median of the keys to be sorted ensures
quicksort achieves optimal performance. Hoare suggested a random pivot, but the
added cost of a random number generator would severely slow the algorithm.
Sedgewick recommends that the median of three keys is chosen, the three keys
being the first, middle and last in the list. This reduces the chance of a worst
case occurring significantly, notably in the case of sorted and almost sorted
lists.

An additional benefit of Sedgewick's median of three implementation is that it
provides a sentinel for the partition. It is now possible to remove the bounds
check from the innermost loops of \cc{partition}, saving a lot of instructions.

Between these three improvements, Sedgewick estimates a reduction in
instruction count of roughly 30\%. The size of the auxiliary stack is also
reduced significantly, and the possibility of stack overflow, in the worst case
of the recursive version, is removed.

Quicksort has been well studied, and several important improvements have been
noted. In \cite{BentleyMcIlroy93}, using a median of nine partition is discussed,
and modern discussions on quicksort refer to the problem of equal keys, which is
discussed there, as well as by Sedgewick in \cite{SedgewickBentley02}. These
improvements are not included here for several reasons. Firstly, a quicksort
which takes into account equal keys puts the equal keys at the front of the
array, then copies them into place later. These copies would represent a lot of
cache misses, and the results would overwhelm the improvements LaMarca makes in
the next phase of the algorithm. Secondly, it is not as important to write a
brilliant version of quicksort as it is to analyse LaMarca's improvements.

\section{Memory-Tuned Quicksort}
LaMarca's first step was to increase the temporal locality of the algorithm by
performing insertion sort as soon as the partitioning stops. Once a partition is
reduced below the size of the threshold, insertion sort is preformed
immediately. By doing this, the keys being sorted are in the cache already. By
leaving the insertion sort until the end, as Sedgewick does, the reduction in
instruction count is made up by the increase in cache misses, once the
lists being sorted are larger than the cache.

As discussed above, it is necessary to insert either a bounds check or a
sentinel.  If the partition to be sorted is the left most partition, then the
smallest key in the partition will be chosen as the sentinel. No other
partition needs a sentinel. A possible way to avoid the sentinel in that
partition is to place zero at $0^{th}$ position in the list to be sorted, and
sort whichever number was in this position back at the end. This sort could be
done with one iteration of insertion sort, and would have just one branch miss.
However, the number of level 2 cache misses would be great in this case.
However, the cost of this would be substantial.  This could be reduced by
choosing the key to remove from a small list, such as the eight keys at the
start, middle and end of the array. This would introduce a constant number of
cache misses, but would significantly reduce the chances of having to sort
across the entire array at the end. This, however, is not implemented; instead a
check is done to fetch a sentinel from the appropriate subfile if required. This
check is expensive, but less so than either a bounds check or putting a sentinel
into every subfile.

\section{Multi-Quicksort}

In the case that the array to be sorted fits in the cache, then quicksort has
excellent cache properties. Once it no longer fits in the cache, however, the
number of misses increases dramatically. To fix this problem, LaMarca uses a
technique similar to that used in multi-mergesort. Instead of multi-merging a
series of sorted lists at the end, a \n{multi-way partition} is done at the start,
moving all keys into cache sized containers which can be sorted individually.

At the start of the algorithm, a set of pivots is chosen and sorted, and keys
are put into containers based on which two pivots their value lies between. A
minimal and maximal key are put on the edges of this list to avoid bounds checks
as the list is iterated over. Because of the way the pivots were chosen, two
pivots which happen to be close together would result in a very small number of
keys being put into their container. While this is not a bad thing, this will
result in a larger number of keys in another container. Conversely, a large
space between two consecutive pivots will mean that the keys in this
container will not all fit into the cache at once.

There are two possible solutions to this. The first is to choose the pivots
well, so that they are very evenly distributed. The second is to choose the
number of pivots so that the average size of the containers is not larger than
the cache. LaMarca presents calculations showing that by choosing the average
subset size to be $C/3$, where $C$ is the number of keys which fit in the
cache\footnote{For a 2MB cache, this is 524,144.}.

Since the maximum number of keys which must fit in a container is not known, it
is necessary to implement containers using linked lists. Each list contains a
block of a few thousand keys. LaMarca notes very little performance difference
between 100 and 5,000 keys in each block. The size is therefore chosen so that
each list fits exactly into a page of memory. When a list is filled, another
list is linked to it, and these lists hold all the keys in the container.

The initial implementation of this allocated a new list every time an old one
was filled. This lowers performance significantly, as it results in a lot of
system calls, which increase instruction count and cache misses. Instead, it is
possible to work out the maximum possible number of lists required to create the
containers, and allocate them at the start. When a list runs out of space, it is
linked to an unused list from the large allocation. This wastes space, since it
is almost guaranteed that less lists will be required than available, though the
number of wasted lists allocated will be less than the number of pivots, and
therefore not exceptionally high.

The entire array is iterated though a key at a time. For each key, the list
of pivots are searched until the appropriate container is found, and the key is
added to that container. Sequentially searching the lists increases the
instruction count by 50\% over the memory-tuned version. For this reason, the
search was replaced with an efficient binary search, which reduced by half the
extra instructions.

Once the entire array has been put into containers, each container is emptied
in turn and the keys put back into the array. When a container is empty, the
pivot greater than that container is put into position, which is guaranteed to
be its final position.  The indices of these positions is pushed onto the
stack, and the subarray is then sorted.

While emptying the containers back into the array, an opportunity is taken to
find the smallest key in the leftmost partition. This is placed as a sentinel,
ensuring that every partition has a sentinel, and that no bounds checks are
required.

\section{Results}

\subsection{Test Parameters}
The threshold chosen was 10, simply as this is the number used in Sedgewick's
book.

The number of keys in the linked list is 1022 since a word is needed for the
count, and another as a pointer to the next list. 1024 is the number of 32-bit
words which fit in a 4KB page of memory. In earlier versions where each page was
\cc{malloc}ed individually, the number of keys in the linked list was 1018,
as 4 words were used by the particular version of \cc{malloc} used by
SimpleScalar\footnote{This particular \cc{malloc} forms part of
\cc{glibc}, the GNU C library, version 1.09.}.

A 2MB cache was assumed, so the number of keys which fit in the cache for this
in-place sort is 524,288. Multi-quicksort began once the number of keys was
larger than 174,762.

\subsection{Expected Performance}
The instruction count for each quicksort is expected to be $O(NlogN)$,
though not as high as the other algorithms due to the efficient inner loop.
The number of instructions for multi-quicksort should increase by about 50\% in
the case of the sequential sort, and 25\% in the case of the binary sort.

LaMarca notes that quicksort is very cache efficient already. Once it no longer
fits in the cache, however, the number of misses should increase significantly.
This is especially true in the case of the base quicksort, which does an
insertion sort across then entire array at the end.

The number of branches should increase in multi-quicksort along with the
instruction count. When a binary search is used, the increase in misses is
expected to be significant. With a sequential search, however, the number of
misses should not increase at all, and should flat-line from that point on.

\subsection{Simulation Results}

\plot{quick}{}{}{Results shown are for a Direct Mapped cache.}{}{}{}{}

The instruction count, shown in Figure \ref{quick instructions} is as
predicted. The expected increases in instruction count for multi-quicksort was
due to a test with a smaller data set and smaller constants. The actual results
were higher than expected: being 63\% and 40\% greater, for
sequential and binary search versions, respectively.
%TODO check

The cache results in Figure \ref{quick level 2 misses} were similar to
predictions, except that the multi-quicksorts did not behave as expected, and
performed worse than the memory-tuned version. Not only do they have more
misses per key, but the rate at which the misses increase with size is also
worse for the multi quicksorts. The tuned version performed better than the
base version, as predicted.

A surprising result is that direct mapped caches often perform better than fully
associative ones. The reason for this is not clear, but it is obvious from the
graph that this is true in the case of all the quicksorts.

The branch prediction results in Figure \ref{quick branch misses} were exactly
as expected. The binary search produced a dramatic increase in the branch
misses, though not when the two-level adaptive predictor was used. A short spike
was recorded in the bimodal results, and its not clear if this is an anomaly or
not. If it is, then the two-level adaptive predictor performs worse than the
bimodal, yet again.

The sequential search stops the increase of branch misses, since every key will
only take one miss to find its proper place. The results of this on cycle count,
shown in Figure \ref{quick cycles} are enough to make up for the increase in
instruction count, which was significantly reduced by using the binary search.
The tuned quicksort, however, had the lower running time and even base
quicksort out-ran the multi-quicksorts.

LaMarca's results are again similar to the results presented here, but not
identical. The shape of the instruction count graph is the same in both cases,
but that of the cache misses is not. The ratio of the cache misses between base
and cache quicksort are the same in both sets of results, but the number of
cache misses in multi-quicksort is much higher here than it is in LaMarca's
findings. In addition, the shape is different. LaMarca's multi-quicksort's
cache misses plateau, while the results here show an increasing slope.

A previous result, however, with a different implementation showed the same
shape, but different results to LaMarca's. That implementation waited until all
the partitioned keys were put back into the array, instead of being sorted
straight away. Those results plateaued once mutli-quicksort took effect.

\section{Predictor results}
\sixplot{base_quicksort-0.eps}{base_quicksort9-0.eps}{cache_quicksort-0.eps}{multi_quicksort-0.eps}{sequential_multi_quicksort-0.eps}{}{base, cache and multi quicksorts and base quicksort9}
\sixplot{base_quicksort1-0.eps}{base_quicksort-0.eps}{base_quicksort5-0.eps}{base_quicksort7-0.eps}{base_quicksort9-0.eps}{}{base quicksort without a median, with median of 3, and with pseudo median of 5, 7 and 9}

%TODO a bit to rewrite

Theres two important bits: comparison between the types of sorts, and comparisons between the different number of things tried as medians.

Firstly, cache quicksort and base quicksort only idffer in how they deal with
the insertion. Base quicksort and base quicksort9 have similar numbers of
insertion branches, and a very similar ratio (2/1 correct/incorrect) of
mispredictions. Comparing base quicksort to cahce quicksort (they both use
median of 3), there is actually a decrease in the number of branches handled by
insertion sort, instead of an increase. An increase is expected (thats why
sedgewick moves the insertion to the end), though when you take into account
the sentinel and the outer loop, there probably is an increase in instructions
at least. Base quicksort has 35\% miss rate, but cache has a 34\%. Not a big
difference, but the point is that there is a reduction in branches, but of the
reduced branches, 50\% were unpredictable. I dont think theres anything
interesting in this fact, however.

When going to multi-quicksort, there si quite a drop in the number fo branches.
Once the sort is larger than the cache, the number fo steps increases linearly
(since it does a sort of counting like radix sort). Theres a decrease of 19 \%.
Since the sort at this point is NlogM where M is the size of the cache (in
keys), there shouldnt be a change in misprediction rates. And there sint.
Before: 63\%. After, 63\%. So no difference.

Theres a lot of branches in the multi quicksort. WE saw in other results that
the mere instruction cost of this bogs down to algorithm significantly. in the
binary search version, there are more than twice as many mispredictions,
despite the fact that there are less than half the branches. (note that i dont
think this was a wonderful binary search, its heavily biased towards the
checking the left). binary search predicts correctly 57 percent of the time,
meaning that it mostly changes direction haphazardly during the search. No
wonder it performs so badly.

Note that sequential multi-quicksort has less mispredictions than cache quicksort.

Part 2: comparing different ways of randomly partitioning:

Ways of getting the partition:
1 - just take the rightmost element. Note that there are no sentinels, so you have to put a bounds check back in
3 - median of(0,1,2) halves
5 - median of(0, (median of(1,2,3), 4) quarters
7 - median of(median of(0,1,2), 3, median of(4,5,6)) sixths
9 - median of(median of(0,1,2), median of(3,4,5), median of(6,7,8)) eights

misprediction rates:
1 - 29.4\%
3 - 35.9\%
5 - 38.4\%
7 - 40.0\%
9 - 41.7\%

The misprediction rates point to the problem. They're just not being evenly divided. The number of times the partitions end is the same, roughly, each time. However, the number of actual branches is much higher in the case of quicksort1, and decreases all the way down. At the same time, the number of mispredictions creeps up. And hence the big change in the rate of mispredictions. As predicted, the better the median, the better the partition, and the less correct predictions pretending to be useful. There must be a 50\% rate waiting to get out, so a little more than median of 9 maybe. Or a lower cut-off point (it cuts out for partitions of less than 40 keys, and uses plain old median of 3).

the rate or number of insertion branches doesnt change at all.

%TODO up to here

\section{Future Work}
The threshold in quicksort is the point at which a simpler sort is used to sort
the partition. A larger threshold will results in fewer checks for a sentinel. A
smaller threshold means that the instruction count due to the insertion sort
will be lower, but more partition steps may be taken. A smaller threshold also
makes it more likely that the keys to be sorted all fit in a single cache
block. However, the partition is likely to occur across a cache block boundary,
though this is not necessarily a bad thing; the next partition to be sorted will
be adjacent to the current block, and will already be in memory. Observing how
results, especially instruction count and branch misses, vary with changes 
to the threshold may be interesting.

Two areas of current research into quicksort are those of equal keys, and
median-of-greater-than-3. The former affects the algorithmic properties, and
some steps may need to be varied to avoid extra cache misses as a result
of extra copying. The latter simply affects the instruction count, though the
extra comparisons are likely to be mispredicted. These results could be
measured, giving greater insight to this debate.

The memory-tuned quicksort is the only algorithm here that requires a sentinel
check. The branch predictor will only mispredict this once, so the instruction
count increase is only that of removing a single check which occurs relatively
few ($N / THRESHOLD$)\footnote{This is a minimum. The figure may be higher since
partitions are likely to be smaller than the threshold.} times. Still, the
problem is interesting and has applications in other algorithm designs.

Finally, in multi-quicksort, it may be possible to choose better pivots.
Choosing better (ie, more evenly distributed pivots) would reduce the number of
containers. The advantage of this is that it reduces linearly the number of
pivots to be searched by the sequential search. The most this can be reduced by
is a factor of 3, indicating perfectly distributed pivots. This would reduce the
instruction overhead of the multi-quicksort considerably.

### 2005-05-26 paper/docs/radix.tex
make way for the new and improved fs

--- 2005-05-26 docs/radix.tex
@@
 
+\subsection{Cache Radixsort}
+Many improvements exist to radixsort, and some are suggested by Sedgewick and
+LaMarca. Some of these improvements help to lower cache misses and instruction
+count, both of which slow down radixsort significantly. LaMarca suggested
+that while the current byte is being put in place, the counting of the next byte
+can be done. This increases the memory requirements, but this is not
+significant.
+
+The first improvement is to reduce the number of runs across the array per
+step.  In the \n{base radixsort}, presented above, there is one step to count,
+one to copy, and one to write back (which accesses both arrays). LaMarca
+suggests counting for the next step in the previous count. In fact, all the
+counting can be done in a single pass at the start, instead of four passes at
+various stages, decreasing instruction count and cache misses. The downside is
+the fourfold increase in memory required for the arrays of counts, with no
+benefits.
+
+The step to copy the array back can also be removed. In mergesort, copying back
+to the source array was changed to copying back and forth between the source and 
+auxiliary arrays. The same could be potentially done with radixsort, without any
+increase in complexity. The number of steps is likely to be even, and is known
+at compile time, so the number of times the copy step is done could be reduced
+to zero. If the number of steps is odd, which is unlikely when sorting
+integers, then one copy at the end would be required. This is not needed in the
+implementation presented here.
+
+Cache radixsort should traverse the array only nine times: Once at the start to
+count the least significant bits; six times as it copies using the three least
+significant bytes (counting the next least significant byte at the same time),
+and a final two as it copies using the most significant byte.
+
 \section{Results}
 \subsection{Test parameters}
-Keys are to be divided into segments 8 bits long, resulting in four passes of
-the array. The radix used is therefore $2^8 = 256$.
+Keys are to be divided into segments 8 bits long, resulting in four sorting
+steps to completely sort the array. The radix used is therefore $2^8 = 256$.
 
@@
 Radixsort is expected to have very good performance, compared to comparison
-based sorts. This is due to that fact that radixsort is $O(N)$; for an array
-as small as 4096 records, it involves 12 times fewer steps than the $O(NlogN)$ sorts.
+based sorts. This is due to that fact that radixsort is $O(N)$; for an array as
+small as 4096 records, it involves 12 times fewer steps than the $O(NlogN)$
+sorts.
 
-The cache performance of radixsort is not expected to be good, however.
+The cache performance of base radixsort is not expected to be good, however.
 Quicksort and mergesort, for example, have very good temporal locality. Keys are
@@
 
+Due to the reduced copying, cache radixsort should have a lower instruction
+count and fewer cache misses that base radixsort. There should be nine passes
+over the array, leading to just over one miss per key, as opposed to two misses
+per key with base radixsort. The complexity and branch prediction results
+should remain the same.
+
 \subsection{Simulation results}
@@
 
-The instruction count of radixsort, shown in Figure \ref{radix instructions} is
+The instruction count of base radixsort, shown in Figure \ref{radix instructions} is
 very low. It is lower than even quicksort for the smallest set, and it
 flat-lines immediately (after amortisation of the loop overhead). The number of
-cache misses in Figure \ref{radix level 2 misses} is exactly the same, until the
-array no longer fits into the cache. The number of misses then leaps up to 2.5
-per
-key, which is roughly what was predicted.
+cache misses in Figure \ref{radix level 2 misses} is exactly the same, until
+the array no longer fits into the cache. The number of misses then leaps up to
+2.5 per key, which is roughly what was predicted.
+
+Cache radixsort performs as predicted aswell. By sorting keys between two
+arrays instead of copying a partially sorted array back, the instruction count
+has been reduced by more than 10\%, and the number of level 2 cache misses
+have been more than halved, to just over one miss per key.
 
@@
 sorts considered had nearly that number of misses. However, it is not
-unexpected.
+unexpected. The number of branches reduces when copying back is removed, and
+while this reduces the number of mispredictions, there are so few to begin with
+that this result is not significant.
 
@@
 never increase beyond 350 cycles per key, meaning that for larger sets it will
-perform significantly better than quicksort. The reason for the high number of
-cycles is that radixsort was not optimised. Possible optimisations are discussed
-below, and should significantly reduce the number of cache misses, the main
-cause of radixsorts poor performance.
+perform significantly better than quicksort. Cache radixsort never increases
+beyond 300 cycles, though it has relatively more cycles when the arrays are
+smaller.
+
+The most curious result is that all simulations show that cache radixsort
+performing better than the base radixsort, yet the Pentium 4 results show that
+base radixsort performs better for smaller arrays.
 
@@
 \label{radix future work}
-Many improvements exist to radixsort, and some are suggested by Sedgewick and
-LaMarca. Some of these improvements help to lower cache misses and instruction
-count, both of which slow down radixsort significantly. LaMarca suggested
-that while the current byte is being put in place, the counting of the next byte
-can be done. This increases the memory requirements, but this is not
-significant.
-
-The first improvement is to reduce the number of runs across the array per step.
-At the moment, there is one step to count, one to copy, and one to write back
-(causing potentially two misses per write). LaMarca suggests counting for the next
-step in the previous count. In fact, all the counting can be done in a single
-pass at the start, instead of four passes at various stages, decreasing instruction
-count and cache misses. The downside is the fourfold increase in memory required
-for the arrays of counts, but this is small compared to the main and auxiliary
-arrays.
-
-The step to copy the array back can also be removed. In mergesort, copying back
-to the source array was changed to copying back and forth between the source and 
-auxiliary arrays. The same could be potentially done with radixsort, without any
-increase in complexity. The number off steps in likely to be even, and is known
-at compile time, so the number of times the copy step is done could be reduced
-to zero. If the number of steps is odd, which is exceptionally unlikely when
-sorting integers, then one copy at the end would be required.
-
 Combining other sorts with radixsort could also be possible. An optimisation to
@@
 
-Additional performance improvements could possibly be gained using a different
-radix. This radix was chosen simply because Sedgewick used it. Another radix may
-lead to better results.
-
 The radixsort presented above is a Least Significant Digit radixsort. A Most

### 2005-05-26 paper/docs/result_list.tex
make way for the new and improved fs

\afterpage{
\smallplot{on2_insertsort_}{insertion sort}{}{}{}{}{}
\smallplot{on2_selectsort_}{selection sort}{}{}{}{}{}
\smallplot{on2_bubblesort_}{bubblesort}{}{}{}{}{}
\smallplot{on2_bubblesort2_}{bubblesort2}{}{}{}{}{}
\smallplot{on2_shakersort_}{shakersort}{}{}{}{}{}
\smallplot{on2_shakersort2_}{shakersort2}{}{}{}{}{}
%
\smallplot{base_heapsort_}{base heapsort}{}{}{}{}{}
\smallplot{cache4_heapsort_}{cache heapsort (4-heap)}{}{}{}{}{}
\smallplot{cache4_heapsort_small_cache_line}{cache heapsort (4-heap) (with 16-byte cache line)}{}{}{}{}{}
\smallplot{cache_heapsort_}{cache heapsort (8-heap)}{}{}{}{}{}
%
\smallplot{algorithm_n}{algorithm N}{}{}{}{}{}
\smallplot{algorithm_s}{algorithm S}{}{}{}{}{}
\smallplot{base_mergesort_}{base mergesort}{}{}{}{}{}
\smallplot{double_tiled_mergesort_}{double tiled mergesort}{}{}{}{}{}
\smallplot{double_multi_mergesort_}{double multi-mergesort}{}{}{}{}{}
%
\smallplot{base_quicksort_}{base quicksort}{}{}{}{}{}
\smallplot{cache_quicksort_}{memory-tuned quicksort}{}{}{}{}{}
\smallplot{multi_quicksort_}{multi-quicksort (binary search)}{}{}{}{}{}
\smallplot{multi_quicksort_seq_}{multi-quicksort (sequential search)}{}{}{}{}{}
%
\smallplot{base_radixsort_}{base radixsort}{}{}{}{}{}
\smallplot{cache_radixsort_}{cache radixsort}{}{}{}{}{}
%
\smallplot{gonnet_shellsort_}{gonnet shellsort}{}{}{}{}{}
\smallplot{improved_gonnet_shellsort_}{improved gonnet shellsort}{}{}{}{}{}
}

### 2005-05-26 paper/docs/shell.tex
make way for the new and improved fs

\label{Shellsort}

Shellsort is a derivative of insertion sort designed to offer a compromise
between the performance of mergesort and radixsort, and the space efficiency of
selection sort, bubblesort and insertion sort. It was designed by Donald Shell
in 1959, and described in \cite{Shell59}\footnote{Shell describes the former
sorts as ``Merging Pairs'' and ``floating decimal sort'', respectively, and the
latter sorts as ``finding the smallest'', ``interchanging pairs'' and ``sifting'',
respectively.}.
%TODO whats partial sort?

Shellsort begins by considering the array to be made up of $N/2$ subarrays of
length two, where the elements of the subarray are spaced $N/2$ keys apart.
These subarrays are then sorted using insertion sort. The first iteration will
be a simple exchange, but the keys which are exchanged move across the array by
$N/2$ spaces. The sort then considers $N/4$ arrays, who's elements are spaced
$N/4$ keys apart, which are individually sorted by insertion sort, then $N/8$
and so on. The final iteration will be a standard insertion sort, after which
the array will be sorted. A problem with this implementation is that since the
increments are powers of two, keys with odd indices are not compared with keys
with even indices until the final insertion sort. This can lengthen the final
insertion sort's running time to $O(N^2)$.

Since first published, there has been much discussion about the best
\textit{increments} to use in shellsort. Shell's original sort, described above,
has the increments $N/2$, $N/4$, $N/8$, $N/16$ ........ $1$. \cite{Sedgewick96}
discusses many increments used, and provides code for comparing these. Based
on running the \textit{driver} program\footnote{see
http://www.cs.princeton.edu/\textasciitilde{}rs/shell/}, Gonnet's shellsort was
seen to be the fastest on the authors local machine. As a result, this was used
to test the performance of shellsort. This uses increments which are $5/11$ths
of the previous increment.
%TODO do we refer to increments from the top or bottom

Two versions of Gonnet's sort were used. The first was taken directly from the
code presented by Sedgewick, and only slightly modified to fit into the framework used.
The second was modified to reduce the complexity of the loop bounds, and to
remove a sentinel check in the inner loop. Code for these is shown in Figures
\ref{Gonnet Shellsort code} and \ref{Improved Gonnet Shellsort code}.

\code{Gonnet Shellsort code}{old_shellsort.c}
\code{Improved Gonnet Shellsort code}{shellsort.c}

\section{Results}
\subsection{Test parameters}
The increments in Gonnet's shellsort are $N*(5/11)$, $N*(5/11)^2$, $N*(5/11)^3$,
...... $1$.

\subsection{Expected Performance}
Shellsort has been much analysed, but very little is known about the complexity
of the algorithm. The instruction count is expected to be $O(NlogN)$, as this
has been indicated by previous experient. The cycle count is expected to be of
the same order of magnitude as the $O(NlogN)$ sorts for the same reason.
 %TODO find some experiements that showed this.

The cache performance of shellsort should be regular. Each subarray is sorted in
turn without reference to the next. Since each element of a subarray is quite
far from its next element, an entire cache line must be loaded for each key,
which will be used once per iteration of the sort across the subarray. If there
are a large number of subarrays, as during the iterations with large increments,
there will be no spatial reuse, and little temporal reuse. Once the increment is
small enough that each subarray fits within the cache (ie quickly for L2 cache,
and much more slowly for the L1 cache), then temporal reuse will occur. Once an
entire subarray fits within an eighth of the size of the cache (eight being the
number of keys that fit in a cache line), then spatial reuse will occur aswell.
This reuse will only occur within a single iteration of shellsort. Once the next
increment begins, nothing in the cache is likely to be valid, unless the entire
array fits within the cache.

The branch performance of shellsort should be predictable. There are
$log_{5/11}N$ iterations, each with $N$ branch misses\footnote{See Section
\ref{insertion is predictable} for a discussion of the properties of insertion
sort.}. It is expected, therefore, that there will be $log_{5/11}N$ misses per
key.

\subsection{Simulation results}
\plot{shell}{}{}{}{}{}{}

Shellsort's simulation results have the characterist shape and range of an
$O(NlogN)$ sort, albeit, not a very good one. It performs better than a
heapsort, but considerably worse than quicksort. However, it easier to program
than either of these two sort.

Shellsort performs roughly double the number of branches and instructions of
quicksort, and has more than twice as many branch mispredictions and nearly
four times as many cache misses. It's cycle count is over three times that of
quicksort, though it is more than 50 percent faster than the best of the
heapsort.

\section{Predictor results}

\sixplot{gonnet_shellsort-0.eps}{gonnet_improved_shellsort-0.eps}{}{}{}{}{Shellsorts}

%TODO better writing

In the unimporved version, half the bracnhes are mispredicted. Assuming taken would provide better results. Each key moves only 1.03 spots on average. Once the insertion sort is seperated, we see that this is because the insertion sort tips it in favour of 1 spot, and in actual fact this should be 1.07 spots, on average, though the difference is not very significant. The improved insertion sort is better behaved, moving each key on average two spots. But the prediction ratio is much worse than this. There should be N mispredictions, and instead there are 1.3xN mispredictions. THis is because the counter doesnt really get time to settle properly.

%TODO end here

\section{Future Work}
Only Gonnet's way of producing increments is measured here. Using different
increments and different schemes for generating increments would produce
different results.

### 2005-05-26 paper/docs/software_predictor.tex
make way for the new and improved fs

\afterpage{
\sixplot{base_heapsort-0.eps}{cache4_heapsort-0.eps}{cache4_heapsort-1.eps}{cache_heapsort-0.eps}{cache_heapsort-1.eps}{}{base and cache heapsorts}

\sixplot{base_quicksort-0.eps}{base_quicksort9-0.eps}{cache_quicksort-0.eps}{multi_quicksort-0.eps}{sequential_multi_quicksort-0.eps}{}{base, cache and multi quicksorts and base quicksort9}
\sixplot{base_quicksort1-0.eps}{base_quicksort-0.eps}{base_quicksort5-0.eps}{base_quicksort7-0.eps}{base_quicksort9-0.eps}{}{base quicksort without a median, with median of 3, and with pseudo median of 5, 7 and 9}

\sixplot{algorithm_n-0.eps}{algorithm_n-1.eps}{algorithm_s-0.eps}{algorithm_s-1.eps}{}{}{Knuth's Mergesorts}
\sixplot{base_mergesort-0.eps}{base_mergesort-1.eps}{base_mergesort-2.eps}{base_mergesort-3.eps}{base_mergesort-4.eps}{base_mergesort-5.eps}{base mergesort}
\sixplot{double_tiled_mergesort-0.eps}{double_tiled_mergesort-1.eps}{double_tiled_mergesort-2.eps}{double_tiled_mergesort-3.eps}{double_tiled_mergesort-4.eps}{double_tiled_mergesort-5.eps}{double tiled mergesort - part 1}
\sixplot{double_tiled_mergesort-6.eps}{double_tiled_mergesort-7.eps}{double_multi_mergesort-0.eps}{double_multi_mergesort-1.eps}{double_multi_mergesort-2.eps}{double_multi_mergesort-3.eps}{double tiled mergesort - part 2, and double multi mergesort - part 1}
\sixplot{double_multi_mergesort-4.eps}{double_multi_mergesort-5.eps}{double_multi_mergesort-6.eps}{double_multi_mergesort-7.eps}{double_multi_mergesort-8.eps}{double_multi_mergesort-9.eps}{double multi-mergesort - part 2}

\sixplot{gonnet_shellsort-0.eps}{gonnet_improved_shellsort-0.eps}{}{}{}{}{Shellsorts}
}

### 2005-05-26 base_heapsort.c

@@ -1,0 +1,185 @@
+#include "branch_sorts.h"
+
+#include <stdio.h>
+
+#ifdef _USE_SOFTWARE_PREDICTOR
+static struct predictor* up_down0;
+static struct predictor* up_down1;
+static struct predictor* up_down2;
+#endif
+
+static void
+fixDown(unsigned int a[], int parent, int N)
+{
+	int child;
+	unsigned int v = a[parent];
+	while(1) /* see if it has any children */
+	{
+		if (2*parent <= N) branch_taken(up_down0);
+		else
+		{
+			branch_not_taken(up_down0);
+			break;
+		}
+		child = 2*parent;
+
+		/* its important that this be less for my improvement */
+		if (less(a[child], a[child+1])) /* see if it has more than 1 child - check which child is larger */
+		{
+			branch_taken(up_down1);
+			child++;
+		}
+		else branch_not_taken(up_down1);
+
+		if (!less(v, a[child])) /* when the larger child isnt large enough to be promoted, stop */
+		{
+			branch_taken(up_down2);
+			break;
+		}
+		else branch_not_taken(up_down2);
+	
+		/* move down */ 
+		a[parent] = a[child];
+
+		parent = child;
+	}
+	
+	a[parent] = v;
+
+}
+
+/*
+static int
+is_a_heap(unsigned int a[], int l, int r)
+{
+	int N = r-l+1;
+	int k;
+	for(k = 1; k <= N/2; k++)
+	{
+		int k_pos = k - 1;
+		int l_pos = k * 2 - 1;
+		int r_pos = k * 2;
+		int l = l_pos + 1;
+		int r = r_pos + 1;
+
+		if (a[k_pos] < a[l_pos])
+		{
+			printf("FAIL: node %d (a[%d]=%d) is less than left child %d (a[%d]=%d)\n", k, k_pos, a[k_pos], l, l_pos, a[l_pos]);
+			return 0;
+		}
+		if (a[k_pos] < a[r_pos])
+		{
+			printf("FAIL: node %d (a[%d]=%d) is less than right child %d (a[%d]=%d)\n", k, k_pos, a[k_pos], r, r_pos, a[r_pos]);
+			return 0;
+		}
+	}
+	return 1;
+}*/
+
+
+void
+base_heapsort(unsigned int a[], int N)
+{
+	int k;
+
+	int is_even = !(N & 1);
+
+#ifdef _USE_SOFTWARE_PREDICTOR
[…]

### 2005-05-26 base_mergesort.c

@@ -1,0 +1,482 @@
+#include <stdlib.h>
+#include <string.h>
+#include <stdio.h>
+
+#include "base_sorts.h"
+
+void
+base_mergesort(unsigned int a[], int N)
+{
+	unsigned int* aux = malloc(N * sizeof(unsigned int)); /* make it twice the size to use the notation */
+
+	int i,j; /* indices for the first array */
+	int k; /* indices for second array */
+	
+	int d; /* d => direction, if (f == 0) keep going */
+
+	int next_count;
+
+	int track; /* this keeps track of i so we know where to merge next */
+	/* an alternative would be to add count/2 and clear the lsb. */
+
+	unsigned int* temp_pointer;
+	unsigned int temp;
+	int min;
+
+	unsigned int* source = a;
+	unsigned int* target = aux;
+
+	describe_predictor(&global_predictor[0], "presort 0");
+	describe_predictor(&global_predictor[1], "presort 1");
+	describe_predictor(&global_predictor[2], "presort 2");
+	describe_predictor(&global_predictor[3], "presort 3");
+	describe_predictor(&global_predictor[4], "presort 4");
+	describe_predictor(&global_predictor[5], "presort 5");
+	describe_predictor(&global_predictor[6], "presort 6");
+	describe_predictor(&global_predictor[7], "presort 7");
+	describe_predictor(&global_predictor[8], "presort 8");
+	describe_predictor(&global_predictor[9], "presort 9");
+	describe_predictor(&global_predictor[10], "presort 10");
+	describe_predictor(&global_predictor[11], "presort 11");
+	describe_predictor(&global_predictor[12], "middle");
+	describe_predictor(&global_predictor[13], "next");
+	describe_predictor(&global_predictor[14], "end");
+	describe_predictor(&global_predictor[15], "setup");
+	describe_predictor(&global_predictor[16], "equal");
+#ifdef _USE_ROLLED_LOOPS
+	describe_predictor(&global_predictor[17], "left");
+	describe_predictor(&global_predictor[18], "right");
+#else
+	describe_predictor(&global_predictor[17], "left 0");
+	describe_predictor(&global_predictor[18], "left 1");
[…]

### 2005-05-26 base_quicksort9.c

@@ -1,0 +1,171 @@
+#include <stdlib.h>
+#include <stdio.h>
+
+#include "base_sorts.h"
+#include "quick.c"
+
+#define THRESHHOLD 14
+
+
+static void
+stackinit(unsigned int N)
+{
+
+	int lg2N = 32; /* im hard coding this and i dont care */
+	unsigned int msbN = 1 << (lg2N - 1);
+	while(msbN > N)
+	{
+		lg2N--;
+		msbN = msbN >> 1;
+	}
+
+/*	printf("\nAllocating a stack %d ints big (lg2N == %d)\n", (lg2N + 2) << 2, lg2N); */
+	
+	stack = malloc(((lg2N + 2) << 2) * sizeof(int));
+}
+
+inline static void
+insertion_sentinel(unsigned int a[], int N)
+{
+	int i;
+	unsigned min;
+
+	min = 0;
+	for(i = 1; i < N; i++)
+	{
+		if(less(a[i], a[min]))
+		{
+			min = i;
+		}
+	}
+	exch(a[0], a[min]);
+
+}
+
+inline static void
+insertion(unsigned int a[], int N)
+{
+	int i;
+	unsigned int v;
+	for(i = 2; i < N; i++)
+	{
+		int j = i;
+		v = a[i];
+
+		while(less(v, a[j-1]))
+		{
+			branch_taken(&global_predictor[3]);
+			a[j] = a[j-1];
+			j--;
+		}
+		branch_not_taken(&global_predictor[3]);
+		a[j] = v;
+	}
+}
+
+
+inline static int
+med3(unsigned int array[], int l, int m, int r)
+{
+	unsigned int a = array[l];
+	unsigned int b = array[m];
+	unsigned int c = array[r];
+	if (a < b)
+	{
+		if (b < c) return m;
+		else if (a < c) return r;
+		else return l;
+	}
+	else
+	{
+		if (c < b) return m;
+		else if (c < a) return r;
+		else return l;
+	}
+}
+
+/* this is bentley and mcilroys version */
+/*static char *med3(char *a, char *b, char *c, int (*cmp)())
+{ return cmp(a, b) < 0 ?
+(cmp(b, c) < 0 ? b : cmp(a, c) < 0 ? c : a)
+: (cmp(b, c) > 0 ? b : cmp(a, c) > 0 ? c : a);
+*/
+
+void
+base_quicksort9(unsigned int a[], int N)
+{
+	int l, r;
+	int i;
+	int m;
[…]

### 2005-05-26 cache4_heapsort.c

@@ -1,0 +1,230 @@
+#include "cache_sorts.h"
+
+#include <stdlib.h>
+#include <stdio.h>
+#include <string.h>
+#include <malloc.h>
+#include <limits.h>
+
+#undef HEAP_SIZE
+#define HEAP_SIZE 4 /* never set this to 1 */
+
+/* these days we index from 0 - N-1
+* the rightmost child is (parent + 1)*heapsize
+* the parent = (child - 1) / heapsize 
+*
+* actually, we go from leftchild, which is (parent * heapsize) + 1
+* parent = (child - 1) / heapsize
+*/
+
+/*
+static int
+check_heap_property(unsigned int a[], int N)
+{
+	int parent;
+	int i;
+	for(parent = 0; parent < N/HEAP_SIZE+1; parent++)
+	{
+		int l = parent*HEAP_SIZE + 1;
+//		printf("checking children of %d (a[parent] = %d\n", parent, a[parent]);
+		for(i = 0; i < HEAP_SIZE; i++)
+		{
+//			printf("checking child %d (a[%d] = %d)\n", HEAP_SIZE-i-1, r-i, a[r-i]);
+			if (l+i >= N)
+			{
+//				printf("finsihing\n");
+				break;
+			}
+			if (a[parent] > a[l+i])
+			{
+				printf("FAIL: node %d (%d) is greater than child %d (node %d (%d)) \n", parent, a[parent], i, l+i, a[l+i]);
+				return -1;
+			}
+		}	
+	}
+	return 0;
+}
+*/
+
+
+static void
+fix_up(unsigned int heap[], int child)
+{
+	int parent;
+
+	unsigned int v = heap[child];
+	while(less(v, heap[parent = ((child-1)/HEAP_SIZE)]))
+	{
+		branch_taken(&global_predictor[6]);
+		heap[child] = heap[parent];
+		heap[parent] = v;
+		child = parent;
+	}
+	branch_not_taken(&global_predictor[6]);
+}
+
+static void
+fix_down(unsigned int heap[], int parent, int N)
+{
+	int child;
+	int child2;
+	unsigned int v = heap[parent];
+/*	printf("fixdown when N = %d\n", N); */
+	while(1) /* see if it has any children */
+	{
+		if (parent*HEAP_SIZE+1 < N) branch_taken(&global_predictor[0]);
+		else
+		{
+			branch_not_taken(&global_predictor[0]);
+			break;
+		}
[…]

### 2005-05-26 docs/code/adt.h

@@ -1,0 +1,10 @@
+#define Item unsigned int
+#define key(A) (A)
+#define less(A,B) (key(A) < key(B))
+#define exch(A, B)
+do	{
+	unsigned int t = (A);
+ 	(A) = (B);
+	(B) = t;
+}
+while(0)

### 2005-05-26 docs/code/double_multi_mergesort.c

@@ -1,0 +1,101 @@
+void double_multi_mergesort(Item a[], int N)
+{
+	/* assume that code is merged into cache sized segments at this point.
+	 * This is now merged in one step, by the following code
+	 */
+	int K = (((N-1) >> (LIMIT_BITS))+1);
+
+	int heap_length = 0;
+	/* this changes to an even number (rounding up) */
+	indices = memalign(HEAP_SIZE, ((K+1)^1) * sizeof(int)); 
+	heap_data = memalign(HEAP_SIZE, (K+1) * HEAP_SIZE * sizeof(Item));
+	heap = heap_data + (HEAP_SIZE - 1);
+	memset(heap_data, 0xff, (K+1) * HEAP_SIZE * sizeof(Item));
+	last_added_data = memalign(HEAP_SIZE, (K+1) * sizeof(Item));
+	last_added_indices = memalign(HEAP_SIZE, (K) * sizeof(Item));
+
+	last_added = &last_added_data[1];
+	last_added_data[0] = 0;
+
+	/* we add 8 from each to the Q */
+	for(i = 0; i < K; i++) /* we do the last one ourselves */
+	{
+		/* this may cause k misses in the bimodal predictor. 
+		 * but k is so small thats its not worth the effort */
+		if (i & 1) /* this is a reverse list */
+		{
+			if (i == K-1) indices[i] = N-1; /* last mini-list */
+			else indices[i] = ((i+1) << LIMIT_BITS) - 1;
+
+			add_heap_line(&aux[indices[i]], heap, -1, heap_length, HEAP_SIZE);
+			indices[i] -= HEAP_SIZE;
+
+			last_added[i] = aux[indices[i]+1];
+			last_added_indices[i] = i;
+
+		}
+		else /* foward list */
+		{
+			indices[i] = i << LIMIT_BITS;
+			if (i == K-1)
+			{
+				int space_to_end = HEAP_SIZE;
+				if (N - indices[i] < HEAP_SIZE ) space_to_end = N - indices[i];
+
+				/* this makes a sort of sentinel ,which simplifies the logic below */
+				indices[i+1] = N-1; 
+				add_heap_line(&aux[indices[i]], heap, 1, heap_length, space_to_end); 
+				indices[i] += space_to_end;
+			}
+			else
+			{
+				add_heap_line(&aux[indices[i]], heap, 1, heap_length, HEAP_SIZE);
+				indices[i] += HEAP_SIZE;
+			}
+
[…]

### 2005-05-26 docs/code/multi_quicksort.c

@@ -1,0 +1,203 @@
+#define KEYS_PER_LIST 1018
+struct List
+{
+	unsigned int blocks[KEYS_PER_LIST];
+	struct List* previous;
+	int count;
+};
+
+static struct List **lists = NULL;
+static int list_count = 0;
+static void* list_memory = NULL;
+static struct List* next_available_list = NULL; 
+
+static void listinit(int number_of_pivots, int N)
+{
+	int i;
+	long num_lists_required = number_of_pivots + (N / KEYS_PER_LIST);
+	lists = malloc(sizeof(struct List*) * number_of_pivots);
+	list_memory = malloc(sizeof(struct List) * num_lists_required);
+	next_available_list = list_memory;
+
+	for(i = 0; i < number_of_pivots; i++)
+	{
+		lists[i] = next_available_list++;
+		lists[i]->count = 0;
+		lists[i]->previous = lists[i];
+	}
+	list_count = number_of_pivots;
+}
+
+static void listclear()
+{
+	if (lists == NULL) return;
+
+	free(list_memory);
+	free(lists);
+	lists = NULL;
+}
+
+/* these functinos are not symetric. you fill it first. 
+ * then you pop it. nothing more is required, or supported */
+static inline void
+add_to_list(unsigned int list_index, unsigned int key)
+{
+	struct List* list = lists[list_index];
+
+/*	printf("adding key %d to list %d\n", key, list_index); */
+	if (list->count == KEYS_PER_LIST) /* if we're full up */
+	{
+		struct List* old_list = list;
+		list = next_available_list++;
+
+		list->previous = old_list;
+
+		list->count = 0;
+		lists[list_index] = list;
+	}
+
+	list->blocks[list->count] = key;
+	list->count++;
+
+}
+
+/* an empty list has count 0. that is all anybody needs to know */
+static inline unsigned int 
+pop_list(unsigned int list_index)
+{
+	struct List* list = lists[list_index];
+	unsigned int result;
+
+	list->count--;
+	result = list->blocks[list->count];
+
+	/* if we've run on */
+	if (list->count == 0)
+	{
+		list = list->previous;
+		if (list != NULL) lists[list_index] = list;
+	}
+
+	return result;
+}
+
+
+
+void
[…]

### 2005-05-26 docs/code/on2_insertsort.c

@@ -1,0 +1,22 @@
+void insertion(Item a[], int N)
+{
+	// get a sentinel
+	int min = 0;
+	for(int i = 1; i < N; i++)
+		if(less(a[i], a[min])) min = i;
+	exch(a[0], a[min]);
+
+	// sort the array
+	for(int i = 1; i < N; i++)
+	{
+		int j = i;
+		Item v = a[i];
+
+		while(less(v, a[j-1]))
+		{
+			a[j] = a[j-1];
+			j--;
+		}
+		a[j] = v;
+	}
+}

### 2005-05-26 double_multi_mergesort.c

@@ -1,0 +1,1334 @@
+#include <stdlib.h>
+#include <string.h>
+#include <stdio.h>
+#include <malloc.h>
+#include <limits.h>
+
+#include "cache_sorts.h"
+
+#define LIMIT OUT_OF_PLACE_LIMIT
+#define LIMIT_BITS OUT_OF_PLACE_LIMIT_BITS
+
+#define ODD_COUNT 8
+#define EVEN_COUNT 4
+
+static void
+merge(unsigned int source[], int N, int starting_size, unsigned int target[]);
+
+static void
+merge_reverse(unsigned int source[], int N, int starting_size, unsigned int target[]);
+
+static void
+presort(unsigned int a[], int N);
+static void
+presort_flexible(unsigned int a[], int N);
+
+static int presort_count = 8;
+static int double_presort_count = 16;
+
+static void
+fix_up(unsigned int heap[], int child)
+{
+	int parent;
+
+	unsigned int v = heap[child];
+	while(less(v, heap[parent = ((child-1)/HEAP_SIZE)]))
+	{
+		branch_taken(&global_predictor[46]);
+		heap[child] = heap[parent];
+		heap[parent] = v;
+		child = parent;
+	}
+	branch_not_taken(&global_predictor[46]);
+}
+
+static void
+fix_down(unsigned int heap[], int parent, int N)
+{
+	int child;
+	int child2;
+	unsigned int v = heap[parent];
+/*	printf("fixdown when N = %d\n", N); */
+	while(1) /* see if it has any children */
+	{
+		if (parent*HEAP_SIZE+1 < N) branch_taken(&global_predictor[47]);
+		else
+		{
+			branch_not_taken(&global_predictor[47]);
+			break;
+		}
+		child = parent*HEAP_SIZE + 1; /* this gets the rightmost node */
+		child2 = child;
+
+		if (less(heap[child2 + 0], heap[child]))
+		{
+			child = child2 + 0;
+			branch_taken(&global_predictor[48]);
+		}
+		else branch_not_taken(&global_predictor[48]);
+		if (less(heap[child2 + 1], heap[child]))
+		{
+			child = child2 + 1;
+			branch_taken(&global_predictor[49]);
+		}
+		else branch_not_taken(&global_predictor[49]);
+		if (less(heap[child2 + 2], heap[child]))
+		{
+			child = child2 + 2;
+			branch_taken(&global_predictor[50]);
+		}
[…]

### 2005-05-26 double_tiled_mergesort.c

@@ -1,0 +1,1050 @@
+#include <stdlib.h>
+#include <string.h>
+#include <stdio.h>
+#include <malloc.h>
+
+#include "cache_sorts.h"
+
+#define LIMIT OUT_OF_PLACE_LIMIT
+#define LIMIT_BITS OUT_OF_PLACE_LIMIT_BITS
+
+#define ODD_COUNT 8
+#define EVEN_COUNT 4
+
+static void
+merge(unsigned int source[], int N, int starting_size, unsigned int target[]);
+
+static void
+merge_reverse(unsigned int source[], int N, int starting_size, unsigned int target[]);
+
+static void
+presort(unsigned int a[], int N);
+static void
+presort_flexible(unsigned int a[], int N);
+
+static int presort_count = 8;
+static int double_presort_count = 16;
+
+inline static void set_presort_count(int count)
+{
+	presort_count = count;
+	double_presort_count = count * 2;
+}
+/*
+static void
+check_merge_reverse(unsigned int a[], int N)
+{
+	int i;
+	for(i = 0; i < N-1; i++)
+	{
+		if (a[i+1] > a[i])
+		{
+			printf("bad reverse: A: %p, a[%u] = %u, a[%u] = %u\n", a, i, a[i], i+1, a[i+1]);
+			return;
+		}
+	}
+}
+static void
+check_merge(unsigned int a[], int N)
+{
+	int i;
+	for(i = 0; i < N-1; i++)
+	{
+		if (a[i] > a[i+1])
+		{
+			printf("bad forwards: A: %p, a[%u] = %u, a[%u] = %u\n", a, i, a[i], i+1, a[i+1]);
+			return;
+		}
+	}
+}
+*/
+
+static int get_count(const int N)
+{
+
+	int lg2N = 32; /* im hard coding this and i dont care */
+	unsigned int msbN = 1 << (lg2N - 1);
+	while(msbN >= N)
+	{
+		lg2N--;
+		msbN = msbN >> 1;
+	}
+	return lg2N;
+}
+
+void
+double_tiled_mergesort(unsigned int a[], int N)
+{
+	/* track keeps an eye on where i started last */
+	/* outer track keeps and eye on which cache_sized/2 segemnt i started in last */
+	int i,j; /* indices for the first array */
+	
+	int level1_count;
+	int level2_count;
+	int extra_level1_count = 0;
+	int extra_level2;
+	int final_extra;
+
+
[…]

### 2005-05-26 fastsort_perfex/branch_sorts.h

@@ -1,0 +1,13 @@
+#ifndef _BRANCH_SORTS_H_
+#define _BRANCH_SORTS_H_
+
+#include "utils.h"
+
+void branch_heapsort(unsigned int a[], int N);
+void branch_quicksort(unsigned int a[], int N);
+void branch_radixsort(unsigned int a[], int N);
+void branch_mergesort(unsigned int a[], int N);
+
+void multi_quicksort_seq(unsigned int a[], int N);
+
+#endif

### 2005-05-26 fastsort_perfex/main.c

@@ -1,0 +1,233 @@
+/***************************************************************************
+                          main.c  -  description
+                             -------------------
+    begin                : Sat Oct 11 2003
+    copyright            : (C) 2003 by Paul Biggar
+    email                : biggarp@tcd.ie
+ ***************************************************************************/
+
+#include <stdio.h> // fopen, fgets, fclose
+#include <errno.h> // errno
+#include <assert.h> // assert
+#include <string.h> // strerror
+#include <stdlib.h> // atoi, malloc, free
+#include <sys/types.h> // time
+#include <sys/time.h> // time
+
+#include "on2_sorts.h"
+#include "base_sorts.h"
+#include "branch_sorts.h"
+#include "cache_sorts.h"
+#include "visual_sorts.h"
+#include "utils.h"
+
+#define RANDOM_SIZE (4194304)
+//#define RANDOM_SIZE (534523)
+//#define RANDOM_SIZE (262144)
+//#define RANDOM_SIZE (4096)
+//#define RANDOM_SIZE (128)
+
+#define RUN_VISUAL 0
+#define VISUAL_SIZE (41)
+#define SEED 0
+
+unsigned int sorted_array[RANDOM_SIZE];
+unsigned int random_array[RANDOM_SIZE];
+unsigned int random_array2[RANDOM_SIZE];
+
+FILE* visual_log = NULL;
+
+static struct timeval t1;
+static struct timeval t2;
+
+
+static inline long
+get_milliseconds(struct timeval *tp1, struct timeval *tp2)
+{
+	return ( (tp2->tv_sec - tp1->tv_sec) * 1000000) + (tp2->tv_usec - tp1->tv_usec);
+}
+
+static void inline 
+start_timer()
+{
+	if (gettimeofday(&t1, NULL) != 0)  printf("Bad result: \"%s\"", strerror(errno));
+}
+
+static void inline
+stop_timer()
+{
+	if (gettimeofday(&t2, NULL) != 0) printf("Bad result: \"%s\"", strerror(errno));
+}
+
+static void inline
+print_timer(const char* description)
+{
+	printf("%s: %ld ticks\n", description, get_milliseconds(&t1, &t2));
+}
+
+static void inline
+test_array(const char* description)
+{
[…]

### 2005-05-26 fastsort_perfex/multi_quicksort.c

@@ -1,0 +1,415 @@
+#include <stdlib.h>
+#include <stdio.h>
+#include <string.h>
+#include <limits.h>
+
+#include "cache_sorts.h"
+
+#define THRESHHOLD 10
+
+static int* stack = NULL;
+static int stack_index = 0;
+
+/* C is the number of keys that fit in the cache */
+#define C IN_PLACE_LIMIT
+#define C_DIV_3 (C / 3)
+static void
+stackinit(unsigned int N)
+{
+
+	int lg2N = 32; /* im hard coding this and i dont care */
+	unsigned int msbN = 1 << (lg2N - 1);
+	while(msbN > N)
+	{
+		lg2N--;
+		msbN = msbN >> 1;
+	}
+
+	if (N > 2 * C) /* make space for all the pivots being pushed */
+	{
+		lg2N +=  N / C_DIV_3;
+	}
+
+/*	printf("\nAllocating a stack %d ints big (lg2N == %d)\n", (lg2N + 2) << 2, lg2N); */
+
+	/* +2 for initial values, *2 cuase we push 2 ints a time */
+	/* we only really need about 3 less, cause of the threshhold, but this doesnt hurt */
+	stack = malloc(((lg2N + 2) * 2) * sizeof(int));
+}
+
+inline static void
+push(int valueA, int valueB)
+{
+	stack[stack_index++] = valueA;
+	stack[stack_index++] = valueB;
+}
+
+inline static int 
+pop()
+{
+	stack_index--;
+	return stack[stack_index];
+}
+
+inline static void
+stackclear()
+{
+	if (stack != NULL) free(stack);
+	stack = NULL;
+}
+
+inline static int
+stackempty()
+{
+	return (stack_index==0);
+}
+
+
+
+static int
+partition(unsigned int a[], int l, int r)
+{
+	unsigned int v;
+	int i, j;
+
+	v = a[r];
+	i = l - 1;
+	j = r;
+
+	for(;;)
+	{
+		while (less(a[++i], v))
+			;
+		while (less(v, a[--j]))
+			;
+		
+
+		if (i >= j)
+			break;
+
+		exch(a[i], a[j]);
+	}
+
+	exch(a[i], a[r]);
+
+	return i;
+}
+
+/*static void
+print_stack()
+{
+	printf("\nThe stack is: ");
+	for(int i = 0; i < stack_index; i++)
+	{
+		printf(" %d, ", stack[i]);
+	}
+}
+*/
+
+/* this is included so that we can replace the < N with <= N
[…]

### 2005-05-26 fastsort_perfex/on2_bubblesort.c

@@ -1,0 +1,24 @@
+#include "on2_sorts.h"
+#include <stdio.h>
+
+void
+on2_bubblesort(unsigned int a[], int N)
+{
+	int i, j;
+	int sorted;
+
+	for (i = 0; i < N-1; i++)
+	{
+		sorted = 1;
+		for (j = N-1; j > i; j--)
+		{
+			if (less(a[j],a[j-1]))
+			{
+				exch(a[j-1],a[j]);
+				sorted = 0;
+			}
+			
+		}
+		if (sorted) return;
+	}
+}

### 2005-05-26 knuth_other_base_mergesort.c

@@ -1,0 +1,168 @@
+#include <stdlib.h>
+#include <string.h>
+#include <stdio.h>
+
+#include "base_sorts.h"
+
+void knuth_other_base_mergesort(unsigned int a[], int N)
+{
+	unsigned int* aux = malloc(N * sizeof(unsigned int));
+	int s; /* s keeps track of the target, we can write something to remove this */
+	int i,j,k,l;
+	int d; /* direction counter */
+	
+	int temp;
+
+	int p, q, r; /* p keeps count of the number of items in this merge, q and r are counters for ascending and descending */
+	unsigned int* source;
+	unsigned int* target;
+
+	describe_predictor(&global_predictor[0], "S2");
+	describe_predictor(&global_predictor[1], "S3");
+	describe_predictor(&global_predictor[2], "S5");
+	describe_predictor(&global_predictor[3], "S6");
+	describe_predictor(&global_predictor[4], "S7");
+	describe_predictor(&global_predictor[5], "S8");
+	describe_predictor(&global_predictor[6], "S10");
+	describe_predictor(&global_predictor[7], "S11");
+	describe_predictor(&global_predictor[8], "S12");
+	describe_predictor(&global_predictor[9], "S13");
+/*S1:*/
+	s = 0; /* we should be able to elimnate s from here */
+	p = 1;
+
+S2:
+	i = 0;
+	j = N-1;
+	k = -1;
+	l = N;
+	d = 1;
+	q = p;
+	r = p;
+
+	if (s == 0)
+	{
+		branch_taken(&global_predictor[0]);
+		source = a;
+		target = aux;
+	}
+	else
+	{
+		branch_not_taken(&global_predictor[0]);
+		source = aux;
+		target = a;
+	}
+
+S3:
+	if (source[i] > source[j])
+	{
+		branch_taken(&global_predictor[1]);
+		goto S8;
+	}
+	branch_not_taken(&global_predictor[1]);
+
+/*S4:*/
+	k = k + d;
+	target[k] = source[i];
+
+/*S5:*/
+	i = i+1;
+	q = q-1;
+	if (q > 0)
+	{
+		branch_taken(&global_predictor[2]);
+		goto S3;
+	}
+	branch_not_taken(&global_predictor[2]);
+
[…]

### 2005-05-26 multi_mergesort.c

@@ -1,0 +1,797 @@
+#include <stdlib.h>
+#include <string.h>
+#include <stdio.h>
+#include <malloc.h>
+#include <limits.h>
+
+#include "cache_sorts.h"
+
+#define LIMIT OUT_OF_PLACE_LIMIT
+#define LIMIT_BITS OUT_OF_PLACE_LIMIT_BITS
+
+static void
+fix_up(unsigned int heap[], int child)
+{
+	int parent;
+
+	unsigned int v = heap[child];
+	while(less(v, heap[parent = ((child-1)/HEAP_SIZE)]))
+	{
+		branch_taken(&global_predictor[33]);
+		heap[child] = heap[parent];
+		heap[parent] = v;
+		child = parent;
+	}
+	branch_not_taken(&global_predictor[33]);
+}
+
+static void
+fix_down(unsigned int heap[], int parent, int N)
+{
+	int child;
+	int child2;
+	unsigned int v = heap[parent];
+/*	printf("fixdown when N = %d\n", N); */
+	while(1) /* see if it has any children */
+	{
+		if (parent*HEAP_SIZE+1 < N) branch_taken(&global_predictor[34]);
+		else
+		{
+			branch_not_taken(&global_predictor[34]);
+			break;
+		}
+		child = parent*HEAP_SIZE + 1; /* this gets the rightmost node */
+		child2 = child;
+
+		if (less(heap[child2 + 0], heap[child]))
+		{
+			child = child2 + 0;
+			branch_taken(&global_predictor[35]);
+		}
+		else branch_not_taken(&global_predictor[35]);
+		if (less(heap[child2 + 1], heap[child]))
+		{
+			child = child2 + 1;
+			branch_taken(&global_predictor[36]);
+		}
+		else branch_not_taken(&global_predictor[36]);
+		if (less(heap[child2 + 2], heap[child]))
+		{
+			child = child2 + 2;
+			branch_taken(&global_predictor[37]);
+		}
+		else branch_not_taken(&global_predictor[37]);
+		if (less(heap[child2 + 3], heap[child]))
+		{
+			child = child2 + 3;
+			branch_taken(&global_predictor[38]);
+		}
+		else branch_not_taken(&global_predictor[38]);
+
+#if (HEAP_SIZE == 8)
+		if (less(heap[child2 + 4], heap[child]))
+		{
+			child = child2 + 4;
+			branch_taken(&global_predictor[39]);
+		}
[…]

### 2005-05-26 old/algorithmN.c

@@ -1,0 +1,161 @@
+#include <stdlib.h>
+#include <string.h>
+#include <stdio.h>
+
+#include "base_sorts.h"
+#include "insertion.c"
+
+void
+base_mergesort(unsigned int a[], int N)
+{
+	unsigned int* aux = malloc(N * sizeof(unsigned int)); /* make it twice the size to use the notation */
+
+	int s = 0; /* this picks which area we write to */
+	
+	int i,j; /* indices for the first array */
+	int k, l; /* indices for second array */
+	
+	int d, f; /* d => direction, if (f == 0) keep going */
+
+	unsigned int temp;
+
+	unsigned int* source;
+	unsigned int* target;
+	
+N2:	/* Prepare for pass */
+
+	i = 0;
+	j = N-1;
+	k = 0;
+	l = N-1;
+	d = 1;
+	f = 1;
+
+	if (s == 0)
+	{
+		source = a;
+		target = aux;
+	}
+	else
+	{
+		source = aux;
+		target = a;
+	}
+	
+
+/*	printf("2,3\n"); */
+
+N3: /* compare Ki, Kj */
+
+	if (source[i] > source[j])
+	{
+/*		printf("3,8\n"); */
+		goto N8;
+	}
+
+	if (i == j)
+	{
+		target[k] = source[i];
+/*		printf("3,13\n"); */
+		goto N13;
+	}
+
+/*N4:  transmit Ri */
+
+/*	printf("3,4\n"); */
+
+	target[k] = source[i];
+	k = k + d; /* increment in the correct direction */
+
+/*N5: // Stepdown? */
+
+	i++;
+	if (source[i-1] <= source[i])
+	{
+/*		printf("4,3\n"); */
+		goto N3;
+	}
+
+/*	printf("4,6\n"); */
+N6: 
+	target[k] = source[j];
+	k = k + d;
+
+/*N7: // stepdown? */
+
+	j--;
+	if (source[j+1] <= source[j])
+	{
+/*		printf("6,6\n"); */
+		goto N6;
+	}
+	else
+	{
+/*		printf("6,12\n"); */
+		goto N12;
+	}
+
+N8: /* transmit Rj */
+
+	target[k]  = source[j];
+	k = k + d; /* increment in the correct direction */
+
+/*N9: // Stepdown? */
+
+	j--;
+	if (source[j+1] <= source[j])
+	{
[…]

### 2005-05-26 old/optimized_cache_heapsort.c

@@ -1,0 +1,193 @@
+#include "cache_sorts.h"
+
+#include <stdlib.h>
+#include <stdio.h>
+#include <string.h>
+#include <limits.h>
+
+#define HEAP_SIZE 4 /* never set this to 1 */
+
+/* these days we index from 0 - N-1
+* the rightmost child is (parent + 1)*heapsize
+* the parent = (child - 1) / heapsize 
+*
+* actually, we go from leftchild, which is (parent * heapsize) + 1
+* parent = (child - 1) / heapsize
+*/
+
+/*
+static int
+check_heap_property(unsigned int a[], int N)
+{
+	int parent;
+	int i;
+	for(parent = 0; parent < N/HEAP_SIZE+1; parent++)
+	{
+		int l = parent*HEAP_SIZE + 1;
+//		printf("checking children of %d (a[parent] = %d\n", parent, a[parent]);
+		for(i = 0; i < HEAP_SIZE; i++)
+		{
+//			printf("checking child %d (a[%d] = %d)\n", HEAP_SIZE-i-1, r-i, a[r-i]);
+			if (l+i >= N)
+			{
+//				printf("finsihing\n");
+				break;
+			}
+			if (a[parent] > a[l+i])
+			{
+				printf("FAIL: node %d (%d) is greater than child %d (node %d (%d)) \n", parent, a[parent], i, l+i, a[l+i]);
+				return -1;
+			}
+		}	
+	}
+	return 0;
+}
+*/
+
+
+static void
+fix_up(unsigned int heap[], int child)
+{
+	int parent;
+/*	printf("fixup: %d\n ", child); */
+/*	printf("next child = %d, parent = %d, heap[child] = %d, heap[parent] = %d\n", child, ((child)/HEAP_SIZE), heap[child], heap[((child)/HEAP_SIZE)]);*/
+	while(less(heap[child], heap[parent = ((child-1)/HEAP_SIZE)]))
+	{
+/*		printf("child = %d, parent = %d, heap[child] = %d, heap[parent] = %d\n", child, parent, heap[child], heap[parent]);*/
+
+		exch(heap[child], heap[parent]);
+		child = parent;
+/*		printf("next child = %d, parent = %d, heap[child] = %d, heap[parent] = %d\n", child, ((child)/HEAP_SIZE), heap[child], heap[((child)/HEAP_SIZE)]);*/
+	}
[…]

### 2005-05-26 on2_bubblesort2.c

@@ -1,0 +1,21 @@
+#include "on2_sorts.h"
+
+void
+on2_bubblesort2(unsigned int a[], int N)
+{
+	int i = 0, j, k = 0;
+
+	while(i < N-1)
+	{
+		k = N-1;
+		for (j = N-1; j > i; j--)
+		{
+			if (less(a[j],a[j-1]))
+			{
+				k = j;
+				exch(a[j-1],a[j]);
+			}
+		}
+		i = k;
+	}
+}

### 2005-05-26 paper/docs/code/base_quicksort.c

@@ -1,0 +1,107 @@
+#define THRESHHOLD 10
+
+static int* stack = NULL;
+static int stack_index = 0;
+
+static void
+stackinit(unsigned int N)
+{
+
+	int lg2N = 32; /* im hard coding this and i dont care */
+	unsigned int msbN = 1 << (lg2N - 1);
+	while(msbN > N)
+	{
+		lg2N--;
+		msbN = msbN >> 1;
+	}
+	stack = malloc(((lg2N + 2) << 2) * sizeof(int));
+}
+
+inline static void push(int valueA, int valueB)
+{
+	stack[stack_index++] = valueA;
+	stack[stack_index++] = valueB;
+}
+
+inline static int pop()
+{
+	stack_index--;
+	return stack[stack_index];
+}
+
+inline static void stackclear()
+{
+	if (stack != NULL) free(stack);
+	stack = NULL;
+}
+
+inline static int stackempty() { return (stack_index==0); }
+
+static int partition(Item a[], int l, int r)
+{
+
+	Item v = a[r];
+	int i = l - 1;
+	int j = r;
+
+	for(;;)
+	{
+		while (less(a[++i], v))
+			;
+		while (less(v, a[--j]))
+			;
+		
+		if (i >= j) break;
+		exch(a[i], a[j]);
+	}
+	exch(a[i], a[r]);
+
+	return i;
+}
+
+base_quicksort(unsigned int a[], int N)
+{
+	stackinit(N);
+	int r = N-1;
+	int l = 0;
+
+	while(1)
+	{
+		if (r - l <= THRESHHOLD)
+		{
+			if (stackempty()) break;
+
+			l = pop(); r = pop();
+			continue;
+		}
+
+		/* Median of 3 partitioning*/
+		m = (l+r)/2;
+
+		exch(a[m], a[r-1]);
+		compexch(a[l], a[r-1]);
+		compexch(a[l], a[r]);
+		compexch(a[r-1], a[r]);
+
+		int i = partition(a,l+1,r-1);
+
+		if (i-l > r-i)
+		{
+			push(i-1,l);
+			l = i+1;
+		}
+		else
+		{
+			push(r,i+1);
+			r = i-1;
+		}
+	}
+
+	stackclear();
+	/* the +1 isnt immediately obvious. 
+	 * its because THRESHHOLD is the difference between l and r up above */
+	if (THRESHHOLD + 1 > N) insertion_sentinel(a,N);
[…]

### 2005-05-26 paper/docs/code/cache_heapsort.c

@@ -1,0 +1,80 @@
+#define HEAP_SIZE 8 
+fix_up(Item heap[], int child)
+{
+	Item v = heap[child];
+	while(less(v, heap[int parent = ((child-1)/HEAP_SIZE)]))
+	{
+		heap[child] = heap[parent];
+		heap[parent] = v;
+		child = parent;
+	}
+}
+
+fix_down(Item heap[], int parent, int N)
+{
+	Item v = heap[parent];
+	while(int parent*HEAP_SIZE+1 < N) 
+	{
+		int child = parent*HEAP_SIZE + 1;
+		int child2 = child;
+
+		if (less(heap[child2 + 0], heap[child])) child = child2 + 0;
+		if (less(heap[child2 + 1], heap[child])) child = child2 + 1;
+		if (less(heap[child2 + 2], heap[child])) child = child2 + 2;
+		if (less(heap[child2 + 3], heap[child])) child = child2 + 3;
+
+#if (HEAP_SIZE == 8)
+		if (less(heap[child2 + 4], heap[child])) child = child2 + 4;
+		if (less(heap[child2 + 5], heap[child])) child = child2 + 5;
+		if (less(heap[child2 + 6], heap[child])) child = child2 + 6;
+		if (less(heap[child2 + 7], heap[child])) child = child2 + 7;
+#endif
+		/* stop when the larger child is les than the parent */
+		if (!less(heap[child], v)) break;
+
+		/* move down */
+		heap[parent] = heap[child];
+		parent = child;
+	}
+	heap[parent] = v;
+}
+
+void cache_heapsort(Item a[], int N)
+{
+	int start_padding = HEAP_SIZE - 1;
+	int length = 0;
+	int sentinel_count = (HEAP_SIZE) - (N % HEAP_SIZE) + 1;
+	if (sentinel_count == HEAP_SIZE) sentinel_count = 0;
+
+	/* the size is N + padding_at_the_start + padding_at_end, 
+	 * rounded up to next HEAPSIZE 
+	 * */
+	int size_to_allocate = ((N + start_padding + sentinel_count - 1) 
+			& ~(HEAP_SIZE-1)) + HEAP_SIZE;
+
+	Item* heap_data = memalign(HEAP_SIZE, size_to_allocate * sizeof(Item));
+	Item* heap = &heap_data[start_padding];
+
+	/* first build the heap */
+	for(i = 0; i < N; i++) 
[…]

### 2005-05-26 paper/docs/code/old_shellsort.c

@@ -1,0 +1,18 @@
+void gonnet_shellsort(Item a[], int N)
+{
+	int l = 0, r = N-1, i ,h;
+	for (h = r-l+1; h > 0; h = ((h > 1) && (h < 5)) ? 1 : 5*h/11)
+	{
+		for (i = l+h; i <= r; i++)
+		{
+			int j = i; 
+			Item v = a[i]; 
+			while (j >= l+h && less(v, a[j-h]))
+			{ 
+				a[j] = a[j-h]; 
+				j -= h; 
+			}
+			a[j] = v; 
+		}
+	}
+}

### 2005-05-26 paper/docs/code/on2_shakersort.c

@@ -1,0 +1,22 @@
+void shakersort(Item a[], int N)
+{
+	int i, j, k = N-1, sorted;
+	for (i = 0; i < k;  )
+	{
+		for (j = k; j > i; j--)
+			if (less(a[j],a[j-1]))
+				exch(a[j-1],a[j]);
+		i++;
+		sorted = 1;
+		for (j = i ; j < k; j++)
+		{
+			if (less(a[j+1],a[j]))
+			{
+				exch(a[j],a[j+1]);
+				sorted = 0;
+			}
+		}
+		if (sorted) return;
+		k--;
+	}
+}

### 2005-05-26 paper/docs/code/on2_shakersort2.c

@@ -1,0 +1,26 @@
+void shakersort2(Item a[], int N)
+{
+	int i = 0, j, k = N-1, l = N-1;
+	for ( ; i < k; )
+	{
+		for (j = k; j > i; j--)
+		{
+			if (less(a[j],a[j-1]))
+			{
+				exch(a[j-1],a[j]);
+				l = j;
+			}
+		}
+		i = l;
+
+		for (j = i ; j < k; j++)
+		{
+			if (less(a[j+1],a[j]))
+			{
+				exch(a[j],a[j+1]);
+				l = j;
+			}
+		}
+		k = l;
+	}
+}

### 2005-05-26 paper/fastsort_perfex/base_quicksort.c

@@ -1,0 +1,199 @@
+#include <stdlib.h>
+#include <stdio.h>
+
+#include "base_sorts.h"
+
+#define THRESHHOLD 10
+
+static int* stack = NULL;
+static int stack_index = 0;
+
+static void
+stackinit(unsigned int N)
+{
+
+	int lg2N = 32; /* im hard coding this and i dont care */
+	unsigned int msbN = 1 << (lg2N - 1);
+	while(msbN > N)
+	{
+		lg2N--;
+		msbN = msbN >> 1;
+	}
+
+/*	printf("\nAllocating a stack %d ints big (lg2N == %d)\n", (lg2N + 2) << 2, lg2N); */
+	
+	stack = malloc(((lg2N + 2) << 2) * sizeof(int));
+}
+
+inline static void
+push(int valueA, int valueB)
+{
+	stack[stack_index++] = valueA;
+	stack[stack_index++] = valueB;
+}
+
+inline static int 
+pop()
+{
+	stack_index--;
+	return stack[stack_index];
+}
+
+inline static void
+stackclear()
+{
+	if (stack != NULL) free(stack);
+	stack = NULL;
+}
+
+inline static int
+stackempty()
+{
+	return (stack_index==0);
+}
+
+static int
+partition(unsigned int a[], int l, int r)
+{
+	unsigned int v;
+	int i, j;
+
+	v = a[r];
+	i = l - 1;
+	j = r;
+
+	for(;;)
+	{
+		while (less(a[++i], v))
+		{
+			branch_taken(&global_predictor0);
+		}
+		branch_not_taken(&global_predictor0);
+
+		while (less(v, a[--j]))
+		{
+			branch_taken(&global_predictor1);
+		}
+		branch_not_taken(&global_predictor1);
+		
+		if (i >= j)
+			break;
+
+		exch(a[i], a[j]);
+	}
+
+	exch(a[i], a[r]);
+
+	return i;
+}
+
+static void
+insertion_sentinel(unsigned int a[], int N)
+{
+	int i;
+	unsigned min;
+
+	min = 0;
+	for(i = 1; i < N; i++)
+	{
+		if(less(a[i], a[min]))
+		{
+			min = i;
+		}
+	}
+	exch(a[0], a[min]);
+
+}
+
+static void
+insertion(unsigned int a[], int N)
+{
+	int i;
+	unsigned int v;
[…]

### 2005-05-26 paper/fastsort_perfex/base_radixsort.c

@@ -1,0 +1,45 @@
+#include <stdlib.h>
+
+#include "base_sorts.h"
+#include "insertion.c"
+
+#define THRESHHOLD 10
+
+#define bitsword 32
+#define bitsbyte 8
+#define bytesword 4
+#define Radix (1 << bitsbyte)
+
+#define digit(A, B) (((A) >> (bitsword-((B)+1)*bitsbyte)) & (Radix-1))
+
+void
+base_radixsort(unsigned int a[], int N)
+{
+	unsigned int* radix_aux = malloc(N * sizeof(unsigned int));
+	int r = N-1;
+	int l = 0;
+	int i, j, w, count[Radix+1]; /* R is the size of a digit, hence the number of bins */
+
+	for(w = bytesword - 1; w >= 0; w--) /* this is from lsb. 3 == lsByte, 0 == msByte */
+	{
+		for(j = 0; j < Radix; j++) count[j] = 0; /* set the counts to 0 */
+		for(i = l; i <= r; i++) /* count each digit */
+		{
+			count[digit(a[i], w) + 1]++;
+		}
+		for(j = 1; j < Radix; j++) /* make the count cumulatative */
+		{
+			count[j] += count[j-1];
+		}
+		for(i = l; i <= r; i++) /* when we look up the array, we now have its position */
+		{
+			radix_aux[count[digit(a[i], w)]++] = a[i];
+		}
+		for(i = l; i <= r; i++) /* copy it back */
+		{
+			a[i] = radix_aux[i];
+		}
+	}
+
+	free(radix_aux);
+}

### 2005-05-26 paper/fastsort_perfex/cache_radixsort.c

@@ -1,0 +1,66 @@
+#include <stdlib.h>
+#include <string.h>
+
+#include "base_sorts.h"
+#include "insertion.c"
+
+#define THRESHHOLD 10
+
+#define BITSWORD 32
+#define BITSBYTE 8
+#define BYTESWORD 4
+#define R 256
+
+#define digit(A, B) (((A) >> (BITSWORD-((B)+1)*BITSBYTE)) & (R-1))
+
+void
+cache_radixsort(unsigned int a[], int N)
+{
+	unsigned int* aux = malloc(N * sizeof(unsigned int));
+	int i, j, w; /* R is the size of a digit, hence the number of bins */
+	int count0[R+1];
+	int count1[R+1];
+	int* temp;
+
+	int* count = count0;
+	int* next_count = count1;
+
+	unsigned int* s = a;
+	unsigned int* t = aux;
+
+	/* count the digits for the first time */
+	for(j = 0; j < R; j++) count[j] = 0;
+	for(i = 0; i < N; i++) count[digit(s[i], 3) + 1]++;
+	for(j = 1; j < R; j++) count[j] += count[j-1];
+
+	for(w = BYTESWORD - 1; w > 0; w--) /* this is from lsb. 3 == lsByte, 0 == msByte */
+	{
+		for(j = 0; j < R; j++) next_count[j] = 0; /* set the next counts to 0 */
+
+		for(i = 0; i <= N-1; i++) /* when we look up the array, we now have its position */
+		{
+			t[count[digit(s[i], w)]++] = s[i];
+			next_count[digit(s[i], w-1) + 1]++;
+		}
+	
+		for(j = 1; j < R; j++) next_count[j] += next_count[j-1]; /* make the next counts cumulative */
+
+		/* swap the counts */
+		temp = next_count;
+		next_count = count;
+		count = temp;;
+
+		/* swap the arrays */
+		temp = t;
+		t = s;
+		s = temp;
+	}
+
+	/* finish it off */
+	for(i = 0; i <= N-1; i++) /* when we look up the array, we now have its position */
+	{
+		t[count[digit(s[i], w)]++] = s[i];
+	}
+
+	free(aux);
+}

### 2005-05-26 paper/fastsort_perfex/do_nothing.c

@@ -1,0 +1,4 @@
+void
+do_nothing(unsigned int a[], int N)
+{
+}

### 2005-05-26 paper/fastsort_perfex/double_multi_mergesort.c

@@ -1,0 +1,990 @@
+#include <stdlib.h>
+#include <string.h>
+#include <stdio.h>
+#include <malloc.h>
+#include <limits.h>
+
+#include "cache_sorts.h"
+
+#define LIMIT OUT_OF_PLACE_LIMIT
+#define LIMIT_BITS OUT_OF_PLACE_LIMIT_BITS
+
+#define ODD_COUNT 8
+#define EVEN_COUNT 4
+
+static void
+merge(unsigned int source[], int N, int starting_size, unsigned int target[]);
+
+static void
+merge_reverse(unsigned int source[], int N, int starting_size, unsigned int target[]);
+
+static void
+presort(unsigned int a[], int N);
+static void
+presort_flexible(unsigned int a[], int N);
+
+static int presort_count = 8;
+static int double_presort_count = 16;
+
+static void
+fix_up(unsigned int heap[], int child)
+{
+	int parent;
+
+	unsigned int v = heap[child];
+	while(less(v, heap[parent = ((child-1)/HEAP_SIZE)]))
+	{
+		heap[child] = heap[parent];
+		heap[parent] = v;
+		child = parent;
+	}
+}
+
+static void
+fix_down(unsigned int heap[], int parent, int N)
+{
+	int child;
+	int child2;
+	unsigned int v = heap[parent];
+
+	while(parent*HEAP_SIZE+1 < N) /* see if it has any children */
+	{
+		child = parent*HEAP_SIZE + 1; /* this gets the rightmost node */
+		child2 = child;
+
+		if (less(heap[child2 + 0], heap[child])) child = child2 + 0;
+		if (less(heap[child2 + 1], heap[child])) child = child2 + 1;
+		if (less(heap[child2 + 2], heap[child])) child = child2 + 2;
+		if (less(heap[child2 + 3], heap[child])) child = child2 + 3;
+
+#if (HEAP_SIZE == 8)
+		if (less(heap[child2 + 4], heap[child])) child = child2 + 4;
+		if (less(heap[child2 + 5], heap[child])) child = child2 + 5;
+		if (less(heap[child2 + 6], heap[child])) child = child2 + 6;
+		if (less(heap[child2 + 7], heap[child])) child = child2 + 7;
+
+#endif
+
+		/* when the larger child isnt large enough to be promoted, stop */
+		if (!less(heap[child], v)) break;
+
+		/* move down */
+		heap[parent] = heap[child];
+		parent = child;
+	}
[…]

### 2005-05-26 paper/fastsort_perfex/double_tiled_mergesort.c

@@ -1,0 +1,779 @@
+#include <stdlib.h>
+#include <string.h>
+#include <stdio.h>
+#include <malloc.h>
+
+#include "cache_sorts.h"
+
+#define LIMIT OUT_OF_PLACE_LIMIT
+#define LIMIT_BITS OUT_OF_PLACE_LIMIT_BITS
+
+#define ODD_COUNT 8
+#define EVEN_COUNT 4
+
+static void
+merge(unsigned int source[], int N, int starting_size, unsigned int target[]);
+
+static void
+merge_reverse(unsigned int source[], int N, int starting_size, unsigned int target[]);
+
+static void
+presort(unsigned int a[], int N);
+static void
+presort_flexible(unsigned int a[], int N);
+
+static int presort_count = 8;
+static int double_presort_count = 16;
+
+inline static void set_presort_count(int count)
+{
+	presort_count = count;
+	double_presort_count = count * 2;
+}
+/*
+static void
+check_merge_reverse(unsigned int a[], int N)
+{
+	int i;
+	for(i = 0; i < N-1; i++)
+	{
+		if (a[i+1] > a[i])
+		{
+			printf("bad reverse: A: %p, a[%d] = %d, a[%d] = %d\n", a, i, a[i], i+1, a[i+1]);
+			return;
+		}
+	}
+}
+static void
+check_merge(unsigned int a[], int N)
+{
+	int i;
+	for(i = 0; i < N-1; i++)
+	{
+		if (a[i] > a[i+1])
+		{
+			printf("bad forwards: A: %p, a[%d] = %d, a[%d] = %d\n", a, i, a[i], i+1, a[i+1]);
+			return;
+		}
+	}
+}
+*/
+
+
+static int get_count(const int N)
+{
+
+	int lg2N = 32; /* im hard coding this and i dont care */
+	unsigned int msbN = 1 << (lg2N - 1);
+	while(msbN >= N)
+	{
+		lg2N--;
+		msbN = msbN >> 1;
+	}
+	return lg2N;
+}
+
+void
+double_tiled_mergesort(unsigned int a[], int N)
+{
+	/* track keeps an eye on where i started last */
+	/* outer track keeps and eye on which cache_sized/2 segemnt i started in last */
+	int i,j; /* indices for the first array */
+	
+	int level1_count;
+	int level2_count;
+	int extra_level1_count = 0;
+	int extra_level2;
+	int final_extra;
+
+
[…]

### 2005-05-26 paper/fastsort_perfex/main.c

@@ -1,0 +1,191 @@
+/***************************************************************************
+                          main.c  -  description
+                             -------------------
+    begin                : Sat Oct 11 2003
+    copyright            : (C) 2003 by Paul Biggar
+    email                : biggarp@tcd.ie
+ ***************************************************************************/
+
+#include <stdio.h> // fopen, fgets, fclose
+#include <errno.h> // errno
+#include <assert.h> // assert
+#include <string.h> // strerror
+#include <stdlib.h> // atoi, malloc, free
+#include <sys/types.h> // time
+#include <sys/time.h> // time
+
+#include "on2_sorts.h"
+#include "base_sorts.h"
+#include "branch_sorts.h"
+#include "cache_sorts.h"
+#include "visual_sorts.h"
+#include "utils.h"
+#include "paper/predictor.h"
+
+#define RANDOM_SIZE (4194304)
+#define _USE_SOFTWARE_PREDICTOR 0
+//#define RANDOM_SIZE (534523)
+//#define RANDOM_SIZE (262144)
+//#define RANDOM_SIZE (16384)
+//#define RANDOM_SIZE (4096)
+//#define RANDOM_SIZE (128)
+
+#define RUN_VISUAL 1
+#define VISUAL_SIZE (128)
+
+unsigned int sorted_array[RANDOM_SIZE];
+unsigned int random_array[RANDOM_SIZE];
+unsigned int random_array2[RANDOM_SIZE];
+
+FILE* visual_log = NULL;
+
+static struct timeval t1;
+static struct timeval t2;
+
+
+static inline long
+get_milliseconds(struct timeval *tp1, struct timeval *tp2)
+{
+	return ( (tp2->tv_sec - tp1->tv_sec) * 1000000) + (tp2->tv_usec - tp1->tv_usec);
+}
+
+static void inline 
+start_timer()
+{
+	if (gettimeofday(&t1, NULL) != 0)  printf("Bad result: \"%s\"", strerror(errno));
+}
+
+static void inline
+stop_timer()
+{
+	if (gettimeofday(&t2, NULL) != 0) printf("Bad result: \"%s\"", strerror(errno));
+}
+
+static void inline
+print_timer(const char* description)
+{
+	printf("%s: %ld ticks\n", description, get_milliseconds(&t1, &t2));
+}
+
[…]

### 2005-05-26 paper/fastsort_perfex/old_shellsort.c

@@ -1,0 +1,23 @@
+#define Item unsigned int
+#include "utils.h"
+
+void old_shellsort(Item a[], int N)
+{
+	int l = 0;
+	int r = N-1;
+	int i, h; 
+	for (h = r-l+1; h > 0; h = ((h > 1) && (h < 5)) ? 1 : 5*h/11)
+	{
+		for (i = l+h; i <= r; i++)
+		{
+			int j = i; 
+			Item v = a[i]; 
+			while (j >= l+h && less(v, a[j-h]))
+			{ 
+				a[j] = a[j-h]; 
+				j -= h; 
+			}
+			a[j] = v; 
+		}
+	}
+}

### 2005-05-26 paper/fastsort_perfex/on2_insertsort.c

@@ -1,0 +1,8 @@
+#include "insertion.c"
+
+void
+on2_insertsort(unsigned int a[], int N)
+{
+	insertion_sentinel(a, N);
+	insertion(a, N);
+}

### 2005-05-26 paper/fastsort_perfex/on2_shakersort.c

@@ -1,0 +1,32 @@
+#include "on2_sorts.h"
+
+void
+on2_shakersort(unsigned int a[], int N)
+{
+	int i, j;
+	int k = N-1;
+	int sorted;
+
+	for (i = 0; i < k;  )
+	{
+		for (j = k; j > i; j--)
+		{
+			if (less(a[j],a[j-1]))
+			{
+				exch(a[j-1],a[j]);
+			}
+		}
+		i++;
+		sorted = 1;
+		for (j = i ; j < k; j++)
+		{
+			if (less(a[j+1],a[j]))
+			{
+				exch(a[j],a[j+1]);
+				sorted = 0;
+			}
+		}
+		if (sorted) return;
+		k--;
+	}
+}

### 2005-05-26 paper/fastsort_perfex/predictor.c

@@ -1,0 +1,104 @@
+#include "predictor.h"
+
+#include <string.h>
+
+#if _USE_SOFTWARE_PREDICTOR
+	
+static char state_names[4][20] = 
+{
+	"NT saturated",
+	"NT",
+	"T",
+	"T saturated"
+};
+
+void
+init_predictor(struct predictor* pred)
+{
+	memset(pred, 0, sizeof(struct predictor));
+	/*pred->state = NOT_TAKEN_SATURATED; */
+	pred->state = TAKEN;
+}
+
+void 
+branch_taken(struct predictor* pred)
+{
+	int s = pred->state;
+	pred->branch_count[s]++;
+	pred->taken_count[s]++;
+
+	/* check what was predicted */
+	if (pred->state >= TAKEN) pred->correct_count[s]++; 
+	else pred->incorrect_count[s]++;
+
+	/* need to update the predictor? */
+	if (pred->state != TAKEN_SATURATED) pred->state++;
+		
+}
+
+void 
+branch_not_taken(struct predictor* pred)
+{
+	int s = pred->state;
+	pred->branch_count[s]++;
+	pred->not_taken_count[s]++;
+
+	/* check what was predicted */
+	if (pred->state <= NOT_TAKEN) pred->correct_count[s]++; 
+	else pred->incorrect_count[s]++;
+
+	/* need to update the predictor? */
+	if (pred->state != NOT_TAKEN_SATURATED) pred->state--;
+}
+
+void
+dump_predictor_info(FILE* file, struct predictor* pred, char* name)
+{
+	fprintf(file,	"%s:"
+					"states                     %13s %13s %13s %13s %13s\n"
+					"branch_count:              %13d %13d %13d %13d %13d\n"
+					"correct_count:             %13d %13d %13d %13d %13d\n"
+					"incorrect_count:           %13d %13d %13d %13d %13d\n"
+					"taken_count:               %13d %13d %13d %13d %13d\n"
+					"not_taken_count:           %13d %13d %13d %13d %13d\n"
+					"current state:             %13s\n\n",
+					name,
+					state_names[0], state_names[1],
+					state_names[2], state_names[3], "Total",
+					pred->branch_count[0], pred->branch_count[1],
+					pred->branch_count[2], pred->branch_count[3],
+					pred->branch_count[0] + pred->branch_count[1] +
+					pred->branch_count[2] + pred->branch_count[3],
+					pred->correct_count[0], pred->correct_count[1],
+					pred->correct_count[2], pred->correct_count[3],
[…]

### 2005-05-26 paper/fastsort_perfex/predictor.h

@@ -1,0 +1,56 @@
+#ifndef _PREDICTOR_H_
+#define _PREDICTOR_H_
+
+#if _USE_SOFTWARE_PREDICTOR
+#include <stdio.h>
+struct predictor
+{
+	int branch_count[4]; /* # of branches */
+
+	int correct_count[4]; /* # of correct predictions */
+	int incorrect_count[4]; /* # of incorrect predictions */
+
+	int taken_count[4]; /* # of branches rtaken */
+	int not_taken_count[4]; /* # of branches not taken */
+
+	int state; /* the current state of the predictor */
+	/* 0 = saturated not-taken */
+	/* 1 = unsaturated not-taken */
+	/* 2 = unstaturated taken */
+	/* 3 = saturated taken */
+
+};
+
+#define NOT_TAKEN_SATURATED 0
+#define NOT_TAKEN 1
+#define TAKEN 2
+#define TAKEN_SATURATED 3
+
+void init_predictor(struct predictor* pred);
+
+void branch_taken(struct predictor* pred);
+void branch_not_taken(struct predictor* pred);
+
+void dump_predictor_info(FILE* file, struct predictor* pred, char* name);
+
+
+extern struct predictor global_predictor0;
+extern struct predictor global_predictor1;
+extern struct predictor global_predictor2;
+extern struct predictor global_predictor3;
+extern struct predictor global_predictor4;
+extern struct predictor global_predictor5;
+extern struct predictor global_predictor6;
+extern struct predictor global_predictor7;
+extern struct predictor global_predictor8;
+extern struct predictor global_predictor9;
+#else
+
+#define init_predictor(A) do{}while(0)
+#define branch_taken(A)
+#define branch_not_taken(A)
+#define dump_predictor_info(A, B, C) do{}while(0)
+
+#endif
+
+#endif

### 2005-05-26 paper/fastsort_perfex/shellsort.c

@@ -1,0 +1,51 @@
+#define Item unsigned int
+#include "utils.h"
+
+void shellsort(Item a[], int N)
+{
+	int i, h,min, bound; 
+	for (h = 5*N/11; h >= 5; h = 5*h/11)
+	{
+		for (i = h; i <= N-1; i++)
+		{
+			int j = i; 
+			Item v = a[i]; 
+			while (j >= h && less(v, a[j-h]))
+			{ 
+				a[j] = a[j-h]; 
+				j -= h; 
+			}
+			a[j] = v; 
+		}
+	}
+
+	/* put a sentinel in place */
+	/* the sentinel will be in the first h positions */
+	/* for h to be below 5, it must have been less than 11 beforehand */
+	min = 0;
+	bound = 11;
+	if (N < bound) bound = N;
+
+	for(i = 1; i < bound; i++)
+	{
+		if(less(a[i], a[min]))
+		{
+			min = i;
+		}
+	}
+	exch(a[0], a[min]);
+
+
+	for(i = 1; i < N; i++)
+	{
+		int j = i;
+		Item v = a[i];
+
+		while(less(v, a[j-1]))
+		{
+			a[j] = a[j-1];
+			j--;
+		}
+		a[j] = v;
+	}
+}

### 2005-05-26 paper/fastsort_perfex/simpleMain.c

@@ -1,0 +1,53 @@
+#include <stdlib.h>
+#include <stdio.h>
+#include "base_sorts.h"
+#include "cache_sorts.h"
+#include "branch_sorts.h"
+#include "on2_sorts.h"
+
+#ifndef SORT
+#error "You must define the sort function like this: \"gcc -DSORT=base_insertion\""
+#endif
+
+#ifndef RANDOM_SIZE
+#error "You must define the size of the array like this: \"gcc -DRANDOM_SIZE=128\""
+#endif
+
+int main()
+{
+	/* if you segfault here, comment out _FASTSORT_STACK */
+	/* this happens because your system stack ownt take a 2 or 4meg stack variable */
+#define _FASTSORT_STACK
+#ifdef _FASTSORT_STACK
+	unsigned int random_array[RANDOM_SIZE];
+#else
+	unsigned int* random_array = malloc(RANDOM_SIZE * sizeof(unsigned int));
+#endif
+
+	/* fill the array or print error message */
+	int count = 0;
+	FILE* random_file = fopen("paper/10megs.001", "r");
+	if (random)
+	{
+		count = fread(random_array, 1, sizeof(unsigned int) * RANDOM_SIZE, random_file);
+	}
+	if (count < RANDOM_SIZE)
+	{
+		fprintf(stderr, "Random not inited\n"); /* does printing to stderr stop simple? */
+		fprintf(stdout, "Random not inited\n"); /* do this just in case */
+	}
+	fclose(random_file);
+
+
+
+
+	SORT(random_array, RANDOM_SIZE);
+
+#ifndef _FASTSORT_STACK
+	free(random_array);
+#endif
+	return 0;
+
+}
+
+

### 2005-05-26 paper/fastsort_perfex/utils.h

@@ -1,0 +1,63 @@
+#ifndef _UTILS_H_
+#define _UTILS_H_
+
+#include "predictor.h"
+
+#ifndef NULL
+#define NULL 0
+#endif
+
+#define HEAP_SIZE 8
+
+#define ALIGNMENT (HEAP_SIZE * sizeof(unsigned int))
+
+#define LINE_BITS 5
+#define LINE_MASK 0x1F
+/* 32 byte line */ 
+
+#define BLOCK_BITS 16
+#define BLOCK_MASK 0xFFFF
+/* 65536 blocks */
+
+#define BLOCK_AND_LINE_BITS (BLOCK_BITS + LINE_BITS)
+#define BLOCK_AND_LINE_MASK ((BLOCK_MASK << LINE_BITS) | (LINE_MASK))
+
+/* LIMIT isnt the same as 2^(LINE_BITS+BLOCK_BITS)  */
+/* it is: sizeof cache / sizeof(unsigned long) / 2) */
+#define IN_PLACE_LIMIT 524288
+#define IN_PLACE_LIMIT_BITS (19)
+
+#define OUT_OF_PLACE_LIMIT 262144
+#define OUT_OF_PLACE_LIMIT_BITS (18)
+
+static inline unsigned int get_index(unsigned int* address)
+{
+	unsigned int a = (unsigned int)(address);
+	return (a >> LINE_BITS) & BLOCK_MASK;
+}
+
+static inline unsigned int get_offset(unsigned int* address)
+{
+	unsigned int a = (unsigned int)(address);
+	return a & LINE_MASK;
+}
+
+
+#define key(A) (A)
+#define less(A,B) (key(A) < key(B))
+#define exch(A, B) do { unsigned int t = (A); (A) = (B); (B) = t; } while(0)
+#define compexch(A, B) do { if (less((B), (A))) exch((A), (B)); } while(0)
+
+#define OUT(A) do { printf(#A ": 0x%x (%d)\n", A, A); } while(0)
+
+
+int check_sorted(unsigned int a[], int l, int r);
+
+void print_array(void* array, unsigned int count, unsigned int size, const char* description);
+void print_int_array(unsigned int array[], int N, const char* description, int left, int right, int pivot);
+
+void fill_random_array(unsigned int a[], int size, int seed);
+
+int compare_ints(const void* a, const void* b);
+
+#endif

### 2005-05-26 paper/fastsort_perfex/visual_sorts.c

@@ -1,0 +1,1353 @@
+#include "visual_sorts.h"
+#include "base_sorts.h"
+#include "utils.h"
+
+#include <string.h>
+#include <stdlib.h>
+#include <limits.h>
+#include <malloc.h>
+#include <errno.h>
+
+static FILE* visual_log = NULL;
+
+int
+open_visual_log(const char* filename, const char* flags)
+{
+	visual_log = fopen(filename, flags);
+
+	if (visual_log == NULL) 
+	{
+		printf("Error opening the visual log file: error %d \"%s\"", errno, strerror(errno));
+		return 1;
+	}
+
+	fprintf(visual_log, "<html><head><title>Visual_sorts</title></head><body text=white bgcolor=black>");
+
+	return 0;
+}
+
+void 
+close_visual_log()
+{
+	fputs("</body></html>", visual_log);
+	fclose(visual_log);
+}
+
+#define start_sort(Sort_name) do { fprintf(visual_log, "<hr><h1>%s Sort - file: \"%s\"</h1>", (Sort_name), __FILE__); } while(0)
+#define start_vars() do { fprintf(visual_log, "<p>On line %d:<b>", __LINE__); } while(0)
+
+#define add_var(A) do { fprintf(visual_log, " %s=%d", #A, (A)); } while(0)
+#define end_vars() do { fputs("</b></p>", visual_log); } while(0)
+#define add_comment(Comment) do { fprintf(visual_log, "<hr><h1>%s</h1>", (Comment)); } while(0)
+
+
+#define MAX_HEIGHT 128.0
+#define MAX 32768
+#define	HEIGHT_DIV (MAX_HEIGHT / MAX)
+
+static void
+visual_print_array(int a[], int l, int r, int highlight_left, int hightlight_right, int highlight_other)
+{
+	int i;
+	for(i = l; i <= r; i++) // we tile this ourselves cause of the fprintf
+	{
+		// the space is for a gap
+		if (i == highlight_other)
+			fprintf(visual_log, "<img src=m.gif width=2 title=%u height=%u> ", a[i], (int)(a[i] * HEIGHT_DIV));
+		else if((i >= highlight_left) && (i <= hightlight_right))
[…]

### 2005-05-26 perfexMain.c

@@ -1,0 +1,45 @@
+#include <stdlib.h>
+#include <stdio.h>
+#include "base_sorts.h"
+#include "cache_sorts.h"
+#include "branch_sorts.h"
+#include "on2_sorts.h"
+
+#ifndef SORT
+#error "You must define the sort function like this: \"gcc -DSORT=base_insertion\""
+#endif
+
+#ifndef RANDOM_SIZE
+#error "You must define the size of the array like this: \"gcc -DRANDOM_SIZE=128\""
+#endif
+
+#ifndef SEED
+#error "You must define the size of the array like this: \"gcc -DRANDOM_SIZE=128\""
+#endif
+
+int main()
+{
+	/* if you segfault here, comment out _FASTSORT_STACK */
+	/* this happens because your system stack ownt take a 2 or 4meg stack variable */
+#define _FASTSORT_STACK
+#ifdef _FASTSORT_STACK
+	unsigned int random_array[RANDOM_SIZE];
+#else
+	unsigned int* random_array = malloc(RANDOM_SIZE * sizeof(unsigned int));
+#endif
+	int i;
+	srand(SEED);
+	for(i = 0; i < RANDOM_SIZE; i++)
+	{
+		random_array[i] = rand();
+	}
+	SORT(random_array, RANDOM_SIZE);
+
+#ifndef _FASTSORT_STACK
+	free(random_array);
+#endif
+	return 0;
+
+}
+
+

### 2005-05-26 predictor.c

@@ -1,0 +1,121 @@
+#include "predictor.h"
+
+#ifdef _USE_SOFTWARE_PREDICTOR
+#include <string.h>
+
+static char state_names[4][20] = 
+{
+	"NT saturated",
+	"NT",
+	"T",
+	"T saturated"
+};
+
+void
+init_predictor(struct predictor* pred)
+{
+	memset(pred, 0, sizeof(struct predictor));
+	/*pred->state = NOT_TAKEN_SATURATED; */
+	pred->state = TAKEN;
+}
+
+void
+describe_predictor(struct predictor* pred, const char* description)
+{
+	strncpy(pred->description, description, 200);
+}
+
+void 
+branch_taken(struct predictor* pred)
+{
+	int s = pred->state;
+//	pred->branch_count[s]++;
+	pred->taken_count[s]++;
+
+	/* need to update the predictor? */
+	if (s != TAKEN_SATURATED) pred->state++;
+}
+
+void 
+branch_not_taken(struct predictor* pred)
+{
+	int s = pred->state;
+//	pred->branch_count[s]++;
+	pred->not_taken_count[s]++;
+
+	/* need to update the predictor? */
+	if (s != NOT_TAKEN_SATURATED) pred->state--;
+}
+
+void
+add_predictor(struct predictor* predA, struct predictor* predB)
+{
+	predA->taken_count[0] += predB->taken_count[0];
+	predA->taken_count[1] += predB->taken_count[1];
+	predA->taken_count[2] += predB->taken_count[2];
+	predA->taken_count[3] += predB->taken_count[3];
+	predA->not_taken_count[0] += predB->not_taken_count[0];
+	predA->not_taken_count[1] += predB->not_taken_count[1];
+	predA->not_taken_count[2] += predB->not_taken_count[2];
+	predA->not_taken_count[3] += predB->not_taken_count[3];
+}
+
+void
+dump_predictor_info(FILE* file, struct predictor* pred, const char* description)
+{
+	fprintf(file,	"%s - %s:\n"
+					"states                     %13s %13s %13s %13s %13s\n"
+					"branch_count:              %13llu %13llu %13llu %13llu %13llu\n"
+					"correct_count:             %13llu %13llu %13llu %13llu %13llu\n"
+					"incorrect_count:           %13llu %13llu %13llu %13llu %13llu\n"
+					"taken_count:               %13llu %13llu %13llu %13llu %13llu\n"
+					"not_taken_count:           %13llu %13llu %13llu %13llu %13llu\n"
[…]

### 2005-05-26 predictor.h

@@ -1,0 +1,54 @@
+#ifndef _PREDICTOR_H_
+#define _PREDICTOR_H_
+
+#define PREDICTOR_COUNT 64
+
+#ifdef _USE_SOFTWARE_PREDICTOR
+#include <stdio.h>
+struct predictor
+{
+	unsigned long long taken_count[4]; /* # of branches rtaken */
+	unsigned long long not_taken_count[4]; /* # of branches not taken */
+
+	int state; /* the current state of the predictor */
+
+	char description[200];
+	/* 0 = saturated not-taken */
+	/* 1 = unsaturated not-taken */
+	/* 2 = unstaturated taken */
+	/* 3 = saturated taken */
+
+};
+
+#define NOT_TAKEN_SATURATED 0
+#define NOT_TAKEN 1
+#define TAKEN 2
+#define TAKEN_SATURATED 3
+
+void init_predictor(struct predictor* pred);
+
+void branch_taken(struct predictor* pred);
+void branch_not_taken(struct predictor* pred);
+
+void dump_predictor_info(FILE* file, struct predictor* pred, const char* description);
+void describe_predictor(struct predictor* pred, const char*);
+
+
+void add_predictor(struct predictor* predA, struct predictor* predB);
+
+extern struct predictor global_predictor[PREDICTOR_COUNT];
+
+#else
+
+#define init_predictor(A) do{}while(0)
+#define branch_taken(A) do{}while(0)
+#define branch_not_taken(A) do{}while(0)
+#define dump_predictor_info(A, B, C) do{}while(0)
+#define describe_predictor(A, B) do{}while(0)
+#define add_predictor(A, B) do{}while(0)
+
+
+extern int global_predictor[];
+#endif
+
+#endif

### 2005-05-26 quick_perf/base_sorts.h

@@ -1,0 +1,17 @@
+#ifndef _BASE_SORTS_H_
+#define _BASE_SORTS_H_
+
+#include "utils.h"
+
+void base_heapsort(unsigned int a[], int N);
+void base_quicksort(unsigned int a[], int N);
+void base_radixsort(unsigned int a[], int N);
+void base_mergesort(unsigned int a[], int N);
+
+void knuth_base_mergesort(unsigned int a[], int N);
+void knuth_other_base_mergesort(unsigned int a[], int N);
+
+void do_nothing(unsigned int a[], int N);
+void base_quicksort9(unsigned int a[], int N);
+
+#endif

### 2005-05-26 quick_perf/branch_sorts.h

@@ -1,0 +1,9 @@
+#ifndef _BRANCH_SORTS_H_
+#define _BRANCH_SORTS_H_
+
+#include "utils.h"
+
+void multi_quicksort_seq(unsigned int a[], int N);
+void predicated_quicksort(unsigned int a[], int N);
+
+#endif

### 2005-05-26 quick_perf/multi_quicksort_seq.c

@@ -1,0 +1,381 @@
+#include <stdlib.h>
+#include <stdio.h>
+#include <string.h>
+#include <limits.h>
+
+#include "cache_sorts.h"
+#include "quick.c"
+
+#define THRESHHOLD 13
+
+/* C is the number of keys that fit in the cache */
+#define C IN_PLACE_LIMIT
+#define C_DIV_3 (C / 3)
+static void
+stackinit(unsigned int N)
+{
+
+	int lg2N = 32; /* im hard coding this and i dont care */
+	unsigned int msbN = 1 << (lg2N - 1);
+	while(msbN > N)
+	{
+		lg2N--;
+		msbN = msbN >> 1;
+	}
+
+	if (N > 2 * C) /* make space for all the pivots being pushed */
+	{
+		lg2N +=  N / C_DIV_3;
+	}
+
+/*	printf("\nAllocating a stack %d ints big (lg2N == %d)\n", (lg2N + 2) << 2, lg2N); */
+
+	/* +2 for initial values, *2 cuase we push 2 ints a time */
+	/* we only really need about 3 less, cause of the threshhold, but this doesnt hurt */
+	stack = malloc(((lg2N + 2) * 2) * sizeof(int));
+}
+
+
+static void
+insertion(unsigned int a[], int N)
+{
+	int i;
+	unsigned int v;
+/*	printf("N = %d\n", N); */
+	for(i = 1; i <= N; i++)
+	{
+		int j = i;
+		v = a[i];
+
+		while(less(v, a[j-1]))
+		{
+			branch_taken(&global_predictor[3]);
+			a[j] = a[j-1];
+			j--;
+		}
+		branch_not_taken(&global_predictor[3]);
+		a[j] = v;
+	}
+}
+
+/* this is slightly under a page, to make room for extra allocated bits*/
+/* glibc source indicates that 16 bytes are used for the allocation. */
+/*2 words are used by the pointer and the count */
+#define KEYS_PER_LIST 1022
+
+struct List
+{
+	unsigned int blocks[KEYS_PER_LIST];
+	struct List* previous;
+	int count;
+};
+
+static struct List **lists = NULL;
+static int list_count = 0;
+static void* list_memory = NULL;
+static struct List* next_available_list = NULL; 
+
+static void
+listinit(int number_of_pivots, int N)
+{
+	int i;
[…]

### 2005-05-26 quick_perf/quick.c

@@ -1,0 +1,156 @@
+static int* stack = NULL;
+static int stack_index = 0;
+
+inline static void
+push(int valueA, int valueB)
+{
+	stack[stack_index++] = valueA;
+	stack[stack_index++] = valueB;
+}
+
+inline static int 
+pop()
+{
+	stack_index--;
+	return stack[stack_index];
+}
+
+inline static void
+stackclear()
+{
+	if (stack != NULL) free(stack);
+	stack_index = 0;
+	stack = NULL;
+}
+
+inline static int
+stackempty()
+{
+	return (stack_index==0);
+}
+
+/*static void
+print_stack()
+{
+	printf("\nThe stack is: ");
+	for(int i = 0; i < stack_index; i++)
+	{
+		printf(" %d, ", stack[i]);
+	}
+}
+*/
+
+static int
+partition(unsigned int a[], int l, int r)
+{
+	unsigned int v;
+	int i, j;
+
+	v = a[r];
+	i = l - 1;
+	j = r;
+
+	for(;;)
+	{
+		while (less(a[++i], v))
+		{
+			branch_taken(&global_predictor[0]);
+		}
+		branch_not_taken(&global_predictor[0]);
+
+		while (less(v, a[--j]))
+		{
+			branch_taken(&global_predictor[1]);
+		}
+		branch_not_taken(&global_predictor[1]);
+		
+		if (i >= j)
+		{
+			branch_taken(&global_predictor[2]);
+			break;
+		}
+		branch_not_taken(&global_predictor[2]);
+
+		exch(a[i], a[j]);
+	}
+
+	exch(a[i], a[r]);
+
+	return i;
+}
+
+#if 0
+static int
+partition(unsigned int a[], int l, int r)
+{
+
+	unsigned int pivot;
+	int i, j;
+
+	asm /* source first, then destination,  */
+		/* which is different from intel */
+	(
+		/* preamble */
+		"movl	(%%eax, %%edx, 4), %%esi\n\t" /* pivot = a[r] */
+		"movl	(%%eax, %%ecx, 4), %%edi\n\t" /* key = a[l] */
+
+		"incl	%%ecx\n\t" /* i = l + 1 */
+		"decl	%%edx\n\t" /* j = r - 1 */
+
[…]

### 2005-05-26 tiled_mergesort.c

@@ -1,0 +1,546 @@
+#include <stdlib.h>
+#include <string.h>
+#include <stdio.h>
+#include <malloc.h>
+
+#include "cache_sorts.h"
+
+#define LIMIT OUT_OF_PLACE_LIMIT
+
+void
+tiled_mergesort(unsigned int a[], int N)
+{
+	/* track keeps an eye on where i started last */
+	/* outer track keeps and eye on which cache_sized/2 segemnt i started in last */
+	int i,j,track; /* indices for the first array */
+	int k; /* indices for second array */
+	int min;
+	
+	int d, outer_d; /* d => direction */
+
+	int count, next_count; /* how many in this list */
+
+	int base, limit, last;
+
+	unsigned int* temp_pointer;
+	unsigned int temp;
+
+	unsigned int* aux_data;
+	unsigned int* aux;
+	unsigned int *source, *target;
+
+	/* get the address we need*/
+	unsigned int minusA = ((1 << BLOCK_BITS) - get_index(a));
+	aux_data = memalign(ALIGNMENT, (N + 2*LIMIT) * sizeof(unsigned int));
+	aux = (unsigned int*)(((unsigned int)aux_data & (~BLOCK_AND_LINE_MASK)) | (minusA << LINE_BITS));
+	if (aux < aux_data) /* then the new index is less than the old one */
+	{
+		aux = (unsigned int*)((unsigned int)aux + (1 << (BLOCK_AND_LINE_BITS)));
+	}
+
+
+	describe_predictor(&global_predictor[0], "presort 0");
+	describe_predictor(&global_predictor[1], "presort 1");
+	describe_predictor(&global_predictor[2], "presort 2");
+	describe_predictor(&global_predictor[3], "presort 3");
+	describe_predictor(&global_predictor[4], "presort 4");
+	describe_predictor(&global_predictor[5], "presort 5");
+	describe_predictor(&global_predictor[6], "presort 6");
+	describe_predictor(&global_predictor[7], "presort 7");
+	describe_predictor(&global_predictor[8], "presort 8");
+	describe_predictor(&global_predictor[9], "presort 9");
+	describe_predictor(&global_predictor[10], "presort 10");
+	describe_predictor(&global_predictor[11], "presort 11");
+	describe_predictor(&global_predictor[12], "forwards left 0");
+	describe_predictor(&global_predictor[13], "forwards left 1");
[…]

### 2005-05-26 utils.c

@@ -1,0 +1,158 @@
+#include <stdio.h> // printf, putchar
+#include <stdlib.h> // atoi, malloc, free
+#include <limits.h>
+#include <unistd.h>
+#include <fcntl.h>
+
+#include "utils.h"
+
+#define CONSOLE_WIDTH 10
+
+//returns 0 is sorted
+int check_sorted(unsigned int a[], int l, int r)
+{
+	int i;
+	for(i = l+1; i <= r; i++)
+	{
+		if (a[i-1] > a[i])
+		{
+			printf("NOT SORTED: a[%d] > a[%d] (%d > %d)\n", (i-1), i, a[i-1], a[i]);
+			return -1;
+		}
+	}
+	return 0;
+}
+
+#define RESET		0
+#define BRIGHT 		1
+#define DIM		2
+#define UNDERLINE 	3
+#define BLINK		4
+#define REVERSE		7
+#define HIDDEN		8
+
+#define BLACK 		0
+#define RED		1
+#define GREEN		2
+#define YELLOW		3
+#define BLUE		4
+#define MAGENTA		5
+#define CYAN		6
+#define	WHITE		7
+
+void 
+setcolor(int attr, int fg, int bg)
+{
+	printf("%c[%d;%d;%dm", 0x1B, attr, fg + 30, bg + 40);
+}
+
+void
+highlight_text()
+{
+	setcolor(BRIGHT, RED, BLACK);
+}
+
+void
+highlight_text2()
+{
+	setcolor(BRIGHT, CYAN, BLACK);
+}
+
+void
+reset_text()
+{
+	setcolor(RESET, WHITE, BLACK);
+}
+
+
+// prints an array of size count, with each member sized size, prefix by description, and highlighting the bits between left and right(inclusive)
+void
+print_int_array(unsigned int a[], int N, const char* description, int left, int right, int pivot)
+{
+	int i;
+	printf("%s:\n\t", description);
+	highlight_text2();
+	for(i = 0; i < CONSOLE_WIDTH; i++) // print the row numbers
+	{
+		printf("%8d  ", i);
+	}
+	printf("\n\n0\t");
+	reset_text();
+
+	for(i = 0; i < N; i++)
+	{
+//		if (i == pivot) highlight_text2();
+//		else if ((i >= left) && (i <= right)) highlight_text();
+
+		printf("%8u", (a[i]));
+		
+//		if ((i == pivot) || ((i >= left) && (i <= right)))  reset_text();
+		
[…]

### 2005-05-26 paper/docs/report.tex
make way for the new and improved fs
an extra file snuck ast me making changes for the updated listings package

--- 2005-05-09 docs/report.tex
@@
 \documentclass[dvips]{report}
+\usepackage{ifthen}
 \usepackage{varioref}
@@
 
-\lstset{language=c,tabsize=3,basicstyle=\small,visiblespaces=false,visibletabs=false,stringspaces=false,labelstep=1,labelstyle=\small,frame=tblr}
+\newcommand{\sixplot}[7]{
+\afterpage{
+\setcounter{subfigure}{0}
+\thispagestyle{empty}
+\clearpage
+\enlargethispage{14em}
+\vspace*{-5em}
+\begin{figure}[H]
+\begin{changemargin}
+\subfigure[]{\includegraphics[scale=0.74]{../counters/#1}}
+\ifthenelse{\equal{#2}{}}{}{\subfigure[]{\includegraphics[scale=0.74]{../counters/#2}}}
+\ifthenelse{\equal{#3}{}}{}{\subfigure[]{\includegraphics[scale=0.74]{../counters/#3}}}
+\ifthenelse{\equal{#4}{}}{}{\subfigure[]{\includegraphics[scale=0.74]{../counters/#4}}}
+\ifthenelse{\equal{#5}{}}{}{\subfigure[]{\includegraphics[scale=0.74]{../counters/#5}}}
+\ifthenelse{\equal{#6}{}}{}{\subfigure[]{\includegraphics[scale=0.74]{../counters/#6}}}
+\end{changemargin}
+\vspace*{6em}
+\caption{Branch prediction performance for #7. All graphs use the same scale.}
+\label{Branch prediction performance for #7}
+\end{figure}
+\newpage
+}}
+
+\lstset{language=c,tabsize=3,basicstyle=\small,showspaces=false,showtabs=false,showstringspaces=false,stepnumber=1,numberstyle=\small,frame=tblr}
 
@@
 \hyphenation{La-Marca}
+
+%TODO make the page wider
+%TODO aspell in UK english
 
@@
 
-\chapter*{Acknowledgements}
+\chapter*{Acknowledgments}
 I would like to thank my supervisor, Dr. David Gregg, for his help and advice
@@
 
+\chapter{Shellsort}
+
+\input shell.tex
+
 \chapter{Conclusions}
@@
 
-%\appendix
+\appendix
 
-%\chapter{Simulation Result Listing}
-%\input result_list.tex
+\chapter{Simulation Result Listing}
+\input result_list.tex
 
-%\chapter{Code Listing}
-%\input code_list.tex
+%\chapter{Software Predictor Result Listing}
+%\input software_predictor.tex
 
-%\chapter{Bug List}
-%\input bugs.tex
+\chapter{Code Listing}
+\input code_list.tex
 
-%\chapter{SimpleScalar installation}
-%\input simplescalar.tex
+% the bugs are now fixed (actually, one of them was only mistakenly a bg :)
+% the only remaining thing is the compulsory cache subtraction.
+\chapter{Bug List}
+\input bugs.tex
+
+\chapter{SimpleScalar installation}
+\input simplescalar.tex
 

### 2005-05-27 paper/docs/background.tex
make way for the new and improved fs
removed files not for paper removed extra bits

--- 2005-05-09 docs/background.tex
@@
 A \n{stable} sort is one in which records which have the same key stay in the
-same order across the sort. All the elementary sorts are stable, as is
-mergesort, heapsort and radixsort. Quicksort is not stable.
+same order across the sort. All the elementary sorts are stable, as are
+mergesort and radixsort. Quicksort, heapsort and shellsort are not stable.
 
@@
 sound and graphics cards, and storage devices such as hard drivers and floppy
-disk controllers.}. Several optimisations in later chapters use the address to
+disk controllers.}. Several optimizations in later chapters use the address to
 offset arrays so that conflicts are not caused. However, because a program

### 2005-05-27 paper/docs/report.tex
removed files not for paper removed extra bits

@@
 \title{Fast Sorting on Modern Computers}
-\author{Paul Biggar\\ B.A. (Mod.) Computer Science \\ Final Year Project April 2004 \\ Supervisor: Dr. David Gregg}
+\author{Paul Biggar, David Gregg}
 \date{}
@@
 
-\begin{titlepage}
-        \begin{center}
-                \vspace*{95mm}
-                \LARGE Fast Sorting on Modern Computers\\
-                \vspace{1em}
-                \large Paul Biggar\\
-                \large B.A. (Mod.) Computer Science \\
-                \large Final Year Project April 2004\\
-                \large Supervisor: Dr. David Gregg\\
-        \end{center}
-\end{titlepage}
-
 %\maketitle
-
-\chapter*{Declaration}
-
-I hereby declare that this thesis is entirely my own work and that it has not
-been submitted as an exercise for a degree at any other university.
-
-\begin{center}
-\vspace*{2in}
-
-\underline{\hspace*{3in}} \today
-
-Paul Biggar
-\end{center}
-
-\chapter*{Acknowledgments}
-I would like to thank my supervisor, Dr. David Gregg, for his help and advice
-during this project. I'd also like to thank those who read my first draft, and
-my friends and family for their support during the last year, and my entire time
-at college. Thank you.
-\newpage
 
@@
 
-%\chapter{Software Predictor Result Listing}
+%TODO \chapter{Software Predictor Result Listing}
 %\input software_predictor.tex
 
-\chapter{Code Listing}
-\input code_list.tex
-
-% the bugs are now fixed (actually, one of them was only mistakenly a bg :)
+% TODO the bugs are now fixed (actually, one of them was only mistakenly a bg :)
 % the only remaining thing is the compulsory cache subtraction.
@@
 \input bugs.tex
-
-\chapter{SimpleScalar installation}
-\input simplescalar.tex
 

### 2005-05-27 paper/docs/conclusions.tex
make way for the new and improved fs
removed files not for paper removed extra bits
done some spellchecking

--- 2005-05-09 docs/conclusions.tex
@@
 The quicksort results were also very similar, though the multi-quicksort results
-had a slightly different shape to LaMarca's. However, the instruction count
-results for each quicksorts agreed with LaMarca's results, as did cache miss
+had a very different shape to LaMarca's. However, the instruction count
+results for each quicksort agreed with LaMarca's results, as did cache miss
 results of the base and cache-conscious versions.
@@
 sorts and their derivatives, measured using Pentium 4 performance counters.
-Neither the elementary sorts nor heapsorts are on the chart, as their results
+None of the elementary sorts or heapsorts are on the chart, as their results
 are significantly worse than the other results, which makes it difficult to
-accurately see those results.
+accurately see them.
 
@@
 quicksort, though radixsort is very close, and the radixsort considered here was
-not optimised. It is expected that an optimised radixsort would beat quicksort
+not optimized. It is expected that an optimized radixsort would beat quicksort
 for smaller sets.
+% TODO redo above paragraph
 

### 2005-05-27 paper/docs/heap.tex
done some spellchecking

@@
 pretend the heap is slightly smaller, and sift up the final key later, as it
-might be with a staticly sized heap.
+might be with a statically sized heap.
 
@@
 
-In the graph, the base heapsort results flat-line for a step longer than the
+In the graph, the base heapsort's results flat-line for a step longer than the
 other heapsorts' results. This is due to being in-place, since twice as many

### 2005-05-27 paper/docs/merge.tex
done some spellchecking

@@
 
-The final impriovement was to fully sort keys in the level 1 cache, before
+The final improvement was to fully sort keys in the level 1 cache, before
 continuing to level 2 cache as in the tiled mergesort. The expected improvement
@@
 
-The reuslts show that the seconsary loops are always taken. Taken here means broken out of, hence seldom used. The main action occurs in N3, the main comparator, which has a 50\% prediction ratio.
+The results show that the secondary loops are always taken. Taken here means broken out of, hence seldom used. The main action occurs in N3, the main comparator, which has a 50\% prediction ratio.
 
@@
 Figure c shows base mergesort.
-This shows a lower number of branches and a higher precentage of correct predictions.
+This shows a lower number of branches and a higher percentage of correct predictions.
 The lower number of branches is probably due to the presort.
@@
 This implies that doing the presort removes incorrect predictions (or, more to the point, moves incorrect predictions to another part of the algorithm).
-Note too, that both sides are used equally, which doesnt happen later.
+Note too, that both sides are used equally, which doesn't happen later.
 Two thirds of the branches are not-taken, which means loops execute twice before breaking.
 this is the same results that we found in quicksort later.
-Im sure knuth explains this somewhere in the fine print.
+I'm sure knuth explains this somewhere in the fine print.
 

### 2005-05-28 paper/docs/background.tex
after spell check
spell checked updateed intro read through method removed extras, which was part of the demo added labels

@@
+\label{background}
 \section{Sorting}
@@
 be sorted. This may require the results to be copied back at the end. An
-in-place sort does not, though it may use a large stack, especially in recursive
-algorithms.
+in-place sort does not require this extra array, though it may use a large
+stack, especially in recursive algorithms.
 
 A \n{stable} sort is one in which records which have the same key stay in the
-same order across the sort. All the elementary sorts are stable, as are
+same relative order during the sort. All the elementary sorts are stable, as are
 mergesort and radixsort. Quicksort, heapsort and shellsort are not stable.
@@
 These are replaced with their actual functions by the C preprocessor. The code
-used for ADTs is in Figure \vref{Definitions used by the ADT}. Items in this
-project were \cc{unsigned integer}s, and were compared using a simple
-\n{less-than} function\footnote{This abstraction broke down in several cases,
-such as where a \n{greater-than} function was required, such as in quicksort}.
-In this case, a record and key are equivalent, and the word \n{key} is used to
-refer to items to be sorted.
+used for ADTs is in Figure \vref{Definitions used by the ADT}. The tests
+presented in this report use \cc{unsigned integer}s as Items, which were
+compared using a simple \n{less-than} function\footnote{This abstraction broke
+down in several cases, such as where a \n{greater-than} function was required,
+i.e. in quicksort.}.  In this case, a record and key are equivalent, and the
+word \n{key} is used to refer to items to be sorted.
 
@@
 array before putting them in place. The most important of these is radixsort,
-discussed in Chapter \ref{Radixsort}.
+discussed in Chapter \ref{radix}.
 
@@
 increasing size and \n{latency}\footnote{Latency is the time between a request
-being made, and its successful execution.}. For example, it may take twice as
+being made and its successful execution.}. For example, it may take twice as
 long to access the level 1 cache as it does to access a register. It may take
@@
 branch. Frequently, the program counter is also included, either concatenated or
-exclusive-ored with the branch history to form the table index.
+exclusive-ORed with the branch history to form the table index.
 
@@
 later chapters.
+%TODO what's shellsort in Big O notation
+TODO what's shellsort in Big O notation
 

### 2005-05-28 paper/docs/conclusions.tex
spell checked updateed intro read through method removed extras, which was part of the demo added labels

@@
+\label{conclusions}
 This section presents a summary of the results from previous chapters, lists

### 2005-05-28 paper/docs/heap.tex
spell checked updateed intro read through method removed extras, which was part of the demo added labels

@@
+\label{heap}
 %\section{Description}
@@
 
-\subsubsection{Out-of-place sorting}
+\subsubsection{Out-of-place Sorting}
 Up until this point, the heap property used has been that a parent will be
@@
 
-\section{Base heapsort}
+\section{Base Heapsort}
 LaMarca recommends to use of the Repeated-Adds algorithm for reducing

### 2005-05-28 paper/docs/intro.tex
spell checked updateed intro read through method removed extras, which was part of the demo added labels

@@
+\label{intro}
 Most major sorting algorithms in computer science date to the 50s and 60s.
@@
 LaMarca and Ladner \cite{LaMarca96} break this trend by discussing the cache
-implications of several algorithms, including sorts. This project attempts to
-investigate these claims and to test and analyse the high performance sorts
-based on them.
+implications of several algorithms, including sorts. This report 
+investigates these claims and tests and analyses high performance sorts
+based upon thier work.
 
-LaMarca devotes a chapter of his thesis to sorting. A chapter is spent
+LaMarca devotes a chapter of his thesis to sorting. Another chapter is spent
 discussing implicit heaps, which are later used in heapsort. In
@@
 
-In addition to repeating LaMarca's work, this project also discusses hardware
+In addition to repeating LaMarca's work, this report also discusses hardware
 branch prediction techniques and their effects on sorting. When algorithms are
@@
 
-Several important results are discovered during the course of the project. The
+Several important discovered results are reported here. The
 branch prediction rates of insertion sort and selection sort, for example, and
@@
 to branch misses. Several steps to remove instructions from standard algorithms
-are devised, such as double tiling mergesort, and removing a sentinel and bounds
+are devised, such as double tiling mergesort, removing a sentinel and bounds
 check from heapsort.
 
-Chapter 2 provides background information on caches and branch predictors. It
+Chapter \ref{background} provides background information on caches and branch predictors. It
 also discusses the theory of sorting, and conventions used in the algorithms in
@@
 
-Chapter 3 discusses the method and framework used in this project, including the
+Chapter \ref{method}  discusses the method and framework used in this project, including the
 tools used and the reasons for using them.
 
-Chapter 4 discusses the elementary sorts and their implementations. Results are
+Chapter \ref{ordernsquared}  discusses the elementary sorts and their implementations. Results are
 included for later comparison.
 
-Chapters 5 through 7 discuss base algorithms and cache-conscious improvements
+Chapters \ref{heap} through \ref{quick} discuss base algorithms and cache-conscious improvements
 for heapsort, mergesort and quicksort. Results are presented and explained, and
@@
 
-Chapter 8 presents a radixsort implementation and discusses its performance in
+Chapter \ref{radix} presents a radixsort implementation and discusses its performance in
 the areas of instruction count, cache-consciousness and branch prediction.
 
-Chapter 9 presents conclusions and contributions developed over the course of
-this project.
+%TODO shell
+TODO Chapter \ref{shell}
+
+Chapter \ref{conclusions} presents conclusions, and lists contributions
+described by this report.

### 2005-05-28 paper/docs/merge.tex
spell checked updateed intro read through method removed extras, which was part of the demo added labels

@@
 %\section{Description}
-\label{Mergesort}
+\label{merge}
 Merging is a technique which takes two sorted lists, and combines them into a
@@
 \subsection{Base Mergesort}
+\label{base mergesort}
 Three optimisations are described by LaMarca for the base mergesort. Firstly, a

### 2005-05-28 paper/docs/method.tex
spell checked updateed intro read through method removed extras, which was part of the demo added labels

@@
+\label{method}
+%TODO rewrite this
 This section discusses the method and framework used to complete this project.
@@
 \section{Method}
-
+i
+%TODO remove this
 The following is the initial project description:
@@
 computer was being used by multiple users and processes, leading to context
-switches,sharing of resources and scheduling. Despite these inaccuracies, it was
+switches, sharing of resources and scheduling. Despite these inaccuracies, it was
 possible to use the results relative to each other, and to determine whether,
@@
 
-This shows two stages of a bitonic mergesort. The top of the image shows the
-source array being merged into the target array below. The source array shows
-three large segments of the array and four smaller segments. The first two large
-segments have been merged into one segment in the target array. The third of
-these has been merged into the corresponding target array segment. The first of
-the smaller segments has begun being merged with the other small segments. The
-rest of the target array is uninitialised.  Below that is a small amount of
-debugging information. $i == j$ is a key point in the program and is printed as
-debugging information. In the two arrays below this, the merge continues.
+This shows two stages of a bitonic mergesort which is described in Section \ref{base
+mergesort}. The top of the image shows the source array being merged into the
+target array below. The source array shows three large segments of the array and
+four smaller segments. The first two large segments have been merged into one
+segment in the target array. The third of these has been merged into the
+corresponding target array segment. The first of the smaller segments has begun
+being merged with the other small segments. The rest of the target array is
+uninitialised.  Below that is a small amount of debugging information. $i == j$
+is a key point in the program and is printed as debugging information. In the
+two arrays below this, the merge continues.
 
@@
 Descriptions of some of the terms here is saved for discussion in Chapter
-\ref{Mergesort}.
+\ref{merge}.
 
@@
 A tiled mergesort was then written, based on LaMarca's guidance. Firstly, this
-was aligned in memory so that the number misses due to conflicts would be
+was aligned in memory so that the number of misses due to conflicts would be
 reduced.  Then it fully sorted segments of the array, each half the size of the
@@
 reduce the instruction count, as well as the level 2 cache miss count. These
-results are discussed in more detail in \ref{Mergesort results}.
+results are discussed in more detail in Section \ref{Mergesort results}.
 
@@
 \section{Future work}
-SimpleScalar ran very slowly, with each simulation run taking at least $2
-\frac{1}{2}$ hours. Several steps could be taken to lower this time. Firstly,
-compiling SimpleScalar with the \cc{-03} flag, which performs a significant
-number of optimisations on the program. This was not done because of the
-difficulty of compiling the program in the first place.
-
-\cc{lib-cheetah} is a SimpleScalar library used to simulate multiple
-caches in the same simulation. Over half the simulation time was taken by four
-cache simulations, and running them concurrently might have significantly
-decreased the running time. This was not a significant worry during this project,
-but decreasing the running time would be necessary if the either the number of
-simulations or the number of sorts was to change dramatically.
-
 An improvement that would have yielded interesting results would be to do simulations
@@
 division of flow control misses and comparative misses. However, this would not
-always be possible; sorts like quicksort and heapsort mix use their data to
+always be possible; sorts like quicksort and heapsort use their data to
 control the flow of the program. In addition, bubblesort, shakersort and
 selection sort have (analytically) predictable flow control, while radixsort has
-no misses at all. Mergesort may benefit from this, however, especially versions
+no comparison misses at all. Mergesort may benefit from this, however, especially versions
 with difficult flow control diagrams, such as \n{algorithm N} and \n{algorithm

### 2005-05-28 paper/docs/method.tex
cleaned up the method

@@
 \label{method}
-%TODO rewrite this
-This section discusses the method and framework used to complete this project.
-From the point of view of a Final Year Project, this is a description of what I
-did all year, what I spent my time on, what went right and wrong during the
-project, and what I learned. The first section discusses the steps involved in
-creating the project, programming the sorts and getting results. The second
-section discusses tools that I used or created in the course of the project.
+This section discusses the method and framework used to complete this research.
+The first section discusses the steps involved in creating a testing framework,
+programming the sorts and getting results. The second section discusses tools
+that I used or created to do this.
 
 \section{Method}
-i
-%TODO remove this
-The following is the initial project description:
-\begin{quotation}
-Sorting is one of the most important and studied problems in computer science.
-Many good algorithms exist which offer various trade-offs in efficiency,
-simplicity, and memory use.  One weakness of classical sorting algorithms is
-that they are not designed for modern computers.
 
-Cache memories, pipelines, and branch prediction have greatly changed the
-relative strengths of different algorithms. For example, comparison based
-sorting algorithms (such as Quicksort) can be inefficient on modern machines,
-because they involve difficult to predict branches.
-
-The goal of this project is to design, implement and experiment with sorting
-algorithms designed to make the best use of modern architectural features.
-Several cache-conscious sorting algorithms have been proposed, and the first
-step would be to implement these algorithms and compare their performance with
-traditional algorithms.  The next step is to consider modifications to the
-algorithms to make better use of the processor pipeline and multiple execution
-units and reduce branch mispredictions. The project will also involve a detailed
-study of the behaviour of the algorithms using a processor simulator or
-performance counters on a real machine.
-
-To do this project, you should have a reasonable knowledge of computer
-architecture and traditional sorting algorithms. You should also be enthusiastic
-about algorithm design, and making things go faster.  All programming is in C or
-C++.
-\end{quotation}
-After discussion with my supervisor, the following project goals were agreed upon:
+The following were the objectives of the research:
 \begin{itemize}
-\item Implement heapsort, mergesort, quicksort, radixsort, and the four
+\item Implement heapsort, mergesort, quicksort, radixsort, shellsort and the four
 elementary sorts: insertion sort, selection sort, bubblesort and shakersort.
@@
 and change in branch prediction accuracy as a result of these changes.
+\item Determine if any changes can be made to improved the sorts' performance.
 \end{itemize}
 
-Several steps were necessary to prepare for the project. The first of these was
-to compile SimpleScalar. A guide to repeating this is included in Section
-\ref{SimpleScalar Install}. A framework was required by which I could test sorts
-as they were developed. This is discussed in Section \ref{Main}. It was also necessary
-to fully understand the technologies to be used in the project. Several texts
-were used for this: cache and branch predictors are both in
-\cite{HennessyPatterson90}; branch predictors in
-\cite{Uht97}, as well as in \cite{McFarling93}
-and \cite{Smith81}; sorts are discussed in \cite{Sedgewick02} and
-\cite{Knuth98}; cache-conscious sorts are discussed \cite{LaMarca96},
-\cite{LaMarca99} and \cite{LaMarcaHeap96}.
+% TODO maybe remoe this?
+Several steps were necessary to prepare for the research.  A framework was
+required by which sorts could be tested as they were being developed. This is
+discussed in Section \ref{Main}. It was also necessary to fully understand the
+technologies to be used in the project. Several texts were used for this: cache
+and branch predictors are both in \cite{HennessyPatterson90}; branch predictors
+in \cite{Uht97}, as well as in \cite{McFarling93} and \cite{Smith81}; sorts are
+discussed in \cite{Sedgewick02} and \cite{Knuth98}; cache-conscious sorts are
+discussed \cite{LaMarca96}, \cite{LaMarca99} and \cite{LaMarcaHeap96}.
 
-There were several steps involved in writing each iteration of the algorithms.
-The elementary sorts and radixsort had just one of these iterations; the other sorts had more
-than one\footnote{Quicksort, for example, had four: base quicksort, tuned
-quicksort, multi-quicksort and sequential multi-quicksort.}. A flowchart of these
-steps appears in Figure \vref{Flowchart}.
+There were many steps involved in writing each iteration of the algorithms.
+Insertion and selection sort had just one of these iterations; the other
+sorts had more than one\footnote{Quicksort, for example, had four: base
+quicksort, tuned quicksort, multi-quicksort and sequential multi-quicksort.}. A
+flowchart of these steps appears in Figure \vref{Flowchart}.
 
+%TODO this is going to have to be updated
 \begin{figure}
@@
 
-The first step was to write the algorithm. \cite{Sedgewick02} was the primary
+The first step in each case was to write the algorithm. \cite{Sedgewick02} was the primary
 source of these, though in several cases LaMarca specified a particular version
@@
 a certain range was implied by LaMarca's descriptions. The relationship between
-one iteration of an algorithm, or between base versions of several algorithms was
-known ahead of time. One configuration of the \cc{simple} script (see
-Section \ref{Simple Script}) ran quick tests using the \cc{sim-fast} simulator,
-giving instruction counts which could be compared to other versions.
+one iteration of an algorithm, or between base versions of several algorithms
+was known ahead of time. One configuration of the \cc{simple} script ran quick
+tests using the \cc{sim-fast} simulator, giving instruction counts which could
+be compared to other versions.
 
@@
 More details of this are in Section \ref{Gnuplot}. The next iteration of the
-algorithm was then begun, and the process repeated for each of the
-algorithm's iterations.
+algorithm was then begun, and the process repeated for each of the algorithm's
+iterations.
 
@@
 
-\subsection{SimpleScalar}
-\cc{SimpleScalar} is an architecture simulator. It works by compiling code into its
-native instruction set, PISA, then emulating a processor as it runs the program.
-SimpleScalar can be found at \n{http://www.simplescalar.com}. Documentation on how
-to use and modify SimpleScalar are \cite{Burger97}, \cite{Austin02},
-\cite{Tutorialv4}, \cite{Tutorialv2} and \cite{SimpleScalarUserGuide}.
-
-The installation of SimpleScalar can be difficult. A step by step description
-for installing it on an x86 linux machine is in Section \ref{SimpleScalar
-Install}.
+\subsection{SimpleScalar} \cc{SimpleScalar} is an architecture simulator. It
+works by compiling code into its native instruction set, PISA, then emulating a
+processor as it runs the program.  SimpleScalar can be found at
+\n{http://www.simplescalar.com}. Documentation on how to use and modify
+SimpleScalar are \cite{Burger97}, \cite{Austin02}, \cite{Tutorialv4},
+\cite{Tutorialv2} and \cite{SimpleScalarUserGuide}.
 
 SimpleScalar works by compiling code into its own instruction set. It provides a
-compiler - a port of gcc 2.63 - with related tools such as a linker and assembler.
-Several simulators are then provided: \cc{sim-fast} only counts instruction,
-quickly; \cc{sim-cache} simulates a cache and counts level 1 and level 2 cache
-misses; \cc{sim-bpred} simulates a branch predictor and counts branch hits and
-misses. Other simulators are provided, but they were not used in this project.
+compiler - a port of gcc 2.63 - with related tools such as a linker and
+assembler.  Several simulators are then provided: \cc{sim-fast} only counts
+instruction, quickly; \cc{sim-cache} simulates a cache and counts level 1 and
+level 2 cache misses; \cc{sim-bpred} simulates a branch predictor and counts
+branch hits and misses. Other simulators are provided, but they were not used in
+this project.
 
-The output of SimpleScalar is human readable (see Section \ref{Sample
-Simulation} for an example) but is not conducive to collecting large amounts of
-data. To that end, a script was written to run each required set of simulations. The
-script was written in perl, and recorded data from the simulations into a single
-file for later retrieval. The source code to the script, \cc{simple}, is in
-section \ref{Simple Script}.  A sample list of commands executed by \cc{simple} are
-in section \ref{Simple List}. The data file produced was suitable to reading
-with a \cc{diff} program.
+The output of SimpleScalar is human readable, but is not conducive to collecting
+large amounts of data. To that end, a script was written to run each required
+set of simulations. It recorded data from the simulations into a single file for
+later retrieval. This file was suitable to reading with a \cc{diff} program.
 
@@
 scaled to so that the y-axis could be in the form of \n{plotted data per key},
-both to be more readable, and to be consistent with LaMarca's graphs. The source
-code to the script to do this is in Figure \ref{Convert Script}. Once the data
-was converted, a script to run gnuplot (see Section \ref{Sample Gnuplot Script}
-for an example) was required. The number of graphs required made this an
+both to be more readable, and to be consistent with LaMarca's graphs. Once the data
+was converted, it was necessary to generate a gnuplot script for each graph.
+The number of graphs required made this an
 impossible manual task, and so another script had to be written.  Some graphs
@@
 
-\subsection{Incidental Tools}
-The algorithms in this project were written in C using \cc{vim}, a powerful text
-editor. \cc{gcc} was used to compile the framework, when not using SimpleScalar.
-The report was written using \LaTeX, and diagrams were written is
-\cc{postscript}, or drawn using \cc{OpenOffice.org}, an open-source office
-suite. Many files had to be edited in the same way, and this was accomplished
-either using \cc{vim} macros, or by writing a short \cc{perl} script.
-
 \section{Implementation of sorts}
-
-The \cc{C} code written for this project was approximately 8000 lines long.
-Appendix \ref{Sort Listings} lists many of the key parts of this code.
 
@@
 
-Each of the three $O(NlogN)$ sorts was treated in this fashion. The mergesorts
-accounted for over 3000 lines of the 8000 lines of code in this project. Key
-portions of this code are shown in Section \ref{Sort Listings}.
+%TODO another sentence here
+Each of the four $O(NlogN)$ sorts and radixsort, was treated in this fashion.
 

### 2005-05-30 paper/docs/ordernsquared.tex
spell checked updateed intro read through method removed extras, which was part of the demo added labels
finished 4 chapters

@@
+\label{ordernsquared}
 Elementary sorts are also referred to as $O(N^2)$ sorts. This is due to the
@@
 
-The level 1 cache should have equally bad performance as a result. Since it is
-smaller than the array, the performance of the level 1 cache should simulate the
+The level 1 cache should have bad performance as a result. Since it is smaller
+than the array, the performance of the level 1 cache should simulate the
 performance of the level 2 cache.
@@
 
+%TODO look at knuths number here
+TODO Knuth has a formula which I will look up.
 \section{Insertion sort}
@@
 Bubblesort2 incorporates an improvement to end bubblesort early. The bubblesort
-implementation above ends early if it detects that the array is fully sorted.
-It keeps a binary flag to indicate if there were any exchanges along the way.
-Instead, bubblesort2 keeps track of the last exchange made, and starts the next
-iteration from there. This means that any pockets of sorted keys are skipped
-over, not just if they occur at the end. Bubblesort2 code is in Figure
-\vref{Bubblesort2 code}
+implementation described in the previous section ends early if it detects that
+the array is fully sorted.  It keeps a binary flag to indicate if there were any
+exchanges along the way.  Instead, bubblesort2 keeps track of the last exchange
+made, and starts the next iteration from there. This means that any pockets of
+sorted keys are skipped over, not just if they occur at the end. Bubblesort2
+code is shown in Figure \vref{Bubblesort2 code}.
 
@@
 become sorted in this manner, with the larger keys slowly moving right and the
-smaller keys moving left. Sample code for shakersort is in Figure
+smaller keys moving left. Sample code for shakersort is shown in Figure
 \vref{Shakersort code}.
@@
 probably slightly reduce the instruction count as the sort is unlikely to end
-early for random keys. In addition, the worst case for bubblesort is not a
-worst case for shakersort and is less likely.
+early for random keys.
 
@@
 array is not more than twice the size of the cache, then the keys in the centre
-of the array are not ejected from the array at all.
+of the array are not ejected from the cache at all.
 
@@
 
+%TODO check that this is put in after the titlepage
+%TODO check that the captions are the same as the captions at the end of the
+%other chapters
 % I'm doing a slightly different result here. Level 2 doesn't interest us, so put
@@
 
-Insertion sort performs best of the four sorts in relation to cache misses. This
+Insertion sort performs best of these sorts in relation to cache misses. This
 is due to the fact that the keys it considered most recently are used again for

### 2005-05-30 paper/docs/heap.tex
added blurb and sample code for the software predictors

@@
 Heapsort is an in-place sort which works in $O(NlogN)$ time in the average and
-worst case. It begins by creating a structure known as a \n{heap}, which stores
-its largest key at the top, then removing keys from the top and placing them in
-their final position in the array.
+worst case. It begins by creating a structure known as a \n{heap}, a binary tree
+in which a parent is always larger its child, and the largest key is at the top
+of the heap. It then repeatedly removes the largest key from the top and places
+it in its final position in the array, fixing the heap as it goes.
 
@@
 referred to as fixing up and down, and algorithms that do this are called
-\n{fix-up} or \n{sift-up}, and \n{fix-down} or \n{sift-down}} keys up the heap.
+\n{fix-up} or \n{sift-up}, and \n{fix-down} or \n{sift-down}.} keys up the heap.
 Assuming that a heap already exists, adding a key and re-heapifying will
@@
 
-The second step is to destroy the heap, leaving a sorted array.
+The second step, described next, is to destroy the heap, leaving a sorted array.
 
@@
 relatively simply with no extra instructions. If the last key in the heap is
-left in the heap while its being sifted down the heap, it will provide a
-sentinel for the sift down operation. The key from the top of the heap is then
+left in place aswell as being sifted down the heap, it will provide a
+sentinel for the sift-down operation. The key from the top of the heap is then
 put into place at the end. This technique reduces instruction count by
@@
 touched when a new key is added to the heap. In the case of Floyd's method, the
-keys being heapified are not related to the last keys that were heapified. Reuse
+key being heapified is not related to the last key that was heapified. Reuse
 only occurs upon reaching a much higher level of the heap. For large heaps, these
@@
 The Repeated-Adds algorithm, however, displays excellent temporal locality. A
-key has a 50\% chance of having same parent, a 75\% of the same grandparent and
-so on. The number of instructions, however, increases significantly, and the
-sort becomes out-of-place.
+key has a 50\% chance of having the same parent, a 75\% of the same grandparent
+and so on. The number of instructions, however, increases significantly, and the
+sort becomes out-of-place. For this reason, the Repeated-Adds algorithm is used
+for cache heapsort.
 
@@
 destroying it. Less than 10\% of heapsort's time is spent building the heap.
-Therefore optimisations must be applied when destroying the heap.
+Therefore optimizations must be applied when destroying the heap.
 
@@
 this increase will be an order of magnitude lower than the bounds check.
+
+%TODO this isnt true is it? placing the nsentinel is cheaper than a bounds check
+%(a write to a variable thats going to be written to in a second vs a comparison?)
+%and there arent an extra logN instructions, theres logN times as many
 
@@
 keys fit in the cache without a corresponding increase in misses. For the same
-reason, the base heapsorts have half as many level 2 cache misses of the later,
-out-of-place sorts (See Figure \ref{heap level 2 misses}).
+reason, the base heapsorts have approximately half as many level 2 cache misses
+as the out-of-place sorts (See Figure \ref{heap level 2 misses}).
 
@@
 the case of the 2- and 8-heaps, the bimodal predictor works better than the
-two-level adaptive predictor. The warm-up time is visible in the base heapsort,
-though it's amortised for larger data sets. There is no difference between
-predictors for the 4-heap.
+two-level adaptive predictor. The warm-up time of the table-driven predictors is
+visible in the base heapsort results, though it's amortised for larger data
+sets.
 
@@
 cache line, a 4-way associative 8KB level 1 data cache, and a maximum branch
-misprediction delay of 20 cycles.\footnote{This is determined from the Pentium
+misprediction delay of 20 cycles\footnote{This is determined from the Pentium
 4's 20 stage pipeline. This number is not exact, and is probably slightly lower,
@@
 instruction cache miss, which is exceptionally unlikely for small routines like
-sorts).} Its static branch prediction strategy is backward taken and forward
+sorts).}. Its static branch prediction strategy is backward taken and forward
 not taken; its dynamic strategy is unspecified: though \cite{Intel248966-010}
@@
 
-The figure shows that the base heapsort performs best while the data set fits in
+The figure shows that the base heapsort performs best when the data set fits in
 the cache. After that point, the 8-heap performs best, and its cycle count increases
 with a smaller slope. This shows the importance of LaMarca's cache-conscious
-design even eight years after his thesis. It also shows the importance of reduced
+design even nine years after his thesis. It also shows the importance of reduced
 branch mispredictions: the base heapsort has more instructions and level 1 cache
-misses, but still outperforms the 4-heap, before level 2 cache misses take over.
+misses, but still outperforms the 4-heap, until level 2 cache misses take over.
 However, this is hardly conclusive: the lower level 2 cache misses may account
@@
 direct mapped cache. The Pentium 4's cache is 8-way associative, and so the
-unoptimised algorithm would not suffer as much.
+unoptimized algorithm would not suffer as much.
 
-\section{Predictor results}
+\section{Predictor Results}
 
@@
 in the creation and destruction of the heap. The base heapsort results in
-Figure \ref{base and cache heapsorts-1} show that the time taken to create the
-heap is not significant relative to the time taken to sort from the heap.
+Figure \ref{base and cache heapsorts-1} show
+that the time taken to create the heap is not significant relative to the time
+taken to sort from the heap.
 
-The most interesting thing is the relation between the predictivity of the
+The most interesting thing is the relation between the predictablility of the
 comparisons in the destruction of the 2-, 4- and 8-heaps, which are shown in
@@
 comparing simulated results from there to the sorts created here might have
-interesting results.
+interesting results and would explain the discreptancies highlighted in this
+chapter.
 

### 2005-05-30 paper/docs/method.tex
added blurb and sample code for the software predictors

@@
 
-The first step in each case was to write the algorithm. \cite{Sedgewick02} was the primary
-source of these, though in several cases LaMarca specified a particular version
-which he used. When a step involved in an algorithm wasn't clear, the paper
-describing the algorithm was used to clarify the meaning.
+The first step in each case was to write the algorithm. \cite{Sedgewick02} was
+the primary source of these, though in several cases LaMarca specified a
+particular version which he used. When a step involved in an algorithm wasn't
+clear, the paper describing the algorithm was used to clarify the meaning.
 
@@
 iterations.
+
+Finally, a software predictor was added in order to pin-point the cause of the
+branch mispredictions. This meant altering the code of the algorithms to call
+functions which updated simulated branch predictors. Each important branch was
+allocated a predictor, which was updated independantly of the other predictors.
+This allowed the performance to be analysed down to the individual branch, and
+blame apportioned appropriately.
 
@@
 SimpleScalar results.
+
+\subsection{Software Predictor}
+The software predictor used was a series of macros and functions which
+maintained the state of a branch predictor at a particular point in the program.
+Statistics about each branch were collected, and dumped into text files, from
+which scripts created graphs to display them. Some sample code is shown in
+Figure \ref{Sample code using the software predictor}.
+
+\code{Sample code using the software predictor}{quick.c}
 

### 2005-05-30 paper/docs/code/quick.c

@@ -1,0 +1,39 @@
+static int
+partition(unsigned int a[], int l, int r)
+{
+	unsigned int v;
+	int i, j;
+
+	v = a[r];
+	i = l - 1;
+	j = r;
+
+	for(;;)
+	{
+		while (less(a[++i], v))
+		{
+			branch_taken(&global_predictor[0]);
+		}
+		branch_not_taken(&global_predictor[0]);
+
+		while (less(v, a[--j]))
+		{
+			branch_taken(&global_predictor[1]);
+		}
+		branch_not_taken(&global_predictor[1]);
+		
+		if (i >= j)
+		{
+			branch_taken(&global_predictor[2]);
+			break;
+		}
+		branch_not_taken(&global_predictor[2]);
+
+		exch(a[i], a[j]);
+	}
+
+	exch(a[i], a[r]);
+
+	return i;
+}
+

### 2005-05-30 paper/docs/heap.tex
finished heap section

@@
 in the creation and destruction of the heap. The base heapsort results in
-Figure \ref{base and cache heapsorts-1} show
-that the time taken to create the heap is not significant relative to the time
-taken to sort from the heap.
+Figure \ref{base and cache heapsorts-1} show that the time taken to create the
+heap is not significant relative to the time taken to sort from the heap.
 
 The most interesting thing is the relation between the predictablility of the
-comparisons in the destruction of the 2-, 4- and 8-heaps, which are shown in
-Figure \ref{Branch prediction performance for base and cache heapsorts}.
-Overall, the 4-heap is half as high as the 2-heap, and twice as high as the
-8-heap, but has to make more comparisons. However, there are three comparisons
-in the 4-heap and seven comparisons in the 8-heap. In total, there are 50\% more
-comparisons in the 4-heap, and 150\% more in the 8-heap.
+comparisons during the destruction of the 2-, 4- and 8-heaps. Compare the
+columns marked `\cc{Destruction, child comparison}' and `\cc{Destruction, child
+promotion}' in Figure \ref{base and cache heapsorts-1} (the 2-heap), with the
+columns marked `\cc{comparison 1}' to `\cc{child promotion}' in Figure \ref{base and
+cache heapsorts-2} (the 4-heap) and Figures \ref{base and cache heapsorts-3} and
+\ref{base and cache heapsorts-4} (the 8-heap).
 
-These branches become more predictable. The first predicts correctly 50\% of the
-time, while the next predicts correctly 60\% of the time, then 71\%, 77\%, and
-so on in a curve. This curve is shown in Figure TODO.
-%TODO make a pretty graph for this.
+Though the number of times each comparison is used shrinks
+as the size of the fanout increased, overall the number of branches increases.
+In total, there are 50\% more comparisons in the 4-heap than the 2-heap, and
+150\% more in the 8-heap.
 
-Also notable from the results is that branches are more often not-taken then
-they are correctly predicted, implying that a semi-static predictor based on
-hints from the compiler would do a better job at making correct predictions.
+More importantly, the accuracy of the each
+successive comparison increases in the same way as selection sort (see Section
+\ref{selection branches logn}), becoming more predictable as the fanout
+increases. For example, in the 8-heap, shown in Figures \ref{base and cache heapsorts-3}
+and \ref{base and cache heapsorts-4},
+the first comparison is predicted correctly 50\% of the
+time, while the next is predicted correctly 61\% of the time, then 71\%, 77\%, and
+so on in a curve, which is shown in Figure \ref{cache heapsort curve}.
+
+\begin{figure}
+\includegraphics{../counters/curve.eps}
+\caption{Predictability of branches in cache heapsort, with a fanout of eight}
+\label{cache heapsort curve}
+\end{figure}
+
+\label{heapsort entropy}
+We observe here that the more branches there are, the more are correctly
+predicted, and that there is a threshold below which the number of branches will
+not drop. This is discussed in more Section \ref{quicksort entropy}.
+
+Also notable from the results is that branches are slightly more often not-taken
+then they are correctly predicted, implying that a semi-static predictor based
+on hints from the compiler could do a better job at making correct predictions.
 

### 2005-05-31 paper/docs/heap.tex
changed optimization spelling

@@
 
-The only other optimisation used is to unroll the loop while destroying the
+The only other optimization used is to unroll the loop while destroying the
 heap. This has minimal affect, however.
@@
 In the base heapsort, an improvement was added that removed the need for a
-sentinel. Since the 4- and 8-heaps were unsorted, this optimisation had to be
+sentinel. Since the 4- and 8-heaps were unsorted, this optimization had to be
 removed. Leaving the key being copied in place was still possible, but it needs

### 2005-05-31 paper/docs/merge.tex
changed optimization spelling

@@
 single sorted list. Starting at the smallest key on both lists, it writes the
-smaller key to a third list. It then compares the smallest key on one list with
-the next smallest on the other, and writes the smaller to the auxiliary list.
-The merging continues until both lists are combined into a single sorted list.
+smaller key to a third list. It then compares the next smallest key on that list
+with the next smallest on the other, and writes the smaller to the auxiliary
+list. The merging continues until both lists are combined into a single sorted
+list.
 
@@
 likewise for all other pairs. These arrays are then merged with their neighbours
-into arrays of length four, and so on, until the arrays are completely sorted.
+into arrays of length four, then eight, and so on until the arrays are
+completely sorted.
 
@@
 
-The number of merges required is $log_2(N)$, and if this number is even then
-the final merge puts the array in the correct place. Otherwise, the result will
-need to be copied back into the initial array.
+The number of merges required is $log_2(N)$, and if this number is even then the
+final merge puts the array in the correct place. Otherwise, the sorted array
+will need to be copied back into the initial array.
 
@@
 While this results in good performance, the merge is not perfectly regular, and
-it is difficult or impossible to perform several of the optimisations LaMarca
+it is difficult or impossible to perform several of the optimizations LaMarca
 describes on it.
@@
 is a very difficult algorithm to modify. Knuth describes it in assembly, and
-making the optimisations that LaMarca describes on it is more difficult than
+making the optimizations that LaMarca describes on it is more difficult than
 rewriting the algorithm more simply. The rewritten algorithm, referred to from
@@
 \label{base mergesort}
-Three optimisations are described by LaMarca for the base mergesort. Firstly, a
+Three optimizations are described by LaMarca for the base mergesort. Firstly, a
 bitonic sort, as described in \cite{Sedgewick02}, removes instructions from the
@@
 
-Overall, these three optimisations reduced the instruction count by 30\%.
+Overall, these three optimizations reduced the instruction count by 30\%.
 
@@
 simulations presented in this project address the cache by virtual address, and
-so this optimisation is effective in this case.
+so this optimization is effective in this case.
 
@@
 being the fastest. Note that the alignment of the mergesort may not work as
-desired due to the size of the cache. This was optimised for a 2MB cache, while
+desired due to the size of the cache. This was optimized for a 2MB cache, while
 the Pentium 4 cache is only 256KB. This means that the two arrays are aligned to

### 2005-05-31 paper/docs/method.tex
changed optimization spelling

@@
 As a result, it was necessary to use the original version, and to try to perform
-optimisations on it. However, several factors made this difficult or impossible.
+optimizations on it. However, several factors made this difficult or impossible.
 LaMarca unrolled the inner loop of mergesort, but it is not clear which is the
@@
 initial version, which performed the same number of instructions as the same as
-algorithm N, and the optimised version, was about 30\%.
+algorithm N, and the optimized version, was about 30\%.
 

### 2005-05-31 paper/docs/merge.tex
more fixes

@@
 slow the algorithm. This problem is shared by all other versions of mergesort
-considered here, except algorithm N. However, since this project uses exact
-powers of two as set sizes, this problem does not affect the results. LaMarca,
+considered here, except algorithm N. However, since our tests use exact
+powers of two as set sizes, this problem does not affect them. LaMarca,
 however, used set sizes such as 4096000, which at times would suffer from this
@@
 Finally, the inner loop is \n{unrolled}\footnote{Unrolling a loop means
-executing several steps of the loop between iterations, reducing the number of
-iterations, and lowering the loop overhead, as well as allowing for better
-scheduling of the instructions.}.  There are actually two inner loops, and they
-are both unrolled 8 times. The compiler is also set to unroll loops. However,
-the observed difference between unrolling manually and compiler unrolling is up
-to 10\% of instruction count, so manual unrolling is used.
+executing several steps of the loop between iterations and reducing the number
+of iterations, which lowers the loop overhead and allows for better scheduling
+of the instructions.}. There are actually two inner loops, and they are both
+unrolled 8 times. The compiler is also set to unroll loops. However, the
+observed difference between unrolling manually and compiler unrolling is up to
+10\% of instruction count, so manual unrolling is used.
 
@@
 C$.
+
+%TODO what does this formula mean. look up LaMarcas paper
 

### 2005-05-31 paper/docs/background.tex
mostly removed the use of the word project, and 'I'

@@
 
-The ADT interface in this project is taken from \cite{Sedgewick02}. An \cc{Item}
+The ADT interface used here is taken from \cite{Sedgewick02}. An \cc{Item}
 is a record, and a function is defined to extract its key.  Comparisons are done
@@
 address may not directly relate to the index used to address the cache. This
-problem is not encountered in simulations in this project, as SimpleScalar
+problem is not encountered in our simulations, as SimpleScalar
 addresses its caches by virtual address.
@@
 elements of an array. This is called \n{spatial locality}. Many of the
-algorithms in this project are designed to take advantage of locality. Spatial
+algorithms presented here are designed to take advantage of locality. Spatial
 locality is exploited by ensuring that once a cache line is loaded, each key in
@@
 
-For this project, I split branches into two types. The first is a \n{flow
+In this report, we split branches into two types. The first is a \n{flow
 control branch}. Flow control refers to controlling the flow of execution of a
@@
 There are several kinds of branch predictors: \n{static}, \n{semi-static} and
-\n{dynamic}. This project is only interested in dynamic predictors, due to
-their high levels of accuracy. The other types are none-the-less discussed
+\n{dynamic}. This report only deals with dynamic predictors, which are of
+interest due to their high accuracy. The other types are none-the-less discussed
 below, as an introduction to dynamic predictors. A good discussion on predictor

### 2005-05-31 paper/docs/heap.tex
mostly removed the use of the word project, and 'I'

@@
 \section{Results}
-All simulations in this project are performed on 32 byte cache lines. As a
-result, 8-heaps are used for heapsort, since 8 4-byte \n{int}s fit the cache line.
+All simulations presented here were performed on 32 byte cache lines. As a
+result, 8-heaps were used for heapsort, since 8 4-byte \n{int}s fit the cache line.
 LaMarca used 64 bit ints, and while this doesn't affect results, it means that
 fewer keys fit into his cache lines. As a result, he used a 4-heap. To try and
-emulate his results, extra tests were done, using a 4-heap with a 32 byte
+emulate his results, extra tests were done using a 4-heap with a 32 byte
 cache line, and using a 4-heap using a 16 byte cache line, so that the entire

### 2005-05-31 paper/docs/merge.tex
have to rewrite a section of the merges code. starting now

@@
 occur at all. The solution is to apply a technique used in compilers called
-\n{tiling}. Using this technique, the number of keys which fit in the cache will
+\n{tiling}. Using this technique, an entire cache sized portion of the array will
 be fully sorted before moving on to the next set. When all the sets have been
@@
 continuing to level 2 cache as in the tiled mergesort. The expected improvement
-is not as significant as in the case of the level 2 cache, since is the penalty
+is not as significant as in the case of the level 2 cache, since the penalty
 for level 1 misses is not great. For this reason, this sort is called Double
@@
 unused key from each list. When there is a small number of lists, this isn't a
-problem, however, when the lists grow big, the overhead involved in this is
-substantial. It is necessary to check each of the lists, and each key searched
-would cause a cache miss, since each array segment is aligned, and map to
-the same cache block.
+problem, however, when the number of lists increases significanty, the overhead
+involved in this is substantial. It is necessary to check each of the lists, and
+each key searched would cause a cache miss, since each array segment is aligned,
+and map to the same cache block.
 
@@
 to the priority queue. There will still be conflict misses, but there will be no
-thrashing since key overwritten in the cache will not be needed again.
+thrashing since the keys overwritten in the cache will not be needed again.
 
@@
 this case only one key is out of place, and can be put in place with a single
-insertion sort run. This is linear at worst, and has excellent branch prediction
-properties, as shown earlier.
+insertion sort run. This is linear at worst, and will have only 1 branch
+misprediction.
 
@@
 eight keys, even though it had just added eight. In this manner, the size of the
-heap could be overrun. With random keys in the array, this is unlikely, and so
-this is not addressed.
+heap could be overrun. With random keys over a large range, as in our tests,
+this is unlikely, and so this is not addressed.
 
@@
 \subsection{Test Parameters}
-The level 1 cache is 8, so 2048 keys can be sorted in the level 1 cache.
-The number of keys to be sorted in the level 2 cache is 262144.  When arrays
-were pre-sorted, they were pre-sorted to a size four, or eight if the number of passes
-was going to be odd. These were reversed for the double multi-mergesort; since
-it does a k-way merge into the source array, the sorted cache sized segments
-need to be in the auxiliary array. An 8-heap was used for the multi-mergesorts.
+The level 1 cache is 8k, so 2048 keys can be sorted in the level 1 cache.
+262144 keys can be sorted in the level 2 cache. When arrays were pre-sorted,
+they were pre-sorted to a size four, or eight if the number of passes was going
+to be odd. These were reversed for the double multi-mergesort; since it does a
+k-way merge into the source array, the sorted cache sized segments need to be in
+the auxiliary array. An 8-heap was used for the multi-mergesorts.
 
@@
 
+%TODO look over lamarcas results
 The prediction results in Figure \ref{Branch Simulation results for mergesort}

### 2005-06-07 old/double_tiled_mergesort.c
removed old tarred gzips reran mergesort simulations removed .eps files from the repository made some fixes to the scripts deleted some directories which werent used, or which were old

@@ -1,1098 +1,0 @@
-#include <stdlib.h>
-#include <string.h>
-#include <stdio.h>
-#include <malloc.h>
-
-#include "cache_sorts.h"
-
-#define LIMIT OUT_OF_PLACE_LIMIT
-#define LIMIT_BITS OUT_OF_PLACE_LIMIT_BITS
-static void
-merge_flexible(unsigned int source[], int N, int starting_size, unsigned int target[]);
-static void
-merge(unsigned int source[], int N, int starting_size, unsigned int target[]);
-
-static void
-merge_reverse(unsigned int source[], int N, int starting_size, unsigned int target[]);
-static void
-merge_reverse_flexible(unsigned int source[], int N, int starting_size, unsigned int target[]);
-
-static void
-presort(unsigned int a[], int N);
-static void
-presort_flexible(unsigned int a[], int N);
-
-static int get_count(const int N)
-{
-
-	int lg2N = 32; /* im hard coding this and i dont care */
-	unsigned int msbN = 1 << (lg2N - 1);
-	while(msbN >= N)
-	{
-		lg2N--;
-		msbN = msbN >> 1;
-	}
-	return lg2N;
-}
-
-void
-double_tiled_mergesort(unsigned int a[], int N)
-{
-	/* track keeps an eye on where i started last */
-	/* outer track keeps and eye on which cache_sized/2 segemnt i started in last */
-	int i,j; /* indices for the first array */
-	
-	int level1_count;
-	int level2_count;
-	int extra_level1_count = 0;
-	int extra_level2;
-	int final_extra;
-
-
-	unsigned int* aux_data;
-	unsigned int* aux;
-
-	unsigned int* level2_start;
-	unsigned int* level2_aux_start;
-
-	unsigned int minusA;
-
-
-	/* get the address we need*/
-	minusA = ((1 << BLOCK_BITS) - get_index(a));
-	aux_data = memalign(ALIGNMENT, (N + 2*LIMIT) * sizeof(unsigned int));
-	aux = (unsigned int*)(((unsigned int)aux_data & (~BLOCK_AND_LINE_MASK)) | (minusA << LINE_BITS));
-	if (aux < aux_data) /* then the new index is less than the old one */
-	{
-		aux = (unsigned int*)((unsigned int)aux + (1 << (BLOCK_AND_LINE_BITS)));
-	}
-
-	if (N < 2048) /* fits in the level 1 cache */
-	{
-		presort_flexible(a, N);
-		merge_flexible(a, N, 4, aux);
[…]

### 2005-06-07 paper/fastsort_perfex/base_quicksort.c

@@ -1,199 +1,0 @@
-#include <stdlib.h>
-#include <stdio.h>
-
-#include "base_sorts.h"
-
-#define THRESHHOLD 10
-
-static int* stack = NULL;
-static int stack_index = 0;
-
-static void
-stackinit(unsigned int N)
-{
-
-	int lg2N = 32; /* im hard coding this and i dont care */
-	unsigned int msbN = 1 << (lg2N - 1);
-	while(msbN > N)
-	{
-		lg2N--;
-		msbN = msbN >> 1;
-	}
-
-/*	printf("\nAllocating a stack %d ints big (lg2N == %d)\n", (lg2N + 2) << 2, lg2N); */
-	
-	stack = malloc(((lg2N + 2) << 2) * sizeof(int));
-}
-
-inline static void
-push(int valueA, int valueB)
-{
-	stack[stack_index++] = valueA;
-	stack[stack_index++] = valueB;
-}
-
-inline static int 
-pop()
-{
-	stack_index--;
-	return stack[stack_index];
-}
-
-inline static void
-stackclear()
-{
-	if (stack != NULL) free(stack);
-	stack = NULL;
-}
-
-inline static int
-stackempty()
-{
-	return (stack_index==0);
-}
-
-static int
-partition(unsigned int a[], int l, int r)
-{
-	unsigned int v;
-	int i, j;
-
-	v = a[r];
-	i = l - 1;
-	j = r;
-
-	for(;;)
-	{
-		while (less(a[++i], v))
-		{
-			branch_taken(&global_predictor0);
-		}
-		branch_not_taken(&global_predictor0);
-
-		while (less(v, a[--j]))
-		{
-			branch_taken(&global_predictor1);
-		}
-		branch_not_taken(&global_predictor1);
-		
-		if (i >= j)
-			break;
-
-		exch(a[i], a[j]);
-	}
-
-	exch(a[i], a[r]);
-
-	return i;
-}
-
-static void
-insertion_sentinel(unsigned int a[], int N)
-{
-	int i;
-	unsigned min;
-
-	min = 0;
-	for(i = 1; i < N; i++)
-	{
-		if(less(a[i], a[min]))
-		{
-			min = i;
-		}
-	}
-	exch(a[0], a[min]);
-
-}
-
-static void
-insertion(unsigned int a[], int N)
-{
-	int i;
-	unsigned int v;
[…]

### 2005-06-07 paper/fastsort_perfex/cache_radixsort.c

@@ -1,66 +1,0 @@
-#include <stdlib.h>
-#include <string.h>
-
-#include "base_sorts.h"
-#include "insertion.c"
-
-#define THRESHHOLD 10
-
-#define BITSWORD 32
-#define BITSBYTE 8
-#define BYTESWORD 4
-#define R 256
-
-#define digit(A, B) (((A) >> (BITSWORD-((B)+1)*BITSBYTE)) & (R-1))
-
-void
-cache_radixsort(unsigned int a[], int N)
-{
-	unsigned int* aux = malloc(N * sizeof(unsigned int));
-	int i, j, w; /* R is the size of a digit, hence the number of bins */
-	int count0[R+1];
-	int count1[R+1];
-	int* temp;
-
-	int* count = count0;
-	int* next_count = count1;
-
-	unsigned int* s = a;
-	unsigned int* t = aux;
-
-	/* count the digits for the first time */
-	for(j = 0; j < R; j++) count[j] = 0;
-	for(i = 0; i < N; i++) count[digit(s[i], 3) + 1]++;
-	for(j = 1; j < R; j++) count[j] += count[j-1];
-
-	for(w = BYTESWORD - 1; w > 0; w--) /* this is from lsb. 3 == lsByte, 0 == msByte */
-	{
-		for(j = 0; j < R; j++) next_count[j] = 0; /* set the next counts to 0 */
-
-		for(i = 0; i <= N-1; i++) /* when we look up the array, we now have its position */
-		{
-			t[count[digit(s[i], w)]++] = s[i];
-			next_count[digit(s[i], w-1) + 1]++;
-		}
-	
-		for(j = 1; j < R; j++) next_count[j] += next_count[j-1]; /* make the next counts cumulative */
-
-		/* swap the counts */
-		temp = next_count;
-		next_count = count;
-		count = temp;;
-
-		/* swap the arrays */
-		temp = t;
-		t = s;
-		s = temp;
-	}
-
-	/* finish it off */
-	for(i = 0; i <= N-1; i++) /* when we look up the array, we now have its position */
-	{
-		t[count[digit(s[i], w)]++] = s[i];
-	}
-
-	free(aux);
-}

### 2005-06-07 paper/fastsort_perfex/double_multi_mergesort.c

@@ -1,990 +1,0 @@
-#include <stdlib.h>
-#include <string.h>
-#include <stdio.h>
-#include <malloc.h>
-#include <limits.h>
-
-#include "cache_sorts.h"
-
-#define LIMIT OUT_OF_PLACE_LIMIT
-#define LIMIT_BITS OUT_OF_PLACE_LIMIT_BITS
-
-#define ODD_COUNT 8
-#define EVEN_COUNT 4
-
-static void
-merge(unsigned int source[], int N, int starting_size, unsigned int target[]);
-
-static void
-merge_reverse(unsigned int source[], int N, int starting_size, unsigned int target[]);
-
-static void
-presort(unsigned int a[], int N);
-static void
-presort_flexible(unsigned int a[], int N);
-
-static int presort_count = 8;
-static int double_presort_count = 16;
-
-static void
-fix_up(unsigned int heap[], int child)
-{
-	int parent;
-
-	unsigned int v = heap[child];
-	while(less(v, heap[parent = ((child-1)/HEAP_SIZE)]))
-	{
-		heap[child] = heap[parent];
-		heap[parent] = v;
-		child = parent;
-	}
-}
-
-static void
-fix_down(unsigned int heap[], int parent, int N)
-{
-	int child;
-	int child2;
-	unsigned int v = heap[parent];
-
-	while(parent*HEAP_SIZE+1 < N) /* see if it has any children */
-	{
-		child = parent*HEAP_SIZE + 1; /* this gets the rightmost node */
-		child2 = child;
-
-		if (less(heap[child2 + 0], heap[child])) child = child2 + 0;
-		if (less(heap[child2 + 1], heap[child])) child = child2 + 1;
-		if (less(heap[child2 + 2], heap[child])) child = child2 + 2;
-		if (less(heap[child2 + 3], heap[child])) child = child2 + 3;
-
-#if (HEAP_SIZE == 8)
-		if (less(heap[child2 + 4], heap[child])) child = child2 + 4;
-		if (less(heap[child2 + 5], heap[child])) child = child2 + 5;
-		if (less(heap[child2 + 6], heap[child])) child = child2 + 6;
-		if (less(heap[child2 + 7], heap[child])) child = child2 + 7;
-
-#endif
-
-		/* when the larger child isnt large enough to be promoted, stop */
-		if (!less(heap[child], v)) break;
-
-		/* move down */
-		heap[parent] = heap[child];
-		parent = child;
-	}
[…]

### 2005-06-07 paper/fastsort_perfex/insertion.c

@@ -1,38 +1,0 @@
-#include "utils.h"
-	
-static void
-insertion_sentinel(unsigned int a[], int N)
-{
-	int i;
-	unsigned min;
-
-	min = 0;
-	for(i = 1; i < N; i++)
-	{
-		if(less(a[i], a[min]))
-		{
-			min = i;
-		}
-	}
-	exch(a[0], a[min]);
-
-}
-
-static void
-insertion(unsigned int a[], int N)
-{
-	int i;
-	unsigned int v;
-	for(i = 1; i < N; i++)
-	{
-		int j = i;
-		v = a[i];
-
-		while(less(v, a[j-1]))
-		{
-			a[j] = a[j-1];
-			j--;
-		}
-		a[j] = v;
-	}
-}

### 2005-06-07 paper/fastsort_perfex/multi_quicksort_seq.c

@@ -1,406 +1,0 @@
-#include <stdlib.h>
-#include <stdio.h>
-#include <string.h>
-#include <limits.h>
-
-#include "cache_sorts.h"
-
-#define THRESHHOLD 10
-
-static int* stack = NULL;
-static int stack_index = 0;
-
-/* C is the number of keys that fit in the cache */
-#define C IN_PLACE_LIMIT
-#define C_DIV_3 (C / 3)
-static void
-stackinit(unsigned int N)
-{
-
-	int lg2N = 32; /* im hard coding this and i dont care */
-	unsigned int msbN = 1 << (lg2N - 1);
-	while(msbN > N)
-	{
-		lg2N--;
-		msbN = msbN >> 1;
-	}
-
-	if (N > 2 * C) /* make space for all the pivots being pushed */
-	{
-		lg2N +=  N / C_DIV_3;
-	}
-
-/*	printf("\nAllocating a stack %d ints big (lg2N == %d)\n", (lg2N + 2) << 2, lg2N); */
-
-	/* +2 for initial values, *2 cuase we push 2 ints a time */
-	/* we only really need about 3 less, cause of the threshhold, but this doesnt hurt */
-	stack = malloc(((lg2N + 2) * 2) * sizeof(int));
-}
-
-inline static void
-push(int valueA, int valueB)
-{
-	stack[stack_index++] = valueA;
-	stack[stack_index++] = valueB;
-}
-
-inline static int 
-pop()
-{
-	stack_index--;
-	return stack[stack_index];
-}
-
-inline static void
-stackclear()
-{
-	if (stack != NULL) free(stack);
-	stack = NULL;
-}
-
-inline static int
-stackempty()
-{
-	return (stack_index==0);
-}
-
-
-
-static int
-partition(unsigned int a[], int l, int r)
-{
-	unsigned int v;
-	int i, j;
-
-	v = a[r];
-	i = l - 1;
-	j = r;
-
-	for(;;)
-	{
-		while (less(a[++i], v))
-			;
-		while (less(v, a[--j]))
-			;
-		
-
-		if (i >= j)
-			break;
-
-		exch(a[i], a[j]);
-	}
-
-	exch(a[i], a[r]);
-
-	return i;
-}
-
-/*static void
-print_stack()
-{
-	printf("\nThe stack is: ");
-	for(int i = 0; i < stack_index; i++)
-	{
-		printf(" %d, ", stack[i]);
-	}
-}
-*/
-
-/* this is included so that we can replace the < N with <= N
[…]

### 2005-06-07 paper/fastsort_perfex/on2_bubblesort2.c

@@ -1,21 +1,0 @@
-#include "on2_sorts.h"
-
-void
-on2_bubblesort2(unsigned int a[], int N)
-{
-	int i = 0, j, k = 0;
-
-	while(i < N-1)
-	{
-		k = N-1;
-		for (j = N-1; j > i; j--)
-		{
-			if (less(a[j],a[j-1]))
-			{
-				k = j;
-				exch(a[j-1],a[j]);
-			}
-		}
-		i = k;
-	}
-}

### 2005-06-07 paper/fastsort_perfex/predictor.c

@@ -1,104 +1,0 @@
-#include "predictor.h"
-
-#include <string.h>
-
-#if _USE_SOFTWARE_PREDICTOR
-	
-static char state_names[4][20] = 
-{
-	"NT saturated",
-	"NT",
-	"T",
-	"T saturated"
-};
-
-void
-init_predictor(struct predictor* pred)
-{
-	memset(pred, 0, sizeof(struct predictor));
-	/*pred->state = NOT_TAKEN_SATURATED; */
-	pred->state = TAKEN;
-}
-
-void 
-branch_taken(struct predictor* pred)
-{
-	int s = pred->state;
-	pred->branch_count[s]++;
-	pred->taken_count[s]++;
-
-	/* check what was predicted */
-	if (pred->state >= TAKEN) pred->correct_count[s]++; 
-	else pred->incorrect_count[s]++;
-
-	/* need to update the predictor? */
-	if (pred->state != TAKEN_SATURATED) pred->state++;
-		
-}
-
-void 
-branch_not_taken(struct predictor* pred)
-{
-	int s = pred->state;
-	pred->branch_count[s]++;
-	pred->not_taken_count[s]++;
-
-	/* check what was predicted */
-	if (pred->state <= NOT_TAKEN) pred->correct_count[s]++; 
-	else pred->incorrect_count[s]++;
-
-	/* need to update the predictor? */
-	if (pred->state != NOT_TAKEN_SATURATED) pred->state--;
-}
-
-void
-dump_predictor_info(FILE* file, struct predictor* pred, char* name)
-{
-	fprintf(file,	"%s:"
-					"states                     %13s %13s %13s %13s %13s\n"
-					"branch_count:              %13d %13d %13d %13d %13d\n"
-					"correct_count:             %13d %13d %13d %13d %13d\n"
-					"incorrect_count:           %13d %13d %13d %13d %13d\n"
-					"taken_count:               %13d %13d %13d %13d %13d\n"
-					"not_taken_count:           %13d %13d %13d %13d %13d\n"
-					"current state:             %13s\n\n",
-					name,
-					state_names[0], state_names[1],
-					state_names[2], state_names[3], "Total",
-					pred->branch_count[0], pred->branch_count[1],
-					pred->branch_count[2], pred->branch_count[3],
-					pred->branch_count[0] + pred->branch_count[1] +
-					pred->branch_count[2] + pred->branch_count[3],
-					pred->correct_count[0], pred->correct_count[1],
-					pred->correct_count[2], pred->correct_count[3],
[…]

### 2005-06-07 paper/fastsort_perfex/predictor.h

@@ -1,56 +1,0 @@
-#ifndef _PREDICTOR_H_
-#define _PREDICTOR_H_
-
-#if _USE_SOFTWARE_PREDICTOR
-#include <stdio.h>
-struct predictor
-{
-	int branch_count[4]; /* # of branches */
-
-	int correct_count[4]; /* # of correct predictions */
-	int incorrect_count[4]; /* # of incorrect predictions */
-
-	int taken_count[4]; /* # of branches rtaken */
-	int not_taken_count[4]; /* # of branches not taken */
-
-	int state; /* the current state of the predictor */
-	/* 0 = saturated not-taken */
-	/* 1 = unsaturated not-taken */
-	/* 2 = unstaturated taken */
-	/* 3 = saturated taken */
-
-};
-
-#define NOT_TAKEN_SATURATED 0
-#define NOT_TAKEN 1
-#define TAKEN 2
-#define TAKEN_SATURATED 3
-
-void init_predictor(struct predictor* pred);
-
-void branch_taken(struct predictor* pred);
-void branch_not_taken(struct predictor* pred);
-
-void dump_predictor_info(FILE* file, struct predictor* pred, char* name);
-
-
-extern struct predictor global_predictor0;
-extern struct predictor global_predictor1;
-extern struct predictor global_predictor2;
-extern struct predictor global_predictor3;
-extern struct predictor global_predictor4;
-extern struct predictor global_predictor5;
-extern struct predictor global_predictor6;
-extern struct predictor global_predictor7;
-extern struct predictor global_predictor8;
-extern struct predictor global_predictor9;
-#else
-
-#define init_predictor(A) do{}while(0)
-#define branch_taken(A)
-#define branch_not_taken(A)
-#define dump_predictor_info(A, B, C) do{}while(0)
-
-#endif
-
-#endif

### 2005-06-07 paper/docs/merge.tex
removed old tarred gzips reran mergesort simulations removed .eps files from the repository made some fixes to the scripts deleted some directories which werent used, or which were old
renamed the logs directory, updated merge a tad

@@
 
-\plot{merge}{}{}{Results shown are for a Direct Mapped cache.}{Results shown are
-for a Direct Mapped cache.}{}{}
+\plot{merge}{}{}{Results shown are for a Direct Mapped cache.}{Except where
+marked, results shown are for a Direct Mapped cache.}{}{}
 
@@
 cache blocks, without the need for padding.
+
+%TODO
+TODO what about level 1 cache misses if we realigned by 1024?

### 2005-06-09 tech_report/docs/code/simple_code.tex
inputs are now the single tiled mergesort ones moved paper to tech_report to allow new paper directory to actually be for the paper

% Remember to use the lgrind style

\Head{}
\File{simple}{2004}{4}{28}{3:23}{9563}
\L{\LB{\C{}\#!/usr/bin/perl_-w\CE{}}}
\L{\LB{\V{use}_\V{Getopt}::\V{Std};}}
\L{\LB{\N{\$command}\V{\_line}_=_\N{\$0};}}
\L{\LB{\K{foreach}_(@\V{ARGV})_\{_\N{\$command}\V{\_line}_.=_\S{}\3_\$\_\3\SE{};_\}}}
\L{\LB{}}
\L{\LB{\C{}\#_-------------------------------------------------------\CE{}}}
\L{\LB{\C{}\#_}\Tab{40}{Do_the_options\CE{}}}
\L{\LB{\C{}\#_-------------------------------------------------------\CE{}}}
\L{\LB{}}
\L{\LB{\V{my}_\%\V{options}=();}}
\L{\LB{\V{getopts}(\S{}\3c:L:ihlm:M:s\3\SE{},\2\%\V{options});}}
\L{\LB{}}
\L{\LB{\K{if}_(\K{defined}(\N{\$options}\{\S{}\3h\3\SE{}\})_\|\,\|_(\V{scalar}(@\V{ARGV})_==_\N{0}))}}
\L{\LB{\{}}
\L{\LB{}\Tab{8}{\K{die}_(\S{}\3simple_[options]_sort\2noptions_are:\2n\2t}}
\L{\LB{}\Tab{16}{\-c_comment_\-_comment_with_which_to_name_log\2n\2t}}
\L{\LB{}\Tab{16}{\-l_list_commands_simple_will_execute\2n\2t}}
\L{\LB{}\Tab{16}{\-s_do_a_short_run\2n\2t}}
\L{\LB{}\Tab{16}{\-h_this_help_message\2n\2t}}
\L{\LB{}\Tab{16}{\-L_length_\-_cache_line_length\2n\2t}}
\L{\LB{}\Tab{16}{\-i_just_count_instructions\2n\2t}}
\L{\LB{}\Tab{16}{\-m_\-_specify_minimum_sort_size\2n\2t}}
\L{\LB{}\Tab{16}{\-M_\-_specify_maximum_sort_size\2n\3\SE{});}}
\L{\LB{\}}}
\L{\LB{}}
\L{\LB{\C{}\#_i_means_just_get_an_instruction_count,_nothing_else\CE{}}}
\L{\LB{\N{\$just}\V{\_instruction}_=_\N{0};}}
\L{\LB{\K{if}_(\K{defined}(\N{\$options}\{\S{}\3i\3\SE{}\}))_\{_\N{\$just}\V{\_instruction}_=_\N{1};_\}_}}
\L{\LB{}}
\L{\LB{\C{}\#redfine_the_max_and_minimum_number_of_items\CE{}}}
\L{\LB{\N{\$min}\V{\_num\_items}_=_\N{4096};}}
\L{\LB{\N{\$max}\V{\_num\_items}_=_\N{4194304};}}
\L{\LB{}}
\L{\LB{\K{if}_(\K{defined}(\N{\$options}\{\S{}\3m\3\SE{}\}))_\{_\N{\$min}\V{\_num\_items}_=_\N{\$options}\{\S{}\3m\3\SE{}\};_\}}}
\L{\LB{\K{if}_(\K{defined}(\N{\$options}\{\S{}\3M\3\SE{}\}))_\{_\N{\$max}\V{\_num\_items}_=_\N{\$options}\{\S{}\3M\3\SE{}\};_\}}}
\L{\LB{}}
\L{\LB{\N{\$$\ell$ine}\V{\_length}_=_\N{32};\C{}_\#_the_length_of_a_cache_line\CE{}}}
\L{\LB{\K{if}_(\K{defined}(\N{\$options}\{\S{}\3L\3\SE{}\}))_\{_\N{\$$\ell$ine}\V{\_length}_=_\N{\$options}\{\S{}\3L\3\SE{}\};_\}}}
\L{\LB{}}
\L{\LB{\C{}\#_check_if_a_short_run_is_desired\CE{}}}
\L{\LB{\N{\$short}\V{\_run}_=_\N{0};}}
\L{\LB{\K{if}_(\K{defined}(\N{\$options}\{\S{}\3s\3\SE{}\}))_\{_\N{\$short}\V{\_run}_=_\N{1};_\}}}
\L{\LB{}}
\L{\LB{\C{}\#_check_if_just_the_list_of_commands_is_wanted\CE{}}}
\L{\LB{\N{\$just}\V{\_list}_=_\N{0};}}
\L{\LB{\K{if}_(\K{defined}(\N{\$options}\{\S{}\3l\3\SE{}\}))_\{_\N{\$just}\V{\_list}_=_\N{1};_\}}}
\L{\LB{}}
\L{\LB{\C{}\#_set_the_comment\CE{}}}
\L{\LB{\N{\$comment}_=_\K{localtime}();}}
\L{\LB{\K{if}_(\K{defined}(\N{\$options}\{\S{}\3c\3\SE{}\}))_\{_\N{\$comment}_=_\N{\$options}\{\S{}\3c\3\SE{}\};_\}}}
\L{\LB{\N{\$comment}_=\~{}_\K{s}/[\2\K{s}:]/\V{\_}/\V{g};}}
\L{\LB{}}
\L{\LB{\C{}\#_----------------------------------------------------------\CE{}}}
\L{\LB{\C{}\#_}\Tab{56}{set_the_constants\CE{}}}
\L{\LB{\C{}\#_----------------------------------------------------------\CE{}}}
\L{\LB{}}
\L{\LB{\C{}\#_the_simulator_details\CE{}}}
\L{\LB{@\V{simulators}_=_(_\S{}\3sim\-cache\3\SE{},_\S{}\3sim\-bpred\3\SE{}_)_;}}
\L{\LB{}}
\L{\LB{\C{}\#_put_in_the_keywords\CE{}}}
\L{\LB{\N{\$keywords}\{\S{}\3sim\-cache\3\SE{}\}_=_}}
\L{\LB{}\Tab{8}{[}}
\L{\LB{}\Tab{16}{\S{}\3sim\_num\_insn\3\SE{},_\S{}\3sim\_num\_refs\3\SE{},_\S{}\3dl1.accesses\3\SE{},_\S{}\3dl1.hits\3\SE{},}}
\L{\LB{}\Tab{16}{\S{}\3dl1.misses\3\SE{},_\S{}\3dl1.replacements\3\SE{},_\S{}\3dl1.writebacks\3\SE{},_\S{}\3dl1.miss\_rate\3\SE{},}}
\L{\LB{}\Tab{16}{\S{}\3dl1.repl\_rate\3\SE{},_\S{}\3dl2.accesses\3\SE{},_\S{}\3dl2.hits\3\SE{},_\S{}\3dl2.misses\3\SE{},}}
\L{\LB{}\Tab{16}{\S{}\3dl2.replacements\3\SE{},_\S{}\3dl2.writebacks\3\SE{},_\S{}\3dl2.miss\_rate\3\SE{},_\S{}\3dl2.repl\_rate\3\SE{},}}
\L{\LB{}\Tab{16}{\S{}\3mem.page\_mem\3\SE{}}}
\L{\LB{}\Tab{8}{];}}
\L{\LB{\N{\$keywords}\{\S{}\3sim\-bpred\3\SE{}\}_=_}}
\L{\LB{}\Tab{8}{[}}
\L{\LB{}\Tab{16}{\S{}\3sim\_num\_branches\3\SE{},_\S{}\3sim\_IPB\3\SE{},_\S{}\3bpred\_bimod.lookups\3\SE{},}}
\L{\LB{}\Tab{16}{\S{}\3bpred\_bimod.updates\3\SE{},_\S{}\3bpred\_bimod.addr\_hits\3\SE{},_\S{}\3bpred\_bimod.dir\_hits\3\SE{},}}
\L{\LB{}\Tab{16}{\S{}\3bpred\_bimod.misses\3\SE{},_\S{}\3bpred\_bimod.bpred\_addr\_rate\3\SE{},}}
\L{\LB{}\Tab{16}{\S{}\3bpred\_bimod.bpred\_dir\_rate\3\SE{},_\S{}\3bpred\_2lev.lookups\3\SE{},}}
\L{\LB{}\Tab{16}{\S{}\3bpred\_2lev.updates\3\SE{},_\S{}\3bpred\_2lev.addr\_hits\3\SE{},_\S{}\3bpred\_2lev.dir\_hits\3\SE{},}}
\L{\LB{}\Tab{16}{\S{}\3bpred\_2lev.misses\3\SE{},_\S{}\3bpred\_2lev.bpred\_addr\_rate\3\SE{},}}
\L{\LB{}\Tab{16}{\S{}\3bpred\_2lev.bpred\_dir\_rate\3\SE{}}}
\L{\LB{}\Tab{8}{];}}
\L{\LB{}}
\L{\LB{\C{}\#\<name\>:\<nsets\>:\<bsize\>:\<assoc\>:\<repl\>\CE{}}}
\L{\LB{\C{}\#_cache_sizes_are_2_and_4_MB,_with_a_line_size_of_32_Bytes.\CE{}}}
\L{\LB{\C{}\#_This_mean_either_the_associativity_or_\#sets_is_65536/131072_and_l1_is_256.\CE{}}}
\L{\LB{\C{}\#_we_make_l1_FA\CE{}}}
\L{\LB{}}
\L{\LB{\N{\$size}\V{\_4\_meg}_=_\N{4}_*_\N{1024}_*_\N{1024}_/_\N{\$$\ell$ine}\V{\_length};}}
\L{\LB{\N{\$size}\V{\_2\_meg}_=_\N{2}_*_\N{1024}_*_\N{1024}_/_\N{\$$\ell$ine}\V{\_length};}}
\L{\LB{\N{\$size}\V{\_8\_k}_=_\N{8}_*_\N{1024}_/_\N{\$$\ell$ine}\V{\_length};}}
\L{\LB{}}
\L{\LB{\N{\$params}\{\S{}\3sim\-cache\3\SE{}\}[\N{0}]_=_\S{}\3\-cache:dl1_dl1:\$size\_8\_k:\$line\_length:1:l}}
\L{\LB{}\Tab{16}{\-cache:dl2_dl2:\$size\_4\_meg:\$line\_length:1:l}}
\L{\LB{}\Tab{16}{\-cache:il1_il1:\$size\_8\_k:\$line\_length:1:l_\-cache:il2_dl2\3\SE{};_}}
\L{\LB{}\Tab{8}{}}
\L{\LB{\N{\$params}\{\S{}\3sim\-cache\3\SE{}\}[\N{1}]_=_\S{}\3\-cache:dl1_dl1:\$size\_8\_k:\$line\_length:1:l}}
\L{\LB{}\Tab{16}{\-cache:dl2_dl2:\$size\_2\_meg:\$line\_length:1:l}}
\L{\LB{}\Tab{16}{\-cache:il1_il1:\$size\_8\_k:\$line\_length:1:l_\-cache:il2_dl2\3\SE{};}}
\L{\LB{\N{\$params}\{\S{}\3sim\-cache\3\SE{}\}[\N{2}]_=_\S{}\3\-cache:dl1_dl1:1:\$line\_length:\$size\_8\_k:l}}
\L{\LB{}\Tab{16}{\-cache:dl2_dl2:1:\$line\_length:\$size\_4\_meg:l_}}
\L{\LB{}\Tab{16}{\-cache:il1_il1:1:\$line\_length:\$size\_8\_k:l_\-cache:il2_dl2\3\SE{};}}
\L{\LB{\N{\$params}\{\S{}\3sim\-cache\3\SE{}\}[\N{3}]_=_\S{}\3\-cache:dl1_dl1:1:\$line\_length:\$size\_8\_k:l}}
\L{\LB{}\Tab{16}{\-cache:dl2_dl2:1:\$line\_length:\$size\_2\_meg:l}}
\L{\LB{}\Tab{16}{\-cache:il1_il1:1:\$line\_length:\$size\_8\_k:l_\-cache:il2_dl2\3\SE{};}}
\L{\LB{}}
\L{\LB{\N{\$params}\{\S{}\3sim\-bpred\3\SE{}\}[\N{0}]_=_\S{}\3\-bpred_bimod_\-bpred:bimod_512\3\SE{};}}
\L{\LB{\N{\$params}\{\S{}\3sim\-bpred\3\SE{}\}[\N{1}]_=_\S{}\3\-bpred_bimod_\-bpred:bimod_1024\3\SE{};}}
\L{\LB{\N{\$params}\{\S{}\3sim\-bpred\3\SE{}\}[\N{2}]_=_\S{}\3\-bpred_bimod_\-bpred:bimod_2048\3\SE{};}}
\L{\LB{}}
\L{\LB{\N{\$params}\{\S{}\3sim\-bpred\3\SE{}\}[\N{3}]_=_\S{}\3\-bpred_2lev_\-bpred:2lev_1_512_10_1\3\SE{};}}
\L{\LB{\N{\$params}\{\S{}\3sim\-bpred\3\SE{}\}[\N{4}]_=_\S{}\3\-bpred_2lev_\-bpred:2lev_1_1024_10_1\3\SE{};}}
\L{\LB{\N{\$params}\{\S{}\3sim\-bpred\3\SE{}\}[\N{5}]_=_\S{}\3\-bpred_2lev_\-bpred:2lev_1_2048_10_1\3\SE{};}}
\L{\LB{\N{\$params}\{\S{}\3sim\-bpred\3\SE{}\}[\N{6}]_=_\S{}\3\-bpred_2lev_\-bpred:2lev_1_4096_10_1\3\SE{};}}
\L{\LB{}}
\L{\LB{\K{if}_(\N{\$short}\V{\_run})}}
\L{\LB{\{}}
\L{\LB{\C{}}\Tab{8}{\#_reduce_the_number_of_items\CE{}}}
\L{\LB{}\Tab{8}{\N{\$max}\V{\_num\_items}_=_\N{32768};\C{}_\#_this_is_caled_down_from_(4096_*_1024)_ie_by_128\CE{}}}
\L{\LB{}}
\L{\LB{\C{}}\Tab{8}{\#_now_simulate_a_smaller_cache\CE{}}}
\L{\LB{}}
\L{\LB{}\Tab{8}{\N{\$params}\{\S{}\3sim\-cache\3\SE{}\}[\N{0}]_=_\S{}\3\-cache:dl1_dl1:16:32:1:l}}
\L{\LB{}\Tab{24}{\-cache:dl2_dl2:256:32:1:l_\-cache:il1_il1:16:32:1:l}}
\L{\LB{}\Tab{24}{\-cache:il2_il2:256:32:1:l\3\SE{};_}}
\L{\LB{}\Tab{8}{\N{\$params}\{\S{}\3sim\-cache\3\SE{}\}[\N{1}]_=_\S{}\3\-cache:dl1_dl1:16:32:1:l_}}
\L{\LB{}\Tab{24}{\-cache:dl2_dl2:512:32:1:l_\-cache:il1_il1:16:32:1:l}}
\L{\LB{}\Tab{24}{\-cache:il2_il2:512:32:1:l\3\SE{};}}
\L{\LB{}\Tab{8}{\N{\$params}\{\S{}\3sim\-cache\3\SE{}\}[\N{2}]_=_\S{}\3\-cache:dl1_dl1:1:32:16:l}}
\L{\LB{}\Tab{24}{\-cache:dl2_dl2:1:32:256:l_\-cache:il1_il1:16:32:1:l}}
\L{\LB{}\Tab{24}{\-cache:il2_il2:1:32:256:l\3\SE{};}}
\L{\LB{}\Tab{8}{\N{\$params}\{\S{}\3sim\-cache\3\SE{}\}[\N{3}]_=_\S{}\3\-cache:dl1_dl1:1:32:16:l_}}
\L{\LB{}\Tab{24}{\-cache:dl2_dl2:1:32:512:l_\-cache:il1_il1:16:32:1:l}}
\L{\LB{}\Tab{24}{\-cache:il2_il2:1:32:512:l\3\SE{};}}
\L{\LB{}\Tab{8}{}}
\L{\LB{\}}}
\L{\LB{}}
\L{\LB{\N{\$number}\V{\_of\_items}_=_\N{\$min}\V{\_num\_items};}}
\L{\LB{}}
\L{\LB{\K{while}(\N{\$number}\V{\_of\_items}_\<_\N{\$max}\V{\_num\_items})}}
\L{\LB{\{}}
\L{\LB{}\Tab{8}{\K{push}(@\V{ns},_\N{\$number}\V{\_of\_items});}}
\L{\LB{}\Tab{8}{\N{\$number}\V{\_of\_items}_*=_\N{2};}}
\L{\LB{\}}}
\L{\LB{\K{push}(@\V{ns},_\N{\$max}\V{\_num\_items});}}
\L{\LB{}}
\L{\LB{\N{\$$\ell$og}\V{\_directory}_=_\S{}\3./logs/\3\SE{};}}
\L{\LB{}}
\L{\LB{\C{}\#_compiler_settings\CE{}}}
\L{\LB{\N{\$CC}_=_\S{}\3sslittle\-na\-sstrix\-gcc\3\SE{};}}
\L{\LB{\N{\$CFLAGS}_=_\S{}\3\-Wall_\-O3_\-funroll\-loops_\-finline\-functions\3\SE{};}}
\L{\LB{\N{\$SIMPLEFLAGS}_=_\S{}\3\3\SE{};\C{}_\#-lm\CE{}}}
\L{\LB{}}
\L{\LB{\K{select}_\V{STDOUT};}}
\L{\LB{}}
\L{\LB{\N{\$dashes}_=_\S{}\3\-\-\-\-\-\-\-\-\-\-\-\-\-\-\-\-\-\-\-\-\-\-\-\-\-\-\-\-\-\-\-\-\-\-\-\-\-\-\-\-\-\-\-\-\-\-\-\-\-\-\-\-\-\-\-\2n\3\SE{};}}
\L{\LB{\C{}\#_each_sort,_do_a_cache_for_each_size,_then_a_bpred_for_each_size\CE{}}}
\L{\LB{}}
\L{\LB{}}
\L{\LB{\K{if}_(\N{\$just}\V{\_instruction})}}
\L{\LB{\{}}
\L{\LB{}\Tab{8}{\N{\$sortname}_=_\N{\$ARGV}[\N{0}];}}
\L{\LB{}\Tab{8}{@\V{seeds}_=_(_\S{}\30\3\SE{},_\S{}\320\3\SE{},_\S{}\31234\3\SE{},_\S{}\363345\3\SE{},_\S{}\323\3\SE{}_);}}
\L{\LB{\C{}}\Tab{8}{\#_we_max_out_at_32768_unless_you_specify_more_with_the_-m.\CE{}}}
\L{\LB{\C{}}\Tab{8}{\#_you_can_specify_less_with_the_-M\CE{}}}
\L{\LB{}\Tab{8}{\K{if}_(\N{\$max}\V{\_num\_items}_\>=_\N{32768})}}
\L{\LB{}\Tab{8}{\{}}
\L{\LB{}\Tab{16}{\N{\$max}\V{\_num\_items}_=_\N{32768};}}
\L{\LB{}\Tab{8}{\}}}
\L{\LB{}\Tab{8}{\K{if}_(\N{\$min}\V{\_num\_items}_\>_\N{\$max}\V{\_num\_items})}}
\L{\LB{}\Tab{8}{\{}}
\L{\LB{}\Tab{16}{\N{\$max}\V{\_num\_items}_=_\N{\$min}\V{\_num\_items};}}
\L{\LB{}\Tab{8}{\}}}
\L{\LB{}}
\L{\LB{}\Tab{8}{\K{foreach}_\N{\$seed}_(@\V{seeds})}}
\L{\LB{}\Tab{8}{\{}}
\L{\LB{}\Tab{16}{@\V{instruction\_count}_=_();}}
\L{\LB{}\Tab{16}{\N{\$comp}\V{\_command}_=_\S{}\3\$CC_\$CFLAGS_\$SIMPLEFLAGS_\-funroll\-loops}}
\L{\LB{}\Tab{16}{\-finline\-functions_\-DRANDOM\_SIZE=\$max\_num\_items_\-DSEED=\$seed}}
\L{\LB{}\Tab{16}{\-DSORT=\$sortname_\$sortname.c_simpleMain.c_\-o_special\_file\3\SE{};}}
\L{\LB{\C{}}\Tab{16}{\#_run_the_command\CE{}}}
\L{\LB{\C{}}\Tab{16}{\#_capture_the_compiler_errors_(STDERR)\CE{}}}
\L{\LB{}\Tab{16}{\N{\$comp}\V{\_output}_=_{`}\N{\$comp}\V{\_command}_\N{2}\>\&\N{1}_\N{1}\>/\V{dev}/\V{null}{`};_}}
\L{\LB{}}
\L{\LB{\C{}}\Tab{16}{\#_if_it_doesnt_compile_stop_and_print_out_errors\CE{}}}
\L{\LB{}\Tab{16}{\K{if}_(\N{\$comp}\V{\_output}_=\~{}_/./)}}
\L{\LB{}\Tab{16}{\{}}
\L{\LB{}\Tab{24}{\K{die}(\S{}\3This_command_returned_errors:}}
\L{\LB{}\Tab{32}{\2\3\$comp\_command\2\3\2n\$comp\_output\3\SE{});}}
\L{\LB{}\Tab{16}{\}}}
\L{\LB{}\Tab{16}{\N{\$sim}\V{\_command}_=_\S{}\3sim\-fast_special\_file\3\SE{};}}
\L{\LB{}\Tab{16}{@\V{sim\_output}_=_{`}\N{\$sim}\V{\_command}_\N{2}\>\&\N{1}_\N{1}\>/\V{dev}/\V{null}{`};}}
\L{\LB{}}
\L{\LB{}\Tab{16}{@\V{instruction\_count}_=_\K{grep}(/\V{sim\_num\_insn}/,_@\V{sim\_output});}}
\L{\LB{}\Tab{16}{\K{print}_\N{\$instruction}\V{\_count}[\N{0}];}}
\L{\LB{}\Tab{8}{\}}}
\L{\LB{}}
\L{\LB{}\Tab{8}{\K{system}(\S{}\3rm_special\_file\3\SE{});}}
\L{\LB{}\Tab{8}{\K{exit};}}
\L{\LB{\}}}
\L{\LB{}}
\L{\LB{}}
\L{\LB{\V{SORT}:_\K{foreach}_\N{\$sortname}_(@\V{ARGV})}}
\L{\LB{\{}}
\L{\LB{}\Tab{8}{\N{\$sortfi$\ell$e}_=_\N{\$sortname.int}(\K{rand}(\N{2147483648})).\S{}\3.c\3\SE{};}}
\L{\LB{}\Tab{8}{\N{\$object}\V{\_file}_=_\N{\$sortname.int}(\K{rand}(\N{2147483648}));}}
\L{\LB{}\Tab{8}{\K{system}(\S{}\3cp_\$sortname.c_\$sortfile\3\SE{});}}
\L{\LB{}}
\L{\LB{}}
\L{\LB{\C{}}\Tab{8}{\#_make_a_temp_file\CE{}}}
\L{\LB{}\Tab{8}{\N{\$n}\V{\_count}_=_\N{0};}}
\L{\LB{}\Tab{8}{\K{foreach}_\N{\$n}_(@\V{ns})\C{}_\#_ther_should_be_num\_sim\_runs_*_count(\$n)_columns.\CE{}}}
\L{\LB{}\Tab{8}{\{}}
\L{\LB{}\Tab{16}{\N{\$comp}\V{\_command}_=_\S{}\3\$CC_\$CFLAGS_\$SIMPLEFLAGS_\-DRANDOM\_SIZE=\$n_\-DSEED=0}}
\L{\LB{}\Tab{24}{\-DSORT=\$sortname_\$sortfile_simpleMain.c_\-o_\$object\_file\3\SE{};}}
\L{\LB{}}
\L{\LB{}\Tab{16}{\K{print}_\N{\$comp}\V{\_command}.\S{}\3\2n\3\SE{};}}
\L{\LB{}\Tab{16}{\K{push}_(@\V{command\_list},_\N{\$comp}\V{\_command});}}
\L{\LB{}}
\L{\LB{\C{}}\Tab{16}{\#_run_the_command\CE{}}}
\L{\LB{\C{}}\Tab{16}{\#_capture_the_compiler_errors_(STDERR)\CE{}}}
\L{\LB{}\Tab{16}{\N{\$comp}\V{\_output}_=_{`}\N{\$comp}\V{\_command}_\N{2}\>\&\N{1}_\N{1}\>/\V{dev}/\V{null}{`};_}}
\L{\LB{}}
\L{\LB{\C{}}\Tab{16}{\#_if_it_doesnt_compile_stop_and_print_out_errors\CE{}}}
\L{\LB{}\Tab{16}{\K{if}_(\N{\$comp}\V{\_output}_=\~{}_/./)}}
\L{\LB{}\Tab{16}{\{}}
\L{\LB{}\Tab{24}{\K{print}_\N{\$comp}\V{\_output};}}
\L{\LB{}\Tab{24}{\K{next}_\V{SORT};}}
\L{\LB{}\Tab{16}{\}}}
\L{\LB{}}
\L{\LB{}\Tab{16}{\K{foreach}_\N{\$simu$\ell$ator}_(@\V{simulators})\C{}_\#_run_each_of_the_simulators\CE{}}}
\L{\LB{}\Tab{16}{\{}}
\L{\LB{}\Tab{24}{\N{\$order}\V{\_count}_=_\N{0};}}
\L{\LB{}\Tab{24}{\N{\$run}\V{\_count}_=_\N{0};}}
\L{\LB{}\Tab{24}{\K{foreach}_\N{\$param}_(@\{\N{\$params}\{\N{\$simu$\ell$ator}\}\})}}
\L{\LB{}\Tab{24}{\{}}
\L{\LB{}\Tab{32}{\N{\$sim}\V{\_command}_=_\S{}\3\$simulator_\$param_\$object\_file\3\SE{};}}
\L{\LB{}\Tab{32}{\K{push}_(@\V{command\_list},_\N{\$sim}\V{\_command});}}
\L{\LB{}\Tab{32}{\K{print}_\N{\$sim}\V{\_command}.\S{}\3\2n\3\SE{};}}
\L{\LB{}\Tab{32}{\K{if}_(\N{\$just}\V{\_list})_\{_\K{next};_\}_}}
\L{\LB{}}
\L{\LB{}\Tab{32}{@\V{sim\_output}_=_{`}\V{nice}_\-\N{19}_\N{\$sim}\V{\_command}_\N{2}\>\&\N{1}_\N{1}\>/\V{dev}/\V{null}{`};}}
\L{\LB{}}
\L{\LB{\C{}}\Tab{32}{\#_ignore_the_headers_we_dont_want\CE{}}}
\L{\LB{}\Tab{32}{\K{while}(\N{\$$\ell$ine}_=_\K{shift}(@\V{sim\_output}))}}
\L{\LB{}\Tab{32}{\{}}
\L{\LB{}\Tab{40}{\K{if}_(\N{\$$\ell$ine}_=\~{}_/\V{sim}:_\2*\2*_\V{simulation}_\V{statistics}_\2*\2*/)}}
\L{\LB{}\Tab{48}{\{_\K{last};_\}}}
\L{\LB{}\Tab{40}{\K{if}_(\N{\$$\ell$ine}_=\~{}_/\V{fatal}/)}}
\L{\LB{}\Tab{48}{\{_\K{die}(\S{}\3Fatal_error_with_}}
\L{\LB{}\Tab{56}{command_\$sim\_command\3\SE{});_\}}}
\L{\LB{}\Tab{32}{\}}}
\L{\LB{}}
\L{\LB{\C{}}\Tab{32}{\#_fill_the_array_with_data_in_the_correct_order\CE{}}}
\L{\LB{}\Tab{32}{\K{while}(\K{shift}(@\V{sim\_output})_=\~{}_/(\2\V{S}+)\2\K{s}+(\2\V{S}+).*/)_}}
\L{\LB{}\Tab{32}{\{}}
\L{\LB{\C{}}\Tab{40}{\#_once_for_each_simulator\CE{}}}
\L{\LB{}\Tab{40}{\K{if}_(!(\N{\$resu$\ell$ts}\{\N{\$simu$\ell$ator}\}\{\N{\$1}\}))_}}
\L{\LB{}\Tab{40}{\{}}
\L{\LB{}\Tab{48}{\N{\$order}\{\N{\$simu$\ell$ator}\}[\N{\$order}\V{\_count}++]_=_\N{\$1};_}}
\L{\LB{}\Tab{40}{\}}}
\L{\LB{}\Tab{40}{@\{_\N{\$resu$\ell$ts}\{\N{\$simu$\ell$ator}\}\{\N{\$1}\}_\}[\N{\$n}\V{\_count}_*}}
\L{\LB{}\Tab{48}{\V{scalar}(@\{\N{\$params}\{\N{\$simu$\ell$ator}\}\})_+_\N{\$run}\V{\_count}]}}
\L{\LB{}\Tab{48}{=}\Tab{51}{\N{\$2};}}
\L{\LB{}\Tab{32}{\}}}
\L{\LB{}\Tab{32}{\N{\$run}\V{\_count}++;}}
\L{\LB{}\Tab{24}{\}}}
\L{\LB{}\Tab{16}{\}}}
\L{\LB{}\Tab{16}{\N{\$n}\V{\_count}++;}}
\L{\LB{}\Tab{16}{\K{system}(\S{}\3rm_\-f_\$object\_file\3\SE{});}}
\L{\LB{}\Tab{8}{\}}}
\L{\LB{}}
\L{\LB{}\Tab{8}{\K{if}_(\N{\$just}\V{\_list})_}}
\L{\LB{}\Tab{8}{\{}}
\L{\LB{}\Tab{16}{\K{system}(\S{}\3rm_\$sortfile\3\SE{});}}
\L{\LB{}\Tab{16}{\K{exit};}}
\L{\LB{}\Tab{8}{\}}}
\L{\LB{}}
\L{\LB{}\Tab{8}{\K{push}_(@\V{my\_output},_\S{}\3\2n\$dashes_Results_for_\$sortname\2n\$dashes\3\SE{});}}
\L{\LB{}\Tab{8}{\K{foreach}_\N{\$simu$\ell$ator}_(@\V{simulators})}}
\L{\LB{}\Tab{8}{\{}}
\L{\LB{}\Tab{16}{@\V{output}_=_();}}
\L{\LB{}\Tab{16}{\K{for}_(\N{\$i}_=_\N{0};_\N{\$i}_\<_\V{scalar}(@\{\N{\$order}\{\N{\$simu$\ell$ator}\}\});_\N{\$i}++)}}
\L{\LB{}\Tab{16}{\{}}
\L{\LB{}\Tab{24}{\N{\$keyword}_=_\N{\$order}\{\N{\$simu$\ell$ator}\}[\N{\$i}];}}
\L{\LB{}\Tab{24}{\N{\$output}\V{\_string}_=_\K{sprintf}(\S{}\3\%\-36s_\3\SE{},_\N{\$keyword});}}
\L{\LB{}\Tab{24}{@\V{result\_array}_=_@\{_\N{\$resu$\ell$ts}\{\N{\$simu$\ell$ator}\}\{\N{\$keyword}\}_\};}}
\L{\LB{}\Tab{24}{\K{foreach}_\N{\$resu$\ell$t}_(@\V{result\_array})}}
\L{\LB{}\Tab{24}{\{}}
\L{\LB{}\Tab{32}{\K{if}_(\K{defined}(\N{\$resu$\ell$t}))_}}
\L{\LB{}\Tab{40}{\{_\N{\$output}\V{\_string}_.=_\K{sprintf}(\S{}\3\%13s_\|\3\SE{},_\N{\$resu$\ell$t});\}}}
\L{\LB{}\Tab{32}{\K{else}_\{_\N{\$output}\V{\_string}_.=_\S{}\3xxxxxxxxxxxxx_\|\3\SE{};_\}}}
\L{\LB{}\Tab{24}{\}}}
\L{\LB{}\Tab{24}{\N{\$output}\V{\_string}_.=_\S{}\3\2n\3\SE{};}}
\L{\LB{}\Tab{24}{\K{push}_(@\V{output},_\N{\$output}\V{\_string});}}
\L{\LB{}\Tab{16}{\}}}
\L{\LB{}\Tab{16}{}}
\L{\LB{\C{}}\Tab{16}{\#_do_headers_here\CE{}}}
\L{\LB{}\Tab{16}{\N{\$header}_=_\S{}\3\$dashes\3\SE{};}}
\L{\LB{}\Tab{16}{\N{\$header}_.=_\K{sprintf}(\S{}\3\%\-36s_\3\SE{},_\N{\$simu$\ell$ator});}}
\L{\LB{}\Tab{16}{\K{foreach}_\N{\$n}_(@\V{ns})}}
\L{\LB{}\Tab{16}{\{}}
\L{\LB{}\Tab{24}{\K{foreach}_\N{\$param}_(@\{\N{\$params}\{\N{\$simu$\ell$ator}\}\})}}
\L{\LB{}\Tab{24}{\{}}
\L{\LB{}\Tab{32}{\N{\$header}_.=_\K{sprintf}(\S{}\3\%13s_\|\3\SE{},_\N{\$n});}}
\L{\LB{}\Tab{24}{\}}}
\L{\LB{}\Tab{16}{\}}}
\L{\LB{}\Tab{16}{\K{foreach}_\N{\$param}_(@\{\N{\$params}\{\N{\$simu$\ell$ator}\}\})}}
\L{\LB{}\Tab{16}{\{}}
\L{\LB{}\Tab{24}{\N{\$header}_.=_\S{}\3\2n\$param\3\SE{};}}
\L{\LB{}\Tab{16}{\}}}
\L{\LB{}\Tab{16}{\N{\$header}_.=_\S{}\3\2n\$dashes\3\SE{};}}
\L{\LB{}\Tab{16}{}}
\L{\LB{}\Tab{16}{\K{push}_(@\V{log\_output},_\N{\$header});}}
\L{\LB{}\Tab{16}{\K{push}_(@\V{log\_output},_@\V{output});}}
\L{\LB{}\Tab{16}{}}
\L{\LB{}\Tab{16}{\N{\$keyword}\V{\_string}_=_\K{join}_(\S{}\3\|\3\SE{},_@\{\N{\$keywords}\{\N{\$simu$\ell$ator}\}\});}}
\L{\LB{}\Tab{16}{@\V{temp}_=_\K{grep}_\{_/\N{\$keyword}\V{\_string}/;_\}_@\V{output};}}
\L{\LB{}\Tab{16}{}}
\L{\LB{}\Tab{16}{\K{push}_(@\V{my\_output},_\N{\$header});}}
\L{\LB{}\Tab{16}{\K{push}_(@\V{my\_output},_@\V{temp});}}
\L{\LB{}\Tab{16}{}}
\L{\LB{}\Tab{8}{\}}}
\L{\LB{}\Tab{8}{}}
\L{\LB{}\Tab{8}{}}
\L{\LB{\C{}}\Tab{8}{\#_we_store_the_complete_log\CE{}}}
\L{\LB{}\Tab{8}{\N{\$$\ell$ogname}_=_\S{}\3\$sortname\2\_\$comment\3\SE{};}}
\L{\LB{}\Tab{8}{\K{open}(\V{LOG},_\S{}\3\>\3\SE{}.\N{\$$\ell$og}\V{\_directory}.\N{\$$\ell$ogname});}}
\L{\LB{}\Tab{8}{\K{print}_\V{LOG}_\S{}\3\3\SE{}.\K{localtime}().\S{}\3\2n\3\SE{};}}
\L{\LB{}\Tab{8}{\K{print}_\V{LOG}_\S{}\3Run_as:_\2\3\$command\_line\2\3\2n\3\SE{};}}
\L{\LB{}\Tab{8}{\K{print}_\V{LOG}_@\V{log\_output};}}
\L{\LB{}}
\L{\LB{}\Tab{8}{\K{print}_\V{LOG}_\S{}\3\2nCommand_list:_\2n\3\SE{};}}
\L{\LB{}\Tab{8}{\K{foreach}_\N{\$command}_(@\V{command\_list})}}
\L{\LB{}\Tab{8}{\{}}
\L{\LB{}\Tab{16}{\K{print}_\V{LOG}_\S{}\3\2t\3\SE{}.\N{\$command.}\S{}\3\2n\3\SE{};}}
\L{\LB{}\Tab{8}{\}}}
\L{\LB{}}
\L{\LB{}\Tab{8}{\K{close}(\V{LOG});}}
\L{\LB{}}
\L{\LB{\C{}}\Tab{8}{\#_remove_the_temp\CE{}}}
\L{\LB{}\Tab{8}{\K{system}(\S{}\3rm_\$sortfile\3\SE{});}}
\L{\LB{}}
\L{\LB{\}}}
\L{\LB{\K{print}_@\V{my\_output};}}

### 2005-06-09 tech_report/docs/method.tex
inputs are now the single tiled mergesort ones moved paper to tech_report to allow new paper directory to actually be for the paper

--- 2005-05-31 paper/docs/method.tex
@@
 programming the sorts and getting results. The second section discusses tools
-that I used or created to do this.
+that were used or created to do this.
 
@@
 
-Early in the project, it was decided that exceeding the bounds of an array
+Early on, it was decided that exceeding the bounds of an array
 during a sort was to be forbidden. No serious sorting library or application
@@
 branch hits and misses. Other simulators are provided, but they were not used in
-this project.
+our tests.
 
@@
 
-The main program provided the framework for the project. Imaginatively called
-\cc{main.c}, this program provided the means for testing and (roughly)
-timing each sort. Figure \vref{Key sections of Main.c} contains this code.
+The main program provided the framework for the development of the sorts.
+Imaginatively called \cc{main.c}, this program provided the means for testing
+and (roughly) timing each sort. Figure \vref{Key sections of Main.c} contains
+this code.
 
@@
 because perfex measures results on a real machine, only one type of cache and
-branch predictor could have been measured. Since access was
-provided late in the project, limited measurements were taken to supplement the
-SimpleScalar results.
+branch predictor could have been measured. We have provided limited
+measurements to supplement the SimpleScalar results.
 
@@
 sort are described in assembly, and while they can be written in C, the result
-is a very low-level version. I attempted to rewrite the algorithm without
-\cc{goto} statements and labels, but instead using standard flow control
+is a very low-level version. An attempt was made to rewrite the algorithm
+without \cc{goto} statements and labels, instead using standard flow control
 statements such as \cc{if}, \cc{while} and \cc{do-while}. Despite the time spent
@@
 cache lines were used in SimpleScalar. The addition of real world data would
-strengthen the results from this project.
+strengthen our results.

### 2005-06-10 fyp_report/base_heapsort.tex
renamed docs to fyp_report

Intro:

One of the sorts used by lamarca is the heap sort. He dedicates a whole chapter
to heaps and stategies to improve their performace in the cache. First we need
someting to compare it to, and for this he optimises a heapsort for
instructions. Our base version ran at about x for data size N, and the latest
version did y.

Heaps
A heap is a specific type of binary tree, where each item has the heap property.
Typically, the heap property means that each child node of a binary tree has a
lower value than its parent, but the opposite is also a valid heap.

Indexing
A heap tree has an interesting property which is not shared by other types of
binary trees, and that is that it is full. Each level of the tree, except the
lowest, is full, and the lowest is filled from left to right. This means that a
heap tree will fit into an array of the same size.
An array that is indexed from 1 to N can be thought of as a heap where a node
indexed i has a parent indexed floor(i/2), and 2 children indexed i*2 and
i*2+1.  These operations are simple on any machine, and can be substituted with
i << 1 and i >> 1.
In C, arrays are indexed from 0 - N-1, breaking this indexing scheme. This can
be fixed by using a pointer to a[-1] instead of a. Hence the index 1 is a[-1+1],
and N indexes a[N-1]

Constructing a Heap (heapifying)
There are two methods to construct a heap, fixing up and fixing down. These are
also used to maintained the heap property after adding or removing items. For
this explanation we shall assume that the heap property is that a parent is
larger than its children.
Fix-up (sift up)
Sifting an item up means comparing it with its parent, and swapping it if it is
larger. This continues up the heap until 
Fix-down (sift down)

O(NlogN)
A heap has two 
Creating the heap is cheap relative the the second part. This takes N/2*(logN-1)
steps. In the second part, there are logN swaps and comparisons in a fix down,
which occurs nearly N times.  This means a near constant ONlogN, with only a
slightly slower pathalogical case.

Floyds Improvement
Floyd observed, as we will shortly, that when sifting a small item down the
heap, it goes nearly the whole way down, before making its way slowly back up.
We can avoid half the comparisons on the way down with this method. Sedgewick
notes that this is a good idea when comparisons are expensive relative to
exchanges. He notes that string comparisons are more expensive than string
exchanges (which simply comprise pointer copies). In our case of using unsigned
ints, this approach is not obviously better.
Lamarca however refers to the in place sorting method as Floyds approach.

Sorting with Heaps
Because of the heap property, a properly constructed heap will either have the
largest or the smallest item on the top. This leads to 2 ways of sorting. Both
begin by constructing a heap, followed by removing the top element and putting
it in its final position. There is an in-place method and an out of place.

In place
In this in place sort we use a top heavy heap. The heap is construted by fixing
down from the first item with a child. Once the heap in contructed, we
exchange the first item, the largest item, with the last, which is quite
relatively small. The large item is in its final position, and we reduce the
size of the heap (ie pretend its not there). We then sift the small item down the
heap, and repeat the process for all items.

Out of place
For this we use a heap with smaller items on top. The heap is constructed using
the 'Repeated Adds' method, where an item is added to the bottom of the heap,
then sifted up to fix the heap, then the process repeated for all items. Once
the heap is constructed,  the smallest item is moved to array for sorting, the
item at the end is moved to the top, then sifted down the heap. This is repeated
for all items.

How does Lamarca do it?

Whats wrong with this?

Why does he advocate this way?

Whats the performance/

What did you add?

Which directnio is it made in?

Talk some performance

### 2005-06-10 fyp_report/code/cycles_script.tex
renamed docs to fyp_report

% Remember to use the lgrind style

\Head{}
\File{cycles\_script}{2004}{4}{29}{0:06}{773}
\L{\LB{\K{set}_\V{key}_\V{top}_\V{left}_\V{Left}_\V{reverse}}}
\L{\LB{\K{set}_\V{xrange}_[\N{4096}:\N{4194304}]}}
\L{\LB{\K{set}_\V{yrange}_[\N{0}:]}}
\L{\LB{\K{set}_\V{xtics}_(\N{8192},\N{16384},\N{32768},\N{65536},\N{131072},\N{262144},\N{524288},\N{1048576},\N{2097152})}}
\L{\LB{\K{set}_\V{format}_\V{x}_\S{}\3\%8.0f\3\SE{}}}
\L{\LB{\K{set}_\V{logscale}_\V{axis}}}
\L{\LB{\K{set}_\V{data}_\V{style}_\V{linespoints}}}
\L{\LB{\K{set}_\V{term}_\V{postscript}_\V{eps}}}
\L{\LB{\K{set}_\V{output}_\S{}{'}.\,./docs/plots/all\_cycles.eps{'}\SE{}}}
\L{\LB{\K{set}_\V{xlabel}_\S{}{'}set_size_in_keys{'}\SE{}}}
\L{\LB{\K{set}_\V{ylabel}_\S{}{'}Cycles_per_key{'}\SE{}}}
\L{\LB{\V{plot}_\2}}
\L{\LB{\S{}{'}processed\_cycles\_data{'}\SE{}_\V{using}_\N{1}:\N{5}_\V{title}_\S{}{'}base_quicksort{'}\SE{},\2}}
\L{\LB{\S{}{'}processed\_cycles\_data{'}\SE{}_\V{using}_\N{1}:\N{6}_\V{title}_\S{}{'}cache_quicksort{'}\SE{},\2}}
\L{\LB{\S{}{'}processed\_cycles\_data{'}\SE{}_\V{using}_\N{1}:\N{7}_\V{title}_\S{}{'}multi_quicksort{'}\SE{},\2}}
\L{\LB{\S{}{'}processed\_cycles\_data{'}\SE{}_\V{using}_\N{1}:\N{10}_\V{title}_\S{}{'}base_mergesort{'}\SE{},\2}}
\L{\LB{\S{}{'}processed\_cycles\_data{'}\SE{}_\V{using}_\N{1}:\N{14}_\V{title}_\S{}{'}double_tiled_mergesort{'}\SE{},\2}}
\L{\LB{\S{}{'}processed\_cycles\_data{'}\SE{}_\V{using}_\N{1}:\N{16}_\V{title}_\S{}{'}double_multi\-mergesort{'}\SE{},\2}}
\L{\LB{\S{}{'}processed\_cycles\_data{'}\SE{}_\V{using}_\N{1}:\N{9}_\V{title}_\S{}{'}base_radixsort{'}\SE{}}}

### 2005-06-10 fyp_report/slide7.tex
renamed docs to fyp_report

\begin{slide}
\begin{stitle}
Results
\end{stitle}

\begin{itemize}
\item Caches
	\begin{itemize}
	\item LaMarca's cache assertions are correct
	\plot{.82}{demo1}
	\pagebreak
	\plot{1.05}{demo2}
	\pagebreak
	\item Fully Associative can be worse than Direct Mapped
	\plot{.95}{demo3}
	\pagebreak
	\end{itemize}
\item Branch Predictors
	\begin{itemize}
	\item Two-level predictor needs large tables to rival bimodal
	\plot{.9}{demo4}
	\pagebreak
	\item Bimodal predictors perform the same regardless of size, for small functions, while two-level adaptive predictors don't
	\plot{.9}{demo4}
	\pagebreak
	\item Avoiding flow control misses is possible
		\begin{itemize}
		\item Radixsort
		\plotdouble{.51}{demo6-1}{demo6-2}
		\pagebreak
		\end{itemize}
	\item Some standard algorithms have very good prediction rates
		\begin{itemize}
		\item Insertion sort has N branch misses
		\plotdouble{.51}{demo7-1}{demo7-2}
		%insertion sort vs selection and bubble
		% put these two beside each other
		\pagebreak
		\item Binary search is worse than sequential search for small lists
		\plotdouble{.51}{demo8-1}{demo8-2}
		\pagebreak
		%binary vs sequential, next to each other again
		\item Radixsort has no comparison misses
		\plot{.95}{demo6-2}
		\pagebreak
		\end{itemize}
	\item Unsorted lists perform badly
		\begin{itemize}
		\item 4 / 8 heaps - Very complex to replace with sorted lists
		\plotdouble{.51}{demo9-2}{demo9-1}
		\pagebreak
		%compare 2 heap to 4/8 heap
		\end{itemize}
	\end{itemize}
\end{itemize}

\end{slide}

### 2005-06-14 docs/fyp_report/quick.tex
moved doc dirs into docs/ moved scripts into scripts/ moves counters/ into data/

--- 2005-05-26 paper/docs/quick.tex
@@
 
-Once the entire array has been put into containers, each container is emptied
-in turn and the keys put back into the array. When a container is empty, the
-pivot greater than that container is put into position, which is guaranteed to
-be its final position.  The indices of these positions is pushed onto the
-stack, and the subarray is then sorted.
+Once the entire array has been put into containers, each container is emptied in
+turn and the keys put back into the array. When a container is empty, the pivot greater than that
+container is put into position, which is guaranteed to be its final position.
+The indices of these positions is pushed onto the stack. The stack must
+therefore be increased in size before this happens. By pushing the last
+array copied onto the stack, an entire set of cache misses is avoided.
+Unfortunately, since a multi-partition is only done when the array is
+more than twice the size of the cache, this reduction in misses in not
+significant.
 
-While emptying the containers back into the array, an opportunity is taken to
+While emptying the containers back into the array, an opportunity is
+taken\footnote{Results do not reflect this step. See \vref{bug1}.} to
 find the smallest key in the leftmost partition. This is placed as a sentinel,
@@
 
-The instruction count, shown in Figure \ref{quick instructions} is as
-predicted. The expected increases in instruction count for multi-quicksort was
-due to a test with a smaller data set and smaller constants. The actual results
-were higher than expected: being 63\% and 40\% greater, for
-sequential and binary search versions, respectively.
-%TODO check
+The instruction count, shown in Figure \ref{quick instructions} is as predicted. The expected increases in instruction
+count for multi-quicksort was due to a test with a smaller data set and smaller
+constants. The actual results were not as high as the expected ones, but closer
+to 20\% and 45\% greater, for sequential and binary search versions,
+respectively.
 
@@
 predictions, except that the multi-quicksorts did not behave as expected, and
-performed worse than the memory-tuned version. Not only do they have more
-misses per key, but the rate at which the misses increase with size is also
-worse for the multi quicksorts. The tuned version performed better than the
-base version, as predicted.
+performed worse than the memory-tuned version. However, at the edge of the graph
+their slopes are considerably less than the slopes of the other sorts, and may
+result in lower misses for larger data sets. The tuned version performed better
+than the base version, as predicted.
 
@@
 shown in Figure \ref{quick cycles} are enough to make up for the increase in
-instruction count, which was significantly reduced by using the binary search.
-The tuned quicksort, however, had the lower running time and even base
-quicksort out-ran the multi-quicksorts.
+instruction count and level 1 accesses, both of which were significantly reduced
+by using the binary search. The tuned quicksort, however, had the lower running
+time and even base quicksort out-ran the multi-quicksorts.
 
@@
 cache misses in multi-quicksort is much higher here than it is in LaMarca's
-findings. In addition, the shape is different. LaMarca's multi-quicksort's
-cache misses plateau, while the results here show an increasing slope.
-
-A previous result, however, with a different implementation showed the same
-shape, but different results to LaMarca's. That implementation waited until all
-the partitioned keys were put back into the array, instead of being sorted
-straight away. Those results plateaued once mutli-quicksort took effect.
-
-\section{Predictor results}
-\sixplot{base_quicksort-0.eps}{base_quicksort9-0.eps}{cache_quicksort-0.eps}{multi_quicksort-0.eps}{sequential_multi_quicksort-0.eps}{}{base, cache and multi quicksorts and base quicksort9}
-\sixplot{base_quicksort1-0.eps}{base_quicksort-0.eps}{base_quicksort5-0.eps}{base_quicksort7-0.eps}{base_quicksort9-0.eps}{}{base quicksort without a median, with median of 3, and with pseudo median of 5, 7 and 9}
-
-%TODO a bit to rewrite
-
-Theres two important bits: comparison between the types of sorts, and comparisons between the different number of things tried as medians.
-
-Firstly, cache quicksort and base quicksort only idffer in how they deal with
-the insertion. Base quicksort and base quicksort9 have similar numbers of
-insertion branches, and a very similar ratio (2/1 correct/incorrect) of
-mispredictions. Comparing base quicksort to cahce quicksort (they both use
-median of 3), there is actually a decrease in the number of branches handled by
-insertion sort, instead of an increase. An increase is expected (thats why
-sedgewick moves the insertion to the end), though when you take into account
-the sentinel and the outer loop, there probably is an increase in instructions
-at least. Base quicksort has 35\% miss rate, but cache has a 34\%. Not a big
-difference, but the point is that there is a reduction in branches, but of the
-reduced branches, 50\% were unpredictable. I dont think theres anything
-interesting in this fact, however.
-
-When going to multi-quicksort, there si quite a drop in the number fo branches.
-Once the sort is larger than the cache, the number fo steps increases linearly
-(since it does a sort of counting like radix sort). Theres a decrease of 19 \%.
-Since the sort at this point is NlogM where M is the size of the cache (in
-keys), there shouldnt be a change in misprediction rates. And there sint.
-Before: 63\%. After, 63\%. So no difference.
-
-Theres a lot of branches in the multi quicksort. WE saw in other results that
-the mere instruction cost of this bogs down to algorithm significantly. in the
-binary search version, there are more than twice as many mispredictions,
-despite the fact that there are less than half the branches. (note that i dont
-think this was a wonderful binary search, its heavily biased towards the
-checking the left). binary search predicts correctly 57 percent of the time,
-meaning that it mostly changes direction haphazardly during the search. No
-wonder it performs so badly.
-
-Note that sequential multi-quicksort has less mispredictions than cache quicksort.
-
-Part 2: comparing different ways of randomly partitioning:
-
-Ways of getting the partition:
-1 - just take the rightmost element. Note that there are no sentinels, so you have to put a bounds check back in
-3 - median of(0,1,2) halves
-5 - median of(0, (median of(1,2,3), 4) quarters
-7 - median of(median of(0,1,2), 3, median of(4,5,6)) sixths
-9 - median of(median of(0,1,2), median of(3,4,5), median of(6,7,8)) eights
-
-misprediction rates:
-1 - 29.4\%
-3 - 35.9\%
-5 - 38.4\%
-7 - 40.0\%
-9 - 41.7\%
-
-The misprediction rates point to the problem.
-They're just not being evenly divided.
-The number of times the partitions end is the same, roughly, each time.
-However, the number of actual branches is much higher in the case of quicksort1, and decreases all the way down.
-At the same time, the number of mispredictions creeps up.
-And hence the big change in the rate of mispredictions.
-As predicted, the better the median, the better the partition, and the less correct predictions pretending to be useful.
-There must be a 50\% rate waiting to get out, so a little more than median of 9 maybe.
-Or a lower cut-off point (it cuts out for partitions of less than 40 keys, and uses plain old median of 3).
-
-the rate or number of insertion branches doesnt change at all.
-
-%TODO up to here
+findings. As with mergesort though, while the results are different, the shape
+is similar: multi-quicksort has a much flatter slope than the other sorts.  This
+difference is likely to be due to the implementation of multi-quicksort; a
+slightly different approach, which is described immediately below, may have the
+effect of changing the results sufficiently that they may agree with LaMarca's.
 
 \section{Future Work}
+In multi-quicksort, it may be possible to reduce the cache misses by sorting
+each containerful of keys as soon as it is put back into the array. This step would
+significantly increase the temporal locality, and may reduce the number of cache
+misses by half.
+
 The threshold in quicksort is the point at which a simpler sort is used to sort

### 2005-06-14 docs/fyp_report/simplescalar.tex
moved doc dirs into docs/ moved scripts into scripts/ moves counters/ into data/

\label{SimpleScalar Install}

This section details how to install SimpleScalar on a Pentium 4 running RedHat
Linux 7. The reason for its inclusion is that the installation was difficult,
and solutions to problems encountered were not easily available on the
SimpleScalar website, nor were emails answered by the authors.

The primary guide this is based on is \cite{SimpleScalarInstall}. However, that
guide is more general than this.

\section{Download}
Go to \textit{http://www.simplescalar.com/tools}. The simulations in this
project used the standard packages provided on this page. Links to other
packages here were not used. Download simplesim-3v0d.tgz, simplesim-3v0d.tgz
and simplesim-3v0d.tgz. Unzip the tools. The following directories should be
present: binutils-2.5.2,  f2c-1994.09.27,  gcc-2.6.3,  glibc-1.09,
simplesim-3.0, ssbig-na-sstrix and sslittle-na-sstrix.

\section{Compile}
Run the following commands:
\begin{verbatim}
cd binutils-2.5.2
./configure --host=i386-redhat-linux --target=sslittle-na-sstrix 
	--with-gnu-as --with-gnu-ld --prefix=/home/ciaran/simple

make
make install
cd ../simplesim-3.0
make
cd ../gcc-2.6.3
./configure --host=i386-redhat-linux --target=sslittle-na-sstrix 
	--with-gnu-as --with-gnu-ld --prefix=/home/ciaran/simple
make LANGUAGES=c
\end{verbatim}

The following error may occur at this point:
\begin{verbatim}
cccp.c:194: conflicting types for `sys_errlist'
/usr/include/stdio.h:554: previous declaration of `sys_errlist'
\end{verbatim}

Comment out line 194 of \texttt{cccp.c}.

This error then occurs:
\begin{verbatim}
gcc -c -DCROSS_COMPILE -DIN_GCC -DPOSIX 
	-g -I. -I. -I./config sdbout.c
sdbout.c:57:18: syms.h: No such file or directory
\end{verbatim}

This can be resolved\footnote{Credit for this is due to
http://moss.csc.ncsu.edu/~mueller/rt/rt02/g1/sim2install.txt} by changing line
57 of \texttt{sdbout.c} from \verb!#include <syms.h>! to \verb!#include "gsyms.h"!.

\begin{verbatim}
bc-typecd.def: In function `bc_emit_instruction':
bc-typecd.def:17: `SFtype' is promoted to `double' 
					when passed through `...'
bc-typecd.def:17: (so you should pass `double' not 
					`SFtype' to `va_arg')
\end{verbatim}

In line 17 of \texttt{bc-typecd.def}, change \texttt{SFtype} to \texttt{DFtype}.

\begin{verbatim}
gcc.c:172: conflicting types for `sys_errlist'
/usr/include/stdio.h:554: previous declaration of `sys_errlist'
\end{verbatim}

Comment out line 172 of \texttt{gcc.c}

The tools should now be compiled. For speed, the architecture type could be
changed to \texttt{i686} and the \texttt{-O3} flag could be set.

### 2005-06-14 docs/fyp_report/slide2.tex
moved doc dirs into docs/ moved scripts into scripts/ moves counters/ into data/

\begin{slide}

\begin{stitle}
Purpose
\end{stitle}

\begin{itemize}
\item Implement and test LaMarca's cache-conscious sorts
\begin{itemize}
\item expect same results as LaMarca's
\end{itemize}
\item Measure cache and branch prediction results
\begin{itemize}
\item Fully Associative vs. Direct Mapped Caches
\item Bimodal vs. 2 Level Adaptive Predictors
\end{itemize}
\item Do changes to algorithms:
\begin{itemize}
\item cause different branch effects?
\item increase/decrease branch misses?
\item affect different predictors differently?
\item \bfseries{Why?}
\end{itemize}
\end{itemize}

\end{slide}

### 2005-06-14 docs/tech_report/docs/code/sample_output.tex
moved doc dirs into docs/ moved scripts into scripts/ moves counters/ into data/

% Remember to use the lgrind style

\Head{}
\File{sample\_output}{2004}{4}{28}{4:03}{4376}
\L{\LB{\V{sim}\-\V{bpred}:_\V{SimpleScalar}/\V{PISA}_\V{Tool}_\V{Set}_\V{version}_\N{3.0}_\V{of}_\V{August},_\N{2003.}}}
\L{\LB{\V{Copyright}_(\V{c})_\N{1994}\-\N{2003}_\V{by}_\V{Todd}_\V{M}._\V{Austin},_\V{Ph}.\V{D}._\V{and}_\V{SimpleScalar},_\V{LLC}.}}
\L{\LB{\V{All}_\V{Rights}_\V{Reserved}._\V{This}_\V{version}_\V{of}_\V{SimpleScalar}_\V{is}_\V{licensed}_\K{for}_\V{academic}}}
\L{\LB{\V{non}\-\V{commercial}_\V{use}.}\Tab{21}{\V{No}_\V{portion}_\V{of}_\V{this}_\V{work}_\V{may}_\V{be}_\V{used}_\V{by}_\V{any}_\V{commercial}}}
\L{\LB{\V{entity},_\V{or}_\K{for}_\V{any}_\V{commercial}_\V{purpose},_\V{without}_\V{the}_\V{prior}_\V{written}_\V{permission}}}
\L{\LB{\V{of}_\V{SimpleScalar},_\V{LLC}_(\V{info}@\V{simplescalar}.\V{com}).}}
\L{\LB{}}
\L{\LB{\V{sim}:_\V{command}_\V{line}:_\V{sim}\-\V{bpred}_\V{base\_heapsort454574856}_}}
\L{\LB{}}
\L{\LB{\V{sim}:_\V{simulation}_\V{started}_@_\V{Wed}_\V{Apr}_\N{28}_\N{04}:\N{07}:\N{08}_\N{2004},_\V{options}_\V{follow}:}}
\L{\LB{}}
\L{\LB{\V{sim}\-\V{bpred}:_\V{This}_\V{simulator}_\V{implements}_\V{a}_\V{branch}_\V{predictor}_\V{analyzer}.}}
\L{\LB{}}
\L{\LB{\C{}\#_-config}\Tab{30}{\#_load_configuration_from_a_file\CE{}}}
\L{\LB{\C{}\#_-dumpconfig}\Tab{30}{\#_dump_configuration_to_a_file\CE{}}}
\L{\LB{\C{}\#_-h}\Tab{24}{false_\#_print_help_message}\Tab{54}{\CE{}}}
\L{\LB{\C{}\#_-v}\Tab{24}{false_\#_verbose_operation}\Tab{54}{\CE{}}}
\L{\LB{\C{}\#_-d}\Tab{24}{false_\#_enable_debug_message}\Tab{54}{\CE{}}}
\L{\LB{\C{}\#_-i}\Tab{24}{false_\#_start_in_Dlite_debugger\CE{}}}
\L{\LB{\-\V{seed}}\Tab{28}{\N{1}_\C{}\#_random_number_generator_seed_(0_for_timer_seed)\CE{}}}
\L{\LB{\C{}\#_-q}\Tab{24}{false_\#_initialize_and_terminate_immediately\CE{}}}
\L{\LB{\C{}\#_-chkpt}\Tab{23}{\<null\>_\#_restore_EIO_trace_execution_from_\<fname\>\CE{}}}
\L{\LB{\C{}\#_-redir:sim}\Tab{23}{\<null\>_\#_redirect_simulator_output_to_file_(non-interactive_only)\CE{}}}
\L{\LB{\C{}\#_-redir:prog}\Tab{23}{\<null\>_\#_redirect_simulated_program_output_to_file\CE{}}}
\L{\LB{\-\V{nice}}\Tab{28}{\N{0}_\C{}\#_simulator_scheduling_priority\CE{}}}
\L{\LB{\-\V{max}:\V{inst}}\Tab{28}{\N{0}_\C{}\#_maximum_number_of_inst{'}s_to_execute\CE{}}}
\L{\LB{\-\V{bpred}}\Tab{24}{\V{bimod}_\C{}\#_branch_predictor_type_\{nottaken\|taken\|bimod\|2lev\|comb\}\CE{}}}
\L{\LB{\-\V{bpred}:\V{bimod}}\Tab{15}{\N{2048}_\C{}\#_bimodal_predictor_config_(\<table_size\>)\CE{}}}
\L{\LB{\-\V{bpred}:\N{2$\ell$ev}}\Tab{15}{\N{1}_\N{1024}_\N{8}_\N{0}_\C{}\#_2-level_predictor_config_(\<l1size\>_\<l2size\>_\<hist\_size\>_\<xor\>)\CE{}}}
\L{\LB{\-\V{bpred}:\V{comb}}\Tab{15}{\N{1024}_\C{}\#_combining_predictor_config_(\<meta\_table\_size\>)\CE{}}}
\L{\LB{\-\V{bpred}:\V{ras}}\Tab{27}{\N{8}_\C{}\#_return_address_stack_size_(0_for_no_return_stack)\CE{}}}
\L{\LB{\-\V{bpred}:\V{btb}}\Tab{15}{\N{512}_\N{4}_\C{}\#_BTB_config_(\<num\_sets\>_\<associativity\>)\CE{}}}
\L{\LB{}}
\L{\LB{}\Tab{2}{\V{Branch}_\V{predictor}_\V{configuration}_\V{examples}_\K{for}_\N{2}\-\V{level}_\V{predictor}:}}
\L{\LB{}\Tab{4}{\V{Configurations}:}\Tab{22}{\V{N},_\V{M},_\V{W},_\V{X}}}
\L{\LB{}\Tab{6}{\V{N}}\Tab{10}{\C{}\#_entries_in_first_level_(\#_of_shift_register(s))\CE{}}}
\L{\LB{}\Tab{6}{\V{W}}\Tab{10}{\V{width}_\V{of}_\K{shift}_\V{register}(\V{s})}}
\L{\LB{}\Tab{6}{\V{M}}\Tab{10}{\C{}\#_entries_in_2nd_level_(\#_of_counters,_or_other_FSM)\CE{}}}
\L{\LB{}\Tab{6}{\V{X}}\Tab{10}{(\V{yes}\-\N{1}/\V{no}\-\N{0})_\V{xor}_\V{history}_\V{and}_\V{address}_\K{for}_\N{2nd}_\V{level}_\V{index}}}
\L{\LB{}\Tab{4}{\V{Sample}_\V{predictors}:}}
\L{\LB{}\Tab{6}{\V{GAg}}\Tab{14}{:_\N{1},_\V{W},_\N{2}\5\V{W},_\N{0}}}
\L{\LB{}\Tab{6}{\V{GAp}}\Tab{14}{:_\N{1},_\V{W},_\V{M}_(\V{M}_\>_\N{2}\5\V{W}),_\N{0}}}
\L{\LB{}\Tab{6}{\V{PAg}}\Tab{14}{:_\V{N},_\V{W},_\N{2}\5\V{W},_\N{0}}}
\L{\LB{}\Tab{6}{\V{PAp}}\Tab{14}{:_\V{N},_\V{W},_\V{M}_(\V{M}_==_\N{2}\5(\V{N}+\V{W})),_\N{0}}}
\L{\LB{}\Tab{6}{\V{gshare}}\Tab{14}{:_\N{1},_\V{W},_\N{2}\5\V{W},_\N{1}}}
\L{\LB{}\Tab{2}{\V{Predictor}_{`}\V{comb}\S{}{'}_combines_a_bimodal_and_a_2\-level_predictor.}}
\L{\LB{}}
\L{\LB{}}
\L{\LB{}}
\L{\LB{sim:_**_starting_functional_simulation_w/_predictors_**}}
\L{\LB{}}
\L{\LB{sim:_**_simulation_statistics_**}}
\L{\LB{sim\_num\_insn}\Tab{28}{1045897_\#_total_number_of_instructions_executed}}
\L{\LB{sim\_num\_refs}\Tab{29}{264304_\#_total_number_of_loads_and_stores_executed}}
\L{\LB{sim\_elapsed\_time}\Tab{34}{1_\#_total_simulation_time_in_seconds}}
\L{\LB{sim\_inst\_rate}\Tab{23}{1045897.0000_\#_simulation_speed_(in_insts/sec)}}
\L{\LB{sim\_num\_branches}\Tab{29}{167628_\#_total_number_of_branches_executed}}
\L{\LB{sim\_IPB}\Tab{29}{6.2394_\#_instruction_per_branch}}
\L{\LB{bpred\_bimod.lookups}\Tab{29}{167628_\#_total_number_of_bpred_lookups}}
\L{\LB{bpred\_bimod.updates}\Tab{29}{167628_\#_total_number_of_updates}}
\L{\LB{bpred\_bimod.addr\_hits}\Tab{29}{139898_\#_total_number_of_address\-predicted_hits}}
\L{\LB{bpred\_bimod.dir\_hits}\Tab{29}{140006_\#_total_number_of_direction\-predicted_hits}}
\L{\LB{bpred\_bimod.misses}\Tab{30}{27622_\#_total_number_of_misses}}
\L{\LB{bpred\_bimod.jr\_hits}\Tab{31}{8551_\#_total_number_of_address\-predicted_hits_for_JRs}}
\L{\LB{bpred\_bimod.jr\_seen}\Tab{31}{8552_\#_total_number_of_JR{'}\SE{}\V{s}_\V{seen}}}
\L{\LB{\V{bpred\_bimod}.\V{jr\_non\_ras\_hits}.\V{PP}_\N{0}_\C{}\#_total_number_of_address-predicted_hits_for_non-RAS_JRs\CE{}}}
\L{\LB{\V{bpred\_bimod}.\V{jr\_non\_ras\_seen}.\V{PP}_\N{0}_\C{}\#_total_number_of_non-RAS_JR{'}s_seen\CE{}}}
\L{\LB{\V{bpred\_bimod}.\V{bpred\_addr\_rate}}\Tab{31}{\N{0.8346}_\C{}\#_branch_address-prediction_rate\CE{}}}
\L{\LB{\V{bpred\_bimod}.\V{bpred\_dir\_rate}}\Tab{30}{\N{0.8352}_\C{}\#_branch_direction-prediction_rate\CE{}}}
\L{\LB{\V{bpred\_bimod}.\V{bpred\_jr\_rate}}\Tab{29}{\N{0.9999}_\C{}\#_JR_address-prediction_rate\CE{}}}
\L{\LB{\V{bpred\_bimod}.\V{bpred\_jr\_non\_ras\_rate}.\V{PP}_\<\V{error}:_\V{divide}_\V{by}_\N{0}\>_\C{}\#_non-RAS_JR_addr-pred_rate\CE{}}}
\L{\LB{\V{bpred\_bimod}.\V{retstack\_pushes}}\Tab{36}{\N{8554}_\C{}\#_total_number_of_address_pushed_onto_ret-addr_stack\CE{}}}
\L{\LB{\V{bpred\_bimod}.\V{retstack\_pops}}\Tab{34}{\N{8552}_\C{}\#_total_number_of_address_popped_off_of_ret-addr_stack\CE{}}}
\L{\LB{\V{bpred\_bimod}.\V{used\_ras}.\V{PP}}\Tab{32}{\N{8552}_\C{}\#_total_number_of_RAS_predictions_used\CE{}}}
\L{\LB{\V{bpred\_bimod}.\V{ras\_hits}.\V{PP}}\Tab{32}{\N{8551}_\C{}\#_total_number_of_RAS_hits\CE{}}}
\L{\LB{\V{bpred\_bimod}.\V{ras\_rate}.\V{PP}}\Tab{27}{\N{0.9999}_\C{}\#_RAS_prediction_rate_(i.e.,_RAS_hits/used_RAS)\CE{}}}
\L{\LB{}}

### 2005-06-14 docs/tech_report/docs/intro.tex
moved doc dirs into docs/ moved scripts into scripts/ moves counters/ into data/

--- 2005-05-28 paper/docs/intro.tex
@@
 also discusses the theory of sorting, and conventions used in the algorithms in
-this project.
+this report.
 
-Chapter \ref{method}  discusses the method and framework used in this project, including the
+Chapter \ref{method} discusses the method and framework used in this research, including the
 tools used and the reasons for using them.

### 2005-06-14 docs/tech_report/docs/quick.tex
moved doc dirs into docs/ moved scripts into scripts/ moves counters/ into data/

--- 2005-05-26 paper/docs/quick.tex
@@
 %\section{Description}
-
+\label{quick}
 Quicksort is an in-place, unstable sort. It was first described in
@@
 
-LaMarca uses the optimised quicksort described by Sedgewick for his base
+LaMarca uses the optimized quicksort described by Sedgewick for his base
 quicksort. This is an iterative quicksort which uses an explicit stack. In the
@@
 initialised at the start of the algorithm. To reduce the size of the stack,
-Sedgewick proposes sorting the smallest partition first. This optimisation
+Sedgewick proposes sorting the smallest partition first. This optimization
 reduces the maximum size of the stack to $O(logN)$. In practice, two \cc{int}s,

### 2005-06-15 docs/tech_report/merge.tex
made the thing compile now

--- 2005-06-07 paper/docs/merge.tex
@@
 
-\plot{merge}{}{}{Results shown are for a Direct Mapped cache.}{Except where
-marked, results shown are for a Direct Mapped cache.}{}{}
+\plot{merge}{}{}{Except where marked, results shown are for a Direct Mapped cache.}{
+Results shown are for a Direct Mapped cache.}{}{}
 
@@
 
-\afterpage{
-\setcounter{subfigure}{0}
-\thispagestyle{empty}
-\clearpage
-\enlargethispage{14em}
-\vspace*{-5em}
-\begin{figure}[H]
-\begin{changemargin}
-\subfigure[]{\includegraphics[scale=0.74]{../counters/algorithm_n-0.eps}}
-\subfigure[]{\includegraphics[scale=0.74]{../counters/algorithm_s-0.eps}}
-\subfigure[]{\includegraphics[scale=0.74]{../counters/base_mergesort-0.eps}}
-\subfigure[]{\includegraphics[scale=0.74]{../counters/double_tiled_mergesort-0.eps}}
-\subfigure[]{\includegraphics[scale=0.74]{../counters/double_multi_mergesort-0.eps}}
-\subfigure[]{\includegraphics[scale=0.74]{../counters/double_multi_mergesort-1.eps}}
-\end{changemargin}
-\vspace*{6em}
-\caption{Branch prediction performance for mergesorts. All graphs use the same scale.}
-\label{Branch prediction performance for mergesorts}
-\end{figure}
-\newpage
-}
+\sixplot{algorithm_n_0.eps}{algorithm_s_0.eps}{base_mergesort_0.eps}{double_tiled_mergesort_0.eps}{double_multi_mergesort_0.eps}{double_multi_mergesort_0.eps}{mergesorts}
+%\afterpage{
+%\setcounter{subfigure}{0}
+%\thispagestyle{empty}
+%\clearpage
+%\enlargethispage{14em}
+%\vspace*{-5em}
+%\begin{figure}[H]
+%\begin{changemargin}
+%\subfigure[]{\includegraphics[scale=0.74]{../counters/algorithm_n-0.eps}}
+%\subfigure[]{\includegraphics[scale=0.74]{../counters/algorithm_s-0.eps}}
+%\subfigure[]{\includegraphics[scale=0.74]{../counters/base_mergesort_0.eps}}
+%\subfigure[]{\includegraphics[scale=0.74]{../counters/double_tiled_mergesort_0.eps}}
+%\subfigure[]{\includegraphics[scale=0.74]{../counters/double_multi_mergesort_0.eps}}
+%\subfigure[]{\includegraphics[scale=0.74]{../counters/double_multi_mergesort-1.eps}}
+%\end{changemargin}
+%\vspace*{6em}
+%\caption{Branch prediction performance for mergesorts. All graphs use the same scale.}
+%\label{Branch prediction performance for mergesorts}
+%\end{figure}
+%\newpage
+%}
 

### 2005-06-15 docs/tech_report/quick.tex
made the thing compile now

--- 2005-06-14 docs/tech_report/docs/quick.tex
@@
 \section{Predictor results}
-\sixplot{base_quicksort-0.eps}{base_quicksort9-0.eps}{cache_quicksort-0.eps}{multi_quicksort-0.eps}{sequential_multi_quicksort-0.eps}{}{base, cache and multi quicksorts and base quicksort9}
-\sixplot{base_quicksort1-0.eps}{base_quicksort-0.eps}{base_quicksort5-0.eps}{base_quicksort7-0.eps}{base_quicksort9-0.eps}{}{base quicksort without a median, with median of 3, and with pseudo median of 5, 7 and 9}
+\sixplot{base_quicksort_0.eps}{base_quicksort9_0.eps}{cache_quicksort_0.eps}{multi_quicksort_0.eps}{sequential_multi_quicksort_0.eps}{}{base, cache and multi quicksorts and base quicksort9}
+\sixplot{base_quicksort1_0.eps}{base_quicksort_0.eps}{base_quicksort5_0.eps}{base_quicksort7_0.eps}{base_quicksort9_0.eps}{}{base quicksort without a median, with median of 3, and with pseudo median of 5, 7 and 9}
 

### 2005-06-15 docs/tech_report/result_list.tex
made the thing compile now

\label{result_list}
\afterpage{
\smallplot{on2_insertsort}{insertion sort}{}{}{}{}{}
\smallplot{on2_selectsort}{selection sort}{}{}{}{}{}
\smallplot{on2_bubblesort}{bubblesort}{}{}{}{}{}
\smallplot{on2_bubblesort2}{bubblesort2}{}{}{}{}{}
\smallplot{on2_shakersort}{shakersort}{}{}{}{}{}
\smallplot{on2_shakersort2}{shakersort2}{}{}{}{}{}
%
\smallplot{base_heapsort}{base heapsort}{}{}{}{}{}
\smallplot{cache4_heapsort}{cache heapsort (4-heap)}{}{}{}{}{}
\smallplot{cache4_heapsort_small_cache_line}{cache heapsort (4-heap) (with 16-byte cache line)}{}{}{}{}{}
\smallplot{cache_heapsort}{cache heapsort (8-heap)}{}{}{}{}{}
%
\smallplot{algorithm_n}{algorithm N}{}{}{}{}{}
\smallplot{algorithm_s}{algorithm S}{}{}{}{}{}
\smallplot{base_mergesort}{base mergesort}{}{}{}{}{}
\smallplot{double_tiled_mergesort}{double tiled mergesort}{}{}{}{}{}
\smallplot{double_multi_mergesort}{double multi-mergesort}{}{}{}{}{}
%
\smallplot{base_quicksort}{base quicksort}{}{}{}{}{}
\smallplot{cache_quicksort}{memory-tuned quicksort}{}{}{}{}{}
\smallplot{multi_quicksort}{multi-quicksort (binary search)}{}{}{}{}{}
\smallplot{multi_quicksort_seq}{multi-quicksort (sequential search)}{}{}{}{}{}
%
\smallplot{base_radixsort}{base radixsort}{}{}{}{}{}
\smallplot{cache_radixsort}{cache radixsort}{}{}{}{}{}
%
\smallplot{gonnet_shellsort}{gonnet shellsort}{}{}{}{}{}
\smallplot{improved_gonnet_shellsort}{improved gonnet shellsort}{}{}{}{}{}
}

### 2005-06-15 docs/tech_report/software_predictor.tex
made the thing compile now

\afterpage{
\sixplot{base_heapsort_0.eps}{cache4_heapsort_0.eps}{cache4_heapsort_1.eps}{cache_heapsort_0.eps}{cache_heapsort_1.eps}{}{base and cache heapsorts}

\sixplot{base_quicksort_0.eps}{base_quicksort9_0.eps}{cache_quicksort_0.eps}{multi_quicksort_0.eps}{sequential_multi_quicksort_0.eps}{}{base, cache and multi quicksorts and base quicksort9}
\sixplot{base_quicksort1_0.eps}{base_quicksort_0.eps}{base_quicksort5_0.eps}{base_quicksort7_0.eps}{base_quicksort9_0.eps}{}{base quicksort without a median, with median of 3, and with pseudo median of 5, 7 and 9}

\sixplot{algorithm_n_0.eps}{algorithm_n_1.eps}{algorithm_s_0.eps}{algorithm_s_1.eps}{}{}{Knuth's Mergesorts}
\sixplot{base_mergesort_0.eps}{base_mergesort_1.eps}{base_mergesort_2.eps}{base_mergesort_3.eps}{base_mergesort_4.eps}{base_mergesort_5.eps}{base mergesort}
\sixplot{double_tiled_mergesort_0.eps}{double_tiled_mergesort_1.eps}{double_tiled_mergesort_2.eps}{double_tiled_mergesort_3.eps}{double_tiled_mergesort_4.eps}{double_tiled_mergesort_5.eps}{double tiled mergesort - part 1}
\sixplot{double_tiled_mergesort_6.eps}{double_tiled_mergesort_7.eps}{double_multi_mergesort_0.eps}{double_multi_mergesort_1.eps}{double_multi_mergesort_2.eps}{double_multi_mergesort_3.eps}{double tiled mergesort - part 2, and double multi mergesort - part 1}
\sixplot{double_multi_mergesort_4.eps}{double_multi_mergesort_5.eps}{double_multi_mergesort_6.eps}{double_multi_mergesort_7.eps}{double_multi_mergesort_8.eps}{double_multi_mergesort_9.eps}{double multi-mergesort - part 2}

\sixplot{gonnet_shellsort_0.eps}{gonnet_improved_shellsort_0.eps}{}{}{}{}{Shellsorts}
}

### 2005-06-16 docs/tech_report/merge.tex
fixed a demo script that was overwriting a merge result fixed the writing in merge sort for the level 1 results

@@
 here on as `base mergesort', is a simple higher level version of algorithm S,
-and performs only slightly more slowly.
+and should perform only slightly more slowly.
 
@@
 
-\section{Double Tiled Mergesort}
+\section{Tiled Mergesort}
 LaMarca's first problem with mergesort is the placement of the arrays with
@@
 the cache is allocated, which can be expensive for small arrays. The address of
-the auxiliary array is then offset such that $index(source) + index(auxiliary) =
-C$.
-
-%TODO what does this formula mean. look up LaMarcas paper
+the auxiliary array is then offset such that the block index in the source array
+is $CacheSize/2$ from the block index in the target array. The is achieved by
+inverting the most significant bit of the block index in the target array.
 
@@
 
-The final improvement was to fully sort keys in the level 1 cache, before
-continuing to level 2 cache as in the tiled mergesort. The expected improvement
-is not as significant as in the case of the level 2 cache, since the penalty
-for level 1 misses is not great. For this reason, this sort is called Double
-Tiled Mergesort instead of Tiled Mergesort, however, it is referred to as both
-in this report.
+\section{Double Tiled Mergesort}
+Offsetting the target array to avoid conflicts reduces significantly the number
+of level 2 cache misses. However, it does this at the expense of increasing the
+number of level 1 cache misses. To avoid this, we furthur offset the target
+array by $Level1CacheSize/2$. It is expected that the number of level 1 misses
+should be greatly decreased, at the expense of very few level 2 misses. The
+expected improvement is not as great as between base mergesort and tiled
+mergesort, but some improvement should be seen.
 
-\section{Double Multi-mergesort}
+\section{Multi-mergesort}
 
-Double Multi-mergesort has two phases. The first is the same as tiled mergesort, and
-sorts the array into lists sized $C/2$. The second phase attempts to improve
-temporal locality by merging these lists together in a single merge step. There
-are $k$ lists and this is called a \n{k-way merge}. It is possible to do a
-search across each of these keys, maintaining a list of indices to the smallest
-unused key from each list. When there is a small number of lists, this isn't a
-problem, however, when the number of lists increases significanty, the overhead
-involved in this is substantial. It is necessary to check each of the lists, and
-each key searched would cause a cache miss, since each array segment is aligned,
-and map to the same cache block.
+Multi-mergesort has two phases. The first is the same as tiled mergesort, and
+sorts the array into lists half the size of the cache. The second phase attempts
+to improve temporal locality by merging these lists together in a single merge
+step. There are $k$ lists and this is called a \n{k-way merge}. It is possible
+to do a search across each of these keys, maintaining a list of indices to the
+smallest unused key from each list. When there is a small number of lists, this
+isn't a problem, however, when the number of lists increases significanty, the
+overhead involved in this is substantial. It is necessary to check each of the
+lists, and each key searched would cause a cache miss, since each array segment
+is aligned, and map to the same cache block.
 
@@
 
+\section{Double Multi-mergesort}
+
+Double Multi-mergesort combines the double tiling of Double Tiled mergesort with
+Multi-mergesort.
+
 \section{Results}
 \subsection{Test Parameters}
-The level 1 cache is 8k, so 2048 keys can be sorted in the level 1 cache.
-262144 keys can be sorted in the level 2 cache. When arrays were pre-sorted,
-they were pre-sorted to a size four, or eight if the number of passes was going
-to be odd. These were reversed for the double multi-mergesort; since it does a
-k-way merge into the source array, the sorted cache sized segments need to be in
-the auxiliary array. An 8-heap was used for the multi-mergesorts.
+The level 1 cache is 8k, so 1024 keys from each array can be sorted in the level
+1 cache. 262144 keys can be sorted in the level 2 cache. When arrays were
+pre-sorted, they were pre-sorted to a size four, or eight if the number of
+passes was going to be odd. These were reversed for the double multi-mergesort;
+since it does a k-way merge into the source array, the sorted cache sized
+segments need to be in the auxiliary array. An 8-heap was used for the
+multi-mergesorts.
 
@@
 
-Algorithm N and algorithm S are included in the results for reference, but are
-not discussed here.
+Algorithm N is not optimized for random data, and should not perform as well as
+algorithm S. Algorithm S is expected to have a lower instruction count than base
+mergesort. Neither algorithm is expected to have good cache performance,
 
@@
 according to input. Multi-mergesort should have a significantly higher
-instruction count once its threshold is hit.
+instruction count once its threshold is hit. The double merges have exactly the
+same flow control as the single merges, so will only differ in cache misses.
 
@@
 correspondingly lower. Due to their design, the double mergesorts should have
-better performance than their counterparts.
+better performance than their counterparts, both in level 1 misses, and overall,
+though the difference is not expected to be great.
 
@@
 
-\plot{merge}{}{}{Except where marked, results shown are for a Direct Mapped cache.}{
+\plot{merge}{}{}{Results shown are for a Direct Mapped cache.}{
 Results shown are for a Direct Mapped cache.}{}{}
 
-The number of instructions, shown in Figure \ref{merge instructions} is exactly
-as predicted, with a 50\% spike at multi-mergesort. Level 2 misses, shown in
-Figure \ref{merge level 2 misses} are also as expected, although the
-multi-mergesort does not show the same improvement over tiled mergesort as in
-LaMarca's results.
+As expected, algorithm N has lower instruction count than algorithm S, which
+also slightly outperforms base mergesort.  The number of instructions, shown in
+Figure \ref{merge instructions} is exactly as predicted, with a 50\% spike at
+multi-mergesort. Level 2 misses, shown in Figure \ref{merge level 2 misses} are
+also as expected, although the multi-mergesort does not show the same
+improvement over tiled mergesort as in LaMarca's results.
 
-%TODO look over lamarcas results
 The prediction results in Figure \ref{Branch Simulation results for mergesort}
@@
 
+As predicted, the rate of level 1 misses is greatly reduced by the double
+tiling, as shown in Figure \ref{merge level 1 misses}. Double multi-mergesort
+has consistantly four times fewer level 1 misses than multi-mergesort, and both
+double tiled sorts perform significantly better than the unoptimizied sorts.
+
+The level 2 cache results are startingly similar to LaMarca's. The shape of the
+Figure \ref{merge level 2 misses} is exactly that provided by LaMarca. The
+ratio's between the each of the sorts is also the same, though the numbers are
+not identical. As we use a different base algorithm from LaMarca, this is not
+unthinkable.
+
 Despite the improvement in cache misses due to multi-mergesort, it still cannot
-account for the increase in instruction count as a result, and it has a higher
+compensate for the increase in instruction count as a result, and it has a higher
 cycle count on the Pentium 4 (see Figure \ref{merge cycles}) as a result. The
@@
 
-The results presented here differ from LaMarca's results. In the case of
-instruction count, both sets are nearly identical: the graphs are the same shape
-and have similar ratios between the base and multi-mergesorts. The cache results
-have a similar shape to LaMarca's, with a large increase in misses as soon as
-the arrays no longer fit in the cache, and similar ratios between the base and
-tiled versions.  But while LaMarca reports a very small increase in level 2
-misses once multi-mergesort's data set is larger than the cache, the results
-here show nearly the same jump as tiled mergesort. This results in half as many
-misses as the base mergesort, while LaMarca shows nearly 10 times fewer misses
-for his multi-mergesort.
-
-As a result of this discrepancy, it is not a surprise that LaMarca's results
-indicate that multi-mergesort is far faster than base-mergesort, but the same
-results are not shown here. The increase in instruction count cripples
-multi-mergesort, and tiled mergesort emerges the victor.
-
-This is not to say that LaMarca's results disagree with those here. In fact, it
-is noticeable that multi-mergesort's cache misses plateau at the edge of the
-graph, while other mergesorts' misses continue to climb. This is in keeping with
-both LaMarca's theory and results.
+The cache misses end up being the deciding factor between the performance of
+double tiled mergesort and base mergesort, the next best contender. Dispite the
+reduction in level 2 cache misses, the sharp increase in level 1 misses result
+is poor performance for tiled and multi-mergesort. Multi-mergsort is further
+encumbered with a much higher instruction count. Double tiled mergesort, with
+its low instruction count, and low cache miss rate in both level 1 and two,
+emerges the winner.
 

### 2005-06-16 docs/tech_report/merge.tex
wrote up the software predictor section

@@
 \sixplot{algorithm_n_0.eps}{algorithm_s_0.eps}{base_mergesort_0.eps}{double_tiled_mergesort_0.eps}{double_multi_mergesort_0.eps}{double_multi_mergesort_0.eps}{mergesorts}
-%\afterpage{
-%\setcounter{subfigure}{0}
-%\thispagestyle{empty}
-%\clearpage
-%\enlargethispage{14em}
-%\vspace*{-5em}
-%\begin{figure}[H]
-%\begin{changemargin}
-%\subfigure[]{\includegraphics[scale=0.74]{../counters/algorithm_n-0.eps}}
-%\subfigure[]{\includegraphics[scale=0.74]{../counters/algorithm_s-0.eps}}
-%\subfigure[]{\includegraphics[scale=0.74]{../counters/base_mergesort_0.eps}}
-%\subfigure[]{\includegraphics[scale=0.74]{../counters/double_tiled_mergesort_0.eps}}
-%\subfigure[]{\includegraphics[scale=0.74]{../counters/double_multi_mergesort_0.eps}}
-%\subfigure[]{\includegraphics[scale=0.74]{../counters/double_multi_mergesort-1.eps}}
-%\end{changemargin}
-%\vspace*{6em}
-%\caption{Branch prediction performance for mergesorts. All graphs use the same scale.}
-%\label{Branch prediction performance for mergesorts}
-%\end{figure}
-%\newpage
-%}
 
-%TODO This is a bit to rewrite
+Figure \ref{mergesorts-1} shows that algorithm N has the same number of branches
+and branch mispredictions as algorithm S, shown in \ref{mergesorts-2}. Base
+mergesort has the same number of mispredictions (see Figure \ref{mergesorts-3},
+indicating a possible minimum number. It has a lower number of branches,
+probably due to performing a persort first. The presort has approximately 6.8
+misses per key (not shown in the graphs). In later mergesorts, an insertion sort
+is used instead.
+%TODO ohw much better does this do? if not, move to other section
 
-figure a shows Algorithm N, knuths natural merge. N3, N5 and N9 indicate parts of the flow chart. 'N3 i==j' is the second branch in N3.
+Double-tiled mergesort (which has the same results are tiled mergesort) has a
+predilication for left and for merging forwards. This is likely to be due to the
+fact that the final merge goes forwards, and that left occurs first in the code.
 
-The results show that the secondary loops are always taken. Taken here means broken out of, hence seldom used. The main action occurs in N3, the main comparator, which has a 50\% prediction ratio.
+In both double tiled mergesort and base mergesort, approximately two thirds are
+of the loops are taken, unlike Knuth's sorts. However, this appears to indicate
+an increase in branches, as opposed to a decrease in mispredictions. Overall
+this stay is line with the idea that there is a minimum number of branch misses,
+below which comparative sorts will not drop.
 
-Algorithm S in graph B shows the same. I never really understood how it worked though.
+Double multi-mergesort sorts the arrays fully in the cache, leaving them
+bitonically arranged. This means that there is no bias for the forward branches,
+as there was in tiled mergesort. This is shown in Figures \ref{mergesorts-5} and
+\ref{mergesorts-6}. While merging the number of branches is 20\% lower than base
+mergesort, while the number of mispredictions in only 10\% lower. This implies
+that most of the mispredictions occur early in the merging process, and not in
+the final merge, which multi-mergesort leaves out.
 
-Figure c shows base mergesort.
-This shows a lower number of branches and a higher percentage of correct predictions.
-The lower number of branches is probably due to the presort.
-there is a higher percentage of correct predictions, but the actual number of correct predictions in the same.
-This implies that doing the presort removes incorrect predictions (or, more to the point, moves incorrect predictions to another part of the algorithm).
-Note too, that both sides are used equally, which doesn't happen later.
-Two thirds of the branches are not-taken, which means loops execute twice before breaking.
-this is the same results that we found in quicksort later.
-I'm sure knuth explains this somewhere in the fine print.
-
-Figure d shows double tiled mergesort.
-The total number fo branches is a 10\% increase over the base mergesort.
-Not sure why this is.
-15\% increase in correct predictions so this incrase into predictable range, which should be avoidable (according to what we guess in quicksort).
-
-Figure e and f show the double mulitmergesort results.
-This sorts fully within the cache only, reducing the number of branches relating to this.
-The reason that the previous balance was so muhc in favour of forwards was that the final sort at the end was forwards.
-Here though, the arrays are left bitonically, so there is an even distribution betwwen forwards and backwards.
-There si still the bias towards left rather than right.
-Left comes first in the code, but this doesnt quite explain why this happens.
-However, the number fo misses each way is identical.
-
-The total branches is a decrease of about 20\% on base mergesort.
-The number of mispredictions decreases by less than 10\%.
-Howver, all these branches are made up by the increase in branches due to the heap.
-The number of extra brnahc misses due to this is 26\% of the branches misses in base mergesort.
-In total, the num ber fo branches isa equivalent however.
-So no loss by doing this (except lots of extra instructions, as seen before).
-
-Note that here and in the heaps, comparison 0 does nothing. THe compiler ought to trim it out, except that now it actually does something (increases the branch counter).
-
-finally, in base mergesort, there is an inline presort used.
-For efficiency, an insertion sort is used later.
-However, there are 6.8 misses per key in the inline presort, which is less than the insertion sort does.
-Its not as useful for presorting to 8, however.
-THese results hold forwards and in reverse.
-
-%TODO up to here
+Overall, when the k-ary heap is used, there are the same number of branch
+misses. This means that using the technique to reduce cache misses does not
+adversly affect branch prediction rates (it does, however, still affect the
+instruction count).  The rate of branch misses is much the same as that in
+heapsort, and progresses in the same manner (see Figure \ref{cache heapsort
+curve}). Note that \n{comparison 0} occurs every time, and would be optimized
+out by the compiler.
 
@@
 cache blocks, without the need for padding.
-
-%TODO
-TODO what about level 1 cache misses if we realigned by 1024?

### 2005-06-16 docs/tech_report/merge.tex
did a spell check on mergesort fixed the left and forward thing in mergesort the first few fixed in quicksort

@@
 of level 2 cache misses. However, it does this at the expense of increasing the
-number of level 1 cache misses. To avoid this, we furthur offset the target
+number of level 1 cache misses. To avoid this, we further offset the target
 array by $Level1CacheSize/2$. It is expected that the number of level 1 misses
@@
 smallest unused key from each list. When there is a small number of lists, this
-isn't a problem, however, when the number of lists increases significanty, the
+isn't a problem, however, when the number of lists increases significantly, the
 overhead involved in this is substantial. It is necessary to check each of the
@@
 tiling, as shown in Figure \ref{merge level 1 misses}. Double multi-mergesort
-has consistantly four times fewer level 1 misses than multi-mergesort, and both
-double tiled sorts perform significantly better than the unoptimizied sorts.
+has consistently four times fewer level 1 misses than multi-mergesort, and both
+double tiled sorts perform significantly better than the unoptimized sorts.
 
-The level 2 cache results are startingly similar to LaMarca's. The shape of the
+The level 2 cache results are startlingly similar to LaMarca's. The shape of the
 Figure \ref{merge level 2 misses} is exactly that provided by LaMarca. The
 ratio's between the each of the sorts is also the same, though the numbers are
-not identical. As we use a different base algorithm from LaMarca, this is not
-unthinkable.
+not identical, due to different \c{int} sizes in our tests.
 
@@
 The cache misses end up being the deciding factor between the performance of
-double tiled mergesort and base mergesort, the next best contender. Dispite the
+double tiled mergesort and base mergesort, the next best contender. Despite the
 reduction in level 2 cache misses, the sharp increase in level 1 misses result
-is poor performance for tiled and multi-mergesort. Multi-mergsort is further
+is poor performance for tiled and multi-mergesort. Multi-mergesort is further
 encumbered with a much higher instruction count. Double tiled mergesort, with
-its low instruction count, and low cache miss rate in both level 1 and two,
+its low instruction count, and low cache miss rate in both levels 1 and 2,
 emerges the winner.
@@
 indicating a possible minimum number. It has a lower number of branches,
-probably due to performing a persort first. The presort has approximately 6.8
+probably due to performing a presort first. The presort has approximately 6.8
 misses per key (not shown in the graphs). In later mergesorts, an insertion sort
@@
 Double-tiled mergesort (which has the same results are tiled mergesort) has a
-predilication for left and for merging forwards. This is likely to be due to the
-fact that the final merge goes forwards, and that left occurs first in the code.
+predilection for merging forwards. This is likely to be due to the fact that the
+final merge goes forwards. The left merge step is also favoured, but this is
+merely due to the fact that it occurs first in the code. While there are more
+left branches, the actual number of branches taken are the same.
 
@@
 misses. This means that using the technique to reduce cache misses does not
-adversly affect branch prediction rates (it does, however, still affect the
+adversely affect branch prediction rates (it does, however, still affect the
 instruction count).  The rate of branch misses is much the same as that in

### 2005-06-16 docs/tech_report/quick.tex
did a spell check on mergesort fixed the left and forward thing in mergesort the first few fixed in quicksort

@@
 smallest of the first \cc{threshold} keys is used as a sentinel, in keeping
-with the policy of not exceeding the bounds of the array.
+with our policy of not exceeding the bounds of the array.
 
@@
 cache misses, and the results would overwhelm the improvements LaMarca makes in
-the next phase of the algorithm. Secondly, it is not as important to write a
-brilliant version of quicksort as it is to analyse LaMarca's improvements.
+the next phase of the algorithm. Secondly, when sorting random keys over a
+large range, the number of equal keys will be vewry low. It is therefore not
+worth our while to investigate this.
 

### 2005-06-16 docs/tech_report/quick.tex
added sum counters fixed minor erros in quick.tex dont know whats wrong with processed_cycles_data dont know the rst either

@@
 \label{quick}
+
 Quicksort is an in-place, unstable sort. It was first described in
@@
 \section{Memory-Tuned Quicksort}
+
 LaMarca's first step was to increase the temporal locality of the algorithm by
@@
 done with one iteration of insertion sort, and would have just one branch miss.
-However, the number of level 2 cache misses would be great in this case.
-However, the cost of this would be substantial.  This could be reduced by
+However, the number of level 2 cache misses would be great in this case,
+and the cost of this would be substantial. This could be reduced by
 choosing the key to remove from a small list, such as the eight keys at the
@@
 subset size to be $C/3$, where $C$ is the number of keys which fit in the
-cache\footnote{For a 2MB cache, this is 524,144.}.
+cache.
 
@@
 \subsection{Test Parameters}
+
 The threshold chosen was 10, simply as this is the number used in Sedgewick's
@@
 
-A 2MB cache was assumed, so the number of keys which fit in the cache for this
-in-place sort is 524,288. Multi-quicksort began once the number of keys was
-larger than 174,762.
+A 2MB cache was assumed, so the number of 32 bit integer keys which fit in the
+cache for this in-place sort is 524,288. Multi-quicksort began once the number
+of keys was larger than 174,762.
 
 \subsection{Expected Performance}
-The instruction count for each quicksort is expected to be $O(NlogN)$,
-though not as high as the other algorithms due to the efficient inner loop.
-The number of instructions for multi-quicksort should increase by about 50\% in
-the case of the sequential sort, and 25\% in the case of the binary sort.
+
+The instruction count for each quicksort is expected to be $O(NlogN)$, though
+not as high as the other algorithms due to the efficient inner loop.  The number
+of instructions for multi-quicksort should increase by about 50\% in the case of
+the sequential sort, and 25\% in the case of the binary sort.
 
@@
 The number of branches should increase in multi-quicksort along with the
-instruction count. When a binary search is used, the increase in misses is
-expected to be significant. With a sequential search, however, the number of
+instruction count. When a binary search is used, the increase in mispredictions
+is expected to be significant. With a sequential search, however, the number of
 misses should not increase at all, and should flat-line from that point on.
@@
 
-A previous result, however, with a different implementation showed the same
-shape, but different results to LaMarca's. That implementation waited until all
-the partitioned keys were put back into the array, instead of being sorted
-straight away. Those results plateaued once mutli-quicksort took effect.
+\section{Predictor results}
 
-\section{Predictor results}
 \sixplot{base_quicksort_0.eps}{base_quicksort9_0.eps}{cache_quicksort_0.eps}{multi_quicksort_0.eps}{sequential_multi_quicksort_0.eps}{}{base, cache and multi quicksorts and base quicksort9}
+
 \sixplot{base_quicksort1_0.eps}{base_quicksort_0.eps}{base_quicksort5_0.eps}{base_quicksort7_0.eps}{base_quicksort9_0.eps}{}{base quicksort without a median, with median of 3, and with pseudo median of 5, 7 and 9}

### 2005-06-16 code/sorts/on2_bubblesort.c

@@ -6,12 +6,16 @@ on2_bubblesort(unsigned int a[], int N)
 	int i, j;
 	int sorted;
 
+#ifdef _USE_SOFTWARE_PREDICTOR
 	char string[10];
+#endif
 
 	for (i = 0; i < N-1; i++)
 	{
+#ifdef _USE_SOFTWARE_PREDICTOR
 		sprintf(string, "%d", i);
 		describe_predictor(&global_predictor[i], string);
+#endif
 
 		sorted = 1;
 		for (j = N-1; j > i; j--)

### 2005-06-16 code/sorts/on2_bubblesort2.c

@@ -5,16 +5,26 @@ on2_bubblesort2(unsigned int a[], int N)
 {
 	int i = 0, j, k = 0;
 
+#ifdef _USE_SOFTWARE_PREDICTOR
+	char string[10];
+#endif
+
 	while(i < N-1)
 	{
+#ifdef _USE_SOFTWARE_PREDICTOR
+		sprintf(string, "%d", i);
+		describe_predictor(&global_predictor[i], string);
+#endif
 		k = N-1;
 		for (j = N-1; j > i; j--)
 		{
 			if (less(a[j],a[j-1]))
 			{
+				branch_taken(&global_predictor[i]);
 				k = j;
 				exch(a[j-1],a[j]);
 			}
+			branch_not_taken(&global_predictor[i]);
 		}
 		i = k;
 	}

### 2005-06-23 docs/tech_report/background.tex
fixed some TODOs

--- 2005-05-31 paper/docs/background.tex
@@
 later chapters.
-%TODO what's shellsort in Big O notation
-TODO what's shellsort in Big O notation
+
+Not all algorithms have an easily derivable complexity. Many studies have
+attempted to tie down the complexity of shellsort, and it is not easily
+definable in terms of the catageories above (see \ref{TODO}).
 
 An important thing to note is that the time the algorithms take to execute are
-proportion to their Order, but different constants of proportionality are used
-for different algorithms. An algorithm with $10N$ instructions is $O(N)$, and is
-not differentiated from an algorithm with $1000N$ instructions by this
+proportion to their complexity, but different constants of proportionality are
+used for different algorithms. An algorithm with $10N$ instructions is $O(N)$,
+and is not differentiated from an algorithm with $1000N$ instructions by this
 notation.

### 2005-06-23 docs/tech_report/intro.tex
fixed some TODOs

--- 2005-06-14 docs/tech_report/docs/intro.tex
@@
 
-Chapters \ref{heap} through \ref{quick} discuss base algorithms and cache-conscious improvements
+Chapters \ref{heap} through \ref{radix} discuss base algorithms and cache-conscious improvements
 for heapsort, mergesort and quicksort. Results are presented and explained, and
@@
 
-Chapter \ref{radix} presents a radixsort implementation and discusses its performance in
-the areas of instruction count, cache-consciousness and branch prediction.
-
-%TODO shell
-TODO Chapter \ref{shell}
+Chapter \ref{shell} presents two shellsort implementations and discusses their
+performance in the areas of instruction count, cache-consciousness and branch
+prediction.
 

### 2005-06-24 docs/tech_report/heap.tex
made the thing compile now
managed to compile the docs, without the 16 byte cache line

--- 2005-05-31 paper/docs/heap.tex
@@
 LaMarca used 64 bit ints, and while this doesn't affect results, it means that
-fewer keys fit into his cache lines. As a result, he used a 4-heap. To try and
-emulate his results, extra tests were done using a 4-heap with a 32 byte
-cache line, and using a 4-heap using a 16 byte cache line, so that the entire
-cache line is used. \verb!UINT_MAX! was used as the sentinel.
+fewer keys fit into his cache lines. Since modern cache lines are typically
+longer than 4 bytes, we stick with 8-byte cache lines. \verb!UINT_MAX! was used
+as the sentinel.
 
@@
 using each cache line, taking advantage of the spatial locality of the cache
-line. However, the 4-heap should perform just as well with a 16 byte cache line.
+line.
 
@@
 
-\plot{heap}{}{}{Cache heapsort (4-heap) was also tested with a 16 byte cache line}{Cache heapsort (4-heap) was also tested with a 16 byte cache line}{}{}
+\plot{heap}{}{}{}{}{}{}
 
@@
 
-\sixplot{base_heapsort-0.eps}{cache4_heapsort-0.eps}{cache_heapsort-0.eps}{cache_heapsort-1.eps}{}{}{base and cache heapsorts}
+\sixplot{base_heapsort_0.eps}{cache4_heapsort_0.eps}{cache_heapsort_0.eps}{cache_heapsort_1.eps}{}{}{base and cache heapsorts}
 
@@
 \begin{figure}
-\includegraphics{../counters/curve.eps}
+\includegraphics{plots/curve.eps}
 \caption{Predictability of branches in cache heapsort, with a fanout of eight}
@@
 chapter.
-
-Comparing the 4-heap with a 16 byte cache line to the 4-heap with a 32 byte
-cache line is not as interesting as it would be to compare it to an 8-heap on a
-16 byte cache line.
-
-Also, it would be interesting to use performance counters to find out whether
-the cache or the branch predictors are responsible for the base heapsort being
-faster than the 4-heap, on the Pentium 4.

### 2005-07-12 docs/tech_report/result_list.tex
removed word gonnet from document
managed to compile the docs, without the 16 byte cache line
updated gnuplot scripts, added scripts for newer sorts updated docs to inculde new sorts in the result list fixed bw

@@
 \smallplot{on2_bubblesort}{bubblesort}{}{}{}{}{}
-\smallplot{on2_bubblesort2}{bubblesort2}{}{}{}{}{}
+\smallplot{on2_bubblesort2}{improved bubblesort}{}{}{}{}{}
 \smallplot{on2_shakersort}{shakersort}{}{}{}{}{}
-\smallplot{on2_shakersort2}{shakersort2}{}{}{}{}{}
+\smallplot{on2_shakersort2}{improved shakersort}{}{}{}{}{}
 %
@@
 \smallplot{cache4_heapsort}{cache heapsort (4-heap)}{}{}{}{}{}
-\smallplot{cache4_heapsort_small_cache_line}{cache heapsort (4-heap) (with 16-byte cache line)}{}{}{}{}{}
 \smallplot{cache_heapsort}{cache heapsort (8-heap)}{}{}{}{}{}
@@
 \smallplot{base_mergesort}{base mergesort}{}{}{}{}{}
+\smallplot{tiled_mergesort}{tiled mergesort}{}{}{}{}{}
+\smallplot{multi_mergesort}{multi-mergesort}{}{}{}{}{}
 \smallplot{double_tiled_mergesort}{double tiled mergesort}{}{}{}{}{}
@@
 %
-\smallplot{base_quicksort}{base quicksort}{}{}{}{}{}
+\smallplot{base_quicksort1}{base quicksort (no median)}{}{}{}{}{}
+\smallplot{base_quicksort}{base quicksort (median-of-3)}{}{}{}{}{}
+\smallplot{base_quicksort5}{base quicksort (pseudo-median-of-5)}{}{}{}{}{}
+\smallplot{base_quicksort7}{base quicksort (pseudo-median-of-7)}{}{}{}{}{}
+\smallplot{base_quicksort9}{base quicksort (pseudo-median-of-9)}{}{}{}{}{}
 \smallplot{cache_quicksort}{memory-tuned quicksort}{}{}{}{}{}
@@
 \smallplot{cache_radixsort}{cache radixsort}{}{}{}{}{}
+\smallplot{aligned_cache_radixsort}{aligned cache radixsort}{}{}{}{}{}
 %
-\smallplot{gonnet_shellsort}{gonnet shellsort}{}{}{}{}{}
-\smallplot{improved_gonnet_shellsort}{improved gonnet shellsort}{}{}{}{}{}
+\smallplot{shellsort}{shellsort}{}{}{}{}{}
+\smallplot{improved_shellsort}{improved shellsort}{}{}{}{}{}
 }

### 2005-07-16 docs/tech_report/quick.tex
fixed the quicksort predictor scripts reconsidered the special directories some changes to main (some bad, but its late and i dont care)
finsihed move, but not rename

@@
 the next phase of the algorithm. Secondly, when sorting random keys over a
-large range, the number of equal keys will be vewry low. It is therefore not
-worth our while to investigate this.
+large range, the number of equal keys will be vewry low. Using a median of
+greater than three is discussed in section \ref{quick-predictors2}.
 
@@
 graph that this is true in the case of all the quicksorts.
+%TODO why is this? the only reason I've come up with is that in the direct
+%ampped cache, the same area is thrashed, while keeping keys that we;re going to
+%reuse is a separte bit. in the fully assoc cache, those keys get washed out by
+%the keys that arent going to be reused.
 
@@
 as expected. The binary search produced a dramatic increase in the branch
-misses, though not when the two-level adaptive predictor was used. A short spike
-was recorded in the bimodal results, and its not clear if this is an anomaly or
-not. If it is, then the two-level adaptive predictor performs worse than the
-bimodal, yet again.
+misses, when using a bimodal predictor. When using a 2-level adaptive predictor,
+however, the number of misses is, in general, slightly higher, though it does
+not spike due to the binary search. While this performs worse than the bimodal
+predictor, it appears to show that 2-level adaptive predictors can adapt to the
+erratic branches of a binary search.
 
 The sequential search stops the increase of branch misses, since every key will
-only take one miss to find its proper place. The results of this on cycle count,
-shown in Figure \ref{quick cycles} are enough to make up for the increase in
-instruction count, which was significantly reduced by using the binary search.
-The tuned quicksort, however, had the lower running time and even base
+only take one miss to find its proper place. The results of this on cycle
+count, shown in Figure \ref{quick cycles} are enough to make up for the increase
+in instruction count, which was significantly reduced by using the binary
+search.  The tuned quicksort, however, had the lower running time and even base
 quicksort out-ran the multi-quicksorts.
@@
 cache misses plateau, while the results here show an increasing slope.
-
+%TODO is a pseudo-median explained
 \section{Predictor results}
 
+\subsection{Different sorts}
+
 \sixplot{base_quicksort_0.eps}{base_quicksort9_0.eps}{cache_quicksort_0.eps}{multi_quicksort_0.eps}{sequential_multi_quicksort_0.eps}{}{base, cache and multi quicksorts and base quicksort9}
-
-\sixplot{base_quicksort1_0.eps}{base_quicksort_0.eps}{base_quicksort5_0.eps}{base_quicksort7_0.eps}{base_quicksort9_0.eps}{}{base quicksort without a median, with median of 3, and with pseudo median of 5, 7 and 9}
-
-%TODO a bit to rewrite
-
-Theres two important bits: comparison between the types of sorts, and comparisons between the different number of things tried as medians.
 
@@
 
-Part 2: comparing different ways of randomly partitioning:
+\subsection{Different medians}
 
-Ways of getting the partition:
-1 - just take the rightmost element. Note that there are no sentinels, so you have to put a bounds check back in
-3 - median of(0,1,2) halves
-5 - median of(0, (median of(1,2,3), 4) quarters
-7 - median of(median of(0,1,2), 3, median of(4,5,6)) sixths
-9 - median of(median of(0,1,2), median of(3,4,5), median of(6,7,8)) eights
+\label{quick-predictors2}
+\sixplot{base_quicksort1_0.eps}{base_quicksort_0.eps}{base_quicksort5_0.eps}{base_quicksort7_0.eps}{base_quicksort9_0.eps}{}{base quicksort without a median, with median of 3, and with pseudo medians of 5, 7 and 9}
 
-misprediction rates:
-1 - 29.4\%
-3 - 35.9\%
-5 - 38.4\%
-7 - 40.0\%
-9 - 41.7\%
+%TODO a bit to rewrite
+\label{quicksort entropy}
+
+We have observed before (see Section \ref{heapsort entropy}), that the more
+branches an algorithm has, the more of them will be correctly predicted. we
+observe that using a better pivot reduces the number of exchanges made by
+quicksort. We expect to find, however, that using a better pivot reduces the
+number of correct predictions, and not the number of incorrect predictions.
+
+Each of the \code{while} loops in \code{partition} is a branch, and each
+iteration of that loop corresponds to that branch being taken, while each
+exchange corresponds to that branch being not-taken. So the fewer
+
+%TODO to whom do we owe the idea of a pseudo median? Bently/McIlroy?
+%TODO when does it cut out, and why did we choose tere (benlty?)
+To test this theory, we create variations of quicksort using different medians.
+We tested no median (or median-of-one), median-of-three, and pseudo-medians-of
+-five, -seven and -nine. A pseudo-median-of-five is achieved by choosing five
+keys, evenly spaces across the array, including the first and last key. The
+median of the middle three keys is determined, and the pivot is chosen as the
+median of the chosen key, the first key and the last key. The median-of-seven is chosen
+similarly, using seven evenly spaced keys, and finding the median of the first
+three and the median of the last three, and choosing as the pivot the median of
+these keys and the middle key. The median of nine determines the median of three
+median-of-threes, involving the first, middle and last three keys from nine
+evenly spaced keys.
+
+We measured the number of predictions at the same points as before, aswell as
+when the predictors were being calculated, expecting that a reduction in misses
+in the partitioning due to using a better median would be made up for by an
+increase in number of mispredictions while finding the median. The results are
+shown in Figure \ref{Branch prediction performance for base quicksort without a
+median, with median of 3, and with pseudo medians of 5, 7 and 9}.
+
+Picking a better pivot reduces the number of brnahces, but doesnt reduce th
+enumber of mispredictions, in fact, it increases them.
 

### 2005-07-18 docs/tech_report/quick.tex
updated the scripts in the docs directory. removed a debug line. updated the data file. made small fixes to quick.tex

@@
 
-Each of the \code{while} loops in \code{partition} is a branch, and each
+Each of the \cc{while} loops in \cc{partition} is a branch, and each
 iteration of that loop corresponds to that branch being taken, while each

### 2005-07-18 docs/tech_report/quick.tex
finished the different medians sction ofd quicksort branch predictors

@@
 
+%TODO a bit to rewrite
 \subsection{Different sorts}
@@
 
+%TO here
+
 \subsection{Different medians}
@@
 
-%TODO a bit to rewrite
 \label{quicksort entropy}
 
-We have observed before (see Section \ref{heapsort entropy}), that the more
-branches an algorithm has, the more of them will be correctly predicted. we
-observe that using a better pivot reduces the number of exchanges made by
-quicksort. We expect to find, however, that using a better pivot reduces the
-number of correct predictions, and not the number of incorrect predictions.
+In Section \ref{heapsort entropy}, we observed that when an algorithm has more
+branches than are strictly necessary complete, these extra branches are
+predicted correctly. In quicksort, we expect to obtain similar results.
 
-Each of the \cc{while} loops in \cc{partition} is a branch, and each
-iteration of that loop corresponds to that branch being taken, while each
-exchange corresponds to that branch being not-taken. So the fewer
+%HOW
 
+We use several variations of quicksort to test this theory. \ref{bently, i
+ithkn} suggests partitioning using a median-of-9 instead of a median-of-three. We
+created sorts using no median, a median-of-3, and pseudo-medians-of-five, -seven and
+-nine. The pseudo-medians used repeated medians-of-three instead of a real
+median, as suggestion by \ref{bently, again, i think}. 
+%use a cut off
+
+We measured the number of branches and mispredictions due to the partitioning,
+the final insertion sort, and the partitioning. It is expected to have little or
+no variance in either the number of branches and misses due to insertion sort,
+but that less time will be spent partitioning to arrays as a better median is
+found, and a corresponsing increase in time spent finding the better median.
+
+The results of these tests are shown in Figure \ref{Branch prediction
+performance for base quicksort without a median, with median of 3, and with
+pseudo medians of 5, 7 and 9}. There is little variance in insertion sort misses
+between the sorts using pseudo-medians, but the median-of-three quicksort has
+6\% fewer branches, and a corresponsing decrease in mispredictions.
 %TODO to whom do we owe the idea of a pseudo median? Bently/McIlroy?
 %TODO when does it cut out, and why did we choose tere (benlty?)
-To test this theory, we create variations of quicksort using different medians.
-We tested no median (or median-of-one), median-of-three, and pseudo-medians-of
--five, -seven and -nine. A pseudo-median-of-five is achieved by choosing five
-keys, evenly spaces across the array, including the first and last key. The
-median of the middle three keys is determined, and the pivot is chosen as the
-median of the chosen key, the first key and the last key. The median-of-seven is chosen
-similarly, using seven evenly spaced keys, and finding the median of the first
-three and the median of the last three, and choosing as the pivot the median of
-these keys and the middle key. The median of nine determines the median of three
-median-of-threes, involving the first, middle and last three keys from nine
-evenly spaced keys.
 
-We measured the number of predictions at the same points as before, aswell as
-when the predictors were being calculated, expecting that a reduction in misses
-in the partitioning due to using a better median would be made up for by an
-increase in number of mispredictions while finding the median. The results are
-shown in Figure \ref{Branch prediction performance for base quicksort without a
-median, with median of 3, and with pseudo medians of 5, 7 and 9}.
+Between a median-of-three and a pseudo-median-of-nine partitioning, there is a
+10\% reduction in branches. At the same time, there is a 20\% reduction in
+correct predictions, accounting for the entire reduction in branhces, and a 4\%
+increase in mispredictions. This is due to better partitioning. Less time is
+spent copying keys unnecessarily, resulting in a less predictable branch. The
+corresponding increase in mispredictions while calculating the median equates to
+50\% of the extra mispredictions due to partitioning.
 
-Picking a better pivot reduces the number of brnahces, but doesnt reduce th
-enumber of mispredictions, in fact, it increases them.
+This indicates the same pattern as we saw during heapsort.%
+% and mergesort?
+The less entropy due to the partitioning key, the less wasted exchanging of keys,
+which leads to a general decrease in branches. This decrease means that the
+fewer keys are exchanged due to partitioning, leading to less predictable
+partitioning. In effect, there are fewer branches pretending to be useful.
 
-The misprediction rates point to the problem.
-They're just not being evenly divided.
-The number of times the partitions end is the same, roughly, each time.
-However, the number of actual branches is much higher in the case of quicksort1, and decreases all the way down.
-At the same time, the number of mispredictions creeps up.
-And hence the big change in the rate of mispredictions.
-As predicted, the better the median, the better the partition, and the less correct predictions pretending to be useful.
-There must be a 50\% rate waiting to get out, so a little more than median of 9 maybe.
-Or a lower cut-off point (it cuts out for partitions of less than 40 keys, and uses plain old median of 3).
-
-the rate or number of insertion branches doesnt change at all.
-
-%TODO up to here
+The pseudo-median-of-five partitioning performs slightly better than the
+median-of-three, and pseudo-median-of-seven performs slightly better again, in
+terms of branch prediction. Figure \ref{TODO} shows the cycle count of each of
+the quicksorts, showing that a pseduo-median-of-seven offers the greatest
+tradeoff between the reduced number of branches and instructions, and the
+increase in branch midpredictions.
 

### 2005-07-19 docs/tech_report/radix.tex
added aligned cache radix sort to discussion. Made minor fixes

--- 2005-05-26 paper/docs/radix.tex
@@
-\label{Radixsort}
+\label{radix}
 %\section{Description}
@@
 counters is created to count the number of keys with a particular least
-significant byte. Then the counters are summed in turn to store the cumulative
-count of keys that belong before them in the array. Now each byte can be put in
+significant byte. The counters are then summed in turn to store the cumulative
+count of keys with a smaller least significant byte. Now each key can be put in
 place by accessing the counter indexed by its least significant byte.
 
-These are written into an auxiliary array, which is written back before the next
-iteration using the second least significant byte. Since every key is processed
-in order, when the next most significant byte is sorted, the order of the less
-significant bytes are maintained, so that once the most significant byte is
-sorted, the entire array is.
+These are written into an auxiliary array, which is written back to the initial
+array before the next iteration. The process is then repeated using the second
+least significant byte.  Since every key is processed in order, when the next
+most significant byte is sorted, the order of the less significant bytes are
+maintained, so that once the most significant byte is sorted, the entire array
+is.
 
 This type of radixsort is called \n{Least Significant Digit} radixsort.
-Other implementations are possible, and several improvements also exist, but as
-radixsort is included here for comparative purposes, these are not implemented.
-They are, however, discussed in Section \ref{radix future work}.
+Other implementations are possible, inclduing a \n{Most significant bit}
+version, which behaves as a divide and conquer algortithm. These are discussed
+in Section \ref{radix future work}.
 
@@
 one to copy, and one to write back (which accesses both arrays). LaMarca
-suggests counting for the next step in the previous count. In fact, all the
-counting can be done in a single pass at the start, instead of four passes at
-various stages, decreasing instruction count and cache misses. The downside is
-the fourfold increase in memory required for the arrays of counts, with no
-benefits.
+suggests counting for the next step in the previous count, which we implement.
+
+In fact, all the counting could be done in a single pass at the start, instead
+of four passes at various stages, decreasing instruction count and cache misses.
+The downside is the fourfold increase in memory required for the arrays of
+counts, with no benefits.
 
@@
 to the source array was changed to copying back and forth between the source and
-auxiliary arrays. The same could be potentially done with radixsort, without any
-increase in complexity. The number of steps is likely to be even, and is known
-at compile time, so the number of times the copy step is done could be reduced
-to zero. If the number of steps is odd, which is unlikely when sorting
-integers, then one copy at the end would be required. This is not needed in the
-implementation presented here.
+auxiliary arrays. The same could be done with radixsort, without any increase in
+complexity. The number of steps is likely to be even, and is known at compile
+time, so the number of times the copy step is done could be reduced to zero. If
+the number of steps is odd, which is unlikely when sorting integers, then one
+copy at the end would be required. This is not needed in the implementation
+presented here.
 
@@
 and a final two as it copies using the most significant byte.
+
+\subsection{Aligned Cache Radixsort}
+In Section \ref{tiled mergesort}, mergesort is improved by aligning its array
+with reference to it auxilliary array. Specifically, extra memory is allocated
+for the auxilliary array, which is then offset so that the start of the two
+arrays are exactly half the length of the cache apart. We further added an
+alignment for the level 1 cache to significantly reduce level 1 misses at the
+expense of a small number of level 2  misses. We adapt this strategy here, and
+call it Aligned Cache Radixsort.
 
@@
 small as 4096 records, it involves 12 times fewer steps than the $O(NlogN)$
-sorts.
+sorts. For an array as large as 4194304, an $(O(NlogN)$ sort will take 22 times
+as many steps to complete. However, it is not expected that radixsort will
+perform many times better than quicksort, the reigning champion. A `step' for
+quicksort is as simple as a comparison and a possible exchange every few steps.
+For radixsort, a step comprises being counted four times and copied four times,
+with the counting involving several layers of indirection, and causing a
+complete iteration across the array at the start. LaMarca's results
+show the instruction count of quicksort and radixsort being in the same order of
+magnitude, and we expect the same.
 
 The cache performance of base radixsort is not expected to be good, however.
-Quicksort and mergesort, for example, have very good temporal locality. Keys are
+Quicksort and mergesort have very good temporal locality. Keys are
 used repeatedly before being swapped out of the cache. In radixsort, however,
@@
 can result in thrashing.
+
+The thrashing can also occur in aligned radixsort. Unlike mergesort, we cannot
+guarantee that the keys being considered are at the same positions in both
+arrays. In fact, we have no idea where is being accessed in the source array, so
+it is unlikely that the alignment will have any effect what-so-ever.
 
@@
 in exceptional circumstances\footnote{See Section \ref{insertion is
-predictable}.}. As a result, only flow control branches occur in radixsort, and
-they are all predictable.
+predictable}.}. Radixsort has no comparative branches at all. As a result, only
+flow control branches occur in radixsort, and they are all predictable.
 
@@
 \label{radix future work}
-Combining other sorts with radixsort could also be possible. An optimisation to
+Combining other sorts with radixsort could also be possible. An optimization to
 mergesort was to do a k-way merge at the end, but this dramatically increased

### 2005-07-19 docs/tech_report/radix.tex
fixed the key position in a graph rewrote some of the results in radixosrt

@@
 the array no longer fits into the cache. The number of misses then leaps up to
-2.5 per key, which is roughly what was predicted.
+2.5 per key, which is slightly more than was predicted. The extra is likely due
+to the unpredictable way that the source array is accessed as the keys are
+copied to the auxilliary array.
 
@@
 has been reduced by more than 10\%, and the number of level 2 cache misses
-have been more than halved, to just over one miss per key.
+have been more than halved, to one and a quarter miss per key. This is exactly
+what is expected, due to there being nine passes over the array, and a one miss
+per pass per cache line.
+
+The number of misses while the array fits in the cache is also obvious: there
+are two compulsory misses per key - one for each array - which divided by the
+keys per cache line results in exactly one miss in four.
+
+Aligned cache radixsort performs exactly the same as cache radixsort, indicating
+that the alignment, and the potentially large memory overhead, are not
+necessary.
 
 The number of branch misses, shown in Figure \ref{radix branch misses} was also
-extremely low. Despite 10 branches per key, there was less than .08 of a branch
+extremely low. Despite 2 branches per key, there was less than .08 of a branch
 miss per key. The number of branches itself is quite low, as several of the
@@
 the results, shown in Figure \ref{radix cycles}, indicate that radixsort will
-never increase beyond 350 cycles per key, meaning that for larger sets it will
-perform significantly better than quicksort. Cache radixsort never increases
-beyond 300 cycles, though it has relatively more cycles when the arrays are
+not increase much beyond 350 cycles per key, meaning that for larger sets it will
+perform significantly better than quicksort. Cache radixsort does not reach 
+250 cycles per key, though it has relatively more cycles when the arrays are
 smaller.
-
-The most curious result is that all simulations show that cache radixsort
-performing better than the base radixsort, yet the Pentium 4 results show that
-base radixsort performs better for smaller arrays.
 
@@
 jumps to around 2.5, where they stay. 
-
+%TODO rewrite this bit a little
+%TODO cache radixsort is faster than quicksort in all cases
 The cycles per key in both cases are unsurprising based on instruction count and
@@
 \label{radix future work}
-Combining other sorts with radixsort could also be possible. An optimization to
-mergesort was to do a k-way merge at the end, but this dramatically increased
-the instruction count. Using a type of radixsort here would be complicated, but
-may be possible. Alternatively, sorting inside the cache using radixsort is far
-faster than merging. The k-way merge could then be performed at the end. In
-practice, though, radixsort is far faster than mergesort, has a lower
-instruction count, and less cache misses. While these steps could improve
-mergesort, they are more likely to slow radixsort, with no advantage.
-
-Radixsort could also be faster than multi-quicksort's initial step. Replacing
-this with a single radix step would split the array into 256 containers, each of
-which could be quicksorted. Alternatively, a smaller radix, based on the size of
-the array, could be generated at run time. Using LaMarca's analysis,
-$C/3$ containers need to be generated. With $N$ as the number of keys in the
-array, and $C$ is the number of keys which fit in the level 2 cache, the radix
-can be calculated with the formula:
+Combining other sorts with radixsort could be possible.  Radixsort should be be
+faster than multi-quicksort's initial step. Replacing this with a single radix
+step would split the array into 256 containers, each of which could be
+quicksorted. Alternatively, a smaller radix, based on the size of the array,
+could be generated at run time. Using LaMarca's analysis, $C/3$ containers need
+to be generated. With $N$ as the number of keys in the array, and $C$ is the
+number of keys which fit in the level 2 cache, the radix can be calculated with
+the formula:
 $$2^{\lceil{}log_2(\frac{3N}{CacheSize})\rceil{}}$$

### 2005-07-24 docs/tech_report/shell.tex
made the thing compile now
removed word gonnet from document
updated heap scripts to remove 16 byte heapsort updated shell to remove reference to gonnet_ removed all as its done by makefile now
added some heading to conclusings
removed the word gonnet from code
finished up shellsort

--- 2005-05-26 paper/docs/shell.tex
@@
-\label{Shellsort}
+\label{shell}
 
@@
 respectively.}.
-%TODO whats partial sort?
 
-Shellsort begins by considering the array to be made up of $N/2$ subarrays of
+Shell begins by considering the array to be made up of $N/2$ subarrays of
 length two, where the elements of the subarray are spaced $N/2$ keys apart.
@@
 has the increments $N/2$, $N/4$, $N/8$, $N/16$ ........ $1$. \cite{Sedgewick96}
-discusses many increments used, and provides code for comparing these. Based
-on running the \textit{driver} program\footnote{see
+discusses many increments used, and provides code for comparing these. Based on
+running the \cc{driver} program\footnote{See
 http://www.cs.princeton.edu/\textasciitilde{}rs/shell/}, Gonnet's shellsort was
 seen to be the fastest on the authors local machine. As a result, this was used
-to test the performance of shellsort. This uses increments which are $5/11$ths
-of the previous increment.
-%TODO do we refer to increments from the top or bottom
+to test the performance of shellsort. This uses increments which are
+five-elevenths of the previous increment.
 
 Two versions of Gonnet's sort were used. The first was taken directly from the
-code presented by Sedgewick, and only slightly modified to fit into the framework used.
-The second was modified to reduce the complexity of the loop bounds, and to
-remove a sentinel check in the inner loop. Code for these is shown in Figures
-\ref{Gonnet Shellsort code} and \ref{Improved Gonnet Shellsort code}.
+code presented by Sedgewick, and only slightly modified to fit into the
+framework used.  The second was modified to reduce the complexity of the loop
+bounds, and to remove a sentinel check in the inner loop. Code for
+\cc{shellsort} and \cc{improved\_shellsort} is shown in Figures \ref{Shellsort
+code} and \ref{Improved Shellsort code}.
 
-\code{Gonnet Shellsort code}{old_shellsort.c}
-\code{Improved Gonnet Shellsort code}{shellsort.c}
+\code{Shellsort code}{shellsort.c}
+\code{Improved Shellsort code}{improved_shellsort.c}
 
@@
 \subsection{Test parameters}
-The increments in Gonnet's shellsort are $N*(5/11)$, $N*(5/11)^2$, $N*(5/11)^3$,
-...... $1$.
+The increments in Gonnet's shellsort are $\frac{5}{11}N$, $(\frac{5}{11})^2N$,
+$(\frac{5}{11})^3N$ ... $1$;
 
 \subsection{Expected Performance}
-Shellsort has been much analysed, but very little is known about the complexity
-of the algorithm. The instruction count is expected to be $O(NlogN)$, as this
-has been indicated by previous experient. The cycle count is expected to be of
-the same order of magnitude as the $O(NlogN)$ sorts for the same reason.
- %TODO find some experiements that showed this.
+Shellsort has been much analysed, and it has been shown that it's complexity is
+difficult to prove\footnote{See TODOA, TODOB, TODOC and TODOD.}. The instruction
+count and cycle counts are expected to be similar to $O(NlogN)$, as this has
+been indicated by previous experient.
 
@@
 are a large number of subarrays, as during the iterations with large increments,
-there will be no spatial reuse, and little temporal reuse. Once the increment is
-small enough that each subarray fits within the cache (ie quickly for L2 cache,
-and much more slowly for the L1 cache), then temporal reuse will occur. Once an
-entire subarray fits within an eighth of the size of the cache (eight being the
-number of keys that fit in a cache line), then spatial reuse will occur aswell.
-This reuse will only occur within a single iteration of shellsort. Once the next
-increment begins, nothing in the cache is likely to be valid, unless the entire
-array fits within the cache.
+there will be no spatial reuse, and little temporal reuse, so long as the
+subarray is large than the cache. Once the increment is small enough that each
+subarray fits within the cache (i.e. quickly for L2 cache, and much more slowly
+for the L1 cache), then temporal reuse will occur. Once an entire subarray fits
+within an eighth of the size of the cache (eight being the number of keys
+that fit in a cache line), then spatial reuse will occur aswell.  This reuse
+will only occur within a single iteration; once the increment with the next
+increment begins, most of the contents of the cache will be invalidated, unless
+the entire array fits within the cache.
 
 The branch performance of shellsort should be predictable. There are
-$log_{5/11}N$ iterations, each with $N$ branch misses\footnote{See Section
+$log_{(\frac{11}{5})}N$ iterations, each with $N$ branch misses\footnote{See Section
 \ref{insertion is predictable} for a discussion of the properties of insertion
-sort.}. It is expected, therefore, that there will be $log_{5/11}N$ misses per
+sort.}. It is expected, therefore, that there will be $log_{(\frac{11}{5})}N$ misses per
 key.
@@
 quicksort, though it is more than 50 percent faster than the best of the
-heapsort.
+heapsorts.a
+
+%TODO
+TODO are there log5_11 N misses
+Regenerate data before looking at this
 
@@
 
-\sixplot{gonnet_shellsort-0.eps}{gonnet_improved_shellsort-0.eps}{}{}{}{}{Shellsorts}
+\sixplot{shellsort_0.eps}{improved_shellsort_0.eps}{}{}{}{}{Shellsorts}
 
-%TODO better writing
+In shellsort, almost exactly half of the important branches are mispredicted.
+Simply predicting that all branches are taken would result in a slightly higher
+prediction ratio of 50.8\%. Similarly, for improved shellsort, slightly more
+than half of the branches are correctly predicted, at 50.3\%, rising to 53\%
+using a strategy of predicting all branches as taken.
 
-In the unimporved version, half the bracnhes are mispredicted.
-Assuming taken would provide better results.
-Each key moves only 1.03 spots on average.
-Once the insertion sort is seperated, we see that this is because the insertion sort tips it in favour of 1 spot, and in actual fact this should be 1.07 spots, on average, though the difference is not very significant.
-The improved insertion sort is better behaved, moving each key on average two spots.
-But the prediction ratio is much worse than this.
-There should be N mispredictions, and instead there are 1.3xN mispredictions.
-THis is because the counter doesnt really get time to settle properly.
+The results also show that in shellsort, a key moves only 1.03 places down the
+array, on average, during each iteration. In improved shellsort, each key moves
+over 1.07\% places each iteration, which in the final insertion sort, each key
+moves an average of 2.1 places.
 
-%TODO end here
+The expected number of branch mispredictions is $log_{(\frac{11}{5})}N$. In
+actual fact, it is almost exactly correct, being slightly lower (18.48
+mispredictions per key, instead of 19.34, for 4194304 keys).
 

### 2005-07-24 docs/tech_report/code/improved_shellsort.c

@@ -1,4 +1,4 @@
-void improved_gonnet_shellsort(Item a[], int N)
+void improved_shellsort(Item a[], int N)
 {
 	int i, h, min, bound; 
 

### 2005-07-24 docs/tech_report/code/shellsort.c

@@ -1,4 +1,4 @@
-void gonnet_shellsort(Item a[], int N)
+void shellsort(Item a[], int N)
 {
 	int l = 0, r = N-1, i ,h;
 	for (h = r-l+1; h > 0; h = ((h > 1) && (h < 5)) ? 1 : 5*h/11)

### 2005-07-24 docs/tech_report/method.tex
updated heap scripts to remove 16 byte heapsort updated shell to remove reference to gonnet_ removed all as its done by makefile now
added testing bit

--- 2005-06-09 tech_report/docs/method.tex
@@
 
-\subsection{perfex}
-\cc{Perfex} is a tool used to access performance counters on a Pentium 4. By default,
-it prints out the number of cycles\footnote{The processor automatically
-maintains this figure in a 40-bit register called P4\_TSC (Time Stamp Counter),
-and perfex prints this value after running the program} which a program takes to
-execute. This was used to create the graphs comparing the running time of sorts.
-The gnuplot scripts for this were short and easily written.
+\subsection{PapiEx}
+\cc{PapiEx} is a tool used to access performance counters on a Pentium 4. It
+uses the PAPI framework to count certain characteristics across a program's
+execution. We used it to count the number of cycles a sort took to run. This
+provided an accurate measure of how fast a sort executes on a Pentium 4. The
+PapiEx homepage is at \n{http://icl.cs.utk.edu/~mucci/papiex/}.
 
-Perfex was an alternative to using SimpleScalar. By using its hardware
-registers, actual real world results could have been measured. Using perfex
-instead of SimpleScalar was an option reviewed at the start, but the difficulty
-involved in setting up a machine to use it was considerable. Additionally,
-because perfex measures results on a real machine, only one type of cache and
-branch predictor could have been measured. We have provided limited
-measurements to supplement the SimpleScalar results.
+PapiEx was used to create the graphs comparing the running time of sorts.
+The gnuplot scripts for this were short and easily written by hand.
+
+PapiEx was an alternative to using SimpleScalar. By using its hardware
+registers, actual real world results could have been measured. Using PapiEx
+instead of SimpleScalar was an option reviewed at the start, but because PapiEx
+measures results on a real machine, only one type of cache and branch predictor
+could have been measured. We have instead provided limited measurements to
+supplement the SimpleScalar results.
 
@@
 The software predictor used was a series of macros and functions which
-maintained the state of a branch predictor at a particular point in the program.
-Statistics about each branch were collected, and dumped into text files, from
-which scripts created graphs to display them. Some sample code is shown in
-Figure \ref{Sample code using the software predictor}.
+maintained the state of a bimodal branch predictor at a particular point in the
+program. Statistics about each branch were collected, and dumped into text
+files, from which scripts created graphs to display them. Some sample code is
+shown in Figure \ref{Sample code using the software predictor}.
 
@@
 %TODO another sentence here
-Each of the four $O(NlogN)$ sorts and radixsort, was treated in this fashion.
+Each of the four $O(NlogN)$ sorts, radixsort and shellsort, was treated in this fashion.
+
+\section{Testing details}
+%TODO a lot of this may be in the first sorts chapter, maybe heapsort.
+discuss: BIG + random.org, 64 bit in lamarca vs 32 bit, do_nothing and how it
+affects cache hits, scaling in the software branch predictors. How to compare
+and contrast varying results (branch vs softwre branhc; P4 vs simplescalar).
+What factors are used for testing: /proc/cpuinfo output, cahce and branch
+predictor data from P4, simplescalar inputs, software predictor, verfication of
+sorts.
 

### 2005-07-26 docs/tech_report/ordernsquared.tex
renamed tech_report_docs to tech_report
minor changes to text started lists of results and contributions

--- 2005-05-30 paper/docs/ordernsquared.tex
@@
 for every key to be sorted, every other key is checked.  Four of these sorts are
-discussed here. They are insertion sort, selection sort, bubblesort and
-shakersort.
+discussed here: insertion sort, selection sort, bubblesort and shakersort,
+along with improved versions of bubblesort and shakersort.
 
@@
 TODO Knuth has a formula which I will look up.
+
 \section{Insertion sort}
@@
 all metrics. Instruction count, were there no conditions that allowed keys to be
-skipped, would be the same, as no extra instructions are added. The flow of the
-algorithm is the same, though there should be fewer iterations, especially at
-the start, when iterations are longer, so cache accesses, and branch misses
-should be reduced along with instruction count.
+skipped, would be the similar, since few extra instructions are added. The flow
+of the algorithm is the same, though there should be fewer iterations,
+especially at the start, when iterations are longer, so cache accesses, and
+branch misses should be reduced along with instruction count.
 
@@
 for every move in bubblesort, since each move is disjoint from the next. In
-shakersort, however, each in the correct direction will follow a move in the
-same direction, each of which should be predicted. This works for a lot of
-keys, which are more likely to move in continuous movements than they are in
-bubblesort, where only small keys on the right move continuously. The
+shakersort, however, each move in the correct direction will follow a move in the
+same direction, each of which should be correctly predicted. This works for a
+lot of keys, which are more likely to move in continuous movements than they are
+in bubblesort, where only small keys on the right move continuously. The
 performance should improve noticeably as a result.
@@
 bubblesort in bubblesort2 have increased the instruction count, contrary to
-predictions, and similarly in the shakersort2.
+predictions, and similarly in the shakersort2. It appears the cost of
+skipping over a segment of the array is not worth the additional book-keeping
+cost of determining when this can be done.
 
@@
 place. Bubblesort, with similar behaviour, is only slightly better. Shakersort
-has a small amount of temporal reuse (see Section \ref{temporal locality}),
-reducing its number of misses. Shakersort2 is placed roughly between insertion
-sort and shakersort, showing that ending early is helpful in its case. This does
-not appear to be the case with bubblesort2 though, which performs identically to
-bubblesort.
+has a small amount of temporal reuse, reducing its number of misses. Shakersort2
+is placed roughly between insertion sort and shakersort, showing that ending
+early is helpful in its case. This does not appear to be the case with
+bubblesort2 though, which performs identically to bubblesort.
 

### 2005-07-26 docs/tech_report/heap.tex
minor changes as rereading

@@
 fewer keys fit into his cache lines. Since modern cache lines are typically
-longer than 4 bytes, we stick with 8-byte cache lines. \verb!UINT_MAX! was used
+longer than 4 ints, we stick with 32 byte cache lines. \verb!UINT_MAX! was used
 as the sentinel.
@@
 smallest are not predictable. In a 2-heap, there is one comparison between
-children, meaning that half the branches are predictable. An 8-heap has seven
-comparisons between children, with the result that only an eighth of the branches
-are predictable.
+children, meaning that half the branches will probably be correctly predicted.
+An 8-heap has seven comparisons between children, and will probably have about
+mispredictions, as it behaves similarly to selection sort (See \ref{selection
+branches logn}). It is possible, therefore, that the 8-heap will have slightly
+fewer mispredictions due to the combination of the shallower heap and the larger
+fan-out.
 
@@
 This continues into the level 2 cache, where the flattest slope belongs to the
-8-heap. The 4-heap with the 32-byte line has fewer misses than the 16-byte
-version, which is surprising since the 16-byte version can fit twice as many
-blocks in the cache, and uses each block fully.  This may be due to the extra
-compulsory misses due to having twice as many blocks to load. The difference
-between fully associative and direct mapped is only minor.
+8-heap. The difference between fully associative and direct mapped is only
+minor, with the fully associative performing slightly better than the direct
+mapped cache.
 
@@
 keys fit in the cache without a corresponding increase in misses. For the same
-reason, the base heapsorts have approximately half as many level 2 cache misses
-as the out-of-place sorts (See Figure \ref{heap level 2 misses}).
+reason, the base heapsorts have no compulsary misses\footnote{See Section
+\ref{TODO}.}, while the out-of-place sorts have one compulsory miss in eight,
+when the auxilliary heap is being populated.
 
-The branch prediction behaves similarly to the expected results, in that the
-prediction ratio gets worse as the fanout gets larger. However, the difference
-is minor between the 4-heap and 8-heap, and the difference for larger fanout
-would be less still. The difference between the 2-heap and 4-heap result is
-interesting though, the 2-heap has more branches, but less misses per key.
+The branch prediction behaves differently to the expected results, as the
+prediction ratio gets worse as the fanout gets larger. The reason we predicted
+differently was due to selection sort's $O(log N)$ mispredictions; however, this
+is over the entire length of the array, and it is not accurate to extrapolate
+these results over a sequence only seven keys long.
+
+The difference between the 4-heap and 8-heap is minor , and the difference for
+larger fanout would be less still. The difference between the 2-heap and 4-heap
+result is interesting though, the 2-heap has more branches, but less misses per
+key.
 
@@
 
+%TODO move this to testing section
 Figure \ref{heap cycles} shows the actual time taken for heap sort on a Pentium

### 2005-07-27 docs/tech_report/merge.tex
reread the merge - updated results and created expected vs actual added a label to heap added a label to smallplot in report change multi mergesort to multi-mergesort

@@
 it is difficult or impossible to perform several of the optimizations LaMarca
-describes on it.
+describes on it. We attempted to contact him in order to discuss this, but
+recieved no response. It is likely that the improvements were actually applied
+to Knuth's algorithm S.
 
@@
 
-Knuth's algorithm S is a \n{straight} merge, and works as described above. It has
-15\% lower instruction count than Algorithm N\footnote{This is observed on
-random data, which algorithm N is not designed for. On more regular data, or
-semi-sorted data, this will probably not occur.}. Algorithm S, like algorithm N,
-is a very difficult algorithm to modify. Knuth describes it in assembly, and
-making the optimizations that LaMarca describes on it is more difficult than
-rewriting the algorithm more simply. The rewritten algorithm, referred to from
-here on as `base mergesort', is a simple higher level version of algorithm S,
-and should perform only slightly more slowly.
-
-A disadvantage of this method is that if an array has slightly more than $2^x$
+Algorithm S is a \n{straight} merge, and works as described above.  A
+disadvantage of this method is that if an array has slightly more than $2^x$
 keys, an entire extra merge step would be required, which would considerably
 slow the algorithm. This problem is shared by all other versions of mergesort
-considered here, except algorithm N. However, since our tests use exact
-powers of two as set sizes, this problem does not affect them. LaMarca,
-however, used set sizes such as 4096000, which at times would suffer from this
-type of problem.
+considered here, except algorithm N. However, since our tests use exact powers
+of two as set sizes, this problem does not affect them. LaMarca, however, used
+set sizes such as 4096000, which at times would suffer from this type of
+problem.
+
+Algorithm S has 15\% lower instruction count than Algorithm N\footnote{This is
+observed on random data, which algorithm N is not designed for. On more regular
+data, or semi-sorted data, this will probably not occur.}. Algorithm S, like
+algorithm N, is a very difficult algorithm to modify. Knuth describes it in
+assembly, and making the optimizations that LaMarca describes on it is more
+difficult than rewriting the algorithm more simply. The rewritten algorithm,
+referred to from here on as `base mergesort', is a simple higher level version
+of algorithm S, and should perform only slightly more slowly.
 
@@
 
-Overall, these three optimizations reduced the instruction count by 30\%.
+Overall, these three optimizations reduce the instruction count by 30\%.
 
@@
 parts of the cache. To do this, an additional amount of memory the same size as
-the cache is allocated, which can be expensive for small arrays. The address of
+the cache is allocated. The address of
 the auxiliary array is then offset such that the block index in the source array
@@
 
-The next problem considered is that of temporal locality. Once a key is loaded
-into the array, it's only used once per merge step. If the data does not fit in
-half the cache, then conflict misses will occur, and no temporal reuse will
-occur at all. The solution is to apply a technique used in compilers called
-\n{tiling}. Using this technique, an entire cache sized portion of the array will
-be fully sorted before moving on to the next set. When all the sets have been
-sorted in the manner, they are all merged together as usual.
+The next problem LaMarca considers is that of temporal locality. Once a key is
+loaded into the array, it's only used once per merge step. If the data does not
+fit in half the cache, then conflict misses will occur, and no temporal reuse
+will occur at all. The solution is to apply a technique used in compilers called
+\n{tiling}. Using this technique, an entire cache sized portion of the array
+will be fully sorted before moving on to the next set. When all the sets have
+been sorted in the manner, they are all merged together as usual.
 
@@
 
-To avoid this, a priority queue is used. The smallest key from each list is
+To avoid this, LaMarca uses a priority queue. The smallest key from each list is
 added to the queue, and then put into the final target array. This reduces the
@@
 LaMarca used a 4-heap for his priority queue, and an 8-heap is used here, for
-the same reasons as before. The heap is aligned, and a cache line
-is put into the array each time. The smallest key is at the top of the heap,
-and this is put into its final position in the array. The heap is then fixed.
+the same reasons as before (see Section \ref{cache heapsort}.). The heap is
+aligned, and a cache line is put into the array each time. The smallest key is
+at the top of the heap, and this is put into its final position in the array.
+The heap is then fixed.
 
@@
 algorithm S. Algorithm S is expected to have a lower instruction count than base
-mergesort. Neither algorithm is expected to have good cache performance,
+mergesort. Neither algorithm is expected to have good cache performance.
 
@@
 fully associative caches should not have these misses, and the results should be
-correspondingly lower. Due to their design, the double mergesorts should have
-better performance than their counterparts, both in level 1 misses, and overall,
-though the difference is not expected to be great.
+correspondingly lower. The double mergesorts should have
+better performance than their counterparts in level 1 misses, though there
+should be a slight increase in level 2 misses. It is expected that they should
+perform slightly faster as a result.
 
@@
 
-As expected, algorithm N has lower instruction count than algorithm S, which
+As expected, algorithm S has lower instruction count than algorithm N, and 
 also slightly outperforms base mergesort.  The number of instructions, shown in
 Figure \ref{merge instructions} is exactly as predicted, with a 50\% spike at
-multi-mergesort. Level 2 misses, shown in Figure \ref{merge level 2 misses} are
+multi-mergesort.
+
+Level 2 misses, shown in Figure \ref{merge level 2 misses} are
 also as expected, although the multi-mergesort does not show the same
-improvement over tiled mergesort as in LaMarca's results.
+improvement over tiled mergesort as in LaMarca's results, as our tiled mergesort
+performs better. The cache performance of algorithms N and S are poor, as
+expected. They perform roughly the same as base mergesort.
+
+It is notable that our prediction of half a miss per key for multi-mergesort,
+which was taken from LaMarca, is exactly right. However, our assertion that the
+fully associative cache will perform slightly better is false. In actual fact,
+the direct mapped cache performs significantly better (see Figure
+\vref{multi-mergesort level 2 misses}). This is due to the replacement policy:
+our fully associative cache removes the least recently used keys from the cache
+to make room, while the direct mapped caches are designed to make best use of
+the keys while they are in the cache. They have, effectively, an optimised
+replacement policy.
 
@@
 multi-mergesorts is considerably lessened by the two level adaptive predictor.
-The reason for this is not clear.
+This shows there must be regular access patterns in the k-way merge at the end
+of multi-mergesort, which are predictable using a branch history.
 
@@
 has consistently four times fewer level 1 misses than multi-mergesort, and both
-double tiled sorts perform significantly better than the unoptimized sorts.
+double tiled sorts perform significantly better than the unoptimized sorts. The
+expected increase in level 2 misses did not materialise, and is in fact
+negligible.
 
@@
 Figure \ref{merge level 2 misses} is exactly that provided by LaMarca. The
-ratio's between the each of the sorts is also the same, though the numbers are
-not identical, due to different \c{int} sizes in our tests.
+ratio's between the each of the sorts is also the same, as each of our results
+is exactly half that of LaMarca, due to different \cc{int} sizes in our tests.
 

### 2005-07-27 docs/tech_report/report.tex
made the thing compile now
reread the merge - updated results and created expected vs actual added a label to heap added a label to smallplot in report change multi mergesort to multi-mergesort

--- 2005-05-27 paper/docs/report.tex
@@
 \subfigure[Level 1 cache misses per key]	{\includegraphics[scale=0.74]{plots/#1_-_Level_1_misses.eps}}
-\subfigure[Level 2 cache misses per key]	{\includegraphics[scale=0.74]{plots/#1_-_Level_2_misses.eps}}
+\subfigure[Level 2 cache misses per key]	{\label{#1 level 2 misses}\includegraphics[scale=0.74]{plots/#1_-_Level_2_misses.eps}}
 \subfigure[Branches per key]				{\includegraphics[scale=0.74]{plots/#1_-_Branches.eps}}
@@
 \begin{changemargin}
-\subfigure[]{\includegraphics[scale=0.74]{../counters/#1}}
-\ifthenelse{\equal{#2}{}}{}{\subfigure[]{\includegraphics[scale=0.74]{../counters/#2}}}
-\ifthenelse{\equal{#3}{}}{}{\subfigure[]{\includegraphics[scale=0.74]{../counters/#3}}}
-\ifthenelse{\equal{#4}{}}{}{\subfigure[]{\includegraphics[scale=0.74]{../counters/#4}}}
-\ifthenelse{\equal{#5}{}}{}{\subfigure[]{\includegraphics[scale=0.74]{../counters/#5}}}
-\ifthenelse{\equal{#6}{}}{}{\subfigure[]{\includegraphics[scale=0.74]{../counters/#6}}}
+\subfigure[]{\label{#7-1}\includegraphics[scale=0.74]{plots/counter_#1}}
+\ifthenelse{\equal{#2}{}}{}{\subfigure[]{\label{#7-2}\includegraphics[scale=0.74]{plots/counter_#2}}}
+\ifthenelse{\equal{#3}{}}{}{\subfigure[]{\label{#7-3}\includegraphics[scale=0.74]{plots/counter_#3}}}
+\ifthenelse{\equal{#4}{}}{}{\subfigure[]{\label{#7-4}\includegraphics[scale=0.74]{plots/counter_#4}}}
+\ifthenelse{\equal{#5}{}}{}{\subfigure[]{\label{#7-5}\includegraphics[scale=0.74]{plots/counter_#5}}}
+\ifthenelse{\equal{#6}{}}{}{\subfigure[]{\label{#7-6}\includegraphics[scale=0.74]{plots/counter_#6}}}
 \end{changemargin}
@@
 \title{Fast Sorting on Modern Computers}
-\author{Paul Biggar, David Gregg}
+\author{Paul Biggar \and David Gregg}
 \date{}
@@
 %TODO aspell in UK english
+%TODO sometimes I say 3 sorts, sometimes 4. It's now 5.
+%TODO do a final check for use of the word I
+%TODO final check for the word 'project'
 
@@
 
-%\maketitle
+\begin{titlepage}
+        \begin{center}
+                \vspace*{95mm}
+                \LARGE Fast Sorting on Modern Computers\\
+                \vspace{1em}
+                \large Paul Biggar, David Gregg\\
+                \large Department of Computer Science\\
+				\large University of Dublin, Trinity College
+        \end{center}
+\end{titlepage}
 
 \begin{abstract}
-This project attempts to create and analyse sorts designed to take advantage of
-features of modern processors. The work on cache-conscious sorting by LaMarca
-and Ladner is the starting point; this is implemented and analysed using a
-hardware simulator. Results relating to a variety of cache and branch predictor
-configurations are obtained and analysed for a variety of sorts, beginning with
-insertion and selection sorts, bubblesort and shakersort, and continuing to
-heapsort, mergesort, quicksort and radixsort. Actual and expected results are
-compared together and to the results of LaMarca's research.
+This report describes the creation and analysis of sorts designed to take
+advantage of features of modern processors. The work on cache-conscious sorting
+by LaMarca and Ladner is the starting point; this is implemented and analysed
+using a hardware simulator. Results relating to a variety of cache and branch
+predictor configurations are obtained and analysed for a variety of sorts,
+beginning with insertion and selection sorts, bubblesort and shakersort, and
+continuing to heapsort, mergesort, quicksort and radixsort. Actual and expected
+results are compared together and to the results of LaMarca's research.
 \end{abstract}

### 2005-07-27 docs/tech_report/heap.tex
reread the merge - updated results and created expected vs actual added a label to heap added a label to smallplot in report change multi mergesort to multi-mergesort
finished the comparison to lamarcs results for heapsort

@@
 child. With this property, a heap can be stored in an array without any wasted
-space, and without any links to other nodes. A heap done in the fashion is
-called an \n{implicit heap}.
+space, and the links are implied by a nodes position, rather than by a pointer
+or reference. A heap done in the fashion is called an \n{implicit heap}.
 
@@
 can be done due to the nature of the indices of parents and their children. In
-the case of an array indexed between $0$ and $N-1$, then a 
-parent whose index is $i$ has two children with indices of
-$2i+1$ and $2i+2$. Conversely, the index of a parent whose child has an index
-of $j$ is $\lfloor{}(j+1)/2-1\rfloor{}$. It is also possible to index the array between $1$
-and $N$, in which case the children have indices of $2i$ and $2i+1$, while the parent
-has an index of $\lfloor{}j/2\rfloor{}$. Using this scheme can reduces instruction cost, but
+the case of an array indexed between $0$ and $N-1$, a parent whose index is
+$i$ has two children with indices of $2i+1$ and $2i+2$. Conversely, the index of
+a parent whose child has an index of $j$ is $\lfloor{}(j+1)/2-1\rfloor{}$. It is
+also possible to index the array between $1$ and $N$, in which case the children
+have indices of $2i$ and $2i+1$, while the parent has an index of
+$\lfloor{}j/2\rfloor{}$. Using this scheme can reduces instruction cost, but
 results is slightly more difficult pointer arithmetic, which cannot be performed
@@
 \section{Cache Heapsort}
-
+\label{cache heapsort}
 LaMarca notes that the Repeated-Adds algorithm has a much lower number of cache
@@
 $\lceil{}log_4(N)\rceil{} = \lceil{}\tfrac{1}{2}log_2(N)\rceil{}$. The height of
-an 8-heap is half again the height of a 4-heap. However, the number of
-comparisons needed in each level is doubled, since all the children will be
-unsorted. LaMarca claims that as long as the size of the fanout is less than
-12, a reduced instruction count will be observed.
+an 8-heap is reduced again to $\lceil{}log_8(N)\rceil{} =
+\lceil{}\tfrac{1}{3}log_2(N)\rceil{}$. However, the number of comparisons needed
+in each level is doubled, since all the children will be unsorted. LaMarca
+claims that as long as the size of the fanout is less than 12, a reduced
+instruction count will be observed.
 
@@
 this increase will be an order of magnitude lower than the bounds check.
-
-%TODO this isnt true is it? placing the nsentinel is cheaper than a bounds check
-%(a write to a variable thats going to be written to in a second vs a comparison?)
-%and there arent an extra logN instructions, theres logN times as many
 
@@
 line.
+
+LaMarca reports base heapsort has very poor temporal locality, and that the
+cache heapsort will have half the cache misses of base heapsort.
 
@@
 increased number of comparisons of the 8-heap increases the instruction count to
-even higher than base heapsort, despite the reduction in the height of the heap.
-This is shown in Figure \ref{heap instructions}.
+even higher than base heapsort, due to the disparity between the increase in
+comparisons in set of children, and the lesser reduction in the height of the
+heap. This is shown in Figure \ref{heap instructions}.
 
@@
 difference between the two types of heap. The slope of the base heapsort result
-is much greater than all the others, with the 8-heap having the fewest misses.
-This continues into the level 2 cache, where the flattest slope belongs to the
-8-heap. The difference between fully associative and direct mapped is only
-minor, with the fully associative performing slightly better than the direct
-mapped cache.
+is much greater than all the others, showing base heapsort's poor temporal
+locality, with the 8-heap having the fewest misses. This continues into the
+level 2 cache, where the flattest slope belongs to the 8-heap. The difference
+between fully associative and direct mapped is only minor, with the fully
+associative performing slightly better than the direct mapped cache.
 
@@
 
-The difference between the 4-heap and 8-heap is minor , and the difference for
+The difference between the 4-heap and 8-heap is minor, and the difference for
 larger fanout would be less still. The difference between the 2-heap and 4-heap

### 2005-07-28 docs/tech_report/quick.tex
added quicksort results

@@
 the next phase of the algorithm. Secondly, when sorting random keys over a
-large range, the number of equal keys will be vewry low. Using a median of
-greater than three is discussed in section \ref{quick-predictors2}.
+large range, the number of equal keys will be very low. Using a median of
+greater than three is discussed in Section \ref{quick-predictors2}.
 
@@
 check is done to fetch a sentinel from the appropriate subfile if required. This
-check is expensive, but less so than either a bounds check or putting a sentinel
-into every subfile.
+check isn't very expensive, and is significantly cheaper than either a bounds
+check or putting a sentinel into every subfile.
 
@@
 subset size to be $C/3$, where $C$ is the number of keys which fit in the
-cache.
+cache, on average only 5\% of subsets will be larger than the cache.
 
@@
 of instructions for multi-quicksort should increase by about 50\% in the case of
-the sequential sort, and 25\% in the case of the binary sort.
+the sequential sort, and 25\% in the case of the binary sort. Cache quicksort
+should also have a slightly higher instruction count than base quicksort, due to
+the change in use of insertion sort.
 
-LaMarca notes that quicksort is very cache efficient already. Once it no longer
-fits in the cache, however, the number of misses should increase significantly.
-This is especially true in the case of the base quicksort, which does an
-insertion sort across then entire array at the end.
+LaMarca notes that quicksort is very cache efficient already. It has excellent
+spatial and temporal locality. Once it no longer fits in the cache, however, the
+number of misses should increase significantly. This is especially true in the
+case of the base quicksort, which does an insertion sort across then entire
+array at the end.
+
+Cache quicksort should have fewer misses than base quicksort as soon as the
+array is larger than the cache. It is expected that these misses will make up
+for the increase in instruction count.
+
+LaMarca estimates a large reduction in level 2 misses in multi-quicksort, and
+puts a figure on it of four misses per cache block, meaning, in our case, half a
+miss per key.
 
@@
 
-The instruction count, shown in Figure \ref{quick instructions} is as
-predicted. The expected increases in instruction count for multi-quicksort was
-due to a test with a smaller data set and smaller constants. The actual results
-were higher than expected: being 63\% and 40\% greater, for
-sequential and binary search versions, respectively.
-%TODO check
+The instruction count, shown in Figure \ref{quick instructions} is moslty as
+predicted. The increase in instructions in both multi-mergesorts is as
+predicted, with the binary search version increasing only 25\% against the
+sequential search version's 50\% increase. However, the increase we predicted in
+the instrcution count of cache quicksort based on the results of both Sedgewick
+and LaMarca -  that is, once the efficient insertion sort was replaced with many
+small insertion sorts - we observe that the instruction count remains exactly
+the same.
 
@@
 predictions, except that the multi-quicksorts did not behave as expected, and
-performed worse than the memory-tuned version. Not only do they have more
-misses per key, but the rate at which the misses increase with size is also
-worse for the multi quicksorts. The tuned version performed better than the
-base version, as predicted.
+performed as badly as the memory-tuned version. However, the rate at which the
+cahce misses are increasing is very low, being almost a flat line on our graph.
+LaMarca's calculation of four misses per cache block is very accurate, as we
+show 0.55 misses per key, or 4.4 misses per block.
 
-A surprising result is that direct mapped caches often perform better than fully
-associative ones. The reason for this is not clear, but it is obvious from the
-graph that this is true in the case of all the quicksorts.
-%TODO why is this? the only reason I've come up with is that in the direct
-%ampped cache, the same area is thrashed, while keeping keys that we;re going to
-%reuse is a separte bit. in the fully assoc cache, those keys get washed out by
-%the keys that arent going to be reused.
+The tuned quicksort performed better than the base version, as predicted. Both the
+tuned version and the base quicksort both have a significant increase once the
+arrays no longer fit in the cache, also as predicted.
 
-The branch prediction results in Figure \ref{quick branch misses} were exactly
-as expected. The binary search produced a dramatic increase in the branch
-misses, when using a bimodal predictor. When using a 2-level adaptive predictor,
-however, the number of misses is, in general, slightly higher, though it does
-not spike due to the binary search. While this performs worse than the bimodal
-predictor, it appears to show that 2-level adaptive predictors can adapt to the
+The results show again that direct mapped caches can perform better than fully
+associative ones. This is true in the case of all the quicksorts, as can be seen
+in Figures \ref{base quicksort (no median) level 2 misses} to
+\ref{multi-quicksort (sequential search) level 2 misses} graph that this is true
+in the case of all the quicksorts.
+
+The branch prediction results are shown in Figure \ref{quick branch misses}. As
+expected, there was a a dramatic increase in the binary search multi-quicksort
+branch misses, but only when using a bimodal predictor. When using a 2-level
+adaptive predictor, the number of misses is, in general, slightly higher, but it
+does not spike due to the binary search. While it performs worse than the
+bimodal predictor, it appears that 2-level adaptive predictors can adapt to the
 erratic branches of a binary search.
@@
 in instruction count, which was significantly reduced by using the binary
-search.  The tuned quicksort, however, had the lower running time and even base
-quicksort out-ran the multi-quicksorts.
+search.
+
+Tuned quicksort had fewer branch misses than base quicksort, due to the fact
+that fewer comparitive branches are executed. In base quicksort, we sort the
+entire array using insertion sort. In tuned quicksort, we don't sort the pivots,
+but only the keys between them.
+
+Despite the sequentially search multi-quicksort performing better than its
+binary searched cousin, and its good cache performance, it still failed to
+approach the performance of either base or tuned quicksort. In fact, base
+quicksort was the fastest of all the quicksorts in our performance counter test,
+even though our simulator results show that it has more level 1 misses, level 2
+misses, and branch misses than tuned quicksort, and the same number of
+instructions.
 
@@
 but that of the cache misses is not. The ratio of the cache misses between base
-and cache quicksort are the same in both sets of results, but the number of
-cache misses in multi-quicksort is much higher here than it is in LaMarca's
-findings. In addition, the shape is different. LaMarca's multi-quicksort's
-cache misses plateau, while the results here show an increasing slope.
-%TODO is a pseudo-median explained
+and cache quicksort are the same in both sets of results, but these are lower
+than the same misses in LaMarca's results. However, the multi-quicksort results
+appear to be the same, though our results are not conclusive either way.
+
 \section{Predictor results}

### 2005-07-28 docs/tech_report/merge.tex
finsied radixsort

@@
 \section{Tiled Mergesort}
+\label{tiled mergesort}
 LaMarca's first problem with mergesort is the placement of the arrays with
@@
 \section{Double Tiled Mergesort}
+\label{double tiled mergesort}
 Offsetting the target array to avoid conflicts reduces significantly the number

### 2005-07-28 docs/tech_report/radix.tex
finsied radixsort

@@
 sorting on that basis. Radixsort is a \n{counting based} sort. It works by
-separating all the keys into small parts, and processing these parts without
+separating the keys into small parts, and processing these parts without
 regard for other keys. For radix of two, these small parts are bits, and for a
@@
 The steps to sort by radix are simple. First a radix is chosen, based on the
-size of the part to be sorted. For 32 bits integers, a byte is used. This means
+size of the part to be sorted. For 32 bit integers, a byte is used. This means
 the radix is 256, as there are 256 possible byte values. An array of 256
@@
 count of keys with a smaller least significant byte. Now each key can be put in
-place by accessing the counter indexed by its least significant byte.
+place by accessing the counter indexed by its least significant byte, after
+which that counter is incremented.
 
-These are written into an auxiliary array, which is written back to the initial
+The keys are written into an auxiliary array, which is written back to the initial
 array before the next iteration. The process is then repeated using the second
@@
 
+\section{Base Radixsort}
+
 This type of radixsort is called \n{Least Significant Digit} radixsort.
-Other implementations are possible, inclduing a \n{Most significant bit}
-version, which behaves as a divide and conquer algortithm. These are discussed
+Other implementations are possible, including a \n{most significant bit}
+version, which behaves as a divide and conquer algorithm. These are discussed
 in Section \ref{radix future work}.
 
-Note that this method is only suitable for unsigned ints. Because a signed int's
-sign is contained in the most significant bit, this sort would place negative
-numbers after the positive numbers.
+Note that this method is only suitable for \cc{unsigned int}s. Because a
+\cc{signed int}'s sign is contained in the most significant bit, this sort would
+place negative numbers after the positive numbers.
 
-\subsection{Cache Radixsort}
+\section{Cache Radixsort}
 Many improvements exist to radixsort, and some are suggested by Sedgewick and
 LaMarca. Some of these improvements help to lower cache misses and instruction
-count, both of which slow down radixsort significantly. LaMarca suggested
-that while the current byte is being put in place, the counting of the next byte
-can be done. This increases the memory requirements, but this is not
-significant.
+count, both of which slow down radixsort significantly. LaMarca suggests an
+improvement by Friend\footnote{See \ref{TODO friend[8]}.} that while the current
+byte is being put in place, the counting of the next byte can be done. This
+increases the memory requirements, but this is not significant.
 
@@
 step.  In the \n{base radixsort}, presented above, there is one step to count,
-one to copy, and one to write back (which accesses both arrays). LaMarca
-suggests counting for the next step in the previous count, which we implement.
+one to copy, and one to write back (which accesses both arrays). We take
+LaMarca's suggestion and begin the counting of the next least significant byte
+in the previous iteration.
 
@@
 auxiliary arrays. The same could be done with radixsort, without any increase in
-complexity. The number of steps is likely to be even, and is known at compile
-time, so the number of times the copy step is done could be reduced to zero. If
-the number of steps is odd, which is unlikely when sorting integers, then one
-copy at the end would be required. This is not needed in the implementation
-presented here.
+complexity. The number of steps is likely to be even, and depends on the radix.
+It is known at compile time, so the number of times the copy step is done could
+be reduced to zero since we choose an even radix. If the number of steps is odd,
+which is unlikely when sorting integers, then one copy at the end would be
+required. This is not needed in the implementation presented here.
 
 Cache radixsort should traverse the array only nine times: Once at the start to
-count the least significant bits; six times as it copies using the three least
-significant bytes (counting the next least significant byte at the same time),
-and a final two as it copies using the most significant byte.
+count the least significant bytes; six times (once per array per pass) as it
+copies using the three least significant bytes (counting the next least
+significant byte at the same time), and a final two as it copies using the most
+significant byte.
 
@@
 arrays are exactly half the length of the cache apart. We further added an
-alignment for the level 1 cache to significantly reduce level 1 misses at the
-expense of a small number of level 2  misses. We adapt this strategy here, and
-call it Aligned Cache Radixsort.
+alignment for the level 1 cache (see Section \ref{double tiled mergesort}) to
+significantly reduce level 1 misses at the expense of a small number of level 2
+misses. We adapt this strategy here, and call it Aligned Cache Radixsort.
 
@@
 as many steps to complete. However, it is not expected that radixsort will
-perform many times better than quicksort, the reigning champion. A `step' for
-quicksort is as simple as a comparison and a possible exchange every few steps.
-For radixsort, a step comprises being counted four times and copied four times,
-with the counting involving several layers of indirection, and causing a
-complete iteration across the array at the start. LaMarca's results
-show the instruction count of quicksort and radixsort being in the same order of
-magnitude, and we expect the same.
+perform many times better than quicksort, the reigning champion of $O(NlogN)$
+sorts. A `step' for quicksort is as simple as a comparison each step and a
+possible exchange every few steps.  For radixsort, a step comprises being
+counted four times and copied four times, with the counting involving several
+layers of indirection, and causing a complete iteration across the array at the
+start. LaMarca's results show the instruction count of quicksort and radixsort
+being in the same order of magnitude, and we expect the same.
 
@@
 arrays. In fact, we have no idea where is being accessed in the source array, so
-it is unlikely that the alignment will have any effect what-so-ever.
+it is unlikely that the alignment will have a great effect.
 
 The number of iterations across the array is constant, however, proportional to
-$(bitsPerInt/log_2(Radix))$. For a 32 bit integer, this is four steps and each step
-has four passes over the array, for a total of 16 passes over the array.  With
-eight keys per cache line, there should be about two misses per key, when the
-array doesn't fit into the array, though this depends on the placing of the
-auxiliary array.
+$(bitsPerInt/log_2(Radix))$. For a 32 bit integer and a radix of 256, this is
+four steps and each step has four passes over the array, for a total of 16
+passes over the array.  With eight keys per cache line, there should be about
+two misses per key, when the array doesn't fit into the array, though this
+depends on the placing of the auxiliary array.
 
@@
 per key with base radixsort. The complexity and branch prediction results
-should remain the same.
+should remain the same. The instruction count should, however, reduce by nearly
+a half to $fract{9}{16}$ths of base radixsort.
+
+Based on LaMarca's results, showing quicksort to be faster than radixsort by a
+small amount, we expect cache radixsort to run faster than quicksort.
 
@@
 very low. It is lower than even quicksort for the smallest set, and it
-flat-lines immediately (after amortisation of the loop overhead). The number of
-cache misses in Figure \ref{radix level 2 misses} is exactly the same, until
-the array no longer fits into the cache. The number of misses then leaps up to
-2.5 per key, which is slightly more than was predicted. The extra is likely due
-to the unpredictable way that the source array is accessed as the keys are
-copied to the auxilliary array.
+flat-lines immediately (after slight amortisation of the loop overhead). The
+number of cache misses in Figure \ref{radix level 2 misses} is only slightly
+more than quicksort - due to being out-of-place - until the array no longer fits
+into the cache. The number of misses then leaps up to 2.5 per key, which is
+slightly more than was predicted. The extra is likely due to the unpredictable
+way that the source array is accessed as the keys are copied to the auxilliary
+array.
 
@@
 has been reduced by more than 10\%, and the number of level 2 cache misses
-have been more than halved, to one and a quarter miss per key. This is exactly
-what is expected, due to there being nine passes over the array, and a one miss
-per pass per cache line.
+have been halved, to one and a quarter miss per key. This is exactly
+as expected, due to there being nine passes over the array, and a one miss
+per pass per cache line (coming to slightly under one and a quarter. The extra
+is again probably due to the unpredictable order of the copying).
 
-The number of misses while the array fits in the cache is also obvious: there
+The number of misses while the array fits in the cache is obvious: there
 are two compulsory misses per key - one for each array - which divided by the
-keys per cache line results in exactly one miss in four.
+keys per cache line results in exactly one miss in four (although our graph
+obscures this, due to the initial subtraction of compulsary misses).
 
-Aligned cache radixsort performs exactly the same as cache radixsort, indicating
-that the alignment, and the potentially large memory overhead, are not
-necessary.
+Aligned cache radixsort performs exactly the same as cache radixsort in both
+level 1 and level 2 misses, indicating that the alignment, and the associated
+memory overhead, are not necessary.
 
 The number of branch misses, shown in Figure \ref{radix branch misses} was also
-extremely low. Despite 2 branches per key, there was less than .08 of a branch
+extremely low. Despite three branches per key, there was less than .02 of a branch
 miss per key. The number of branches itself is quite low, as several of the
@@
 
-With the low number of branch misses, a much better overall performance is
-expected than is achieved. When in the cache, it performs only slightly worse
-than quicksort, but once the data set grows larger, the gap widens.  However,
-the results, shown in Figure \ref{radix cycles}, indicate that radixsort will
-not increase much beyond 350 cycles per key, meaning that for larger sets it will
-perform significantly better than quicksort. Cache radixsort does not reach 
-250 cycles per key, though it has relatively more cycles when the arrays are
-smaller.
+Again we see that the bimodal branch predictor performs better than the two
+level adaptive. In this case it is contrary to expectations, however, as
+there are no comparative branches. However, there are so few misses that these
+results do not show anything; in the worst case there are slightly over 200
+misses in total.
+
+With the low number of cache misses, low instruction count and negligible branch
+mispredictions, it is no surprise that cache radixsort performs as well as it
+does. As shown in Figure \ref{radix cycles}, cache radixsorts performance is
+almost linear, with only a slight increase when the arrays no longer fit in the
+cache.  Even then, the increase in cache misses are more than made up for by the
+lack of branch mispredictions, and quicksort ends up taking almost 50\% percent
+longer to sort more than 4 million keys.
 
@@
 falls before flattening out. By contrast, the graph here is almost completely
-flat. A similar thing happens in both cache miss graphs: a sharp dip in
-LaMarca's results, and a flat-line with a slight slope in these results. Also,
-in both cases, once the array no longer fits in the cache, the number of misses
-jumps to around 2.5, where they stay. 
-%TODO rewrite this bit a little
-%TODO cache radixsort is faster than quicksort in all cases
-The cycles per key in both cases are unsurprising based on instruction count and
-cache miss results: LaMarca's have a sharp fall before rising slightly once the
-data set no longer fits in the cache; the result presented here flat-line before rising
-slightly for the same reason. While both results agree for large data sets, they do not
-for smaller sets.
+flat. He shows a similar result in both cache miss graphs: a sharp dip, followed
+by a nearly flat-line. Also, in both cases, once the array no longer fits in the
+cache, the number of misses jumps to around 2.5, where they stay.
+
+However, LaMarca's conclusion - that radixsort is not as fast as quicksort - can
+be easily seen from the base radixsort results. Once an radixsort optimised for
+the cache is presented, these no longer hold, and the lack of branch
+misprediction results in a new fastest sort.
 
@@
 The radixsort presented above is a Least Significant Digit radixsort. A Most
-Significant Digit radixsort behaves differently. Once the initial sort is done, each
-bin in recursed into, and continually sorted in this manner. This results in a
-higher instruction count, and increased stack usage due to the recursion.
+Significant Digit radixsort behaves differently. Once the initial sort is done,
+each bin is recursed into, and continually sorted in this manner. This results
+in a higher instruction count, and increased stack usage due to the recursion.
 However, there should be a large increase in temporal reuse due to this, which

### 2005-07-28 docs/tech_report/shell.tex
finished shell

@@
 
-Shell begins by considering the array to be made up of $N/2$ subarrays of
-length two, where the elements of the subarray are spaced $N/2$ keys apart.
-These subarrays are then sorted using insertion sort. The first iteration will
-be a simple exchange, but the keys which are exchanged move across the array by
-$N/2$ spaces. The sort then considers $N/4$ arrays, who's elements are spaced
-$N/4$ keys apart, which are individually sorted by insertion sort, then $N/8$
-and so on. The final iteration will be a standard insertion sort, after which
-the array will be sorted. A problem with this implementation is that since the
-increments are powers of two, keys with odd indices are not compared with keys
-with even indices until the final insertion sort. This can lengthen the final
-insertion sort's running time to $O(N^2)$.
+Shell begins by considering the array to be made up of $\frac{N}{2}$ subarrays
+of length two, where the elements of the subarray are spaced $\frac{N}{2}$ keys
+apart.  These subarrays are then sorted using insertion sort. The first
+iteration will be a simple exchange, but the keys which are exchanged move
+across the array by $\frac{N}{2}$ spaces. The sort then considers $\frac{N}{4}$
+arrays, whose elements are spaced $\frac{N}{4}$ keys apart, which are
+individually sorted by insertion sort, then $\frac{N}{8}$ and so on. The final
+iteration will be a standard insertion sort, after which the array will be
+sorted. A problem with this implementation is that since the increments are
+powers of two, keys with odd indices are not compared with keys with even
+indices until the final insertion sort. This can lengthen the final insertion
+sort's running time to $O(N^2)$.
 
@@
 \textit{increments} to use in shellsort. Shell's original sort, described above,
-has the increments $N/2$, $N/4$, $N/8$, $N/16$ ........ $1$. \cite{Sedgewick96}
-discusses many increments used, and provides code for comparing these. Based on
-running the \cc{driver} program\footnote{See
+has the increments $\frac{N}{2}$, $\frac{N}{4}$, $\frac{N}{8}$, $\frac{N}{16}$
+........ $1$. \cite{Sedgewick96} discusses many increments used, and provides
+code for comparing these. Based on running the \cc{driver} program\footnote{See
 http://www.cs.princeton.edu/\textasciitilde{}rs/shell/}, Gonnet's shellsort was
@@
 to test the performance of shellsort. This uses increments which are
-five-elevenths of the previous increment.
+$\frac{5}{11}$ths of the previous increment.
 
@@
 
+Improved shellsort should have a slighlty lower instruction count, and hence a
+lower cycle count, than shellsort. It's cache performance should be the same,
+and it should have the same number of comparitive branches, and hence a similar
+number of mispredictions, though its branch count should be higher than
+shellsort.
+
 \subsection{Simulation results}
@@
 Shellsort's simulation results have the characterist shape and range of an
-$O(NlogN)$ sort, albeit, not a very good one. It performs better than a
+$O(NlogN)$ sort, albeit, not a very good one. The slopes of its graphs are
+slighlty steeper, as can be seen in Figures \ref{shell cycles}, \ref{shell
+instruction count}, \ref{shell level 1 misses} and \ref{shell branch
+mispredictions}. It performs better than a
 heapsort, but considerably worse than quicksort. However, it easier to program
@@
 quicksort, though it is more than 50 percent faster than the best of the
-heapsorts.a
+heapsorts.
+
+Shellsort's cache misses are certainly in the range of an $O(NlogN)$ sort, and
+are not considerably worse than cache heapsort, and are in fact better than base
+heapsort. This occurs without any cache improvements being performed upon them.
+Shellsort's branch mispredictions, are, as predicted $O(Nlog_{\frac{5}{11}}N)$.
+
+Improved shellsort does improve on shellsort, having approximately 10\% fewer
+instructions. It also has slightly fewer cache misses and branch mispredictions.
+Overall, this leads to a reduction in cycle count of between 35 and 70\%, as
+seen in Figure \ref{shell cycle count}.
 
 %TODO
-TODO are there log5_11 N misses
-Regenerate data before looking at this
-
 \section{Predictor results}

### 2005-07-29 docs/tech_report/conclusions.tex
added some heading to conclusings
removed the word gonnet from code
finished up shellsort
added quicksort results
added comparisons with lamarca to the conclusions

--- 2005-05-27 paper/docs/conclusions.tex
@@
-This section presents a summary of the results from previous chapters, lists
-important results and discoveries, and suggests contributions made to computer
-science by this project.
+\label{conclusions}
+This section presents a summary of the results from previous chapters, compares
+our results with LaMarca's, lists important results and discoveries, and
+suggests contributions made to computer science by this research.
+
+\section{Results summary}
+
+\subsection{Order N squared sorts}
+\subsection{Heapsorts}
+\subsection{Mergesorts}
+\subsection{Quicksorts}
+\subsection{Radixsorts}
+\subsection{Shellsorts}
+
+\section{Comparison with LaMarca's results}
+
+Over the course of this research. The vast majority of our results have
+correlated exactly with LaMarca's. Many of our graphs are exactly the same
+shape to LaMarca's, can be calculated exactly or alomst exactly using his
+methods, or can be seen to be equivalent when scaled by an appropriate ratio
+(due to different test parameters).
+
+LaMarca's heapsort graphs were equivalent to ours in the areas of level 1
+misses, instruction count and cycle count. Like LaMarca, we found that a four
+heap lowered our instruction count. Our level 2 misses also concurred with
+LaMarca, with only one difference: cache heapsort, being an out-of-place sort,
+fills the cache fully with a smaller number of keys than base heapsort, which is
+in place. LaMarca's results do not show this. Nor do they show a difference in
+cache misses between these two sorts when the array does fit in the cache.
+LaMarca's base heapsort have the characteristics of an out-of-place heapsort.
+However, our results show that his implicit heap improvements have led to a
+significantly faster heapsort.
+
+LaMarca's initial improvements to base mergesort were due to reduce the
+instruction count. We noted a reduction with each of the steps he prescribed. We
+note too that each of his improvement reduce level 2 cache misses, and the shape
+of our level 2 cache miss graphs exactly match his. We also note similar
+characterists: LaMarca mentions a wobble in the instruction count graphs, due to
+having to copy the final merge back; we note this same wobble.
+
+However, we disagree with his results in several key places. Firstly, we note
+that an inlined sorting method does not perform as fast as insertion sort. We
+find it difficult to perform his improvements on either algorithm N or algorithm
+S, and write our own mergesort instead. We expected that mergesort would have
+half the instruction count of heapsortl we found instead they were very similar.
+
+LaMarca did not include details of his tagging system in multi-mergesort. As a
+result, we used a search, which greatly increased the instruction count. As a
+result, mutli-mergesort did not perform as well as the mergesorts which had only
+been tiled.
+
+We note also that aligning the level 2 cache alone leads to worse
+performance than not making any improvements at all, due to the thrashing of the
+level 1 cache. When we aligned the arrays for the level 1 cache, the results
+were much closer to LaMarca's results, and showed what an improvement his
+techniques brought to those algorithms.
+
+Our quicksort results, again, we very similar to LaMarca's. We showed an
+increase is instruction count of 25\% due to multi-quicksort - very close to
+LaMarca's 20\%.  The shape of our instruction count graphs were the same, and
+switching to a binary search in multi-quicksort reduces the instruction count.
+Like him, we found that the base quicksort had very few cache misses, and that
+not using Sedgewick's insertion sort improvement reduces cache misses. We found
+that multi-quicksort's k-way sort changes the slope of the graph, but our
+result were not conclusive as to whether this would eventually make
+multi-quicksort faster than the other versions. Our results do show it being
+slower for all our test cases, due to the increase in instruction count.
+
+LaMarca predicts two cache misses per key, which translated to one miss per key
+with our 32 bit integers. We found this calculation to be true, as is his
+calculation of a decrease of one cache miss in four: we note a reduction of one
+cache miss in eight.
+
+Where LaMarca found that using Sedgewicks insertion sort optimisation decreased
+instruction count, we found the opposite. We also found that despite
+improvements due to cache performance and instruction count, our base quicksort
+performed faster than our tuned version. Finally, we found that although
+changing multi-mergesort's sequential search to a binary search decreased the
+instruction count, it greatly increased the number of branch mispredictions
+incurred, and slowed the sort down accordingly.
+
+LaMarca determines that radixsort could not be improved in the same was as the
+comparison based sorts, and that due to poor cache performance, it ran more
+slowly than quicksort. We showed, however, that one of the improvements applied
+to mergesort could be applied to it, and that its resulting cache performance is
+sufficiently good that its low instruction count and lack of branch
+mispredictions make it the fastest sort in our survey.
+
+\section{Important results and contributions}
+
+%TODO lamarca does an overall comparison. we should do the same
+
+\section{Instrucyion count results}
 
@@
 
+% TODO more details in this
 The test results in this project mostly agreed with LaMarca's results.  The
@@
 
+%TODO claims that it backed
+% TODO claims that it refuted or found problems with
+% - cache in heapsort
+% - algorithm N, can those improvements really be made
+
 Several discoveries were made and analysed: insertion sort has one misprediction
@@
 
+%TODO add novel contributions
+% - avoid copying back in mergesort
+% - remove bounds check without sentinel in base heapsort
+% - double-tiling
+
+%TODO look over these
 In addition, several ideas were conceived although not implemented, which relate

### 2005-07-29 docs/tech_report/heap.tex
added comparisons with lamarca to the conclusions

@@
 same amount of misses in both cases. Similarly, the results here indicate that
-base heapsort experiences a sharp increase in cache misses. This increase comes
-when the set size is twice that of cache mergesort when its sharp increase
-occurs. LaMarca's results do not show this.
+base heapsort experiences a sharp increase in cache misses. Cache heaport also
+has a similar increase, though LaMarca's results show that these occur at the
+same position in the graph, in other words, when th set size is the same. Our
+results, however, show that this occurs earlier in cache heapsort, due to the
+fact that it is an out of place sort.
 

### 2005-07-29 docs/tech_report/conclusions.tex
nearly finished

@@
 
-\subsection{Order N squared sorts}
+\subsection{Elementary Sorts}
+
+We found that insertion sort has the best performance of any of the elementary
+sorts, with fewer instructions, cache misses and exactly one branch
+misprediction per key. Selection sort also has very few branch mispredictions,
+and our results match Knuth's calculations. We find that reducing the number of
+runs in bubblesort and shakersort at the cost of additional instructions is
+counter productive, and that their branch prediction performance is incredibly
+poor.
+
 \subsection{Heapsorts}
+
+The cache optimized heaps perform better than the non-optimized versions. At the
+cost of a slightly increased instruction count, the number of branch misses can
+be greatly reduced, and the overall performance of our 8-heap is significantly
+better than the unoptimized version. This is despite our finding of that
+increasing the fanout increases branch mispredictions.
+
+Additionally, we created an improvement to heapsort which removed the need for a
+sentinel, without introducing any more instructions.
+
 \subsection{Mergesorts}
+
+We found that algorithm S has a lower instruction count than algorithm N, but
+that this does not necessarily translate to better performance. Our results show
+also that the level 1 cache needs to be taken into account, that the extra
+instruction in multi-mergesort make it slower than double tiled mergesort, and
+that multi-mergesort has regular branch predictive access patterns during the
+k-way merge, which can be taken advantage of by a 2-level adaptive predictor.
+
 \subsection{Quicksorts}
+
+Our results show base quicksort being the fastest of the quicksorts. We note our
+lowest instruction count occurs with the median-of-7 partitioning. We also see
+that using a binary search slows multi-quicksort down compared to a sequential
+search, but that its extra instructions make it slower than tuned quicksort.
+Despite being slower in our tests, pur simulations show tuned quicksort to have
+fewer cache misses, instructions and branch mispredictions than base quicksort.
+
 \subsection{Radixsorts}
+
+We were able to reduce the number of cache misses considerably using a technique
+adapted from LaMarca's tiled mergesort, and managed to make cache radixsort the
+fastest sort in our tests. This is due to its low instruction count, few cache
+misses and lack of branch mispredictions. Although we attempted to align it
+using another technique from mergesort, we found this decreased its performance.
+
 \subsection{Shellsorts}
 
-\section{Comparison with LaMarca's results}
+Based on our tests, shellsort performs very similarly to an $O(NlogN)$ sort,
+outperforming heapsort. We note that our improved version is considerably
+faster, mostly due to its reduced instruction count.
+
+We were also able to measure it's branch mispredictions, and determine a formula
+to calculate them.
+
+\section {Branch Results}
+
+No results for the branch prediction performance of sorting algorithm yet
+exists. We were able to find some interesting results.
+
+Insertion sort has exaclty one miss per key. Using this result we were able to
+make several other sorts faster by replacing their simple inline sorts with
+insertion sorts.
+
+Selection sort also has a very low number of branch mispredictions, the results
+of which match up well to Knuth's calculation. We found aswell that bubblesort
+and shakersort have incredibly bad branch prediction performance, bring
+$O(N^2)$.
+
+We found a pattern in branch misprediction when using a fanout in our 8-heap,
+and that this pattern matched closely that of selection sort, discussed above.
+
+TODO merge
+
+TODO quick
+
+TODO radix
+
+TODO shell
+
+%TODO results from software predictors
+
+\section{Comparison with LaMarca's Results}
 
@@
 
-\section{Instrucyion count results}
+\section{Instruction count results}
 

### 2005-07-30 docs/tech_report/method.tex
minor changes to references

@@
 \section{Testing details}
+
 %TODO a lot of this may be in the first sorts chapter, maybe heapsort.
-discuss: BIG + random.org, 64 bit in lamarca vs 32 bit, do_nothing and how it
-affects cache hits, scaling in the software branch predictors. How to compare
-and contrast varying results (branch vs softwre branhc; P4 vs simplescalar).
-What factors are used for testing: /proc/cpuinfo output, cahce and branch
-predictor data from P4, simplescalar inputs, software predictor, verfication of
-sorts.
+%discuss: BIG + random.org, 64 bit in lamarca vs 32 bit, do_nothing and how it
+%affects cache hits, scaling in the software branch predictors. How to compare
+%and contrast varying results (branch vs softwre branhc; P4 vs simplescalar).
+%What factors are used for testing: /proc/cpuinfo output, cahce and branch
+%predictor data from P4, simplescalar inputs, software predictor, verfication of
+%sorts.
 

### 2005-07-30 docs/tech_report/shell.tex
minor changes to references

@@
 slighlty steeper, as can be seen in Figures \ref{shell cycles}, \ref{shell
-instruction count}, \ref{shell level 1 misses} and \ref{shell branch
-mispredictions}. It performs better than a
+instructions}, \ref{shell level 1 misses} and \ref{shell branch
+misses}. It performs better than a
 heapsort, but considerably worse than quicksort. However, it easier to program
@@
 Overall, this leads to a reduction in cycle count of between 35 and 70\%, as
-seen in Figure \ref{shell cycle count}.
+seen in Figure \ref{shell cycles}.
 

### 2005-07-30 docs/tech_report/background.tex
minor changes to references
fixing up lots of things

@@
 
-\section{Big O notation}
+\section{Big O Notation}
 Big O notation is used to describe the complexity of an algorithm. It says that
@@
 expensive than quicksort, due to the cost of each step. These are discussed in
-later chapters.
-
-Not all algorithms have an easily derivable complexity. Many studies have
-attempted to tie down the complexity of shellsort, and it is not easily
-definable in terms of the catageories above (see \ref{TODO}).
+later chapters. The shellsort we use, called \n{Gonnet Shellsort} has a
+complexity of TODO (see \ref{TODO}).
 

### 2005-07-30 docs/tech_report/conclusions.tex
fixing up lots of things

@@
 \end{changemargin}
-\caption{\label{All cycles}Cycles per key of several major sorts and their variations - this was
+\caption[Cycles per key of several major sorts and their variations]{\label{All cycles}Cycles per key of several major sorts and their variations - this was
 measured on a Pentium 4 using hardware performance counters.}

### 2005-07-30 docs/tech_report/heap.tex
fixing up lots of things

@@
 
-\subsection{Implicit heaps}
+\subsection{Implicit Heaps}
 
@@
 
-\subsection{Sorting with heaps}
+\subsection{Sorting with Heaps}
 To sort an array with a heap, it is necessary to first create the heap. This is
@@
 
-\subsubsection{In-place sorting}
+\subsubsection{In-place Sorting}
 For an in-place sort, a heap is created as specified above. Then the root of the

### 2005-07-30 docs/tech_report/method.tex
fixing up lots of things

@@
 
-\section{Implementation of sorts}
+\section{Sort Implementation}
 
@@
 
-\section{Future work}
+\section{Future Work}
 An improvement that would have yielded interesting results would be to do simulations

### 2005-07-30 docs/tech_report/report.tex
minor changes to references
fixing up lots of things

@@
 \begin{changemargin}
-\subfigure[Instructions per key] {\includegraphics[scale=0.74]{plots/#1_-_Instructions.eps}}\\
+\subfigure[Cycles per key] {\includegraphics[scale=0.74]{plots/#1-_cycles.eps}}
+\subfigure[Instructions per key] {\includegraphics[scale=0.74]{plots/#1_-_Instructions.eps}}
 \subfigure[Level 1 cache misses per key]	{\includegraphics[scale=0.74]{plots/#1_-_Level_1_misses.eps}}
-\subfigure[Level 2 cache misses per key]	{\label{#1 level 2 misses}\includegraphics[scale=0.74]{plots/#1_-_Level_2_misses.eps}}
+\subfigure[Level 2 cache misses per key]	{\label{#2 level 2 misses}\includegraphics[scale=0.74]{plots/#1_-_Level_2_misses.eps}}
 \subfigure[Branches per key]				{\includegraphics[scale=0.74]{plots/#1_-_Branches.eps}}
@@
 \vspace*{6em}
-\caption{Branch prediction performance for #7. All graphs use the same scale.}
+\caption[Branch prediction performance for #7]{Branch prediction performance for #7. All graphs use the same scale.}
 \label{Branch prediction performance for #7}

### 2005-07-31 docs/tech_report/code/cycles_script.tex
done filename renaming inside files

--- 2005-06-10 fyp_report/code/cycles_script.tex
@@
 \L{\LB{\S{}{'}processed\_cycles\_data{'}\SE{}_\V{using}_\N{1}:\N{5}_\V{title}_\S{}{'}base_quicksort{'}\SE{},\2}}
-\L{\LB{\S{}{'}processed\_cycles\_data{'}\SE{}_\V{using}_\N{1}:\N{6}_\V{title}_\S{}{'}cache_quicksort{'}\SE{},\2}}
+\L{\LB{\S{}{'}processed\_cycles\_data{'}\SE{}_\V{using}_\N{1}:\N{6}_\V{title}_\S{}{'}memory_tuned_quicksort{'}\SE{},\2}}
 \L{\LB{\S{}{'}processed\_cycles\_data{'}\SE{}_\V{using}_\N{1}:\N{7}_\V{title}_\S{}{'}multi_quicksort{'}\SE{},\2}}
 \L{\LB{\S{}{'}processed\_cycles\_data{'}\SE{}_\V{using}_\N{1}:\N{10}_\V{title}_\S{}{'}base_mergesort{'}\SE{},\2}}
-\L{\LB{\S{}{'}processed\_cycles\_data{'}\SE{}_\V{using}_\N{1}:\N{14}_\V{title}_\S{}{'}double_tiled_mergesort{'}\SE{},\2}}
+\L{\LB{\S{}{'}processed\_cycles\_data{'}\SE{}_\V{using}_\N{1}:\N{14}_\V{title}_\S{}{'}double_aligned_mergesort{'}\SE{},\2}}
 \L{\LB{\S{}{'}processed\_cycles\_data{'}\SE{}_\V{using}_\N{1}:\N{16}_\V{title}_\S{}{'}double_multi\-mergesort{'}\SE{},\2}}

### 2005-07-31 docs/tech_report/merge.tex
added comparisons with lamarca to the conclusions
fixing up lots of things
done filename renaming inside files

@@
 
-\section{Predictor results}
+\section{Predictor Results}
 
-\sixplot{algorithm_n_0.eps}{algorithm_s_0.eps}{base_mergesort_0.eps}{double_tiled_mergesort_0.eps}{double_multi_mergesort_0.eps}{double_multi_mergesort_0.eps}{mergesorts}
+\sixplot{algorithm_n_0.eps}{algorithm_s_0.eps}{base_mergesort_0.eps}{double_aligned_mergesort_0.eps}{double_aligned_multi_mergsort_0.eps}{double_aligned_multi_mergsort_0.eps}{mergesorts}
 
@@
 is used instead.
-%TODO ohw much better does this do? if not, move to other section
 
@@
 
-\section{Future work}
+\section{Future Work}
 Several avenues could be pursued in order to improve the algorithms presented

### 2005-07-31 docs/tech_report/quick.tex
minor edit
fixing up lots of things
done filename renaming inside files

@@
 
-\section{Predictor results}
+\section{Predictor Results}
 
 %TODO a bit to rewrite
-\subsection{Different sorts}
+\subsection{Different Sorts}
 
-\sixplot{base_quicksort_0.eps}{base_quicksort9_0.eps}{cache_quicksort_0.eps}{multi_quicksort_0.eps}{sequential_multi_quicksort_0.eps}{}{base, cache and multi quicksorts and base quicksort9}
+\sixplot{base_quicksort_0.eps}{base_quicksort9_0.eps}{memory_tuned_quicksort_0.eps}{multi_quicksort_0.eps}{sequential_multi_quicksort_0.eps}{}{base, cache and multi quicksorts and base quicksort9}
 
-Firstly, cache quicksort and base quicksort only idffer in how they deal with
+Firstly, cache quicksort and base quicksort only differ in how they deal with
 the insertion. Base quicksort and base quicksort9 have similar numbers of
@@
 
-%TO here
+%TODO to  here
 
-\subsection{Different medians}
+\subsection{Different Medians}
 

### 2005-07-31 docs/tech_report/result_list.tex
fixing up lots of things
done filename renaming inside files

@@
 \afterpage{
-\smallplot{on2_insertsort}{insertion sort}{}{}{}{}{}
-\smallplot{on2_selectsort}{selection sort}{}{}{}{}{}
-\smallplot{on2_bubblesort}{bubblesort}{}{}{}{}{}
-\smallplot{on2_bubblesort2}{improved bubblesort}{}{}{}{}{}
-\smallplot{on2_shakersort}{shakersort}{}{}{}{}{}
-\smallplot{on2_shakersort2}{improved shakersort}{}{}{}{}{}
+\smallplot{insertsort}{insertion sort}{}{}{}{}{}
+\smallplot{selectsort}{selection sort}{}{}{}{}{}
+\smallplot{bubblesort}{bubblesort}{}{}{}{}{}
+\smallplot{bubblesort2}{improved bubblesort}{}{}{}{}{}
+\smallplot{shakersort}{shakersort}{}{}{}{}{}
+\smallplot{shakersort2}{improved shakersort}{}{}{}{}{}
 %
 \smallplot{base_heapsort}{base heapsort}{}{}{}{}{}
-\smallplot{cache4_heapsort}{cache heapsort (4-heap)}{}{}{}{}{}
-\smallplot{cache_heapsort}{cache heapsort (8-heap)}{}{}{}{}{}
+\smallplot{memory_tuned4_heapsort}{cache heapsort (4-heap)}{}{}{}{}{}
+\smallplot{memory_tuned_heapsort}{cache heapsort (8-heap)}{}{}{}{}{}
 %
-\smallplot{algorithm_n}{algorithm N}{}{}{}{}{}
-\smallplot{algorithm_s}{algorithm S}{}{}{}{}{}
+\smallplot{algorithm_n}{algorithm n}{}{}{}{}{}
+\smallplot{algorithm_s}{algorithm s}{}{}{}{}{}
 \smallplot{base_mergesort}{base mergesort}{}{}{}{}{}
@@
 \smallplot{multi_mergesort}{multi-mergesort}{}{}{}{}{}
-\smallplot{double_tiled_mergesort}{double tiled mergesort}{}{}{}{}{}
-\smallplot{double_multi_mergesort}{double multi-mergesort}{}{}{}{}{}
+\smallplot{double_aligned_mergesort}{double tiled mergesort}{}{}{}{}{}
+\smallplot{double_aligned_multi_mergsort}{double multi-mergesort}{}{}{}{}{}
 %
@@
 \smallplot{base_quicksort9}{base quicksort (pseudo-median-of-9)}{}{}{}{}{}
-\smallplot{cache_quicksort}{memory-tuned quicksort}{}{}{}{}{}
+\smallplot{memory_tuned_quicksort}{memory-tuned quicksort}{}{}{}{}{}
 \smallplot{multi_quicksort}{multi-quicksort (binary search)}{}{}{}{}{}
@@
 \smallplot{base_radixsort}{base radixsort}{}{}{}{}{}
-\smallplot{cache_radixsort}{cache radixsort}{}{}{}{}{}
-\smallplot{aligned_cache_radixsort}{aligned cache radixsort}{}{}{}{}{}
+\smallplot{memory_tuned_radixsort}{cache radixsort}{}{}{}{}{}
+\smallplot{aligned_memory_tuned_radixsort}{aligned cache radixsort}{}{}{}{}{}
 %

### 2005-07-31 docs/tech_report/software_predictor.tex
removed word gonnet from document
done filename renaming inside files

@@
 \afterpage{
-\sixplot{base_heapsort_0.eps}{cache4_heapsort_0.eps}{cache4_heapsort_1.eps}{cache_heapsort_0.eps}{cache_heapsort_1.eps}{}{base and cache heapsorts}
+\sixplot{base_heapsort_0.eps}{memory_tuned4_heapsort_0.eps}{memory_tuned4_heapsort_1.eps}{memory_tuned_heapsort_0.eps}{memory_tuned_heapsort_1.eps}{}{base and cache heapsorts}
 
-\sixplot{base_quicksort_0.eps}{base_quicksort9_0.eps}{cache_quicksort_0.eps}{multi_quicksort_0.eps}{sequential_multi_quicksort_0.eps}{}{base, cache and multi quicksorts and base quicksort9}
+\sixplot{base_quicksort_0.eps}{base_quicksort9_0.eps}{memory_tuned_quicksort_0.eps}{multi_quicksort_0.eps}{sequential_multi_quicksort_0.eps}{}{base, cache and multi quicksorts and base quicksort9}
 \sixplot{base_quicksort1_0.eps}{base_quicksort_0.eps}{base_quicksort5_0.eps}{base_quicksort7_0.eps}{base_quicksort9_0.eps}{}{base quicksort without a median, with median of 3, and with pseudo median of 5, 7 and 9}
@@
 \sixplot{base_mergesort_0.eps}{base_mergesort_1.eps}{base_mergesort_2.eps}{base_mergesort_3.eps}{base_mergesort_4.eps}{base_mergesort_5.eps}{base mergesort}
-\sixplot{double_tiled_mergesort_0.eps}{double_tiled_mergesort_1.eps}{double_tiled_mergesort_2.eps}{double_tiled_mergesort_3.eps}{double_tiled_mergesort_4.eps}{double_tiled_mergesort_5.eps}{double tiled mergesort - part 1}
-\sixplot{double_tiled_mergesort_6.eps}{double_tiled_mergesort_7.eps}{double_multi_mergesort_0.eps}{double_multi_mergesort_1.eps}{double_multi_mergesort_2.eps}{double_multi_mergesort_3.eps}{double tiled mergesort - part 2, and double multi mergesort - part 1}
-\sixplot{double_multi_mergesort_4.eps}{double_multi_mergesort_5.eps}{double_multi_mergesort_6.eps}{double_multi_mergesort_7.eps}{double_multi_mergesort_8.eps}{double_multi_mergesort_9.eps}{double multi-mergesort - part 2}
+\sixplot{double_aligned_mergesort_0.eps}{double_aligned_mergesort_1.eps}{double_aligned_mergesort_2.eps}{double_aligned_mergesort_3.eps}{double_aligned_mergesort_4.eps}{double_aligned_mergesort_5.eps}{double tiled mergesort - part 1}
+\sixplot{double_aligned_mergesort_6.eps}{double_aligned_mergesort_7.eps}{double_aligned_multi_mergsort_0.eps}{double_aligned_multi_mergsort_1.eps}{double_aligned_multi_mergsort_2.eps}{double_aligned_multi_mergsort_3.eps}{double tiled mergesort - part 2, and double multi mergesort - part 1}
+\sixplot{double_aligned_multi_mergsort_4.eps}{double_aligned_multi_mergsort_5.eps}{double_aligned_multi_mergsort_6.eps}{double_aligned_multi_mergsort_7.eps}{double_aligned_multi_mergsort_8.eps}{double_aligned_multi_mergsort_9.eps}{double multi-mergesort - part 2}
 
-\sixplot{gonnet_shellsort_0.eps}{gonnet_improved_shellsort_0.eps}{}{}{}{}{Shellsorts}
+\sixplot{shellsort_0.eps}{improved_shellsort_0.eps}{}{}{}{}{Shellsorts}
 }

### 2005-07-31 code/sorts/aligned_memory_tuned_radixsort.c

@@ -17,7 +17,7 @@
 #define digit(A, B) (((A) >> (BITSWORD-((B)+1)*BITSBYTE)) & (R-1))
 
 void
-aligned_cache_radixsort(unsigned int a[], int N)
+aligned_memory_tuned_radixsort(unsigned int a[], int N)
 {
 	unsigned int* aux_data;
 	unsigned int* aux;

### 2005-07-31 code/sorts/double_aligned_multi_mergsort.c

@@ -211,7 +211,7 @@ static int get_count(const int N)
 }
 
 void
-double_multi_mergesort(unsigned int a[], int N)
+double_aligned_multi_mergsort(unsigned int a[], int N)
 {
 	/* track keeps an eye on where i started last */
 	/* outer track keeps and eye on which cache_sized/2 segemnt i started in last */

### 2005-07-31 code/sorts/improved_bubblesort.c

@@ -1,7 +1,7 @@
 #include "on2_sorts.h"
 
 void
-on2_bubblesort2(unsigned int a[], int N)
+bubblesort2(unsigned int a[], int N)
 {
 	int i = 0, j, k = 0;
 

### 2005-07-31 code/sorts/insertsort.c

@@ -1,7 +1,7 @@
 #include "insertion.c"
 
 void
-on2_insertsort(unsigned int a[], int N)
+insertsort(unsigned int a[], int N)
 {
 	describe_predictor(&global_predictor[0], "Comparator");
 	describe_predictor(&global_predictor[1], "Sentinel comparator");

### 2005-07-31 code/sorts/memory_tuned_heapsort.c

@@ -149,7 +149,7 @@ fix_down(unsigned int heap[], int parent, int N)
 }
 
 void
-cache_heapsort(unsigned int a[], int N)
+memory_tuned_heapsort(unsigned int a[], int N)
 {
 	int start_padding, sentinel_count, size_to_allocate, i, length;
 	unsigned int *heap ,*heap_data;

### 2005-07-31 code/sorts/memory_tuned_quicksort.c

@@ -66,7 +66,7 @@ insertion(unsigned int a[], int N)
 }
 
 void
-cache_quicksort(unsigned int a[], int N)
+memory_tuned_quicksort(unsigned int a[], int N)
 {
 	int l, r;
 	int i;

### 2005-07-31 code/sorts/memory_tuned_radixsort.c

@@ -13,7 +13,7 @@
 #define digit(A, B) (((A) >> (BITSWORD-((B)+1)*BITSBYTE)) & (R-1))
 
 void
-cache_radixsort(unsigned int a[], int N)
+memory_tuned_radixsort(unsigned int a[], int N)
 {
 	unsigned int* aux = malloc(N * sizeof(unsigned int));
 	int i, j, w; /* R is the size of a digit, hence the number of bins */

### 2005-07-31 docs/tech_report/code/main.c

@@ -42,5 +42,5 @@ test_sort(void sort(Item*, int), char* description) {
 	}
 	printf("Testing of %s complete\n", description);
 }
-test_sort(on2_insertsort, "O(N squared) Insertion");
-time_sort(on2_insertsort, "O(N squared) Insertion");
+test_sort(insertsort, "O(N squared) Insertion");
+time_sort(insertsort, "O(N squared) Insertion");

### 2005-07-31 docs/tech_report/conclusions.tex
the grand reaming is almost complete

@@
 also that the level 1 cache needs to be taken into account, that the extra
-instruction in multi-mergesort make it slower than double tiled mergesort, and
-that multi-mergesort has regular branch predictive access patterns during the
-k-way merge, which can be taken advantage of by a 2-level adaptive predictor.
+instruction in multi-mergesort make it slower than double-aligned tiled
+mergesort, and that multi-mergesort has regular branch predictive access
+patterns during the k-way merge, which can be taken advantage of by a 2-level
+adaptive predictor.
 
@@
 We were able to reduce the number of cache misses considerably using a technique
-adapted from LaMarca's tiled mergesort, and managed to make cache radixsort the
+adapted from LaMarca's tiled mergesort, and managed to make memory-tuned radixsort the
 fastest sort in our tests. This is due to its low instruction count, few cache
@@
 heap lowered our instruction count. Our level 2 misses also concurred with
-LaMarca, with only one difference: cache heapsort, being an out-of-place sort,
+LaMarca, with only one difference: memory-tuned heapsort, being an out-of-place sort,
 fills the cache fully with a smaller number of keys than base heapsort, which is
@@
 improvements due to cache performance and instruction count, our base quicksort
-performed faster than our tuned version. Finally, we found that although
+performed faster than our memory-tuned version. Finally, we found that although
 changing multi-mergesort's sequential search to a binary search decreased the
@@
 The best stable sort is radixsort. The best comparison-based stable sort is
-double tiled mergesort. The best in-place sort is quicksort. The best in-place
+double-aligned tiled mergesort. The best in-place sort is quicksort. The best in-place
 stable sort is, sadly, insertion sort. The best elementary sort is also

### 2005-07-31 docs/tech_report/heap.tex
done filename renaming inside files
the grand reaming is almost complete

@@
 
-\section{Cache Heapsort}
-\label{cache heapsort}
+\section{Memory-tuned Heapsort}
+\label{memory-tuned heapsort}
 LaMarca notes that the Repeated-Adds algorithm has a much lower number of cache
@@
 sort becomes out-of-place. For this reason, the Repeated-Adds algorithm is used
-for cache heapsort.
+for memory-tuned heapsort.
 
@@
 LaMarca reports base heapsort has very poor temporal locality, and that the
-cache heapsort will have half the cache misses of base heapsort.
+memory-tuned heapsort will have half the cache misses of base heapsort.
 
@@
 same position in the graph, in other words, when th set size is the same. Our
-results, however, show that this occurs earlier in cache heapsort, due to the
+results, however, show that this occurs earlier in memory-tuned heapsort, due to the
 fact that it is an out of place sort.
@@
 Finally, though this is likely to be related the previous differences, base
-heapsort performs slightly better for small sets than the cache heapsorts do in
-the results presented here. LaMarca's results indicate that the cache heapsort
+heapsort performs slightly better for small sets than the memory-tuned heapsort's do in
+the results presented here. LaMarca's results indicate that the memory-tuned heapsort
 has better results the entire way through the cache. This may relate to the
@@
 
-\sixplot{base_heapsort_0.eps}{cache4_heapsort_0.eps}{cache_heapsort_0.eps}{cache_heapsort_1.eps}{}{}{base and cache heapsorts}
+\sixplot{base_heapsort_0.eps}{memory_tuned4_heapsort_0.eps}{memory_tuned_heapsort_0.eps}{memory_tuned_heapsort_1.eps}{}{}{base and memory-tuned heapsorts}
 
@@
 in the creation and destruction of the heap. The base heapsort results in
-Figure \ref{base and cache heapsorts-1} show that the time taken to create the
-heap is not significant relative to the time taken to sort from the heap.
+Figure \ref{base and memory-tuned heapsorts-1} show that the time taken to
+create the heap is not significant relative to the time taken to sort from the
+heap.
 
@@
 columns marked `\cc{Destruction, child comparison}' and `\cc{Destruction, child
-promotion}' in Figure \ref{base and cache heapsorts-1} (the 2-heap), with the
-columns marked `\cc{comparison 1}' to `\cc{child promotion}' in Figure \ref{base and
-cache heapsorts-2} (the 4-heap) and Figures \ref{base and cache heapsorts-3} and
-\ref{base and cache heapsorts-4} (the 8-heap).
+promotion}' in Figure \ref{base and memory-tuned heapsort-1} (the 2-heap), with
+the columns marked `\cc{comparison 1}' to `\cc{child promotion}' in Figure
+\ref{base and memory-tuned heapsorts-2} (the 4-heap) and Figures \ref{base and
+memory-tuned heapsort-3} and \ref{base and memory-tuned heapsort-4} (the
+8-heap).
 
-Though the number of times each comparison is used shrinks
-as the size of the fanout increased, overall the number of branches increases.
-In total, there are 50\% more comparisons in the 4-heap than the 2-heap, and
-150\% more in the 8-heap.
+Though the number of times each comparison is used shrinks as the size of the
+fanout increased, overall the number of branches increases.  In total, there are
+50\% more comparisons in the 4-heap than the 2-heap, and 150\% more in the
+8-heap.
 
-More importantly, the accuracy of the each
-successive comparison increases in the same way as selection sort (see Section
-\ref{selection branches logn}), becoming more predictable as the fanout
-increases. For example, in the 8-heap, shown in Figures \ref{base and cache heapsorts-3}
-and \ref{base and cache heapsorts-4},
-the first comparison is predicted correctly 50\% of the
-time, while the next is predicted correctly 61\% of the time, then 71\%, 77\%, and
-so on in a curve, which is shown in Figure \ref{cache heapsort curve}.
+More importantly, the accuracy of the each successive comparison increases in
+the same way as selection sort (see Section \ref{selection branches logn}),
+becoming more predictable as the fanout increases. For example, in the 8-heap,
+shown in Figures \ref{base and memory-tuned heapsort-3} and \ref{base and
+memory-tuned heapsort-4}, the first comparison is predicted correctly 50\% of
+the time, while the next is predicted correctly 61\% of the time, then 71\%,
+77\%, and so on in a curve, which is shown in Figure \ref{memory-tuned heapsort
+curve}.
 
@@
 \includegraphics{plots/curve.eps}
-\caption{Predictability of branches in cache heapsort, with a fanout of eight}
-\label{cache heapsort curve}
+\caption{Predictability of branches in memory-tuned heapsort, with a fanout of eight}
+\label{memory-tuned heapsort curve}
 \end{figure}

### 2005-07-31 docs/tech_report/merge.tex
the grand reaming is almost complete

@@
 
-\section{Double Tiled Mergesort}
-\label{double tiled mergesort}
+\section{Double-aligned Tiled Mergesort}
+\label{double-aligned tiled mergesort}
 Offsetting the target array to avoid conflicts reduces significantly the number
@@
 LaMarca used a 4-heap for his priority queue, and an 8-heap is used here, for
-the same reasons as before (see Section \ref{cache heapsort}.). The heap is
-aligned, and a cache line is put into the array each time. The smallest key is
-at the top of the heap, and this is put into its final position in the array.
+the same reasons as before (see Section \ref{memory-tuned heapsort}.). The heap
+is aligned, and a cache line is put into the array each time. The smallest key
+is at the top of the heap, and this is put into its final position in the array.
 The heap is then fixed.
@@
 pre-sorted, they were pre-sorted to a size four, or eight if the number of
-passes was going to be odd. These were reversed for the double multi-mergesort;
-since it does a k-way merge into the source array, the sorted cache sized
-segments need to be in the auxiliary array. An 8-heap was used for the
-multi-mergesorts.
+passes was going to be odd. These were reversed for the double-aligned
+multi-mergesort; since it does a k-way merge into the source array, the sorted
+cache sized segments need to be in the auxiliary array. An 8-heap was used for
+the multi-mergesorts.
 
@@
 cycle count on the Pentium 4 (see Figure \ref{merge cycles}) as a result. The
-other sorts generally perform in $O(NlogN)$ time, with double tiled mergesort
+other sorts generally perform in $O(NlogN)$ time, with double-aligned tiled mergesort
 being the fastest. Note that the alignment of the mergesort may not work as
@@
 The cache misses end up being the deciding factor between the performance of
-double tiled mergesort and base mergesort, the next best contender. Despite the
-reduction in level 2 cache misses, the sharp increase in level 1 misses result
-is poor performance for tiled and multi-mergesort. Multi-mergesort is further
-encumbered with a much higher instruction count. Double tiled mergesort, with
-its low instruction count, and low cache miss rate in both levels 1 and 2,
-emerges the winner.
+double-aligned tiled mergesort and base mergesort, the next best contender.
+Despite the reduction in level 2 cache misses, the sharp increase in level 1
+misses result is poor performance for tiled and multi-mergesort. Multi-mergesort
+is further encumbered with a much higher instruction count. Double tiled
+mergesort, with its low instruction count, and low cache miss rate in both
+levels 1 and 2, emerges the winner.
 
@@
 
-In both double tiled mergesort and base mergesort, approximately two thirds are
+In both double-aligned tiled mergesort and base mergesort, approximately two thirds are
 of the loops are taken, unlike Knuth's sorts. However, this appears to indicate
@@
 instruction count).  The rate of branch misses is much the same as that in
-heapsort, and progresses in the same manner (see Figure \ref{cache heapsort
+heapsort, and progresses in the same manner (see Figure \ref{memory-tuned heapsort (8-heap)
 curve}). Note that \n{comparison 0} occurs every time, and would be optimized

### 2005-07-31 docs/tech_report/ordernsquared.tex
minor changes as rereading
fixing up lots of things
done filename renaming inside files
the grand reaming is almost complete

@@
 
-\section{Selection sort}
+\section{Selection Sort}
 
@@
 
-\code{Selection sort code}{on2_selectsort.c}
+\code{Selection sort code}{selectsort.c}
 \subsection{Testing}
@@
 
-\section{Insertion sort}
+\section{Insertion Sort}
 
@@
 
-\code{Insertion sort code}{on2_insertsort.c}
+\code{Insertion sort code}{insertsort.c}
 \subsection{Testing}
@@
 
-\code{Bubblesort code}{on2_bubblesort.c}
+\code{Bubblesort code}{bubblesort.c}
 \subsection{Testing}
@@
 
-\section{Bubblesort2}
-Bubblesort2 incorporates an improvement to end bubblesort early. The bubblesort
-implementation described in the previous section ends early if it detects that
-the array is fully sorted.  It keeps a binary flag to indicate if there were any
-exchanges along the way.  Instead, bubblesort2 keeps track of the last exchange
-made, and starts the next iteration from there. This means that any pockets of
-sorted keys are skipped over, not just if they occur at the end. Bubblesort2
-code is shown in Figure \vref{Bubblesort2 code}.
+\section{Improved Bubblesort}
+Improved bubblesort incorporates an improvement to end bubblesort early. The
+bubblesort implementation described in the previous section ends early if it
+detects that the array is fully sorted.  It keeps a binary flag to indicate if
+there were any exchanges along the way.  Instead, improved bubblesort keeps
+track of the last exchange made, and starts the next iteration from there. This
+means that any pockets of sorted keys are skipped over, not just if they occur
+at the end. Improved bubblesort code is shown in Figure \vref{Improved bubblesort code}.
 
-\code{Bubblesort2 code}{on2_bubblesort2.c}
+\code{Improved bubblesort code}{improved_bubblesort.c}
 \subsection{Testing}
 \subsubsection{Expected Performance}
-It is expected that the bubblesort2 should perform faster than the original in
-all metrics. Instruction count, were there no conditions that allowed keys to be
-skipped, would be the similar, since few extra instructions are added. The flow
-of the algorithm is the same, though there should be fewer iterations,
+It is expected that improved bubblesort should perform faster than the original
+in all metrics. Instruction count, were there no conditions that allowed keys to
+be skipped, would be the similar, since few extra instructions are added. The
+flow of the algorithm is the same, though there should be fewer iterations,
 especially at the start, when iterations are longer, so cache accesses, and
@@
 
-\code{Shakersort code}{on2_shakersort.c}
+\code{Shakersort code}{shakersort.c}
 \subsection{Testing}
@@
 
-\section{Shakersort2}
-Shakersorts incorporates the same improvements over shakersort as bubblesort2
+\section{Improved Shakersort}
+Shakersorts incorporates the same improvements over shakersort as improved bubblesort
 does over bubblesort, that is, it skips over pockets of sorted keys in both
-directions, as well as ending early. The code for shakersort2 is in Figure
-\vref{Shakersort2 code}.
+directions, as well as ending early. The code for improved shakersort is in Figure
+\vref{Improved shakersort code}.
 
-\code{Shakersort2 code}{on2_shakersort2.c}
+\code{Improved shakersort code}{improved_shakersort.c}
 \subsection{Testing}
@@
 selection sort, but shakersort has a significant number less. The improvement to
-bubblesort in bubblesort2 have increased the instruction count, contrary to
-predictions, and similarly in the shakersort2. It appears the cost of
+bubblesort in improved bubblesort have increased the instruction count, contrary to
+predictions, and similarly in the improved shakersort. It appears the cost of
 skipping over a segment of the array is not worth the additional book-keeping
@@
 place. Bubblesort, with similar behaviour, is only slightly better. Shakersort
-has a small amount of temporal reuse, reducing its number of misses. Shakersort2
-is placed roughly between insertion sort and shakersort, showing that ending
-early is helpful in its case. This does not appear to be the case with
-bubblesort2 though, which performs identically to bubblesort.
+has a small amount of temporal reuse, reducing its number of misses. Improved
+shakersort is placed roughly between insertion sort and shakersort, showing that
+ending early is helpful in its case. This does not appear to be the case with
+improved bubblesort though, which performs identically to bubblesort.
 
@@
 many as selection and bubblesort, and shakersort being half-way between the
-two. The number of misses for bubblesort and shakersort are huge; nearly one in
-four comparisons are misses. As the data sets gets large, there are nearly
-10,000 branch misses involved in sorting a single key. The improved bubblesort
-had exactly the same number of misses and branches. The improved shakersort has
-a lot less branches, but the same number of misses. This is because skipping
-over sorted keys is predictable from the first branch miss.
+two. The number of misses for bubblesort and shakersort are huge.  As the data
+sets gets large, there are nearly 10,000 branch misses involved in sorting a
+single key. The improved bubblesort had exactly the same number of misses and
+branches. The improved shakersort has a lot less branches, but the same number
+of misses. This is because skipping over sorted keys is predictable from the
+first branch miss.
 

### 2005-07-31 docs/tech_report/quick.tex
the grand reaming is almost complete

@@
 
-Cache quicksort should have fewer misses than base quicksort as soon as the
-array is larger than the cache. It is expected that these misses will make up
-for the increase in instruction count.
+Memory-tuned quicksort should have fewer misses than base quicksort as soon as
+the array is larger than the cache. It is expected that these misses will make
+up for the increase in instruction count.
 
@@
 
-The instruction count, shown in Figure \ref{quick instructions} is moslty as
-predicted. The increase in instructions in both multi-mergesorts is as
+The instruction count, shown in Figure \ref{quick instructions} is mostly as
+predicted. The increase in instructions in both multi-quicksorts is as
 predicted, with the binary search version increasing only 25\% against the
 sequential search version's 50\% increase. However, the increase we predicted in
-the instrcution count of cache quicksort based on the results of both Sedgewick
+the instruction count of memory-tuned quicksort based on the results of both Sedgewick
 and LaMarca -  that is, once the efficient insertion sort was replaced with many
@@
 
-The tuned quicksort performed better than the base version, as predicted. Both the
-tuned version and the base quicksort both have a significant increase once the
-arrays no longer fit in the cache, also as predicted.
+The memory-tuned quicksort performed better than the base version, as predicted.
+Both the memory-tuned version and the base quicksort both have a significant
+increase once the arrays no longer fit in the cache, also as predicted.
 
@@
 associative ones. This is true in the case of all the quicksorts, as can be seen
-in Figures \ref{base quicksort (no median) level 2 misses} to
+in Figures \ref{base quicksort (median-of-3) (no median) level 2 misses} to
 \ref{multi-quicksort (sequential search) level 2 misses} graph that this is true
@@
 
-Tuned quicksort had fewer branch misses than base quicksort, due to the fact
+Memory-tuned quicksort had fewer branch misses than base quicksort, due to the fact
 that fewer comparitive branches are executed. In base quicksort, we sort the
@@
 but that of the cache misses is not. The ratio of the cache misses between base
-and cache quicksort are the same in both sets of results, but these are lower
+and memory-tuned quicksort are the same in both sets of results, but these are lower
 than the same misses in LaMarca's results. However, the multi-quicksort results
@@
 
-\sixplot{base_quicksort_0.eps}{base_quicksort9_0.eps}{memory_tuned_quicksort_0.eps}{multi_quicksort_0.eps}{sequential_multi_quicksort_0.eps}{}{base, cache and multi quicksorts and base quicksort9}
+\sixplot{base_quicksort_0.eps}{base_quicksort9_0.eps}{memory_tuned_quicksort_0.eps}{multi_quicksort_0.eps}{sequential_multi_quicksort_0.eps}{}{base,
+cache and multi quicksorts and base quicksort (pseudo-median-of-9)}
 
-Firstly, cache quicksort and base quicksort only differ in how they deal with
-the insertion. Base quicksort and base quicksort9 have similar numbers of
-insertion branches, and a very similar ratio (2/1 correct/incorrect) of
-mispredictions. Comparing base quicksort to cahce quicksort (they both use
+Firstly, memory-tuned quicksort and base quicksort (median-of-3) only differ in how they deal with
+the insertion. Base quicksort (median-of-3) and base quicksort
+(pseudo-median-of-9) have similar numbers of insertion branches, and a very
+similar ratio (2/1 correct/incorrect) of mispredictions. Comparing base
+quicksort to memory-quicksort (they both use
 median of 3), there is actually a decrease in the number of branches handled by
@@
 
-Note that sequential multi-quicksort has less mispredictions than cache quicksort.
+Note that sequential multi-quicksort has less mispredictions than memory-tuned quicksort.
 
@@
 \label{quick-predictors2}
-\sixplot{base_quicksort1_0.eps}{base_quicksort_0.eps}{base_quicksort5_0.eps}{base_quicksort7_0.eps}{base_quicksort9_0.eps}{}{base quicksort without a median, with median of 3, and with pseudo medians of 5, 7 and 9}
+\sixplot{base_quicksort1_0.eps}{base_quicksort_0.eps}{base_quicksort5_0.eps}{base_quicksort7_0.eps}{base_quicksort9_0.eps}{}{base quicksort without a median, with median of 3, and with pseudo-medians of 5, 7 and 9}
 
@@
 We use several variations of quicksort to test this theory. \ref{bently, i
-ithkn} suggests partitioning using a median-of-9 instead of a median-of-three. We
+ithkn} suggests partitioning using a median-of-9 instead of a median-of-3. We
 created sorts using no median, a median-of-3, and pseudo-medians-of-five, -seven and
@@
 This indicates the same pattern as we saw during heapsort.%
-% and mergesort?
+% and mergesort? TODO
 The less entropy due to the partitioning key, the less wasted exchanging of keys,
@@
 
-The pseudo-median-of-five partitioning performs slightly better than the
-median-of-three, and pseudo-median-of-seven performs slightly better again, in
+The pseudo-median-of-5 partitioning performs slightly better than the
+median-of-3, and pseudo-median-of-7 performs slightly better again, in
 terms of branch prediction. Figure \ref{TODO} shows the cycle count of each of
@@
 pivots to be searched by the sequential search. The most this can be reduced by
-is a factor of 3, indicating perfectly distributed pivots. This would reduce the
-instruction overhead of the multi-quicksort considerably.
+is a factor of three, indicating perfectly distributed pivots. This would reduce
+the instruction overhead of the multi-quicksort considerably.

### 2005-07-31 docs/tech_report/radix.tex
fixing up lots of things
the grand reaming is almost complete

@@
 
-\subsection{Aligned Cache Radixsort}
+\subsection{Aligned Memory-tuned Radixsort}
 In Section \ref{tiled mergesort}, mergesort is improved by aligning its array
@@
 arrays are exactly half the length of the cache apart. We further added an
-alignment for the level 1 cache (see Section \ref{double tiled mergesort}) to
-significantly reduce level 1 misses at the expense of a small number of level 2
-misses. We adapt this strategy here, and call it Aligned Cache Radixsort.
+alignment for the level 1 cache (see Section \ref{double-aligned tiled
+mergesort}) to significantly reduce level 1 misses at the expense of a small
+number of level 2 misses. We adapt this strategy here, and call it Aligned Cache
+Radixsort.
 
 \section{Results}
-\subsection{Test parameters}
+\subsection{Test Parameters}
 Keys are to be divided into segments 8 bits long, resulting in four sorting
@@
 
-Due to the reduced copying, cache radixsort should have a lower instruction
+Due to the reduced copying, memory-tuned radixsort should have a lower instruction
 count and fewer cache misses that base radixsort. There should be nine passes
@@
 Based on LaMarca's results, showing quicksort to be faster than radixsort by a
-small amount, we expect cache radixsort to run faster than quicksort.
+small amount, we expect memory-tuned radixsort to run faster than quicksort.
 
-\subsection{Simulation results}
+\subsection{Simulation Results}
 \plot{radix}{}{}{}{}{}{}
@@
 
-Aligned cache radixsort performs exactly the same as cache radixsort in both
+Aligned memory-tuned radixsort performs exactly the same as memory-tuned radixsort in both
 level 1 and level 2 misses, indicating that the alignment, and the associated
@@
 With the low number of cache misses, low instruction count and negligible branch
-mispredictions, it is no surprise that cache radixsort performs as well as it
-does. As shown in Figure \ref{radix cycles}, cache radixsorts performance is
+mispredictions, it is no surprise that memory-tuned radixsort performs as well as it
+does. As shown in Figure \ref{radix cycles}, memory-tuned radixsorts performance is
 almost linear, with only a slight increase when the arrays no longer fit in the

### 2005-07-31 docs/tech_report/result_list.tex
the grand reaming is almost complete

@@
 \smallplot{bubblesort}{bubblesort}{}{}{}{}{}
-\smallplot{bubblesort2}{improved bubblesort}{}{}{}{}{}
+\smallplot{improved_bubblesort}{improved bubblesort}{}{}{}{}{}
 \smallplot{shakersort}{shakersort}{}{}{}{}{}
-\smallplot{shakersort2}{improved shakersort}{}{}{}{}{}
+\smallplot{improved_shakersort}{improved shakersort}{}{}{}{}{}
 %
 \smallplot{base_heapsort}{base heapsort}{}{}{}{}{}
-\smallplot{memory_tuned4_heapsort}{cache heapsort (4-heap)}{}{}{}{}{}
-\smallplot{memory_tuned_heapsort}{cache heapsort (8-heap)}{}{}{}{}{}
+\smallplot{memory_tuned4_heapsort}{memory-tuned heapsort (4-heap)}{}{}{}{}{}
+\smallplot{memory_tuned_heapsort}{memory-tuned heapsort (8-heap)}{}{}{}{}{}
 %
@@
 \smallplot{multi_mergesort}{multi-mergesort}{}{}{}{}{}
-\smallplot{double_aligned_mergesort}{double tiled mergesort}{}{}{}{}{}
-\smallplot{double_aligned_multi_mergsort}{double multi-mergesort}{}{}{}{}{}
+\smallplot{double_aligned_mergesort}{double-aligned tiled mergesort}{}{}{}{}{}
+\smallplot{double_aligned_multi_mergsort}{double-aligned multi-mergesort}{}{}{}{}{}
 %
@@
 \smallplot{base_radixsort}{base radixsort}{}{}{}{}{}
-\smallplot{memory_tuned_radixsort}{cache radixsort}{}{}{}{}{}
-\smallplot{aligned_memory_tuned_radixsort}{aligned cache radixsort}{}{}{}{}{}
+\smallplot{memory_tuned_radixsort}{memory-tuned radixsort}{}{}{}{}{}
+\smallplot{aligned_memory_tuned_radixsort}{aligned memory-tuned radixsort}{}{}{}{}{}
 %

### 2005-07-31 docs/tech_report/shell.tex
fixing up lots of things
the grand reaming is almost complete

@@
 \section{Results}
-\subsection{Test parameters}
+\subsection{Test Parameters}
 The increments in Gonnet's shellsort are $\frac{5}{11}N$, $(\frac{5}{11})^2N$,
@@
 
-\subsection{Simulation results}
+\subsection{Simulation Results}
 \plot{shell}{}{}{}{}{}{}
@@
 slighlty steeper, as can be seen in Figures \ref{shell cycles}, \ref{shell
-instructions}, \ref{shell level 1 misses} and \ref{shell branch
-misses}. It performs better than a
-heapsort, but considerably worse than quicksort. However, it easier to program
-than either of these two sort.
+instructions}, \ref{shell level 1 misses} and \ref{shell branch misses}. It
+performs better than a heapsort, but considerably worse than quicksort. However,
+it easier to program than either of these two sort.
 
@@
 Shellsort's cache misses are certainly in the range of an $O(NlogN)$ sort, and
-are not considerably worse than cache heapsort, and are in fact better than base
-heapsort. This occurs without any cache improvements being performed upon them.
-Shellsort's branch mispredictions, are, as predicted $O(Nlog_{\frac{5}{11}}N)$.
+are not considerably worse than memory-tuned heapsort, and are in fact better
+than base heapsort. This occurs without any cache improvements being performed
+upon them.  Shellsort's branch mispredictions, are, as predicted
+$O(Nlog_{\frac{5}{11}}N)$.
 
@@
 %TODO
-\section{Predictor results}
+\section{Predictor Results}
 

### 2005-07-31 docs/tech_report/software_predictor.tex
the grand reaming is almost complete

@@
 \afterpage{
-\sixplot{base_heapsort_0.eps}{memory_tuned4_heapsort_0.eps}{memory_tuned4_heapsort_1.eps}{memory_tuned_heapsort_0.eps}{memory_tuned_heapsort_1.eps}{}{base and cache heapsorts}
+\sixplot{base_heapsort_0.eps}{memory_tuned4_heapsort_0.eps}{memory_tuned4_heapsort_1.eps}{memory_tuned_heapsort_0.eps}{memory_tuned_heapsort_1.eps}{}{base and memory-tuned heapsorts}
 
-\sixplot{base_quicksort_0.eps}{base_quicksort9_0.eps}{memory_tuned_quicksort_0.eps}{multi_quicksort_0.eps}{sequential_multi_quicksort_0.eps}{}{base, cache and multi quicksorts and base quicksort9}
-\sixplot{base_quicksort1_0.eps}{base_quicksort_0.eps}{base_quicksort5_0.eps}{base_quicksort7_0.eps}{base_quicksort9_0.eps}{}{base quicksort without a median, with median of 3, and with pseudo median of 5, 7 and 9}
-
-\sixplot{algorithm_n_0.eps}{algorithm_n_1.eps}{algorithm_s_0.eps}{algorithm_s_1.eps}{}{}{Knuth's Mergesorts}
+\sixplot{base_quicksort_0.eps}{base_quicksort9_0.eps}{memory_tuned_quicksort_0.eps}{multi_quicksort_0.eps}{sequential_multi_quicksort_0.eps}{}{base, memory-tuned and multi quicksorts and base quicksort (pseudo-median-of-9)}
+\sixplot{base_quicksort1_0.eps}{base_quicksort_0.eps}{base_quicksort5_0.eps}{base_quicksort7_0.eps}{base_quicksort9_0.eps}{}{base quicksort without a median, with median of 3, and with pseudo-medians of 5, 7 and 9}
+\sixplot{algorithm_n_0.eps}{algorithm_n_1.eps}{algorithm_s_0.eps}{algorithm_s_1.eps}{}{}{Knuth's mergesorts}
 \sixplot{base_mergesort_0.eps}{base_mergesort_1.eps}{base_mergesort_2.eps}{base_mergesort_3.eps}{base_mergesort_4.eps}{base_mergesort_5.eps}{base mergesort}
-\sixplot{double_aligned_mergesort_0.eps}{double_aligned_mergesort_1.eps}{double_aligned_mergesort_2.eps}{double_aligned_mergesort_3.eps}{double_aligned_mergesort_4.eps}{double_aligned_mergesort_5.eps}{double tiled mergesort - part 1}
-\sixplot{double_aligned_mergesort_6.eps}{double_aligned_mergesort_7.eps}{double_aligned_multi_mergsort_0.eps}{double_aligned_multi_mergsort_1.eps}{double_aligned_multi_mergsort_2.eps}{double_aligned_multi_mergsort_3.eps}{double tiled mergesort - part 2, and double multi mergesort - part 1}
-\sixplot{double_aligned_multi_mergsort_4.eps}{double_aligned_multi_mergsort_5.eps}{double_aligned_multi_mergsort_6.eps}{double_aligned_multi_mergsort_7.eps}{double_aligned_multi_mergsort_8.eps}{double_aligned_multi_mergsort_9.eps}{double multi-mergesort - part 2}
+\sixplot{double_aligned_mergesort_0.eps}{double_aligned_mergesort_1.eps}{double_aligned_mergesort_2.eps}{double_aligned_mergesort_3.eps}{double_aligned_mergesort_4.eps}{double_aligned_mergesort_5.eps}{double-aligned tiled mergesort - part 1}
+\sixplot{double_aligned_mergesort_6.eps}{double_aligned_mergesort_7.eps}{double_aligned_multi_mergsort_0.eps}{double_aligned_multi_mergsort_1.eps}{double_aligned_multi_mergsort_2.eps}{double_aligned_multi_mergsort_3.eps}{double-aligned tiled mergesort - part 2, and double multi mergesort - part 1}
+\sixplot{double_aligned_multi_mergsort_4.eps}{double_aligned_multi_mergsort_5.eps}{double_aligned_multi_mergsort_6.eps}{double_aligned_multi_mergsort_7.eps}{double_aligned_multi_mergsort_8.eps}{double_aligned_multi_mergsort_9.eps}{double-aligned multi-mergesort - part 2}
 

### 2005-07-31 docs/tech_report/code/improved_bubblesort.c

@@ -1,4 +1,4 @@
-void bubblesort2(Item a[], int N)
+void improved_bubblesort(Item a[], int N)
 {
 	int i = 0, j, k = 0;
 	while(i < N-1)

### 2005-07-31 docs/tech_report/code/improved_shakersort.c

@@ -1,4 +1,4 @@
-void shakersort2(Item a[], int N)
+void improved_shakersort(Item a[], int N)
 {
 	int i = 0, j, k = N-1, l = N-1;
 	for ( ; i < k; )

### 2005-07-31 docs/tech_report/code/multi_quicksort.c

@@ -199,5 +199,5 @@ multi_quicksort(unsigned int a[], int N)
 		l = 0;
 	}
 
-	/* as cache quicksort from here */
+	/* as memory-tuned quicksort from here */
 }

### 2005-07-31 docs/tech_report/conclusions.tex
some spell checks, removed unneccessary files, rename elementary

@@
 
-\subsection{Heapsorts}
+\subsection{Heapsort}
 
@@
 
-\subsection{Mergesorts}
+\subsection{Mergesort}
 
@@
 
-\subsection{Quicksorts}
+\subsection{Quicksort}
 
@@
 that using a binary search slows multi-quicksort down compared to a sequential
-search, but that its extra instructions make it slower than tuned quicksort.
-Despite being slower in our tests, pur simulations show tuned quicksort to have
-fewer cache misses, instructions and branch mispredictions than base quicksort.
+search, but that its extra instructions make it slower than memory-tuned
+quicksort.  Despite being slower in our tests, our simulations show tuned
+quicksort to have fewer cache misses, instructions and branch mispredictions
+than base quicksort.
 
-\subsection{Radixsorts}
+\subsection{Radixsort}
 
@@
 
-\subsection{Shellsorts}
+\subsection{Shellsort}
 
@@
 
-Insertion sort has exaclty one miss per key. Using this result we were able to
+Insertion sort has exactly one miss per key. Using this result we were able to
 make several other sorts faster by replacing their simple inline sorts with
@@
 Selection sort also has a very low number of branch mispredictions, the results
-of which match up well to Knuth's calculation. We found aswell that bubblesort
+of which match up well to Knuth's calculation. We found as well that bubblesort
 and shakersort have incredibly bad branch prediction performance, bring
@@
 correlated exactly with LaMarca's. Many of our graphs are exactly the same
-shape to LaMarca's, can be calculated exactly or alomst exactly using his
+shape to LaMarca's, can be calculated exactly or almost exactly using his
 methods, or can be seen to be equivalent when scaled by an appropriate ratio
@@
 of our level 2 cache miss graphs exactly match his. We also note similar
-characterists: LaMarca mentions a wobble in the instruction count graphs, due to
+characteristics: LaMarca mentions a wobble in the instruction count graphs, due to
 having to copy the final merge back; we note this same wobble.
@@
 However, we disagree with his results in several key places. Firstly, we note
-that an inlined sorting method does not perform as fast as insertion sort. We
+that an in-lined sorting method does not perform as fast as insertion sort. We
 find it difficult to perform his improvements on either algorithm N or algorithm
@@
 result, we used a search, which greatly increased the instruction count. As a
-result, mutli-mergesort did not perform as well as the mergesorts which had only
+result, multi-mergesort did not perform as well as the mergesorts which had only
 been tiled.
@@
 
-Where LaMarca found that using Sedgewicks insertion sort optimisation decreased
+Where LaMarca found that using Sedgewick's insertion sort optimisation decreased
 instruction count, we found the opposite. We also found that despite

### 2005-07-31 docs/tech_report/elementary.tex
spell checking. just noticed the word mergsort, so need to fix it

--- 2005-07-31 docs/tech_report/ordernsquared.tex
@@
 \section{Improved Shakersort}
-Shakersorts incorporates the same improvements over shakersort as improved bubblesort
-does over bubblesort, that is, it skips over pockets of sorted keys in both
-directions, as well as ending early. The code for improved shakersort is in Figure
-\vref{Improved shakersort code}.
+Improved shakersort incorporates the same improvements over shakersort as
+improved bubblesort does over bubblesort, that is, it skips over pockets of
+sorted keys in both directions, as well as ending early. The code for improved
+shakersort is in Figure \vref{Improved shakersort code}.
 

### 2005-07-31 docs/tech_report/heap.tex
spell checking. just noticed the word mergsort, so need to fix it

@@
 relatively simply with no extra instructions. If the last key in the heap is
-left in place aswell as being sifted down the heap, it will provide a
+left in place as well as being sifted down the heap, it will provide a
 sentinel for the sift-down operation. The key from the top of the heap is then
@@
 keys fit in the cache without a corresponding increase in misses. For the same
-reason, the base heapsorts have no compulsary misses\footnote{See Section
+reason, the base heapsorts have no compulsory misses\footnote{See Section
 \ref{TODO}.}, while the out-of-place sorts have one compulsory miss in eight,
-when the auxilliary heap is being populated.
+when the auxiliary heap is being populated.
 
@@
 same amount of misses in both cases. Similarly, the results here indicate that
-base heapsort experiences a sharp increase in cache misses. Cache heaport also
+base heapsort experiences a sharp increase in cache misses. Cache heapsort also
 has a similar increase, though LaMarca's results show that these occur at the
-same position in the graph, in other words, when th set size is the same. Our
+same position in the graph, in other words, when the set size is the same. Our
 results, however, show that this occurs earlier in memory-tuned heapsort, due to the
@@
 
-The most interesting thing is the relation between the predictablility of the
+The most interesting thing is the relation between the predictability of the
 comparisons during the destruction of the 2-, 4- and 8-heaps. Compare the
@@
 comparing simulated results from there to the sorts created here might have
-interesting results and would explain the discreptancies highlighted in this
+interesting results and would explain the discrepancies highlighted in this
 chapter.

### 2005-07-31 docs/tech_report/merge.tex
spell checking. just noticed the word mergsort, so need to fix it

@@
 describes on it. We attempted to contact him in order to discuss this, but
-recieved no response. It is likely that the improvements were actually applied
+received no response. It is likely that the improvements were actually applied
 to Knuth's algorithm S.
@@
 
-\sixplot{algorithm_n_0.eps}{algorithm_s_0.eps}{base_mergesort_0.eps}{double_aligned_mergesort_0.eps}{double_aligned_multi_mergsort_0.eps}{double_aligned_multi_mergsort_0.eps}{mergesorts}
+\sixplot{algorithm_n_0.eps}{algorithm_s_0.eps}{base_mergesort_0.eps}{double_aligned_mergesort_0.eps}{double_aligned_multi_mergesort_0.eps}{double_aligned_multi_mergesort_0.eps}{mergesorts}
 

### 2005-07-31 docs/tech_report/software_predictor.tex
spell checking. just noticed the word mergsort, so need to fix it

@@
 \sixplot{double_aligned_mergesort_0.eps}{double_aligned_mergesort_1.eps}{double_aligned_mergesort_2.eps}{double_aligned_mergesort_3.eps}{double_aligned_mergesort_4.eps}{double_aligned_mergesort_5.eps}{double-aligned tiled mergesort - part 1}
-\sixplot{double_aligned_mergesort_6.eps}{double_aligned_mergesort_7.eps}{double_aligned_multi_mergsort_0.eps}{double_aligned_multi_mergsort_1.eps}{double_aligned_multi_mergsort_2.eps}{double_aligned_multi_mergsort_3.eps}{double-aligned tiled mergesort - part 2, and double multi mergesort - part 1}
-\sixplot{double_aligned_multi_mergsort_4.eps}{double_aligned_multi_mergsort_5.eps}{double_aligned_multi_mergsort_6.eps}{double_aligned_multi_mergsort_7.eps}{double_aligned_multi_mergsort_8.eps}{double_aligned_multi_mergsort_9.eps}{double-aligned multi-mergesort - part 2}
+\sixplot{double_aligned_mergesort_6.eps}{double_aligned_mergesort_7.eps}{double_aligned_multi_mergesort_0.eps}{double_aligned_multi_mergesort_1.eps}{double_aligned_multi_mergesort_2.eps}{double_aligned_multi_mergesort_3.eps}{double-aligned tiled mergesort - part 2, and double multi mergesort - part 1}
+\sixplot{double_aligned_multi_mergesort_4.eps}{double_aligned_multi_mergesort_5.eps}{double_aligned_multi_mergesort_6.eps}{double_aligned_multi_mergesort_7.eps}{double_aligned_multi_mergesort_8.eps}{double_aligned_multi_mergesort_9.eps}{double-aligned multi-mergesort - part 2}
 

### 2005-07-31 docs/tech_report/method.tex
the grand reaming is almost complete
spell checking

@@
 functions which updated simulated branch predictors. Each important branch was
-allocated a predictor, which was updated independantly of the other predictors.
+allocated a predictor, which was updated independently of the other predictors.
 This allowed the performance to be analysed down to the individual branch, and
@@
 actually increased slightly as a result of this, indicating that algorithm N
-does a better job of pre-sorting the array than the inlined pre-sort did.
+does a better job of pre-sorting the array than the in-lined pre-sort did.
 
@@
 which had been developed as part of heapsort.
-
+%TODO this isnt what happened exactly
 Once these sorts were written it was attempted to improve them slightly.
-Instead of a tiled mergesort, a double tiled mergesort was written.  This
+Instead of a tiled mergesort, a double-aligned tiled mergesort was written.  This
 attempted to completely sort in the level 1 cache, then to completely sort in

### 2005-07-31 docs/tech_report/quick.tex
spell checking

@@
 Sedgewick's second improvement is the use of an elementary sort for small
-subfiles.  Once a partition is smaller than an arbitrary threshold, insertion
+sub-files.  Once a partition is smaller than an arbitrary threshold, insertion
 sort is used to sort the partition. Sedgewick states that roughly the same
@@
 across the entire array at the end. This, however, is not implemented; instead a
-check is done to fetch a sentinel from the appropriate subfile if required. This
+check is done to fetch a sentinel from the appropriate sub-file if required. This
 check isn't very expensive, and is significantly cheaper than either a bounds
-check or putting a sentinel into every subfile.
+check or putting a sentinel into every sub-file.
 
@@
 be its final position.  The indices of these positions is pushed onto the
-stack, and the subarray is then sorted.
+stack, and the sub-array is then sorted.
 
@@
 performed as badly as the memory-tuned version. However, the rate at which the
-cahce misses are increasing is very low, being almost a flat line on our graph.
+cache misses are increasing is very low, being almost a flat line on our graph.
 LaMarca's calculation of four misses per cache block is very accurate, as we
@@
 Memory-tuned quicksort had fewer branch misses than base quicksort, due to the fact
-that fewer comparitive branches are executed. In base quicksort, we sort the
+that fewer comparative branches are executed. In base quicksort, we sort the
 entire array using insertion sort. In tuned quicksort, we don't sort the pivots,
@@
 but that less time will be spent partitioning to arrays as a better median is
-found, and a corresponsing increase in time spent finding the better median.
+found, and a corresponding increase in time spent finding the better median.
 
@@
 10\% reduction in branches. At the same time, there is a 20\% reduction in
-correct predictions, accounting for the entire reduction in branhces, and a 4\%
+correct predictions, accounting for the entire reduction in branches, and a 4\%
 increase in mispredictions. This is due to better partitioning. Less time is
@@
 terms of branch prediction. Figure \ref{TODO} shows the cycle count of each of
-the quicksorts, showing that a pseduo-median-of-seven offers the greatest
-tradeoff between the reduced number of branches and instructions, and the
-increase in branch midpredictions.
+the quicksorts, showing that a pseudo-median-of-seven offers the greatest
+trade-off between the reduced number of branches and instructions, and the
+increase in branch mispredictions.
 

### 2005-07-31 docs/tech_report/radix.tex
spell checking

@@
 In Section \ref{tiled mergesort}, mergesort is improved by aligning its array
-with reference to it auxilliary array. Specifically, extra memory is allocated
-for the auxilliary array, which is then offset so that the start of the two
+with reference to it auxiliary array. Specifically, extra memory is allocated
+for the auxiliary array, which is then offset so that the start of the two
 arrays are exactly half the length of the cache apart. We further added an
@@
 slightly more than was predicted. The extra is likely due to the unpredictable
-way that the source array is accessed as the keys are copied to the auxilliary
+way that the source array is accessed as the keys are copied to the auxiliary
 array.
 
-Cache radixsort performs as predicted aswell. By sorting keys between two
+Cache radixsort performs as predicted as well. By sorting keys between two
 arrays instead of copying a partially sorted array back, the instruction count
@@
 keys per cache line results in exactly one miss in four (although our graph
-obscures this, due to the initial subtraction of compulsary misses).
+obscures this, due to the initial subtraction of compulsory misses).
 
@@
 mispredictions, it is no surprise that memory-tuned radixsort performs as well as it
-does. As shown in Figure \ref{radix cycles}, memory-tuned radixsorts performance is
+does. As shown in Figure \ref{radix cycles}, memory-tuned radixsort's performance is
 almost linear, with only a slight increase when the arrays no longer fit in the

### 2005-07-31 docs/tech_report/shell.tex
spell checking

@@
 
-Shell begins by considering the array to be made up of $\frac{N}{2}$ subarrays
-of length two, where the elements of the subarray are spaced $\frac{N}{2}$ keys
-apart.  These subarrays are then sorted using insertion sort. The first
+Shell begins by considering the array to be made up of $\frac{N}{2}$ sub-arrays
+of length two, where the elements of the sub-array are spaced $\frac{N}{2}$ keys
+apart.  These sub-arrays are then sorted using insertion sort. The first
 iteration will be a simple exchange, but the keys which are exchanged move
@@
 count and cycle counts are expected to be similar to $O(NlogN)$, as this has
-been indicated by previous experient.
+been indicated by previous experiment.
 
-The cache performance of shellsort should be regular. Each subarray is sorted in
-turn without reference to the next. Since each element of a subarray is quite
+The cache performance of shellsort should be regular. Each sub-array is sorted in
+turn without reference to the next. Since each element of a sub-array is quite
 far from its next element, an entire cache line must be loaded for each key,
-which will be used once per iteration of the sort across the subarray. If there
-are a large number of subarrays, as during the iterations with large increments,
+which will be used once per iteration of the sort across the sub-array. If there
+are a large number of sub-arrays, as during the iterations with large increments,
 there will be no spatial reuse, and little temporal reuse, so long as the
-subarray is large than the cache. Once the increment is small enough that each
-subarray fits within the cache (i.e. quickly for L2 cache, and much more slowly
-for the L1 cache), then temporal reuse will occur. Once an entire subarray fits
+sub-array is large than the cache. Once the increment is small enough that each
+sub-array fits within the cache (i.e. quickly for L2 cache, and much more slowly
+for the L1 cache), then temporal reuse will occur. Once an entire sub-array fits
 within an eighth of the size of the cache (eight being the number of keys
-that fit in a cache line), then spatial reuse will occur aswell.  This reuse
+that fit in a cache line), then spatial reuse will occur as well.  This reuse
 will only occur within a single iteration; once the increment with the next
@@
 
-Improved shellsort should have a slighlty lower instruction count, and hence a
+Improved shellsort should have a slightly lower instruction count, and hence a
 lower cycle count, than shellsort. It's cache performance should be the same,
-and it should have the same number of comparitive branches, and hence a similar
+and it should have the same number of comparative branches, and hence a similar
 number of mispredictions, though its branch count should be higher than
@@
 
-Shellsort's simulation results have the characterist shape and range of an
+Shellsort's simulation results have the characteristic shape and range of an
 $O(NlogN)$ sort, albeit, not a very good one. The slopes of its graphs are
-slighlty steeper, as can be seen in Figures \ref{shell cycles}, \ref{shell
+slightly steeper, as can be seen in Figures \ref{shell cycles}, \ref{shell
 instructions}, \ref{shell level 1 misses} and \ref{shell branch misses}. It

### 2005-07-31 docs/tech_report/heap.tex
more minor renaming

@@
 same amount of misses in both cases. Similarly, the results here indicate that
-base heapsort experiences a sharp increase in cache misses. Cache heapsort also
-has a similar increase, though LaMarca's results show that these occur at the
-same position in the graph, in other words, when the set size is the same. Our
-results, however, show that this occurs earlier in memory-tuned heapsort, due to the
-fact that it is an out of place sort.
+base heapsort experiences a sharp increase in cache misses. Memory-tuned
+heapsort also has a similar increase, though LaMarca's results show that these
+occur at the same position in the graph, in other words, when the set size is
+the same. Our results, however, show that this occurs earlier in memory-tuned
+heapsort, due to the fact that it is an out of place sort.
 

### 2005-07-31 docs/tech_report/merge.tex
more minor renaming

@@
 
-\section{Double Multi-mergesort}
+\section{Double-aligned Multi-mergesort}
 
-Double Multi-mergesort combines the double tiling of Double Tiled mergesort with
-Multi-mergesort.
+Double-aligned Multi-mergesort combines the double alignment of Double-aligned
+tiled mergesort with multi-mergesort.
 
@@
 As predicted, the rate of level 1 misses is greatly reduced by the double
-tiling, as shown in Figure \ref{merge level 1 misses}. Double multi-mergesort
+tiling, as shown in Figure \ref{merge level 1 misses}. Double-aligned multi-mergesort
 has consistently four times fewer level 1 misses than multi-mergesort, and both
@@
 misses result is poor performance for tiled and multi-mergesort. Multi-mergesort
-is further encumbered with a much higher instruction count. Double tiled
+is further encumbered with a much higher instruction count. Double-aligned tiled
 mergesort, with its low instruction count, and low cache miss rate in both
@@
 
-Double-tiled mergesort (which has the same results are tiled mergesort) has a
+Double-aligned tiled mergesort (which has the same results are tiled mergesort) has a
 predilection for merging forwards. This is likely to be due to the fact that the

### 2005-07-31 docs/tech_report/radix.tex
minor edits
more minor renaming

@@
 
-Cache radixsort should traverse the array only nine times: Once at the start to
-count the least significant bytes; six times (once per array per pass) as it
-copies using the three least significant bytes (counting the next least
+Memory-tuned radixsort should traverse the array only nine times: Once at the
+start to count the least significant bytes; six times (once per array per pass)
+as it copies using the three least significant bytes (counting the next least
 significant byte at the same time), and a final two as it copies using the most
@@
 mergesort}) to significantly reduce level 1 misses at the expense of a small
-number of level 2 misses. We adapt this strategy here, and call it Aligned Cache
-Radixsort.
+number of level 2 misses. We adapt this strategy here, and call it Aligned
+Memory-tuned Radixsort.
 
@@
 should remain the same. The instruction count should, however, reduce by nearly
-a half to $fract{9}{16}$ths of base radixsort.
+a half to $\frac{9}{16}^ths$ of base radixsort.
 
@@
 
-Cache radixsort performs as predicted as well. By sorting keys between two
+Memory-tuned radixsort performs as predicted as well. By sorting keys between two
 arrays instead of copying a partially sorted array back, the instruction count

### 2005-08-01 docs/tech_report/conclusions.tex

@@
 
-\section {Branch Results}
+\section{Branch Results}
+%TODO this section has to be reread by David
 
@@
 of which match up well to Knuth's calculation. We found as well that bubblesort
-and shakersort have incredibly bad branch prediction performance, bring
-$O(N^2)$.
+and shakersort have incredibly bad branch prediction performance. Both are
+$O(N^2)$ sorts, and the number of branches and branch misses have the same
+complexity.
+%TODO i dont like how this reads
 
-We found a pattern in branch misprediction when using a fanout in our 8-heap,
-and that this pattern matched closely that of selection sort, discussed above.
+TODO bubblesort
 
-TODO merge
+In heapsort, we found a pattern of branch mispredictions when using a fanout in
+our 8-heap, and that this pattern matched closely that of selection sort,
+discussed above. Despite this property, the number of mispredictions using an
+8-heap was greater than when using a 4-heap.
+
+Using a k-way merge, multi-mergesort had a great number of extra branch
+predictions. While many of these were mispredicted using the bimodal predictor,
+the 2 level adaptive predictor was able to predict a higher
+proportion correctly.
 

### 2005-08-01 docs/tech_report/merge.tex
renamed double_aligned_tiled_mergesort

@@
 
-\sixplot{algorithm_n_0.eps}{algorithm_s_0.eps}{base_mergesort_0.eps}{double_aligned_mergesort_0.eps}{double_aligned_multi_mergesort_0.eps}{double_aligned_multi_mergesort_0.eps}{mergesorts}
+\sixplot{algorithm_n_0.eps}{algorithm_s_0.eps}{base_mergesort_0.eps}{double_aligned_tiled_mergesort_0.eps}{double_aligned_multi_mergesort_0.eps}{double_aligned_multi_mergesort_0.eps}{mergesorts}
 

### 2005-08-01 docs/tech_report/result_list.tex
spell checking. just noticed the word mergsort, so need to fix it
renamed double_aligned_tiled_mergesort

@@
 \smallplot{multi_mergesort}{multi-mergesort}{}{}{}{}{}
-\smallplot{double_aligned_mergesort}{double-aligned tiled mergesort}{}{}{}{}{}
-\smallplot{double_aligned_multi_mergsort}{double-aligned multi-mergesort}{}{}{}{}{}
+\smallplot{double_aligned_tiled_mergesort}{double-aligned tiled mergesort}{}{}{}{}{}
+\smallplot{double_aligned_multi_mergesort}{double-aligned multi-mergesort}{}{}{}{}{}
 %

### 2005-08-01 docs/tech_report/software_predictor.tex
renamed double_aligned_tiled_mergesort

@@
 \sixplot{base_mergesort_0.eps}{base_mergesort_1.eps}{base_mergesort_2.eps}{base_mergesort_3.eps}{base_mergesort_4.eps}{base_mergesort_5.eps}{base mergesort}
-\sixplot{double_aligned_mergesort_0.eps}{double_aligned_mergesort_1.eps}{double_aligned_mergesort_2.eps}{double_aligned_mergesort_3.eps}{double_aligned_mergesort_4.eps}{double_aligned_mergesort_5.eps}{double-aligned tiled mergesort - part 1}
-\sixplot{double_aligned_mergesort_6.eps}{double_aligned_mergesort_7.eps}{double_aligned_multi_mergesort_0.eps}{double_aligned_multi_mergesort_1.eps}{double_aligned_multi_mergesort_2.eps}{double_aligned_multi_mergesort_3.eps}{double-aligned tiled mergesort - part 2, and double multi mergesort - part 1}
+\sixplot{double_aligned_tiled_mergesort_0.eps}{double_aligned_tiled_mergesort_1.eps}{double_aligned_tiled_mergesort_2.eps}{double_aligned_tiled_mergesort_3.eps}{double_aligned_tiled_mergesort_4.eps}{double_aligned_tiled_mergesort_5.eps}{double-aligned tiled mergesort - part 1}
+\sixplot{double_aligned_tiled_mergesort_6.eps}{double_aligned_tiled_mergesort_7.eps}{double_aligned_multi_mergesort_0.eps}{double_aligned_multi_mergesort_1.eps}{double_aligned_multi_mergesort_2.eps}{double_aligned_multi_mergesort_3.eps}{double-aligned tiled mergesort - part 2, and double multi mergesort - part 1}
 \sixplot{double_aligned_multi_mergesort_4.eps}{double_aligned_multi_mergesort_5.eps}{double_aligned_multi_mergesort_6.eps}{double_aligned_multi_mergesort_7.eps}{double_aligned_multi_mergesort_8.eps}{double_aligned_multi_mergesort_9.eps}{double-aligned multi-mergesort - part 2}

### 2005-08-01 code/sorts/cache_sorts.h

@@ -9,7 +9,7 @@ void memory_tuned4_heapsort(unsigned int a[], int N); /* aligned 4 heap */
 
 void tiled_mergesort(unsigned int a[], int N); /* level 2 first, then continue merging */
 void multi_mergesort(unsigned int a[], int N); /* level 2, then k-way mere using heapsort */
-void double_aligned_mergesort(unsigned int a[], int N); /* level 1, level 2, then continue merging */
+void double_aligned_tiled_mergesort(unsigned int a[], int N); /* level 1, level 2, then continue merging */
 void double_aligned_multi_mergesort(unsigned int a[], int N); /* level 1, level 2, then k-way mere using heapsort */
 
 void memory_tuned_quicksort(unsigned int a[], int N); /* insertion sort at the end */

### 2005-08-01 code/sorts/double_aligned_tiled_mergesort.c

@@ -75,7 +75,7 @@ static int get_count(const int N)
 }
 
 void
-double_aligned_mergesort(unsigned int a[], int N)
+double_aligned_tiled_mergesort(unsigned int a[], int N)
 {
 	/* track keeps an eye on where i started last */
 	/* outer track keeps and eye on which cache_sized/2 segemnt i started in last */

### 2005-08-01 docs/tech_report/code/double_aligned_tiled_mergesort.c

@@ -74,7 +74,7 @@ static int get_count(const int N)
 }
 
 void
-double_aligned_mergesort(unsigned int a[], int N)
+double_aligned_tiled_mergesort(unsigned int a[], int N)
 {
 	/* track keeps an eye on where i started last */
 	/* outer track keeps and eye on which cache_sized/2 segemnt i started in last */

### 2005-08-01 docs/tech_report/quick.tex
more minor renaming
enambled main again
more minor edits.

@@
 
-An additional benefit of Sedgewick's median of three implementation is that it
+An additional benefit of Sedgewick's median-of-3 implementation is that it
 provides a sentinel for the partition. It is now possible to remove the bounds
@@
 Quicksort has been well studied, and several important improvements have been
-noted. In \cite{BentleyMcIlroy93}, using a median of nine partition is discussed,
+noted. In \cite{BentleyMcIlroy93}, using a median-of-9 partition is discussed,
 and modern discussions on quicksort refer to the problem of equal keys, which is
@@
 
-\section{Memory-Tuned Quicksort}
+\section{Memory-tuned Quicksort}
 
@@
 of instructions for multi-quicksort should increase by about 50\% in the case of
-the sequential sort, and 25\% in the case of the binary sort. Cache quicksort
-should also have a slightly higher instruction count than base quicksort, due to
-the change in use of insertion sort.
+the sequential sort, and 25\% in the case of the binary sort. Memory-tuned
+quicksort should also have a slightly higher instruction count than base
+quicksort, due to the change in use of insertion sort.
 
@@
 associative ones. This is true in the case of all the quicksorts, as can be seen
-in Figures \ref{base quicksort (median-of-3) (no median) level 2 misses} to
+in Figures \ref{base quicksort (no median) level 2 misses} to
 \ref{multi-quicksort (sequential search) level 2 misses} graph that this is true
@@
 
-\sixplot{base_quicksort_0.eps}{base_quicksort9_0.eps}{memory_tuned_quicksort_0.eps}{multi_quicksort_0.eps}{sequential_multi_quicksort_0.eps}{}{base,
+\sixplot{base_quicksort__median_of_3__0.eps}{base_quicksort__pseudo_median_of_9__0.eps}{memory_tuned_quicksort_0.eps}{multi_quicksort__binary_search__0.eps}{multi_quicksort__sequential_search__0.eps}{}{base,
 cache and multi quicksorts and base quicksort (pseudo-median-of-9)}
@@
 \label{quick-predictors2}
-\sixplot{base_quicksort1_0.eps}{base_quicksort_0.eps}{base_quicksort5_0.eps}{base_quicksort7_0.eps}{base_quicksort9_0.eps}{}{base quicksort without a median, with median of 3, and with pseudo-medians of 5, 7 and 9}
+\sixplot{base_quicksort__no_median__0.eps}{base_quicksort__median_of_3__0.eps}{base_quicksort__pseudo_median_of_5__0.eps}{base_quicksort__pseudo_median_of_7__0.eps}{base_quicksort__pseudo_median_of_9__0.eps}{}{base quicksort without a median, with median of 3, and with pseudo-medians of 5, 7 and 9}
 
@@
 ithkn} suggests partitioning using a median-of-9 instead of a median-of-3. We
-created sorts using no median, a median-of-3, and pseudo-medians-of-five, -seven and
--nine. The pseudo-medians used repeated medians-of-three instead of a real
+created sorts using no median, a median-of-3, and pseudo-medians-of-5, -7 and
+-9. The pseudo-medians used repeated medians-of-3 instead of a real
 median, as suggestion by \ref{bently, again, i think}. 
@@
 pseudo medians of 5, 7 and 9}. There is little variance in insertion sort misses
-between the sorts using pseudo-medians, but the median-of-three quicksort has
+between the sorts using pseudo-medians, but the median-of-3 quicksort has
 6\% fewer branches, and a corresponsing decrease in mispredictions.
@@
 
-Between a median-of-three and a pseudo-median-of-nine partitioning, there is a
+Between a median-of-3 and a pseudo-median-of-9 partitioning, there is a
 10\% reduction in branches. At the same time, there is a 20\% reduction in
@@
 terms of branch prediction. Figure \ref{TODO} shows the cycle count of each of
-the quicksorts, showing that a pseudo-median-of-seven offers the greatest
+the quicksorts, showing that a pseudo-median-of-7 offers the greatest
 trade-off between the reduced number of branches and instructions, and the

### 2005-08-01 docs/tech_report/heap.tex
minor edits to fix latex errors

@@
 columns marked `\cc{Destruction, child comparison}' and `\cc{Destruction, child
-promotion}' in Figure \ref{base and memory-tuned heapsort-1} (the 2-heap), with
+promotion}' in Figure \ref{base and memory-tuned heapsorts-1} (the 2-heap), with
 the columns marked `\cc{comparison 1}' to `\cc{child promotion}' in Figure
 \ref{base and memory-tuned heapsorts-2} (the 4-heap) and Figures \ref{base and
-memory-tuned heapsort-3} and \ref{base and memory-tuned heapsort-4} (the
+memory-tuned heapsorts-3} and \ref{base and memory-tuned heapsorts-4} (the
 8-heap).
@@
 becoming more predictable as the fanout increases. For example, in the 8-heap,
-shown in Figures \ref{base and memory-tuned heapsort-3} and \ref{base and
-memory-tuned heapsort-4}, the first comparison is predicted correctly 50\% of
+shown in Figures \ref{base and memory-tuned heapsorts-3} and \ref{base and
+memory-tuned heapsorts-4}, the first comparison is predicted correctly 50\% of
 the time, while the next is predicted correctly 61\% of the time, then 71\%,

### 2005-08-01 docs/tech_report/quick.tex
minor edits to fix latex errors

@@
 performance for base quicksort without a median, with median of 3, and with
-pseudo medians of 5, 7 and 9}. There is little variance in insertion sort misses
+pseudo-medians of 5, 7 and 9}. There is little variance in insertion sort misses
 between the sorts using pseudo-medians, but the median-of-3 quicksort has

### 2005-08-01 docs/tech_report/merge.tex
minor edits to fix latex errors
added testing comments

@@
 \subsection{Test Parameters}
+%TESTING
 The level 1 cache is 8k, so 1024 keys from each array can be sorted in the level
@@
 being the fastest. Note that the alignment of the mergesort may not work as
+%TESTING
 desired due to the size of the cache. This was optimized for a 2MB cache, while
@@
 instruction count).  The rate of branch misses is much the same as that in
-heapsort, and progresses in the same manner (see Figure \ref{memory-tuned heapsort (8-heap)
-curve}). Note that \n{comparison 0} occurs every time, and would be optimized
-out by the compiler.
+heapsort, and progresses in the same manner (see Figure \ref{memory-tuned
+heapsort curve}). Note that \n{comparison 0} occurs every time, and would be
+optimized out by the compiler.
 

### 2005-08-01 docs/tech_report/method.tex
more minor renaming
added testing comments

@@
 
+%TESTING - also what version
 The SimpleScalar settings were chosen to be similar to LaMarca's simulations.
@@
 
+%TESTING - what versino
 \subsection{Gnuplot}
@@
 Imaginatively called \cc{main.c}, this program provided the means for testing
-and (roughly) timing each sort. Figure \vref{Key sections of Main.c} contains
-this code.
+and (roughly) timing each sort. Figure \vref{Key sections of Main.c} contains a
+sample of this code.
 
@@
 
+%TESTING - what version
 \subsection{Valgrind}
@@
 
+%TESTING - what version
 \subsection{PapiEx}
@@
 %TODO this isnt what happened exactly
+
 Once these sorts were written it was attempted to improve them slightly.
-Instead of a tiled mergesort, a double-aligned tiled mergesort was written.  This
-attempted to completely sort in the level 1 cache, then to completely sort in
-the level 2 cache. Double multi-mergesort was a combination of this and of
-multi-mergesort. The rewrite was based on the base mergesort, and was written
-far more cleanly, being more readable and efficient. These were also found to
-reduce the instruction count, as well as the level 2 cache miss count. These
-results are discussed in more detail in Section \ref{Mergesort results}.
+Instead of a tiled mergesort, a double-aligned tiled mergesort was written.
+This attempted to completely sort in the level 1 cache, then to completely sort
+in the level 2 cache. Double-aligned multi-mergesort was a combination of this
+and of multi-mergesort. The rewrite was based on the base mergesort, and was
+written far more cleanly, being more readable and efficient. These were also
+found to reduce the instruction count, as well as the level 2 cache miss count.
+These results are discussed in more detail in Section \ref{Mergesort results}.
 
@@
 
-%TODO another sentence here
-Each of the four $O(NlogN)$ sorts, radixsort and shellsort, was treated in this fashion.
+Radixsort, shellsort and each of the four $O(NlogN)$ sorts were treated in this
+fashion.
 
@@
 
+%TESTING
 %TODO a lot of this may be in the first sorts chapter, maybe heapsort.
@@
 strengthen our results.
+%TODO not true. we have it

### 2005-08-01 docs/tech_report/heap.tex
added testing comments
started testing

@@
 
-%TODO move this to testing section
 Figure \ref{heap cycles} shows the actual time taken for heap sort on a Pentium
-4 1.8GHz machine, which has an 8-way associative 256KB level 2 cache with a 64 byte
-cache line, a 4-way associative 8KB level 1 data cache, and a maximum branch
-misprediction delay of 20 cycles\footnote{This is determined from the Pentium
-4's 20 stage pipeline. This number is not exact, and is probably slightly lower,
-with the possibility of occasionally rising higher (in the case of an
-instruction cache miss, which is exceptionally unlikely for small routines like
-sorts).}. Its static branch prediction strategy is backward taken and forward
-not taken; its dynamic strategy is unspecified: though \cite{Intel248966-010}
-mentions a branch history so it is likely a two-level adaptive predictor.
-
-The figure shows that the base heapsort performs best when the data set fits in
-the cache. After that point, the 8-heap performs best, and its cycle count increases
-with a smaller slope. This shows the importance of LaMarca's cache-conscious
-design even nine years after his thesis. It also shows the importance of reduced
-branch mispredictions: the base heapsort has more instructions and level 1 cache
-misses, but still outperforms the 4-heap, until level 2 cache misses take over.
-However, this is hardly conclusive: the lower level 2 cache misses may account
-for this.
+4 machine.  The figure shows that the base heapsort performs best when the data
+set fits in the cache. After that point, the 8-heap performs best, and its cycle
+count increases with a smaller slope. This shows the importance of LaMarca's
+cache-conscious design even nine years after his thesis. It also shows the
+importance of reduced branch mispredictions: the base heapsort has more
+instructions and level 1 cache misses, but still outperforms the 4-heap, until
+level 2 cache misses take over.  However, this is hardly conclusive: the lower
+level 2 cache misses may account for this.
 

### 2005-08-01 docs/tech_report/method.tex
started testing

@@
 SimpleScalar works by compiling code into its own instruction set. It provides a
-compiler - a port of gcc 2.63 - with related tools such as a linker and
-assembler.  Several simulators are then provided: \cc{sim-fast} only counts
-instruction, quickly; \cc{sim-cache} simulates a cache and counts level 1 and
-level 2 cache misses; \cc{sim-bpred} simulates a branch predictor and counts
-branch hits and misses. Other simulators are provided, but they were not used in
-our tests.
+compiler - a port of gcc - with related tools such as a linker and assembler.
+Several simulators are then provided: \cc{sim-fast} only counts instruction,
+quickly; \cc{sim-cache} simulates a cache and counts level 1 and level 2 cache
+misses; \cc{sim-bpred} simulates a branch predictor and counts branch hits and
+misses. Other simulators are provided, but they were not used in our tests.
 
@@
 
-%TESTING - also what version
-The SimpleScalar settings were chosen to be similar to LaMarca's simulations.
-LaMarca used the \n{Atom} simulator, which was used to test software for the DEC
-AlphaStation. It simulated a 2MB cache with a 32 byte cache line. The remainder
-of the settings were taken from the DEC Alpha: 8KB level 1 data cache, 8KB level
-1 instruction cache and shared instruction and data level 2 cache.
-
-%TESTING - what versino
 \subsection{Gnuplot}
 \label{Gnuplot}
-\cc{Gnuplot} is a tool for plotting data files. It provides visualisations of data in
-two and three dimensions, and is used to create all of the graphs in this
-document. Its homepage is \n{http://www.gnuplot.info/}.
+\cc{Gnuplot} is a tool for plotting data files. It provides visualisations of
+data in two and three dimensions, and is used to create all of the graphs in
+this document. Its homepage is \n{http://www.gnuplot.info/}.
 
@@
 scaled to so that the y-axis could be in the form of \n{plotted data per key},
-both to be more readable, and to be consistent with LaMarca's graphs. Once the data
-was converted, it was necessary to generate a gnuplot script for each graph.
-The number of graphs required made this an
-impossible manual task, and so another script had to be written.  Some graphs
-still needed to be tweaked by hand, due to difficulties in the script, but the
-task was much shortened by this, and the tweaking was mostly accomplished using
-macros in the development platform.
+both to be more readable, and to be consistent with LaMarca's graphs. Once the
+data was converted, it was necessary to generate a gnuplot script for each
+graph.  The number of graphs required made this an impossible manual task, and
+so another script had to be written.  Some graphs still needed to be tweaked by
+hand, due to difficulties in the script, but the task was much shortened by
+this, and the tweaking was mostly accomplished using macros in the development
+platform.
 
@@
 
-%TESTING - what version
 \subsection{Valgrind}
@@
 
-%TESTING - what version
 \subsection{PapiEx}
@@
 
+\subsection{}
+simplescalar version and settings used
+The SimpleScalar settings were chosen to be similar to LaMarca's simulations.
+LaMarca used the \n{Atom} simulator, which was used to test software for the DEC
+AlphaStation. It simulated a 2MB cache with a 32 byte cache line. The remainder
+of the settings were taken from the DEC Alpha: 8KB level 1 data cache, 8KB level
+1 instruction cache and shared instruction and data level 2 cache.
+
+\gnuplot version
+valgrinf version
+papiex version
+
+cache line length. - vs pentium 4
+
+penitum 4:
+1.8, 8-way 256K level 2 64 byte line (note i wrote 128 byte xsomewhere. is this
+right)
+4-way 8kb level 1 cache
+
+branch delay 20 cycles
+
+1.8GHz machine, which has an 8-way associative 256KB level 2 cache with a 64 byte
+cache line, a 4-way associative 8KB level 1 data cache, and a maximum branch
+misprediction delay of 20 cycles\footnote{This is determined from the Pentium
+4's 20 stage pipeline. This number is not exact, and is probably slightly lower,
+with the possibility of occasionally rising higher (in the case of an
+instruction cache miss, which is exceptionally unlikely for small routines like
+sorts).}. Its static branch prediction strategy is backward taken and forward
+not taken; its dynamic strategy is unspecified: though \cite{Intel248966-010}
+mentions a branch history so it is likely a two-level adaptive predictor.
+
+64 bit vs 32 bit
+
 \section{Future Work}

### 2005-08-02 docs/tech_report/heap.tex
minor changes mergeed in david's heap changes

@@
 
-\section{Predictor Results}
+\section{A more detailed look at branch prediction}
 
 \sixplot{base_heapsort_0.eps}{memory_tuned4_heapsort_0.eps}{memory_tuned_heapsort_0.eps}{memory_tuned_heapsort_1.eps}{}{}{base and memory-tuned heapsorts}
+
+In order to explore the behaviour of the individual branches in
+heapsort, we performed a second set of branch prediction
+experiments. Rather than using SimpleScalar, we instead added
+additional statements to the heapsort source code to simulate a
+separate bimodal predictor for each branch. This allowed us to measure
+each branch individually, whereas SimpleScalar gives only a total
+figure for all branches. An additional advantage is that the resulting
+compiled simulation is much faster than a full microarchitecture
+simulator, and we were thus able to take the average of thousands of
+runs (\textbf{TODO find out exact number}), rather than the small
+number of runs possible with SimpleScalar. For these experiments,
+randomly generated array of size \textbf{xxxxx} were used.
+
+\textbf{[David] I am confused by this chart, because the labelling is
+a bit unclear. There appears to be a base heapsort, a 4-way cache
+heapsort and an 8-way cache heapsort. But the labelling doesn't tell
+us this.}
+
+Figure \ref{base and cache heapsorts-1}(a) shows the behaviour of each
+branch in the base heapsort algorithm. The ``creation, has children'' branch
+is from line \textbf{xxxxx} of Figure \ref{Simple Base Heapsort}.
+
+\textbf{[David] We need to find some way to describe what each of these
+branches is. I don't know myself, so the general reader really, really
+won't know.}
 

### 2005-08-02 docs/tech_report/elementary.tex
added testing comments
merged david's elementary changes

@@
 \subsection{Testing}
+%TESTING
 All elementary sorts are only sorted up to 65536 keys in the simulations, and
@@
 cost of comparing them. Conversely, for large keys with small records, the cost
-of selection sort is far higher than for other simple sorts.
+of selection sort may be higher than for other simple sorts.
 
@@
 \label{selection branches logn}
-The branch prediction performance is not expected to be bad. Flow control
-predictions are very straightforward, and should result in very few misses.
-Comparison predictions, however, are very numerous. Each array traversal has an
-average of $N/2$ comparisons, and approximately $log_2(N/2)$ of them should be
-taken, while the rest are not-taken. The taken predictions will be misses,
-while the rest should be hits, so it is expected that there will be $logN$
-misses per key.
+The branch prediction performance is not expected to be bad. Flow
+control predictions are very straightforward, and should result in
+very few misses.  Comparison predictions, however, are very
+numerous. Each traversal of the array has an average of $N/2$
+comparisons, as we search for the smallest key in the unsorted part of
+the array. During the search, the algorithm compares each key with the
+smallest key seen so far. In a random array, we would expect that the
+candidate for the smallest key (the left-to-right minimum) would
+change frequently towards the start of the array, making the
+comparison branch rather unpredictable. However, as the traversal
+continues, the left-to-right minimum will become smaller, and thus
+more difficult to displace. It will become very unusual to find a
+smaller key, and thus the branch will almost always resolve in the
+same direction.  Therefore, the comparison branch will quickly become
+very predictable.
 
-This number is not immediately obvious. Consider a single step of the selection
-sort. The first key to be sorted is, on average, larger than half the keys and
-smaller than the other half, so it will have to travel across half the array to
-find a key smaller than it, on average. Once it finds a smaller key, this will
-be larger than a quarter of the keys in the array, and so it will need to travel
-half of the remaining array, on average, before it finds it. The search will
-continue like this until the entire array has been covered. Comparing two keys
-and discovering that the new key is not smaller than the current smallest is the
-norm; every time a smaller key is found a branch misprediction will occur (if
-two or more keys in a row are successively smaller, this may not be the case).
-The length of the array to be considered in $N/2$ on average, and an average
-total of $Nlog_2(N/2)$ branch misses will occur. This assumes randomly sorted
-keys.
+Knuth \cite{Knuth1} analyses the number of changes to the
+left-to-right minimum when searching for the minimum element of an
+array. He shows that for $N$ elements, the number of changes is $H_N -
+1$, where $H_N$ is the harmonic series:
 
-%TODO look at knuths number here
-TODO Knuth has a formula which I will look up.
+\displaymath{H_N = 1 + \frac{1}{2} + \frac{1}{3} + ... + \frac{1}{N} =  \sum_{k=1}^{N} \frac{1}{k},  \hspace{0.4cm}  N \geq 1}
+
+Within practical sorting ranges, $H_N$ grows at a rate similar to
+$log(N)$. Thus, in the average case, we can expect that the comparison
+branch will mostly go in the same direction, making the branch highly
+predictable.
+
+Knuth also presents the number of changes to the left-to-right minimum
+in the best case (0 changes) and worst cases ($N-1$ changes).  In the
+worst case, the array is sorted in reverse order, and so the
+left-to-right minimum changes on every element. Interestingly, the
+worst cases for the number of changes is not the worst case for branch
+mispredictions. If the minimum changes for every element, then the
+comparison branch will always resolve in the same direction, and thus
+be almost perfectly predictable.  From the point of view of branch
+prediction, frequent changes (preferably with no simple pattern) are
+the main indicator of poor performance, rather than the absolute
+number of times that the branch resolves in each direction. The worst
+case for selection sort would probably be an array which is partially
+sorted in reverse order --- the left-to-right minimum would change
+frequently, but not predictably so.
 
@@
 The branch predictive performance is also expected to be similar. While
-selection sort has $O(NlogN)$ misses, each of bubblesort's comparisons has
-the potential to be a miss, and not just in the pathological case. However, as
-the sort continues and the array becomes more sorted, the number of misses
-should decrease. The number of flow control misses should be very low.
+selection sort has $O(NH_N)$ mispredictions, each of bubblesort's comparisons
+has the potential to be a misprediction, and not just in the pathological case.
+However, as the sort continues and the array becomes more sorted, the number of
+mispredictions should decrease. The number of flow control mispredictions should
+be very low.
 
@@
 detects that the array is fully sorted. It keeps a binary flag to indicate if
-there were any exchanges along the way.  Instead, improved bubblesort keeps
-track of the last exchange made, and starts the next iteration from there. This
-means that any pockets of sorted keys are skipped over, not just if they occur
-at the end. Improved bubblesort code is shown in Figure \vref{Improved bubblesort code}.
+there were any exchanges along the way. Instead, improved bubblesort keeps track
+of the last exchange made, and starts the next iteration from there. This means
+that any pockets of sorted keys are skipped over, not just if they occur at the
+end. Improved bubblesort code is shown in Figure \vref{Improved bubblesort
+code}.
 
@@
 
-As expected, selection sort is the worst. This is due to the fact that it
-considers the entire unsorted segment of the array for each key to be put in
-place. Bubblesort, with similar behaviour, is only slightly better. Shakersort
-has a small amount of temporal reuse, reducing its number of misses. Improved
-shakersort is placed roughly between insertion sort and shakersort, showing that
-ending early is helpful in its case. This does not appear to be the case with
-improved bubblesort though, which performs identically to bubblesort.
+As expected, selection sort has the worst cache performance. This is due to the
+fact that it considers the entire unsorted segment of the array for each key to
+be put in place. Bubblesort, with similar behaviour, is only slightly better.
+Shakersort has a small amount of temporal reuse, reducing its number of misses.
+Improved shakersort is placed roughly between insertion sort and shakersort,
+showing that ending early is helpful in its case. This does not appear to be the
+case with improved bubblesort though, which performs identically to bubblesort.
 
@@
 sorts, while Figure \ref{Branch Simulation results for elementary sorts} shows
-the branch misses per key.  Branch prediction performance is not surprising.
-Each sort has a similar number of branches, with insertion sort having half as
-many as selection and bubblesort, and shakersort being half-way between the
-two. The number of misses for bubblesort and shakersort are huge.  As the data
-sets gets large, there are nearly 10,000 branch misses involved in sorting a
-single key. The improved bubblesort had exactly the same number of misses and
-branches. The improved shakersort has a lot less branches, but the same number
-of misses. This is because skipping over sorted keys is predictable from the
-first branch miss.
+the branch misses per key. Due to huge variations in predictability, we split
+the numbers of branch mispredictions into two charts. Selection sort causes
+around $H_N$ mispredictions per key, which is remarkably small. The comparison
+branch in the inner loop is highly predictable on random data, because a good
+candidate minimum is usually found quickly.
 
 Selection sort, meanwhile, has the same number of branches as bubblesort, but
-only $O(logN)$ misses per key. The most interesting result, though, is insertion
+only $O(H_N)$ misses per key. The most interesting result, though, is insertion
 sort. This has exactly one miss per key. Although this result was not predicted
@@
 establishes insertion sort as the best of the simple sorts, and it is used later
-in quicksort and mergesort as a result of this. \label{insertion is
-predictable}
+in quicksort and mergesort as a result of this. \label{insertion is predictable}
 
-Another interesting result comes from a comparison between shakersort and
-selection sort. The number of cycles to sort an array (shown in Figure
-\ref{elementary cycles}) is nearly twice as high on shakersort as it is for
-selection sort. Shakersort has a significantly lower instruction count and cache
-miss count than selection sort, but has several orders of magnitude worse branch
-misses. This shows how important it is to consider branch misses in an
-algorithm, especially comparative ones. It also establishes that exchange-based
-algorithms, such as bubblesort and shakersort, will perform considerably worse
-than algorithms which have a correlation between successive branches, as in
-insertion and selection sorts.
+The number of misses for bubblesort and shakersort are so huge that we had to
+plot them on a separate chart. Nearly one in four comparisons are misses. As the
+data sets gets large, there are nearly 10,000 branch misses involved in sorting
+a single key. The improved bubblesort had exactly the same number of misses and
+branches. The improved shakersort has a lot less branches, but the same number
+of misses. This is because skipping over sorted keys is predictable from the
+first branch miss.
+
+The branch behaviour of bubble sort is particularly interesting when
+compared to selection sort. Although we usually think of the two
+algorithms as being different, they are actually very similar. Like
+selection sort, bubble sort finds the smallest (or largest) key in the
+unsorted part of the array and moves it to start (end) of the array.
+Thus, we would expect that bubbble sort's branch prediction behaviour
+would be similar to that of selection sort.
+
+However, in the proces of moving the smallest element into position,
+bubble sort also rearranges other array elements. Selection sort
+sweeps across the array and always keeps track of the left-to-right
+minimum up to the current point. However, bubble sort actually moves
+this minimum, exchanging pairwise with each element it encounters
+which is larger. Thus, even the unsorted part of the array becomes
+partially sorted. We might expect that bubble sort's comparison branch
+for a partially sorted array might be more predictable than for a
+random one, but it is not. With a random array, the smaller elements
+are equally likely to appear anywhere in the array. With a partially
+sorted array, the smaller elements tend to appear towards the start of
+the array which is, in bubble sort, the last place that we look. Thus,
+we change the candidate minimum much more often than if we were
+searching a randomly ordered array.
 

### 2005-08-02 docs/tech_report/conclusions.tex
fixed small errors. didnt really get rid of any TODOs

@@
 No results for the branch prediction performance of sorting algorithm yet
-exists. We were able to find some interesting results.
+exists. We discovered some interesting new results.
 
@@
 insertion sort; as such it should be only used for very small data sets.  The
-best sort for very large data sets is radixsort. The best overall sort was
-quicksort, though radixsort is very close, and the radixsort considered here was
-not optimized. It is expected that an optimized radixsort would beat quicksort
-for smaller sets.
+best sort for very large data sets is radixsort. The fastest overall sort was
+memory-tuned radixsort, though quicksort is very close.
 % TODO redo above paragraph

### 2005-08-02 docs/tech_report/elementary.tex
fixed small errors. didnt really get rid of any TODOs

@@
 
-\displaymath{H_N = 1 + \frac{1}{2} + \frac{1}{3} + ... + \frac{1}{N} =  \sum_{k=1}^{N} \frac{1}{k},  \hspace{0.4cm}  N \geq 1}
+$$H_N = 1 + \frac{1}{2} + \frac{1}{3} + ... + \frac{1}{N} =  \sum_{k=1}^{N}
+\frac{1}{k},  \hspace{0.4cm}  N \geq 1$$
 
@@
 The instruction count of shakersort is expected to be exactly the same as
-bubblesort. The version of shakersort used here, however, does not incorporate
-the improvement to end the sort early if the array is already sorted. This will
-probably slightly reduce the instruction count as the sort is unlikely to end
-early for random keys.
-
-The cache performance of shakersort is expected to be superior to that of
-bubblesort. A small amount of temporal reuse occurs at the ends of the array,
-as each key is used twice once its loaded into the cache. In addition, when the
-array is not more than twice the size of the cache, then the keys in the centre
-of the array are not ejected from the cache at all.
+bubblesort, while its cache performance is expected to be superior. A small
+amount of temporal reuse occurs at the ends of the array, as each key is used
+twice once its loaded into the cache. In addition, when the array is not more
+than twice the size of the cache, then the keys in the centre of the array are
+not ejected from the cache at all.
 

### 2005-08-02 docs/tech_report/heap.tex
fixed small errors. didnt really get rid of any TODOs

@@
 $\lfloor{}j/2\rfloor{}$. Using this scheme can reduces instruction cost, but
-results is slightly more difficult pointer arithmetic, which cannot be performed
+results in slightly more difficult pointer arithmetic, which cannot be performed
 in some languages\footnote{In C this can be done by \cc{Item a[] =
@@
 An 8-heap has seven comparisons between children, and will probably have about
+3 %TODO does this mean that knuths harmonics applies to this sort of prediction
+% thing TODO
 mispredictions, as it behaves similarly to selection sort (See \ref{selection
@@
 prediction ratio gets worse as the fanout gets larger. The reason we predicted
-differently was due to selection sort's $O(log N)$ mispredictions; however, this
+differently was due to selection sort's $O(H_N)$ mispredictions; however, this
 is over the entire length of the array, and it is not accurate to extrapolate
@@
 
-\sixplot{base_heapsort_0.eps}{memory_tuned4_heapsort_0.eps}{memory_tuned_heapsort_0.eps}{memory_tuned_heapsort_1.eps}{}{}{base and memory-tuned heapsorts}
-
+\sixplot{base_heapsort_0.eps}{memory_tuned4_heapsort_0.eps}{memory_tuned_heapsort_0.eps}{memory_tuned_heapsort_1.eps}{}{}{base
+and memory-tuned heapsorts}
+%TESTING TODO
+TODO move to testing
 In order to explore the behaviour of the individual branches in
@@
 
-\textbf{[David] I am confused by this chart, because the labelling is
-a bit unclear. There appears to be a base heapsort, a 4-way cache
-heapsort and an 8-way cache heapsort. But the labelling doesn't tell
-us this.}
-
 Figure \ref{base and cache heapsorts-1}(a) shows the behaviour of each
@@
 
+%TODO
 \textbf{[David] We need to find some way to describe what each of these

### 2005-08-02 docs/tech_report/method.tex
minor changes
fixed small errors. didnt really get rid of any TODOs

@@
 
-% TODO maybe remoe this?
 Several steps were necessary to prepare for the research.  A framework was
@@
 provided an accurate measure of how fast a sort executes on a Pentium 4. The
-PapiEx homepage is at \n{http://icl.cs.utk.edu/~mucci/papiex/}.
+PapiEx homepage is at \n{http://icl.cs.utk.edu/\textasciitilde{}mucci/papiex/}.
 
@@
 the number of cache misses in the final merge steps of tiled mergesort by
-merging all the sorted arrays in one step. This required the use of an 8-heap,
-which had been developed as part of heapsort.
-%TODO this isnt what happened exactly
+merging all the sorted arrays in one step. This required the use of an k-ary
+heap, which had been developed as part of heapsort.
 
+%TODO this bit also has to be checked once we fix mergesort. if w fix mergesort
 Once these sorts were written it was attempted to improve them slightly.

### 2005-08-02 docs/tech_report/quick.tex
added testing comments
fixed small errors. didnt really get rid of any TODOs

@@
 In the case that the array to be sorted fits in the cache, then quicksort has
-excellent cache properties. Once it no longer fits in the cache, however, the
+excellent cache properties. This applies equally whether discussing a level 1
+cache or a level 2 cache. Once the array is larger than the cache, however, the
 number of misses increases dramatically. To fix this problem, LaMarca uses a
 technique similar to that used in multi-mergesort. Instead of multi-merging a
-series of sorted lists at the end, a \n{multi-way partition} is done at the start,
-moving all keys into cache sized containers which can be sorted individually.
+series of sorted lists at the end, a \n{multi-way partition} is done at the
+start, moving all keys into cache sized containers which can be sorted
+individually.
 
@@
 \subsection{Test Parameters}
-
+%TESTING
 The threshold chosen was 10, simply as this is the number used in Sedgewick's
@@
 
-Despite the sequentially search multi-quicksort performing better than its
+Despite the sequential search multi-quicksort performing better than its
 binary searched cousin, and its good cache performance, it still failed to
@@
 
-\sixplot{base_quicksort__median_of_3__0.eps}{base_quicksort__pseudo_median_of_9__0.eps}{memory_tuned_quicksort_0.eps}{multi_quicksort__binary_search__0.eps}{multi_quicksort__sequential_search__0.eps}{}{base,
+\sixplot{base_quicksort__median_of_3__0.eps}{memory_tuned_quicksort_0.eps}{multi_quicksort__binary_search__0.eps}{multi_quicksort__sequential_search__0.eps}{}{}{base,
 cache and multi quicksorts and base quicksort (pseudo-median-of-9)}
@@
 count increase is only that of removing a single check which occurs relatively
-few ($N / THRESHOLD$)\footnote{This is a minimum. The figure may be higher since
-partitions are likely to be smaller than the threshold.} times. Still, the
+few ($N / THRESHOLD$) times\footnote{This is a minimum. The figure may be higher since
+partitions are likely to be smaller than the threshold.}. Still, the
 problem is interesting and has applications in other algorithm designs.

### 2005-08-02 docs/tech_report/radix.tex
fixed small errors. didnt really get rid of any TODOs

@@
 
-\section{Cache Radixsort}
+\section{Memory-tuned Radixsort}
 Many improvements exist to radixsort, and some are suggested by Sedgewick and
@@
 
-Due to the reduced copying, memory-tuned radixsort should have a lower instruction
-count and fewer cache misses that base radixsort. There should be nine passes
-over the array, leading to just over one miss per key, as opposed to two misses
-per key with base radixsort. The complexity and branch prediction results
-should remain the same. The instruction count should, however, reduce by nearly
-a half to $\frac{9}{16}^ths$ of base radixsort.
+Due to the reduced copying, memory-tuned radixsort should have a lower
+instruction count and fewer cache misses that base radixsort. There should be
+nine passes over the array, leading to just over one miss per key, as opposed to
+two misses per key with base radixsort. The complexity and branch prediction
+results should remain the same. The instruction count should, however, reduce by
+nearly a half to $\frac{9}{16}^{th}s$ of base radixsort.
 
@@
 
-The instruction count of base radixsort, shown in Figure \ref{radix instructions} is
-very low. It is lower than even quicksort for the smallest set, and it
-flat-lines immediately (after slight amortisation of the loop overhead). The
-number of cache misses in Figure \ref{radix level 2 misses} is only slightly
+The instruction count of base radixsort, shown in Figure \ref{radix
+instructions} is very low. It is lower than even quicksort for the smallest set,
+and it flat-lines immediately (after slight amortisation of the loop overhead).
+The number of cache misses in Figure \ref{radix level 2 misses} is only slightly
 more than quicksort - due to being out-of-place - until the array no longer fits

### 2005-08-02 docs/tech_report/heap.tex
fixed up the references

@@
 keys fit in the cache without a corresponding increase in misses. For the same
-reason, the base heapsorts have no compulsory misses\footnote{See Section
-\ref{TODO}.}, while the out-of-place sorts have one compulsory miss in eight,
-when the auxiliary heap is being populated.
+reason, the base heapsorts have no compulsory misses, while the out-of-place
+sorts have one compulsory miss in eight, when the auxiliary heap is being
+populated.
 
@@
 
-Figure \ref{base and cache heapsorts-1}(a) shows the behaviour of each
+Figure \ref{base and memory-tuned heapsorts-1}(a) shows the behaviour of each
 branch in the base heapsort algorithm. The ``creation, has children'' branch
@@
 won't know.}
+
+%TODO read over the merged bits. might not be quite right
 

### 2005-08-02 docs/tech_report/quick.tex
fixed up the references

@@
 
-We use several variations of quicksort to test this theory. \ref{bently, i
-ithkn} suggests partitioning using a median-of-9 instead of a median-of-3. We
-created sorts using no median, a median-of-3, and pseudo-medians-of-5, -7 and
--9. The pseudo-medians used repeated medians-of-3 instead of a real
-median, as suggestion by \ref{bently, again, i think}. 
-%use a cut off
+We use several variations of quicksort to test this theory.
+\cite{BentleyMcIlroy93} suggests partitioning using a median-of-9 instead of a
+median-of-3. We created sorts using no median, a median-of-3, and
+pseudo-medians-of-5, -7 and -9. As suggested, our pseudo-medians used repeated
+medians-of-3 instead of a real median, in order to reduce the cost of their
+calculation. It was also suggested that once the partition sized were small
+enough, there was no more advantage to be gained by using a median of greater
+than three, and so a median-of-3 was used below a threshhold of 40.
 
@@
 6\% fewer branches, and a corresponsing decrease in mispredictions.
-%TODO to whom do we owe the idea of a pseudo median? Bently/McIlroy?
-%TODO when does it cut out, and why did we choose tere (benlty?)
 
@@
 increase in branch mispredictions.
+
+%TODO do we add more graphs here?
 

### 2005-08-02 docs/tech_report/radix.tex
fixed up the references

@@
 count, both of which slow down radixsort significantly. LaMarca suggests an
-improvement by Friend\footnote{See \ref{TODO friend[8]}.} that while the current
+improvement by Friend\footnote{See \cite{Friend56:152}.} that while the current
 byte is being put in place, the counting of the next byte can be done. This

### 2005-08-02 docs/tech_report/method.tex
fixed up the references
wrote about testing

--- 2005-08-01 docs/tech_report/method.tex
@@
 
-%TODO this is going to have to be updated
+%TODO this is going to have to be updated - or is it
 \begin{figure}
@@
 
+Our tests were designed so that we could compare our results to LaMarca's, and
+so our results true in the general case. Both our testing method and our data
+sets were chosen with these goals in mind.
+
+\subsection{SimpleScalar Simulations}
+
+We tested each of our sorts using SimpleScalar, using set sizes from 4096 to
+4194304, increasing by a factor of two each time. Each of these set sizes was
+simulated using \cc{sim-cache} and \cc{sim-bpred} in a variety of
+configurations.
+
+The SimpleScalar settings were chosen to be similar to LaMarca's simulations.
+LaMarca used the \n{Atom} simulator, which was used to test software for the DEC
+AlphaStation. It simulated a 2MB direct-mapped cache with a 32 byte cache line.
+Other settings were also chosen to correspond to the DEC Alpha: 8KB level 1 data
+cache, 8KB level 1 instruction cache and shared instruction and data level 2
+cache.
+
+In order to be able to extrapolate more information, we increased the number of
+cache simulations. We simulated both 2MB and 4MB caches, configured as both
+direct-mapped and fully-associative.
+
+We chose the a similar approach for our branch predictor caches. We aimed to
+simulate bimodal predictors as well as 2 level adaptive, in order to be able to
+compare them together. We simulated bimodal predictors with a 512, 1024 and
+2048 byte table, and 2 level adaptive predictors with 512, 1024, 2048 and 4096
+entries.
+
+\subsection{Data}
+
+Both our SimpleScalar simulations and our software predictor tests used truly
+random data from \n{http://www.random.org}. The software predictor used ten
+sections containing 4194304 \cc{unsigned integer}s, and the SimpleScalar
+simulations used the same ten sections. When SimpleScalar simulations were run
+on sized smaller than 4194304 keys, only the left hand side of the sections were
+used.
+
+\subsection{PaPiex}
+
+Our papiex tests ran significantly faster than either of our other simulations,
+and so we were able to run them over a significantly greater quantity of data.
+Each sort, and each size from 4096 to 4194304 key, was run 1024 times, using the
+system's random number generator to generate data, using seeds of between 0 and
+1023.
+
+%TODO did i say 128 byte elsewhere?
+These simulations were run a on Pentium 4 1.8Ghz, which had an 8-way, 256K level
+2 cache, with a 64-byte cache line. The level 1 caches are seperate 8K 4-way
+associative caches.
+
+The instruction pipeline of the Pentium 4 is 20 stages long, and so the branch
+misprediction penalty of the Pentium 4 approaches 20 cycles. The Pentium 4's
+static branch prediction strategy is backward taken and forward not taken; its
+dynamic strategy is unspecified: though \cite{Intel248966-010} mentions a branch
+history so it is likely a two-level adaptive predictor.
+
+\subsection{32-bit Integers}
+LaMarca used 64-bit \cc{integer}s, as his test machine was a 64-bit system.
+Since our machine is 32-bit, and SimpleScalar uses 32-bit addressing on our
+architecture, LaMarca's results will not be identical to ours. Rather, the
+number of cache misses in his results should be halved to be compared to our;
+when this is not the case, it is indicated.
+
+\subsection{Filling Arrays}
+
+It takes time to fill arrays, and this can distort the relationship between the
+results for small sizes, and larger sizes, for which this time is amortized. As
+a result, we measured this time, and subtracted it from our results. None of the
+graphs show this time.
+
+However, this practice also distorts the results. It removes compulsory cache
+misses from the results, so long as the data fits in the cache. When the data
+does not fit in the cache, the keys from the start of the array are faulted,
+with the effect that capacity misses reoccur, which are not discounted. As a
+result, on the left hand side of cache graphs, a number of compulsory misses are
+removed, though they are not on the right hand side. Similarly, LaMarca does not
+do this, and so this needs to be taken into account when comparing his results
+to ours.
+
 \section{Future Work}

### 2005-08-05 docs/tech_report/quick.tex
removed last unknown reference fixed grpahics

@@
 median-of-3, and pseudo-median-of-7 performs slightly better again, in
-terms of branch prediction. Figure \ref{TODO} shows the cycle count of each of
-the quicksorts, showing that a pseudo-median-of-7 offers the greatest
-trade-off between the reduced number of branches and instructions, and the
-increase in branch mispredictions.
+terms of branch prediction. Figure \ref{quick_median_cycles} shows the cycle
+count of each of the quicksorts, showing that a median-of-3 offers the greatest
+trade-off between the number of instructions and the increase in branch
+mispredictions.
 
-%TODO do we add more graphs here?
+\begin{figure}
+\includegraphics{plots/quick_median_cycles.eps}
+\caption{Empiric cycles count for quicksort with varying medians}
+\label{quick_median_cycles}
+\end{figure}
 
@@
 count increase is only that of removing a single check which occurs relatively
-few ($N / THRESHOLD$) times\footnote{This is a minimum. The figure may be higher since
-partitions are likely to be smaller than the threshold.}. Still, the
+few ($N / THRESHOLD$) times\footnote{This is a minimum. The figure may be higher
+since partitions are likely to be smaller than the threshold.}. Still, the
 problem is interesting and has applications in other algorithm designs.

### 2005-08-05 docs/tech_report/quick.tex
merged in david'd bit to quicksort. added a reference to mergresort. have yet to add replacement policy explanation

@@
 across the entire array at the end. This, however, is not implemented; instead a
-check is done to fetch a sentinel from the appropriate sub-file if required. This
-check isn't very expensive, and is significantly cheaper than either a bounds
-check or putting a sentinel into every sub-file.
+check is done to fetch a sentinel from the appropriate sub-file if required.
+This check isn't very expensive, and is significantly cheaper than either a
+bounds check or putting a sentinel into every sub-file.
 
@@
 
-In the case that the array to be sorted fits in the cache, then quicksort has
-excellent cache properties. This applies equally whether discussing a level 1
-cache or a level 2 cache. Once the array is larger than the cache, however, the
+In the case that the array to be sorted fits in a particular level of
+the cache, then quicksort has excellent cache properties at that
+level. Once the array is larger than the cache, however, the
 number of misses increases dramatically. To fix this problem, LaMarca uses a
@@
 \subsection{Test Parameters}
-%TESTING
-The threshold chosen was 10, simply as this is the number used in Sedgewick's
-book.
+
+When the size of a partition to be sorted is less than a certain
+threshold value, insertion sort is used instead of quicksort. The
+chosen threshold was 10, simply because this is the number used in
+Sedgewick's book.
 
@@
 
-LaMarca notes that quicksort is very cache efficient already. It has excellent
+LaMarca notes that quicksort is already very cache efficient as it has excellent
 spatial and temporal locality. Once it no longer fits in the cache, however, the
@@
 \ref{multi-quicksort (sequential search) level 2 misses} graph that this is true
-in the case of all the quicksorts.
+in the case of all the quicksorts. As in Section \ref{mergesort replacement
+policy}, this is due to the replacement policy of the fully-associative cache.
 
@@
 
-\section{Predictor Results}
+\section{Branch predictor results}
 
-%TODO a bit to rewrite
-\subsection{Different Sorts}
+\subsection{More detailed analysis of branch prediction}
+
+% \textbf{[David] I'm not sure what to do with this section. Part of the
+% problem is that the pictures are very difficult to understand. Another
+% problem is that we are trying to show differences that are almost
+% invisible. The median of 9 picture should definitely be removed from
+% this subsection. However, I suspect that the whole of this subsection should
+% go. What do we want to say? One really puzzling result is that there
+% seem to be no median branches at all for cache quicksort. This looks
+% strikingly like a programming error.}
+
+% \textbf{[David] On thing that's badly needed in this section is
+% a detailed description of what these funny charts all mean. However,
+% the appropriate place for most of this is the place where the first of
+% these charts appeared.}
+
+% \sixplot{base_quicksort_0.eps}{base_quicksort9_0.eps}{cache_quicksort_0.eps}{multi_quicksort_0.eps}{sequential_multi_quicksort_0.eps}{}{base, cache and multi quicksorts and base quicksort9}
 
 \sixplot{base_quicksort__median_of_3__0.eps}{memory_tuned_quicksort_0.eps}{multi_quicksort__binary_search__0.eps}{multi_quicksort__sequential_search__0.eps}{}{}{base,
-cache and multi quicksorts and base quicksort (pseudo-median-of-9)}
+memory-tuned and multi quicksorts}
 
-Firstly, memory-tuned quicksort and base quicksort (median-of-3) only differ in how they deal with
-the insertion. Base quicksort (median-of-3) and base quicksort
-(pseudo-median-of-9) have similar numbers of insertion branches, and a very
-similar ratio (2/1 correct/incorrect) of mispredictions. Comparing base
-quicksort to memory-quicksort (they both use
-median of 3), there is actually a decrease in the number of branches handled by
-insertion sort, instead of an increase. An increase is expected (thats why
-sedgewick moves the insertion to the end), though when you take into account
-the sentinel and the outer loop, there probably is an increase in instructions
-at least. Base quicksort has 35\% miss rate, but cache has a 34\%. Not a big
-difference, but the point is that there is a reduction in branches, but of the
-reduced branches, 50\% were unpredictable. I dont think theres anything
-interesting in this fact, however.
+Figure \ref{base, cache and multi quicksorts}(a) shows the behaviour
+of the indiviual branches in base quicksort. The $i$ branch is the
+loop control branch for the index sweeping from left-to-right up the
+array during partition. The $j$ branch is the corresponding loop
+sweeping down the array. The partition step terminates when $i \geq
+j$, and the $partition end$ branch checks that condition. The
+$insertion$ branch is the comparison branch in the insertion sort
+algorithm. Recall that our quicksort algorithm does not sort
+partitions which are smaller than 10 elements, but instead does a
+final pass of insertion sort over the whole array to sort these small
+partitions. Finally, the $median$ branch is the sum of the results for
+the branches involved in computing the median-of-three for each
+partition.
 
-When going to multi-quicksort, there si quite a drop in the number fo branches.
-Once the sort is larger than the cache, the number fo steps increases linearly
-(since it does a sort of counting like radix sort). Theres a decrease of 19 \%.
-Since the sort at this point is NlogM where M is the size of the cache (in
-keys), there shouldnt be a change in misprediction rates. And there sint.
-Before: 63\%. After, 63\%. So no difference.
+The only difference between cache quicksort and base quicksort is that
+cache quicksort applies insertion sort to small partitions
+immediately, whereas base quicksort does a single insertion sort pass
+at the end. The result is that both algorithms perform almost
+identically from the viewpoint of branch prediction. Cache quicksort
+uses a very slightly smaller number of insertion sort comparison
+branches, because it sorts only within partitions and does not attempt
+to move array elements that have acted as pivots. (The reason that
+quicksort works is that after partitioning, the pivot is in its
+correct location, and must never move again). In practice, this
+reduction in branches is probably offset by the increase in
+mispredictions in the outer loop exit branch of the many insertion
+sorts.
 
-Theres a lot of branches in the multi quicksort. WE saw in other results that
-the mere instruction cost of this bogs down to algorithm significantly. in the
-binary search version, there are more than twice as many mispredictions,
-despite the fact that there are less than half the branches. (note that i dont
-think this was a wonderful binary search, its heavily biased towards the
-checking the left). binary search predicts correctly 57 percent of the time,
-meaning that it mostly changes direction haphazardly during the search. No
-wonder it performs so badly.
+Base quicksort has a 35\% branch misprediction rate, whereas the rate
+for cache quicksort is only 34\%. The main reason for this is the
+reduction in branches by not attempting to move the pivot elements.
+Attempting to move these elements usually results in a branch
+misprediction. The main reason is that most other elements move at
+least one position, and thus cause the insertion sort loop to iterate
+at least once. But the pivot elements never move --- they are already
+in the correct place.
 
-Note that sequential multi-quicksort has less mispredictions than memory-tuned quicksort.
+When we look at multi-quicksort, there is a 19\% drop in the number of
+iterations of the $i$ and $j$ loops, with a corresponding reduction in
+the number of mispredictions. Interestingly, the percentage
+misprediction rate of the $i$ and $j$ branches does not change --- it
+is 37\% in both cases.
 
-%TODO to  here
+% Once the sort is larger than the cache, the number fo steps increases linearly
+% (since it does a sort of counting like radix sort). Theres a decrease of 19 \%.
+% Since the sort at this point is NlogM where M is the size of the cache (in
+% keys), there shouldnt be a change in misprediction rates. And there sint.
+% Before: 63\%. After, 63\%. So no difference.
 
+Multi-quicksort does, however, introduce a lot of additional
+branches. We already saw in Figure \ref{quick instructions}, that the
+binary search and sequential search variants increase the number of
+executed instructions by around 25\% and 50\% respectively. Figure
+\ref{base, cache and multi quicksorts} shows additional measures of the $binary
+left$ and $binary right$ branches, which are the branches within the
+binary search code. The number of binary search branches is much
+greater than the reduction in $i$ and $j$ branches. Furthermore, these
+binary search branches are highly (43\%) unpredictable.
+
+The variation of multi-quicksort that uses sequential search dramatically
+increases the number of executed branches, but these branches are highly
+predictable. This version actually causes fewer branch mispredictions
+than cache quicksort.
+
+% Theres a lot of branches in the multi quicksort. WE saw in other results that
+% the mere instruction cost of this bogs down to algorithm significantly. in the
+% binary search version, there are more than twice as many mispredictions,
+% despite the fact that there are less than half the branches. (note that i dont
+% think this was a wonderful binary search, its heavily biased towards the
+% checking the left). binary search predicts correctly 57 percent of the time,
+% meaning that it mostly changes direction haphazardly during the search. No
+% wonder it performs so badly.
+
+% Note that sequential multi-quicksort has less mispredictions than cache quicksort.
 \subsection{Different Medians}
@@
 \label{quicksort entropy}
+When we originally looked at quicksort, we expected that the $i$ and
+$j$ branches in quicksort would resolve in each direction. In fact,
+this is not the case. The main reason is that the chosen pivot is
+almost never the true median of the section of the array to be
+partitioned. Thus, the $i$ and $j$ branches are almost always biased,
+which makes them predictable. According to Knuth, if the pivot element
+is chosen randomly, then the $i$ and $j$ branches will be taken (i.e.
+the $i$ or $j$ loop will execute) twice as often as they are not taken
+(i.e. the loop exits).
 
-In Section \ref{heapsort entropy}, we observed that when an algorithm has more
-branches than are strictly necessary complete, these extra branches are
-predicted correctly. In quicksort, we expect to obtain similar results.
+In order to investigate this, we implemented a simple version of
+quicksort that uses the middle element as the pivot rather than a
+median-of-three\footnote{It is important to note that selecting
+the pivot based on only a single element can be particularly inefficient
+because it removes the natural sentinels at the top and bottom
+of the regions to be partitioned that we get from median-of-three. In
+Figure \ref{base quicksort without a median, with median of 3, and with pseudo medians of 5, 7 and 9} we see that this results in an additional branch to handle
+the edgecase without a sentinel. Although this branch is almost 100\%
+predictable, it is executed a very large number of times.}. The arrays we sort are random, so the value of this
+element will also be random. Figure \ref{base quicksort without a median, with median of 3, and with pseudo medians of 5, 7 and 9}(a)
+shows the behaviour of each of the important branches in this code. We
+see that the $i$ and $j$ branches are indeed taken about twice as
+often as not-taken, and the $i$ and $j$ branches are correctly
+predicted around 71\% of the time.
 
-%HOW
+The result that the branches are more predictable (71\%) than they are
+biased (66\%) is a particularly interesting finding. The reason is that
+the bias in the results is an average figure over a large number of
+partitions that quicksort performs to sort the array. On a particular
+partition, the bias will be more pronounced because the $i$ and $j$
+branches will be biased in opposite directions.
 
-We use several variations of quicksort to test this theory.
-\cite{BentleyMcIlroy93} suggests partitioning using a median-of-9 instead of a
-median-of-3. We created sorts using no median, a median-of-3, and
-pseudo-medians-of-5, -7 and -9. As suggested, our pseudo-medians used repeated
-medians-of-3 instead of a real median, in order to reduce the cost of their
-calculation. It was also suggested that once the partition sized were small
-enough, there was no more advantage to be gained by using a median of greater
-than three, and so a median-of-3 was used below a threshhold of 40.
+For example, consider the case where the pivot element is larger than
+the true median. In this case, the final resting place of the pivot
+will be towards the right-hand end of the region to be
+partitioned. Thus, the $i$ loop will execute a large number of times,
+and thus the $i$ loop will be biased towards being taken. The $j$
+loop, on the other hand, will be biased toward not-taken, because the
+average element is likely to be smaller than the pivot. Therefore, the
+$j$ loop will execute only a small number of times. Thus, we see that
+in a particular partition, we do not normally get an equal number of
+executions of the $i$ and $j$ loops. Instead we get a large number of
+executions in one loop (which makes its branch relatively
+predictable), whereas the other loop often does not enter at all
+(which makes its branch also predictable, although executed a much
+smaller number of times). Thus, although overall the $i$ and $j$
+branches are taken twice as often as not-taken, on a particular
+partition one will usually be biased in the not-taken direction, and
+as a result will be executed a relativelty small number of times.
 
-We measured the number of branches and mispredictions due to the partitioning,
-the final insertion sort, and the partitioning. It is expected to have little or
-no variance in either the number of branches and misses due to insertion sort,
-but that less time will be spent partitioning to arrays as a better median is
-found, and a corresponding increase in time spent finding the better median.
+In the preceding paragraphs we have seen that the $i$ and $j$ branches
+are predictable when we have a pivot that is larger or smaller than
+the true median of the array section to be partitioned. An interesting
+question is the effect of choosing a better approximation of the true
+median on branch prediction. Figure \ref{base quicksort without a
+median, with median of 3, and with pseudo medians of 5, 7 and 9}(b)
+shows the behavior of these branches when median-of-three is used to
+select the pivot. Perhaps the biggest effect is that the number of
+executions of the $i$ and $j$ branches falls by around 15\%. This fall
+is partly offset by the comparison branches to compute the median, but
+these account for only around 1.25\% of the original branches from the
+no-median quicksort. Thus, there is a reduction of almost 14\% in the
+number of executed comparison branches. Even more interesting, however,
+is the bias in the remaining $i$ and $j$ branches. The no-median $i$ and
+$j$ branches are 66\% biased, but the median-of-three branches are
+only 60\% biased in the taken direction. The accuracy with which
+these branches are correctly predicted falls from 71\% to 64\%.
 
-The results of these tests are shown in Figure \ref{Branch prediction
-performance for base quicksort without a median, with median of 3, and with
-pseudo-medians of 5, 7 and 9}. There is little variance in insertion sort misses
-between the sorts using pseudo-medians, but the median-of-3 quicksort has
-6\% fewer branches, and a corresponsing decrease in mispredictions.
+The reason for the reduction in bias and prediction accuracy is that
+median-of-three gives a much better estimate of the true median of the
+section to be partitioned. The final resting place of the pivot will
+usually be closer to the middle of the section, to there is less bias
+in the branches. Thus, although median-of-three gives around a 14\%
+reduction in executed comparison branches, each of those branches is
+much less likely to be correctly predicted. Overall there is actually
+a 6\% \textit{increase} in the total number of branch mispredictions.
 
-Between a median-of-3 and a pseudo-median-of-9 partitioning, there is a
-10\% reduction in branches. At the same time, there is a 20\% reduction in
-correct predictions, accounting for the entire reduction in branches, and a 4\%
-increase in mispredictions. This is due to better partitioning. Less time is
-spent copying keys unnecessarily, resulting in a less predictable branch. The
-corresponding increase in mispredictions while calculating the median equates to
-50\% of the extra mispredictions due to partitioning.
+To further investigate this phenomenon, we measured the effects of
+using pseudo-medians-of-five, -seven and -nine to choose the pivot. A
+pseudo-median uses repeated medians-of-three for parts of the
+computation to reduce the number of comparisons in computing the
+pivot. The reductions in executed comparison branches are 16.5\%,
+17.8\% and 19.6\% respectively. However, the biases of the $i$ and $j$
+branches in these algorithms are only 57.8\%, 56.3\% and 55.1\%. The
+corresponding predictabilities are 61.6\%, 59.9\% and 58.3\%. Overall,
+the increases in comparison branch mispredictions (including computing
+the pseudo-median) are 9.2\%, 11.8\% and 14.3\%.
 
-This indicates the same pattern as we saw during heapsort.%
-% and mergesort? TODO
-The less entropy due to the partitioning key, the less wasted exchanging of keys,
-which leads to a general decrease in branches. This decrease means that the
-fewer keys are exchanged due to partitioning, leading to less predictable
-partitioning. In effect, there are fewer branches pretending to be useful.
+Picking a better estimate of the true median substantially reduces the
+number of executed branches, but each of those branches is much more
+unpredictable. Intuitively, this makes sense. Sorting the array
+involves putting an order on it, and thus reducing its entropy. If we
+want to sort it in a smaller number of steps, then each step must
+remove more randomness. Thus, we would expect that the outcome of the
+branch that removes the randomness would itself be more random.
 
-The pseudo-median-of-5 partitioning performs slightly better than the
-median-of-3, and pseudo-median-of-7 performs slightly better again, in
-terms of branch prediction. Figure \ref{quick_median_cycles} shows the cycle
-count of each of the quicksorts, showing that a median-of-3 offers the greatest
-trade-off between the number of instructions and the increase in branch
-mispredictions.
+Figure \ref{TODO} shows the cycle count of each of the quicksorts,
+showing that a pseduo-median-of-seven offers the greatest tradeoff
+between the reduced number of branches and instructions, and the
+increase in branch mispredictions.
 
+An additional advantage of using a higher order median is that it
+dramatically reduces the probability of quicksort exhibiting worst
+case ($O(N^2)$) time complexity. Although the average case reduction
+in running time is small, reducing the chance of worst-case
+performance, which might cause the program to pause for minutes or
+hours, is very valuable.
+
+%TODO add a reference to this
 \begin{figure}
@@
 
-The memory-tuned quicksort is the only algorithm here that requires a sentinel
+Memory-tuned quicksort is the only algorithm here that requires a sentinel
 check. The branch predictor will only mispredict this once, so the instruction

### 2005-08-09 docs/tech_report/quick.tex
moved some things off the TODO list

@@
 Quicksort has been well studied, and several important improvements have been
-noted. In \cite{BentleyMcIlroy93}, using a median-of-9 partition is discussed,
-and modern discussions on quicksort refer to the problem of equal keys, which is
-discussed there, as well as by Sedgewick in \cite{SedgewickBentley02}. These
-improvements are not included here for several reasons. Firstly, a quicksort
-which takes into account equal keys puts the equal keys at the front of the
-array, then copies them into place later. These copies would represent a lot of
-cache misses, and the results would overwhelm the improvements LaMarca makes in
-the next phase of the algorithm. Secondly, when sorting random keys over a
-large range, the number of equal keys will be very low. Using a median of
-greater than three is discussed in Section \ref{quick-predictors2}.
+noted. In \cite{BentleyMcIlroy93}, using a median-of-9 partition is discussed.
+This is discussed below. Modern discussions on quicksort refer to the problem of
+equal keys, which is discussed there, as well as by Sedgewick in
+\cite{SedgewickBentley02}. These improvements are not included here for several
+reasons. Firstly, a quicksort which takes into account equal keys puts the equal
+keys at the front of the array, then copies them into place later. These copies
+would represent a lot of cache misses, and the results may overwhelm the
+improvements LaMarca makes in the next phase of the algorithm. Secondly, when
+sorting random keys over a large range, the number of equal keys will be very
+low. Using a median of greater than three is discussed in Section
+\ref{quick-predictors2}.
 
@@
 
+%TODO
 % \textbf{[David] I'm not sure what to do with this section. Part of the

### 2005-08-09 docs/tech_report/shell.tex
minor edits
fixed small errors. didnt really get rid of any TODOs
moved some things off the TODO list

@@
 Since first published, there has been much discussion about the best
-\textit{increments} to use in shellsort. Shell's original sort, described above,
+\n{increments} to use in shellsort. Shell's original sort, described above,
 has the increments $\frac{N}{2}$, $\frac{N}{4}$, $\frac{N}{8}$, $\frac{N}{16}$
@@
 to test the performance of shellsort. This uses increments which are
-$\frac{5}{11}$ths of the previous increment.
+$\frac{5}{11}^{th}s$ of the previous increment.
 
@@
 \subsection{Expected Performance}
-Shellsort has been much analysed, and it has been shown that it's complexity is
-difficult to prove\footnote{See TODOA, TODOB, TODOC and TODOD.}. The instruction
-count and cycle counts are expected to be similar to $O(NlogN)$, as this has
-been indicated by previous experiment.
+Shellsort is designed to perform similarly to an $O(NlogN)$ sort. The instruction
+count and cycle counts are therefore expected to be similar to $O(NlogN)$.
 
@@
 performs better than a heapsort, but considerably worse than quicksort. However,
-it easier to program than either of these two sort.
+it easier to program than either of these two sorts.
 
@@
 
+%TODO see figure something
 In shellsort, almost exactly half of the important branches are mispredicted.

### 2005-08-13 docs/tech_report/heap.tex
changes for testing

@@
 An 8-heap has seven comparisons between children, and will probably have about
-3 %TODO does this mean that knuths harmonics applies to this sort of prediction
-% thing TODO
-mispredictions, as it behaves similarly to selection sort (See \ref{selection
-branches logn}). It is possible, therefore, that the 8-heap will have slightly
-fewer mispredictions due to the combination of the shallower heap and the larger
-fan-out.
+2.58 mispredictions, as it behaves similarly to selection sort (See \ref{selection
+branches harmonics}). It is possible, therefore, that the 8-heap will have slightly
+fewer mispredictions (about 36\%) due to the combination of the shallower heap
+and the larger fan-out.
 
@@
 and memory-tuned heapsorts}
-%TESTING TODO
-TODO move to testing
-In order to explore the behaviour of the individual branches in
-heapsort, we performed a second set of branch prediction
-experiments. Rather than using SimpleScalar, we instead added
-additional statements to the heapsort source code to simulate a
-separate bimodal predictor for each branch. This allowed us to measure
-each branch individually, whereas SimpleScalar gives only a total
-figure for all branches. An additional advantage is that the resulting
-compiled simulation is much faster than a full microarchitecture
-simulator, and we were thus able to take the average of thousands of
-runs (\textbf{TODO find out exact number}), rather than the small
-number of runs possible with SimpleScalar. For these experiments,
-randomly generated array of size \textbf{xxxxx} were used.
 
 Figure \ref{base and memory-tuned heapsorts-1}(a) shows the behaviour of each
-branch in the base heapsort algorithm. The ``creation, has children'' branch
-is from line \textbf{xxxxx} of Figure \ref{Simple Base Heapsort}.
+branch in the base heapsort algorithm. The `\cc{Creation, has children}' branch
+refers to the check as to whether a key has children, during the creation of the
+heap The `\cc{Creation, child comparison}' branch is the comparison between the
+two children in the 2-heap, and the `\cc{Creation, child promotion}' refers to
+checking if the child is larger than it's parent. The same checks are later made
+during the destruction of the heap, and are referred to as `\cc{Destruction, has
+children}', `\cc{Destruction, child comparison}' and `\cc{Desruction child
+promotion}'.
 
-%TODO 
-\textbf{[David] We need to find some way to describe what each of these
-branches is. I don't know myself, so the general reader really, really
-won't know.}
+In the memory-tuned heapsort, the creation phase is simpler, involving merely
+one branch, called `\cc{fix-up}' here. The other columns refer to the
+destruction phase, where most of the work takes place in memory-tuned heapsoirt.
+When a 4-heap is now used, there are three comparisons instead of just one.
+Likewise, in an (8-heap), there are seven comparisons.
 

### 2005-08-13 docs/tech_report/merge.tex
merged in david'd bit to quicksort. added a reference to mergresort. have yet to add replacement policy explanation
changes for testing

@@
 \subsection{Test Parameters}
-%TESTING
-The level 1 cache is 8k, so 1024 keys from each array can be sorted in the level
+The level 1 cache is 8K, so 1024 keys from each array can be sorted in the level
 1 cache. 262144 keys can be sorted in the level 2 cache. When arrays were
@@
 multi-mergesort; since it does a k-way merge into the source array, the sorted
-cache sized segments need to be in the auxiliary array. An 8-heap was used for
+cache-sized segments need to be in the auxiliary array. An 8-heap was used for
 the multi-mergesorts.
@@
 
+\label{mergesort replacement policy}
 It is notable that our prediction of half a miss per key for multi-mergesort,
@@
 being the fastest. Note that the alignment of the mergesort may not work as
-%TESTING
 desired due to the size of the cache. This was optimized for a 2MB cache, while

### 2005-08-13 docs/tech_report/method.tex
changes for testing

@@
 
-%TODO this is going to have to be updated - or is it
 \begin{figure}
@@
 
-%TESTING
-%TODO a lot of this may be in the first sorts chapter, maybe heapsort.
-%discuss: BIG + random.org, 64 bit in lamarca vs 32 bit, do_nothing and how it
-%affects cache hits, scaling in the software branch predictors. How to compare
-%and contrast varying results (branch vs softwre branhc; P4 vs simplescalar).
-%What factors are used for testing: /proc/cpuinfo output, cahce and branch
-%predictor data from P4, simplescalar inputs, software predictor, verfication of
-%sorts.
-
 Our tests were designed so that we could compare our results to LaMarca's, and
-so our results true in the general case. Both our testing method and our data
-sets were chosen with these goals in mind.
+so that our results would hold in the general case. Both our testing method and
+our data sets were chosen with these goals in mind.
 
@@
 
-We chose the a similar approach for our branch predictor caches. We aimed to
+We chose a similar approach for our branch predictor caches. We aimed to
 simulate bimodal predictors as well as 2 level adaptive, in order to be able to
@@
 entries.
+
+\subsection{Software Predictor}
+In order to explore the behaviour of the individual branches, we performed a
+second set of branch prediction experiments. Rather than using SimpleScalar, we
+instead added additional statements to the source code to simulate a separate
+bimodal predictor for each interesting branch. This allowed us to measure each
+branch individually, whereas SimpleScalar gives only a total figure for all
+branches.
+
+The software predictor results are averaged over ten runs, using arrays
+containing 4194304 keys. To compare these results to the SimpleScalar ones, it
+is necessary to divide the number of branch mispredictions by 4194304.
+
+\subsection{PaPiex}
+Our papiex tests ran significantly faster than either of our other simulations,
+and so we were able to run them over a significantly greater quantity of data.
+Each sort and each size from 4096 to 4194304 keys was run 1024 times, using the
+system's random number generator to generate data, using seeds of between 0 and
+1023. The results displayed are averaged over the 1024 runs.
+
+These simulations were run a on Pentium 4 1.6GHz, which had an 8-way, 256K level
+2 cache, with a 64-byte cache line. The level 1 caches are 8K 4-way associative
+caches, with seperate instruction and data level 1 caches.
+
+The instruction pipeline of the Pentium 4 is 20 stages long, and so the branch
+misprediction penalty of the Pentium 4 approaches 20 cycles. The Pentium 4's
+static branch prediction strategy is backward taken and forward not taken; its
+dynamic strategy is unspecified: though \cite{Intel248966-010} mentions a branch
+history so it is likely to be a two-level adaptive predictor.
 
@@
 
-\subsection{PaPiex}
-
-Our papiex tests ran significantly faster than either of our other simulations,
-and so we were able to run them over a significantly greater quantity of data.
-Each sort, and each size from 4096 to 4194304 key, was run 1024 times, using the
-system's random number generator to generate data, using seeds of between 0 and
-1023.
-
-%TODO did i say 128 byte elsewhere?
-These simulations were run a on Pentium 4 1.8Ghz, which had an 8-way, 256K level
-2 cache, with a 64-byte cache line. The level 1 caches are seperate 8K 4-way
-associative caches.
-
-The instruction pipeline of the Pentium 4 is 20 stages long, and so the branch
-misprediction penalty of the Pentium 4 approaches 20 cycles. The Pentium 4's
-static branch prediction strategy is backward taken and forward not taken; its
-dynamic strategy is unspecified: though \cite{Intel248966-010} mentions a branch
-history so it is likely a two-level adaptive predictor.
-
 \subsection{32-bit Integers}
@@
 architecture, LaMarca's results will not be identical to ours. Rather, the
-number of cache misses in his results should be halved to be compared to our;
+number of cache misses in his results should be halved to be compared to ours;
 when this is not the case, it is indicated.
@@
 It takes time to fill arrays, and this can distort the relationship between the
-results for small sizes, and larger sizes, for which this time is amortized. As
-a result, we measured this time, and subtracted it from our results. None of the
-graphs show this time.
+results for small set sizes, and larger set sizes, for which this time is
+amortized. As a result, we measured this time, and subtracted it from our
+results.
 
@@
 result, on the left hand side of cache graphs, a number of compulsory misses are
-removed, though they are not on the right hand side. Similarly, LaMarca does not
-do this, and so this needs to be taken into account when comparing his results
-to ours.
+removed, though they are not on the right hand side. LaMarca does not do this,
+and so this needs to be taken into account when comparing his results to ours.
 
 \section{Future Work}
-An improvement that would have yielded interesting results would be to do simulations
-which only consisted of flow control, with no comparisons. This would allow a
-division of flow control misses and comparative misses. However, this would not
-always be possible; sorts like quicksort and heapsort use their data to
-control the flow of the program. In addition, bubblesort, shakersort and
+An improvement that would have yielded interesting results would be to do
+simulations which only consisted of flow control, with no comparisons. This
+would allow a division of flow control misses and comparative misses. However,
+this would not always be possible; sorts like quicksort and heapsort use their
+data to control the flow of the program. In addition, bubblesort, shakersort and
 selection sort have (analytically) predictable flow control, while radixsort has
-no comparison misses at all. Mergesort may benefit from this, however, especially versions
-with difficult flow control diagrams, such as \n{algorithm N} and \n{algorithm
-S} (see Section \ref{Algorithm N}).
+no comparison misses at all. Mergesort may benefit from this, however,
+especially versions with difficult flow control diagrams, such as \n{algorithm
+N} and \n{algorithm S} (see Section \ref{Algorithm N}).
 
@@
 sort may indicate areas for potential improvement.
-
-As well as the number of cycles, Pentium 4 performance registers can also provide
-most of the same information as SimpleScalar, with different parameters. For
-example, the cache line of a Pentium 4 machine is 128 bytes whereas 32 byte
-cache lines were used in SimpleScalar. The addition of real world data would
-strengthen our results.
-%TODO not true. we have it

### 2005-08-13 docs/tech_report/quick.tex
changes for testing

@@
 
-\subsection{More detailed analysis of branch prediction}
+\subsection{A more detailed look at branch prediction}
 
@@
 
-Figure \ref{TODO} shows the cycle count of each of the quicksorts,
+Figure \ref{quick_median_cycles} shows the cycle count of each of the quicksorts,
 showing that a pseduo-median-of-seven offers the greatest tradeoff
@@
 
-%TODO add a reference to this
 \begin{figure}

### 2005-08-13 docs/tech_report/merge.tex
fixed up the graphs, involving changes to code, gnuplot scripts, tex files and the README

@@
 
-\sixplot{algorithm_n_0.eps}{algorithm_s_0.eps}{base_mergesort_0.eps}{double_aligned_tiled_mergesort_0.eps}{double_aligned_multi_mergesort_0.eps}{double_aligned_multi_mergesort_0.eps}{mergesorts}
+\sixplot{algorithm_n_0.eps}{algorithm_s_0.eps}{base_mergesort_0.eps}{double_aligned_tiled_mergesort_0.eps}{double_aligned_multi_mergesort_0.eps}{double_aligned_multi_mergesort_1.eps}{mergesorts}
 

### 2005-08-13 docs/tech_report/quick.tex
fixed up the graphs, involving changes to code, gnuplot scripts, tex files and the README

@@
 
-% \sixplot{base_quicksort_0.eps}{base_quicksort9_0.eps}{cache_quicksort_0.eps}{multi_quicksort_0.eps}{sequential_multi_quicksort_0.eps}{}{base, cache and multi quicksorts and base quicksort9}
-
 \sixplot{base_quicksort__median_of_3__0.eps}{memory_tuned_quicksort_0.eps}{multi_quicksort__binary_search__0.eps}{multi_quicksort__sequential_search__0.eps}{}{}{base,
-memory-tuned and multi quicksorts}
+memory-tuned and multi- quicksorts}
 

### 2005-08-13 code/main.c

@@ -235,8 +235,8 @@ main(int argc, char** args)
 	test_sort(algorithm_n, "algorithm N");
 	test_sort(algorithm_s, "algorithm S");
 	test_sort(base_mergesort, "base mergesort");
-*/	test_sort(tiled_mergesort, "tiled mergesort");
-/*	test_sort(multi_mergesort, "multi-mergesort");
+	test_sort(tiled_mergesort, "tiled mergesort");
+	test_sort(multi_mergesort, "multi-mergesort");
 	test_sort(double_aligned_tiled_mergesort, "double-aligned tiled mergesort");
 	test_sort(double_aligned_multi_mergesort, "double-aligned multi-mergesort");
 
@@ -289,8 +289,8 @@ main(int argc, char** args)
 	time_sort(algorithm_n, "algorithm N");
 	time_sort(algorithm_s, "algorithm S");
 	time_sort(base_mergesort, "base mergesort");
-*/	time_sort(tiled_mergesort, "tiled mergesort");
-/*	time_sort(multi_mergesort, "multi-mergesort");
+	time_sort(tiled_mergesort, "tiled mergesort");
+	time_sort(multi_mergesort, "multi-mergesort");
 	time_sort(double_aligned_tiled_mergesort, "double-aligned tiled mergesort");
 	time_sort(double_aligned_multi_mergesort, "double-aligned multi-mergesort");
 
@@ -352,9 +352,9 @@ main(int argc, char** args)
 	predictor_run(memory_tuned_heapsort, 11, "memory-tuned heapsort");
 	predictor_run(memory_tuned4_heapsort, 7, "memory-tuned4 heapsort");
 
-	predictor_run(algorithm_n, 8, "algorithm N");
+*/	predictor_run(algorithm_n, 8, "algorithm N");
 	predictor_run(algorithm_s, 10, "algorithm S");
-	predictor_run(base_mergesort, 33, "base mergesort");
+/*	predictor_run(base_mergesort, 33, "base mergesort");
 	predictor_run(tiled_mergesort, 46, "tiled mergesort");
 	predictor_run(multi_mergesort, 59, "multi-mergesort");
 	predictor_run(double_aligned_tiled_mergesort, 46, "double-aligned tiled mergesort");

### 2005-08-13 code/sorts/algorithm_s.c

@@ -173,4 +173,14 @@ S13:
 	}
 
 	free(aux);
+
+	/* clear uninteresting predictors */
+	init_predictor(&global_predictor[0]);
+	init_predictor(&global_predictor[3]);
+	init_predictor(&global_predictor[4]);
+	init_predictor(&global_predictor[6]);
+	init_predictor(&global_predictor[7]);
+	init_predictor(&global_predictor[8]);
+	init_predictor(&global_predictor[9]);
+
 }

### 2005-08-14 docs/tech_report/report.tex
some spell checks, removed unneccessary files, rename elementary
more minor edits.
minor changes mergeed in david's heap changes
changes for testing
fixed abstract and title page

@@
-\documentclass[dvips]{report}
+\documentclass{report}
 \usepackage{ifthen}
@@
 
+%cc is used for
+% - unix commans or programs: gcc, sim-bpred etc
 \newcommand{\cc}[1]{\texttt{#1}}
+
+%n is used for
+% - proper names: Atom, SimpleScalar, Papi
 \newcommand{\n}[1]{\textit{#1}}
+
 \newcommand{\code}[2]{\begin{figure}\lstinputlisting{code/#2}\caption{#1}\label{#1}\end{figure}}
@@
 
-\title{Fast Sorting on Modern Computers}
-\author{Paul Biggar \and David Gregg}
-\date{}
-
 \bibliographystyle{these}
@@
        \begin{center}
-                \vspace*{95mm}
-                \LARGE Fast Sorting on Modern Computers\\
+				\vspace*{40mm}
+                \Huge Sorting in the Presence of Branch Prediction and Caches\\
                 \vspace{1em}
-                \large Paul Biggar, David Gregg\\
-                \large Department of Computer Science\\
-				\large University of Dublin, Trinity College
+                \Large Fast Sorting on Modern Computers\\
+                \vspace{2em}
+                \Large Paul Biggar \hspace{1cm} David Gregg\\
+				\vspace{2em}
+				\large Technical Report TR-CS-2005-XX\\
+                \large Department of Computer Science,\\
+				\large University of Dublin, Trinity College,\\
+				\large Dublin 2, Ireland.
         \end{center}
@@
 \begin{abstract}
-This report describes the creation and analysis of sorts designed to take
-advantage of features of modern processors. The work on cache-conscious sorting
-by LaMarca and Ladner is the starting point; this is implemented and analysed
-using a hardware simulator. Results relating to a variety of cache and branch
-predictor configurations are obtained and analysed for a variety of sorts,
-beginning with insertion and selection sorts, bubblesort and shakersort, and
-continuing to heapsort, mergesort, quicksort and radixsort. Actual and expected
-results are compared together and to the results of LaMarca's research.
+Sorting is one of the most important and studied problems in computer
+science. Many good algorithms exist which offer various trade-offs in
+efficiency, simplicity and memory use. However most of these
+algorithms were discovered decades ago at a time when computer
+architectures were much simpler than today. Branch prediction and
+cache memories are two developments in computer architecture that have
+a particularly large impact on the performance of sorting algorithms.
+
+This report describes a study of the behaviour of sorting algorithms
+on branch predictors and caches. Our work on branch prediction is
+almost entirely new, and finds a number of important results. In
+particular we show that insertion sort causes the fewest branch
+mispredictions of any comparison-based algorithm, that optimizations,
+such as the choice of the pivot in quicksort, can have a large impact
+on the predictability of branches, and that advanced two-level branch
+predictors are usually worse at predicting branches in sorting
+algorithms than simpler branch predictors. In many cases it is
+possible to draw links between classical theoretical analyses of
+algorithms and their branch prediction behaviour.
+
+The other main work described in this report is an analysis of the
+behaviour of sorting algorithms on modern caches. Over the last decade
+there has been considerable interest in optimizing sorting algorithms
+to reduce the number of cache misses. We experimentally study the
+cache performance of both classical sorting algorithms, and a variety
+of cache-optimized algorithms, primarily those proposed by LaMarca and
+Ladner. Our experiments cover a much wider range of algorithms than
+other work, including the $O(N^2)$ sorts and shellsort, all within a
+single framework. We discover a number of new results, particularly
+relating to the branch prediction behaviour of cache-optimized sorts.
+
+We also developed a number of other improvements to the algorithms,
+such as removing the need for a sentinel in heapsort. Overall, we
+found that a cache-optimized radix sort was the fastest algorithm for
+all large arrays; the absence of comparison branches means that the
+algorithm causes almost no branch mispredictions.
 \end{abstract}
@@
 
-\input ordernsquared.tex
+\input elementary.tex
 

### 2005-08-14 docs/tech_report/elementary.tex
fixed up the references
changes for testing
fixed all references

@@
 \subsection{Testing}
-%TESTING
+
 All elementary sorts are only sorted up to 65536 keys in the simulations, and
@@
 
-\label{selection branches logn}
+\label{selection branches harmonics}
 The branch prediction performance is not expected to be bad. Flow
@@
 
-Knuth \cite{Knuth1} analyses the number of changes to the
+Knuth \cite{Knuth97} analyses the number of changes to the
 left-to-right minimum when searching for the minimum element of an
@@
 
-%TODO check that this is put in after the titlepage
-%TODO check that the captions are the same as the captions at the end of the
-%other chapters
 % I'm doing a slightly different result here. Level 2 doesn't interest us, so put

### 2005-08-14 docs/tech_report/heap.tex
fixed all references

@@
 
-\section{A more detailed look at branch prediction}
+\section{A More Detailed Look at Branch Prediction}
 
@@
 More importantly, the accuracy of the each successive comparison increases in
-the same way as selection sort (see Section \ref{selection branches logn}),
+the same way as selection sort (see Section \ref{selection branches harmonics}),
 becoming more predictable as the fanout increases. For example, in the 8-heap,

### 2005-08-14 docs/tech_report/quick.tex
fixed all references

@@
 
-\section{Branch predictor results}
-
-\subsection{A more detailed look at branch prediction}
+\section{A More Detailed Look at Branch Prediction}
 
@@
 
-Figure \ref{base, cache and multi quicksorts}(a) shows the behaviour
-of the indiviual branches in base quicksort. The $i$ branch is the
-loop control branch for the index sweeping from left-to-right up the
-array during partition. The $j$ branch is the corresponding loop
-sweeping down the array. The partition step terminates when $i \geq
-j$, and the $partition end$ branch checks that condition. The
-$insertion$ branch is the comparison branch in the insertion sort
-algorithm. Recall that our quicksort algorithm does not sort
-partitions which are smaller than 10 elements, but instead does a
-final pass of insertion sort over the whole array to sort these small
-partitions. Finally, the $median$ branch is the sum of the results for
-the branches involved in computing the median-of-three for each
-partition.
+Figure \ref{base, memory-tuned and multi- quicksorts-1} shows the behaviour of
+the indiviual branches in base quicksort. The \cc{i} branch is the loop control
+branch for the index sweeping from left-to-right up the array during partition.
+The \cc{j} branch is the corresponding loop sweeping down the array. The
+partition step terminates when $i \geq j$, and the \cc{partition end} branch
+checks that condition. The \cc{insertion} branch is the comparison branch in the
+insertion sort algorithm. Recall that our quicksort algorithm does not sort
+partitions which are smaller than 10 elements, but instead does a final pass of
+insertion sort over the whole array to sort these small partitions. Finally, the
+\cc{median} branch is the sum of the results for the branches involved in
+computing the median-of-three for each partition.
 
@@
 When we look at multi-quicksort, there is a 19\% drop in the number of
-iterations of the $i$ and $j$ loops, with a corresponding reduction in
-the number of mispredictions. Interestingly, the percentage
-misprediction rate of the $i$ and $j$ branches does not change --- it
-is 37\% in both cases.
+iterations of the \cc{i} and \cc{j} loops, with a corresponding reduction in the
+number of mispredictions. Interestingly, the percentage misprediction rate of
+the \cc{i} and \cc{j} branches does not change --- it is 37\% in both cases.
 
@@
 
-Multi-quicksort does, however, introduce a lot of additional
-branches. We already saw in Figure \ref{quick instructions}, that the
-binary search and sequential search variants increase the number of
-executed instructions by around 25\% and 50\% respectively. Figure
-\ref{base, cache and multi quicksorts} shows additional measures of the $binary
-left$ and $binary right$ branches, which are the branches within the
-binary search code. The number of binary search branches is much
-greater than the reduction in $i$ and $j$ branches. Furthermore, these
-binary search branches are highly (43\%) unpredictable.
+Multi-quicksort does, however, introduce a lot of additional branches. We
+already saw in Figure \ref{quick instructions}, that the binary search and
+sequential search variants increase the number of executed instructions by
+around 25\% and 50\% respectively. Figure \ref{base, memory-tuned and multi-
+quicksorts-3} shows additional measures of the \cc{binary left} and \cc{binary
+right} branches, which are the branches within the binary search code. The
+number of binary search branches is much greater than the reduction in \cc{i}
+and \cc{j} branches. Furthermore, these binary search branches are highly (43\%)
+unpredictable.
 
@@
 \label{quicksort entropy}
-When we originally looked at quicksort, we expected that the $i$ and
-$j$ branches in quicksort would resolve in each direction. In fact,
-this is not the case. The main reason is that the chosen pivot is
-almost never the true median of the section of the array to be
-partitioned. Thus, the $i$ and $j$ branches are almost always biased,
-which makes them predictable. According to Knuth, if the pivot element
-is chosen randomly, then the $i$ and $j$ branches will be taken (i.e.
-the $i$ or $j$ loop will execute) twice as often as they are not taken
-(i.e. the loop exits).
+When we originally looked at quicksort, we expected that the \cc{i} and \cc{j}
+branches in quicksort would resolve in each direction. In fact, this is not the
+case. The main reason is that the chosen pivot is almost never the true median
+of the section of the array to be partitioned. Thus, the \cc{i} and \cc{j}
+branches are almost always biased, which makes them predictable. According to
+Knuth, if the pivot element is chosen randomly, then the \cc{i} and \cc{j}
+branches will be taken (i.e.  the \cc{i} or \cc{j} loop will execute) twice as
+often as they are not taken (i.e. the loop exits).
 
-In order to investigate this, we implemented a simple version of
-quicksort that uses the middle element as the pivot rather than a
-median-of-three\footnote{It is important to note that selecting
-the pivot based on only a single element can be particularly inefficient
-because it removes the natural sentinels at the top and bottom
-of the regions to be partitioned that we get from median-of-three. In
-Figure \ref{base quicksort without a median, with median of 3, and with pseudo medians of 5, 7 and 9} we see that this results in an additional branch to handle
-the edgecase without a sentinel. Although this branch is almost 100\%
-predictable, it is executed a very large number of times.}. The arrays we sort are random, so the value of this
-element will also be random. Figure \ref{base quicksort without a median, with median of 3, and with pseudo medians of 5, 7 and 9}(a)
-shows the behaviour of each of the important branches in this code. We
-see that the $i$ and $j$ branches are indeed taken about twice as
-often as not-taken, and the $i$ and $j$ branches are correctly
-predicted around 71\% of the time.
+In order to investigate this, we implemented a simple version of quicksort that
+uses the middle element as the pivot rather than a median-of-three\footnote{It
+is important to note that selecting the pivot based on only a single element can
+be particularly inefficient because it removes the natural sentinels at the top
+and bottom of the regions to be partitioned that we get from median-of-three. In
+Figure \ref{quick-predictors2} we see that this results in an additional branch
+to handle the edgecase without a sentinel. Although this branch is almost 100\%
+predictable, it is executed a very large number of times.}. The arrays we sort
+are random, so the value of this element will also be random. Figure \ref{base
+quicksort without a median, with median of 3, and with pseudo-medians of 5, 7
+and 9-1} shows the behaviour of each of the important branches in this code. We
+see that the \cc{i} and \cc{j} branches are indeed taken about twice as often as
+not-taken, and the \cc{i} and \cc{j} branches are correctly predicted around
+71\% of the time.
 
-The result that the branches are more predictable (71\%) than they are
-biased (66\%) is a particularly interesting finding. The reason is that
-the bias in the results is an average figure over a large number of
-partitions that quicksort performs to sort the array. On a particular
-partition, the bias will be more pronounced because the $i$ and $j$
-branches will be biased in opposite directions.
+The result that the branches are more predictable (71\%) than they are biased
+(66\%) is a particularly interesting finding. The reason is that the bias in the
+results is an average figure over a large number of partitions that quicksort
+performs to sort the array. On a particular partition, the bias will be more
+pronounced because the \cc{i} and \cc{j} branches will be biased in opposite
+directions.
 
-For example, consider the case where the pivot element is larger than
-the true median. In this case, the final resting place of the pivot
-will be towards the right-hand end of the region to be
-partitioned. Thus, the $i$ loop will execute a large number of times,
-and thus the $i$ loop will be biased towards being taken. The $j$
-loop, on the other hand, will be biased toward not-taken, because the
-average element is likely to be smaller than the pivot. Therefore, the
-$j$ loop will execute only a small number of times. Thus, we see that
-in a particular partition, we do not normally get an equal number of
-executions of the $i$ and $j$ loops. Instead we get a large number of
-executions in one loop (which makes its branch relatively
-predictable), whereas the other loop often does not enter at all
-(which makes its branch also predictable, although executed a much
-smaller number of times). Thus, although overall the $i$ and $j$
-branches are taken twice as often as not-taken, on a particular
-partition one will usually be biased in the not-taken direction, and
-as a result will be executed a relativelty small number of times.
+For example, consider the case where the pivot element is larger than the true
+median. In this case, the final resting place of the pivot will be towards the
+right-hand end of the region to be partitioned. Thus, the \cc{i} loop will
+execute a large number of times, and thus the \cc{i} loop will be biased towards
+being taken. The \cc{j} loop, on the other hand, will be biased toward
+not-taken, because the average element is likely to be smaller than the pivot.
+Therefore, the \cc{j} loop will execute only a small number of times. Thus, we
+see that in a particular partition, we do not normally get an equal number of
+executions of the \cc{i} and \cc{j} loops. Instead we get a large number of
+executions in one loop (which makes its branch relatively predictable), whereas
+the other loop often does not enter at all (which makes its branch also
+predictable, although executed a much smaller number of times). Thus, although
+overall the \cc{i} and \cc{j} branches are taken twice as often as not-taken, on
+a particular partition one will usually be biased in the not-taken direction,
+and as a result will be executed a relativelty small number of times.
 
-In the preceding paragraphs we have seen that the $i$ and $j$ branches
-are predictable when we have a pivot that is larger or smaller than
-the true median of the array section to be partitioned. An interesting
-question is the effect of choosing a better approximation of the true
-median on branch prediction. Figure \ref{base quicksort without a
-median, with median of 3, and with pseudo medians of 5, 7 and 9}(b)
-shows the behavior of these branches when median-of-three is used to
-select the pivot. Perhaps the biggest effect is that the number of
-executions of the $i$ and $j$ branches falls by around 15\%. This fall
-is partly offset by the comparison branches to compute the median, but
-these account for only around 1.25\% of the original branches from the
-no-median quicksort. Thus, there is a reduction of almost 14\% in the
-number of executed comparison branches. Even more interesting, however,
-is the bias in the remaining $i$ and $j$ branches. The no-median $i$ and
-$j$ branches are 66\% biased, but the median-of-three branches are
-only 60\% biased in the taken direction. The accuracy with which
-these branches are correctly predicted falls from 71\% to 64\%.
+In the preceding paragraphs we have seen that the \cc{i} and \cc{j} branches are
+predictable when we have a pivot that is larger or smaller than the true median
+of the array section to be partitioned. An interesting question is the effect of
+choosing a better approximation of the true median on branch prediction. Figure
+\ref{base quicksort without a median, with median of 3, and with pseudo-medians
+of 5, 7 and 9-2} shows the behavior of these branches when median-of-three is
+used to select the pivot. Perhaps the biggest effect is that the number of
+executions of the \cc{i} and \cc{j} branches falls by around 15\%. This fall is
+partly offset by the comparison branches to compute the median, but these
+account for only around 1.25\% of the original branches from the no-median
+quicksort. Thus, there is a reduction of almost 14\% in the number of executed
+comparison branches. Even more interesting, however, is the bias in the
+remaining \cc{i} and \cc{j} branches. The no-median \cc{i} and \cc{j} branches
+are 66\% biased, but the median-of-three branches are only 60\% biased in the
+taken direction. The accuracy with which these branches are correctly predicted
+falls from 71\% to 64\%.
 
 The reason for the reduction in bias and prediction accuracy is that
-median-of-three gives a much better estimate of the true median of the
-section to be partitioned. The final resting place of the pivot will
-usually be closer to the middle of the section, to there is less bias
-in the branches. Thus, although median-of-three gives around a 14\%
-reduction in executed comparison branches, each of those branches is
-much less likely to be correctly predicted. Overall there is actually
-a 6\% \textit{increase} in the total number of branch mispredictions.
+median-of-three gives a much better estimate of the true median of the section
+to be partitioned. The final resting place of the pivot will usually be closer
+to the middle of the section, to there is less bias in the branches. Thus,
+although median-of-three gives around a 14\% reduction in executed comparison
+branches, each of those branches is much less likely to be correctly predicted.
+Overall there is actually a 6\% \n{increase} in the total number of branch
+mispredictions.
 
-To further investigate this phenomenon, we measured the effects of
-using pseudo-medians-of-five, -seven and -nine to choose the pivot. A
-pseudo-median uses repeated medians-of-three for parts of the
-computation to reduce the number of comparisons in computing the
-pivot. The reductions in executed comparison branches are 16.5\%,
-17.8\% and 19.6\% respectively. However, the biases of the $i$ and $j$
-branches in these algorithms are only 57.8\%, 56.3\% and 55.1\%. The
-corresponding predictabilities are 61.6\%, 59.9\% and 58.3\%. Overall,
-the increases in comparison branch mispredictions (including computing
-the pseudo-median) are 9.2\%, 11.8\% and 14.3\%.
+To further investigate this phenomenon, we measured the effects of using
+pseudo-medians-of-five, -seven and -nine to choose the pivot. A pseudo-median
+uses repeated medians-of-three for parts of the computation to reduce the number
+of comparisons in computing the pivot. The reductions in executed comparison
+branches are 16.5\%, 17.8\% and 19.6\% respectively. However, the biases of the
+$i$ and $j$ branches in these algorithms are only 57.8\%, 56.3\% and 55.1\%. The
+corresponding predictabilities are 61.6\%, 59.9\% and 58.3\%. Overall, the
+increases in comparison branch mispredictions (including computing the
+pseudo-median) are 9.2\%, 11.8\% and 14.3\%.
 
-Picking a better estimate of the true median substantially reduces the
-number of executed branches, but each of those branches is much more
-unpredictable. Intuitively, this makes sense. Sorting the array
-involves putting an order on it, and thus reducing its entropy. If we
-want to sort it in a smaller number of steps, then each step must
-remove more randomness. Thus, we would expect that the outcome of the
-branch that removes the randomness would itself be more random.
+Picking a better estimate of the true median substantially reduces the number of
+executed branches, but each of those branches is much more unpredictable.
+Intuitively, this makes sense. Sorting the array involves putting an order on
+it, and thus reducing its entropy. If we want to sort it in a smaller number of
+steps, then each step must remove more randomness. Thus, we would expect that
+the outcome of the branch that removes the randomness would itself be more
+random.
 
-Figure \ref{quick_median_cycles} shows the cycle count of each of the quicksorts,
-showing that a pseduo-median-of-seven offers the greatest tradeoff
-between the reduced number of branches and instructions, and the
-increase in branch mispredictions.
+Figure \ref{quick_median_cycles} shows the cycle count of each of the
+quicksorts, showing that a pseduo-median-of-seven offers the greatest tradeoff
+between the reduced number of branches and instructions, and the increase in
+branch mispredictions.
 
-An additional advantage of using a higher order median is that it
-dramatically reduces the probability of quicksort exhibiting worst
-case ($O(N^2)$) time complexity. Although the average case reduction
-in running time is small, reducing the chance of worst-case
-performance, which might cause the program to pause for minutes or
-hours, is very valuable.
+An additional advantage of using a higher order median is that it dramatically
+reduces the probability of quicksort exhibiting worst case ($O(N^2)$) time
+complexity. Although the average case reduction in running time is small,
+reducing the chance of worst-case performance, which might cause the program to
+pause for minutes or hours, is very valuable.
 

### 2005-08-14 docs/tech_report/shell.tex
fixed all references

@@
 %TODO
-\section{Predictor Results}
+\section{A More Detailed Look at Branch Prediction}
 
@@
 than half of the branches are correctly predicted, at 50.3\%, rising to 53\%
-using a strategy of predicting all branches as taken.
+using a strategy of predicting all branches as taken. These results can be seen
+in Figures \ref{Branch prediction performance for Shellsorts}.
 
@@
 
-The expected number of branch mispredictions is $log_{(\frac{11}{5})}N$. In
-actual fact, it is almost exactly correct, being slightly lower (18.48
-mispredictions per key, instead of 19.34, for 4194304 keys).
+The expected number of branch mispredictions is $log_{(\frac{11}{5})}N$.  This
+is almost exactly correct, being slightly lower (18.48 mispredictions per key,
+instead of 19.34, for 4194304 keys).
 

### 2005-08-14 docs/tech_report/background.tex
fixed small errors. didnt really get rid of any TODOs
fixed up the references
tidied up some more stuff

@@
 increments the count, to a maximum of 11, and each not-taken prediction
-decrements the counter, to a minimum of zero.
+decrements the counter, to a minimum of 00.
 
@@
 expensive than quicksort, due to the cost of each step. These are discussed in
-later chapters. The shellsort we use, called \n{Gonnet Shellsort} has a
-complexity of TODO (see \ref{TODO}).
+later chapters. The complexity of shellsort, meanwhile, is quite complex, and is
+discussed in \cite{Sedgewick96}.
 

### 2005-08-14 docs/tech_report/quick.tex
tidied up some more stuff

@@
 
-The results show again that direct mapped caches can perform better than fully
-associative ones. This is true in the case of all the quicksorts, as can be seen
-in Figures \ref{base quicksort (no median) level 2 misses} to
-\ref{multi-quicksort (sequential search) level 2 misses} graph that this is true
-in the case of all the quicksorts. As in Section \ref{mergesort replacement
-policy}, this is due to the replacement policy of the fully-associative cache.
+The results show again (see Section \ref{mergesort replacement policy}) that
+direct mapped caches can perform better than fully associative ones. This is
+true in the case of all the quicksorts, as can be seen in Figures \ref{base
+quicksort (no median) level 2 misses} to \ref{multi-quicksort (sequential
+search) level 2 misses} graph that this is true in the case of all the
+quicksorts. As in Section \ref{mergesort replacement policy}, this is due to the
+replacement policy of the fully-associative cache.
 

### 2005-08-14 docs/tech_report/background.tex
fixed use of =comparison-based (no hyphen) and fixed up nearly all \n vs \cc usage

@@
 used for ADTs is in Figure \vref{Definitions used by the ADT}. The tests
-presented in this report use \cc{unsigned integer}s as Items, which were
+presented in this report use \n{unsigned integer}s as Items, which were
 compared using a simple \n{less-than} function\footnote{This abstraction broke
@@
 
-Sorts which are not comparison based also exist. These sorts are called counting
-based sorts, as they count the number of keys to be put in each segment of an
-array before putting them in place. The most important of these is radixsort,
-discussed in Chapter \ref{radix}.
+Sorts which are not comparison-based also exist. These sorts are called
+counting-based sorts, as they count the number of keys to be put in each segment
+of an array before putting them in place. The most important of these is
+radixsort, discussed in Chapter \ref{radix}.
 

### 2005-08-14 docs/tech_report/conclusions.tex
fixed all references
fixed use of =comparison-based (no hyphen) and fixed up nearly all \n vs \cc usage

@@
 LaMarca determines that radixsort could not be improved in the same was as the
-comparison based sorts, and that due to poor cache performance, it ran more
+comparison-based sorts, and that due to poor cache performance, it ran more
 slowly than quicksort. We showed, however, that one of the improvements applied
@@
 \vspace*{-10em}
-\setcounter{figure}{8}
+\setcounter{figure}{9} % note the cheating to make this work
 \begin{figure}[H]
@@
 \end{changemargin}
-\caption[Cycles per key of several major sorts and their variations]{\label{All cycles}Cycles per key of several major sorts and their variations - this was
+\caption[Cycles per key of several major sorts and their variations]{\label{All
+cycles}Cycles per key of several major sorts and their variations - this was
 measured on a Pentium 4 using hardware performance counters.}

### 2005-08-14 docs/tech_report/heap.tex
fixed use of =comparison-based (no hyphen) and fixed up nearly all \n vs \cc usage

@@
 \section{Results}
-All simulations presented here were performed on 32 byte cache lines. As a
-result, 8-heaps were used for heapsort, since 8 4-byte \n{int}s fit the cache line.
-LaMarca used 64 bit ints, and while this doesn't affect results, it means that
-fewer keys fit into his cache lines. Since modern cache lines are typically
-longer than 4 ints, we stick with 32 byte cache lines. \verb!UINT_MAX! was used
+All simulations presented here were performed on 32-byte cache lines. As a
+result, 8-heaps were used for heapsort, since 8 4-byte \cc{int}s fit the cache
+line.  LaMarca used 64 bit ints, and while this doesn't affect results, it means
+that fewer keys fit into his cache lines. Since modern cache lines are typically
+longer than 4 ints, we stick with 32 byte cache lines. \cc{UINT\_MAX} was used
 as the sentinel.
@@
 Figure \ref{base and memory-tuned heapsorts-1}(a) shows the behaviour of each
-branch in the base heapsort algorithm. The `\cc{Creation, has children}' branch
+branch in the base heapsort algorithm. The `\n{Creation, has children}' branch
 refers to the check as to whether a key has children, during the creation of the
-heap The `\cc{Creation, child comparison}' branch is the comparison between the
-two children in the 2-heap, and the `\cc{Creation, child promotion}' refers to
+heap The `\n{Creation, child comparison}' branch is the comparison between the
+two children in the 2-heap, and the `\n{Creation, child promotion}' refers to
 checking if the child is larger than it's parent. The same checks are later made
-during the destruction of the heap, and are referred to as `\cc{Destruction, has
-children}', `\cc{Destruction, child comparison}' and `\cc{Desruction child
+during the destruction of the heap, and are referred to as `\n{Destruction, has
+children}', `\n{Destruction, child comparison}' and `\n{Desruction child
 promotion}'.
@@
 In the memory-tuned heapsort, the creation phase is simpler, involving merely
-one branch, called `\cc{fix-up}' here. The other columns refer to the
+one branch, called `\n{fix-up}' here. The other columns refer to the
 destruction phase, where most of the work takes place in memory-tuned heapsoirt.
@@
 comparisons during the destruction of the 2-, 4- and 8-heaps. Compare the
-columns marked `\cc{Destruction, child comparison}' and `\cc{Destruction, child
+columns marked `\n{Destruction, child comparison}' and `\n{Destruction, child
 promotion}' in Figure \ref{base and memory-tuned heapsorts-1} (the 2-heap), with
-the columns marked `\cc{comparison 1}' to `\cc{child promotion}' in Figure
+the columns marked `\n{comparison 1}' to `\n{child promotion}' in Figure
 \ref{base and memory-tuned heapsorts-2} (the 4-heap) and Figures \ref{base and

### 2005-08-14 docs/tech_report/method.tex
fixed use of =comparison-based (no hyphen) and fixed up nearly all \n vs \cc usage

@@
 Once the algorithm was written, a testing framework was used. This tested for
-edge-cases in the algorithms by testing sorts sized from $8$ to $255$ keys, with
-different random data sets. Testing was also performed on a very large data set.
-This test was timed and the results were considered. The results of this test
-were not accurate: the program took up to a minute to run, during which time the
-computer was being used by multiple users and processes, leading to context
-switches, sharing of resources and scheduling. Despite these inaccuracies, it was
-possible to use the results relative to each other, and to determine whether,
-for example, memory-tuned quicksort was running faster than base quicksort.
+edge-cases in the algorithms by testing sorts sized from \cc{8} to \cc{255}
+keys, with different random data sets. Testing was also performed on a very
+large data set.  This test was timed and the results were considered. The
+results of this test were not accurate: the program took up to a minute to run,
+during which time the computer was being used by multiple users and processes,
+leading to context switches, sharing of resources and scheduling. Despite these
+inaccuracies, it was possible to use the results relative to each other, and to
+determine whether, for example, memory-tuned quicksort was running faster than
+base quicksort.
 
@@
 
-This shows two stages of a bitonic mergesort which is described in Section \ref{base
-mergesort}. The top of the image shows the source array being merged into the
-target array below. The source array shows three large segments of the array and
-four smaller segments. The first two large segments have been merged into one
-segment in the target array. The third of these has been merged into the
-corresponding target array segment. The first of the smaller segments has begun
-being merged with the other small segments. The rest of the target array is
-uninitialised.  Below that is a small amount of debugging information. $i == j$
-is a key point in the program and is printed as debugging information. In the
+This shows two stages of a bitonic mergesort which is described in Section
+\ref{base mergesort}. The top of the image shows the source array being merged
+into the target array below. The source array shows three large segments of the
+array and four smaller segments. The first two large segments have been merged
+into one segment in the target array. The third of these has been merged into
+the corresponding target array segment. The first of the smaller segments has
+begun being merged with the other small segments. The rest of the target array
+is uninitialised.  Below that is a small amount of debugging information. $i ==
+j$ is a key point in the program and is printed as debugging information. In the
 two arrays below this, the merge continues.
@@
 
-\subsection{SimpleScalar} \cc{SimpleScalar} is an architecture simulator. It
+\subsection{SimpleScalar} \n{SimpleScalar} is an architecture simulator. It
 works by compiling code into its native instruction set, PISA, then emulating a
@@
 \label{Gnuplot}
-\cc{Gnuplot} is a tool for plotting data files. It provides visualisations of
+\n{Gnuplot} is a tool for plotting data files. It provides visualisations of
 data in two and three dimensions, and is used to create all of the graphs in
@@
 Imaginatively called \cc{main.c}, this program provided the means for testing
-and (roughly) timing each sort. Figure \vref{Key sections of Main.c} contains a
+and (roughly) timing each sort. Figure \vref{Key sections of main.c} contains a
 sample of this code.
 
-\code{Key sections of Main.c}{main.c}
+\code{Key sections of main.c}{main.c}
 
@@
 \label{Valgrind}
-\cc{Valgrind} is a set of tools which are used for debugging and profiling.  By
+\n{Valgrind} is a set of tools which are used for debugging and profiling.  By
 default it works as a memory checker, and detects errors such as the use of
@@
 random data from \n{http://www.random.org}. The software predictor used ten
-sections containing 4194304 \cc{unsigned integer}s, and the SimpleScalar
+sections containing 4194304 \n{unsigned integer}s, and the SimpleScalar
 simulations used the same ten sections. When SimpleScalar simulations were run
@@
 \subsection{32-bit Integers}
-LaMarca used 64-bit \cc{integer}s, as his test machine was a 64-bit system.
+LaMarca used 64-bit integers, as his test machine was a 64-bit system.
 Since our machine is 32-bit, and SimpleScalar uses 32-bit addressing on our

### 2005-08-14 docs/tech_report/quick.tex
fixed use of =comparison-based (no hyphen) and fixed up nearly all \n vs \cc usage

@@
 considerations for a fast C implementation. \cite{SedgewickBentley02} discusses
-equal keys and proves that quicksort has the lowest complexity of any comparison
-based sort.
+equal keys and proves that quicksort has the lowest complexity of any
+comparison-based sort.
 
@@
 the end. In this implementation, the sentinel is placed at the front, and the
-smallest of the first \cc{threshold} keys is used as a sentinel, in keeping
+smallest of the first \cc{THRESHOLD} keys is used as a sentinel, in keeping
 with our policy of not exceeding the bounds of the array.
@@
 SimpleScalar\footnote{This particular \cc{malloc} forms part of
-\cc{glibc}, the GNU C library, version 1.09.}.
+\n{glibc}, the GNU C library, version 1.09.}.
 

### 2005-08-14 docs/tech_report/radix.tex
fixed use of =comparison-based (no hyphen) and fixed up nearly all \n vs \cc usage

@@
 
-All other sorts considered in this project are \n{comparison based} sorts.
+All other sorts considered in this project are \n{comparison-based} sorts.
 They compare two keys at a time, and determine the larger and smaller key,
-sorting on that basis. Radixsort is a \n{counting based} sort. It works by
+sorting on that basis. Radixsort is a \n{counting-based} sort. It works by
 separating the keys into small parts, and processing these parts without
@@
 \subsection{Expected Performance}
-Radixsort is expected to have very good performance, compared to comparison
-based sorts. This is due to that fact that radixsort is $O(N)$; for an array as
-small as 4096 records, it involves 12 times fewer steps than the $O(NlogN)$
-sorts. For an array as large as 4194304, an $(O(NlogN)$ sort will take 22 times
-as many steps to complete. However, it is not expected that radixsort will
-perform many times better than quicksort, the reigning champion of $O(NlogN)$
-sorts. A `step' for quicksort is as simple as a comparison each step and a
-possible exchange every few steps.  For radixsort, a step comprises being
-counted four times and copied four times, with the counting involving several
-layers of indirection, and causing a complete iteration across the array at the
-start. LaMarca's results show the instruction count of quicksort and radixsort
-being in the same order of magnitude, and we expect the same.
+Radixsort is expected to have very good performance, compared to
+comparison-based sorts. This is due to that fact that radixsort is $O(N)$; for
+an array as small as 4096 records, it involves 12 times fewer steps than the
+$O(NlogN)$ sorts. For an array as large as 4194304, an $(O(NlogN)$ sort will
+take 22 times as many steps to complete. However, it is not expected that
+radixsort will perform many times better than quicksort, the reigning champion
+of $O(NlogN)$ sorts. A `step' for quicksort is as simple as a comparison each
+step and a possible exchange every few steps.  For radixsort, a step comprises
+being counted four times and copied four times, with the counting involving
+several layers of indirection, and causing a complete iteration across the array
+at the start. LaMarca's results show the instruction count of quicksort and
+radixsort being in the same order of magnitude, and we expect the same.
 
@@
 The number of branch prediction misses should be exceptionally low. In
-comparison based sorts, a branch due to a comparison can not be predicted except
+comparison-based sorts, a branch due to a comparison can not be predicted except
 in exceptional circumstances\footnote{See Section \ref{insertion is

### 2005-08-14 docs/tech_report/shell.tex
finished \cc vs \n search

@@
 in 1959, and described in \cite{Shell59}\footnote{Shell describes the former
-sorts as ``Merging Pairs'' and ``floating decimal sort'', respectively, and the
-latter sorts as ``finding the smallest'', ``interchanging pairs'' and ``sifting'',
-respectively.}.
+sorts as \n{Merging Pairs} and \n{Floating Decimal Sort}, respectively, and the
+latter sorts as \n{Finding the Smallest}, \n{Interchanging Pairs} and
+\n{Sifting}, respectively.}.
 
@@
 framework used.  The second was modified to reduce the complexity of the loop
-bounds, and to remove a sentinel check in the inner loop. Code for
-\cc{shellsort} and \cc{improved\_shellsort} is shown in Figures \ref{Shellsort
-code} and \ref{Improved Shellsort code}.
+bounds, and to remove a sentinel check in the inner loop. Code for shellsort and
+improved shellsort is shown in Figures \ref{Shellsort code} and \ref{Improved
+Shellsort code}.
 

### 2005-08-14 docs/tech_report/elementary.tex
fixed use of =comparison-based (no hyphen) and fixed up nearly all \n vs \cc usage
fixed usage of '32 byte' without a hyphen

@@
-\label{ordernsquared}
-Elementary sorts are also referred to as $O(N^2)$ sorts. This is due to the
-manner in which they check every key in the array in order to sort a single key;
-for every key to be sorted, every other key is checked.  Four of these sorts are
-discussed here: insertion sort, selection sort, bubblesort and shakersort,
-along with improved versions of bubblesort and shakersort.
+\label{ordernsquared} Elementary sorts are also referred to as $O(N^2)$ sorts.
+This is due to the manner in which they check every key in the array in order to
+sort a single key; for every key to be sorted, every other key is checked.  Four
+of these sorts are discussed here: insertion sort, selection sort, bubblesort
+and shakersort, along with improved versions of bubblesort and shakersort.
 
@@
 \subfigure[Level 1 cache misses per key - this was simulated using SimpleScalar \cc{sim-cache}, simulating an 8KB data
-cache with a 32 byte cache line and separate instruction cache.]
+cache with a 32-byte cache line and separate instruction cache.]
 {\label{elementary level 1 misses}\includegraphics[scale=\myscale]{plots/on2_-_Level_1_misses.eps}}

### 2005-08-14 docs/tech_report/report.tex
fixed all references
fixed use of =comparison-based (no hyphen) and fixed up nearly all \n vs \cc usage
finished \cc vs \n search
fixed usage of '32 byte' without a hyphen

@@
 
+\newenvironment{spacedpars}{%for the abstract
+\setlength{\parskip}{1em}
+\setlength{\parindent}{0ex}
+}
+
 %cc is used for
-% - unix commans or programs: gcc, sim-bpred etc
+% - unix commans or programs: gcc, sim-bpred, sim-cache, espresso, li, simple,
+%   sim-fast, diff, valgrind, cachegrind, driver
+% - code or keywords: less, Item, exch, if, while, unsigned int, int, qsort,
+%   goto, do-while, THRESHOLD, partition, malloc, shellsort, improved shellsort
+% - file names: main.c
+% - #defines: UNIT_MAX
 \newcommand{\cc}[1]{\texttt{#1}}
@@
 %n is used for
-% - proper names: Atom, SimpleScalar, Papi
+% - proper names: Atom, SimpleScalar, Papi, GnuPlot, ValGrind, glibc
+% - technical names: unsigned integer, less-than, greater-than, in-place, out-of
+%   place, Abstract Data Type, records, cache hierarchy, latency, hit, miss,
+%   compulsory, cold-start, capacity, conflict, cache line, cache block, offset,
+%   tag, virtaul addresses, bus address, temporal locality, locality, spatial
+%   locality, taken, not taken, flow control branch, comparaitve branch, static,
+%   semi-static, dynamic, 1-bit, bimodal, two-level adaptive, heap, children,
+%   implicit heap, heapifying, sifting, fix-up, fix-down, sift-up, sift-down,
+%   Repeated-adds, Remove-min, d-heap, fanout, natural, straight, unrolled,
+%   Memory management unit, tiling, k-way merge, trashing divide and conquer,
+%   multi-way partition, comparision based, counting based, increments
+% - web addresses: http://www.gnuplot.com, http://valgrind.kde.org
+% - sort names, when highlighted: algorithm S, algorithm N
+% - branch predictor names: 'Creation, child promotion', etc
 \newcommand{\n}[1]{\textit{#1}}
@@
 \subfigure[Level 1 cache misses per key - this was simulated using SimpleScalar \cc{sim-cache}, simulating an 8KB data
-cache with a 32 byte cache line and separate instruction cache. #4]
+cache with a 32-byte cache line and separate instruction cache. #4]
 {\label{#1 level 1 misses}\includegraphics[scale=\myscale]{plots/#1_-_Level_1_misses.eps}}\\
 \subfigure[Level 2 cache misses per key - this was simulated using SimpleScalar \cc{sim-cache}, simulating a 2MB data
-and instruction cache and a 32 byte cache line. #5]
+and instruction cache and a 32-byte cache line. #5]
 {\label{#1 level 2 misses}\includegraphics[scale=\myscale]{plots/#1_-_Level_2_misses.eps}}\\
@@
 
+\begin{spacedpars}
 \begin{abstract}
@@
 \end{abstract}
+\end{spacedpars}
 

### 2005-08-14 docs/tech_report/conclusions.tex
renamed cases of 2[ \-]level adaptive to two-level

@@
 mergesort, and that multi-mergesort has regular branch predictive access
-patterns during the k-way merge, which can be taken advantage of by a 2-level
+patterns during the k-way merge, which can be taken advantage of by a two-level
 adaptive predictor.
@@
 predictions. While many of these were mispredicted using the bimodal predictor,
-the 2 level adaptive predictor was able to predict a higher
+the two-level adaptive predictor was able to predict a higher
 proportion correctly.

### 2005-08-14 docs/tech_report/method.tex
fixed usage of '32 byte' without a hyphen
renamed cases of 2[ \-]level adaptive to two-level

@@
 LaMarca used the \n{Atom} simulator, which was used to test software for the DEC
-AlphaStation. It simulated a 2MB direct-mapped cache with a 32 byte cache line.
+AlphaStation. It simulated a 2MB direct-mapped cache with a 32-byte cache line.
 Other settings were also chosen to correspond to the DEC Alpha: 8KB level 1 data
@@
 We chose a similar approach for our branch predictor caches. We aimed to
-simulate bimodal predictors as well as 2 level adaptive, in order to be able to
+simulate bimodal predictors as well as two-level adaptive, in order to be able to
 compare them together. We simulated bimodal predictors with a 512, 1024 and
-2048 byte table, and 2 level adaptive predictors with 512, 1024, 2048 and 4096
+2048 byte table, and two-level adaptive predictors with 512, 1024, 2048 and 4096
 entries.

### 2005-08-14 docs/tech_report/quick.tex
renamed cases of 2[ \-]level adaptive to two-level

@@
 expected, there was a a dramatic increase in the binary search multi-quicksort
-branch misses, but only when using a bimodal predictor. When using a 2-level
+branch misses, but only when using a bimodal predictor. When using a two-level
 adaptive predictor, the number of misses is, in general, slightly higher, but it
 does not spike due to the binary search. While it performs worse than the
-bimodal predictor, it appears that 2-level adaptive predictors can adapt to the
-erratic branches of a binary search.
+bimodal predictor, it appears that two-level adaptive predictors can adapt to
+the erratic branches of a binary search.
 

### 2005-08-14 docs/tech_report/radix.tex
renamed cases of 2[ \-]level adaptive to two-level
fixed cache misses slope vs amortization thing

@@
 
-Again we see that the bimodal branch predictor performs better than the two
-level adaptive. In this case it is contrary to expectations, however, as
+Again we see that the bimodal branch predictor performs better than the
+two-level adaptive. In this case it is contrary to expectations, however, as
 there are no comparative branches. However, there are so few misses that these
@@
 flat. He shows a similar result in both cache miss graphs: a sharp dip, followed
-by a nearly flat-line. Also, in both cases, once the array no longer fits in the
-cache, the number of misses jumps to around 2.5, where they stay.
+by a nearly flat-line. This may be due to amortization, in which case the trend
+would not be expected in our graphs due to subtraction of the time taken to fill
+the array.
+
+Also, in both cases in LaMarca's graphs, once the array no longer fits in the
+cache, the number of misses jumps to around 3.25 (which would be equivalent to
+1.625 misses per key in our results), where they stay. Our results
+disagree, showing approximately 2.5 misses per key for base radixsort.
 
 However, LaMarca's conclusion - that radixsort is not as fast as quicksort - can
-be easily seen from the base radixsort results. Once an radixsort optimised for
-the cache is presented, these no longer hold, and the lack of branch
+be easily explained from the base radixsort results. Once an radixsort optimised
+for the cache is presented, these no longer hold, and the lack of branch
 misprediction results in a new fastest sort.

### 2005-08-14 docs/tech_report/elementary.tex
fixed captitalization in captions

@@
 \vspace*{6em}
-\caption{Simulated Instruction count and empiric cycle count for elementary sorts}
-\label{Simulated Instruction count and empiric cycle count for elementary sorts}
+\caption{Simulated instruction count and empiric cycle count for elementary sorts}
+\label{Simulated instruction count and empiric cycle count for elementary sorts}
 \end{figure}
@@
 \vspace*{6em}
-\caption{Cache and Branch prediction simulation results for elementary sorts}
-\label{Cache and Branch prediction simulation results for elementary sorts}
+\caption{Cache and branch prediction simulation results for elementary sorts}
+\label{Cache and branch prediction simulation results for elementary sorts}
 \end{figure}
@@
 \vspace*{6em}
-\caption{Branch Simulation results for elementary sorts}
-\label{Branch Simulation results for elementary sorts}
+\caption{Branch simulation results for elementary sorts}
+\label{Branch simulation results for elementary sorts}
 \end{figure}
@@
 Figure \ref{elementary branches} shows the number of branches per key of these
-sorts, while Figure \ref{Branch Simulation results for elementary sorts} shows
+sorts, while Figure \ref{Branch simulation results for elementary sorts} shows
 the branch misses per key. Due to huge variations in predictability, we split

### 2005-08-14 docs/tech_report/heap.tex
fixed usage of '32 byte' without a hyphen
fixed captitalization in captions

@@
 instruction count. Regardless, the base algorithm used here is Floyd's in-place
-sort. Code for a base heapsort is in Figure \ref{Simple Base Heapsort}.
+sort. Code for a base heapsort is in Figure \ref{Simple base heapsort}.
 
-\code{Simple Base Heapsort}{base_heapsort.c}
+\code{Simple base heapsort}{base_heapsort.c}
 
@@
 that fewer keys fit into his cache lines. Since modern cache lines are typically
-longer than 4 ints, we stick with 32 byte cache lines. \cc{UINT\_MAX} was used
+longer than 4 ints, we stick with 32-byte cache lines. \cc{UINT\_MAX} was used
 as the sentinel.

### 2005-08-14 docs/tech_report/merge.tex
fixed all references
fixed use of =comparison-based (no hyphen) and fixed up nearly all \n vs \cc usage
fixed captitalization in captions

@@
 
-The prediction results in Figure \ref{Branch Simulation results for mergesort}
+The prediction results in Figure \ref{Branch simulation results for mergesort}
 are as expected. Very little variation exists between the types of mergesort. It
@@
 
-\section{Predictor Results}
+\section{A More Detailed Look at Branch Prediction}
 
@@
 heapsort, and progresses in the same manner (see Figure \ref{memory-tuned
-heapsort curve}). Note that \n{comparison 0} occurs every time, and would be
+heapsort curve}). Note that `\n{comparison 0}' occurs every time, and would be
 optimized out by the compiler.

### 2005-08-14 docs/tech_report/elementary.tex
fixed the captions for the graphs, and a few of the graphs themselves

@@
 \begin{changemargin}
-\subfigure[Level 1 cache misses per key - this was simulated using SimpleScalar \cc{sim-cache}, simulating an 8KB data
-cache with a 32-byte cache line and separate instruction cache.]
+\subfigure[Level 1 cache misses per key - this was simulated using SimpleScalar
+\cc{sim-cache}, simulating an 8KB data cache with a 32-byte cache line and
+separate instruction cache. Results shown are for a direct mapped cache.]
 {\label{elementary level 1 misses}\includegraphics[scale=\myscale]{plots/on2_-_Level_1_misses.eps}}
@@
 \centering
-\subfigure[Branches misses per key - this was simulated using \cc{sim-bpred}, with bimodal and two-level adaptive
-predictors. The simulated two-level adaptive predictor used a 10 bit branch history register which was XOR-ed with the
-program counter.]
+\subfigure[Branches misses per key - this was simulated using \cc{sim-bpred},
+with bimodal and two-level adaptive predictors. The simulated two-level adaptive
+predictor used a 10 bit branch history register which was XOR-ed with the
+program counter. Results shown are for a 512 byte bimodal predictor.]
 {\includegraphics[scale=\myscale]{plots/on2_-_Branch_misses1.eps}}
-\subfigure[Branches misses per key - this was simulated using \cc{sim-bpred}, with bimodal and two-level adaptive
-predictors. The simulated two-level adaptive predictor used a 10 bit branch history register which was XOR-ed with the
-program counter.]
+\subfigure[Branches misses per key - this was simulated using \cc{sim-bpred},
+with bimodal and two-level adaptive predictors. The simulated two-level adaptive
+predictor used a 10 bit branch history register which was XOR-ed with the
+program counter. Results shown are for a 512 byte bimodal predictor.]
 {\includegraphics[scale=\myscale]{plots/on2_-_Branch_misses2.eps}}

### 2005-08-14 docs/tech_report/report.tex
fixed captitalization in captions
fixed the captions for the graphs, and a few of the graphs themselves

@@
 
+% widens the margins for the graphs
 \newenvironment{changemargin}{%
@@
 
-%cc is used for
+%\cc{x} is used for
 % - unix commans or programs: gcc, sim-bpred, sim-cache, espresso, li, simple,
@@
 
-%n is used for
+%\n{x} is used for
 % - proper names: Atom, SimpleScalar, Papi, GnuPlot, ValGrind, glibc
@@
 
+% to include a file as code
 \newcommand{\code}[2]{\begin{figure}\lstinputlisting{code/#2}\caption{#1}\label{#1}\end{figure}}
 
+% used for scaling graphs
 \newcommand{\myscale}{1.2}
+
+% plots graphs across 3 pages
 \newcommand{\plot}[7]{
@@
 \vspace*{7em}
-\caption{Simulated Instruction count and empiric cycle count for #1sort}
-\label{Simulated Instruction count and empiric cycle count for #1sort}
+\caption{Simulated instruction count and empiric cycle count for #1sort}
+\label{Simulated instruction count and empiric cycle count for #1sort}
 \end{figure}
@@
 \begin{changemargin}
-\subfigure[Level 1 cache misses per key - this was simulated using SimpleScalar \cc{sim-cache}, simulating an 8KB data
-cache with a 32-byte cache line and separate instruction cache. #4]
-{\label{#1 level 1 misses}\includegraphics[scale=\myscale]{plots/#1_-_Level_1_misses.eps}}\\
-\subfigure[Level 2 cache misses per key - this was simulated using SimpleScalar \cc{sim-cache}, simulating a 2MB data
-and instruction cache and a 32-byte cache line. #5]
+\subfigure[Level 1 cache misses per key - this was simulated using SimpleScalar
+\cc{sim-cache}, simulating an 8KB data cache with a 32-byte cache line and
+separate instruction cache. #4] {\label{#1 level 1
+misses}\includegraphics[scale=\myscale]{plots/#1_-_Level_1_misses.eps}}\\
+\subfigure[Level 2 cache misses per key - this was simulated using SimpleScalar
+\cc{sim-cache}, simulating a 2MB data and instruction cache and a 32-byte cache
+line. #5]
 {\label{#1 level 2 misses}\includegraphics[scale=\myscale]{plots/#1_-_Level_2_misses.eps}}\\
@@
 \vspace*{7em}
-\caption{Cache Simulation results for #1sort}
-\label{Cache Simulation results for #1sort}
+\caption{Cache simulation results for #1sort}
+\label{Cache simulation results for #1sort}
 \end{figure}
@@
 {\label{#1 branches}\includegraphics[scale=\myscale]{plots/#1_-_Branches.eps}}
-\subfigure[Branches misses per key - this was simulated using \cc{sim-bpred}, with bimodal and two-level adaptive
-predictors. The simulated two-level adaptive predictor used a 10 bit branch history register which was XOR-ed with the
+\subfigure[Branches misses per key - this was simulated using \cc{sim-bpred},
+with bimodal and two-level adaptive predictors. The simulated two-level adaptive
+predictor used a 10-bit branch history register which was XOR-ed with the
 program counter. #7]
@@
 \vspace*{6em}
-\caption{Branch Simulation results for #1sort}
-\label{Branch Simulation results for #1sort}
+\caption{Branch simulation results for #1sort}
+\label{Branch simulation results for #1sort}
 \end{figure}
@@
 
+% the scaling for the smaller graphs
 \newcommand{\smallscale}{.74}
+
+% plots 6 images on 1 page
 \newcommand{\smallplot}[2]{
@@
 
+% plots 6 or less images on a page, using the counters
 \newcommand{\sixplot}[7]{
@@
 
+% settings for the code listings
 \lstset{language=c,tabsize=3,basicstyle=\small,showspaces=false,showtabs=false,showstringspaces=false,stepnumber=1,numberstyle=\small,frame=tblr}
 
+% i actually have no idea what this does, and no memory of writing it
 \long\def\symbolfootnote[#1]#2{\begingroup\def\thefootnote{\fnsymbol{footnote}}\footnote[#1]{#2}\endgroup}
@@
 
+% how to hyphenate words which arent standard
 \hyphenation{La-Marca}
@@
 
-% TODO the bugs are now fixed (actually, one of them was only mistakenly a bg :)
 % the only remaining thing is the compulsory cache subtraction.

### 2005-08-14 docs/tech_report/heap.tex
fixed the captions for the graphs, and a few of the graphs themselves
added info about all the labels. Also capitalized them all

@@
 Figure \ref{base and memory-tuned heapsorts-1}(a) shows the behaviour of each
-branch in the base heapsort algorithm. The `\n{Creation, has children}' branch
+branch in the base heapsort algorithm. The `\n{Creation, Has Children}' branch
 refers to the check as to whether a key has children, during the creation of the
-heap The `\n{Creation, child comparison}' branch is the comparison between the
-two children in the 2-heap, and the `\n{Creation, child promotion}' refers to
+heap The `\n{Creation, Child Comparison}' branch is the comparison between the
+two children in the 2-heap, and the `\n{Creation, Child Promotion}' refers to
 checking if the child is larger than it's parent. The same checks are later made
-during the destruction of the heap, and are referred to as `\n{Destruction, has
-children}', `\n{Destruction, child comparison}' and `\n{Desruction child
-promotion}'.
+during the destruction of the heap, and are referred to as `\n{Destruction, Has
+Children}', `\n{Destruction, Child Comparison}' and `\n{Destruction Child
+Promotion}'.
 
 In the memory-tuned heapsort, the creation phase is simpler, involving merely
-one branch, called `\n{fix-up}' here. The other columns refer to the
+one branch, called `\n{Fix-up}' here. The other columns refer to the
 destruction phase, where most of the work takes place in memory-tuned heapsoirt.
@@
 comparisons during the destruction of the 2-, 4- and 8-heaps. Compare the
-columns marked `\n{Destruction, child comparison}' and `\n{Destruction, child
-promotion}' in Figure \ref{base and memory-tuned heapsorts-1} (the 2-heap), with
-the columns marked `\n{comparison 1}' to `\n{child promotion}' in Figure
+columns marked `\n{Destruction, Child Comparison}' and `\n{Destruction, Child
+Promotion}' in Figure \ref{base and memory-tuned heapsorts-1} (the 2-heap), with
+the columns marked `\n{Comparison 0}' to `\n{Child Promotion}' in Figure
 \ref{base and memory-tuned heapsorts-2} (the 4-heap) and Figures \ref{base and
@@
 \includegraphics{plots/curve.eps}
-\caption{Predictability of branches in memory-tuned heapsort, with a fanout of eight}
+\caption{Predictability of branches in memory-tuned heapsort, using an 8-heap.}
 \label{memory-tuned heapsort curve}

### 2005-08-14 docs/tech_report/merge.tex
fixed the captions for the graphs, and a few of the graphs themselves
added info about all the labels. Also capitalized them all

@@
 
-\plot{merge}{}{}{Results shown are for a Direct Mapped cache.}{
-Results shown are for a Direct Mapped cache.}{}{}
+\plot{merge}{}{}{Results shown are for a direct mapped cache.}{Results shown are
+for a direct mapped cache.}{}{}
 
@@
 Figure \ref{mergesorts-1} shows that algorithm N has the same number of branches
-and branch mispredictions as algorithm S, shown in \ref{mergesorts-2}. Base
-mergesort has the same number of mispredictions (see Figure \ref{mergesorts-3},
-indicating a possible minimum number. It has a lower number of branches,
-probably due to performing a presort first. The presort has approximately 6.8
-misses per key (not shown in the graphs). In later mergesorts, an insertion sort
-is used instead.
+and branch mispredictions as algorithm S, shown in \ref{mergesorts-2}. The
+labels in the graphs correspond to the names of the steps in Knuth's algorithm,
+which also correspond with the labels in our code.
 
-Double-aligned tiled mergesort (which has the same results are tiled mergesort) has a
-predilection for merging forwards. This is likely to be due to the fact that the
-final merge goes forwards. The left merge step is also favoured, but this is
-merely due to the fact that it occurs first in the code. While there are more
-left branches, the actual number of branches taken are the same.
+Base mergesort has the same number of mispredictions as Knuth's sorts (see
+Figure \ref{mergesorts-3}, indicating a possible minimum number. `\n{Left}' and
+`\n{Right}' refer to interating through the first (or left-hand) and second (or
+right-hand) lists being compared, respectively.
 
-In both double-aligned tiled mergesort and base mergesort, approximately two thirds are
-of the loops are taken, unlike Knuth's sorts. However, this appears to indicate
-an increase in branches, as opposed to a decrease in mispredictions. Overall
-this stay is line with the idea that there is a minimum number of branch misses,
-below which comparative sorts will not drop.
+Base mergesort has a lower number of branches than Knuth's sorts, probably due
+to performing a presort first. The presort has approximately 6.8 misses per key
+(not shown in the graphs). In later mergesorts, an insertion sort is used
+instead.
+
+Double-aligned tiled mergesort (which has the same results are tiled mergesort)
+has a predilection for merging forwards, shown in \ref{mergesorts-4}. The
+`\n{Left}' and `\n{Right}' in the labels refer to the same as in
+\ref{mergesorts-3}, while `\n{Forwards}' and `\n{Reverse}' refer to whether the
+first or second part of the bitonic array segment is being merged.
+
+This predilication is likely to be due to the fact that the final merge goes
+forwards. The left merge step is also favoured, but this is merely due to the
+fact that it occurs first in the code. While there are more left branches, the
+actual number of branches taken are the same.
+
+In both double-aligned tiled mergesort and base mergesort, approximately two
+thirds are of the loops are taken, unlike Knuth's sorts. However, this appears
+to indicate an increase in branches, as opposed to a decrease in mispredictions.
+Overall this stay is line with the idea that there is a minimum number of branch
+misses, below which comparative sorts will not drop.
 
 Double multi-mergesort sorts the arrays fully in the cache, leaving them
-bitonically arranged. This means that there is no bias for the forward branches,
-as there was in tiled mergesort. This is shown in Figures \ref{mergesorts-5} and
-\ref{mergesorts-6}. While merging the number of branches is 20\% lower than base
-mergesort, while the number of mispredictions in only 10\% lower. This implies
-that most of the mispredictions occur early in the merging process, and not in
-the final merge, which multi-mergesort leaves out.
+bitonically arranged. This means that there is no final merge, and hence no bias
+for the forward branches, as there was in tiled mergesort. This is shown in
+Figures \ref{mergesorts-5} and \ref{mergesorts-6}. The labels in these graphs
+are the same as in Figure \ref{mergesorts-4} and \ref{base and memory-tuned
+heapsorts-2}.
+
+While merging, the number of branches is 20\% lower than base mergesort, while
+the number of mispredictions in only 10\% lower. This implies that most of the
+mispredictions occur early in the merging process, and not in the final merge,
+which multi-mergesort leaves out.
 

### 2005-08-14 docs/tech_report/quick.tex
fixed the captions for the graphs, and a few of the graphs themselves
added info about all the labels. Also capitalized them all

@@
 sentinel. If the partition to be sorted is the left most partition, then the
-smallest key in the partition will be chosen as the sentinel. No other
-partition needs a sentinel. A possible way to avoid the sentinel in that
-partition is to place zero at $0^{th}$ position in the list to be sorted, and
-sort whichever number was in this position back at the end. This sort could be
-done with one iteration of insertion sort, and would have just one branch miss.
-However, the number of level 2 cache misses would be great in this case,
-and the cost of this would be substantial. This could be reduced by
-choosing the key to remove from a small list, such as the eight keys at the
-start, middle and end of the array. This would introduce a constant number of
-cache misses, but would significantly reduce the chances of having to sort
-across the entire array at the end. This, however, is not implemented; instead a
-check is done to fetch a sentinel from the appropriate sub-file if required.
-This check isn't very expensive, and is significantly cheaper than either a
-bounds check or putting a sentinel into every sub-file.
+smallest key in the partition will be chosen as the sentinel. No other partition
+needs a sentinel. A possible way to avoid the sentinel in that partition is to
+place zero at $0^{th}$ position in the list to be sorted, and sort whichever
+number was in this position back at the end. This sort could be done with one
+iteration of insertion sort, and would have just one branch miss.  However, the
+number of level 2 cache misses would be great in this case, and the cost of this
+would be substantial. This could be reduced by choosing the key to remove from a
+small list, such as the eight keys at the start, middle and end of the array.
+This would introduce a constant number of cache misses, but would significantly
+reduce the chances of having to sort across the entire array at the end. This,
+however, is not implemented; instead a check is done to fetch a sentinel from
+the appropriate sub-file if required.  This check isn't very expensive, and is
+significantly cheaper than either a bounds check or putting a sentinel into
+every sub-file.
 
@@
 
-\plot{quick}{}{}{Results shown are for a Direct Mapped cache.}{}{}{}{}
+\plot{quick}{}{}{Results shown are for a direct mapped cache.}{Results shown are
+for a direct mapped cache.}{}{}{}
 
@@
 Figure \ref{base, memory-tuned and multi- quicksorts-1} shows the behaviour of
-the indiviual branches in base quicksort. The \cc{i} branch is the loop control
+the indiviual branches in base quicksort. The \n{i} branch is the loop control
 branch for the index sweeping from left-to-right up the array during partition.
-The \cc{j} branch is the corresponding loop sweeping down the array. The
-partition step terminates when $i \geq j$, and the \cc{partition end} branch
-checks that condition. The \cc{insertion} branch is the comparison branch in the
+The \n{j} branch is the corresponding loop sweeping down the array. The
+partition step terminates when \n{i $\geq$ j}, and the \n{Partition End} branch
+checks that condition. The \n{Insertion} branch is the comparison branch in the
 insertion sort algorithm. Recall that our quicksort algorithm does not sort
@@
 insertion sort over the whole array to sort these small partitions. Finally, the
-\cc{median} branch is the sum of the results for the branches involved in
+\n{Median} branch is the sum of the results for the branches involved in
 computing the median-of-three for each partition.
 
-The only difference between cache quicksort and base quicksort is that
-cache quicksort applies insertion sort to small partitions
-immediately, whereas base quicksort does a single insertion sort pass
-at the end. The result is that both algorithms perform almost
-identically from the viewpoint of branch prediction. Cache quicksort
-uses a very slightly smaller number of insertion sort comparison
-branches, because it sorts only within partitions and does not attempt
-to move array elements that have acted as pivots. (The reason that
-quicksort works is that after partitioning, the pivot is in its
-correct location, and must never move again). In practice, this
-reduction in branches is probably offset by the increase in
-mispredictions in the outer loop exit branch of the many insertion
-sorts.
+The only difference between memory-tuned quicksort and base quicksort is that
+memory-tuned quicksort applies insertion sort to small partitions immediately,
+whereas base quicksort does a single insertion sort pass at the end. The result
+is that both algorithms perform almost identically from the viewpoint of branch
+prediction. Memory-tuned quicksort uses a very slightly smaller number of
+insertion sort comparison branches, because it sorts only within partitions and
+does not attempt to move array elements that have acted as pivots. (The reason
+that quicksort works is that after partitioning, the pivot is in its correct
+location, and must never move again). In practice, this reduction in branches is
+probably offset by the increase in mispredictions in the outer loop exit branch
+of the many insertion sorts.
 
-Base quicksort has a 35\% branch misprediction rate, whereas the rate
-for cache quicksort is only 34\%. The main reason for this is the
-reduction in branches by not attempting to move the pivot elements.
-Attempting to move these elements usually results in a branch
-misprediction. The main reason is that most other elements move at
-least one position, and thus cause the insertion sort loop to iterate
-at least once. But the pivot elements never move --- they are already
-in the correct place.
+Base quicksort has a 35\% branch misprediction rate, whereas the rate for
+memory-tuned quicksort is only 34\%. The main reason for this is the reduction
+in branches by not attempting to move the pivot elements.  Attempting to move
+these elements usually results in a branch misprediction. The main reason is
+that most other elements move at least one position, and thus cause the
+insertion sort loop to iterate at least once. But the pivot elements never move
+-- they are already in the correct place.
 
 When we look at multi-quicksort, there is a 19\% drop in the number of
-iterations of the \cc{i} and \cc{j} loops, with a corresponding reduction in the
+iterations of the \n{i} and \n{j} loops, with a corresponding reduction in the
 number of mispredictions. Interestingly, the percentage misprediction rate of
-the \cc{i} and \cc{j} branches does not change --- it is 37\% in both cases.
+the \n{i} and \n{j} branches does not change -- it is 37\% in both cases.
 
@@
 around 25\% and 50\% respectively. Figure \ref{base, memory-tuned and multi-
-quicksorts-3} shows additional measures of the \cc{binary left} and \cc{binary
-right} branches, which are the branches within the binary search code. The
-number of binary search branches is much greater than the reduction in \cc{i}
-and \cc{j} branches. Furthermore, these binary search branches are highly (43\%)
+quicksorts-3} shows additional measures of the \n{Binary Left} and \n{Binary
+Right} branches, which are the branches within the binary search code. The
+number of binary search branches is much greater than the reduction in \n{i}
+and \n{j} branches. Furthermore, these binary search branches are highly (43\%)
 unpredictable.
@@
 increases the number of executed branches, but these branches are highly
-predictable. This version actually causes fewer branch mispredictions
-than cache quicksort.
+predictable. This version actually causes fewer branch mispredictions than
+memory-tuned quicksort.
 
@@
 \label{quicksort entropy}
-When we originally looked at quicksort, we expected that the \cc{i} and \cc{j}
+When we originally looked at quicksort, we expected that the \n{i} and \n{j}
 branches in quicksort would resolve in each direction. In fact, this is not the
 case. The main reason is that the chosen pivot is almost never the true median
-of the section of the array to be partitioned. Thus, the \cc{i} and \cc{j}
+of the section of the array to be partitioned. Thus, the \n{i} and \n{j}
 branches are almost always biased, which makes them predictable. According to
-Knuth, if the pivot element is chosen randomly, then the \cc{i} and \cc{j}
-branches will be taken (i.e.  the \cc{i} or \cc{j} loop will execute) twice as
+Knuth, if the pivot element is chosen randomly, then the \n{i} and \n{j}
+branches will be taken (i.e.  the \n{i} or \n{j} loop will execute) twice as
 often as they are not taken (i.e. the loop exits).
@@
 and 9-1} shows the behaviour of each of the important branches in this code. We
-see that the \cc{i} and \cc{j} branches are indeed taken about twice as often as
-not-taken, and the \cc{i} and \cc{j} branches are correctly predicted around
-71\% of the time.
+see that the \n{i} and \n{j} branches are indeed taken about twice as often as
+not-taken, and the \n{i} and \n{j} branches are correctly predicted around 71\%
+of the time.
 
@@
 performs to sort the array. On a particular partition, the bias will be more
-pronounced because the \cc{i} and \cc{j} branches will be biased in opposite
+pronounced because the \n{i} and \n{j} branches will be biased in opposite
 directions.
@@
 median. In this case, the final resting place of the pivot will be towards the
-right-hand end of the region to be partitioned. Thus, the \cc{i} loop will
-execute a large number of times, and thus the \cc{i} loop will be biased towards
-being taken. The \cc{j} loop, on the other hand, will be biased toward
-not-taken, because the average element is likely to be smaller than the pivot.
-Therefore, the \cc{j} loop will execute only a small number of times. Thus, we
-see that in a particular partition, we do not normally get an equal number of
-executions of the \cc{i} and \cc{j} loops. Instead we get a large number of
-executions in one loop (which makes its branch relatively predictable), whereas
-the other loop often does not enter at all (which makes its branch also
-predictable, although executed a much smaller number of times). Thus, although
-overall the \cc{i} and \cc{j} branches are taken twice as often as not-taken, on
-a particular partition one will usually be biased in the not-taken direction,
-and as a result will be executed a relativelty small number of times.
+right-hand end of the region to be partitioned. Thus, the \n{i} loop will
+execute a large number of times, and thus the \n{i} loop will be biased towards
+being taken. The \n{j} loop, on the other hand, will be biased toward not-taken,
+because the average element is likely to be smaller than the pivot.  Therefore,
+the \n{j} loop will execute only a small number of times. Thus, we see that in a
+particular partition, we do not normally get an equal number of executions of
+the \n{i} and \n{j} loops. Instead we get a large number of executions in one
+loop (which makes its branch relatively predictable), whereas the other loop
+often does not enter at all (which makes its branch also predictable, although
+executed a much smaller number of times). Thus, although overall the \n{i} and
+\n{j} branches are taken twice as often as not-taken, on a particular partition
+one will usually be biased in the not-taken direction, and as a result will be
+executed a relativelty small number of times.
 
-In the preceding paragraphs we have seen that the \cc{i} and \cc{j} branches are
+In the preceding paragraphs we have seen that the \n{i} and \n{j} branches are
 predictable when we have a pivot that is larger or smaller than the true median
@@
 used to select the pivot. Perhaps the biggest effect is that the number of
-executions of the \cc{i} and \cc{j} branches falls by around 15\%. This fall is
+executions of the \n{i} and \n{j} branches falls by around 15\%. This fall is
 partly offset by the comparison branches to compute the median, but these
@@
 comparison branches. Even more interesting, however, is the bias in the
-remaining \cc{i} and \cc{j} branches. The no-median \cc{i} and \cc{j} branches
-are 66\% biased, but the median-of-three branches are only 60\% biased in the
-taken direction. The accuracy with which these branches are correctly predicted
-falls from 71\% to 64\%.
+remaining \n{i} and \n{j} branches. The no-median \n{i} and \n{j} branches are
+66\% biased, but the median-of-three branches are only 60\% biased in the taken
+direction. The accuracy with which these branches are correctly predicted falls
+from 71\% to 64\%.
 

### 2005-08-14 docs/tech_report/shell.tex
fixed captitalization in captions
added info about all the labels. Also capitalized them all

@@
 improved shellsort is shown in Figures \ref{Shellsort code} and \ref{Improved
-Shellsort code}.
+shellsort code}.
 
 \code{Shellsort code}{shellsort.c}
-\code{Improved Shellsort code}{improved_shellsort.c}
+\code{Improved shellsort code}{improved_shellsort.c}
 
@@
 array, on average, during each iteration. In improved shellsort, each key moves
-over 1.07\% places each iteration, which in the final insertion sort, each key
+over 1.07\% places each iteration, while in the final insertion sort, each key
 moves an average of 2.1 places.

### 2005-08-14 docs/tech_report/background.tex
fixed usage of '32 byte' without a hyphen
fixed the captions for the graphs, and a few of the graphs themselves
fixed some missing hyphens in cache types

@@
 causes two separate memory locations to be mapped to the same area of the cache.
-This only occurs in direct mapped or set-associative cache, described below.
+This only occurs in direct-mapped or set-associative cache, described below.
 Finally, a capacity miss occurs when the data sought has been ejected from the
@@
 
-Caches range from direct mapped to fully associative. In a direct mapped cache,
+Caches range from direct-mapped to fully-associative. In a direct-mapped cache,
 every memory address has an exact position in the cache where it can be found.
 Since the cache is smaller than main memory, many addresses map to the same position.
-Fully associative is the opposite of this: data can be stored anywhere in the
+Fully-associative is the opposite of this: data can be stored anywhere in the
 cache. A set-associative cache is a mixture of both: more than one piece of data
@@
 length of the cache line dictate the mapping of memory addresses to cache
-storage. Figure \vref{Addressing} shows how a memory address is divided.
+storage. Figure \vref{Address Division} shows how a memory address is divided.
 
@@
 \includegraphics{cache_addressing}
-\caption{Addressing}
-\label{Addressing}
+\caption{Address Division}
+\label{Address Division}
 \end{figure}
 
-This cache is a 2MB direct mapped cache.  There is a 32 byte cache line, which
+This cache is a 2MB direct-mapped cache.  There is a 32-byte cache line, which
 is indexed by the 5 bit \n{offset}. There are $2^{16}$ cache lines, which are
@@
 the \n{tag}. A cache line is tagged to check if the data stored at an index is
-the same as the data requested. For a fully associative cache, there is no
-index, and all the bits except the offset make up the tag. A fully associative
-cache has no conflict misses for this reason; similarly, a direct mapped cache
+the same as the data requested. For a fully-associative cache, there is no
+index, and all the bits except the offset make up the tag. A fully-associative
+cache has no conflict misses for this reason; similarly, a direct-mapped cache
 has no capacity misses, since even when full, misses are due to conflicts.

### 2005-08-14 docs/tech_report/heap.tex
fixed some missing hyphens in cache types

@@
 level 2 cache, where the flattest slope belongs to the 8-heap. The difference
-between fully associative and direct mapped is only minor, with the fully
-associative performing slightly better than the direct mapped cache.
+between fully-associative and direct-mapped is only minor, with the
+fully-associative performing slightly better than the direct-mapped cache.
 
@@
 associativity: LaMarca's tests were performed on a DEC AlphaStation 250 with a
-direct mapped cache. The Pentium 4's cache is 8-way associative, and so the
+direct-mapped cache. The Pentium 4's cache is 8-way associative, and so the
 unoptimized algorithm would not suffer as much.

### 2005-08-14 docs/tech_report/merge.tex
fixed some missing hyphens in cache types

@@
 slope to take into account the conflict misses of the aligned array segments. The
-fully associative caches should not have these misses, and the results should be
+fully-associative caches should not have these misses, and the results should be
 correspondingly lower. The double mergesorts should have
@@
 
-\plot{merge}{}{}{Results shown are for a direct mapped cache.}{Results shown are
-for a direct mapped cache.}{}{}
+\plot{merge}{}{}{Results shown are for a direct-mapped cache.}{Results shown are
+for a direct-mapped cache.}{}{}
 
@@
 which was taken from LaMarca, is exactly right. However, our assertion that the
-fully associative cache will perform slightly better is false. In actual fact,
-the direct mapped cache performs significantly better (see Figure
+fully-associative cache will perform slightly better is false. In actual fact,
+the direct-mapped cache performs significantly better (see Figure
 \vref{multi-mergesort level 2 misses}). This is due to the replacement policy:
-our fully associative cache removes the least recently used keys from the cache
-to make room, while the direct mapped caches are designed to make best use of
+our fully-associative cache removes the least recently used keys from the cache
+to make room, while the direct-mapped caches are designed to make best use of
 the keys while they are in the cache. They have, effectively, an optimised

### 2005-08-14 docs/tech_report/quick.tex
fixed some missing hyphens in cache types

@@
 
-\plot{quick}{}{}{Results shown are for a direct mapped cache.}{Results shown are
-for a direct mapped cache.}{}{}{}
+\plot{quick}{}{}{Results shown are for a direct-mapped cache.}{Results shown are
+for a direct-mapped cache.}{}{}{}
 
@@
 The results show again (see Section \ref{mergesort replacement policy}) that
-direct mapped caches can perform better than fully associative ones. This is
+direct-mapped caches can perform better than fully-associative ones. This is
 true in the case of all the quicksorts, as can be seen in Figures \ref{base

### 2005-08-14 docs/tech_report/elementary.tex
fixed some missing hyphens in cache types
done spell check. some errors found

@@
 \cc{sim-cache}, simulating an 8KB data cache with a 32-byte cache line and
-separate instruction cache. Results shown are for a direct mapped cache.]
+separate instruction cache. Results shown are for a direct-mapped cache.]
 {\label{elementary level 1 misses}\includegraphics[scale=\myscale]{plots/on2_-_Level_1_misses.eps}}
@@
 unsorted part of the array and moves it to start (end) of the array.
-Thus, we would expect that bubbble sort's branch prediction behaviour
+Thus, we would expect that bubble sort's branch prediction behaviour
 would be similar to that of selection sort.
 
-However, in the proces of moving the smallest element into position,
+However, in the process of moving the smallest element into position,
 bubble sort also rearranges other array elements. Selection sort

### 2005-08-14 docs/tech_report/merge.tex
done spell check. some errors found

@@
 Figure \ref{mergesorts-3}, indicating a possible minimum number. `\n{Left}' and
-`\n{Right}' refer to interating through the first (or left-hand) and second (or
+`\n{Right}' refer to iterating through the first (or left-hand) and second (or
 right-hand) lists being compared, respectively.
@@
 
-This predilication is likely to be due to the fact that the final merge goes
+This predilection is likely to be due to the fact that the final merge goes
 forwards. The left merge step is also favoured, but this is merely due to the

### 2005-08-14 docs/tech_report/quick.tex
done spell check. some errors found

@@
 Figure \ref{base, memory-tuned and multi- quicksorts-1} shows the behaviour of
-the indiviual branches in base quicksort. The \n{i} branch is the loop control
+the individual branches in base quicksort. The \n{i} branch is the loop control
 branch for the index sweeping from left-to-right up the array during partition.
@@
 one will usually be biased in the not-taken direction, and as a result will be
-executed a relativelty small number of times.
+executed a relatively small number of times.
 
@@
 Figure \ref{quick_median_cycles} shows the cycle count of each of the
-quicksorts, showing that a pseduo-median-of-seven offers the greatest tradeoff
+quicksorts, showing that a pseudo-median-of-seven offers the greatest trade-off
 between the reduced number of branches and instructions, and the increase in
@@
 Finally, in multi-quicksort, it may be possible to choose better pivots.
-Choosing better (ie, more evenly distributed pivots) would reduce the number of
+Choosing better (i.e., more evenly distributed pivots) would reduce the number of
 containers. The advantage of this is that it reduces linearly the number of

### 2005-08-14 docs/tech_report/method.tex
done spell check. some errors found
added bit about mispredictions vs misses, and about the scale of software prediction graphs

@@
 files, from which scripts created graphs to display them. Some sample code is
-shown in Figure \ref{Sample code using the software predictor}.
+shown in Figure \ref{Sample code using the software predictor}. Each of the
+graphs produced by the predictors uses the same scale; this way, it is possible
+to visually compare the results from one sort to another.
 
@@
 2 cache, with a 64-byte cache line. The level 1 caches are 8K 4-way associative
-caches, with seperate instruction and data level 1 caches.
+caches, with separate instruction and data level 1 caches.
 

### 2005-08-14 docs/tech_report/conclusions.tex
capitalization fix
done spell check. some errors found
removed the word project, and resolved the occasional TODO

@@
 
-\section{Results summary}
+\section{Results Summary}
 
@@
 S, and write our own mergesort instead. We expected that mergesort would have
-half the instruction count of heapsortl we found instead they were very similar.
+half the instruction count of heapsort we found instead they were very similar.
 
@@
 
-\section{Important results and contributions}
+\section{Important Results and Contributions}
 
@@
 
-\section{Instruction count results}
+\section{Instruction Count Results}
 
@@
 % TODO more details in this
-The test results in this project mostly agreed with LaMarca's results.  The
+The test results in this report mostly agreed with LaMarca's results.  The
 heapsort graphs had nearly exactly the same shape as LaMarca's, and the ratios
@@
 \vspace*{-10em}
-\setcounter{figure}{9} % note the cheating to make this work
+%\setcounter{figure}{9} % note the cheating to make this work
 \begin{figure}[H]
@@
 
-This section lists the contributions of this project. Each of these is novel  
-and was discovered or developed during the course of this project.
-
-The primary contribution of this project is that it repeats the work of LaMarca
+This section lists the contributions of this report. Each of these is novel  
+and was discovered or developed during the course of this report.
+%TODO cant say that
+The primary contribution of this report is that it repeats the work of LaMarca
 on cache-conscious sorting, and validates many of his claims, while questioning

### 2005-08-16 docs/tech_report/merge.tex
removed the word project, and resolved the occasional TODO
merged davids final changes in

@@
 and physical or bus addresses, then this problem is very unlikely. The
-simulations presented in this project address the cache by virtual address, and
+simulations presented in this report address the cache by virtual address, and
 so this optimization is effective in this case.
@@
 
-Figure \ref{mergesorts-1} shows that algorithm N has the same number of branches
-and branch mispredictions as algorithm S, shown in \ref{mergesorts-2}. The
-labels in the graphs correspond to the names of the steps in Knuth's algorithm,
-which also correspond with the labels in our code.
+Figure \ref{mergesorts-1} shows the behaviour of the individual branches within
+algorithm N.  The labels in the graphs correspond to the names of the steps in
+Knuth's algorithm (we also used the same names for the labels of branch targets
+in our C source code. Branch N3 is the main comparison branch; it is used to
+choose the smaller of the elements from the start of the two sub-arrays being
+merged. The direction of this branch is completely random -- it depends entirely
+on the (in this case random) data. Figure \ref{mergesorts-1} also shows that
+this branch is almost entirely unpredictable using a bimodal predictor. The
+result for a two-level predictor is much the same, because there is no pattern
+in the outcome of these branches. Branches `\n{N3 i==j}' (check for the current
+merge being complete), `\n{N5}' and `\n{N9} (both checks for the end of a run of
+increasing elements) are control-flow branches, and are almost entirely
+predictable. Figure \ref{mergesorts-2} shows that the comparison branch in
+Algorithm S is just as unpredictable as that in algorithm N. However, given that
+the merge in algorithm S is entirely regular (see section \ref{Algorithm S}) it
+is possible to eliminate one of the control-flow branches.
 
-Base mergesort has the same number of mispredictions as Knuth's sorts (see
-Figure \ref{mergesorts-3}, indicating a possible minimum number. `\n{Left}' and
-`\n{Right}' refer to iterating through the first (or left-hand) and second (or
-right-hand) lists being compared, respectively.
+Base mergesort is an optimized algorithm that reverses the sorting direction of
+every second sub-array, pre-sorts small sub-arrays, and applies a number of
+other optimizations. As Figure \ref{mergesorts-3} shows, our base mergesort has
+eliminated some of the control-flow branches. The main comparison branch is
+split into two separate branches --- one iterating from left-to-right over the
+increasing sub-array (`\n{Left}'), and the other from right-to-left over the
+decreasing sub-array (`\n{Right}').  However, base mergesort appears to need a
+total of 50\% more comparison branches that either Algorithm S or N.
 
-Base mergesort has a lower number of branches than Knuth's sorts, probably due
-to performing a presort first. The presort has approximately 6.8 misses per key
-(not shown in the graphs). In later mergesorts, an insertion sort is used
-instead.
+In fact, base mergesort eliminated control-flow branches by using the same
+branch that is used for comparison to also deal with the case where one (or
+both) of the merging sub-arrays becomes empty. Thus, the sum of the number of
+executions of the branches `\n{Left}' and `\n{Right}' is almost exactly the same
+as the sum of the number of executions of the branches N3 and N3 i==j in
+Algorithm N. The total number of mispredictions is also almost the same.
 
-Double-aligned tiled mergesort (which has the same results are tiled mergesort)
-has a predilection for merging forwards, shown in \ref{mergesorts-4}. The
-`\n{Left}' and `\n{Right}' in the labels refer to the same as in
-\ref{mergesorts-3}, while `\n{Forwards}' and `\n{Reverse}' refer to whether the
-first or second part of the bitonic array segment is being merged.
+The comparison branch in double-aligned tiled mergesort is divided into four
+separate branches. As in base mergesort, there are `\n{Left}' and `\n{Right}'
+branches. However, for each of these there are separate branches for when
+creating a new sub-array in increasing order (`\n{Forwards}') or decreasing
+order (`\n{Reverse}'). There are slightly more executions of `\n{Forwards}' than
+`\n{Reverse}' because the final array is always sorted into increasing order.
+The left merge step is also favoured, but this is merely due to the fact that it
+occurs first in the code. While there are more left branches, the actual number
+of branches taken are the same.
 
-This predilection is likely to be due to the fact that the final merge goes
-forwards. The left merge step is also favoured, but this is merely due to the
-fact that it occurs first in the code. While there are more left branches, the
-actual number of branches taken are the same.
+Double aligned multi-mergesort is an even more complicated algorithm.  It sorts
+the arrays fully in the cache, leaving them bitonically arranged (sorted into
+sub-arrays, sorted into alternatey increasing and decreasing order). The final
+merge is performed separately, and hence no bias for the forward branches, as
+there was in tiled mergesort. The final sort is performed with a $k$-way merge,
+which is implemented using a heap data structure. This is shown in Figures
+\ref{mergesorts-5} and \ref{mergesorts-6}. The `\n{Comparison}' branches are
+comparisons from the merge. The labels in these graphs are the same as in Figure
+\ref{mergesorts-4} and \ref{base and memory-tuned heapsorts-2}.
 
-In both double-aligned tiled mergesort and base mergesort, approximately two
-thirds are of the loops are taken, unlike Knuth's sorts. However, this appears
-to indicate an increase in branches, as opposed to a decrease in mispredictions.
-Overall this stay is line with the idea that there is a minimum number of branch
-misses, below which comparative sorts will not drop.
+During regular two-way merging, the number of branches is 20\% lower than base
+mergesort, while the number of mispredictions in only 10\% lower. This implies
+that most of the mispredictions occur early in the merging process, and not in
+the final merge, which multi-mergesort leaves out.
 
-Double multi-mergesort sorts the arrays fully in the cache, leaving them
-bitonically arranged. This means that there is no final merge, and hence no bias
-for the forward branches, as there was in tiled mergesort. This is shown in
-Figures \ref{mergesorts-5} and \ref{mergesorts-6}. The labels in these graphs
-are the same as in Figure \ref{mergesorts-4} and \ref{base and memory-tuned
-heapsorts-2}.
-
-While merging, the number of branches is 20\% lower than base mergesort, while
-the number of mispredictions in only 10\% lower. This implies that most of the
-mispredictions occur early in the merging process, and not in the final merge,
-which multi-mergesort leaves out.
-
-Overall, when the k-ary heap is used, there are the same number of branch
-misses. This means that using the technique to reduce cache misses does not
-adversely affect branch prediction rates (it does, however, still affect the
-instruction count).  The rate of branch misses is much the same as that in
-heapsort, and progresses in the same manner (see Figure \ref{memory-tuned
-heapsort curve}). Note that `\n{comparison 0}' occurs every time, and would be
-optimized out by the compiler.
+Overall, when the $k$-ary heap is used, there are approximately the same total
+number of branch mispredictions as in the other mergesort algorithms. Although
+the final $k$-way merge executes a lot more comparison branches, these heap
+branches are quite predictables, as we saw in Section
+\ref{heapsort-branch-prediction}. This means that using the technique to reduce
+cache misses does not adversely affect branch prediction rates (it does,
+however, still affect the instruction count).  The rate of branch misses is much
+the same as that in heapsort, and progresses in the same manner (see Figure
+\ref{memory-tuned heapsort curve}). Note that `\n{comparison 0}' is included for
+illustration purposes only. It is guaranteed to be taken every time, and would
+be optimized out by the compiler.
 

### 2005-08-18 docs/tech_report/heap.tex
done spell check. some errors found
merged davids final changes in
reread predictor section and added comparison between H_N and actual results for short arrays

@@
 \section{A More Detailed Look at Branch Prediction}
+\label{heapsort-branch-prediction}
 
@@
 
-Figure \ref{base and memory-tuned heapsorts-1}(a) shows the behaviour of each
+Figure \ref{base and memory-tuned heapsorts-1} shows the behaviour of each
 branch in the base heapsort algorithm. The `\n{Creation, Has Children}' branch
 refers to the check as to whether a key has children, during the creation of the
-heap The `\n{Creation, Child Comparison}' branch is the comparison between the
+heap. The `\n{Creation, Child Comparison}' branch is the comparison between the
 two children in the 2-heap, and the `\n{Creation, Child Promotion}' refers to
@@
 during the destruction of the heap, and are referred to as `\n{Destruction, Has
-Children}', `\n{Destruction, Child Comparison}' and `\n{Destruction Child
+Children}', `\n{Destruction, Child Comparison}' and `\n{Destruction, Child
 Promotion}'.
@@
 In the memory-tuned heapsort, the creation phase is simpler, involving merely
-one branch, called `\n{Fix-up}' here. The other columns refer to the
-destruction phase, where most of the work takes place in memory-tuned heapsoirt.
-When a 4-heap is now used, there are three comparisons instead of just one.
-Likewise, in an (8-heap), there are seven comparisons.
-
-%TODO read over the merged bits. might not be quite right
+one branch, called `\n{Fix-up}' here. The other columns refer to the destruction
+phase, where most of the work takes place in memory-tuned heapsort.  When a
+4-heap is used (see Figure \ref{base and memory-tuned heapsorts-2}), there are
+three comparisons instead of just one. Similarly, in an 8-heap (Figures
+\ref{base and memory-tuned heapsorts-3} and \ref{base and memory-tuned
+heapsorts-4}), there are seven comparisons.
 
 The results of the software predictor show the relation of the branches
-in the creation and destruction of the heap. The base heapsort results in
-Figure \ref{base and memory-tuned heapsorts-1} show that the time taken to
-create the heap is not significant relative to the time taken to sort from the
-heap.
+in the creation and destruction of the heap. The base heapsort results show that
+the time taken to create the heap is not significant relative to the time taken
+to sort from the heap.
 
@@
 becoming more predictable as the fanout increases. For example, in the 8-heap,
-shown in Figures \ref{base and memory-tuned heapsorts-3} and \ref{base and
-memory-tuned heapsorts-4}, the first comparison is predicted correctly 50\% of
-the time, while the next is predicted correctly 61\% of the time, then 71\%,
-77\%, and so on in a curve, which is shown in Figure \ref{memory-tuned heapsort
-curve}.
+the first comparison is predicted correctly 50\% of the time, while the next is
+predicted correctly 61\% of the time, then 71\%, 78\%, and so on in a curve,
+which is shown in Figure \ref{memory-tuned heapsort curve}.
+
+The relationship between this number and $H_N-1$ is relatively straightforward.
+$H_N-1$ represents the cumulative number of changes in the minimum. Over a long
+array, this approximates the number of branch mispredictions, as almost every
+branch will not be taken, and the predictor will always predict not-taken. The
+graph shows us the rate of correct predictions, and each point is calculated
+using a seperate branch predictor. As a result, we expect $1-y = \frac{1}{x+1}$,
+where $x$ and $y$ are the cartesian coordinates of each point on the graph.
+$1-y$ is the number of mispredictions from the graph, while $\frac{1}{x+1}$ is
+the increase of $H_N$ for step $N$.
+
+When the array is short, as in this example, the branch predictor has too much
+feedback to be as accurate as $H_N$. For $N=2$, we would approximate a 33\%
+misprediction ratio using $H_N$, but we measure 39\%. The approximated and
+actual ratios become closer as $N$ increases, as we would expect, and for $N=7$,
+we approximate 12.5\% and measure 12.7\%.
 
@@
 \label{heapsort entropy}
-We observe here that the more branches there are, the more are correctly
-predicted, and that there is a threshold below which the number of branches will
-not drop. This is discussed in more Section \ref{quicksort entropy}.
-
-Also notable from the results is that branches are slightly more often not-taken
-then they are correctly predicted, implying that a semi-static predictor based
-on hints from the compiler could do a better job at making correct predictions.
+We observe also that the more branches there are, the more are correctly
+predicted, and that there may a threshold below which the number of branches will
+not drop. This is discussed further in Section \ref{quicksort entropy}.
 
@@
 chapter.
+
+It should be possible to more accurately model and therefore predict the branch
+prediction ratio when the array is short. Our results show us what we expect the
+answer to be, but have not generated a formula to predict it.

### 2005-08-18 docs/tech_report/heap.tex
tocuhed up, and added mini-conclusion

@@
 array, this approximates the number of branch mispredictions, as almost every
-branch will not be taken, and the predictor will always predict not-taken. The
-graph shows us the rate of correct predictions, and each point is calculated
-using a seperate branch predictor. As a result, we expect $1-y = \frac{1}{x+1}$,
-where $x$ and $y$ are the cartesian coordinates of each point on the graph.
-$1-y$ is the number of mispredictions from the graph, while $\frac{1}{x+1}$ is
-the increase of $H_N$ for step $N$.
+branch will not be taken, and the predictor will always predict not-taken.
+
+Meanwhile, Figure \ref{memory-tuned heapsort curve} shows us the rate of correct
+predictions, and each point is calculated using a seperate branch predictor. As
+a result, we expect $1-y = \frac{1}{x+1}$, where $x$ and $y$ are the cartesian
+coordinates of each point on the graph.  $1-y$ is the number of mispredictions
+from the graph, while $\frac{1}{x+1}$ is the increase in $H_N$ for step $N$.
 
 When the array is short, as in this example, the branch predictor has too much
-feedback to be as accurate as $H_N$. For $N=2$, we would approximate a 33\%
+feedback to be close to $H_N$. For $N=2$, we would approximate a 33\%
 misprediction ratio using $H_N$, but we measure 39\%. The approximated and
 actual ratios become closer as $N$ increases, as we would expect, and for $N=7$,
-we approximate 12.5\% and measure 12.7\%.
+we approximate 12.5\% and measure 12.7\%. We conclude that initial suggestion is
+correct: while the branch prediction ratio of finding the minimum is very close
+to $H_N-1$, the relationship is less accurate for very short arrays.
 

### 2005-08-18 docs/tech_report/conclusions.tex
fixed up the heap stuff in the conclusion

@@
 our 8-heap, and that this pattern matched closely that of selection sort,
-discussed above. Despite this property, the number of mispredictions using an
-8-heap was greater than when using a 4-heap.
+discussed above. We found that the approximated results using $H_N$ was similar to
+our expected results for large arrays, and followed a similar pattern for small
+arrays.
+
+Also in heapsort, the number of mispredictions decreases depending on how much
+the heap fans out.  Despite this property, the number of mispredictions using an
+8-heap was greater than when using a 4-heap, due to the increased total number
+of branches.
 

### 2005-08-18 docs/tech_report/quick.tex
minor changes to quicksort predictor stuff on reread

@@
 
-Base quicksort has a 35\% branch misprediction rate, whereas the rate for
-memory-tuned quicksort is only 34\%. The main reason for this is the reduction
-in branches by not attempting to move the pivot elements.  Attempting to move
-these elements usually results in a branch misprediction. The main reason is
-that most other elements move at least one position, and thus cause the
-insertion sort loop to iterate at least once. But the pivot elements never move
--- they are already in the correct place.
+Overall, base quicksort has a 35\% branch misprediction rate, whereas the rate
+for memory-tuned quicksort is only 34\%. The main reason for this is the
+reduction in branches by not attempting to move the pivot elements.  Attempting
+to move these elements usually results in a branch misprediction.  The main
+reason is that most other elements move at least one position, and thus cause
+the insertion sort loop to iterate at least once. But the pivot elements never
+move -- they are already in the correct place.
 
@@
 
-The variation of multi-quicksort that uses sequential search dramatically
-increases the number of executed branches, but these branches are highly
-predictable. This version actually causes fewer branch mispredictions than
-memory-tuned quicksort.
+The variation of multi-quicksort that uses sequential search (see Figure
+\ref{base, memory-tuned and multi- quicksorts-4}) dramatically increases the
+number of executed branches, but these branches are highly predictable. This
+version actually causes fewer branch mispredictions than memory-tuned quicksort.
 
-% Theres a lot of branches in the multi quicksort. WE saw in other results that
-% the mere instruction cost of this bogs down to algorithm significantly. in the
-% binary search version, there are more than twice as many mispredictions,
-% despite the fact that there are less than half the branches. (note that i dont
-% think this was a wonderful binary search, its heavily biased towards the
-% checking the left). binary search predicts correctly 57 percent of the time,
-% meaning that it mostly changes direction haphazardly during the search. No
-% wonder it performs so badly.
-
-% Note that sequential multi-quicksort has less mispredictions than cache quicksort.
 \subsection{Different Medians}
@@
 see that the \n{i} and \n{j} branches are indeed taken about twice as often as
-not-taken, and the \n{i} and \n{j} branches are correctly predicted around 71\%
-of the time.
+not-taken, and the \n{i} and \n{j} branches are correctly predicted
+approximately 71\% of the time.
 
@@
 used to select the pivot. Perhaps the biggest effect is that the number of
-executions of the \n{i} and \n{j} branches falls by around 15\%. This fall is
-partly offset by the comparison branches to compute the median, but these
-account for only around 1.25\% of the original branches from the no-median
+executions of the \n{i} and \n{j} branches falls by approximately 15\%. This
+fall is partly offset by the comparison branches to compute the median, but
+these account for only around 1.25\% of the original branches from the no-median
 quicksort. Thus, there is a reduction of almost 14\% in the number of executed
@@
 Figure \ref{quick_median_cycles} shows the cycle count of each of the
-quicksorts, showing that a pseudo-median-of-seven offers the greatest trade-off
+quicksorts, showing that a median-of-three offers the greatest trade-off
 between the reduced number of branches and instructions, and the increase in
-branch mispredictions.
+branch mispredictions. This shows the cost of the extra branch mispredictions.
+It appears that it can be more costly to remove unnecessary entropy than to
+endure it.
 
@@
 Two areas of current research into quicksort are those of equal keys, and
-median-of-greater-than-3. The former affects the algorithmic properties, and
+median-of-greater-than-three. The former affects the algorithmic properties, and
 some steps may need to be varied to avoid extra cache misses as a result

### 2005-08-18 docs/tech_report/conclusions.tex
added quicskort conclusions

@@
 
-TODO quick
+Sedgewick proposed removing individual insertion sorts in quicksort with one
+elegant sweep at the end. LaMarca proposed replacing the large number of
+insertion sorts in order to reduce cache misses. We note that attempting to move
+the pivots, as Sedgewick does, actually increases both the number of branches
+and the rate of mispredictions. We conclude that LaMarca's improvment, which
+slightly improves quicksorts cache performance, also slightly improves it's
+branch prediction performance.
 
-TODO radix
+However, his move to multi-quicksort, while not affecting the rate of branch
+mispredictions in the partitioning phase, results in a large increase in branch
+mispredictions due to the addition of a binary search. We determine that this
+while reducing the instruction count, the increase is actual running time due to
+branch mispredictions is significant.
+
+We also analysed the effects of different medians on quicksort. We deterine that
+not using a median to choose a pivot makes the branches more predictable than
+they are biased, though at a cost of a large increase in instructions (and also
+significantly greater possability of a worst-case $O(N^2)$ sort). We also saw
+that the more keys considered when determining the median, the lower the number
+of branches. However, this can result in an increase in the rate of branch
+mispredictions. We determine that reducing entropy is not the best way of making
+quicksort perform faster, and that using a median-of-three actually achieves the
+higher performance.
+
+As predicted, radixsort is highly predictable, having no comparison
+mispredictions at all. This result, combined with its linear complexity, and its
+improved cache performance, mean that it is the fastest of any of the sorts
+presented here.
 

### 2005-08-18 docs/tech_report/quick.tex
added quicskort conclusions

@@
 
-In the case that the array to be sorted fits in a particular level of
-the cache, then quicksort has excellent cache properties at that
-level. Once the array is larger than the cache, however, the
-number of misses increases dramatically. To fix this problem, LaMarca uses a
-technique similar to that used in multi-mergesort. Instead of multi-merging a
-series of sorted lists at the end, a \n{multi-way partition} is done at the
-start, moving all keys into cache sized containers which can be sorted
-individually.
+In the case that the array to be sorted fits in a particular level of the cache,
+then quicksort has excellent cache properties at that level. Once the array is
+larger than the cache, however, the number of misses increases dramatically. To
+fix this problem, LaMarca uses a technique similar to that used in
+multi-mergesort. Instead of multi-merging a series of sorted lists at the end, a
+\n{multi-way partition} is done at the start, moving all keys into cache sized
+containers which can be sorted individually.
 
@@
 sequential search version's 50\% increase. However, the increase we predicted in
-the instruction count of memory-tuned quicksort based on the results of both Sedgewick
-and LaMarca -  that is, once the efficient insertion sort was replaced with many
-small insertion sorts - we observe that the instruction count remains exactly
-the same.
+the instruction count of memory-tuned quicksort based on the results of both
+Sedgewick and LaMarca - that is, once the efficient insertion sort was replaced
+with many small insertion sorts - we observe that the instruction count remains
+exactly the same.
 
@@
 Overall, base quicksort has a 35\% branch misprediction rate, whereas the rate
-for memory-tuned quicksort is only 34\%. The main reason for this is the
-reduction in branches by not attempting to move the pivot elements.  Attempting
-to move these elements usually results in a branch misprediction.  The main
-reason is that most other elements move at least one position, and thus cause
-the insertion sort loop to iterate at least once. But the pivot elements never
-move -- they are already in the correct place.
+for memory-tuned quicksort is only 34\%, while the number of branches also
+decreases. The main reason for this is the reduction in branches by not
+attempting to move the pivot elements. Attempting to move these elements
+usually results in a branch misprediction. The main reason is that most other
+elements move at least one position, and thus cause the insertion sort loop to
+iterate at least once. But the pivot elements never move -- they are already in
+the correct place.
 

### 2005-08-18 docs/tech_report/conclusions.tex
branch prediction summary for shellsorts

@@
 
-TODO shell
-
-%TODO results from software predictors
+We are able to accurately predict the number of branch mispredictions in
+shellsort, using our results from insertion sort. We also found that the
+improvement mode to shellsort, seperating the final loop from the rest, results
+in keys being moved 1.07 places, up from 1.03, contributing to its improved
+performance.
 

### 2005-08-18 docs/tech_report/shell.tex
branch prediction summary for shellsorts

@@
 
-%TODO
 \section{A More Detailed Look at Branch Prediction}
@@
 
-%TODO see figure something
 In shellsort, almost exactly half of the important branches are mispredicted.
@@
 array, on average, during each iteration. In improved shellsort, each key moves
-over 1.07\% places each iteration, while in the final insertion sort, each key
+over 1.07 places each iteration, while in the final insertion sort, each key
 moves an average of 2.1 places.
@@
 The expected number of branch mispredictions is $log_{(\frac{11}{5})}N$. This
-is almost exactly correct, being slightly lower (18.48 mispredictions per key,
+is almost exactly correct, being slightly lower (18.48 mispredictions per key
 instead of 19.34, for 4194304 keys).

### 2005-08-26 docs/tech_report/background.tex
added bit about mispredictions vs misses, and about the scale of software prediction graphs
fixed from davids notes

@@
 
-Each branch can be resolved in two ways: \n{taken} or \n{not-taken}. If the branch is
-taken, then the program counter moves to the address provided by the branch
-instruction. Otherwise, the program counter increments, and takes the next
-sequential instruction.
+Each branch can be resolved in two ways: \n{taken} or \n{not-taken}. If the
+branch is taken, then the program counter moves to the address provided by the
+branch instruction. Otherwise, the program counter increments, and takes the
+next sequential instruction.
 
@@
 and the next instruction must be fetched.
+
+Branch mispredictions are sometimes called \n{branch misses}, due to the
+similarity between dynamic predictors and caches. In this report, we use the
+terms interchangably.
 
@@
 
-In the sequence T-T-T-N-T-T-T-N, a 2-bit predictor will only mispredict once.
-In general, it will achieve 78 to 89\% accuracy.
+In the sequence T-T-T-N-T-T-T-N, a 2-bit predictor will only mispredict once per
+loop iteration. In general, it will achieve 78 to 89\% accuracy.
 
@@
 instructions each step will take to execute. It is possible, therefore, that an
-algorithm with a lower complexity will complete slower than an algorithm where
-each step is executed very quickly, but which has a greater order of complexity.
+algorithm with a lower complexity will complete more slowly than an algorithm
+where each step is executed very quickly, but which has a greater order of
+complexity.
 

### 2005-08-26 docs/tech_report/elementary.tex
fixed from davids notes

@@
 
-The level 1 cache should have bad performance as a result. Since it is smaller
-than the array, the performance of the level 1 cache should simulate the
-performance of the level 2 cache.
+The level 1 cache should have bad performance as a result. Since the level 1
+cache is smaller that the arrays tested, it should perform analagiously to a
+level 2 cache being tested with a much larger array.
 
@@
 Within practical sorting ranges, $H_N$ grows at a rate similar to
-$log(N)$. Thus, in the average case, we can expect that the comparison
-branch will mostly go in the same direction, making the branch highly
-predictable.
+$log(N)$\footnote{This seems like an error, since $H_N$ is very similar to
+$log_eN$. It is not. We mean that $H_N$ is very similar to $log_2N$.}. Thus, in
+the average case, we can expect that the comparison branch will mostly go in the
+same direction, making the branch highly predictable.
 
@@
 $O(NlogN)$ instructions, but this is small in comparison to the number of
-instructions saved if the sort ended early.
+instructions saved if the sort ends early.
 

### 2005-08-26 docs/tech_report/method.tex
removed the word project, and resolved the occasional TODO
fixed from davids notes

@@
 discussed in Section \ref{Main}. It was also necessary to fully understand the
-technologies to be used in the project. Several texts were used for this: cache
+technologies to be used in the report. Several texts were used for this: cache
 and branch predictors are both in \cite{HennessyPatterson90}; branch predictors
@@
 \subsection{SimpleScalar} \n{SimpleScalar} is an architecture simulator. It
-works by compiling code into its native instruction set, PISA, then emulating a
-processor as it runs the program.  SimpleScalar can be found at
+works by compiling code into its native instruction set, \n{PISA}, then
+emulating a processor as it runs the program. SimpleScalar can be found at
 \n{http://www.simplescalar.com}. Documentation on how to use and modify
@@
 compiler - a port of gcc - with related tools such as a linker and assembler.
-Several simulators are then provided: \cc{sim-fast} only counts instruction,
-quickly; \cc{sim-cache} simulates a cache and counts level 1 and level 2 cache
+Several simulators are then provided: \cc{sim-fast} only counts instructions
+executed; \cc{sim-cache} simulates a cache and counts level 1 and level 2 cache
 misses; \cc{sim-bpred} simulates a branch predictor and counts branch hits and
@@
 \subsection{PapiEx}
-\cc{PapiEx} is a tool used to access performance counters on a Pentium 4. It
-uses the PAPI framework to count certain characteristics across a program's
+\n{PapiEx} is a tool used to access performance counters on a Pentium 4. It uses
+the \n{PAPI} framework to count certain characteristics across a program's
 execution. We used it to count the number of cycles a sort took to run. This

### 2005-08-26 docs/tech_report/report.tex
removed the word project, and resolved the occasional TODO
added the numbers to the counter graphs
fixed from davids notes

@@
 \vspace*{6em}
-\caption[Branch prediction performance for #7]{Branch prediction performance for #7. All graphs use the same scale.}
+\caption[Branch prediction performance for #7]{Branch prediction performance for
+#7. All graphs use the same scale. Above each column is the total number of
+branches and the percentage of correctly predicted branches for a simulated
+bimodal predictor.}
 \label{Branch prediction performance for #7}
@@
 %TODO make the page wider
-%TODO aspell in UK english
-%TODO sometimes I say 3 sorts, sometimes 4. It's now 5.
-%TODO do a final check for use of the word I
-%TODO final check for the word 'project'
 
@@
 particular we show that insertion sort causes the fewest branch
-mispredictions of any comparison-based algorithm, that optimizations,
-such as the choice of the pivot in quicksort, can have a large impact
+mispredictions of any comparison-based algorithm, that optimizations -
+such as the choice of the pivot in quicksort - can have a large impact
 on the predictability of branches, and that advanced two-level branch

### 2005-08-26 docs/tech_report/elementary.tex
more changes due to davids minor edits

@@
 It should be possible to make more cache-conscious versions of these sorts.
-Making a bubblesort which performs a bubble across an entire cache line would
-be an example. However, these would only make a difference on large data sets,
+Making a bubblesort which performs a bubble across an entire cache line would be
+an example. Another would be a cache-conscious selection sort. It might switch
+the direction of the search to reuse lines at the end of the array. A
+cache-conscious bubblesort could also do something similar, which would make it
+into shakersort. However, these would only make a difference on large data sets,
 when better sorts such as quicksort should be used. In addition, the complexity
-of doing this means it may be easier to code a simple $O(NlogN)$ sort, which
-would perform better for the same effort. Due to this, a cache-conscious
-elementary sort is but an academic exercise.
+of improving elementary sorts means it may be easier to code a simple $O(NlogN)$
+sort, which would perform better for the same effort. Due to this, a
+cache-conscious elementary sort is but an academic exercise.

### 2005-08-26 docs/tech_report/heap.tex
fixed minor typo
more changes due to davids minor edits

@@
 to two other nodes, which have the same structure. The first node is called the
-\n{parent}, while the linked nodes are called \n{children}. One node is called the
-\n{root}, and the tree branches out from this, with each level having twice as many
-nodes as the previous level.
+\n{parent}, while the linked nodes are called \n{children}. There is one node
+with no parents, called the \n{root}, from which the tree branches out, with
+each level having twice as many nodes as the previous level.
 
 A heap is a type of binary tree with several important properties. Firstly, the
-the data stored by the parent must always be greater than the data stored by its
-children. Secondly, in a binary tree there is no requirement that the tree be full.
-In a heap, it is important that every node's left child exists before its right
-child. With this property, a heap can be stored in an array without any wasted
-space, and the links are implied by a nodes position, rather than by a pointer
-or reference. A heap done in the fashion is called an \n{implicit heap}.
+data stored by the parent must always be greater than the data stored by its
+children. Secondly, in a binary tree there is no requirement that the tree be
+full.  In a heap, it is important that every node's left child exists before its
+right child. With this property, a heap can be stored in an array without any
+wasted space, and the links are implied by a nodes position, rather than by a
+pointer or reference. A heap done in the fashion is called an \n{implicit heap}.
 
@@
 notes speed improvements only when comparisons are move expensive than swaps,
-such as in the case of sorting strings. For integers, this step should not be
-done.
+such as in the case of sorting strings. In our case, the comparisons are very
+predictable moving down the heap, and so this step should not be done.
 
@@
 as the number of keys is static. When destroying the heap, however, the
-maximal key must be added to every second operation. This can be avoided
-relatively simply with no extra instructions. If the last key in the heap is
-left in place as well as being sifted down the heap, it will provide a
-sentinel for the sift-down operation. The key from the top of the heap is then
-put into place at the end. This technique reduces instruction count by
+maximal key must be added to every second operation. We found this can be
+avoided with an innovative technique, with no extra instructions. If the last
+key in the heap is left in place as well as being sifted down the heap, it will
+provide a sentinel for the sift-down operation. The key from the top of the heap
+is then put into place at the end. This technique reduces instruction count by
 approximately 10\%.
@@
 An 8-heap has seven comparisons between children, and will probably have about
-2.58 mispredictions, as it behaves similarly to selection sort (See \ref{selection
+1.59 mispredictions, as it behaves similarly to selection sort (See \ref{selection
 branches harmonics}). It is possible, therefore, that the 8-heap will have slightly

### 2005-08-26 docs/tech_report/merge.tex
added a label
more changes due to davids minor edits

@@
 \subsection{Algorithm S}
+\label{Algorithm S}
 
@@
 the final $k$-way merge executes a lot more comparison branches, these heap
-branches are quite predictables, as we saw in Section
+branches are quite predictable, as we saw in Section
 \ref{heapsort-branch-prediction}. This means that using the technique to reduce
@@
 \ref{memory-tuned heapsort curve}). Note that `\n{comparison 0}' is included for
-illustration purposes only. It is guaranteed to be taken every time, and would
-be optimized out by the compiler.
+illustration purposes only. It is a comparison between two values, one of which
+has just been assigned to the other, and which are therefore equal. It is
+guaranteed to be taken every time, and would be optimized out by the compiler.
 

### 2005-08-26 docs/tech_report/quick.tex
more changes due to davids minor edits

@@
 
-LaMarca uses the optimized quicksort described by Sedgewick for his base
-quicksort. This is an iterative quicksort which uses an explicit stack. In the
-worst case, this stack must be the same size as the array, and must be
-initialised at the start of the algorithm. To reduce the size of the stack,
-Sedgewick proposes sorting the smallest partition first. This optimization
-reduces the maximum size of the stack to $O(logN)$. In practice, two \cc{int}s,
-the left and the right index, are required for each partition.
+LaMarca uses the optimized quicksort described by Sedgewick in
+\cite{Sedgewick02} for his base quicksort. This is an iterative quicksort which
+uses an explicit stack. In the worst case, this stack must be the same size as
+the array, and must be initialised at the start of the algorithm. To reduce the
+size of the stack, Sedgewick proposes sorting the smallest partition first. This
+optimization reduces the maximum size of the stack to $O(logN)$. In practice,
+two \cc{int}s, the left and the right index, are required for each partition.
 
@@
 sort is used to sort the partition. Sedgewick states that roughly the same
-results are obtained for thresholds between 5 and 25. An additional reduction in
-instruction count can be had by doing the insertion sort over the entire array
-at the end, rather than across each small array as soon as the threshold is
-reached. At the end of the partition sorting, LaMarca places a sentinel at the
-end of the array and uses an insertion sort, eliminating the bounds check from
-the end. In this implementation, the sentinel is placed at the front, and the
-smallest of the first \cc{THRESHOLD} keys is used as a sentinel, in keeping
+results are obtained for thresholds between five and 25. An additional reduction
+in instruction count can be had by doing the insertion sort over the entire
+array at the end, rather than across each small array as soon as the threshold
+is reached. At the end of the partition sorting, LaMarca places a sentinel at
+the end of the array and uses an insertion sort, eliminating the bounds check
+from the end. In this implementation, the sentinel is placed at the front, and
+the smallest of the first \cc{THRESHOLD} keys is used as a sentinel, in keeping
 with our policy of not exceeding the bounds of the array.
@@
 
-LaMarca notes that quicksort is already very cache efficient as it has excellent
-spatial and temporal locality. Once it no longer fits in the cache, however, the
-number of misses should increase significantly. This is especially true in the
-case of the base quicksort, which does an insertion sort across then entire
-array at the end.
+LaMarca notes that quicksort is already very cache efficient\footnote{We
+speculate wildly at this point and suggest that since an average of 25\% of
+computing time is spent sorting (see \cite{TODO}), the development of caches has
+been subconsicously affected by the cache performance of quicksort.}, as it has
+excellent spatial and temporal locality. Once it no longer fits in the cache,
+however, the number of misses should increase significantly. This is especially
+true in the case of the base quicksort, which does an insertion sort across then
+entire array at the end.
 
@@
 When we originally looked at quicksort, we expected that the \n{i} and \n{j}
-branches in quicksort would resolve in each direction. In fact, this is not the
-case. The main reason is that the chosen pivot is almost never the true median
-of the section of the array to be partitioned. Thus, the \n{i} and \n{j}
-branches are almost always biased, which makes them predictable. According to
-Knuth, if the pivot element is chosen randomly, then the \n{i} and \n{j}
-branches will be taken (i.e. the \n{i} or \n{j} loop will execute) twice as
-often as they are not taken (i.e. the loop exits).
+branches in quicksort would resolve in each direction an equal number of times.
+In fact, this is not the case. The main reason is that the chosen pivot is
+almost never the true median of the section of the array to be partitioned.
+Thus, the \n{i} and \n{j} branches are almost always biased, which makes them
+predictable. According to Knuth, if the pivot element is chosen randomly, then
+the \n{i} and \n{j} branches will be taken (i.e. the \n{i} or \n{j} loop will
+execute) twice as often as they are not taken (i.e. the loop exits).
 
@@
 right-hand end of the region to be partitioned. Thus, the \n{i} loop will
-execute a large number of times, and thus the \n{i} loop will be biased towards
-being taken. The \n{j} loop, on the other hand, will be biased toward not-taken,
-because the average element is likely to be smaller than the pivot. Therefore,
-the \n{j} loop will execute only a small number of times. Thus, we see that in a
-particular partition, we do not normally get an equal number of executions of
-the \n{i} and \n{j} loops. Instead we get a large number of executions in one
-loop (which makes its branch relatively predictable), whereas the other loop
-often does not enter at all (which makes its branch also predictable, although
-executed a much smaller number of times). Thus, although overall the \n{i} and
-\n{j} branches are taken twice as often as not-taken, on a particular partition
-one will usually be biased in the not-taken direction, and as a result will be
-executed a relatively small number of times.
+execute a large number of times, and thus the \n{i} branch will be biased
+towards being taken. The \n{j} branch, on the other hand, will be biased toward
+not-taken, because the average element is likely to be smaller than the pivot.
+Therefore, the \n{j} loop will execute only a small number of times. Thus, we
+see that in a particular partition, we do not normally get an equal number of
+executions of the \n{i} and \n{j} loops. Instead we get a large number of
+executions in one loop (which makes its branch relatively predictable), whereas
+the other loop often does not enter at all (which makes its branch also
+predictable, although executed a much smaller number of times). Thus, although
+overall the \n{i} and \n{j} branches are taken twice as often as not-taken, on a
+particular partition one will usually be biased in the not-taken direction, and
+as a result will be executed a relatively small number of times.
 
@@
 results, especially instruction count and branch misses, vary with changes 
-to the threshold may be interesting.
+to the threshold may be interesting. In particular, we expect that a larger
+threshold would be beneficial, as insertion sort causes far fewer branch
+mispredictions than quicksort.
 

### 2005-08-26 docs/tech_report/method.tex
minor edits due to david, and amortizZZZZZ

@@
 results for small set sizes, and larger set sizes, for which this time is
-amortized. As a result, we measured this time, and subtracted it from our
+amortised. As a result, we measured this time, and subtracted it from our
 results.

### 2005-08-26 docs/tech_report/radix.tex
removed the word project, and resolved the occasional TODO
minor edits due to david, and amortizZZZZZ

@@
 
-All other sorts considered in this project are \n{comparison-based} sorts.
+All other sorts considered in this report are \n{comparison-based} sorts.
 They compare two keys at a time, and determine the larger and smaller key,
@@
 
-Memory-tuned radixsort should traverse the array only nine times: Once at the
-start to count the least significant bytes; six times (once per array per pass)
-as it copies using the three least significant bytes (counting the next least
-significant byte at the same time), and a final two as it copies using the most
-significant byte.
+For unsigned 32-bit integers, memory-tuned radixsort should traverse the two
+array only nine times in total: Once at the start to count the least significant
+bytes; six times (once per array per pass) as it copies using the three least
+significant bytes (counting the next least significant byte at the same time),
+and a final two as it copies using the most significant byte.
 
@@
 In Section \ref{tiled mergesort}, mergesort is improved by aligning its array
-with reference to it auxiliary array. Specifically, extra memory is allocated
+with reference to its auxiliary array. Specifically, extra memory is allocated
 for the auxiliary array, which is then offset so that the start of the two
@@
 The number of branch prediction misses should be exceptionally low. In
-comparison-based sorts, a branch due to a comparison can not be predicted except
-in exceptional circumstances\footnote{See Section \ref{insertion is
-predictable}.}. Radixsort has no comparative branches at all. As a result, only
-flow control branches occur in radixsort, and they are all predictable.
+comparison-based sorts, there are generally frequent mispredictions\footnote{See
+Section \ref{insertion is predictable} for an exception.}. Radixsort has no
+comparative branches at all. As a result, only flow control branches occur in
+radixsort, and they are all predictable.
 
@@
 
-Aligned memory-tuned radixsort performs exactly the same as memory-tuned radixsort in both
-level 1 and level 2 misses, indicating that the alignment, and the associated
-memory overhead, are not necessary.
+Aligned memory-tuned radixsort performs exactly the same as memory-tuned
+radixsort in both level 1 and level 2 misses, indicating that the alignment, and
+the associated memory overhead, are not necessary.
 
@@
 miss per key. The number of branches itself is quite low, as several of the
-sorts considered had nearly that number of misses. However, it is not
+sorts considered had more thna that number of mispredictions. However, it is not
 unexpected. The number of branches reduces when copying back is removed, and
@@
 flat. He shows a similar result in both cache miss graphs: a sharp dip, followed
-by a nearly flat-line. This may be due to amortization, in which case the trend
-would not be expected in our graphs due to subtraction of the time taken to fill
-the array.
+by a nearly flat-line. This is due to the instructions, time and compulsory
+misses being amortised over the ever larger array size. The trend is not
+expected in our graphs due to subtraction of the time taken to fill the array.
 

### 2005-08-26 docs/tech_report/conclusions.tex
most of davids edits in

@@
 
-However, his move to multi-quicksort, while not affecting the rate of branch
-mispredictions in the partitioning phase, results in a large increase in branch
-mispredictions due to the addition of a binary search. We determine that this
-while reducing the instruction count, the increase is actual running time due to
-branch mispredictions is significant.
+However, LaMarca's move to multi-quicksort, while not affecting the rate of
+branch mispredictions in the partitioning phase, results in a large increase in
+branch mispredictions due to the addition of a binary search. We determine that
+this while reducing the instruction count, the increase is actual running time
+due to branch mispredictions is significant.
 
@@
 
-LaMarca's initial improvements to base mergesort were due to reduce the
+LaMarca's initial improvements to base mergesort were expected to reduce the
 instruction count. We noted a reduction with each of the steps he prescribed. We

### 2005-08-26 docs/tech_report/shell.tex
most of davids edits in

@@
 Shellsort is a derivative of insertion sort designed to offer a compromise
-between the performance of mergesort and radixsort, and the space efficiency of
-selection sort, bubblesort and insertion sort. It was designed by Donald Shell
-in 1959, and described in \cite{Shell59}\footnote{Shell describes the former
-sorts as \n{Merging Pairs} and \n{Floating Decimal Sort}, respectively, and the
-latter sorts as \n{Finding the Smallest}, \n{Interchanging Pairs} and
-\n{Sifting}, respectively.}.
+between the performance of mergesort and radixsort\footnote{Quicksort and
+heapsort were developed only later, so at the time there was no efficient
+in-place sort.}, and the space efficiency of selection sort, bubblesort and
+insertion sort. It was designed by Donald Shell in 1959, and described in
+\cite{Shell59}\footnote{Shell describes the former sorts as \n{Merging Pairs}
+and \n{Floating Decimal Sort}, respectively, and the latter sorts as \n{Finding
+the Smallest}, \n{Interchanging Pairs} and \n{Sifting}, respectively.}.
 
@@
 http://www.cs.princeton.edu/\textasciitilde{}rs/shell/}, Gonnet's shellsort was
-seen to be the fastest on the authors local machine. As a result, this was used
+seen to be the fastest on the authors' local machine. As a result, this was used
 to test the performance of shellsort. This uses increments which are
@@
 there will be no spatial reuse, and little temporal reuse, so long as the
-sub-array is large than the cache. Once the increment is small enough that each
+sub-array is larger than the cache. Once the increment is small enough that each
 sub-array fits within the cache (i.e. quickly for L2 cache, and much more slowly
 for the L1 cache), then temporal reuse will occur. Once an entire sub-array fits
-within an eighth of the size of the cache (eight being the number of keys
-that fit in a cache line), then spatial reuse will occur as well. This reuse
-will only occur within a single iteration; once the increment with the next
-increment begins, most of the contents of the cache will be invalidated, unless
-the entire array fits within the cache.
+within an eighth of the size of the cache (eight being the number of keys that
+fit in a cache line), then spatial reuse will occur as well. This reuse will
+only occur within a single iteration; once the increment with the next increment
+begins, most of the contents of the cache will be invalidated, unless the entire
+array fits within the cache.
 
@@
 instructions. It also has slightly fewer cache misses and branch mispredictions.
-Overall, this leads to a reduction in cycle count of between 35 and 70\%, as
+Overall, this leads to a reduction in cycle count of between 35\% and 70\%, as
 seen in Figure \ref{shell cycles}.

### 2005-08-27 docs/tech_report/elementary.tex
minor edits and merged in davids writing

@@
-\label{ordernsquared} Elementary sorts are also referred to as $O(N^2)$ sorts.
-This is due to the manner in which they check every key in the array in order to
-sort a single key; for every key to be sorted, every other key is checked. Four
-of these sorts are discussed here: insertion sort, selection sort, bubblesort
-and shakersort, along with improved versions of bubblesort and shakersort.
+\label{ordernsquared}
+Elementary sorts are also referred to as $O(N^2)$ sorts. This is due to the
+manner in which they check every key in the array in order to sort a single key;
+for every key to be sorted, every other key is checked. Four of these sorts are
+discussed here: insertion sort, selection sort, bubblesort and shakersort, along
+with improved versions of bubblesort and shakersort.
 
@@
 The performance of selection sort is not affected by input, so the instruction
-count varies very little. Its cache behaviour is also not affected by input,
-and selection sort performs very badly from a cache perspective. There are $N$
+count varies very little. Its cache behaviour is also not affected by input, and
+selection sort performs very badly from a cache perspective. There are $N$
 traversals of the array, leading to bad temporal reuse. If the array doesn't fit
@@
 
-Knuth also presents the number of changes to the left-to-right minimum
-in the best case (0 changes) and worst cases ($N-1$ changes). In the
-worst case, the array is sorted in reverse order, and so the
-left-to-right minimum changes on every element. Interestingly, the
-worst cases for the number of changes is not the worst case for branch
-mispredictions. If the minimum changes for every element, then the
-comparison branch will always resolve in the same direction, and thus
-be almost perfectly predictable. From the point of view of branch
-prediction, frequent changes (preferably with no simple pattern) are
-the main indicator of poor performance, rather than the absolute
-number of times that the branch resolves in each direction. The worst
-case for selection sort would probably be an array which is partially
-sorted in reverse order --- the left-to-right minimum would change
-frequently, but not predictably so.
+Knuth also presents the number of changes to the left-to-right minimum in the
+best case (0 changes) and worst cases ($N-1$ changes). In the worst case, the
+array is sorted in reverse order, and so the left-to-right minimum changes on
+every element. Interestingly, the worst cases for the number of changes is not
+the worst case for branch mispredictions. If the minimum changes for every
+element, then the comparison branch will always resolve in the same direction,
+and thus be almost perfectly predictable. From the point of view of branch
+prediction, frequent changes (preferably with no simple pattern) are the main
+indicator of poor performance, rather than the absolute number of times that the
+branch resolves in each direction. The worst case for selection sort would
+probably be an array which is partially sorted in reverse order --- the
+left-to-right minimum would change frequently, but not predictably so.
 
@@
 half-exchanges\footnote{A half-exchange is where only one part of the swap is
-completed, with the other value stored in a temporary to be put in place later.}.
-As a result, the instruction count should be high, though it is expected to be
-lower than selection sort.
+completed, with the other value stored in a temporary to be put in place
+later.}. As a result, the instruction count should be high, though it is
+expected to be lower than selection sort.
 
@@
 keys are slowly moved left, so that the unsorted part of the array becomes more
-sorted as bubble sort continues.
+sorted as bubblesort continues.
 
@@
 
-Finally, the branch prediction results are expected to be the same as in bubble
-sort, with one difference. Keys which must travel the entire way across the
-array are taken care of earlier in the sort. If the largest key is in the
+Finally, the branch prediction results are expected to be the same as in
+bubblesort, with one difference. Keys which must travel the entire way across
+the array are taken care of earlier in the sort. If the largest key is in the
 leftmost position, it must be moved the entire way right, which causes misses
 for every move in bubblesort, since each move is disjoint from the next. In
-shakersort, however, each move in the correct direction will follow a move in the
-same direction, each of which should be correctly predicted. This works for a
-lot of keys, which are more likely to move in continuous movements than they are
-in bubblesort, where only small keys on the right move continuously. The
+shakersort, however, each move in the correct direction will follow a move in
+the same direction, each of which should be correctly predicted. This works for
+a lot of keys, which are more likely to move in continuous movements than they
+are in bubblesort, where only small keys on the right move continuously. The
 performance should improve noticeably as a result.
@@
 
-The branch behaviour of bubble sort is particularly interesting when
-compared to selection sort. Although we usually think of the two
-algorithms as being different, they are actually very similar. Like
-selection sort, bubble sort finds the smallest (or largest) key in the
-unsorted part of the array and moves it to start (end) of the array.
-Thus, we would expect that bubble sort's branch prediction behaviour
-would be similar to that of selection sort.
+The branch behaviour of bubblesort is particularly interesting when compared to
+selection sort. Although we usually think of the two algorithms as being
+different, they are actually very similar. Like selection sort, bubblesort finds
+the smallest (or largest) key in the unsorted part of the array and moves it to
+start (end) of the array.  Thus, we would expect that bubblesort's branch
+prediction behaviour would be similar to that of selection sort.
 
-However, in the process of moving the smallest element into position,
-bubble sort also rearranges other array elements. Selection sort
-sweeps across the array and always keeps track of the left-to-right
-minimum up to the current point. However, bubble sort actually moves
-this minimum, exchanging pairwise with each element it encounters
-which is larger. Thus, even the unsorted part of the array becomes
-partially sorted. We might expect that bubble sort's comparison branch
-for a partially sorted array might be more predictable than for a
-random one, but it is not. With a random array, the smaller elements
-are equally likely to appear anywhere in the array. With a partially
-sorted array, the smaller elements tend to appear towards the start of
-the array which is, in bubble sort, the last place that we look. Thus,
-we change the candidate minimum much more often than if we were
-searching a randomly ordered array.
+However, in the process of moving the smallest element into position, bubblesort
+also rearranges other array elements. Selection sort sweeps across the array and
+always keeps track of the left-to-right minimum up to the current point.
+However, bubblesort actually moves this minimum, exchanging pairwise with each
+element it encounters which is larger. Thus, even the unsorted part of the array
+becomes partially sorted. We might expect that bubblesort's comparison branch
+for a partially sorted array might be more predictable than for a random one,
+but it is not. With a random array, the smaller elements are equally likely to
+appear anywhere in the array. With a partially sorted array, the smaller
+elements tend to appear towards the start of the array which is, in bubblesort,
+the last place that we look. Thus, we change the candidate minimum much more
+often than if we were searching a randomly ordered array.
+
+\begin{figure}
+\includegraphics{plots/two_point_counter_bubblesort.eps}
+\caption{Behaviour of branches in bubblesort compared to the ``sweep'' number}
+\label{Predictability of branches in bubblesort compared to sweep number}
+\end{figure}
+
+To demonstrate this effect, we measured how the predictability of bubblesort's
+comparison branch varies with successive sweeps across the array. We added a
+simple software simulation of a bimodal predictor to the source code of base
+bubblesort, and ran it on a large number of randomly generated arrays, each with
+8192 elements.
+
+In Figure \ref{Predictability of branches in bubblesort compared to sweep
+number} we see that on the early iterations the number of correct predictions is
+very close to the number of executed comparison branches. The comparison branch
+is highly predictable because bubblesort's search for the smallest element is
+similar to selection sort. However, with successive sweeps even the unsorted
+part of the array becomes increasingly sorted and the branch becomes more
+unpredictable. We see that the number of executed comparison branches falls
+linearly, but that the number of correctly predicted comparison branches falls
+much more quickly. Eventually, the array becomes so close to fully sorted that
+the branch becomes increasingly predictable in the other direction, and the
+number of correct predictions again approaches the number of executed branches.
+However, during the intermediate stage, a huge number of branch mispredictions
+have already occured. Paradoxically, by partially sorting the array, bubblesort
+greatly increases the number of branch mispredictions and thus greatly increases
+its running time. Bubblesort would be substantially faster if, instead of
+attempting partially sort these intermediate elements, it simply left them
+alone.
+
+A particularly interesting result comes from a comparison between shakersort and
+selection sort. The number of cycles to sort an array (shown in Figure
+\ref{elementary cycles}) is nearly twice as high on shakersort as it is for
+selection sort. Shakersort has a significantly lower instruction count and cache
+miss count than selection sort, but has several orders of magnitude worse branch
+misses. This shows how important it is to consider branch misses in an
+algorithm, especially comparative ones. It also establishes that algorithms such
+as bubblesort and shakersort, will perform considerably worse than algorithms
+which have a correlation between successive branches, as in insertion and
+selection sorts.
 

### 2005-08-27 docs/tech_report/conclusions.tex
started making fixes to conclusions

@@
 instruction in multi-mergesort make it slower than double-aligned tiled
-mergesort, and that multi-mergesort has regular branch predictive access
-patterns during the k-way merge, which can be taken advantage of by a two-level
-adaptive predictor.
+mergesort, and that multi-mergesort has regular access patterns during the k-way
+merge, which can be taken advantage of by a two-level adaptive predictor.
 
@@
 Our results show base quicksort being the fastest of the quicksorts. We note our
-lowest instruction count occurs with the median-of-7 partitioning. We also see
-that using a binary search slows multi-quicksort down compared to a sequential
-search, but that its extra instructions make it slower than memory-tuned
-quicksort.  Despite being slower in our tests, our simulations show tuned
-quicksort to have fewer cache misses, instructions and branch mispredictions
-than base quicksort.
+lowest instruction count occurs with the median-of-three partitioning. We also
+see that using a binary search slows multi-quicksort down compared to a
+sequential search due to branch mispredictions, but that its extra instructions
+make it slower than memory-tuned quicksort. Despite being slower in our tests,
+our simulations show memory-tuned quicksort to have fewer cache misses,
+instructions and branch mispredictions than base quicksort.
 
@@
 \section{Branch Results}
-%TODO this section has to be reread by David
 
@@
 
+\subsection{Elementary Sorts}
 Insertion sort has exactly one miss per key. Using this result we were able to
@@
 complexity.
-%TODO i dont like how this reads
 
@@
 
+\subsection{Heapsort}
 In heapsort, we found a pattern of branch mispredictions when using a fanout in
@@
 
+\subsection{Mergesort}
+
 Using a k-way merge, multi-mergesort had a great number of extra branch
@@
 proportion correctly.
+
+\subsection{Quicksort}
 
@@
 
+\subsection{Radixsort}
+
 As predicted, radixsort is highly predictable, having no comparison
@@
 presented here.
+
+\subsection{Shellsort}
 
@@
 
+\subsection{Heapsort}
+
 LaMarca's heapsort graphs were equivalent to ours in the areas of level 1
@@
 significantly faster heapsort.
+
+\subsection{Mergesort}
 
@@
 
+\subsection{Quicksort}
+
 Our quicksort results, again, we very similar to LaMarca's. We showed an
@@
 
+\subsection{Radixsort}
+
 LaMarca determines that radixsort could not be improved in the same was as the
@@
 mispredictions make it the fastest sort in our survey.
-
-\section{Important Results and Contributions}
-
-%TODO lamarca does an overall comparison. we should do the same
-
-\section{Instruction Count Results}
 

### 2005-08-30 docs/tech_report/shell.tex
fixed shellsorts - numbers and documented errors with h = 5

@@
 framework used. The second was modified to reduce the complexity of the loop
-bounds, and to remove a sentinel check in the inner loop. Code for shellsort and
+bounds, and to remove a bounds check in the inner loop. Code for shellsort and
 improved shellsort is shown in Figures \ref{Shellsort code} and \ref{Improved
 shellsort code}.
+
+A slight difference occurs in the two algorithms. Due to a slight difference
+in behaviour when $h = 5$, the improved shellsort runs faster than shellsort
+when this condition occurs. This is due to shellsort running an extra iteration,
+with $h = 2$, which is skipped by improved shellsort. This occurs when the array
+is sized $16,384$, $32,768$ and $4,194,304$.
 
@@
 The branch performance of shellsort should be predictable. There are
-$log_{(\frac{11}{5})}N$ iterations, each with $N$ branch misses\footnote{See Section
-\ref{insertion is predictable} for a discussion of the properties of insertion
-sort.}. It is expected, therefore, that there will be $log_{(\frac{11}{5})}N$ misses per
-key.
+$log_{(\frac{11}{5})}N$ iterations, each with $N$ branch misses\footnote{See
+Section \ref{insertion is predictable} for a discussion of the properties of
+insertion sort.}. It is expected, therefore, that there will be
+$log_{(\frac{11}{5})}N$ misses per key.
 
@@
 Overall, this leads to a reduction in cycle count of between 35\% and 70\%, as
-seen in Figure \ref{shell cycles}.
+seen in Figure \ref{shell cycles}. While a small amount of this can be
+attributed to the reduction in instruction count due to the removal of the
+bounds check, a large amount cannot. We believe that the rest is due to slight
+differences in the performance of the code generated by the compiler: register
+allocation may be more efficient, for example, or better performing instruction
+scheduling may take place.
 
@@
 
-The results also show that in shellsort, a key moves only 1.03 places down the
-array, on average, during each iteration. In improved shellsort, each key moves
-over 1.07 places each iteration, while in the final insertion sort, each key
-moves an average of 2.1 places.
+In general, the number of taken branches in both sorts would be almost exactly
+the same, with only a minor difference due to the positioning of the sentinel.
+However, during the sorts, the $h = 5$ condition was hit, meaning that there
+were less branches and less taken branches in the improved shellsort, than in
+the shellsort.
+
+The results show that in improved shellsort, each key moves an average 0.915
+places each iteration, while in the final insertion sort, each key moves an
+average of 2.1 places. We can extrapolate a weighted average from this, to
+determine that in shellsort, run under the same conditions, each key would move
+an average of 0.9744.
 
@@
 \section{Future Work}
+
 Only Gonnet's way of producing increments is measured here. Using different
@@
 different results.
+
+We note that improved shellsort performed in cases where $h = 5$ occurred.
+Shellsort should also be investigated with this change made.

### 2005-08-30 docs/tech_report/shell.tex
minor edit on that

@@
 
-A slight difference occurs in the two algorithms. Due to a slight difference
-in behaviour when $h = 5$, the improved shellsort runs faster than shellsort
-when this condition occurs. This is due to shellsort running an extra iteration,
-with $h = 2$, which is skipped by improved shellsort. This occurs when the array
-is sized $16,384$, $32,768$ and $4,194,304$.
+Apart from these changes, the two algorithms algorithms are subtly different.
+Due to a slight difference in behaviour when $h = 5$, the improved shellsort
+runs faster than shellsort when this condition occurs. This is due to shellsort
+running an extra iteration, with $h = 2$, which is skipped by improved
+shellsort. This occurs when the array is sized 16384, 32768 and 4194304.
 
@@
 However, during the sorts, the $h = 5$ condition was hit, meaning that there
-were less branches and less taken branches in the improved shellsort, than in
-the shellsort.
+were less branches and less taken branches in the improved shellsort than in the
+shellsort.
 
@@
 determine that in shellsort, run under the same conditions, each key would move
-an average of 0.9744.
+an average of 0.9744 places.
 

### 2005-08-30 docs/tech_report/heap.tex
fixed heapsort. rewrote a bit of fututre work. changed a H_N to h_N-1

@@
 branches harmonics}). It is possible, therefore, that the 8-heap will have slightly
-fewer mispredictions (about 36\%) due to the combination of the shallower heap
-and the larger fan-out.
+fewer mispredictions due to the combination of the shallower heap and the larger
+fan-out.
 
@@
 prediction ratio gets worse as the fanout gets larger. The reason we predicted
-differently was due to selection sort's $O(H_N)$ mispredictions; however, this
+differently was due to selection sort's $O(H_N-1)$ mispredictions; however, this
 is over the entire length of the array, and it is not accurate to extrapolate
@@
 
-It should be possible to more accurately model and therefore predict the branch
-prediction ratio when the array is short. Our results show us what we expect the
-answer to be, but have not generated a formula to predict it.
+The formula we use to explain selection sort's branch prediction preformance
+breaks down for short arrays, as in our 8-heap. It should be possible to more
+accurately model and therefore predict the branch prediction ratio when the
+array is short. Our results show us what we expect the answer to be, but we have
+not attempted to create a formula to predict it.

### 2005-08-30 docs/tech_report/heap.tex
spell check

@@
 Meanwhile, Figure \ref{memory-tuned heapsort curve} shows us the rate of correct
-predictions, and each point is calculated using a seperate branch predictor. As
-a result, we expect $1-y = \frac{1}{x+1}$, where $x$ and $y$ are the cartesian
+predictions, and each point is calculated using a separate branch predictor. As
+a result, we expect $1-y = \frac{1}{x+1}$, where $x$ and $y$ are the Cartesian
 coordinates of each point on the graph.  $1-y$ is the number of mispredictions
@@
 When the array is short, as in this example, the branch predictor has too much
-feedback to be close to $H_N$. For $N=2$, we would approximate a 33\%
-misprediction ratio using $H_N$, but we measure 39\%. The approximated and
+feedback to be close to $H_N-1$. For $N=2$, we would approximate a 33\%
+misprediction ratio using $H_N-1$, but we measure 39\%. The approximated and
 actual ratios become closer as $N$ increases, as we would expect, and for $N=7$,
@@
 
-The formula we use to explain selection sort's branch prediction preformance
+The formula we use to explain selection sort's branch prediction performance
 breaks down for short arrays, as in our 8-heap. It should be possible to more

### 2005-08-30 docs/tech_report/method.tex
fixed the merge part, and added details to the list of goals

@@
 elementary sorts: insertion sort, selection sort, bubblesort and shakersort.
+\item Implement cache-conscious sorts from \cite{LaMarca96} and
+\cite{LaMarca99}.
 \item Simulate these sorts using the SimpleScalar architecture simulator, using a
 variety of branch predictors and cache configurations.
-\item Implement cache-conscious sorts from \cite{LaMarca96} and
-\cite{LaMarca99}.
+\item Measure the performance of these sorts using the PapiEx performance
+counters.
+\item Simulate bimodal branch predictors in software and use them to measure the
+sorts.
 \item Investigate and analyse the increase in speed, reduction in cache misses, 
-and change in branch prediction accuracy as a result of these changes.
+and change in branch prediction accuracy as a result of the changes made.
 \item Determine if any changes can be made to improved the sorts' performance.
@@
 reduced. Then it fully sorted segments of the array, each half the size of the
-cache, so that temporal locality would be exploited. Finally, all these segments
-were merged together in the remaining merge steps.
+cache, so that temporal locality would be exploited. It also fully sorted
+segements within the level 1 cache. Finally, all these segments were merged
+together in the remaining merge steps.
 
@@
 Instead of a tiled mergesort, a double-aligned tiled mergesort was written.
-This attempted to completely sort in the level 1 cache, then to completely sort
-in the level 2 cache. Double-aligned multi-mergesort was a combination of this
-and of multi-mergesort. The rewrite was based on the base mergesort, and was
-written far more cleanly, being more readable and efficient. These were also
-found to reduce the instruction count, as well as the level 2 cache miss count.
-These results are discussed in more detail in Section \ref{Mergesort results}.
+This aligned the arrays to avoid conflicts in the level 1 cache, at a slight
+cost to level 2 cache performance. Double-aligned multi-mergesort was a
+combination of this and of multi-mergesort. The rewrite was based on the base
+mergesort, and was written far more cleanly, being more readable and efficient.
+These were also found to reduce the instruction count, as well as the level 2
+cache miss count. These results are discussed in more detail in Section
+\ref{Mergesort results}.
 
@@
 
-\section{Testing details}
+\section{Testing Details}
 

### 2005-08-30 docs/tech_report/method.tex
fixed the merge method  more

@@
 
-%TODO this bit also has to be checked once we fix mergesort. if w fix mergesort
 Once these sorts were written it was attempted to improve them slightly.
@@
 cost to level 2 cache performance. Double-aligned multi-mergesort was a
-combination of this and of multi-mergesort. The rewrite was based on the base
-mergesort, and was written far more cleanly, being more readable and efficient.
-These were also found to reduce the instruction count, as well as the level 2
-cache miss count. These results are discussed in more detail in Section
-\ref{Mergesort results}.
+combination of this and of multi-mergesort. We also rewrote base mergesort far
+more cleanly, with the results being more readable and efficient.  These were
+also found to reduce the instruction count, as well as the level 2 cache miss
+count. These results are discussed in more detail in Section \ref{Mergesort
+results}.
 

### 2005-08-30 docs/tech_report/elementary.tex
formatting fixes. -- and "\.  "

@@
 \label{selection branches harmonics}
-The branch prediction performance is not expected to be bad. Flow
-control predictions are very straightforward, and should result in
-very few misses. Comparison predictions, however, are very
-numerous. Each traversal of the array has an average of $N/2$
-comparisons, as we search for the smallest key in the unsorted part of
-the array. During the search, the algorithm compares each key with the
-smallest key seen so far. In a random array, we would expect that the
-candidate for the smallest key (the left-to-right minimum) would
-change frequently towards the start of the array, making the
-comparison branch rather unpredictable. However, as the traversal
-continues, the left-to-right minimum will become smaller, and thus
-more difficult to displace. It will become very unusual to find a
-smaller key, and thus the branch will almost always resolve in the
-same direction. Therefore, the comparison branch will quickly become
-very predictable.
+The branch prediction performance is not expected to be bad. Flow control
+predictions are very straightforward, and should result in very few misses.
+Comparison predictions, however, are very numerous. Each traversal of the array
+has an average of $N/2$ comparisons, as we search for the smallest key in the
+unsorted part of the array. During the search, the algorithm compares each key
+with the smallest key seen so far. In a random array, we would expect that the
+candidate for the smallest key (the left-to-right minimum) would change
+frequently towards the start of the array, making the comparison branch rather
+unpredictable. However, as the traversal continues, the left-to-right minimum
+will become smaller, and thus more difficult to displace. It will become very
+unusual to find a smaller key, and thus the branch will almost always resolve in
+the same direction. Therefore, the comparison branch will quickly become very
+predictable.
 
-Knuth \cite{Knuth97} analyses the number of changes to the
-left-to-right minimum when searching for the minimum element of an
-array. He shows that for $N$ elements, the number of changes is $H_N -
-1$, where $H_N$ is the harmonic series:
+Knuth \cite{Knuth97} analyses the number of changes to the left-to-right minimum
+when searching for the minimum element of an array. He shows that for $N$
+elements, the number of changes is $H_N - 1$, where $H_N$ is the harmonic
+series:
 
@@
 branch resolves in each direction. The worst case for selection sort would
-probably be an array which is partially sorted in reverse order --- the
+probably be an array which is partially sorted in reverse order - the
 left-to-right minimum would change frequently, but not predictably so.

### 2005-08-30 docs/tech_report/merge.tex
formatting fixes. -- and "\.  "

@@
 \ref{insertion is predictable}, it is shown that insertion sort is the fastest
-elementary sort.  Sedgewick's quicksort uses this, and LaMarca recommends the
-use of a simple inline pre-sort, which can be made from insertion sort. To make
-lists of length four, the smallest of the four keys is found, and exchanged
-with the first key, in order to act as a sentinel. The next three keys are
-sorted with insertion sort. The next set of four keys is sorted in reverse
-using the same method. It is possible to hard code this behaviour, in an attempt
-to reduce the instruction count, but in fact the instruction count stays the
-same, and the branch prediction suffers as a result.
+elementary sort. Sedgewick's quicksort uses this, and LaMarca recommends the use
+of a simple inline pre-sort, which can be made from insertion sort. To make
+lists of length four, the smallest of the four keys is found, and exchanged with
+the first key, in order to act as a sentinel. The next three keys are sorted
+with insertion sort. The next set of four keys is sorted in reverse using the
+same method. It is possible to hard code this behaviour, in an attempt to reduce
+the instruction count, but in fact the instruction count stays the same, and the
+branch prediction suffers as a result.
 
@@
 cache block, and the constant access of one followed by the other causes every
-access to be a miss.}.  To avoid this, an entire cache line is added at a time
-to the priority queue. There will still be conflict misses, but there will be no
+access to be a miss.}. To avoid this, an entire cache line is added at a time to
+the priority queue. There will still be conflict misses, but there will be no
 thrashing since the keys overwritten in the cache will not be needed again.
@@
 the array. No more details are provided on this, and a sorted list of length $k$
-is used here, in its place.  If the key at the top of the heap is the same as
-the smallest value in the list, another eight keys are added from the array
-segment to which the number belonged. This continues until the array is full
-sorted.
+is used here, in its place. If the key at the top of the heap is the same as the
+smallest value in the list, another eight keys are added from the array segment
+to which the number belonged. This continues until the array is full sorted.
 
@@
 choose the smaller of the elements from the start of the two sub-arrays being
-merged. The direction of this branch is completely random -- it depends entirely
+merged. The direction of this branch is completely random - it depends entirely
 on the (in this case random) data. Figure \ref{mergesorts-1} also shows that
@@
 eliminated some of the control-flow branches. The main comparison branch is
-split into two separate branches --- one iterating from left-to-right over the
+split into two separate branches - one iterating from left-to-right over the
 increasing sub-array (`\n{Left}'), and the other from right-to-left over the

### 2005-08-30 docs/tech_report/quick.tex
removed a TODO
formatting fixes. -- and "\.  "

@@
 
-%TODO
-% \textbf{[David] I'm not sure what to do with this section. Part of the
-% problem is that the pictures are very difficult to understand. Another
-% problem is that we are trying to show differences that are almost
-% invisible. The median of 9 picture should definitely be removed from
-% this subsection. However, I suspect that the whole of this subsection should
-% go. What do we want to say? One really puzzling result is that there
-% seem to be no median branches at all for cache quicksort. This looks
-% strikingly like a programming error.}
-
-% \textbf{[David] On thing that's badly needed in this section is
-% a detailed description of what these funny charts all mean. However,
-% the appropriate place for most of this is the place where the first of
-% these charts appeared.}
-
 \sixplot{base_quicksort__median_of_3__0.eps}{memory_tuned_quicksort_0.eps}{multi_quicksort__binary_search__0.eps}{multi_quicksort__sequential_search__0.eps}{}{}{base,
@@
 decreases. The main reason for this is the reduction in branches by not
-attempting to move the pivot elements. Attempting to move these elements
-usually results in a branch misprediction. The main reason is that most other
-elements move at least one position, and thus cause the insertion sort loop to
-iterate at least once. But the pivot elements never move -- they are already in
-the correct place.
+attempting to move the pivot elements. Attempting to move these elements usually
+results in a branch misprediction. The main reason is that most other elements
+move at least one position, and thus cause the insertion sort loop to iterate at
+least once. But the pivot elements never move - they are already in the correct
+place.
 
@@
 number of mispredictions. Interestingly, the percentage misprediction rate of
-the \n{i} and \n{j} branches does not change -- it is 37\% in both cases.
+the \n{i} and \n{j} branches does not change - it is 37\% in both cases.
 

### 2005-08-30 docs/tech_report/conclusions.tex
formatting fixes. -- and "\.  "
spells checks

@@
 the pivots, as Sedgewick does, actually increases both the number of branches
-and the rate of mispredictions. We conclude that LaMarca's improvment, which
+and the rate of mispredictions. We conclude that LaMarca's improvement, which
 slightly improves quicksorts cache performance, also slightly improves it's
@@
 
-We also analysed the effects of different medians on quicksort. We deterine that
+We also analysed the effects of different medians on quicksort. We determine that
 not using a median to choose a pivot makes the branches more predictable than
 they are biased, though at a cost of a large increase in instructions (and also
-significantly greater possability of a worst-case $O(N^2)$ sort). We also saw
+significantly greater possibility of a worst-case $O(N^2)$ sort). We also saw
 that the more keys considered when determining the median, the lower the number
@@
 shellsort, using our results from insertion sort. We also found that the
-improvement mode to shellsort, seperating the final loop from the rest, results
+improvement mode to shellsort, separating the final loop from the rest, results
 in keys being moved 1.07 places, up from 1.03, contributing to its improved
@@
 not using Sedgewick's insertion sort improvement reduces cache misses. We found
-that multi-quicksort's k-way sort changes the slope of the graph, but our
-result were not conclusive as to whether this would eventually make
-multi-quicksort faster than the other versions. Our results do show it being
-slower for all our test cases, due to the increase in instruction count.
+that multi-quicksort's k-way sort changes the slope of the graph, but our result
+were not conclusive as to whether this would eventually make multi-quicksort
+faster than the other versions. Our results do show it being slower for all our
+test cases, due to the increase in instruction count.
 

### 2005-08-30 docs/tech_report/elementary.tex
spells checks

@@
 The level 1 cache should have bad performance as a result. Since the level 1
-cache is smaller that the arrays tested, it should perform analagiously to a
+cache is smaller that the arrays tested, it should perform analogously to a
 level 2 cache being tested with a much larger array.
@@
 However, during the intermediate stage, a huge number of branch mispredictions
-have already occured. Paradoxically, by partially sorting the array, bubblesort
+have already occurred. Paradoxically, by partially sorting the array, bubblesort
 greatly increases the number of branch mispredictions and thus greatly increases

### 2005-08-30 docs/tech_report/merge.tex
spells checks

@@
 the arrays fully in the cache, leaving them bitonically arranged (sorted into
-sub-arrays, sorted into alternatey increasing and decreasing order). The final
+sub-arrays, sorted into alternately increasing and decreasing order). The final
 merge is performed separately, and hence no bias for the forward branches, as

### 2005-08-30 docs/tech_report/quick.tex
spells checks

@@
 computing time is spent sorting (see \cite{TODO}), the development of caches has
-been subconsicously affected by the cache performance of quicksort.}, as it has
+been subconsciously affected by the cache performance of quicksort.}, as it has
 excellent spatial and temporal locality. Once it no longer fits in the cache,
@@
 Figure \ref{quick-predictors2} we see that this results in an additional branch
-to handle the edgecase without a sentinel. Although this branch is almost 100\%
+to handle the edge-case without a sentinel. Although this branch is almost 100\%
 predictable, it is executed a very large number of times.}. The arrays we sort

### 2005-08-30 docs/tech_report/quick.tex
filled in 25% reference

@@
 LaMarca notes that quicksort is already very cache efficient\footnote{We
-speculate wildly at this point and suggest that since an average of 25\% of
-computing time is spent sorting (see \cite{TODO}), the development of caches has
-been subconsciously affected by the cache performance of quicksort.}, as it has
-excellent spatial and temporal locality. Once it no longer fits in the cache,
-however, the number of misses should increase significantly. This is especially
-true in the case of the base quicksort, which does an insertion sort across then
-entire array at the end.
+speculate wildly at this point and suggest that since, historically, an average
+of 25\% of computing time has been spent sorting (see \cite[Page 3]{Knuth97}),
+the development of caches has been subconsciously affected by the cache
+performance of quicksort.}, as it has excellent spatial and temporal locality.
+Once it no longer fits in the cache, however, the number of misses should
+increase significantly. This is especially true in the case of the base
+quicksort, which does an insertion sort across then entire array at the end.
 

### 2005-08-30 docs/tech_report/conclusions.tex
tidied up

@@
 We found that insertion sort has the best performance of any of the elementary
-sorts, with fewer instructions, cache misses and exactly one branch
-misprediction per key. Selection sort also has very few branch mispredictions,
-and our results match Knuth's calculations. We find that reducing the number of
-runs in bubblesort and shakersort at the cost of additional instructions is
-counter productive, and that their branch prediction performance is incredibly
-poor.
+sorts, with fewer instructions and cache misses than any other elementary sort,
+and exactly one branch misprediction per key. Selection sort also has very few
+branch mispredictions, and our results match figures created using Knuth's
+harmonics formula. We found that bubblesort and shakersort have poor branch
+prediction performance, and that this cripples their performance, especially
+compared to selection sort which is otherwise similar in behaviour. We
+determined that bubblesort has better branch performance at the start and end of
+its sorting.
+
+We also found that reducing the number of runs performed by shakersort and
+bubblesort is counter productive due to additional book-keeping cost. We found
+also that of the elementary sorts, only insertion sort has good cache
+performance, as the others do not exploit any temporal locality.
 
@@
 The cache optimized heaps perform better than the non-optimized versions. At the
-cost of a slightly increased instruction count, the number of branch misses can
+cost of a slightly increased instruction count, the number of cache misses can
 be greatly reduced, and the overall performance of our 8-heap is significantly
@@
 
-Additionally, we created an improvement to heapsort which removed the need for a
-sentinel, without introducing any more instructions.
+Additionally, we created an improvement to heapsort using a 2-heap, which
+removed the need for a sentinel, without introducing any more instructions. We
+found that a $k$-ary heap has similar branch prediction performance to selection
+sort, though not exactly, in that the fanout becomes more predictable the larger
+it is.
 
@@
 also that the level 1 cache needs to be taken into account, that the extra
-instruction in multi-mergesort make it slower than double-aligned tiled
-mergesort, and that multi-mergesort has regular access patterns during the k-way
-merge, which can be taken advantage of by a two-level adaptive predictor.
+instruction in multi-mergesort make it slower than tiled mergesort, and that
+multi-mergesort has regular access patterns during the k-way merge, which can be
+taken advantage of by a two-level adaptive predictor.
 
@@
 
+Among them is the fact that bimodal branch predictors have performed better than
+2-level adpative ones in almost every case in this research. This holds across
+all comparative branches in all the sorts we surveyed, and we believe that where
+explicit patterns - which can be exploited by 2-level adaptive predictors - do
+not exist, a bimodal predictors perform better.
+
 \subsection{Elementary Sorts}
+
 Insertion sort has exactly one miss per key. Using this result we were able to
@@
 Selection sort also has a very low number of branch mispredictions, the results
-of which match up well to Knuth's calculation. We found as well that bubblesort
-and shakersort have incredibly bad branch prediction performance. Both are
-$O(N^2)$ sorts, and the number of branches and branch misses have the same
-complexity.
+of which can be calculated using Knuth's formula for $H_N-1$. We found as well
+that bubblesort and shakersort have incredibly bad branch prediction
+performance. Both are $O(N^2)$ sorts, and the number of branches and branch
+misses have the same complexity.
 
-TODO bubblesort
+Bubblesort performs similarly to selection sort. It differs in that it partially
+sorts the array as it moves the smallest key down the array, which selection sort
+does not do. This has a very large negative effect for branch prediction
+performance, which is not suffered by selection sort. This occurs in the middle
+of the sort; at the start, it performs like selection sort, being randomly
+arranged; at the end, it is largely sorted and is predictable again.
 
 \subsection{Heapsort}
+
 In heapsort, we found a pattern of branch mispredictions when using a fanout in
 our 8-heap, and that this pattern matched closely that of selection sort,
-discussed above. We found that the approximated results using $H_N$ was similar to
-our expected results for large arrays, and followed a similar pattern for small
-arrays.
+discussed above. We found that though our approximated results using $H_N-1$
+were not identical to our actual results - due to the arrays being very small -
+our actual results did become more predictable in a similar fashion.
 
@@
 branch mispredictions due to the addition of a binary search. We determine that
-this while reducing the instruction count, the increase is actual running time
-due to branch mispredictions is significant.
+while this reduces the instruction count and number of memory accesses, the
+increase is actual running time due to branch mispredictions is significant.
 
@@
 improvement mode to shellsort, separating the final loop from the rest, results
-in keys being moved 1.07 places, up from 1.03, contributing to its improved
-performance.
+in a significantly faster loop, with an extra speed boost from a side-effect
+skipping an iteration with $h = 2$.
 
@@
 
-Over the course of this research. The vast majority of our results have
+Over the course of this research, the vast majority of our results have
 correlated exactly with LaMarca's. Many of our graphs are exactly the same
 shape to LaMarca's, can be calculated exactly or almost exactly using his
-methods, or can be seen to be equivalent when scaled by an appropriate ratio
-(due to different test parameters).
+methods, or can be seen to be equivalent when scaled by an appropriate ratio.
 
@@
 However, we disagree with his results in several key places. Firstly, we note
-that an in-lined sorting method does not perform as fast as insertion sort. We
+that an inlined sorting method does not perform as fast as insertion sort. We
 find it difficult to perform his improvements on either algorithm N or algorithm
@@
 
-\section{Cache Results}
-
-% TODO more details in this
-The test results in this report mostly agreed with LaMarca's results. The
-heapsort graphs had nearly exactly the same shape as LaMarca's, and the ratios
-between the base and cache-conscious versions were nearly identical. The
-multi-mergesort results were slightly different: instruction count was
-identical, but while the cache results of the base and tiled mergesorts were
-identical, multi-mergesort results were not the same. However, the basis for
-those results - a sharp increase in cache misses followed by a flat-lined graph
-- were in both cases.
-
-The quicksort results were also very similar, though the multi-quicksort results
-had a very different shape to LaMarca's. However, the instruction count
-results for each quicksort agreed with LaMarca's results, as did cache miss
-results of the base and cache-conscious versions.
-
-The two sets of radixsort results were slightly similar. The rise and plateau of
-the curve to the right of the graphs was the same in both cases, though the left
-hand side of the graphs did not agree.
-
-Another interesting observation can be made: the results of the simulations on
-the Pentium 4 were not always influenced by the cache misses. Though heapsort
-and radixsort began to perform poorly once the data set no longer fit in the
-cache, neither quicksort nor mergesort were affected. The reason why radixsort
-was affected is obvious - there is no temporal reuse; in heapsort, the full
-cache line is not being used, and spatial locality suffers as a result. It is
-interesting that the other two sorts were unaffected, even though their cache
-miss graphs indicate a significant increase in misses.
-
-Finally, it was shown that in some cases a direct-mapped cache performed better
-than a fully-associative one. The reasons for this were not clear.
-
 \section{Branch Prediction Results}
-
-Insertion sort has only one branch misprediction per key. This shows that it is
-possible to predict comparative branches correctly, though the instruction cost
-of such an algorithm may be high. This is also shown in multi-quicksort: though
-the instruction count and memory reference count are both doubled, it is still
-faster to use a sequential search than a binary search across small lists.
-
-Selection sort has a similar number of instructions and cache misses as
-bubblesort and a larger instruction and cache miss count than shakersort, yet it
-still considerably outperforms both of these. This shows the high cost of branch
-mispredictions on an algorithm.
-
-Radixsort has almost no branch mispredictions at all. The only mispredictions
-are due to its flow control, but it escapes comparative mispredictions, and this
-factor allows it to perform as fast as, and in some cases faster than quicksort,
-despite its high number of cache misses.
-
-The heapsort results show the disadvantage of using unsorted lists. Even though
-there are less branches in the 4-heap version, the number of branch
-mispredictions are higher than in a standard 2-heap.
+%TODO mergre this section
 
@@
 \section{Best Sort}
+
 The results also helped answer the question of which is the best sort. However,
@@
 This section lists the contributions of this report. Each of these is novel  
-and was discovered or developed during the course of this report.
-%TODO cant say that
+and was discovered or developed during the course of this research.
+
 The primary contribution of this report is that it repeats the work of LaMarca

### 2005-08-30 docs/tech_report/heap.tex
formatting fixes. -- and "\.  "
tidied up

@@
 the start of a cache block, then the sibling of the last child in the cache
-block will be in the next cache block.  For a cache block fitting four keys,
-this means that half the children will cross cache block boundaries. This can be
+block will be in the next cache block. For a cache block fitting four keys, this
+means that half the children will cross cache block boundaries. This can be
 avoided by padding the start of the heap. Since the sort is out of place, an
-extra cache block can be allocated, and the root of the heap is placed at the end of
-the first block. Now, all siblings will reside in the same block.
+extra cache block can be allocated, and the root of the heap is placed at the
+end of the first block. Now, all siblings will reside in the same block.
 
@@
 fewer mispredictions due to the combination of the shallower heap and the larger
-fan-out.
+fanout.
 
@@
 shown for the two-level predictor. The smallest table size is shown for the
-bimodal predictor, as all the bimodal table sizes show identical results.  In
-the case of the 2- and 8-heaps, the bimodal predictor works better than the
+bimodal predictor, as all the bimodal table sizes show identical results. In the
+case of the 2- and 8-heaps, the bimodal predictor works better than the
 two-level adaptive predictor. The warm-up time of the table-driven predictors is

### 2005-08-30 docs/tech_report/merge.tex
tidied up

@@
 merge is performed separately, and hence no bias for the forward branches, as
-there was in tiled mergesort. The final sort is performed with a $k$-way merge,
+there was in tiled mergesort. The final sort is performed with a k-way merge,
 which is implemented using a heap data structure. This is shown in Figures
@@
 number of branch mispredictions as in the other mergesort algorithms. Although
-the final $k$-way merge executes a lot more comparison branches, these heap
+the final k-way merge executes a lot more comparison branches, these heap
 branches are quite predictable, as we saw in Section

### 2005-08-30 docs/tech_report/conclusions.tex
lots of minor changes

@@
 while this reduces the instruction count and number of memory accesses, the
-increase is actual running time due to branch mispredictions is significant.
+increase in actual running time due to branch mispredictions is significant.
 
@@
 fills the cache fully with a smaller number of keys than base heapsort, which is
-in place. LaMarca's results do not show this. Nor do they show a difference in
+in-place. LaMarca's results do not show this. Nor do they show a difference in
 cache misses between these two sorts when the array does fit in the cache.
@@
 S, and write our own mergesort instead. We expected that mergesort would have
-half the instruction count of heapsort we found instead they were very similar.
+half the instruction count of heapsort; we found instead they were very similar.
 
-LaMarca did not include details of his tagging system in multi-mergesort. As a
-result, we used a search, which greatly increased the instruction count. As a
+LaMarca did not include details of his tagging system in multi-mergesort.
+Instead we used a search, which greatly increased the instruction count. As a
 result, multi-mergesort did not perform as well as the mergesorts which had only
@@
 LaMarca predicts two cache misses per key, which translated to one miss per key
-with our 32 bit integers. We found this calculation to be true, as is his
+with our 32-bit integers. We found this calculation to be true, as is his
 calculation of a decrease of one cache miss in four: we note a reduction of one
@@
 
-Where LaMarca found that using Sedgewick's insertion sort optimisation decreased
+Where LaMarca found that using Sedgewick's insertion sort optimization decreased
 instruction count, we found the opposite. We also found that despite
@@
 double-aligned tiled mergesort. The best in-place sort is quicksort. The best in-place
-stable sort is, sadly, insertion sort. The best elementary sort is also
-insertion sort; as such it should be only used for very small data sets. The
-best sort for very large data sets is radixsort. The fastest overall sort was
-memory-tuned radixsort, though quicksort is very close.
-% TODO redo above paragraph
+stable sort we examine in this report is insertion sort. The best elementary
+sort is also insertion sort; as such it should be only used for very small data
+sets. The best sort for very large data sets is radixsort. The fastest overall
+sort was memory-tuned radixsort, though quicksort is very close.
 
@@
 
-The primary contribution of this report is that it repeats the work of LaMarca
+The first main contribution of this report is that it repeats the work of LaMarca
 on cache-conscious sorting, and validates many of his claims, while questioning
-others. The next most important contribution is having performed a branch
-prediction analysis of major sorts - which to my knowledge has not been done
-before - and developing a framework by which these results can be repeated and
-expanded upon.
+others. The other main contribution is having performed a branch prediction
+analysis of major sorts - which to our knowledge has not been done before - and
+developing a framework by which these results can be repeated and expanded upon.
 
@@
 Several discoveries were made and analysed: insertion sort has one misprediction
-per key; selection sort has $log_2(N/2)$ branch mispredictions per key - an
+per key; selection sort has $H_N-1$ branch mispredictions per key - an
 explanation of this was also provided; tiling mergesort can be done in two
@@
 % - double-tiling
-
-%TODO look over these
-In addition, several ideas were conceived although not implemented, which relate
-to sorting algorithms: it should be possible to reduce the conflict misses of
-tiled mergesort without padding, by slightly reducing the size of the array
-segments sorted in the first sorting phase; it may be possible to reduce the
-cost of an insertion sort or selection sort by considering more than one key at
-a time; it may be possible to reduce the number of partitions - and hence the
-increase in instruction count - of multi-quicksort by choosing from a larger
-selection of pivots. Finally, it should be possible to reduce by half the number
-of cache misses in radixsort, by counting earlier and by copying back and forth
-between two arrays.

### 2005-08-30 docs/tech_report/method.tex
spell chekc
formatting fixes. -- and "\.  "
spells checks
lots of minor changes

@@
 actually increased slightly as a result of this, indicating that algorithm N
-does a better job of pre-sorting the array than the in-lined pre-sort did.
+does a better job of pre-sorting the array than the inlined pre-sort did.
 
@@
 cache, so that temporal locality would be exploited. It also fully sorted
-segements within the level 1 cache. Finally, all these segments were merged
+segments within the level 1 cache. Finally, all these segments were merged
 together in the remaining merge steps.
@@
 the number of cache misses in the final merge steps of tiled mergesort by
-merging all the sorted arrays in one step. This required the use of an k-ary
+merging all the sorted arrays in one step. This required the use of an $k$-ary
 heap, which had been developed as part of heapsort.

### 2005-08-30 docs/tech_report/quick.tex
lots of minor changes

@@
 block of a few thousand keys. LaMarca notes very little performance difference
-between 100 and 5,000 keys in each block. The size is therefore chosen so that
+between 100 and 5000 keys in each block. The size is therefore chosen so that
 each list fits exactly into a page of memory. When a list is filled, another
@@
 
-A 2MB cache was assumed, so the number of 32 bit integer keys which fit in the
-cache for this in-place sort is 524,288. Multi-quicksort began once the number
-of keys was larger than 174,762.
+A 2MB cache was assumed, so the number of 32-bit integer keys which fit in the
+cache for this in-place sort is 524288. Multi-quicksort began once the number
+of keys was larger than 174762.
 

### 2005-08-30 docs/tech_report/radix.tex
formatting fixes. -- and "\.  "
spells checks
lots of minor changes

@@
 The steps to sort by radix are simple. First a radix is chosen, based on the
-size of the part to be sorted. For 32 bit integers, a byte is used. This means
+size of the part to be sorted. For 32-bit integers, a byte is used. This means
 the radix is 256, as there are 256 possible byte values. An array of 256
@@
 The number of iterations across the array is constant, however, proportional to
-$(bitsPerInt/log_2(Radix))$. For a 32 bit integer and a radix of 256, this is
+$(bitsPerInt/log_2(Radix))$. For a 32-bit integer and a radix of 256, this is
 four steps and each step has four passes over the array, for a total of 16
-passes over the array.  With eight keys per cache line, there should be about
-two misses per key, when the array doesn't fit into the array, though this
-depends on the placing of the auxiliary array.
+passes over the array. With eight keys per cache line, there should be about two
+misses per key, when the array doesn't fit into the array, though this depends
+on the placing of the auxiliary array.
 
@@
 miss per key. The number of branches itself is quite low, as several of the
-sorts considered had more thna that number of mispredictions. However, it is not
+sorts considered had more than that number of mispredictions. However, it is not
 unexpected. The number of branches reduces when copying back is removed, and
@@
 However, LaMarca's conclusion - that radixsort is not as fast as quicksort - can
-be easily explained from the base radixsort results. Once an radixsort optimised
+be easily explained from the base radixsort results. Once an radixsort optimized
 for the cache is presented, these no longer hold, and the lack of branch

### 2005-08-30 docs/tech_report/conclusions.tex
so close to being done i can taste it

@@
 Among them is the fact that bimodal branch predictors have performed better than
-2-level adpative ones in almost every case in this research. This holds across
+2-level adaptive ones in almost every case in this research. This holds across
 all comparative branches in all the sorts we surveyed, and we believe that where
@@
 
-\section{Branch Prediction Results}
-%TODO mergre this section
-
-It is also shown that bimodal branch predictors generally out-perform two-level
-adaptive predictors when resolving comparative misses. This is shown by the
-results of insertion sort, bubblesort, shakersort, heapsort and quicksort. In
-the cases where the two-level adaptive predictor does perform as well, the table
-required is much larger than that required by the bimodal predictors to get the
-same results.
-
 \section{Best Sort}
@@
 
-%TODO claims that it backed
-% TODO claims that it refuted or found problems with
-% - cache in heapsort
-% - algorithm N, can those improvements really be made
+Several discoveries were made and analysed: insertion sort has one misprediction
+per key; selection sort has $H_N-1$ branch mispredictions per key, which applies
+to long arrays, though short arrays become more predictable the longer they
+get, as in heapsort; tiling mergesort can be done in two steps, reducing the
+cache miss count; aligning mergesort for a level 2 cache alone is
+counter-productive - aligning it for level 1 and level 2 is significantly
+faster; it is possible to avoid using a sentinel and still remove the bounds
+check from a standard 2-heap, simply by reordering the steps involved;
+sequential searches are far cheaper than binary searches across small lists, due
+to a low branch misprediction rate. Bubblesort has terrible branch performance
+in the middle, due to being partially sorted - it performs much worse than
+selection sort as a result; multi-mergesort contains regular access patterns
+which can be exploited by a 2-level adaptive branch predictor; in almost all
+other cases, a bimodal predictor out-performs a 2-level adaptive predictor. A
+median-of-3 quicksort outperforms other medians; radix sort has no branch
+mispredictions; shellsort has easily calculable branch mispredictions, and
+performs very similarly to an $O(NlogN)$ so.
 
-Several discoveries were made and analysed: insertion sort has one misprediction
-per key; selection sort has $H_N-1$ branch mispredictions per key - an
-explanation of this was also provided; tiling mergesort can be done in two
-steps, reducing the cache miss count; it is possible to avoid using a
-sentinel and still remove the bounds check from a standard 2-heap, simply by
-reordering the steps involved; sequential searches are far cheaper than binary
-searches across small lists, due to a low branch misprediction rate.
-
-%TODO add novel contributions
-% - avoid copying back in mergesort
-% - remove bounds check without sentinel in base heapsort
-% - double-tiling
+In addition, we provide an analysis of most major sorts currently used, a
+comparison of $O(NlogN)$ sorts, of different medians for quicksort, of $O(N^2)$
+sorts and of cache conscious sorts. We present branch prediction performance
+figures for most major sorts, and adapt cache-conscious strategies from
+mergesort to make an improved radixsort the fastest performing sort in our
+tests.

### 2005-08-31 docs/tech_report/conclusions.tex
david's final changes

@@
 Among them is the fact that bimodal branch predictors have performed better than
-2-level adaptive ones in almost every case in this research. This holds across
+two-level adaptive ones in almost every case in this research. This holds across
 all comparative branches in all the sorts we surveyed, and we believe that where
-explicit patterns - which can be exploited by 2-level adaptive predictors - do
+explicit patterns - which can be exploited by two-level adaptive predictors - do
 not exist, a bimodal predictors perform better.
@@
 predictions. While many of these were mispredicted using the bimodal predictor,
-the two-level adaptive predictor was able to predict a higher
-proportion correctly.
+the two-level adaptive predictor was able to predict a higher proportion
+correctly.
 
@@
 cache misses between these two sorts when the array does fit in the cache.
-LaMarca's base heapsort have the characteristics of an out-of-place heapsort.
+LaMarca's base heapsort has the characteristics of an out-of-place heapsort.
 However, our results show that his implicit heap improvements have led to a
@@
 Several discoveries were made and analysed: insertion sort has one misprediction
-per key; selection sort has $H_N-1$ branch mispredictions per key, which applies
-to long arrays, though short arrays become more predictable the longer they
-get, as in heapsort; tiling mergesort can be done in two steps, reducing the
-cache miss count; aligning mergesort for a level 2 cache alone is
+per key; selection sort has $O(H_N-1)$ branch mispredictions per key, which
+applies to long arrays, though short arrays become more predictable the longer
+they get, as in heapsort; tiling mergesort can be done in two steps, reducing
+the cache miss count; aligning mergesort for a level 2 cache alone is
 counter-productive - aligning it for level 1 and level 2 is significantly
@@
 selection sort as a result; multi-mergesort contains regular access patterns
-which can be exploited by a 2-level adaptive branch predictor; in almost all
-other cases, a bimodal predictor out-performs a 2-level adaptive predictor. A
+which can be exploited by a two-level adaptive branch predictor; in almost all
+other cases, two-level branch predictors are measurably less accurate than
+simple bimodal predictors at predicting branches in sorting algorithms. A
 median-of-3 quicksort outperforms other medians; radix sort has no branch
@@
 mergesort to make an improved radixsort the fastest performing sort in our
-tests.
+tests, and to make mergesort itself significantly faster.

### 2005-08-31 docs/tech_report/elementary.tex
lots of minor changes
david's final changes

@@
 Within practical sorting ranges, $H_N$ grows at a rate similar to
-$log(N)$\footnote{This seems like an error, since $H_N$ is very similar to
-$log_eN$. It is not. We mean that $H_N$ is very similar to $log_2N$.}. Thus, in
-the average case, we can expect that the comparison branch will mostly go in the
-same direction, making the branch highly predictable.
+$log(N)$. Thus, in the average case, we can expect that the comparison branch
+will mostly go in the same direction, making the branch highly predictable.
 
@@
 plot them on a separate chart. Nearly one in four comparisons are misses. As the
-data sets gets large, there are nearly 10,000 branch misses involved in sorting
+data sets gets large, there are nearly 10000 branch misses involved in sorting
 a single key. The improved bubblesort had exactly the same number of misses and

### 2005-08-31 docs/tech_report/heap.tex
david's final changes

@@
 
-The relationship between this number and $H_N-1$ is relatively straightforward.
-$H_N-1$ represents the cumulative number of changes in the minimum. Over a long
-array, this approximates the number of branch mispredictions, as almost every
-branch will not be taken, and the predictor will always predict not-taken.
+The relationship between these numbers and $H_N-1$ is relatively straighforward.
+$H_N-1$ represents the cumulative number of changes in the minimum. There first
+element is guaranteed to be the minimum. The second element has a 50\% chance of
+being smaller than the first. If we examine a third element, then the chances of
+that element being smaller than two other elements is one in three. Therefore,
+this branch will be taken one third of the time. As the number of elements we
+examine increases, the probability of the nth element being smaller than all
+elements seen so far is $\frac{1}{n}$.
 
-Meanwhile, Figure \ref{memory-tuned heapsort curve} shows us the rate of correct
-predictions, and each point is calculated using a separate branch predictor. As
-a result, we expect $1-y = \frac{1}{x+1}$, where $x$ and $y$ are the Cartesian
-coordinates of each point on the graph. $1-y$ is the number of mispredictions
-from the graph, while $\frac{1}{x+1}$ is the increase in $H_N$ for step $N$.
+Meanwhile, Figure \ref{memory-tuned heapsort curve} shows the rate of correct
+predictions. The loop has been fully unrolled, so there is a separate branch
+(with its own predictor) for each of the seven comparisons.  As a result, we
+expect $1-y = \frac{1}{x+1}$, where $x$ and $y$ are the Cartesian coordinates of
+each point on the graph. $1-y$ is the number of mispredictions from the graph,
+while $\frac{1}{x+1}$ is the increase in $H_N$ for step $N$.
+
+An interesting question is whether fully unrolling this loop, and thus
+separating the comparison branch into seven separate branches, has an impact on
+branch prediction. Essentially, there are two effects at work if we replace the
+sequence of branches with a single branch. First, the single branch predictor
+will spend most of its time in the state where it predicts not-taken, rather
+than taken. This is because all branches apart from the first are biased in the
+taken direction. This should improve accuracy. On the other hand, on those
+occasions when the bi-modal branch predictor is pushed towards predicting taken,
+the next branch is highly likely to be mispredicted. This is especially true
+because the first two or three branches are most likely to push the predictor in
+the taken direction, whereas the later branches are those that are mostly likely
+to result in a misprediction in this state (because they are almost always
+not-taken).
+
+We investigated the trade-offs in these two effects and we discovered that using
+a single branch does indeed cause the branches for the ealier comparisons to
+become more predictable. However, this improvement is outweighed by a loss in
+predictability of the later comparisons. The overall prediction accuracy fell by
+around 2\% when we used a single branch, rather than unrolling the loop. This
+result is interesting because it shows that relatively small changes in the way
+that an algorithm is implemented can have an effect on branch prediction
+accuracy. In this case, unrolling does not just remove loop overhead. It also
+improves the predictability of the branches within the loop.
 

### 2005-08-31 docs/tech_report/conclusions.tex
davids fixes to heap sort due to incorrect taken vs not-taken. also, first corrections to conclusions

@@
 \label{conclusions}
-This section presents a summary of the results from previous chapters, compares
-our results with LaMarca's, lists important results and discoveries, and
-suggests contributions made to computer science by this research.
+This section presents a summary of the results from previous chapters, lists
+important branch prediction results and discoveries, compares our results with
+LaMarca's and discusses important cache results, discusses the best sort that we
+found in a number of catagories, and suggests contributions made to computer
+science by this research.
 
@@
 determined that bubblesort has better branch performance at the start and end of
-its sorting.
+its sorting, then in the middle.
 
 We also found that reducing the number of runs performed by shakersort and
-bubblesort is counter productive due to additional book-keeping cost. We found
-also that of the elementary sorts, only insertion sort has good cache
-performance, as the others do not exploit any temporal locality.
+bubblesort is counter-productive due to additional book-keeping cost. We found
+that of the elementary sorts, only insertion sort has good cache performance, as
+the others do not exploit any temporal locality.
 
@@
 removed the need for a sentinel, without introducing any more instructions. We
-found that a $k$-ary heap has similar branch prediction performance to selection
-sort, though not exactly, in that the fanout becomes more predictable the larger
-it is.
+found that a $k$-ary heap has similar, though not the same, branch prediction
+performance as selection sort, in that the fanout becomes more predictable the
+larger it is.
 
@@
 We found that algorithm S has a lower instruction count than algorithm N, but
-that this does not necessarily translate to better performance. Our results show
-also that the level 1 cache needs to be taken into account, that the extra
-instruction in multi-mergesort make it slower than tiled mergesort, and that
-multi-mergesort has regular access patterns during the k-way merge, which can be
-taken advantage of by a two-level adaptive predictor.
+that this does not translate to better performance. Our results show also that
+the level 1 cache needs to be taken into account, both in tiling and in
+alignment; that the extra instructions in multi-mergesort make it slower than
+tiled mergesort, and that multi-mergesort has regular access patterns during the
+k-way merge, which can be taken advantage of by a two-level adaptive predictor.
 
@@
 Our results show base quicksort being the fastest of the quicksorts. We note our
-lowest instruction count occurs with the median-of-three partitioning. We also
-see that using a binary search slows multi-quicksort down compared to a
-sequential search due to branch mispredictions, but that its extra instructions
-make it slower than memory-tuned quicksort. Despite being slower in our tests,
-our simulations show memory-tuned quicksort to have fewer cache misses,
-instructions and branch mispredictions than base quicksort.
+lowest instruction count occurs with median-of-three partitioning. We also see
+that using a binary search slows multi-quicksort down compared to a sequential
+search due to branch mispredictions, but that both have too many extra
+instructions to compete against memory-tuned quicksort. Despite being slower in
+our tests, our simulations show memory-tuned quicksort to have fewer cache
+misses, instructions and branch mispredictions than base quicksort.
 

### 2005-08-31 docs/tech_report/conclusions.tex
finishded reread of conclusions also minor edits and consistancy checks in other chapters

@@
 fastest sort in our tests. This is due to its low instruction count, few cache
-misses and lack of branch mispredictions. Although we attempted to align it
+misses and lack of any branch mispredictions. Although we attempted to align it
 using another technique from mergesort, we found this decreased its performance.
@@
 outperforming heapsort. We note that our improved version is considerably
-faster, mostly due to its reduced instruction count.
+faster, mostly due to its reduced instruction count. We also measured its branch
+mispredictions, and determined a formula to accurately calculate them.
 
-We were also able to measure it's branch mispredictions, and determine a formula
-to calculate them.
+The improvement made to shellsort, separating the final loop from the rest, also
+results in a significantly faster loop, with an extra speed boost from a
+side-effect skipping an iteration with $h = 2$.
 
-\section{Branch Results}
+\section{Branch Prediction Results}
 
@@
 explicit patterns - which can be exploited by two-level adaptive predictors - do
-not exist, a bimodal predictors perform better.
+not exist, bimodal predictors perform better.
 
@@
 were not identical to our actual results - due to the arrays being very small -
-our actual results did become more predictable in a similar fashion.
+our results showed the branches becoming more predictable in a similar fashion.
 
-Also in heapsort, the number of mispredictions decreases depending on how much
-the heap fans out. Despite this property, the number of mispredictions using an
-8-heap was greater than when using a 4-heap, due to the increased total number
-of branches.
+The result of this is that the number of mispredictions decreases depending on
+how much the heap fans out. Despite this property, the number of mispredictions
+using an 8-heap was greater than when using a 4-heap, due to the increased total
+number of branches.
 
@@
 
-Sedgewick proposed removing individual insertion sorts in quicksort with one
-elegant sweep at the end. LaMarca proposed replacing the large number of
-insertion sorts in order to reduce cache misses. We note that attempting to move
-the pivots, as Sedgewick does, actually increases both the number of branches
-and the rate of mispredictions. We conclude that LaMarca's improvement, which
-slightly improves quicksorts cache performance, also slightly improves it's
-branch prediction performance.
+Sedgewick proposed replacing individual insertion sorts in quicksort with one
+elegant sweep at the end. LaMarca proposed replacing this with the previous
+large number of insertion sorts in order to reduce cache misses. We note that
+attempting to move the pivots, as Sedgewick does, actually increases both the
+number of branches and the rate of mispredictions. We conclude that LaMarca's
+improvement, which slightly improves quicksorts cache performance, also slightly
+improves it's branch prediction performance.
 
@@
 that the more keys considered when determining the median, the lower the number
-of branches. However, this can result in an increase in the rate of branch
-mispredictions. We determine that reducing entropy is not the best way of making
-quicksort perform faster, and that using a median-of-three actually achieves the
-higher performance.
+of branches. However, this results in an increase not only in the rate of branch
+mispredictions, but also in the number of actual branch misprediction. We
+determine that reducing entropy is not the best way of making quicksort perform
+faster, and that using a median-of-three actually achieves the higher
+performance.
 
@@
 We are able to accurately predict the number of branch mispredictions in
-shellsort, using our results from insertion sort. We also found that the
-improvement mode to shellsort, separating the final loop from the rest, results
-in a significantly faster loop, with an extra speed boost from a side-effect
-skipping an iteration with $h = 2$.
+shellsort, using our results from insertion sort.
 
-\section{Comparison with LaMarca's Results}
+\section{Cache Results and Comparison with LaMarca's Results}
+
+In several cases, we found that direct-mapped caches performed better than
+fully-associative ones, resulting in fewer misses. This was due to they're
+replacement policy, and that the sorts in question were heavily optimized for
+the direct-mapped caches. None of the changes slowed the fully-associatvive
+caches however.
 
@@
 LaMarca's heapsort graphs were equivalent to ours in the areas of level 1
-misses, instruction count and cycle count. Like LaMarca, we found that a four
-heap lowered our instruction count. Our level 2 misses also concurred with
-LaMarca, with only one difference: memory-tuned heapsort, being an out-of-place sort,
-fills the cache fully with a smaller number of keys than base heapsort, which is
-in-place. LaMarca's results do not show this. Nor do they show a difference in
-cache misses between these two sorts when the array does fit in the cache.
-LaMarca's base heapsort has the characteristics of an out-of-place heapsort.
-However, our results show that his implicit heap improvements have led to a
-significantly faster heapsort.
+misses, instruction count and cycle count. Like LaMarca, we found that a 4-heap
+lowered our instruction count, and we found better results by using an 8-heap
+instead, as an 8-heap filled our cache lines exactly. Our level 2 misses also
+concurred with LaMarca, with only one difference: memory-tuned heapsort, being
+an out-of-place sort, fills the cache fully with a smaller number of keys than
+base heapsort, which is in-place. LaMarca's results do not show this. Nor do
+they show a difference in cache misses between these two sorts when the array
+does fit in the cache. LaMarca's base heapsort has the characteristics of an
+out-of-place heapsort. However, our results show that his implicit heap
+improvements have led to a significantly faster heapsort.
 
@@
 LaMarca's initial improvements to base mergesort were expected to reduce the
-instruction count. We noted a reduction with each of the steps he prescribed. We
-note too that each of his improvement reduce level 2 cache misses, and the shape
-of our level 2 cache miss graphs exactly match his. We also note similar
-characteristics: LaMarca mentions a wobble in the instruction count graphs, due to
-having to copy the final merge back; we note this same wobble.
+instruction count. We noted such a reduction with each of the steps he
+proscribed. We note too that each of his improvement reduce level 2 cache
+misses, and the shape of our level 2 cache miss graphs exactly match his. Our
+graphs also have similar characteristics: LaMarca mentions a wobble in the
+instruction count graphs, due to having to copy the final merge back; we note
+this same wobble.
 
@@
 
-We note also that aligning the level 2 cache alone leads to worse
-performance than not making any improvements at all, due to the thrashing of the
-level 1 cache. When we aligned the arrays for the level 1 cache, the results
-were much closer to LaMarca's results, and showed what an improvement his
-techniques brought to those algorithms.
+We note also that aligning the level 2 cache alone leads to worse performance
+than not making any improvements at all, due to the thrashing of the level 1
+cache. When we aligned the arrays for the level 1 cache, the results were much
+closer to LaMarca's results, and showed what an improvement his techniques
+brought to those algorithms.
+
+In addition to the double alignement, we also added two new improvements. The
+mergesorts were tiled for the level 1 caches as well as the level 2, slightly
+reducing the number of level 1 cache misses, and we removed the need for a step
+to copy the array back from the auxilliary array; instead, we presort a step
+further in cases where this occurs, and find the merge naturally ends in the
+correct destination.
 
@@
 LaMarca's 20\%. The shape of our instruction count graphs were the same, and
-switching to a binary search in multi-quicksort reduces the instruction count.
-Like him, we found that the base quicksort had very few cache misses, and that
-not using Sedgewick's insertion sort improvement reduces cache misses. We found
-that multi-quicksort's k-way sort changes the slope of the graph, but our result
-were not conclusive as to whether this would eventually make multi-quicksort
-faster than the other versions. Our results do show it being slower for all our
-test cases, due to the increase in instruction count.
+switching to a binary search in multi-quicksort reduced the instruction count.
+Like LaMarca, we found that the base quicksort had very few cache misses, and
+that not using Sedgewick's insertion sort improvement reduces cache misses. We
+found that multi-quicksort's k-way sort changed the slope of the graph, but our
+result were not conclusive as to whether this would eventually make
+multi-quicksort faster than the other versions. Our results do show it being
+slower for all our test cases, due to the increase in instruction count.
 
@@
 to mergesort could be applied to it, and that its resulting cache performance is
-sufficiently good that its low instruction count and lack of branch
-mispredictions make it the fastest sort in our survey.
+sufficiently good that - partially due to its low instruction count and lack of
+branch mispredictions - it is the fastest sort in our survey.
 
@@
 so this shall be considered. A comparative chart of the fastest sorts is shown
-in Figure \ref{All cycles}. This shows the cycles per key of each of the major
+in Figure \vref{All cycles}. This shows the cycles per key of each of the major
 sorts and their derivatives, measured using Pentium 4 performance counters.
@@
 
-The best stable sort is radixsort. The best comparison-based stable sort is
-double-aligned tiled mergesort. The best in-place sort is quicksort. The best in-place
-stable sort we examine in this report is insertion sort. The best elementary
-sort is also insertion sort; as such it should be only used for very small data
-sets. The best sort for very large data sets is radixsort. The fastest overall
-sort was memory-tuned radixsort, though quicksort is very close.
+The best stable sort is memory-tuned radixsort. The best comparison-based stable
+sort is double-aligned tiled mergesort. The best in-place sort is base
+quicksort. The best in-place stable sort we examine in this report is insertion
+sort. The best elementary sort is also insertion sort; as such it should be only
+used for very small data sets. The best sort for very large data sets is
+memory-tuned radixsort. The fastest overall sort was memory-tuned radixsort,
+though all the base and memory-tuned quicksort are both very close.
 
@@
 simple bimodal predictors at predicting branches in sorting algorithms. A
-median-of-3 quicksort outperforms other medians; radix sort has no branch
+median-of-three quicksort outperforms other medians; radixsort has no branch
 mispredictions; shellsort has easily calculable branch mispredictions, and
-performs very similarly to an $O(NlogN)$ so.
+performs very similarly to an $O(NlogN)$ sort.
 
@@
 comparison of $O(NlogN)$ sorts, of different medians for quicksort, of $O(N^2)$
-sorts and of cache conscious sorts. We present branch prediction performance
+sorts and of cache-conscious sorts. We present branch prediction performance
 figures for most major sorts, and adapt cache-conscious strategies from
-mergesort to make an improved radixsort the fastest performing sort in our
-tests, and to make mergesort itself significantly faster.
+mergesort to not only significantly increase the performance of mergesort, but
+also to make memory-tuned radixsort the fastest performing sort in our tests.

### 2005-08-31 docs/tech_report/heap.tex
davids fixes to heap sort due to incorrect taken vs not-taken. also, first corrections to conclusions
finishded reread of conclusions also minor edits and consistancy checks in other chapters

@@
 An additional consideration is that the base heapsort has more instructions than
-it's out-of-place equivalent. This means that the 4-heap will have a better
+its out-of-place equivalent. This means that the 4-heap will have a better
 instruction count, as well as better cache performance than the base heapsort.
@@
 two children in the 2-heap, and the `\n{Creation, Child Promotion}' refers to
-checking if the child is larger than it's parent. The same checks are later made
+checking if the child is larger than its parent. The same checks are later made
 during the destruction of the heap, and are referred to as `\n{Destruction, Has
@@
 than taken. This is because all branches apart from the first are biased in the
-taken direction. This should improve accuracy. On the other hand, on those
+not-taken direction. This should improve accuracy. On the other hand, on those
 occasions when the bi-modal branch predictor is pushed towards predicting taken,

### 2005-08-31 docs/tech_report/report.tex
minor edits and merged in davids writing
fixed stupid mistake accidentally committed. added axis labels to bubblesort graphic
so close to being done i can taste it
finishded reread of conclusions also minor edits and consistancy checks in other chapters

@@
 %\n{x} is used for
-% - proper names: Atom, SimpleScalar, Papi, GnuPlot, ValGrind, glibc
+% - proper names: Atom, SimpleScalar, Papi, GnuPlot, ValGrind, glibc, PapiEx
 % - technical names: unsigned integer, less-than, greater-than, in-place, out-of
@@
 				\vspace{2em}
-				\large Technical Report TR-CS-2005-XX\\
+				\large Technical Report TCD-CS-2005-57\\
                 \large Department of Computer Science,\\
 				\large University of Dublin, Trinity College,\\
-				\large Dublin 2, Ireland.
+				\large Dublin 2, Ireland.\\
+				\vspace{2em}
+				\large August, 2005\\
         \end{center}
@@
 
-We also developed a number of other improvements to the algorithms,
-such as removing the need for a sentinel in heapsort. Overall, we
-found that a cache-optimized radix sort was the fastest algorithm for
-all large arrays; the absence of comparison branches means that the
-algorithm causes almost no branch mispredictions.
+We also developed a number of other improvements to the algorithms, such as
+removing the need for a sentinel in heapsort. Overall, we found that a
+cache-optimized radixsort was the fastest algorithm for all large arrays; the
+absence of comparison branches means that the algorithm causes almost no branch
+mispredictions.
+
+% TODO add copy back
 \end{abstract}

### 2005-08-31 docs/tech_report/conclusions.tex
davids email conclusion changes merged in

@@
 determined that bubblesort has better branch performance at the start and end of
-its sorting, then in the middle.
+its sorting, than in the middle.
 
@@
 
-No results for the branch prediction performance of sorting algorithm yet
-exists. We discovered some interesting new results.
+To our knowledge, this is the first large-scale, systematic study of branch
+prediction in sorting algorithms. We discovered some interesting new results.
 
@@
 our 8-heap, and that this pattern matched closely that of selection sort,
-discussed above. We found that though our approximated results using $H_N-1$
-were not identical to our actual results - due to the arrays being very small -
-our results showed the branches becoming more predictable in a similar fashion.
+discussed above. We found that the number of mispredictions was close to, but
+measurably higher than, $H_N-1$. The main reason was that the number of items
+from which the selection was being made was small. Although the comparisons are
+mostly biased in one direction, the early ones are sufficiently unpredictable
+that the branch predictor often predicts the less likely direction. The later
+comparisons are highly predictable, and yield prediction accuracies much closer
+to that indicated by $H_N-1$.
 

### 2005-08-31 docs/tech_report/conclusions.tex
spell checked

@@
 LaMarca's and discusses important cache results, discusses the best sort that we
-found in a number of catagories, and suggests contributions made to computer
+found in a number of categories, and suggests contributions made to computer
 science by this research.
@@
 replacement policy, and that the sorts in question were heavily optimized for
-the direct-mapped caches. None of the changes slowed the fully-associatvive
+the direct-mapped caches. None of the changes slowed the fully-associative
 caches however.
@@
 
-In addition to the double alignement, we also added two new improvements. The
+In addition to the double alignment, we also added two new improvements. The
 mergesorts were tiled for the level 1 caches as well as the level 2, slightly
 reducing the number of level 1 cache misses, and we removed the need for a step
-to copy the array back from the auxilliary array; instead, we presort a step
+to copy the array back from the auxiliary array; instead, we presort a step
 further in cases where this occurs, and find the merge naturally ends in the

### 2005-08-31 docs/tech_report/heap.tex
spell checked

@@
 
-The relationship between these numbers and $H_N-1$ is relatively straighforward.
+The relationship between these numbers and $H_N-1$ is relatively straightforward.
 $H_N-1$ represents the cumulative number of changes in the minimum. There first
@@
 We investigated the trade-offs in these two effects and we discovered that using
-a single branch does indeed cause the branches for the ealier comparisons to
+a single branch does indeed cause the branches for the earlier comparisons to
 become more predictable. However, this improvement is outweighed by a loss in

### 2005-08-31 docs/tech_report/intro.tex
fixed small errors. didnt really get rid of any TODOs
made final pass

@@
 thought of when these sorts were written: branch predictors, for example, were
-not created until the 1970s, and advanced branch prediction techniques were
-not used in consumer processors until very recently. It is hardly surprising,
-then, that sorting algorithms rarely take these architectural features into
-account.
+not created until the 1970s, and advanced branch prediction techniques were not
+used in consumer processors until very recently. It is hardly surprising, then,
+that sorting algorithms rarely take these architectural features into account.
 
@@
 LaMarca and Ladner \cite{LaMarca96} break this trend by discussing the cache
-implications of several algorithms, including sorts. This report 
-investigates these claims and tests and analyses high performance sorts
-based upon thier work.
+implications of several algorithms, including sorts. This report investigates
+these claims and tests and analyses high performance sorts based upon their
+work.
 
@@
 discussing implicit heaps, which are later used in heapsort. In
-\cite{LaMarca99}, radixsort was added for comparison. Radixsort is added here
-for the same purpose. For comparison, elementary sorts - insertion sort,
-selection sort, bubblesort and shakersort - are also included here.
+\cite{LaMarca99}, radixsort was added for comparison. Radixsort and shellsort
+are added here for the same purpose. For comparison, elementary sorts -
+insertion sort, selection sort, bubblesort and shakersort - are also included.
 
@@
 
-Several important discovered results are reported here. The
-branch prediction rates of insertion sort and selection sort, for example, and
-the difference in performance between binary searches and sequential searches due
-to branch misses. Several steps to remove instructions from standard algorithms
-are devised, such as double tiling mergesort, removing a sentinel and bounds
-check from heapsort.
+Several important discovered results are reported here. The branch prediction
+rates of insertion sort and selection sort, for example, and the difference in
+performance between binary searches and sequential searches due to branch
+misses. Several steps to remove instructions from standard algorithms are
+devised, such as double tiling and aligning mergesort, removing a sentinel and
+bounds check from heapsort, and adding a copy-back step to radixsort.
 
-Chapter \ref{background} provides background information on caches and branch predictors. It
-also discusses the theory of sorting, and conventions used in the algorithms in
-this report.
+Chapter \ref{background} provides background information on caches and branch
+predictors. It also discusses the theory of sorting, and conventions used in the
+algorithms in this report.
 
-Chapter \ref{method} discusses the method and framework used in this research, including the
-tools used and the reasons for using them.
+Chapter \ref{method} discusses the method and framework used in this research,
+including the tools used and the reasons for using them.
 
-Chapter \ref{ordernsquared}  discusses the elementary sorts and their implementations. Results are
-included for later comparison.
+Chapter \ref{ordernsquared}  discusses the elementary sorts and their
+implementations. Results are included for later comparison.
 
-Chapters \ref{heap} through \ref{radix} discuss base algorithms and cache-conscious improvements
-for heapsort, mergesort and quicksort. Results are presented and explained, and
-branch prediction implications of the changes are discussed.
+Chapters \ref{heap} through \ref{radix} discuss base algorithms and
+cache-conscious improvements for heapsort, mergesort and quicksort. Results are
+presented and explained, and branch prediction implications of the changes are
+discussed.
 
@@
 
-Chapter \ref{conclusions} presents conclusions, and lists contributions
-described by this report.
+Chapter \ref{conclusions} presents conclusions, summarises results from previous
+chapters, compares our results with LaMarcas, presents a list of cache and
+branch prediction results and improvements, discusses the fastest sort in our
+tests and lists contributions made by this research.

### 2005-08-31 docs/tech_report/report.tex
did the abstract. removed comment from the latex files

@@
 \begin{abstract}
-Sorting is one of the most important and studied problems in computer
-science. Many good algorithms exist which offer various trade-offs in
-efficiency, simplicity and memory use. However most of these
-algorithms were discovered decades ago at a time when computer
-architectures were much simpler than today. Branch prediction and
-cache memories are two developments in computer architecture that have
-a particularly large impact on the performance of sorting algorithms.
+Sorting is one of the most important and studied problems in computer science.
+Many good algorithms exist which offer various trade-offs in efficiency,
+simplicity and memory use. However most of these algorithms were discovered
+decades ago at a time when computer architectures were much simpler than today.
+Branch prediction and cache memories are two developments in computer
+architecture that have a particularly large impact on the performance of sorting
+algorithms.
 
-This report describes a study of the behaviour of sorting algorithms
-on branch predictors and caches. Our work on branch prediction is
-almost entirely new, and finds a number of important results. In
-particular we show that insertion sort causes the fewest branch
-mispredictions of any comparison-based algorithm, that optimizations -
-such as the choice of the pivot in quicksort - can have a large impact
-on the predictability of branches, and that advanced two-level branch
-predictors are usually worse at predicting branches in sorting
-algorithms than simpler branch predictors. In many cases it is
-possible to draw links between classical theoretical analyses of
-algorithms and their branch prediction behaviour.
+This report describes a study of the behaviour of sorting algorithms on branch
+predictors and caches. Our work on branch prediction is almost entirely new, and
+finds a number of important results. In particular we show that insertion sort
+causes the fewest branch mispredictions of any comparison-based algorithm, that
+optimizations - such as the choice of the pivot in quicksort - can have a large
+impact on the predictability of branches, and that advanced two-level branch
+predictors are usually worse at predicting branches in sorting algorithms than
+simpler branch predictors. In many cases it is possible to draw links between
+classical theoretical analyses of algorithms and their branch prediction
+behaviour.
 
-The other main work described in this report is an analysis of the
-behaviour of sorting algorithms on modern caches. Over the last decade
-there has been considerable interest in optimizing sorting algorithms
-to reduce the number of cache misses. We experimentally study the
-cache performance of both classical sorting algorithms, and a variety
-of cache-optimized algorithms, primarily those proposed by LaMarca and
-Ladner. Our experiments cover a much wider range of algorithms than
-other work, including the $O(N^2)$ sorts and shellsort, all within a
-single framework. We discover a number of new results, particularly
-relating to the branch prediction behaviour of cache-optimized sorts.
+The other main work described in this report is an analysis of the behaviour of
+sorting algorithms on modern caches. Over the last decade there has been
+considerable interest in optimizing sorting algorithms to reduce the number of
+cache misses. We experimentally study the cache performance of both classical
+sorting algorithms, and a variety of cache-optimized algorithms proposed by
+LaMarca and Ladner. Our experiments cover a much wider range of algorithms than
+other work, including the $O(N^2)$ sorts, radixsort and shellsort, all within a
+single framework. We discover a number of new results, particularly relating to
+the branch prediction behaviour of cache-optimized sorts.
 
 We also developed a number of other improvements to the algorithms, such as
-removing the need for a sentinel in heapsort. Overall, we found that a
-cache-optimized radixsort was the fastest algorithm for all large arrays; the
-absence of comparison branches means that the algorithm causes almost no branch
+removing the need for a sentinel in classical heapsort. Overall, we found that a
+cache-optimized radixsort was the fastest sort in our study; the absence of
+comparison branches means that the algorithm causes almost no branch
 mispredictions.
 
-% TODO add copy back
 \end{abstract}
@@
 
-%TODO \chapter{Software Predictor Result Listing}
 %\input software_predictor.tex

### 2005-11-01 docs/tech_report/conclusions.tex
did the abstract. removed comment from the latex files
uncommitted changes from ages ago

@@
 In several cases, we found that direct-mapped caches performed better than
-fully-associative ones, resulting in fewer misses. This was due to they're
+fully-associative ones, resulting in fewer misses. This was due to their
 replacement policy, and that the sorts in question were heavily optimized for
@@
 \vspace*{-10em}
-%\setcounter{figure}{9} % note the cheating to make this work
 \begin{figure}[H]
@@
 memory-tuned radixsort. The fastest overall sort was memory-tuned radixsort,
-though all the base and memory-tuned quicksort are both very close.
+though the base and memory-tuned quicksorts are all very close.
 

### 2005-11-01 docs/tech_report/intro.tex
widenned the pages
uncommitted changes from ages ago

@@
 Papers considering sorting performance typically analyse instruction count.
-LaMarca and Ladner \cite{LaMarca96} break this trend by discussing the cache
+LaMarca and Ladner break this trend in \cite{LaMarca96} by discussing the cache
 implications of several algorithms, including sorts. This report investigates
@@
 Chapters \ref{heap} through \ref{radix} discuss base algorithms and
-cache-conscious improvements for heapsort, mergesort and quicksort. Results are
-presented and explained, and branch prediction implications of the changes are
-discussed.
+cache-conscious improvements for heapsort, mergesort, quicksort and radixsort.
+Results are presented and explained, and branch prediction implications of the
+changes are discussed.
 
@@
 Chapter \ref{conclusions} presents conclusions, summarises results from previous
-chapters, compares our results with LaMarcas, presents a list of cache and
+chapters, compares our results with LaMarca's, presents a list of cache and
 branch prediction results and improvements, discusses the fastest sort in our

### 2005-11-01 docs/tech_report/merge.tex
lots of minor changes
did the abstract. removed comment from the latex files
uncommitted changes from ages ago

@@
-%\section{Description}
 \label{merge}
@@
 to make room, while the direct-mapped caches are designed to make best use of
-the keys while they are in the cache. They have, effectively, an optimised
+the keys while they are in the cache. They have, effectively, an optimized
 replacement policy.
@@
 Figure \ref{mergesorts-1} shows the behaviour of the individual branches within
-algorithm N. The labels in the graphs correspond to the names of the steps in
-Knuth's algorithm (we also used the same names for the labels of branch targets
-in our C source code. Branch N3 is the main comparison branch; it is used to
-choose the smaller of the elements from the start of the two sub-arrays being
-merged. The direction of this branch is completely random - it depends entirely
-on the (in this case random) data. Figure \ref{mergesorts-1} also shows that
-this branch is almost entirely unpredictable using a bimodal predictor. The
-result for a two-level predictor is much the same, because there is no pattern
-in the outcome of these branches. Branches `\n{N3 i==j}' (check for the current
-merge being complete), `\n{N5}' and `\n{N9} (both checks for the end of a run of
-increasing elements) are control-flow branches, and are almost entirely
-predictable. Figure \ref{mergesorts-2} shows that the comparison branch in
-Algorithm S is just as unpredictable as that in algorithm N. However, given that
-the merge in algorithm S is entirely regular (see section \ref{Algorithm S}) it
-is possible to eliminate one of the control-flow branches.
+algorithm N. The labels in the graphs of algorithms N and S correspond to the
+names of the steps in Knuth's algorithm (we also used the same names for the
+labels of branch targets in our C source code. Branch `\n{N3}' is the main
+comparison branch; it is used to choose the smaller of the elements from the
+start of the two sub-arrays being merged. The direction of this branch is
+completely random - it depends entirely on the (in this case random) data.
+Figure \ref{mergesorts-1} also shows that this branch is almost entirely
+unpredictable using a bimodal predictor. The result for a two-level predictor is
+much the same, because there is no pattern in the outcome of these branches.
+Branches `\n{N3 i==j}' (check for the current merge being complete), `\n{N5}'
+and `\n{N9} (both checks for the end of a run of increasing elements) are
+control-flow branches, and are almost entirely predictable. Figure
+\ref{mergesorts-2} shows that the comparison branch in Algorithm S is just as
+unpredictable as that in algorithm N. However, given that the merge in algorithm
+S is entirely regular (see section \ref{Algorithm S}) it is possible to
+eliminate one of the control-flow branches.
 

### 2006-01-19 docs/branch_paper/paper.tex
added david's branch paper made the README a lot more accurate. Suffice to say it was well off removed the need for lgrind, which wasnt used, from the report

\documentclass[acmtocl]{acmtrans2m}

\usepackage{graphicx}

\newtheorem{theorem}{Theorem}[section]
\newtheorem{conjecture}[theorem]{Conjecture}
\newtheorem{corollary}[theorem]{Corollary}
\newtheorem{proposition}[theorem]{Proposition}
\newtheorem{lemma}[theorem]{Lemma}
\newdef{definition}[theorem]{Definition}
\newdef{remark}[theorem]{Remark}

\newcommand{\comment}[1]{}

\title{An Experimental Study of Sorting and Branch Prediction}

\author{
PAUL BIGGAR
\and
DAVID GREGG
\\University of Dublin, Trinity College
}

\begin{abstract}

\end{abstract}

\category{E.5}{Data}{Files}[Sorting/Searching]
\category{C.1.1}{Computer Systems Organization}{Processor Architectures, Single Data Stream Architectures}[Pipeline architectures]
\terms{Algorithms, Experimentation, Measurement, Performance}
\keywords{Sorting, Branch prediction}

\begin{document}

\begin{bottomstuff}
Corresponding author's address: David Gregg, Department of Computer
Science, University of Dublin, Trinity College, Dublin 2, Ireland. {\tt
David.Gregg@cd.tcd.ie}.
\end{bottomstuff}

\maketitle

\section{Motivation}
Classical analyses of algorithms make simplifying assumptions about
the cost of different machine instructions. For example, Knuth's Mix
machine code assumes that all operations have unit costs. More
recently researchers have recognized that on modern computers the cost
of accessing memory can vary dramatically depending on whether the
data can be found in the first-level cache, or must be fetched from a
lower level of cache or even main memory.  This has spawned a great
deal of research on cache-efficient searching and sorting \cite{}.

Another type of instruction whose cost can vary dramatically is the
conditional branch. Modern pipelined processors depend on \emph{branch
prediction} for much of their performance.  If the direction of a
conditional branch is correctly predicted ahead of time, the cost of
the conditional branch may be as little as the cost of, say, an
integer add instruction. If, on the other hand, the branch is
mispredicted the the processor must flush its pipeline, and restart
from the correct target of the branch. This cost is typically a large
multiple of the cost of executing a correctly-predicted branch. For
example, the Intel Pentium 4 processor has a 31-stage pipeline,
meaning that a branch misprediction will cost around 30 cycles.  In
contrast, on the same processor the cost of a memory access that
misses the first level cache and hits the second level cache is 23
cycles. Fortunately, the branches in most programs are very
predictable, so branch mispredictions are usually rare (prediction
accuracies of greater than 90\% are typical).

The cost of executing branches is particularly important for sorting
because the inner loops of most sorting algorithms consist of
camparisons of items to be sorted. Thus, the predictability of
these comparison branches is critical to the performance of
sorting algorithms. In this paper we study the predictability
of branches is most of the major sorting algorithms. We also
examine cache-conscious variations of some of the algorithms,
and show how optimizations for the cache impact in branch
mispredictions.

The rest of this paper is organized as follows.

\section{Branch prediction}

\section{Experimental setup}

\section{Elementary sorts}
\label{ordernsquared}
Elementary sorts are also referred to as $O(N^2)$ sorts. These
algorithms have the worst performance for large inputs. However they
have two advantages: they are simple to implement correctly and they
are often faster than more sophisticated algorithms for small input
sets. Four of these sorts are discussed here: selection sort,
insertion sort, bubblesort and shakersort.

Selection sort is a simple algorithm that repeatedly finds the
smallest element of the array and removes it to a sorted section at
the start. It requires approximately $N^2/2$ comparisons and exactly
$N$ exchanges. Insertion sort takes a different strategy and instead
creates an (initially empty) sorted region at the start of the array
and inserts items into their correct position in this sorted region.
Insertion sort performs in linear time on a sorted list. With a random
list, there are, on average, there are $N^2/4$ comparisons and $N^2/4$
half-exchanges. Thus, selection sort performs around twice as many
branches as insertion sort.

Although we know the approximate number of comparisons that each
algorithm performs, and important question is whether there is any
difference of the predictability of the branches in each algorithm.
Figure \ref{} shows the number of mispredictions per key for both the
algorithms using random data, and varying data set sizes, measured
using the Simplescalar processor simulator.

There are three notable features in this data. Firstly, the number of
mispredictions is very low. If the behaviour of these branches were
random, then we would expect that half of the comparison branches
would be mispredictions, that we would expect $O(N^2)$
mispredictions. In fact, the number of mispredictions per key is less
than 12 even for the largest arrays. Secondly, although selection sort
performs twice as many comparison branches it does not cause twice as
many branch mispredictions. For the bimodal predictor, it causes 9-12
times as many mispredictions within the size of input sets that we
examine. Thirdly the two-level branch predictor is slightly worse at
predicting the comparison branches in these algorithms than the simple
bimodal predictor.  This is a remarkable result, because two-level
predictors are similar to, or sigificantly more accurate than bimodal
predictors for almost all other types of branches \cite{Uht}.

\subsection{Insertion sort}
With a bimodal predictor, insertion sort causes almost exactly one
branch misprediction per key, regardless of the size of the array.
Thus, although insertion sort performs around $N/4$ comparisons per
key, only one of these comparisons causes a misprediction. The
reason for this becomes clear when we consider the inner loop
of insertion sort:
\begin{verbatim}
   while( item < a[j-1] ) {
     a[j] = a[j-1];
     j--;
   }
   a[j] = item;
\end{verbatim}

The comparison in insertion sort is also the loop guard.  As long as
the loop continues, this comparison always resolves in the same
direction, and is thus highly predictable. When the location for the
item is finally found, the comparison resolves in the opposite
direction, and a branch misprediction occurs. Thus, for a bimodal
predictor, the number of mispredictions per key is almost exactly
one. Although the results we present are for random data, the same
is true regardless of the input data. Insertion sort will always
cause around one misprediction per key.

The two-level predictor combines a pattern of recent branch outcomes
make a prediction. This goal is to exploit correlations between branch
outcomes. For most programs this is valuable, but there is correlation
of branch outcomes in insertion sort. The additional complexity simply
causes the predictor to find false correlations, with the result that
number of mispredictions is higher than for a simpler, bimodal
predictor. In general, misprediction rates of more than one per key
are caused by imperfections in the predictor, rather than the
unpredictability of the comparison branch in insertion sort.

\subsection{Selection sort}
The comparison branch in selection sort is also highly predictable,
but less so than for insertion sort. The inner loop of selection
sort consists of searching for the smallest element of part of
the array:
\begin{verbatim}
  min = i;
  for ( j = i+1; j < N; j++ )
    if ( a[j] < a[min] ) min = j;
\end{verbatim}
The predictability of the comparison branch depends on the number of
times that the cadidate minimum, $min$, changes. Knuth \cite{Knuth97}
shows that for $N$ elements, the number of changes is $H_N - 1$, where
$H_N$ is the harmonic series:
$$H_N = 1 + \frac{1}{2} + \frac{1}{3} + ... + \frac{1}{N} =  \sum_{k=1}^{N}
\frac{1}{k},  \hspace{0.4cm}  N \geq 1$$

Within practical sorting ranges, $H_N$ grows at a rate similar to
$log_e(N)$. Thus, in the average case, we can expect that the comparison
branch will mostly go in the same direction, making the branch highly
predictable.

The number of changes to the left-to-right minimum in the best case is
zero and in the worst case $N-1$. In the worst case, the array is
sorted in reverse order, and so the left-to-right minimum changes on
every element. Interestingly, the worst cases for the number of
changes is not the worst case for branch mispredictions. If the
minimum changes for every element, then the comparison branch will
always resolve in the same direction, and thus be almost perfectly
predictable. From the point of view of branch prediction, frequent
changes (preferably with no simple pattern) are the main indicator of
poor performance, rather than the absolute number of times that the
branch resolves in each direction. The worst case for selection sort
would be an array which is partially sorted in reverse order; the
left-to-right minimum would change frequently, but not predictably so.

\subsection{Bubble sort}
Bubblesort is a notoriously inefficient algorithm that compares and
exchanges adjacent pairs of elements in the array. One execution of
the inner loop will move the smallest element of the array to the
first position, and partially sort the rest of the array by moving
small elements towards the start. Cocktail shaker sort (or simply
shakersort) alternates the direction of the inner loop to take
more advantage of this partial sorting to reduce the number of
outer-loop iterations.

Figure \ref{} shows the number of branch mispredictions per key for
bubble- and shakersort.  Nearly one in four comparisons in bubblesort
are mispredicted. As the data sets gets large, there are nearly 10000
branch mispredictions per key key. Shakersort executes a lot fewer
branches, and so has fewer mispredictions, but the rate of
misprediction is similar.

The branch behaviour of bubblesort is particularly interesting when
compared to selection sort. Although we usually think of the two
algorithms as being different, they are actually very similar. Like
selection sort, bubblesort finds the smallest key in the
unsorted part of the array and moves it to start of the
array. Thus, we would expect that bubblesort's branch prediction
behaviour would be similar to that of selection sort.

However, in the process of moving the smallest element into position,
bubblesort also rearranges other array elements. Selection sort sweeps
across the array and always keeps track of the left-to-right minimum
up to the current point.  However, bubblesort actually moves this
minimum, exchanging pairwise with each element it encounters which is
larger. Thus, even the unsorted part of the array becomes partially
sorted. We might expect that bubblesort's comparison branch for a
partially sorted array might be more predictable than for a random
one, but it is not. With a random array, the smaller elements are
equally likely to appear anywhere in the array. With a partially
sorted array, the smaller elements tend to appear towards the start of
the array which is, in bubblesort, the last place that we look. Thus,
the candidate minimum is changed much more often than if we were
searching a randomly ordered array.

\begin{figure}
\includegraphics{plots/two_point_counter_bubblesort.eps}
\caption{Behaviour of branches in bubblesort compared to the ``sweep'' number}
\label{Predictability-of-branches-in-bubblesort-compared-to-sweep-number}
\end{figure}

To demonstrate this effect, we measured how the predictability of
bubblesort's comparison branch varies with successive sweeps across
the array. We added a simple software simulation of a bimodal
predictor to the source code of base bubblesort, and ran it on a large
number of randomly generated arrays, each with 8192 elements.

Figure
\ref{Predictability-of-branches-in-bubblesort-compared-to-sweep-number}
shows that on the early iterations the number of correct predictions
is very close to the number of executed comparison branches. The
comparison branch is highly predictable because bubblesort's search
for the smallest element is similar to selection sort. However, with
successive sweeps even the unsorted part of the array becomes
increasingly sorted and the branch becomes more unpredictable. We see
that the number of executed comparison branches falls linearly, but
that the number of correctly predicted comparison branches falls much
more quickly. Eventually, the array becomes so close to fully sorted
that the branch becomes increasingly predictable in the other
direction, and the number of correct predictions again approaches the
number of executed branches. However, during the intermediate stage,
a huge number of branch mispredictions have already
occurred. Paradoxically, by partially sorting the array, bubblesort
greatly increases the number of branch mispredictions and thus greatly
increases its running time.

Overall, our results show that branch prediction is an important
factor in the efficiency of sorting algorithms. For example,
shakersort executes fewer machine instructions and causes fewer cache
misses than selection sort, but is slower due to branch
mispredictions. Insertion sort is confirmed as the most efficient
elementary sorting algorithm. In additon to having the best time
complexity it causes only around one branch misprediction per key even
in the worst case.

\section{Shellsort}
Shellsort \cite{Shell59} was the first in-place sub-$N^2$ sorting
algorithm. The algorithm operates by sorting all elements that are $h$
elements apart using insertion, where $h$ is some increment.  On early
iterations the increment is large, allowing items to move large
distances. The increment is reduced on successive iterations to allow
more local sorting. The final iteration is a standard insertion sort,
after which the array is sorted. We use Gonnet's increments, which set
the next value of the increment to $\frac{5}{11}^{th}s$ of the
previous increment, because we found them to be efficient in practice.
All discussions of complexity are based on these increments.

The branch performance of shellsort should be predictable. There are
$log_{(\frac{11}{5})}N$ increments, each with a separate pass over the
array. On each pass, all keys are insertion sorted into some
sub-array. Given that insertion sort causes around one branch
misprediction per key, we can expect $N$ comparison branch
mispredictions per pass. It is expected, therefore, that there will be
around $log_{(\frac{11}{5})}N$ mispredictions per key. In general, if
there are $i$ increments, shellsort will cause $ni$ mispredictions.

Figure \ref{} shows the number of branches and branch mispredictions
per key for shellsort. For 4194304 keys, we would expect
$log_{(\frac{11}{5})}4194304 = 18.34$ mispredictions per key. The
number of mispredictions we measure in our simulation is actually a
little under 18, which is remarkably close to the predicted number.

To further investigate this we separated the comparison branches from
the other branches in the code in a separate simulation of a bimodal
branch predictor on a large number of randomly generated arrays. We
found that using Gonnet's increments each key moves only an average of
0.9744 places in each insertion sort (although moving one place
actually involves skipping over $h$ other array elements, where $h$ is
the increment). Thus, the situation is somewhat different to insertion
sorting large arrays of random elements where each element can be
expected to move a large number of places. Instead of the comparison
branch almost always resolving in one direction, the actual outcome
will be close to random. We can still expect a number of
mispredictions that is close to one per key, but we are unlikely to
get the almost perfectly predictable behaviour that we see when using
insertion sort where the average number of places moved is larger.

The main apparent weakness of shellsort on modern architectures is its
poor locality of reference in memory access.  By comparing items which
are separated by $h$ keys, there is little reuse of items in the same
cache line. One can easily imagine cases where shellsort would cause
very large numbers of cache misses.  In the worst case, each key would
move a very large number of places in each insertion sort pass,
possibly causing a cache miss for every place that it moves.  To
further investigate this effect, we measured the number of first and
second-level cache misses caused by our implementation of shellsort
(see figure \ref{}). In fact, the number of mispredictions caused by
our shellsort is significant, but not much greater than other sorting
algorithms such as heapsort and merge sort (see sections \ref{} and
\ref{}). The main reason is that using Gonnet's increments on average
keys move only 0.9744 places on each pass, so very large numbers of
references in diverse parts of the array are not needed.

\section{Heapsort}
Heapsort is an in-place sort which works in $O(NlogN)$ time in the
average and worst case. It begins by creating a structure known as a
\n{heap}, a binary tree in which a parent is always larger its child,
and the largest key is at the top of the heap. It then repeatedly
removes the largest key from the top and places it in its final
position in the array, fixing the heap as it goes.

\section{Quicksort}
Quicksort is a divide and conquer algorithm that recursively
partitions its array around a pivot, swapping large keys on the left
hand side with smaller keys on the right hand side. The partition
causes the pivot to be put in its final position, with all greater
keys to the right, and lesser keys on the left. The left and right
partitions are then sorted recursively.

\subsection{Simple Quicksort}
Sedgewick \cite{} has proposed a number of improvements to the
standard quicksort algorithm, which we use in our base quicksort
implementation. First, we use an explicit stack to save space.
Secondly, we always sort the smaller partition of the array first, to
reduce the worst case size of the stack (from $O(N)$ to
$O(logN)$). Thirdly, we not quicksort patitions of less than a certain
\texttt{THRESHOLD} size; instead we do a final insertion pass over the
array after quicksorting. The smallest of the first \texttt{THRESHOLD}
keys is placed in the first location of the array as a
sentinel. Finally, we use median-of-3 to select the pivot, which
greatly reduces the likelihood of very unbalanced partitioning. This
also allows the bounds check to be removed from the inner loop,
greatly reducing the instruction count. Improvements such as these are
found in most efficient implementations of quicksort.

In addition to our base quicksort algorithm, we experimented with
variations which attempt to reduce the number of data cache
misses. LaMarca \cite{} proposed a \textit{memory-tuned} quicksort.
Once a partition is reduced below the size of the threshold, insertion
sort is preformed immediately. By doing this, the keys being sorted
are in the cache already. By leaving the insertion sort until the end,
as Sedgewick does, the reduction in instruction count is made up by
the increase in cache misses, once the lists being sorted are larger
than the cache.

\subsection{Multi-Quicksort}
Generally, quicksort has good locality of reference. It divides the
same region of the array into increasingly small partitions.  In the
case that the array to be sorted fits in a particular level of the
cache, then quicksort has excellent cache properties at that
level. Once the array is larger than the cache, however, the number of
misses increases dramatically. On the early passes, which divide the
array into large sections, there is likely to be little reuse of data
cache entries between passes. Given that there is little reuse between
passes, it would be better to reduce the number of passes by dividing
the array into a larger number of partitions on each pass. LaMarca
uses a technique similar to that used in multi-mergesort, in that a
\n{multi-way partition} is done at the start, moving all keys into
cache sized containers which can be sorted individually.

Implementing a multi-way partition is complicated. At the start of the
algorithm, a set of pivots is chosen and sorted, and keys are put into
containers based on which two pivots their value lies between. A
minimal and maximal key are put on the edges of this list to avoid
bounds checks as the list is iterated over.

The main problem with multi-quicksort is that we don't know how many
keys there will be in each container. We implement the containers as a
linked list of arrays. LaMarca notes very little performance
difference between 100 and 5000 keys in each block. The size is
therefore chosen so that each list fits exactly into a page of
memory. When a list is filled, another list is linked to it, and these
lists hold all the keys in the container. The total number of
containers is chosen so that on average each will hold $C/3$ keys,
where $C$ is the number of keys which fit in the cache. LaMarca shows
that using this formula on average only 5\% of subsets will be larger
than the cache.

The entire array is iterated though a key at a time. For each key, the
list of pivots are searched until the appropriate container is found,
and the key is added to that container. Sequentially searching the
lists increases the instruction count by 50\% over the memory-tuned
version. For this reason, the search was replaced with an efficient
binary search, which reduced by half the extra instructions.

Once the entire array has been put into containers, each container is
emptied in turn and the keys put back into the array. When a container
is empty, the pivot greater than that container is put into position,
which is guaranteed to be its final position. The indices of these
positions is pushed onto the stack, and the sub-array is then sorted.

While emptying the containers back into the array, an opportunity is
taken to find the smallest key in the leftmost partition. This is
placed as a sentinel, ensuring that every partition has a sentinel,
and that no bounds checks are required.

\subsection{Quicksort Results}
Figure \ref{} shows the

\section{Radixsort}

\section{Related work}

% The IBM 7030 "Stretch" supercomputer was the first general-purpose
% pipelined architecture \cite{Bloch59}. Released in 1959 it had a
% four-stage pipeline and many other architectural innovations that are
% now commonplace in desktop and even embedded computers.

We present the first large-scale systematic study of branch prediction
and sorting. However, researchers have been aware for many years that
the branches in comparison-based sorts cause problems on pipelined
architectures. As early as 1972 Knuth commented on the
'ever-increasing number of "pipeline" or "number crunching" computers
that have appeared in recent years' whose 'efficiency deteriorates
noticeably in the presence of conditional branch instructions unless
the branch almost always goes the same way'. He also notes that 'radix
sorting is usually more efficient than any other known method for
internal sorting on such machines' \cite{Knuth72c}. Most likely, Knuth
was referring to the IBM 7030 "Stretch" computer (released in 1961
with a four-stage pipeline), Seymor Cray's Freon-cooled CDC 6600
supercomputer (released 1964), or other early supercomputers which
used pipelining in the execution of general-purpose instructions.

A number of other authors have considered optimized implementations of
sorting algorithms for more recent single-processor machines. Nyberg
et al present a quicksort-based algorithm for RISC architectures
\cite{Nyberg+94}. They found that the main limit on performance was
cache behaviour, and did not mention branch prediction even
once. Bentley implemented a highly-tuned quicksort, to be used in the
\texttt{qsort} function of the standard C library
\cite{Bentley93}. Again, Bentley did not consider branch prediction,
perhaps because the first desktop processors with dynamic branch
predictors (such as the DEC Alpha 21064, MIPS R8000 and Intel Pentium)
were only just appearing around that time.

Agarwal developed an optimized algorithm for sorting random records on
an IBM RS/6000 RISC machine \cite{Agarwal96}. His is the first
published algorithm that we are aware of that deliberately attempts to
improve performance by eliminating branch mispredictions. First, it
uses the seven higher order bits of the key to divide the data into
128 buckets. Each of these buckets is radix-sorted. However, before
the radix sort is reaches the lowest order bits, the probability of
any adjacent pair of keys being out of order is extremely low, because
the problem states that the data is guaranteed to be random. In the
final stage, the algorithm checks that large runs of keys are
correctly ordered using subtraction and bitwise operations, and only
sorts a section if some key is found to be out of order. Although
Agarwal's techniques are interesting, they are very dependent on the
randomness of the data; it is easy to construct commmon cases that
would result in very poor performance. Nonetheless, the work is
interesting because it shows the effectiveness of radix sorting
techniques on pipelined architectures.

A recent trend in optimizing algorithms for FFT, linear algebra, and
digital signal processing has been to automatically generate and
evaluate very large numbers of variants of the algorithm.  A similar
approch is used by \cite{Li+05} to automatically generate sorting
algorithms that are tuned to the architecture of the target machine
and the characteristics of the input data. They identify six sorting
primitives, which can be combined to construct a sorting algorithm.  A
large number of combinations are tested, using a genetic algorithms
search to find the most efficient. The resulting implementations are,
in many cases, significantly faster than commercial sorting libraries.
Although they do not consider branch prediction explicitly, they find
that the implementations that perform best usually use radix sort for
a large part of the sorting process.

Sanders and Winkel investigate the use of \emph{predicated}
instructions on Intel Itanium~2 processors \cite{SandersWinkel05}. In
addition to its normal inputs, a predicated instruction takes a
predicate register input. If the value in the predicate register is
true, the instruction takes effect; otherwise it does not. Predication
is normally implemented by the instruction being executed regardless
of whether the predicate is true. However, its result is only
\textit{written back} to a register or memory if the predicate is
true. Although predicated instructions use execution resources
regardless of whether they are allowed to write back, they allow
conditional branches to be eliminated.  Sanders and Winkel show how
the partition step in quicksort can be rewritten to use predicated
instructions on an Itanium~2 highly instruction-level parallel
machine. Itannium~2 provides large numbers of parallel functional
units and registers which make such trade-offs worthwhile.

We attempted to replicate Sanders and Winkel's work a Pentium~4 based
desktop machine, using conditional move instructions (the only form of
predicated instructions on the Pentium~4). The standard form of the
quicksort loops is highly efficient, but we found that rewriting them
into a single loop suitable for predication made them less so. The
transformation also increased the number of simultaneously live
values, which exceeded the number of available registers. We also
discovered that, at least on the Pentium 4, conditional moves are
surprisingly expensive.

Mudge \textit{et al} examine the behaviour of the $i$ and $j$
comparison branches in quicksort, as an example to demonstrate a
general method for estimating the limit on the predictability of
branches in a program \cite{Mudge+96}. They argue that the optimal
predictor for these branches would keep a running count of the total
number of the proportion of array elements examined so far that are
greater than the pivot. They estimate that this approach will give a
prediction accuracy of around 75\%, which is remarkably close to our
result for these branches, using a bimodal predictor.

\textbf{Describe the paper from Alenex 2005}

To our knowlege, this is the only research, other than our own, that
investigates and attempts to understand the predictability of
comparison branches in sorting, rather than simply attempting to
eliminate such branches.

\section{Conclusions}

\comment{
\begin{acks}
We would like to thank the anonymous reviewers of VEE 2005, whose
comments greatly improved an earlier version of this paper.
\end{acks}
}

\bibliographystyle{acmtrans}
\bibliography{report}
%\begin{thebibliography}{...}
%\end{thebibliography}

\begin{received}
...
\end{received}
\end{document}

### 2006-01-20 docs/branch_paper/paper.tex
done initial pictures

@@
 \usepackage{graphicx}
+\usepackage{subfigure}
 
@@
 \newdef{remark}[theorem]{Remark}
+
+\newcommand{\subscale}{0.50}
+\newcommand{\mainscale}{1}
 
@@
 difference of the predictability of the branches in each algorithm.
-Figure \ref{} shows the number of mispredictions per key for both the
+Figure \ref{Predictability-of-insertion-and-selection-sorts} shows the number of mispredictions per key for both the
 algorithms using random data, and varying data set sizes, measured
 using the Simplescalar processor simulator.
+
+\begin{figure}[h]
+\includegraphics[scale=\mainscale]{plots/bpred_misses_insert_select.eps}
+\caption{TODO: Predictability of insertion and selection sorts}
+\label{Predictability-of-insertion-and-selection-sorts}
+\end{figure}
 
@@
 
-Figure \ref{} shows the number of branch mispredictions per key for
+Figure \ref{Predictability-of-bubblesort-and-shakersort} shows the number of branch mispredictions per key for
 bubble- and shakersort.  Nearly one in four comparisons in bubblesort
@@
 misprediction is similar.
+
+\begin{figure}[h]
+\includegraphics[]{plots/bpred_misses_bubble_shaker.eps}
+\caption{TODO: Predictability of bubble and shaker sorts}
+\label{Predictability-of-bubblesort-and-shakersort}
+\end{figure}
 
@@
 
-\begin{figure}
-\includegraphics{plots/two_point_counter_bubblesort.eps}
+\begin{figure}[h]
+\includegraphics{plots/bpred_bubble_progress.eps}
 \caption{Behaviour of branches in bubblesort compared to the ``sweep'' number}
@@
 second-level cache misses caused by our implementation of shellsort
-(see figure \ref{}). In fact, the number of mispredictions caused by
+(see figure \ref{Cache-misses-in-shellsort}). In fact, the number of mispredictions caused by
 our shellsort is significant, but not much greater than other sorting
@@
 
+\begin{figure}[h]
+\subfigure[Caption1]{\includegraphics[scale=\subscale]{plots/cache_misses_level_2_shell.eps}}
+\subfigure[Caption2]{\includegraphics[scale=\subscale]{plots/cache_misses_level_2_shell.eps}}
+\caption{TODO: Predictability of insertion and selection sorts}
+\label{Cache-misses-in-shellsort}
+\end{figure}
+
 \section{Heapsort}
@@
 average and worst case. It begins by creating a structure known as a
-\n{heap}, a binary tree in which a parent is always larger its child,
+heap, a binary tree in which a parent is always larger its child,
 and the largest key is at the top of the heap. It then repeatedly
@@
 uses a technique similar to that used in multi-mergesort, in that a
-\n{multi-way partition} is done at the start, moving all keys into
+multi-way partition is done at the start, moving all keys into
 cache sized containers which can be sorted individually.

### 2007-01-18 docs/branch_paper/paper.tex
Moved nicks paper into the branch_paper directory

--- 2006-01-19 docs/branch_paper/paper.tex
@@
 \author{
-PAUL BIGGAR
-\and
-DAVID GREGG
+PAUL BIGGAR\footnote[1]{PUT IN IRCSET GRANT DETAILS}, NICHOLAS NASH\footnotemark[1] KEVIN WILLIAMS\footnotemark[1] and DAVID GREGG
 \\University of Dublin, Trinity College
@@
 \begin{abstract}
-
+Sorting is one of the most important and well studied problems in Computer Science. Many good algorithms
+are known which offer various trade-offs in efficiency, simplicity, memory use, and other factors. However,
+these algorithms do not take into account features of modern computer architectures that significantly
+influence performance. Caches and branch predictors are two such features, and while
+there has been a significant
+amount of research into the cache performance of general purpose sorting algorithms, 
+there has been little research on their branch prediction properties. In this paper we empirically examine
+the behaviour of the branches in all the most common sorting algorithms. 
+We also consider the interaction of cache optimization on the predictability of the branches in these algorithms.
+We find insertion sort to have the fewest branch mispredictions of any
+comparison-based sorting algorithm, that bubble and shaker sort operate in a fashion which makes their
+branches highly unpredictable, that the unpredictability of shellsort's branches improves
+its caching behaviour and that several cache optimizations have little effect on mergesort's branch
+mispredictions. We find also that optimizations to quicksort -- for example the choice of pivot -- have a strong
+influence on the predictability of its branches. We point out a simple way of removing branch instructions from 
+a classic heapsort implementation, and show also that unrolling a loop in a cache
+optimized heapsort implementation improves the predicitability of its branches. Finally, we note
+that when sorting random data two-level adaptive branch predictors are usually no better than simpler bimodal predictors. This is despite 
+the fact
+that two-level adaptive predictors are almost always superior to bimodal predictors in general.
 \end{abstract}
@@
 Science, University of Dublin, Trinity College, Dublin 2, Ireland. {\tt
-David.Gregg@cd.tcd.ie}.
+David.Gregg@cs.tcd.ie}.
 \end{bottomstuff}
@@
 Classical analyses of algorithms make simplifying assumptions about
-the cost of different machine instructions. For example, Knuth's Mix
-machine code assumes that all operations have unit costs. More
+the cost of different machine instructions. For example, the RAM model used for 
+establishing asymptotic bounds and Knuth's MIX 
+machine code both make drastically simplifying assumptions about the cost of
+machine instructions \cite{KnuthVol1_97}. 
+More
 recently researchers have recognized that on modern computers the cost
@@
 lower level of cache or even main memory.  This has spawned a great
-deal of research on cache-efficient searching and sorting \cite{}.
+deal of research on cache-efficient searching and sorting \cite{Nyberg+94,Agarwal96,LaMarca96b,LaMarca96a,LaMarca97,Xiao+00,Rahman+01,Wickremesinghe+02}.
 
@@
 integer add instruction. If, on the other hand, the branch is
-mispredicted the the processor must flush its pipeline, and restart
+mispredicted the processor must flush its pipeline, and restart
 from the correct target of the branch. This cost is typically a large
 multiple of the cost of executing a correctly-predicted branch. For
-example, the Intel Pentium 4 processor has a 31-stage pipeline,
+example, Intel Pentium 4 processors \cite{Intel248966-010,Intel249438-01} have pipelines of up 
+to 31 stages,
 meaning that a branch misprediction will cost around 30 cycles.  In
 contrast, on the same processor the cost of a memory access that
-misses the first level cache and hits the second level cache is 23
+misses the first-level cache and hits the second level cache is 23
 cycles. Fortunately, the branches in most programs are very
-predictable, so branch mispredictions are usually rare (prediction
-accuracies of greater than 90\% are typical).
+predictable, so branch mispredictions are usually rare. Indeed, prediction
+accuracies of greater than 90\% are typical \cite{Uht+97} with the
+best predictors.
 
 The cost of executing branches is particularly important for sorting
-because the inner loops of most sorting algorithms consist of
-camparisons of items to be sorted. Thus, the predictability of
+because the inner-loops of most sorting algorithms consist of
+comparisons of items to be sorted. Thus, the predictability of
 these comparison branches is critical to the performance of
 sorting algorithms. In this paper we study the predictability
-of branches is most of the major sorting algorithms. We also
-examine cache-conscious variations of some of the algorithms,
-and show how optimizations for the cache impact in branch
+of branches in most of the major sorting algorithms. 
+We focus on the behaviour of the branches whose outcome depends on a
+comparison of keys presented to the sorting algorithm in its input. Throughout this paper we refer to
+such branches as \textit{comparison branches}. Branches associated with controlling simpler aspects 
+of the control flow of the algorithms are of much less interest because they are generally almost perfectly predictable.
+We also examine cache conscious variations of many of the algorithms,
+and show how optimizations for the cache influence the level of branch
 mispredictions.
 
-The rest of this paper is organized as follows.
+\section{Branch prediction}
+\label{branch_prediction}
 
-\section{Branch prediction}
+Branches are a type of instruction used for defining the flow control of programs. In high level languages they result from 
+the use of statements like \texttt{if}, \texttt{while} and their variants. Branches can be either \textit{taken}, indicating that the
+address they provide is the new value for the program counter, or they can be \textit{not-taken}, in which case sequential execution continues
+as though the branch had not been present.
 
-\section{Experimental setup}
+Branch instructions pose a difficulty for pipelined 
+processors. In a pipelined processor, while the program counter points to a particular instruction, a potentially large
+number of subsequent instructions are in a partially completed state.
+Until it is known whether a branch is taken or not-taken, it is not possible to know what the next instructions to be executed are.
+If the processor cannot tell what these next instructions are, it cannot fill its pipeline.
+To keep the utilization of the pipeline at a reasonable level, modern processors attempt to anticipate the outcome of 
+branch instructions by employing \textit{branch predictors}. A good discussion of pipelining and its associated issues can
+be found in Hennessy and Patterson \citeyear{HennessyPatterson90}.
+
+There are several kinds of branch predictors: static, semi-static and dynamic. We use only dynamic predictors in our experiments, 
+because they are most commonly used in real processors. The statistics in the following paragraphs are taken from those derived
+by Uht \textit{et al} \citeyear{Uht+97}, over the SPECint92 benchmarking suite.
+
+The simplest type of dynamic predictor is referred to 
+as a \textit{1-bit} predictor. It keeps a table recording, at each entry, whether a particular branch was taken or not-taken. 
+Using the table, it predicts 
+that a branch
+will go the same way as it went on its previous execution. 1-bit predictors achieve 77\% to 79\% accuracy.
+
+We refer to a 2-bit dynamic predictor as a \textit{bimodal} predictor. A bimodal predictor operates in the same manner as a 1-bit predictor but each table entry
+can be thought of  as maintaining a counter from 0 to 3. The counter decrements on each
+taken branch (with the exception of when the count is 0), and increments on each not-taken branch (with the exception of when the count is 3). With
+counts equal to 0 or 1 the next branch is predicted as taken, with counts of 2 or 3 the branch is predicted not-taken. 
+When the counter is 0, we say the predictor is in the \textit{strongly taken} state, since the branch must be not-taken twice before the not-taken
+direction will be predicted again. When the counter is 1 we say the predictor is in the \textit{taken} state. Similarly counts of 2 and 3 are referred
+to as \textit{not-taken} and \textit{strongly not-taken} respectively. Bimodal predictors achieve 78\% to 89\% accuracy.
+
+Finally, some branch predictors attempt to exploit correlations in branch outcomes to improve accuracy. A \textit{two-level adaptive} predictor maintains
+a \textit{branch history register}. This register records the outcomes of a number of previous branch instructions. For example, a register contents 
+of \texttt{110010} indicates that the 6 previous 
+branch outcomes were taken, taken, not-taken, not-taken, taken and not-taken.
+For each branch, this register is used to index a table of bimodal predictors.
+The program counter can also be included in the index, either concatenated or XORed with the indexing register. Two-level adpative
+predictors are accurate about 93\% of the time.
+
+In all the above, due to the size of the predictor table branches can collide. That is, two 
+different branches can map to the same table location, often reducing accuracy.
+
+\section{Experimental Setup}
+
+We used truly random data provided by Haahr \citeyear{Random} in all our experiments. We used ten chunks of random data containing $2^{22}$ (4194304) keys.
+Where a particular experiment used only part of the keys in a chunk, it used the left-most keys.
+Our results are averaged over these chunks of data.
+
+Processors do not generally provide access to the behaviour of their caches or branch predictors. However, a software processor simulator can be used to 
+provide measurements 
+of cache and branch prediction behaviour.
+Our branch prediction and caching results were generated using the SimpleScalar PISA simulator version 3 \cite{SimpleScalarTutorialv4}. 
+We used \texttt{sim-cache} and 
+\texttt{sim-bpred} to generate results for caching and branch prediction characteristics respectively.
+We used a variety of cache configurations; generally speaking we used an 8 KB level 1 data cache, 8 KB level 1 instruction cache,
+and a shared 2 MB instruction and data level 2 cache, all with 32 byte cache lines. 
+We also experimented with both direct mapped and fully associative caches. When presenting results
+where the cache configuration is important, we will describe the precise configuration used. We present
+results here only for direct mapped caches. The results for fully associative caches are similar \cite{BiggarGregg05}.
+
+For the measurements taken using SimpleScalar, we used power of two sized sets of random keys ranging from $2^{12}$ to $2^{22}$ keys in size.
+This is with the exception of the quadratic time sorting algorithms, for which the maximum set size used was $2^{16}$ keys. This was done
+because of the very significant time it takes such algorithms to sort large inputs.
+
+It takes time to fill arrays, and this can distort the relationship between the results for small set sizes, and larger set sizes, 
+for which this time is amortised. As a result, we measured this time, and subtracted it from our results.
+However, this practice also distorts the results. It removes compulsory cache misses from the results, so long as the data fits in the cache. 
+When the data does not fit in the cache, the keys from the start of the array are faulted, with the effect that capacity misses reoccur, 
+which are not discounted.
+
+For branch prediction we simulated both bimodal and two-level adaptive predictors. 
+We used power of two sized tables for the branch predictors. 
+We used tables containing between $2^{11}$ and $2^{14}$ predictors.
+When
+presenting branch prediction results, the predictor configurations will be described precisely.
+SimpleScalar's \texttt{sim-bpred} provides total results over
+all the branches in a program. It is often desirable to examine particular branches individually however.
+In order to do so, we added our own simulations of branch predictors to the programs. For our 
+own simulations we average the results of each branch over ten runs of random data containing $2^{22}$ keys.
+
+Finally we used \textit{PapiEx} \cite{PapiExManual} to access performance counters on a Pentium 4 1.6 Ghz. This allows cycle counts (i.e. running times)
+to be
+determined for the sorting algorithms on a real processor.
+The results presented from these tests are averaged over 1024 runs and use the system's random number generator to provide data. The data
+ranged in size from $2^{12}$ to $2^{22}$ keys in size.
+The processor we used had an 8-way associative 256KB level 2 
+cache with 64-byte cache lines. The separate data and instruction level 1 caches are 4-way associative, 8KB in size and have 32-byte cache lines.
+The Pentium 4 uses an unspecified form of dynamic branch prediction. A reference manual \cite{Intel248966-010} mentions a branch history
+register, so it is probable that the scheme is similar to two-level adaptive.
 
 \section{Elementary sorts}
-\label{ordernsquared}
-Elementary sorts are also referred to as $O(N^2)$ sorts. These
-algorithms have the worst performance for large inputs. However they
-have two advantages: they are simple to implement correctly and they
-are often faster than more sophisticated algorithms for small input
-sets. Four of these sorts are discussed here: selection sort,
-insertion sort, bubblesort and shakersort.
 
-Selection sort is a simple algorithm that repeatedly finds the
-smallest element of the array and removes it to a sorted section at
-the start. It requires approximately $N^2/2$ comparisons and exactly
-$N$ exchanges. Insertion sort takes a different strategy and instead
-creates an (initially empty) sorted region at the start of the array
-and inserts items into their correct position in this sorted region.
-Insertion sort performs in linear time on a sorted list. With a random
-list, there are, on average, there are $N^2/4$ comparisons and $N^2/4$
-half-exchanges. Thus, selection sort performs around twice as many
-branches as insertion sort.
+We begin by analysing the branch prediction behaviour of a number
+of common elementary sorting algorithms, each operating in
+$O(n^2)$ time. The main practical advantage of these
+algorithms is their simplicity, which generally allows them to
+operate faster than algorithms with $O(n \lg n)$ time complexity
+for small inputs. Indeed, as we will see later, optimized versions
+of $O(n \lg n)$ sorts use calls to elementary sorts for small inputs.
+Knuth \citeyear{KnuthVol3_98} provides a good exposition of these elementary
+sorting algorithms (as well as basic versions of the sorting algorithms
+we shall examine in later sections).
 
-Although we know the approximate number of comparisons that each
-algorithm performs, and important question is whether there is any
-difference of the predictability of the branches in each algorithm.
-Figure \ref{} shows the number of mispredictions per key for both the
-algorithms using random data, and varying data set sizes, measured
-using the Simplescalar processor simulator.
+\subsection{Selection Sort}
+\label{selection_sort}
 
-There are three notable features in this data. Firstly, the number of
-mispredictions is very low. If the behaviour of these branches were
-random, then we would expect that half of the comparison branches
-would be mispredictions, that we would expect $O(N^2)$
-mispredictions. In fact, the number of mispredictions per key is less
-than 12 even for the largest arrays. Secondly, although selection sort
-performs twice as many comparison branches it does not cause twice as
-many branch mispredictions. For the bimodal predictor, it causes 9-12
-times as many mispredictions within the size of input sets that we
-examine. Thirdly the two-level branch predictor is slightly worse at
-predicting the comparison branches in these algorithms than the simple
-bimodal predictor.  This is a remarkable result, because two-level
-predictors are similar to, or sigificantly more accurate than bimodal
-predictors for almost all other types of branches \cite{Uht}.
+Selection sort is a very simple quadratic time sorting algorithm which begins by sequentially searching
+for the smallest key in its input and moving it into position. The algorithm 
+continues by repeatedly searching for and then moving the smallest key in the unsorted
+section of its input into the appropriate position. The inner-loop of selection sort,
+which is run a total of $n - 1$ times on an input \texttt{a[0..n - 1]} is shown below
 
-\subsection{Insertion sort}
-With a bimodal predictor, insertion sort causes almost exactly one
-branch misprediction per key, regardless of the size of the array.
-Thus, although insertion sort performs around $N/4$ comparisons per
-key, only one of these comparisons causes a misprediction. The
-reason for this becomes clear when we consider the inner loop
-of insertion sort:
 \begin{verbatim}
-   while( item < a[j-1] ) {
-     a[j] = a[j-1];
-     j--;
-   }
-   a[j] = item;
+  min = i;
+  for(j = i + 1; j < n; j++)
+    if(a[j] < a[min]) min = j;
+  swap(a[i], a[min]);
 \end{verbatim}
 
-The comparison in insertion sort is also the loop guard.  As long as
-the loop continues, this comparison always resolves in the same
-direction, and is thus highly predictable. When the location for the
-item is finally found, the comparison resolves in the opposite
-direction, and a branch misprediction occurs. Thus, for a bimodal
-predictor, the number of mispredictions per key is almost exactly
-one. Although the results we present are for random data, the same
-is true regardless of the input data. Insertion sort will always
-cause around one misprediction per key.
+It is easy to see that selection sort performs approximately $n^2/2$ comparisons
+and exactly $n - 1$ exchanges. 
+On the $j^{th}$ iteration of the inner-loop the comparison branch is taken
+if \texttt{a[j]} is the minimum of \texttt{a[i..j]}, which occurs with
+probability $1 / (j - i + 1)$\footnote{Throughout this paper, for the purpose of analysis we make
+the usual assumption that the input to be sorted consists of a random permutation of $\lbrace1, .., n\rbrace$.}. 
+The expected number of times the branch is taken
+on the first iteration is thus given by $H_n - 1$ where 
+$H_n = \sum_{i = 1}^n 1/i$ is the $n^{th}$ harmonic number \cite{KnuthVol1_97}.
+Since $H_n \approx \ln n$, and $\sum_{i = 1}^n H_i = O(n \ln n)$ the comparison branch
+will be highly predictable. As expected, Figure \ref{elementary_sorts}(a) shows that number of mispredictions
+per key grows in a similar fashion to $\ln n$.
 
-The two-level predictor combines a pattern of recent branch outcomes
-make a prediction. This goal is to exploit correlations between branch
-outcomes. For most programs this is valuable, but there is correlation
-of branch outcomes in insertion sort. The additional complexity simply
-causes the predictor to find false correlations, with the result that
-number of mispredictions is higher than for a simpler, bimodal
-predictor. In general, misprediction rates of more than one per key
-are caused by imperfections in the predictor, rather than the
-unpredictability of the comparison branch in insertion sort.
-
-\subsection{Selection sort}
-The comparison branch in selection sort is also highly predictable,
-but less so than for insertion sort. The inner loop of selection
-sort consists of searching for the smallest element of part of
-the array:
-\begin{verbatim}
-  min = i;
-  for ( j = i+1; j < N; j++ )
-    if ( a[j] < a[min] ) min = j;
-\end{verbatim}
-The predictability of the comparison branch depends on the number of
-times that the cadidate minimum, $min$, changes. Knuth \cite{Knuth97}
-shows that for $N$ elements, the number of changes is $H_N - 1$, where
-$H_N$ is the harmonic series:
-$$H_N = 1 + \frac{1}{2} + \frac{1}{3} + ... + \frac{1}{N} =  \sum_{k=1}^{N}
-\frac{1}{k},  \hspace{0.4cm}  N \geq 1$$
-
-Within practical sorting ranges, $H_N$ grows at a rate similar to
-$log_e(N)$. Thus, in the average case, we can expect that the comparison
-branch will mostly go in the same direction, making the branch highly
-predictable.
-
-The number of changes to the left-to-right minimum in the best case is
-zero and in the worst case $N-1$. In the worst case, the array is
-sorted in reverse order, and so the left-to-right minimum changes on
-every element. Interestingly, the worst cases for the number of
-changes is not the worst case for branch mispredictions. If the
-minimum changes for every element, then the comparison branch will
-always resolve in the same direction, and thus be almost perfectly
+If the array is initially in sorted order, then the value of \texttt{min}
+will never change in the inner-loop. Since this implies the comparison branch
+is never taken, it will be almost perfectly predictable. Similarly, 
+if the array is sorted in reverse order, the value of \texttt{min} must
+change at every element. That is, the comparison branch 
+is taken at every element, and the branch is therefore almost perfectly
 predictable.  From the point of view of branch prediction, frequent
@@
 branch resolves in each direction. The worst case for selection sort
-would be an array which is partially sorted in reverse order; the
-left-to-right minimum would change frequently, but not predictably so.
+would be an array which is partially sorted in reverse order; the value of
+\texttt{min} would change frequently, but not predictably so.
+Figure \ref{elementary_sorts}(a) shows the behaviour of selection
+sort's branches.
+
+\subsection{Insertion Sort}
+\label{insertion_sort}
+
+Insertion sort is another simple algorithm with quadratic
+worst case time. It works by iterating the inner-loop below
+a total of $n - 1$ times over its input
+
+\begin{verbatim}
+   item = a[i];
+   while(item < a[i - 1]) 
+   {
+     a[i] = a[i - 1];
+     i--;
+   }
+   a[i] = item;
+\end{verbatim}
+On the $i^{th}$ iteration \texttt{a[0..i - 1]} is sorted, allowing
+\texttt{a[i]} to be placed in the correct position as shown in
+the inner-loop above. On average, insertion sort performs approximately
+$n^2/4$ comparisons and $n^2/4$ assignments in its inner-loop. Thus insertion sort
+performs around half as many branches as selection sort.
+
+A very pleasing property of insertion sort is that
+it generally causes just 
+a single branch misprediction per key, as Figure \ref{elementary_sorts}(a) shows. Typically
+the inner \texttt{while} loop exit causes this misprediction, when
+the correct location for the current item has been found.
+In fact, a static
+predictor would perform just as well as the two predictors shown, since it
+would mispredict exactly once per key.
+
+\begin{figure}
+\centering
+\begin{tabular}{cc}
+\includegraphics[width=0.5\textwidth]{plots/selection_insertion_sort_branches.eps} & \includegraphics[width=0.5\textwidth]{plots/bubble_shaker_sort_branches.eps} \\
+(a) & (b) \\
+\end{tabular}
+\caption{(a) Shows the branch mispredictions for selection and insertion sort using bimodal and two-level adaptive predictors both having 4096 table entries.
+It is noteworthy the simple bimodal predictor has essentially identical performance to the two-level adaptive predictors.
+(b) Shows the branch mispredictions for bubble and shaker sort using bimodal and two-level adaptive predictors, again both having 4096 table entries.
+For bubble sort the bimodal and two-level adaptive predictors have close to identical performance. However, for shaker sort the simple bimodal
+predictor results in fewer mispredictions than the two-level adaptive predictor. We also note that as a result of 
+the partial sorting performed by bubble and shaker sort they incur far more
+branch mispredictions per key than other $O(n^2)$ sorting algorithms.
+\texttt{sim-bpred} was used to simulate the branch predictors in both (a) and (b). The two-level adaptive predictors used 10-bit history registers.}
+\label{elementary_sorts}
+\end{figure}
 
 \subsection{Bubble sort}
-Bubblesort is a notoriously inefficient algorithm that compares and
-exchanges adjacent pairs of elements in the array. One execution of
-the inner loop will move the smallest element of the array to the
-first position, and partially sort the rest of the array by moving
-small elements towards the start. Cocktail shaker sort (or simply
-shakersort) alternates the direction of the inner loop to take
-more advantage of this partial sorting to reduce the number of
-outer-loop iterations.
+\label{bubble_sort}
 
-Figure \ref{} shows the number of branch mispredictions per key for
-bubble- and shakersort.  Nearly one in four comparisons in bubblesort
-are mispredicted. As the data sets gets large, there are nearly 10000
-branch mispredictions per key key. Shakersort executes a lot fewer
+Bubble sort is another simple sorting algorithm, although notoriously
+inefficient. It works by iterating the loop shown below at most $n - 1$
+times on an input \texttt{a[0..n - 1]}.
+
+\begin{verbatim}
+for(j = 0; j < n - i - 1; j++)
+{
+    if(a[j + 1] < a[j])
+       swap(a[j + 1], a[j]);
+}
+\end{verbatim} 
+In this code $i$ is the outer-loop iteration counter, and begins at 0 and counts
+towards $n - 1$. Bubble sort uses
+information yielded by the comparisons which selection
+sort simply discards. After a full iteration of the inner-loop shown larger
+elements have moved to the right, closer to their final location.
+
+Shaker sort is a variation of bubble sort in which there are two inner-loops
+which are alternated. One is the inner-loop shown above. The other
+inner-loop scans right-to-left moving small elements to the left in the
+same manner as moving larger elements to the right has just been described.
+Although shaker sort has the same worst-case time as bubble sort, it is generally
+more efficient in practice.
+
+Figure \ref{elementary_sorts}(b) shows the number of branch mispredictions per key for
+bubble and shaker sort. As the data set gets large, bubble sort causes nearly 10000
+branch mispredictions per key. Shaker sort executes a lot fewer
 branches, and so has fewer mispredictions, but the rate of
@@
 
-The branch behaviour of bubblesort is particularly interesting when
-compared to selection sort. Although we usually think of the two
-algorithms as being different, they are actually very similar. Like
-selection sort, bubblesort finds the smallest key in the
-unsorted part of the array and moves it to start of the
-array. Thus, we would expect that bubblesort's branch prediction
-behaviour would be similar to that of selection sort.
+To understand the high misprediction rate of bubble sort we define $Q^k_j$
+as the probability that $j^{th}$ inner-loop comparison branch is taken on the $k^{th}$ 
+outer-loop iteration, $1 \leq k < n$. It turns out that
 
-However, in the process of moving the smallest element into position,
-bubblesort also rearranges other array elements. Selection sort sweeps
-across the array and always keeps track of the left-to-right minimum
-up to the current point.  However, bubblesort actually moves this
-minimum, exchanging pairwise with each element it encounters which is
-larger. Thus, even the unsorted part of the array becomes partially
-sorted. We might expect that bubblesort's comparison branch for a
-partially sorted array might be more predictable than for a random
-one, but it is not. With a random array, the smaller elements are
-equally likely to appear anywhere in the array. With a partially
-sorted array, the smaller elements tend to appear towards the start of
-the array which is, in bubblesort, the last place that we look. Thus,
-the candidate minimum is changed much more often than if we were
-searching a randomly ordered array.
+\begin{equation}
+\label{bubble_equation}
+Q^k_j = 
+\left\{ 
+\begin{array}{ll}
+  \frac{j}{j + k} & \quad \mbox{if $j \leq n - k$}\\
+  0 & \quad \mbox{otherwise}\\ 
+\end{array} 
+\right. 
+\end{equation}
+
+\noindent
+This is a result of the fact that the $j^{th}$ comparison branch of the inner-loop
+is taken only if \texttt{a[j]} is not larger than all of \texttt{a[0..j - 1]}.
+For this
+to be the case, the $(j + 1)^{th}$ branch in the previous outer-loop iteration must have been taken,
+since otherwise the previous iteration of the outer-loop left the maximum of \texttt{a[0..j - 1]} in \texttt{a[j]}.
+The probability that the $(j + 1)^{th}$ branch of the
+previous outer-loop iteration is taken is $Q^{k - 1}_{j + 1}$. Given that the $(j + 1)^{th}$ branch of
+the previous outer-loop iteration was taken, the probability that \texttt{a[j]} is
+not larger than all of \texttt{a[0..j - 1]} is $1 - 1/(j + 1)$, thus,
+
+\[
+Q^k_j = Q^{k-1}_{j+1}\left(1 - \frac{1}{j + 1} \right)
+\]
+
+\noindent
+The bases $Q^{1}_j = 1 - 1 / (j + 1)$ for $1 \leq j < n$ and $Q^1_n = 0$ give the solution shown in Equation \ref{bubble_equation}.
+It is easy to see that in the first outer-loop iteration ($k = 1$) the comparison branch is very likely to be taken, whereas in the last
+outer-loop iteration ($k = n - 1$), the comparison branch is very unlikely to be taken. Moving between these two extremes causes the intermediate
+branches to be unpredictable. The remainder of this section examines this phenomenon in more detail.
+
+In order to analyse the behaviour of bubble sort when a bimodal predictor is used,
+it is useful to make use of the \textit{steady state predictability} function. This function
+generally gives a good estimate of the probability that a particular branch will be correctly predicted,
+given that we know the probability that the branch will be taken. For a bimodal predictor this function is
+given by
+
+\begin{equation}
+CP(p) = \frac{3p^2 - 3p + 1}{2p^2 - 2p + 1}
+\label{CP_equation}
+\end{equation}
+
+\noindent
+Where $p$ is the probability that the branch in question will be taken. Although the derivation of this
+function is simple, we relegate it to the appendix because we wish
+to focus mainly on our experimental results. $CP$ can be used to examine the predictability of the branches in bubble sort as the outer-loop iterates.
+Bubble sort executes $n - k$ comparison branches on its $k^{th}$ outer-loop iteration. Thus,
+using Equation \ref{bubble_equation} the average probability that the comparison branch is taken on the $k^{th}$ outer-loop iteration
+is
+
+\begin{eqnarray*}
+Q^k_{avg} &=& \frac{1}{n - k} \sum_{j = 1}^{n - k} \frac{j}{j + k} \\
+          &=& 1 + \frac{k}{n - k}(H_k - H_n)\\
+%  &=& \frac{n - k - k(H_n - H_k)}{n - k}\\
+         &\approx& 1 + \frac{k}{n - k}\ln\frac{k}{n}
+\end{eqnarray*}
+
+\noindent
+Where $H_n$ denotes the $n^{th}$ harmonic number. An approximation to the predictability
+of the comparison branch of bubble sort on the $k^{th}$ iteration is then given by $CP(Q^k_{avg})$.
+Figure \ref{bubble_predictability}(a) plots this, showing how the predictability of the comparison branch varies
+with the number of outer-loop iterations. Figure \ref{bubble_predictability}(b)
+shows the measured number of correctly predicted branches as the outer-loop iterates, these results
+correspond approximately with what was obtained analytically in Figure \ref{bubble_predictability}(a).
 
 \begin{figure}
-\includegraphics{plots/two_point_counter_bubblesort.eps}
-\caption{Behaviour of branches in bubblesort compared to the ``sweep'' number}
-\label{Predictability-of-branches-in-bubblesort-compared-to-sweep-number}
+\centering
+\begin{tabular}{cc}
+\includegraphics[width=0.5\textwidth]{plots/bubble_predictability.eps} & \includegraphics[width=0.5\textwidth]{plots/bubble_sweep_misses.eps} \\
+(a) & (b) \\
+\end{tabular}
+\caption{(a) Shows $CP(Q^k_{avg})$ as the sweep number, $k$, increases (i.e. as the outer-loop iterates). This gives a good approximation to the 
+predictability of the comparison branch of bubble sort. (b) Shows the number of correctly predicted branches in bubble sort, as well as the number of executed branches.
+Initially the number of correctly predicted branches is close to the number executed, however, the number of correctly predicted branches rapidly 
+decreases as the sweep number increases. Gradually, the branches become predictable again. These trends correspond to the 
+variation in predictability shown in (a). The results of (b)
+were obtained using our own software simulated bimodal predictor averaged over a large number of arrays with 8192 elements. Note also that in (b) 
+the version of bubble sort used does not terminate early if no swaps are performed for an entire execution of the inner-loop.  }
+\label{bubble_predictability}
 \end{figure}
 
-To demonstrate this effect, we measured how the predictability of
-bubblesort's comparison branch varies with successive sweeps across
-the array. We added a simple software simulation of a bimodal
-predictor to the source code of base bubblesort, and ran it on a large
-number of randomly generated arrays, each with 8192 elements.
+Initially the branches
+are quite predictable, however their predictability reduces quite rapidly as the outer-loop iterates.
+The incremental movement of large keys to the right in bubble sort causes this degredation in the
+predictability of its branches. As the data becomes close to being
+fully sorted the predictability of the branches improves again. The partial sorting bubble sort performs allows it to potentially converge to 
+fully sorted data in fewer outer-loop iterations than selection sort. We can
+detect that the data is sorted early if the branch shown in the inner-loop above
+is never taken for a whole iteration. However, this partial sorting also greatly increases
+the number of branch mispredictions bubble sort incurs, substantially slowing it over selection sort or insertion sort.
 
-Figure
-\ref{Predictability-of-branches-in-bubblesort-compared-to-sweep-number}
-shows that on the early iterations the number of correct predictions
-is very close to the number of executed comparison branches. The
-comparison branch is highly predictable because bubblesort's search
-for the smallest element is similar to selection sort. However, with
-successive sweeps even the unsorted part of the array becomes
-increasingly sorted and the branch becomes more unpredictable. We see
-that the number of executed comparison branches falls linearly, but
-that the number of correctly predicted comparison branches falls much
-more quickly. Eventually, the array becomes so close to fully sorted
-that the branch becomes increasingly predictable in the other
-direction, and the number of correct predictions again approaches the
-number of executed branches. However, during the intermediate stage,
-a huge number of branch mispredictions have already
-occurred. Paradoxically, by partially sorting the array, bubblesort
-greatly increases the number of branch mispredictions and thus greatly
-increases its running time.
+\subsection{Remarks}
 
-Overall, our results show that branch prediction is an important
-factor in the efficiency of sorting algorithms. For example,
-shakersort executes fewer machine instructions and causes fewer cache
-misses than selection sort, but is slower due to branch
-mispredictions. Insertion sort is confirmed as the most efficient
-elementary sorting algorithm. In additon to having the best time
-complexity it causes only around one branch misprediction per key even
-in the worst case.
+Our results clearly demonstrate the importance of branch
+prediction for the elementary sorting algorithms. 
+Our shaker sort implementation has fewer cache misses than our selection sort implementation, as Figure \ref{shaker_plots}(a) shows. 
+It also has a lower instruction count than selection sort. However it is slower due to branch mispredictions, as Figure \ref{shaker_plots}(b) shows.
+Figure \ref{elementary_sorts} shows
+the contrast between the branch prediction performance of selection sort and bubble or shaker sort. Where
+selection sort causes on average fewer than 12 mispredictions per key for a set
+size of 65536 keys, bubble and shaker sort cause many thousands.
+Insertion sort behaves even better than selection sort, causing
+almost a single misprediction per key. Since insertion sort
+also has the fewest cache misses it has much to recommend it
+as an elementary sorting algorithm. It is bubble sort's
+partial sorting behaviour that wreaks havoc with branch prediction,
+adding further weight to Knuth's \citeyear{KnuthVol3_98} point of view that it
+has very little to recommend it, besides a catchy name and some interesting theoretical problems it
+poses.
+
+Finally our results show that over these elementary sorting algorithms two-level adaptive branch predictors are no
+better than simpler bimodal predictors. 
+This is noteworthy, because two-level
+predictors are similar to, or sigificantly more accurate than bimodal
+predictors for almost all other types of branches \cite{Uht+97}.
+In fact, a bimodal predictor out-performs a two-level adaptive predictor for shaker sort, as
+is shown in Figure \ref{elementary_sorts}(b).
+
+\begin{figure}
+\centering
+\begin{tabular}{cc}
+\includegraphics[width=0.5\textwidth]{plots/elementary_cache1_misses.eps} & \includegraphics[width=0.5\textwidth]{plots/elementary_cycles.eps}\\
+(a) & (b) \\
+\end{tabular}
+\caption{(a) Shows the level 1 cache misses of the elementary sorting algorithms. 
+These results were obtained by using \texttt{sim-cache} to simulate an 8KB direct mapped 
+level 1 cache with 32 byte cache lines. Note that there are no level 2 cache misses since the simulated 2nd
+level cache is 2MB in size.
+(b) Shows the number of cycles per key for the elementary sorting algorithms. 
+Despite bubble and shaker sort having better cache performance than selection sort, the vast number of branch mispredictions 
+(see Figure \ref{elementary_sorts}(b)) they cause make them less efficient than selection sort in practice.
+These results were measured using hardware performance counters on a Pentium 4.}
+\label{shaker_plots}
+\end{figure}
 
 \section{Shellsort}
-Shellsort \cite{Shell59} was the first in-place sub-$N^2$ sorting
-algorithm. The algorithm operates by sorting all elements that are $h$
-elements apart using insertion, where $h$ is some increment.  On early
-iterations the increment is large, allowing items to move large
-distances. The increment is reduced on successive iterations to allow
-more local sorting. The final iteration is a standard insertion sort,
-after which the array is sorted. We use Gonnet's increments, which set
-the next value of the increment to $\frac{5}{11}^{th}s$ of the
-previous increment, because we found them to be efficient in practice.
-All discussions of complexity are based on these increments.
 
-The branch performance of shellsort should be predictable. There are
-$log_{(\frac{11}{5})}N$ increments, each with a separate pass over the
-array. On each pass, all keys are insertion sorted into some
-sub-array. Given that insertion sort causes around one branch
-misprediction per key, we can expect $N$ comparison branch
-mispredictions per pass. It is expected, therefore, that there will be
-around $log_{(\frac{11}{5})}N$ mispredictions per key. In general, if
-there are $i$ increments, shellsort will cause $ni$ mispredictions.
+Shellsort \cite{Shell59} was the first in-place sorting algorithm
+with time complexity better than $O(n^2)$.
+Algorithms like selection and bubble sort use each comparison
+to resolve at most one inversion (an inversion is 
+a pair of keys in the input which appear in the wrong order) in the input, and
+it is well known that this restricts their average case running
+time to be in $O(n^2)$. Shellsort uses each comparison to
+resolve more than one inversion, giving better performance.
 
-Figure \ref{} shows the number of branches and branch mispredictions
-per key for shellsort. For 4194304 keys, we would expect
-$log_{(\frac{11}{5})}4194304 = 18.34$ mispredictions per key. The
-number of mispredictions we measure in our simulation is actually a
-little under 18, which is remarkably close to the predicted number.
+Shellsort uses a sequence of decreasing integers $h_1, h_2, \ldots, h_k$
+with $h_k = 1$, called increments. 
+On the $i^{th}$ of its $k$ iterations, shellsort performs
+an insertion sort but treats \texttt{a[j]} and \texttt{a[j + $h_i$]} as though
+they were adjacent (rather than a normal insertion sort, which treats \texttt{a[j]} 
+and \texttt{a[j + 1]} as adjacent).
 
-To further investigate this we separated the comparison branches from
-the other branches in the code in a separate simulation of a bimodal
-branch predictor on a large number of randomly generated arrays. We
-found that using Gonnet's increments each key moves only an average of
-0.9744 places in each insertion sort (although moving one place
-actually involves skipping over $h$ other array elements, where $h$ is
-the increment). Thus, the situation is somewhat different to insertion
-sorting large arrays of random elements where each element can be
-expected to move a large number of places. Instead of the comparison
-branch almost always resolving in one direction, the actual outcome
-will be close to random. We can still expect a number of
-mispredictions that is close to one per key, but we are unlikely to
-get the almost perfectly predictable behaviour that we see when using
-insertion sort where the average number of places moved is larger.
+The choice of increments influences shellsort's asymptotic behaviour.
+We used Gonnet's increments \cite{Gonnet+91} for our implementation, because
+we found them to be efficient empirically. Using
+Gonnet's increments $h_i = \lfloor \frac{5}{11} h_{i - 1} \rfloor$, with $h_1 = \lfloor \frac{5}{11}n \rfloor$ and the
+final increment being 1 as usual.
 
-The main apparent weakness of shellsort on modern architectures is its
-poor locality of reference in memory access.  By comparing items which
-are separated by $h$ keys, there is little reuse of items in the same
-cache line. One can easily imagine cases where shellsort would cause
-very large numbers of cache misses.  In the worst case, each key would
-move a very large number of places in each insertion sort pass,
-possibly causing a cache miss for every place that it moves.  To
-further investigate this effect, we measured the number of first and
-second-level cache misses caused by our implementation of shellsort
-(see figure \ref{}). In fact, the number of mispredictions caused by
-our shellsort is significant, but not much greater than other sorting
-algorithms such as heapsort and merge sort (see sections \ref{} and
-\ref{}). The main reason is that using Gonnet's increments on average
-keys move only 0.9744 places on each pass, so very large numbers of
-references in diverse parts of the array are not needed.
+We use a straightforward implementation of shellsort, except we perform the final iteration
+(i.e. the iteration in which we perform a standard insertion sort because the increment is 1) separately. 
+For this final insertion sort
+we extract $\min(a[0], \ldots, a[9])$ as a sentinel, removing the need
+for a bounds check in its inner-loop. This also simplifies
+the bounds checks on the outer-most loop of the other iterations.
+
+\subsection{Branch Prediction Results}
+
+\begin{figure}
+\centering
+\begin{tabular}{cc}
+\includegraphics[width=0.5\textwidth]{plots/shellsort_branch_misses.eps} & \includegraphics[width=0.5\textwidth]{plots/shell_heap_cache_misses.eps} \\
+(a) & (b) \\
+\end{tabular}
+\caption{(a) Shows the average branches misses per key for shellsort using a bimodal and two-level adaptive predictor, both having 4096 table entries.
+On this occasion, the two-level adaptive predictor out-performs the simpler bimodal predictor. \texttt{sim-bpred}
+was used to simulate these branch predictors.
+(b) Shows that the average number of cache misses per key of our basic heapsort implementation
+is higher than for our basic shellsort implementation. Shellsort incurs only a small number of cache misses
+because on average a key moves just 0.9744 places per iteration. These results use \texttt{sim-cache} to simulate
+an 8KB direct mapped cache with 32 byte cache lines.}
+\label{shellsort_results_figs}
+\end{figure}
+
+The behaviour of the comparison branches in shellsort is
+simple to analyse. There are $\lfloor \log_\frac{11}{5} n \rfloor$ passes, with
+each pass performing an insertion sort (using the current
+increment). As we saw in Section \ref{insertion_sort} insertion
+sort causes about one branch misprediction per key, thus
+we expect about $n \lfloor \log_\frac{11}{5}n \rfloor$ mispredictions in total, and
+about $\lfloor \log_\frac{11}{5}n \rfloor$ mispredictions per key.
+
+Figure \ref{shellsort_results_figs}(a)
+shows the branch mispredictions per key on average for our shellsort
+implementations, with bimodal and two-level adaptive predictors.
+The scale on the x-axis is logarithmic, and the plots are nearly
+straight lines, as we would expect if the branch mispredictions per key
+are given by $\lfloor \log_\frac{11}{5} n \rfloor$. For example,
+the final measurement on the x-axis is for a set size of 4194304 and
+here we observe from the plot approximately 18 misses per key on average,
+and $\lfloor \log_\frac{11}{5} 4194304 \rfloor = 19$.
+
+The comparison branch of the insertion sort inner-loop of shellsort are very unpredictable, unlike the 
+nearly perfectly predictable branches of a standard insertion sort.
+We measured a predictability of about 50\% on average for this branch using a simulated bimodal
+predictor.
+A static predictor which predicts all branches as taken would be ideal for predicting the branches of the
+insertion sorts performed by shellsort.
+We found that on average, using Gonnet's increments, a key moves 
+0.9744 places per iteration (note that a move is of $h$ positions, 
+where $h$ is the current increment). Thus, although the insertion
+sort branch will still cause just a single misprediction per key, 
+its outcome is not nearly as predictable as a standard insertion
+sort, where keys on average are moved a larger number of times
+per inner-loop iteration.
+
+One might expect shellsort's cache performance to be poor, since it repeatedly accesses
+distantly separated keys in its input while the increments are large. Each such access
+could potentially cause a cache miss. However, the fact that on average 
+keys move just 0.9744 places per iteration
+implies that large numbers of references in distant parts
+of the input (which could lead to cache misses) are not generally necessary.
+In fact, the number of cache misses for our shellsort implementations, although significant,
+is smaller than for our basic heapsort implementation as Figure \ref{shellsort_results_figs}(b)
+shows. For our input sizes, which were up to $2^{22}$ keys, our shellsort implementations were on
+average about 50\% faster than the best of our heapsort implementations (see Section \ref{heapsort} for
+further details on our heapsort implementations).
 
 \section{Heapsort}
-Heapsort is an in-place sort which works in $O(NlogN)$ time in the
-average and worst case. It begins by creating a structure known as a
-\n{heap}, a binary tree in which a parent is always larger its child,
-and the largest key is at the top of the heap. It then repeatedly
-removes the largest key from the top and places it in its final
-position in the array, fixing the heap as it goes.
+\label{heapsort}
+
+\begin{figure}
+\centering
+\begin{tabular}{cc}
+\includegraphics[width=0.5\textwidth]{plots/heapsort_branch_counts.eps} & \includegraphics[width=0.5\textwidth]{plots/heapsort_branch_misses.eps}\\
+(a) & (b) \\
+\includegraphics[width=0.5\textwidth]{plots/heapsort_cycle_counts.eps} & \includegraphics[width=0.5\textwidth]{plots/heapsort_8heap_branches.eps} \\
+(c) & (d) \\
+\end{tabular}
+\caption{Branch prediction and performance characteristics of our heapsort implementations. (a) Shows
+the average number of branches per key measured using \texttt{sim-bpred}. (b) Shows the average branch mispredictions per key for tables of 4096
+predictors, again measured using \texttt{sim-bpred}. The two-level adaptive (which used a 10-bit history register) and bimodal predictors have close to identical performance, and to avoid
+cluttering the diagram only one instance of a two-level adaptive predictor is shown, the remainder of the predictors are bimodal. (c) Shows the cycle counts of the algorithms measured using
+Pentium 4 hardware performance counters. Note that the Pentium 4 has 32 byte cache lines in its first level cache, and 64 byte cache lines in its
+second level cache.
+Finally (d) shows the predictability of the unrolled branches associated with an 8-heap, measured with our own software simulated bimodal predictor.}
+\label{heapsort_results}
+\end{figure}
+
+Another well known general purpose sorting algorithm is heapsort \cite{Williams64}. Heapsort's running time is $O(n \lg n)$.
+Heapsort begins by constructing a \textit{heap}: a binary tree in which every level except possibly the deepest is entirely filled, with the deepest level filled from the
+left. In addition, a heap must also satisfy the \textit{heap property}: A key is associated with each node that is always at least as large as the keys 
+associated with all its children nodes. The sequence of keys resulting from a pre-order traversal of a heap's nodes can be used to represent it unambiguously in an array.
+
+An unordered array can be transformed into a heap in $O(n)$ time, using the approach of Floyd \citeyear{Floyd64}.
+Given a node whose children are heaps but who may itself
+violate the heap property, an iteration of Floyd's approach swaps the node with the larger of its children, and then repeats the same procedure 
+with the subtree rooted at that child, until the heap property is satisfied or a leaf is reached. We refer to this as ``sifting down'' a node. 
+To build the entire heap, we first sift-down the deepest nodes having children, followed by their
+parents, and so on until we reach the root.
+
+Once the heap is built, we gradually destroy it in order to sort. We iterate the code below $n - 1$ times.
+
+\begin{verbatim}
+    tmp = a[0];
+    a[0] = a[--n];
+    a[n] = tmp;    
+    sift_down(a, 0, n);
+\end{verbatim}
+
+\noindent
+On the $i^{th}$ iteration we swap the largest key, 
+\texttt{a[0]}, with \texttt{a[n - i]}. 
+After each swap, the heap property is restored by sifting down the new root of the heap (an operation with worst case time $O(\lg n)$). 
+After iteration $i$ the heap contains $n - i$ keys, and is found in \texttt{a[0..n - i - 1]}. Meanwhile 
+\texttt{a[n - i..n - 1]} contains the $i$ largest keys of the input from smallest to largest.
+
+\subsection{Heapsort Variations}
+
+For our base heapsort implementation, we used the approach described in the previous section, but removed a bounds check from the sift down procedure
+by introducing a sentinel at the cost of no extra instructions. The following optimization reduced the instruction count by about 10\%
+on average compared to an implementation of sift-down which uses a bounds check. We have not seen it described elsewhere. In a heap, exactly one node has a single left child, all 
+other nodes have either two children
+or zero children. On every sift-down, we compare each node which has children with each of its children. As a result of the node with a single child, on 
+every sift-down we
+must check that the node being compared to its children has a right child as well as a left child. To avoid this check while building the heap, we can just 
+insert
+a maximal extra node at the end of the heap. When destroying the heap however, every second iteration requires the insertion of 
+this extra node. To avoid this the code for destroying the heap shown in the previous section is changed to
+
+\begin{verbatim}
+    tmp = a[0];
+    a[0] = a[--n];
+    sift_down(a, 0, n);
+    a[n] = tmp;
+\end{verbatim}
+
+\noindent
+By not modifying \texttt{a[n - 1]} until after sifting down,
+it acts as a sentinel for sift-down. This removes the need for the bounds check without introducing any extra instructions.
+
+Aside from this optimization, we also applied caching optimizations of LaMarca and Ladner \citeyear{LaMarca96b} to give several cache-optimized versions of the
+algorithm. Firstly, the heap can be padded so that siblings
+always reside in the same cache line. Another cache optimization is realised by abandoning Floyd's heap building method and simply repeatedly growing the
+heap one array element at a time, sifting each new key upwards as needed. This greatly improves the temporal locality of the algorithm when the data
+does not fit entirely within the cache.
+
+To increase the utilization of cache lines (and hence require fewer cache lines to be
+fetched) we can use a $k$-ary heap instead of the binary heap that has been assumed until now. When using a $k$-ary heap, every node having children (except perhaps one)
+has $k$ children. When sifting up or down a $k$-ary heap, the maximum (or minimum, depending on the heap property being used) child must be found, which
+requires $k - 1$ comparisons. Of course, as $k$ grows the sort degenerates to a selection sort. Note also that for a $k$-ary heap, the optimization described
+above, which avoids bounds checking while sifting down cannot be applied. However, it is still more efficient to add a sentinel on every second iteration of the heap 
+destruction (since this introduces
+$O(n)$ operations whereas the bounds check in the inner-loop introduces $O(n \lg n)$ operations). For modest values of $k$ a performance improvement
+can be expected over a simple binary heap \cite{LaMarca96b}. We experimented with 4-heaps and 8-heaps, combined with the aforementioned cache optimizations.
+
+\subsection{Branch Prediction Results}
+\label{heapsort_branch_results_text}
+
+Figure \ref{heapsort_results}(a) shows the average number of branches per key for our heapsort implementations. We note that the cache-optimized heapsort
+implementations execute fewer branches per key than our base heapsort implementation. Although these algorithms execute fewer branches
+per key on average, they cause a larger total number of branch mispredictions than our base heapsort implementation. Indeed, Figure
+\ref{heapsort_results}(b) shows that the base heapsort implementation incurs the fewest branch mispredictions of any of the algorithms. 
+This is despite the fact that it is executing more branches on average.
+Figure \ref{heapsort_results}(b) also shows that the simpler
+bimodal predictors are generally no worse at predicting branches than the two-level adaptive predictors. With the exception of shellsort, this is in 
+keeping with the trend we observed for 
+the other sorting algorithms examined so far.
+
+Notwithstanding the base heapsort implementation having the lowest misprediction rate,
+Figure \ref{heapsort_results}(c) shows that the most efficient implementation on our hardware configuration to be the 8-heap variation of heapsort,
+followed by the 4-heap variation.
+This suggests the improved cache performance of these algorithms outweighs their increased level of branch mispredictions shown in 
+Figure \ref{heapsort_results}(b).
+
+When using a $k$-ary heap, finding the maximum (or minimum) child at a particular node causes the $i^{th}$ comparison branch to be taken
+with probability $1/(1 + i)$. The comparisons behave just like those in selection sort (see Section \ref{selection_sort}), and are therefore quite
+predictable.  For our 8-heap we unrolled the
+loop to find the minimum child into 7 separate branches. 
+Figure \ref{heapsort_results}(d) shows the behaviour of these 7 branches, for an unrolled loop.
+As expected, the later branches are gradually taken less often on average, with a corresponding increase in their predictability.
+Unrolling this loop
+not only removes the loop over-head, but is also experimentally observed to improve the average prediction accuracy by about 2\%. 
+The average steady state predicitability in the unrolled case, is given by
+
+\[
+\frac{1}{7}\sum^7_{i=1}CP\left(\frac{1}{1 + i}\right) \approx 0.724
+\]
+
+\noindent
+Where $CP$ is the steady state predictability function introduced in Section \ref{bubble_sort}.
+Without unrolling, the comparison branch is taken on average with probability $\frac{1}{7}\sum^7_{i=1}\frac{1}{i + 1}$, and its predictability
+can be estimated as
+
+\[
+CP\left(\frac{1}{7}\sum^7_{i=1}\frac{1}{1 + i} \right) \approx 0.706
+\]
+
+\noindent
+As is observed experimentally, the unrolled loop results in about a 2\% greater prediction accuracy.
+
+\section{Mergesort}
+\label{mergesort}
+
+Mergesort \cite{KnuthVol3_98} is a another $O(n \lg n)$ sorting algorithm, which works by 
+repeatedly merging pairs of sorted lists into a single sorted list of twice the length. These merges can
+be done very simply in linear time and space. Mergesort's
+input can be thought of as $n$ sorted lists of length one. To simplify the description, we 
+assume the number of keys, $n$, is a power of two. Mergesort makes $\lg n$
+passes over the input where the $i^{th}$ pass merges sorted lists of length $2^{i - 1}$ into 
+sorted lists of length $2^i$.
+
+\subsection{Mergesort Variations}
+\label{mergesort_variations}
+
+For our base
+mergesort implementation we apply several optimizations.
+Firstly, our mergesort alternates the merge
+operation between two arrays to avoid unnecessary copying.
+Secondly, a bitonic sort as described by Sedgewick \citeyear{Sedgewick02},
+removes instructions from the inner-loop. A bitonic sort
+sorts every second list in reverse, merging then works
+from the two ends of the array containing the elements to
+be merged, terminating when the indices cross. Since the indices
+cannot go outside the bounds of the array unless they cross first, 
+this eliminates the need for a bounds check in the inner-loop.
+An insertion sort is also initially used to create
+sorted lists of length four, rather than allowing mergesort to
+sort these lists.
+
+We also analysed the branch prediction behaviour in several
+cache conscious mergesort variations, using the work of LaMarca and Ladner \citeyear{LaMarca97} as a reference.
+Mergesort is not in-place\footnote{Worst case linear time in-place merging is possible but is generally inefficient in practice, see for example Katajainen 
+\textit{et al} \citeyear{Katajainen+96}.}
+and uses an auxiliary array 
+for
+its merge operation. With a cache of size $C$, the largest input we can sort in the cache therefore has size $C/2$. To avoid conflict misses, the
+arrays should be allocated so that the cache line index of the start of the source array is $C/2$ from the cache line index of the start of the destination array. 
+This can
+be done for both the level 1 and level 2 cache.
+
+Since each key is only used once per pass
+in mergesort, if the input is larger than the cache then we will have no temporal re-use of data at all. To address this we can \textit{tile}
+mergesort: cache sized blocks of the input are mergesorted, and then merged as usual. Combined with alignment this variation is referred to as
+tiled mergesort. Tiling mergesort in this way does not fully address the problem of its poor temporal re-use, since the merging of the cache sized blocks in tiled 
+mergesort will exhibit no temporal re-use and again have very poor
+cache performance. To address this, a multi-way merge of the cache sized blocks can be used. A multi-way merge combines a number of sorted lists into a 
+single sorted
+list while reading through each of them only once. A heap can be used to help implement the multi-way merge efficiently. 
+This multi-mergesort variation has better cache performance. We refer to this combination of alignment and tiling with multi-way merging
+simply as multi-mergesort.
+
+\subsection{Branch Prediction Results}
+\label{mergesort_results_text}
+
+Figure \ref{mergesort_branch_results}(a) shows the average number of branches per key executed by
+our mergesort implementations. Up to $2^{18}$ keys can be sorted in the simulated 2 MB level 2 cache
+(each key is 4 bytes, and mergesort is not in-place). As a result, when the number of keys being 
+sorted reaches $2^{19}$ there is a dramatic increase in the number of branches executed by multi-mergesort. This
+is because a multi-way merge begins to be used. Figure \ref{mergesort_branch_results}(b) 
+shows that there is a corresponding increase in the number of branch
+mispredictions caused by multi-mergesort as a result of the multi-way merge.
+
+Many of the additional comparison branches introduced by
+the multi-way merge are associated with a $k$-ary heap, and are therefore quite predictable, as we saw in
+Section \ref{heapsort_branch_results_text}. In addition, two-level adaptive predictors outperform bimodal predictors on multi-mergesort's branches.
+Figure \ref{mergesort_branch_results}(b) shows this. Both predictors have 4096 entry tables. Varying the table size
+of the bimodal predictor between 2048 and 8192 entries did not result in any reduction in the number of branch mispredictions.
+On the other hand, increasing the table size of the two-level adaptive predictor gave incremental
+improvements in branch predictability. This suggests that the two-level adaptive predictor is
+able to exploit correlations between branch outcomes which occur during the multi-way merge.
+
+Figure \ref{mergesort_branch_results}(c) shows the simulated level 2 cache misses per key for our mergesort
+implementations. It is noteworthy that although multi-mergesort has the best cache performance, its cache
+performance is only slightly better than tiled mergesort. However, as the set size increases the difference in cache
+performance will slowly become more pronounced, as the trend in Figure \ref{mergesort_branch_results}(c) suggests. 
+Figure \ref{mergesort_branch_results}(d) shows the cycles per key measured for our mergesort implementations
+running on actual hardware. These results show that our less cache conscious tiled mergesort implementation
+out-performs multi-mergesort. This is likely because the set sizes are not large enough to make the superior
+cache performance of multi-mergesort the dominating factor in the comparative performances. Instead, for the 
+data sizes we used, multi-mergesort's increased instruction count and branch mispredictions
+cause its performance to be inferior to tiled mergesort.
+
+\begin{figure}
+\centering
+\begin{tabular}{cc}
+\includegraphics[width=0.5\textwidth]{plots/mergesort_branches_per_key.eps} & \includegraphics[width=0.5\textwidth]{plots/mergesort_branch_misses.eps} \\
+(a) & (b) \\
+\includegraphics[width=0.5\textwidth]{plots/mergesort_level_2_misses.eps} & \includegraphics[width=0.5\textwidth]{plots/mergesort_cycle_counts.eps} \\
+(c) & (d) \\
+\end{tabular}
+\caption{(a) Shows the branches per key executed by our mergesort variations. The sudden increase in the number of executed branches for multi-mergesort
+results from the use of the multi-way merge when the data no longer fits within the cache. (b) Shows the number of mispredictions per key, all predictors
+had 4096 table entries. Bimodal predictors were used for base and tiled mergesort. These results were generated using \texttt{sim-bpred}, the
+two-level adaptive predictor had a 10-bit history register.
+(c) Shows the level 2 cache misses per key, base mergesort has by far the worst cache behaviour, while multi-mergesort
+has slightly better cache performance than tiled mergesort. These results were generated by using \texttt{sim-cache} to simulate a
+direct mapped 2MB cache with 32-byte cache lines. Finally (d) shows the cycles per key of our algorithms measured using Pentium 4 hardware performance
+counters. Tiled mergesort performs the best out of the algorithms, despite its inferior cache performance compared to multi-mergesort.}
+\label{mergesort_branch_results}
+\end{figure}
 
 \section{Quicksort}
-Quicksort is a divide and conquer algorithm that recursively
-partitions its array around a pivot, swapping large keys on the left
-hand side with smaller keys on the right hand side. The partition
-causes the pivot to be put in its final position, with all greater
-keys to the right, and lesser keys on the left. The left and right
-partitions are then sorted recursively.
 
-\subsection{Simple Quicksort}
-Sedgewick \cite{} has proposed a number of improvements to the
-standard quicksort algorithm, which we use in our base quicksort
-implementation. First, we use an explicit stack to save space.
-Secondly, we always sort the smaller partition of the array first, to
-reduce the worst case size of the stack (from $O(N)$ to
-$O(logN)$). Thirdly, we not quicksort patitions of less than a certain
-\texttt{THRESHOLD} size; instead we do a final insertion pass over the
-array after quicksorting. The smallest of the first \texttt{THRESHOLD}
-keys is placed in the first location of the array as a
-sentinel. Finally, we use median-of-3 to select the pivot, which
-greatly reduces the likelihood of very unbalanced partitioning. This
-also allows the bounds check to be removed from the inner loop,
-greatly reducing the instruction count. Improvements such as these are
-found in most efficient implementations of quicksort.
+Quicksort \cite{Hoare62} is an algorithm which selects a key from its input
+called the pivot and then partitions the other keys into two sets,
+one set containing keys at most equal to the pivot and another
+containing keys at least equal to the pivot. The keys of these sets should
+appear respectively to the left and right of pivot in sorted order.
+Therefore they can be sorted independently, each using quicksort.
 
-In addition to our base quicksort algorithm, we experimented with
-variations which attempt to reduce the number of data cache
-misses. LaMarca \cite{} proposed a \textit{memory-tuned} quicksort.
-Once a partition is reduced below the size of the threshold, insertion
-sort is preformed immediately. By doing this, the keys being sorted
-are in the cache already. By leaving the insertion sort until the end,
-as Sedgewick does, the reduction in instruction count is made up by
-the increase in cache misses, once the lists being sorted are larger
-than the cache.
+\subsection{Quicksort Variations}
 
-\subsection{Multi-Quicksort}
-Generally, quicksort has good locality of reference. It divides the
-same region of the array into increasingly small partitions.  In the
-case that the array to be sorted fits in a particular level of the
-cache, then quicksort has excellent cache properties at that
-level. Once the array is larger than the cache, however, the number of
-misses increases dramatically. On the early passes, which divide the
-array into large sections, there is likely to be little reuse of data
-cache entries between passes. Given that there is little reuse between
-passes, it would be better to reduce the number of passes by dividing
-the array into a larger number of partitions on each pass. LaMarca
-uses a technique similar to that used in multi-mergesort, in that a
-\n{multi-way partition} is done at the start, moving all keys into
-cache sized containers which can be sorted individually.
+For our basic quicksort implementation, we make use of a number
+of improvements suggested by Sedgewick \citeyear{Sedgewick78}. Firstly, we
+use an explicit stack instead of recursing, to save space. Secondly, after
+partitioning, we always quicksort the set with the smaller number of elements first. 
+This reduces the worst case stack space to $O(\lg n)$ from $O(n)$.
+Thirdly, inputs of less than a certain \texttt{THRESHOLD} (typically \texttt{THRESHOLD} is 10) in size are not quicksorted
+at all. Instead, insertion sort is run as a post-pass over the output of quicksort.
+The smallest of the first \texttt{THRESHOLD} keys is used as a sentinel for insertion sort.
+Finally, we use median-of-3 pivoting, greatly reducing the chances of very unbalanced
+partitioning and also removing the need for bounds checks in the inner-loop, thus
+greatly reducing the instruction count.
 
-Implementing a multi-way partition is complicated. At the start of the
-algorithm, a set of pivots is chosen and sorted, and keys are put into
-containers based on which two pivots their value lies between. A
-minimal and maximal key are put on the edges of this list to avoid
-bounds checks as the list is iterated over.
+Improvements such as those just described are found in most efficient quicksort
+implementations. We also investigated the branch prediction behaviour of cache
+conscious quicksort implementations. LaMarca and Ladner \citeyear{LaMarca97} suggest that instead of running
+insertion sort as a post-pass over the quicksorted input, each time quicksort
+is presented with an input of size less than \texttt{THRESHOLD} it immediately
+insertion sorts it. Although this increases the instruction count, it is more
+than compensated for by the resultant drop in cache misses. We refer to this
+as memory-tuned quicksort.
 
-The main problem with multi-quicksort is that we don't know how many
-keys there will be in each container. We implement the containers as a
-linked list of arrays. LaMarca notes very little performance
-difference between 100 and 5000 keys in each block. The size is
-therefore chosen so that each list fits exactly into a page of
-memory. When a list is filled, another list is linked to it, and these
-lists hold all the keys in the container. The total number of
-containers is chosen so that on average each will hold $C/3$ keys,
-where $C$ is the number of keys which fit in the cache. LaMarca shows
-that using this formula on average only 5\% of subsets will be larger
-than the cache.
+When quicksort's input fits within a particular level of the cache quicksort
+has good cache properties at that level. To combat a dramatic increase
+in cache misses when the input does not fit in the cache, LaMarca and Ladner also suggest multi-quicksort. 
+Multi-quicksort chooses a set of pivots and performs a multi-way partition which divides the input into
+cache sized containers. Each container is then quicksorted. We leave
+the precise details of performing a multi-way partition to elsewhere \cite{LaMarca96a,BiggarGregg05}, it is sufficient
+to note that we can decide which container a key should be placed in by using either a sequential
+or binary search.
 
-The entire array is iterated though a key at a time. For each key, the
-list of pivots are searched until the appropriate container is found,
-and the key is added to that container. Sequentially searching the
-lists increases the instruction count by 50\% over the memory-tuned
-version. For this reason, the search was replaced with an efficient
-binary search, which reduced by half the extra instructions.
+\begin{figure}
+\centering
+\begin{tabular}{cc}
+\includegraphics[width=0.5\textwidth]{plots/counter_base_quicksort__median_of_3__0.eps} & \includegraphics[width=0.5\textwidth]{plots/counter_memory_tuned_quicksort_0.eps} \\
+(a) Basic quicksort & (b) Memory-tuned quicksort \\ 
+\includegraphics[width=0.5\textwidth]{plots/counter_multi_quicksort__binary_search__0.eps} & \includegraphics[width=0.5\textwidth]{plots/counter_multi_quicksort__sequential_search__0.eps} \\
+(c) Multi-quicksort (binary search) & (d) Multi-quicksort (sequential search)
+\end{tabular}
+\caption{Overview of branch prediction behaviour in our quicksort implementations. Every figure shows the behaviour of the $i$ and $j$ branches when
+using a median-of-3 pivot. As described in Section \ref{quicksort_branch_results_text}, these branches are about 60\% biased and 64\% predictable when using
+the median-of-3. In (a) the \texttt{median} branch is the combined results of the branches which compute the median-of-3 (these branches are also executed for (b), (c) and (d)).
+Comparing (a) with (b), (c) and (d), we see that the \texttt{insertion} branch associated with its insertion sort is slightly less predictable than in 
+the other variations. This is due to it running as a post-pass. Finally, comparing (c) with (d) we see that the binary search branches of (c), 
+\texttt{binary left} and \texttt{binary right}, are very unpredictable compared to the \texttt{sequential} branch of (d).
+}
+\label{branch_prediction_quicksort}
+\end{figure}
 
-Once the entire array has been put into containers, each container is
-emptied in turn and the keys put back into the array. When a container
-is empty, the pivot greater than that container is put into position,
-which is guaranteed to be its final position. The indices of these
-positions is pushed onto the stack, and the sub-array is then sorted.
+\begin{figure}
+\begin{verbatim}
+      pv = a[0];
+      i = l, j = r + 1;
+      while(true)
+      {
+              while(a[++i] < pv) ; // i-loop
+              while(a[--j] > pv) ; // j-loop
+              if(i >= j) break;
+              swap(a[i], a[j]);
+      }
+      swap(a[0], a[j]);
+\end{verbatim}
+\caption{Quicksort's partition inner-loop. We refer to the inner while loops
+as the $i$ and $j$ loops. We refer to their associated branches as
+the $i$ and $j$ branches respectively.}
+\label{partition_code}
+\end{figure}
 
-While emptying the containers back into the array, an opportunity is
-taken to find the smallest key in the leftmost partition. This is
-placed as a sentinel, ensuring that every partition has a sentinel,
-and that no bounds checks are required.
+\subsection{Branch Prediction Results}
+\label{quicksort_branch_results_text}
 
-\subsection{Quicksort Results}
-Figure \ref{} shows the
+Figure \ref{branch_prediction_quicksort} shows the branch prediction results of our quicksort implementations.
+In all figures, the most important branches are the $i$ and $j$ branches (see Figure \ref{partition_code}), since these 
+are in quicksort's inner-loop, and are executed more often than any of the other branches (with the exception of the \texttt{sequential} branch
+of Figure \ref{branch_prediction_quicksort}(d)).
 
-\section{Radixsort}
+The branch prediction behaviour of basic quicksort and memory-tuned quicksort are similar.
+As can be seen in Figure \ref{branch_prediction_quicksort}(b) 
+quicksort has a slightly lower branch misprediction rate on the \texttt{insertion} branch. This
+is because a basic quicksort implementation, which performs insertion sort as a post-pass,
+will generate branch mispredictions on keys chosen as pivots since they never require movement.
+In practice, the over-all reduction in branch misprediction rate from 35\% in basic quicksort
+to 34\% in memory-tuned quicksort is probably offset by the branch misprediction rate of the
+outer-loop exit in the many insertion sorts executed by memory-tuned quicksort. However,
+memory-tuned quicksort has better caching properties, and so from the point of view of both
+branch prediction and caching behaviour this optimization is to be recommended.
+
+Multi-quicksort results in a 19\% reduction in the number of iterations of the $i$ and $j$
+loops of quicksort's partition inner-loop, 
+though their misprediction rate remains the same at 37\%. Figure \ref{branch_prediction_quicksort}(c) shows branch behaviour
+for multi-quicksort when using binary search. The binary search branches are highly unpredictable,
+being incorrectly predicted 43\% of the time. If multi-quicksort uses sequential search, the
+number of executed branches dramatically increases, but these
+branches are correctly predicted 92\% of the time, as shown in Figure \ref{branch_prediction_quicksort}(d).
+
+The implementations of multi-quicksort have the best cache performance of the quicksort variations, as Figure \ref{quicksort_plots}(a) shows.
+However, the multi-quicksort implementations also introduce many extra instructions per key as Figure \ref{quicksort_plots}(b) shows. Naturally, the 
+sequentially searched version introduces more extra instructions than the binary searched version. However, as described in the previous paragraph,
+the branches introduced by binary search are highly unpredictable. As a result, multi-quicksort using binary search has more branch mispredictions than the
+sequential searched version (see Figure \ref{quicksort_plots}(c)), consequently, despite having identical cache performance, binary searched multi-quicksort
+performs worse than the sequentially searched multi-quicksort, as shown in Figure \ref{quicksort_plots}(d).
+Of course, on very large data sets (larger than the $2^{22}$ keys we experimented with), the
+reduced instruction count of binary searched multi-quicksort would eventually allow it to out-perform sequentially searched multi-quicksort.
+
+As Figure \ref{quicksort_plots}(d) shows, our base quicksort implementation performs the best of all. It seems the improved cache performance of
+multi-quicksort is not enough for it to out-perform base quicksort on the data set sizes we experimented with. Examining the trends of 
+Figure \ref{quicksort_plots}(a), it seems likely that multi-quicksort would eventually out-perform base quicksort due to its improved cache performance.
+The fact that base quicksort out-performs memory-tuned quicksort is very difficult to explain. In the simulations memory-tuned quicksort has the same
+cache performance and instruction count, as shown in Figures \ref{quicksort_plots}(a) and \ref{quicksort_plots}(b). Also in the simulations, 
+memory-tuned quicksort has slightly fewer branch mispredictions than base quicksort, as Figure \ref{quicksort_plots}(c) shows. Despite this, base 
+quicksort slightly out-performs memory-tuned quicksort on our non-simulated hardware test as Figure \ref{quicksort_plots}(d) shows.
+
+\begin{figure}
+\centering
+\begin{tabular}{cc}
+\includegraphics[width=0.5\textwidth]{plots/quicksort_cache_misses.eps} & \includegraphics[width=0.5\textwidth]{plots/quicksort_instruction_counts.eps} \\
+(a) & (b) \\ 
+\includegraphics[width=0.5\textwidth]{plots/quicksort_branch_misses.eps} & \includegraphics[width=0.5\textwidth]{plots/quicksort_cycles.eps} \\
+(c) & (d) \\
+\end{tabular}
+\caption{(a) Shows the 2nd level cache misses for our quicksort implementations. As is expected, the multi-quicksort implementations have the best
+cache performance. These results were generated using \texttt{sim-cache} to simulate a 2MB direct-mapped cache with 32-byte cache lines.
+(b) Shows the instruction counts of our quicksort implementations, when the data sets no longer fit within the cache the multi-quicksort implementations
+show a large increase in instruction count, these results were also generated using SimpleScalar. (c) Shows the simulated branch mispredictions per key
+for our quicksort implementations. All predictors use 4096 table entries and are bimodal predictors except where indicated. Multi-quicksort using
+binary search is the only implementation which gives varying results from the use of a two-level adaptive predictor. These results were generated
+using \texttt{sim-bpred}. (d) Shows the cycles per key for our quicksort implementations measured using Pentium 4 hardware performance counters.
+}
+\label{quicksort_plots}
+\end{figure}
+
+\subsection{Pivot Choice}
+\label{quicksort_pivot_choice}
+
+The choice of pivot element has a strong influence on the behaviour of the $i$ and 
+$j$ branches (see Figure \ref{partition_code}). Clearly, if the pivot is chosen as the median of the input then
+either branch is equally likely to be taken (i.e. the loop executes) as not taken (i.e.
+the loop exits). In this case, the branches are also on average approximately 50\% predictable. 
+However, on average over random input, with a fixed choice of pivot the $i$ and
+$j$ branches are about 66\% biased (that is, they are taken about 66\% of the time). It is
+straightforward to show why this is so. Assume that the data to be partitioned is a random permutation of $\lbrace 1, \ldots, n \rbrace$.
+If the chosen pivot has rank $q$ then the $i$ branch will be taken with probability $(q - 1)/n$.
+Moreover, the branch will be executed (but not necessarily taken) a total of $q$ times, since there are exactly $q - 1$ elements
+to the left of the pivot. The average bias is given by averaging over all 
+possible choices of pivot $q$, thus
+
+\begin{equation}
+B_{avg}^n = \frac{2}{n(n + 1)}\sum_{q = 1}^n \left(\frac{q - 1}{n} \right) q = \frac{2}{3} - \frac{2}{3n} 
+\label{bias_equation}\end{equation}
+
+\noindent
+So the bias rapidly approaches 66\% as $n$ increases. With this bias, the $i$ and $j$ branches are about 71\% predictable. 
+The fact that predictability of these branches exceeds their biases
+may at first seem paradoxical. The reason the branches are more predictable than biased is because the bias 
+is an average. The average predictability of the $i$ and $j$ branches can be obtained very simply from
+Equation \ref{bias_equation} using the steady state predictability function, $CP$, introduced in Section \ref{bubble_sort}, thus
+
+\[
+P_{avg}^n = \frac{2}{n(n + 1)}\sum_{q = 1}^n CP\left( \frac{q - 1}{n} \right) q
+\]
+
+\noindent
+As $n \rightarrow \infty$ we have $P_{avg} = \int_0^1 CP(q)\,dq = 3/2 - \pi/4$.
+Thus we see $P_{avg}$ is about 0.71. This estimated average predictability is close to what is observed experimentally when using a fixed pivot
+over random data, as the predictability entry of the first row of Table \ref{median_table} shows.
+
+If the pivot is chosen as the median-of-3, then the $i$ and $j$ branches
+are about 60\% biased in the taken direction, and about 64\% predictable. It is especially
+noteworthy that although median-of-3 gives around a 14\% reduction in the total number of executed
+comparison branches, there is actually a 6\% \textit{increase} in the total
+number of branch mispredictions. As the pivot more closely approximates the median of the input,
+there is a continuing trend of a reduction in the total number of executed branches, as well as reducing bias and predictability 
+with an increasing total number of mispredictions, as Table \ref{median_table} shows.
+
+\begin{table}
+\begin{tabular}{|c|c|c|c|c|}
+\hline
+Median-of & Reduction In $\#$Branches & Bias    & Predictability & Misprediction Increase \\
+\hline                                                                                   
+1         & 0\%                       & 66.1\%  & 70.6\%         & 0\%                    \\
+\hline
+3         & 14\%                      & 60.0\%  & 64.3\%         & 6.1\%                  \\
+\hline
+5         & 16.5\%                    & 57.8\%  & 61.6\%         & 9.2\%                  \\                     
+\hline
+7         & 17.8\%                    & 56.3\%  & 59.9\%         & 11.8\%                 \\
+\hline
+9         & 19.6\%                    & 55.1\% & 58.3\%          & 14.3\%                 \\
+\hline
+\end{tabular}
+\caption{The effect of choosing a pivot closer to the median on the prediction
+of the $i$ and $j$ branches (the results are the same for both). As a higher order median is used the bias of the $i$ and $j$
+branches is reduced, with a corresponding reduction in their predictability.
+The reductions in the number of executed branches are relative to a fixed choice of pivot (i.e. median-of-1). 
+The misprediction increases correspond to the total number
+of branches mispredicted with a higher order median compared to a fixed choice of pivot.
+These measurements were obtained 
+from our own software simulations of a bimodal branch predictor. Note that the small number of branches
+required to compute the median approximations are included in these results.}
+\label{median_table} 
+\end{table}
+
+The use of the median-of-3 or higher order median approximation greatly reduces the
+chances of the worst case behaviour of quicksort, and on average therefore results
+in fewer comparison branches being executed. From an information theoretic
+perspective, each comparison branch resolves more uncertainty, or yields
+more information, and is hence harder to predict. This observation applies also
+to the predictability of the branches used in the binary searched version of 
+multi-quicksort versus the sequentially searched version. There is a trade-off
+to be had between the penalty induced by unpredictable comparisons, versus the 
+increased instruction count caused by making comparisons more predictable.
+
+In work concurrent and independent of our own, Kaligosi and Sanders \citeyear{Kaligosi+06} 
+have investigated the effect of the choice of pivot on the performance of quicksort. They found that
+an artificially skewed pivot gives better performance than using a random pivot or the median-of-3,
+due to the reduction in branch mispredictions. This performance increase is despite the increased instruction count
+such an artificially skewed pivot causes, we describe this work in more detail in Section \ref{RelatedWork}.
+
+\section{Radix Sort}
+\label{Radixsort}
+
+The purpose of this paper is to show the relevance of branch mispredictions to sorting. We now show how to develop a  
+radix sort \cite{Friend56} implementation for 32-bit integers that, in all our tests, is more efficient than all other sorting algorithms presented in this paper. 
+This is in part due to the tiny number of branch mispredictions incurred by radix sorting. It is also in part because radix sort 
+operates in linear worst case time. However, in spite of its better asymptotic performance a ``step'' for our radix sort is quite costly, requiring
+a key to be counted four times and copied four times, with the counting involving several layers of indirection. In contrast, a step for quicksort
+is as simple as a comparison and a possible exchange every few steps.
+
+The cache performance of radix sort is generally poor compared
+to algorithms like mergesort or quicksort. In radix sort, each key is used just once each time it is in the cache, when the data to be sorted
+is larger than the cache every key access will result in a cache miss. On the other hand, mergesort and quicksort have good temporal locality and use keys 
+repeatedly before they are ejected from the cache.
+
+\subsection{Memory-tuned Radix Sort}
+
+Our basic radix sort implementation is a typical least significant digit radix sort. We operate in radix 256, and begin by counting the number of keys
+with a particular least significant byte by indexing a table of 256 entries with the least significant bytes of the keys. The entries of this table are then summed to 
+store the cumulative count of keys with a smaller least significant byte. Now each key is placed in an auxiliary array according to the value retrieved by 
+using its least significant byte to index into the table of cumulative counts. Before the next iteration, the auxiliary array is written back to the 
+original array and the sort can proceed on the next most significant byte. We use this implementation to compare against the simple caching improvements
+we now suggest.
+
+We can firstly remove the copying back and forth between the original and auxiliary array, by just alternating their roles on each step. Since we are
+sorting 32 bit integers, the number of iterations is even and the data will end up sorted in its original location. If the number of steps was odd, an
+additional copy would be required at the end. Since radix sort is out-of-place, we might consider aligning its two arrays as we did with mergesort (see
+Section \ref{mergesort_variations}). However, radix sort does not access its arrays in the same regular fashion that mergesort does and so is unlikely to
+benefit from such a change. Indeed, we implemented a double-aligned radix sort and did not observe any performance gains.
+Finally, we use an improvement suggested by LaMarca and Ladner \citeyear{LaMarca97}, where the counting 
+of the next most significant bytes is begun during their preceding iteration. We refer to this optimized version of radix sort as memory-tuned
+radix sort.
+
+In our basic radix sort implementation, the first array is accessed 3 times and the second array 4 times, in each of the four
+iterations over the four bytes in the 32-bit integers. This gives 20 accesses in total.
+In memory-tuned radix sort each key of the two arrays is accessed only nine times in total.
+At the start, it traverses the original array once to count the least significant bytes. It then iterates four times, positioning
+the keys based on the previously established counts. These four iterations cause
+a total of eight accesses, because each array is accessed once per iteration.
+
+\subsection{Results}
+
+\begin{figure}
+\centering
+\begin{tabular}{cc}
+\includegraphics[width=0.5\textwidth]{plots/radix_cycles.eps} & \includegraphics[width=0.5\textwidth]{plots/radix_versus_quick_level_2_misses.eps} \\
+(a) & (b) \\
+\includegraphics[width=0.5\textwidth]{plots/radix_branch_misses.eps} & \includegraphics[width=0.5\textwidth]{plots/radix_branches.eps} \\
+(c) & (d) \\
+\end{tabular}
+\caption{(a) Shows the cycles per key of our radix sort implementations alongside our base quicksort implementation, measured using Pentium 4
+hardware performance counters. (b) Shows the level 2 cache 
+misses incurred by our radix sort implementations and a multi-quicksort implementation for a 2MB direct mapped cache with 32-byte cache lines. These
+results were generated using \texttt{sim-cache}.
+(c) Shows the average branch mispredictions per key 
+incurred by our radix sort implementations while using bimodal predictors with 4096 table entries, the results for two-level adaptive predictors are extremely
+similar. Finally, (d) shows the average number of branches per key executed by 
+our  implementations. The results of (c) and (d) were generated using \texttt{sim-bpred}.}
+\label{radixsort_results_pictures}
+\end{figure}
+
+Our radix sort implementation requires fewer processor cycles per key than even our base quicksort 
+implementation as Figure \ref{radixsort_results_pictures}(a) shows. The number of level 2 cache misses is only slightly higher than we measured for
+quicksort while the array fits inside the cache, presumably because radix sort is out-of-place.
+
+Figure \ref{radixsort_results_pictures}(b) shows the level 2 cache misses for our basic and memory-tuned implementations.
+We note that, when the array no longer fits inside the cache the memory-tuned implementation has approximately half the number of misses per key
+(1.25) than the basic version (2.5). When the data no longer fits inside the cache there should be about one miss for each key per pass per cache line.
+Each cache line can fit 8 keys, and thus we expect about 2 misses per key since there are 16 passes in total. The 2.5 misses per key which was measured
+likely exceeds this because of the irregular fashion of the array indexing in each pass. Similary, in memory-tuned radix sort we expect about 1.125
+misses per key, because there are 9 passes, again the measured 1.25 misses per key presumably exceeds this because
+of the irregular array indexing.
+
+The large advantage of radix sort is its almost complete lack of branch mispredictions however. Figure \ref{radixsort_results_pictures}(c) shows the tiny
+number of branch mispredictions incurred by radix sort. It also executes a similarly tiny number of branch instructions as 
+Figure \ref{radixsort_results_pictures}(d) shows.
 
 \section{Related work} 
-
+\label{RelatedWork}
 % The IBM 7030 "Stretch" supercomputer was the first general-purpose
@@
 architectures. As early as 1972 Knuth commented on the
-'ever-increasing number of "pipeline" or "number crunching" computers
-that have appeared in recent years' whose 'efficiency deteriorates
+``ever-increasing number of `pipeline' or `number crunching' computers
+that have appeared in recent years'' whose ``efficiency deteriorates
 noticeably in the presence of conditional branch instructions unless
-the branch almost always goes the same way'. He also notes that 'radix
+the branch almost always goes the same way''. He also notes that ``radix
 sorting is usually more efficient than any other known method for
-internal sorting on such machines' \cite{Knuth72c}. Most likely, Knuth
-was referring to the IBM 7030 "Stretch" computer (released in 1961
-with a four-stage pipeline), Seymor Cray's Freon-cooled CDC 6600
+internal sorting on such machines'' (the same comment appears in \cite{KnuthVol3_98}). Most likely, Knuth
+was referring to the IBM 7030 ``Stretch'' computer (released in 1961
+with a four-stage pipeline), Seymour Cray's Freon-cooled CDC 6600
 supercomputer (released 1964), or other early supercomputers which
@@
 sorting algorithms for more recent single-processor machines. Nyberg
-et al present a quicksort-based algorithm for RISC architectures
-\cite{Nyberg+94}. They found that the main limit on performance was
-cache behaviour, and did not mention branch prediction even
-once. Bentley implemented a highly-tuned quicksort, to be used in the
-\texttt{qsort} function of the standard C library
-\cite{Bentley93}. Again, Bentley did not consider branch prediction,
+\textit{et al} \citeyear{Nyberg+94} present a quicksort-based algorithm for RISC
+architectures. They found that the main limit on
+performance was cache behaviour, and did not mention branch prediction
+even once. Bentley and McIlroy \citeyear{Bentley+93} implemented a highly-tuned quicksort,
+to be used in the \texttt{qsort} function of the standard C library. Again, they did not consider branch prediction,
 perhaps because the first desktop processors with dynamic branch
@@
 
-Agarwal developed an optimized algorithm for sorting random records on
-an IBM RS/6000 RISC machine \cite{Agarwal96}. His is the first
+Agarwal \citeyear{Agarwal96} developed an optimized algorithm for sorting random records on
+an IBM RS/6000 RISC machine. His is the first
 published algorithm that we are aware of that deliberately attempts to
@@
 evaluate very large numbers of variants of the algorithm.  A similar
-approch is used by \cite{Li+05} to automatically generate sorting
+approch is used by Li \textit{et al} \citeyear{Li+05} to automatically generate sorting
 algorithms that are tuned to the architecture of the target machine
@@
 
-Sanders and Winkel investigate the use of \emph{predicated}
-instructions on Intel Itanium~2 processors \cite{SandersWinkel05}. In
+Sanders and Winkel \citeyear{Sanders+04} investigate the use of \emph{predicated}
+instructions on Intel Itanium 2 processors. In
 addition to its normal inputs, a predicated instruction takes a
@@
 the partition step in quicksort can be rewritten to use predicated
-instructions on an Itanium~2 highly instruction-level parallel
-machine. Itannium~2 provides large numbers of parallel functional
+instructions on an Itanium 2, a highly instruction-level parallel
+machine. Itanium 2 provides large numbers of parallel functional
 units and registers which make such trade-offs worthwhile.
 
-We attempted to replicate Sanders and Winkel's work a Pentium~4 based
-desktop machine, using conditional move instructions (the only form of
-predicated instructions on the Pentium~4). The standard form of the
-quicksort loops is highly efficient, but we found that rewriting them
-into a single loop suitable for predication made them less so. The
-transformation also increased the number of simultaneously live
-values, which exceeded the number of available registers. We also
-discovered that, at least on the Pentium 4, conditional moves are
+In  
+preliminary experiments, we found that Sanders and Winkel's work is difficult 
+to reproduce on a Pentium 4 based machine. The only predicated instruction
+supported on the Pentium 4 is the conditional move instruction. We found that
+rewriting
+quicksort's loops so that they are suitable for predication
+reduces their efficiency and also increases the number of simultaneously
+live variables. Since the Pentium 4 has only eight general
+purpose registers the number of simultaneously live variables exceeded
+the number of available registers, reducing efficiency. In addition,
+we also discovered that on the Pentium 4 conditional moves are
 surprisingly expensive.
 
-Mudge \textit{et al} examine the behaviour of the $i$ and $j$
-comparison branches in quicksort, as an example to demonstrate a
+Mudge \textit{et al} \citeyear{Mudge+96} examine the behaviour of the $i$ and $j$
+comparison branches in randomized quicksort, as an example to demonstrate a
 general method for estimating the limit on the predictability of
-branches in a program \cite{Mudge+96}. They argue that the optimal
-predictor for these branches would keep a running count of the total
-number of the proportion of array elements examined so far that are
-greater than the pivot. They estimate that this approach will give a
-prediction accuracy of around 75\%, which is remarkably close to our
+branches in a program. They argue that the optimal
+predictor for these branches would 
+keep a running count of the total
+number of array elements examined so far that are
+greater than the pivot. If the majority of the elements examined so far
+are greater than the pivot, the next element is predicted as greater
+than the pivot, and vice versa. They estimate that this approach will give a
+prediction accuracy of 75\%, which is close to our
 result for these branches, using a bimodal predictor.
+This shows that a bimodal predictor is close to optimal
+for randomized quicksort's branches, since $P_{avg}$ of Section \ref{quicksort_branch_results_text}
+is about 0.71.
 
-\textbf{Describe the paper from Alenex 2005}
+Brodal \textit{et al} \citeyear{Brodal+05} examine the adaptiveness of quicksort.
+They find that although randomized quicksort is not adaptive (i.e. its asymptotic analysis does not 
+improve with respect to some measure of presortedness), its performance
+is better on data with a low number of inversions. They justify this by showing 
+that the expected number of element swaps performed by randomized quicksort falls as
+the number of inversions does. The number of branch mispredictions
+is roughly twice the number of element swaps, because two branch mispredictions
+tend to occur when the two while loops of the partition code exit (see Figure \ref{partition_code}).
+Furthermore they note that the branch mispredictions are the dominant part of the running time of
+randomized quicksort. This provides an empirical relation between the presortedness of the input, 
+randomized quicksort's performance, and the number of branch mispredictions.
 
-To our knowlege, this is the only research, other than our own, that
-investigates and attempts to understand the predictability of
-comparison branches in sorting, rather than simply attempting to
-eliminate such branches.
+In closely related work Brodal and Moruz \citeyear{BrodalMoruz05} provide a lower bound on the
+number of branch mispredictions a comparison-based sorting algorithm must incur based on the
+number of comparisons it performs. They also show that a mergesort variation can achieve this lower
+bound based on the number of comparisons it performs, although they do not provide experimental
+performance characteristics. They also extend their results to adaptive sorting algorithms
 
-\section{Conclusions}
+In recent work conducted independently of our own, Kaligosi and Sanders \citeyear{Kaligosi+06}
+investigate the choice of pivot on the performance of quicksort. They examine $\alpha$-skewed
+pivoting, that is, where the pivot has rank $\alpha n$ when sorting $n$ keys. They examine the performance
+of quicksort when the pivot is determined randomly, from the median-of-3, or is the true median
+(i.e. a 1/2-skewed pivot). They show that using a 1/10-skewed pivot gives a performance increase 
+over all the aforementioned pivot choices. This is despite the increase in instruction count that a 
+deliberately skewed pivot causes.
 
-\comment{
-\begin{acks}
-We would like to thank the anonymous reviewers of VEE 2005, whose
-comments greatly improved an earlier version of this paper.
-\end{acks}
-}
+\section{Conclusion}
 
+This paper has presented the first large scale study on the characteristics of branch instructions in 
+all of the most common sorting algorithms. We noted that except where the sorting algorithms involve reasonably complicated
+flow control (i.e. shellsort, multi-mergesort, multi-quicksort), bimodal predictors preform just as well as two-level adaptive predictors. Indeed we
+noted that for shaker sort a bimodal predictor out-performs a two-level adaptive predictor.
+
+Of the elementary sorting algorithms we noted that insertion sort,
+causing just a single misprediction per key, has the best branch prediction characteristics. On the other hand,
+bubble and shaker sort operate in a manner which causes surprisingly unpredictable branches.
+
+Sorting (and indeed, searching) algorithms have been designed to minimize the number of comparisons necessary 
+in the worst case. However, minimizing the number of comparisons makes each comparison less predictable. When a branch is
+mispredicted, there is often a large penalty on modern processors. 
+There is therefore a tension between reducing the number of executed comparisons and the penalty associated with 
+increasing the chance of branch mispredictions. This tension is especially evident in the choice of pivot for quicksort.
+
+Since cache optimizations are now recognized as very important, we 
+have also noted the affect that these cache optimizations have on the branch prediction behaviour of sorting algorithms. In particular we
+provided performance results for cache optimized versions of quicksort, mergesort, heapsort and radix sort.
+
+Our study provides concrete results of the branch prediction characteristics for a large number
+of sorting algorithms, accompanied with the relevant analysis.  These branch prediction results have been obtained for realistic dynamic predictors, using
+a variety of configurations of bimodal and two-level adaptive predictors.
+
+\appendix
+
+\section{Appendix: Steady State Predictability}
+\label{appendix_SSP}
+
+In Sections \ref{bubble_sort}, \ref{heapsort_branch_results_text} and \ref{quicksort_branch_results_text}
+we made use of the steady state predictability function $CP$. When a branch instruction is executed 
+repeatedly with a known probability $p$ of being taken, $CP(p)$ gives a good estimate of the probability
+that it will be correctly predicted.
+
+The bimodal predictor associated with a branch instruction can be thought
+of as a Markov chain. Note that we are thus assuming that successive branches
+are independent. If the branch has probability $p$ of being taken then 
+the transition matrix of the associated Markov chain is
+
+\[
+M = 
+\left[
+\begin{array}{llll}
+p   & p   & 0   & 0 \\
+1-p & 0   & p   & 0 \\
+0   & 1-p & 0   & p \\
+0   & 0   & 1-p & 1-p \\
+\end{array}
+\right]
+\]
+
+\noindent
+The steady state probabilities are given by $s$ such that $Ms = s$.
+Since the steady state probabilities sum to one this can be reduced to solving
+a $3 \times 3$ linear system. The resultant probabilities are given by
+
+\[
+\left[ 
+\begin{array}{c}
+P(ST) \\
+P(T) \\
+P(NT) \\
+P(SNT) \\
+\end{array}
+\right]
+=
+\frac{1}{2p^2 - 2p + 1}
+\left[
+\begin{array}{c}
+p^3 \\
+p^2(1 - p) \\
+p(1 - p)^2 \\
+(1 - p)^3 \\
+\end{array}
+\right]
+\]
+
+\noindent
+Here $P(ST)$, $P(T)$, $P(NT)$ and $P(SNT)$ denote the respective probabilities of the predictor states strongly taken, taken, not-taken
+and strongly not-taken described in Section \ref{branch_prediction}. A branch is correctly predicted with probability $p$ if we are 
+in the strongly taken or taken state. Similarly, a branch is correctly predicted with probability $1 - p$ if we are in the strongly not-taken
+or not-taken state. Thus,
+
+\[
+CP(p) = p\left[P(ST) + P(T)\right] + (1 - p)\left[P(SNT) + P(NT)\right]
+\]
+
+\noindent
+This equation simplifies to Equation \ref{CP_equation} of Section \ref{bubble_sort}.
 \bibliographystyle{acmtrans}
-\bibliography{report}
-%\begin{thebibliography}{...}
-%\end{thebibliography}
-
+\bibliography{paper}
 \begin{received}
-...
 \end{received}

### 2007-01-29 docs/tech_report/radix.tex
did the abstract. removed comment from the latex files
Updated the tech report to revision 1, which corrects 16 to 20. This also fixes the graphs in some places.

@@
 \label{radix}
-%\section{Description}
 
@@
 $(bitsPerInt/log_2(Radix))$. For a 32-bit integer and a radix of 256, this is
-four steps and each step has four passes over the array, for a total of 16
-passes over the array. With eight keys per cache line, there should be about two
+four steps and each step has five passes over the array, for a total of 20
+passes over the array. With eight keys per cache line, there should be about 2.5
 misses per key, when the array doesn't fit into the array, though this depends
@@
 results should remain the same. The instruction count should, however, reduce by
-nearly a half to $\frac{9}{16}^{th}s$ of base radixsort.
+nearly a half to $\frac{9}{20}^{th}s$ of base radixsort.
 
@@
 cache, the number of misses jumps to around 3.25 (which would be equivalent to
-1.625 misses per key in our results), where they stay. Our results
-disagree, showing approximately 2.5 misses per key for base radixsort.
+1.625 misses per key in our results), where they stay. Our results disagree,
+showing approximately 2.5 misses per key for base radixsort, in line with our
+predictions.
 

### 2007-01-29 docs/branch_paper/paper.tex
Formatting changes for paper.tex and paper.bib, updated README, removed unused directory.

@@
 \author{
-PAUL BIGGAR\footnote[1]{PUT IN IRCSET GRANT DETAILS}, NICHOLAS NASH\footnotemark[1] KEVIN WILLIAMS\footnotemark[1] and DAVID GREGG
-\\University of Dublin, Trinity College
+PAUL BIGGAR\footnote[1]{Supported by the Irish Research Council for Science,
+Engineering and Technology (IRCSET).}, NICHOLAS NASH\footnotemark[1], KEVIN
+WILLIAMS\footnote[2]{Supported by the Irish Research Council for Science,
+Engineering and Technology (IRCSET) and IBM.}, and DAVID GREGG \\University of
+Dublin, Trinity College
 }
@@
 \begin{abstract} 
-Sorting is one of the most important and well studied problems in Computer Science. Many good algorithms
-are known which offer various trade-offs in efficiency, simplicity, memory use, and other factors. However,
-these algorithms do not take into account features of modern computer architectures that significantly
-influence performance. Caches and branch predictors are two such features, and while
-there has been a significant
-amount of research into the cache performance of general purpose sorting algorithms, 
-there has been little research on their branch prediction properties. In this paper we empirically examine
-the behaviour of the branches in all the most common sorting algorithms. 
-We also consider the interaction of cache optimization on the predictability of the branches in these algorithms.
-We find insertion sort to have the fewest branch mispredictions of any
-comparison-based sorting algorithm, that bubble and shaker sort operate in a fashion which makes their
-branches highly unpredictable, that the unpredictability of shellsort's branches improves
-its caching behaviour and that several cache optimizations have little effect on mergesort's branch
-mispredictions. We find also that optimizations to quicksort -- for example the choice of pivot -- have a strong
-influence on the predictability of its branches. We point out a simple way of removing branch instructions from 
-a classic heapsort implementation, and show also that unrolling a loop in a cache
-optimized heapsort implementation improves the predicitability of its branches. Finally, we note
-that when sorting random data two-level adaptive branch predictors are usually no better than simpler bimodal predictors. This is despite 
-the fact
-that two-level adaptive predictors are almost always superior to bimodal predictors in general.
+Sorting is one of the most important and well studied problems
+in Computer Science. Many good algorithms are known which offer various
+trade-offs in efficiency, simplicity, memory use, and other factors. However,
+these algorithms do not take into account features of modern computer
+architectures that significantly influence performance. Caches and branch
+predictors are two such features, and while there has been a significant amount
+of research into the cache performance of general purpose sorting algorithms,
+there has been little research on their branch prediction properties. In this
+paper we empirically examine the behaviour of the branches in all the most
+common sorting algorithms. We also consider the interaction of cache
+optimization on the predictability of the branches in these algorithms.  We find
+insertion sort to have the fewest branch mispredictions of any comparison-based
+sorting algorithm, that bubble and shaker sort operate in a fashion which makes
+their branches highly unpredictable, that the unpredictability of shellsort's
+branches improves its caching behaviour and that several cache optimizations
+have little effect on mergesort's branch mispredictions. We find also that
+optimizations to quicksort -- for example the choice of pivot -- have a strong
+influence on the predictability of its branches. We point out a simple way of
+removing branch instructions from a classic heapsort implementation, and show
+also that unrolling a loop in a cache optimized heapsort implementation improves
+the predicitability of its branches. Finally, we note that when sorting random
+data two-level adaptive branch predictors are usually no better than simpler
+bimodal predictors. This is despite the fact that two-level adaptive predictors
+are almost always superior to bimodal predictors in general.
 \end{abstract}
@@
 \section{Motivation}
-Classical analyses of algorithms make simplifying assumptions about
-the cost of different machine instructions. For example, the RAM model used for 
-establishing asymptotic bounds and Knuth's MIX 
-machine code both make drastically simplifying assumptions about the cost of
-machine instructions \cite{KnuthVol1_97}. 
-More
-recently researchers have recognized that on modern computers the cost
-of accessing memory can vary dramatically depending on whether the
-data can be found in the first-level cache, or must be fetched from a
-lower level of cache or even main memory.  This has spawned a great
-deal of research on cache-efficient searching and sorting \cite{Nyberg+94,Agarwal96,LaMarca96b,LaMarca96a,LaMarca97,Xiao+00,Rahman+01,Wickremesinghe+02}.
-
-Another type of instruction whose cost can vary dramatically is the
-conditional branch. Modern pipelined processors depend on \emph{branch
-prediction} for much of their performance.  If the direction of a
-conditional branch is correctly predicted ahead of time, the cost of
-the conditional branch may be as little as the cost of, say, an
-integer add instruction. If, on the other hand, the branch is
-mispredicted the processor must flush its pipeline, and restart
-from the correct target of the branch. This cost is typically a large
-multiple of the cost of executing a correctly-predicted branch. For
-example, Intel Pentium 4 processors \cite{Intel248966-010,Intel249438-01} have pipelines of up 
-to 31 stages,
-meaning that a branch misprediction will cost around 30 cycles.  In
-contrast, on the same processor the cost of a memory access that
-misses the first-level cache and hits the second level cache is 23
-cycles. Fortunately, the branches in most programs are very
-predictable, so branch mispredictions are usually rare. Indeed, prediction
-accuracies of greater than 90\% are typical \cite{Uht+97} with the
-best predictors.
-
-The cost of executing branches is particularly important for sorting
-because the inner-loops of most sorting algorithms consist of
-comparisons of items to be sorted. Thus, the predictability of
-these comparison branches is critical to the performance of
-sorting algorithms. In this paper we study the predictability
-of branches in most of the major sorting algorithms. 
-We focus on the behaviour of the branches whose outcome depends on a
-comparison of keys presented to the sorting algorithm in its input. Throughout this paper we refer to
-such branches as \textit{comparison branches}. Branches associated with controlling simpler aspects 
-of the control flow of the algorithms are of much less interest because they are generally almost perfectly predictable.
-We also examine cache conscious variations of many of the algorithms,
-and show how optimizations for the cache influence the level of branch
-mispredictions.
+Classical analyses of algorithms make simplifying assumptions about the cost of
+different machine instructions. For example, the RAM model used for establishing
+asymptotic bounds and Knuth's MIX machine code both make drastically simplifying
+assumptions about the cost of machine instructions \cite{KnuthVol1_97}. More
+recently researchers have recognized that on modern computers the cost of
+accessing memory can vary dramatically depending on whether the data can be
+found in the first-level cache, or must be fetched from a lower level of cache
+or even main memory.  This has spawned a great deal of research on
+cache-efficient searching and sorting
+\cite{Nyberg+94,Agarwal96,LaMarca96b,LaMarca96a,LaMarca97,Xiao+00,Rahman+01,Wickremesinghe+02}.
+
+Another type of instruction whose cost can vary dramatically is the conditional
+branch. Modern pipelined processors depend on \emph{branch prediction} for much
+of their performance.  If the direction of a conditional branch is correctly
+predicted ahead of time, the cost of the conditional branch may be as little as
+the cost of, say, an integer add instruction. If, on the other hand, the branch
+is mispredicted the processor must flush its pipeline, and restart from the
+correct target of the branch. This cost is typically a large multiple of the
+cost of executing a correctly-predicted branch. For example, Intel Pentium 4
+processors \cite{Intel248966-010,Intel249438-01} have pipelines of up to 31
+stages, meaning that a branch misprediction will cost around 30 cycles.
+Fortunately, the branches in most programs are very predictable, so branch
+mispredictions are usually rare. Indeed, prediction accuracies of greater than
+90\% are typical \cite{Uht+97} with the best predictors.
+
+The cost of executing branches is particularly important for sorting because the
+inner-loops of most sorting algorithms consist of comparisons of items to be
+sorted. Thus, the predictability of these comparison branches is critical to the
+performance of sorting algorithms. In this paper we study the predictability of
+branches in most of the major sorting algorithms.  We focus on the behaviour of
+the branches whose outcome depends on a comparison of keys presented to the
+sorting algorithm in its input. Throughout this paper we refer to such branches
+as \textit{comparison branches}. Branches associated with controlling simpler
+aspects of the control flow of the algorithms are of much less interest because
+they are generally almost perfectly predictable.  We also examine cache
+conscious variations of many of the algorithms, and show how optimizations for
+the cache influence the level of branch mispredictions.
 
@@
 
-Branches are a type of instruction used for defining the flow control of programs. In high level languages they result from 
-the use of statements like \texttt{if}, \texttt{while} and their variants. Branches can be either \textit{taken}, indicating that the
-address they provide is the new value for the program counter, or they can be \textit{not-taken}, in which case sequential execution continues
-as though the branch had not been present.
-
-Branch instructions pose a difficulty for pipelined 
-processors. In a pipelined processor, while the program counter points to a particular instruction, a potentially large
-number of subsequent instructions are in a partially completed state.
-Until it is known whether a branch is taken or not-taken, it is not possible to know what the next instructions to be executed are.
-If the processor cannot tell what these next instructions are, it cannot fill its pipeline.
-To keep the utilization of the pipeline at a reasonable level, modern processors attempt to anticipate the outcome of 
-branch instructions by employing \textit{branch predictors}. A good discussion of pipelining and its associated issues can
-be found in Hennessy and Patterson \citeyear{HennessyPatterson90}.
-
-There are several kinds of branch predictors: static, semi-static and dynamic. We use only dynamic predictors in our experiments, 
-because they are most commonly used in real processors. The statistics in the following paragraphs are taken from those derived
-by Uht \textit{et al} \citeyear{Uht+97}, over the SPECint92 benchmarking suite.
-
-The simplest type of dynamic predictor is referred to 
-as a \textit{1-bit} predictor. It keeps a table recording, at each entry, whether a particular branch was taken or not-taken. 
-Using the table, it predicts 
-that a branch
-will go the same way as it went on its previous execution. 1-bit predictors achieve 77\% to 79\% accuracy.
-
-We refer to a 2-bit dynamic predictor as a \textit{bimodal} predictor. A bimodal predictor operates in the same manner as a 1-bit predictor but each table entry
-can be thought of  as maintaining a counter from 0 to 3. The counter decrements on each
-taken branch (with the exception of when the count is 0), and increments on each not-taken branch (with the exception of when the count is 3). With
-counts equal to 0 or 1 the next branch is predicted as taken, with counts of 2 or 3 the branch is predicted not-taken. 
-When the counter is 0, we say the predictor is in the \textit{strongly taken} state, since the branch must be not-taken twice before the not-taken
-direction will be predicted again. When the counter is 1 we say the predictor is in the \textit{taken} state. Similarly counts of 2 and 3 are referred
-to as \textit{not-taken} and \textit{strongly not-taken} respectively. Bimodal predictors achieve 78\% to 89\% accuracy.
-
-Finally, some branch predictors attempt to exploit correlations in branch outcomes to improve accuracy. A \textit{two-level adaptive} predictor maintains
-a \textit{branch history register}. This register records the outcomes of a number of previous branch instructions. For example, a register contents 
-of \texttt{110010} indicates that the 6 previous 
-branch outcomes were taken, taken, not-taken, not-taken, taken and not-taken.
-For each branch, this register is used to index a table of bimodal predictors.
-The program counter can also be included in the index, either concatenated or XORed with the indexing register. Two-level adpative
-predictors are accurate about 93\% of the time.
-
-In all the above, due to the size of the predictor table branches can collide. That is, two 
-different branches can map to the same table location, often reducing accuracy.
+Branches are a type of instruction used for defining the flow control of
+programs. In high level languages they result from the use of statements like
+\texttt{if}, \texttt{while} and their variants. Branches can be either
+\textit{taken}, indicating that the address they provide is the new value for
+the program counter, or they can be \textit{not-taken}, in which case sequential
+execution continues as though the branch had not been present.
+
+Branch instructions pose a difficulty for pipelined processors. In a pipelined
+processor, while the program counter points to a particular instruction, a
+potentially large number of subsequent instructions are in a partially completed
+state.  Until it is known whether a branch is taken or not-taken, it is not
+possible to know what the next instructions to be executed are.  If the
+processor cannot tell what these next instructions are, it cannot fill its
+pipeline.  To keep the utilization of the pipeline at a reasonable level, modern
+processors attempt to anticipate the outcome of branch instructions by employing
+\textit{branch predictors}. A good discussion of pipelining and its associated
+issues can be found in Hennessy and Patterson \citeyear{HennessyPatterson90}.
+
+There are several kinds of branch predictors: static, semi-static and dynamic.
+We use only dynamic predictors in our experiments, because they are most
+commonly used in real processors. The statistics in the following paragraphs are
+taken from those derived by Uht \textit{et al} \citeyear{Uht+97}, over the
+SPECint92 benchmarking suite.
+
+The simplest type of dynamic predictor is referred to as a \textit{1-bit}
+predictor. It keeps a table recording, at each entry, whether a particular
+branch was taken or not-taken.  Using the table, it predicts that a branch will
+go the same way as it went on its previous execution. 1-bit predictors achieve
+77\% to 79\% accuracy.
+
+We refer to a 2-bit dynamic predictor as a \textit{bimodal} predictor. A bimodal
+predictor operates in the same manner as a 1-bit predictor but each table entry
+can be thought of  as maintaining a counter from 0 to 3. The counter decrements
+on each taken branch (with the exception of when the count is 0), and increments
+on each not-taken branch (with the exception of when the count is 3). With
+counts equal to 0 or 1 the next branch is predicted as taken, with counts of 2
+or 3 the branch is predicted not-taken.  When the counter is 0, we say the
+predictor is in the \textit{strongly taken} state, since the branch must be
+not-taken twice before the not-taken direction will be predicted again. When the
+counter is 1 we say the predictor is in the \textit{taken} state. Similarly
+counts of 2 and 3 are referred to as \textit{not-taken} and \textit{strongly
+not-taken} respectively. Bimodal predictors achieve 78\% to 89\% accuracy.
+
+Finally, some branch predictors attempt to exploit correlations in branch
+outcomes to improve accuracy. A \textit{two-level adaptive} predictor maintains
+a \textit{branch history register}. This register records the outcomes of a
+number of previous branch instructions. For example, a register contents of
+\texttt{110010} indicates that the 6 previous branch outcomes were taken, taken,
+not-taken, not-taken, taken and not-taken.  For each branch, this register is
+used to index a table of bimodal predictors.  The program counter can also be
+included in the index, either concatenated or XORed with the indexing register.
+Two-level adpative predictors are accurate about 93\% of the time.
+
+In all the above, due to the size of the predictor table branches can collide.
+That is, two different branches can map to the same table location, often
+reducing accuracy.
 
@@
 
-We used truly random data provided by Haahr \citeyear{Random} in all our experiments. We used ten chunks of random data containing $2^{22}$ (4194304) keys.
-Where a particular experiment used only part of the keys in a chunk, it used the left-most keys.
-Our results are averaged over these chunks of data.
-
-Processors do not generally provide access to the behaviour of their caches or branch predictors. However, a software processor simulator can be used to 
-provide measurements 
-of cache and branch prediction behaviour.
-Our branch prediction and caching results were generated using the SimpleScalar PISA simulator version 3 \cite{SimpleScalarTutorialv4}. 
-We used \texttt{sim-cache} and 
-\texttt{sim-bpred} to generate results for caching and branch prediction characteristics respectively.
-We used a variety of cache configurations; generally speaking we used an 8 KB level 1 data cache, 8 KB level 1 instruction cache,
-and a shared 2 MB instruction and data level 2 cache, all with 32 byte cache lines. 
-We also experimented with both direct mapped and fully associative caches. When presenting results
-where the cache configuration is important, we will describe the precise configuration used. We present
-results here only for direct mapped caches. The results for fully associative caches are similar \cite{BiggarGregg05}.
-
-For the measurements taken using SimpleScalar, we used power of two sized sets of random keys ranging from $2^{12}$ to $2^{22}$ keys in size.
-This is with the exception of the quadratic time sorting algorithms, for which the maximum set size used was $2^{16}$ keys. This was done
-because of the very significant time it takes such algorithms to sort large inputs.
-
-It takes time to fill arrays, and this can distort the relationship between the results for small set sizes, and larger set sizes, 
-for which this time is amortised. As a result, we measured this time, and subtracted it from our results.
-However, this practice also distorts the results. It removes compulsory cache misses from the results, so long as the data fits in the cache. 
-When the data does not fit in the cache, the keys from the start of the array are faulted, with the effect that capacity misses reoccur, 
-which are not discounted.
-
-For branch prediction we simulated both bimodal and two-level adaptive predictors. 
-We used power of two sized tables for the branch predictors. 
-We used tables containing between $2^{11}$ and $2^{14}$ predictors.
-When
-presenting branch prediction results, the predictor configurations will be described precisely.
-SimpleScalar's \texttt{sim-bpred} provides total results over
-all the branches in a program. It is often desirable to examine particular branches individually however.
-In order to do so, we added our own simulations of branch predictors to the programs. For our 
-own simulations we average the results of each branch over ten runs of random data containing $2^{22}$ keys.
-
-Finally we used \textit{PapiEx} \cite{PapiExManual} to access performance counters on a Pentium 4 1.6 Ghz. This allows cycle counts (i.e. running times)
-to be
-determined for the sorting algorithms on a real processor.
-The results presented from these tests are averaged over 1024 runs and use the system's random number generator to provide data. The data
-ranged in size from $2^{12}$ to $2^{22}$ keys in size.
-The processor we used had an 8-way associative 256KB level 2 
-cache with 64-byte cache lines. The separate data and instruction level 1 caches are 4-way associative, 8KB in size and have 32-byte cache lines.
-The Pentium 4 uses an unspecified form of dynamic branch prediction. A reference manual \cite{Intel248966-010} mentions a branch history
-register, so it is probable that the scheme is similar to two-level adaptive.
+We used truly random data provided by Haahr \citeyear{Random} in all our
+experiments. We used ten chunks of random data containing $2^{22}$ (4194304)
+keys.  Where a particular experiment used only part of the keys in a chunk, it
+used the left-most keys.  Our results are averaged over these chunks of data.
+
+In order to experiment with a variety of cache and branch prediction results we
+used the SimpleScalar PISA processor simulator version 3
+\cite{SimpleScalarTutorialv4}.  We used \texttt{sim-cache} and
+\texttt{sim-bpred} to generate results for caching and branch prediction
+characteristics respectively.
+
+We used a variety of cache configurations; generally speaking we used an 8 KB
+level 1 data cache, 8 KB level 1 instruction cache, and a shared 2 MB
+instruction and data level 2 cache, all with 32 byte cache lines.  We also
+experimented with both direct mapped and fully associative caches. When
+presenting results where the cache configuration is important, we will describe
+the precise configuration used. We present results here only for direct mapped
+caches. The results for fully associative caches are similar, as we show in our
+technical report \cite{BiggarGregg05}, which contains complete results for a
+large number of branch predictor and cache configurations.
+
+For the measurements taken using SimpleScalar, we used power of two sized sets
+of random keys ranging from $2^{12}$ to $2^{22}$ keys in size.  This is with the
+exception of the quadratic time sorting algorithms, for which the maximum set
+size used was $2^{16}$ keys. This was done because of the very significant time
+it takes such algorithms to sort large inputs.
+
+It takes time to fill arrays, and this can distort the relationship between the
+results for small set sizes, and larger set sizes, for which this time is
+amortised. As a result, we measured this time, and subtracted it from our
+results.  However, this practice also distorts the results. It removes
+compulsory cache misses from the results, so long as the data fits in the cache.
+When the data does not fit in the cache, the keys from the start of the array
+are faulted, with the effect that capacity misses reoccur, which are not
+discounted.
+
+For branch prediction we simulated both bimodal and two-level adaptive
+predictors.  We used power of two sized tables for the branch predictors.  We
+used tables containing between $2^{11}$ and $2^{14}$ predictors.  When
+presenting branch prediction results, the predictor configurations will be
+described precisely.  SimpleScalar's \texttt{sim-bpred} provides total results
+over all the branches in a program. It is often desirable to examine particular
+branches individually however.  In order to do so, we added our own simulations
+of branch predictors to the programs. For our own simulations we average the
+results of each branch over ten runs of random data containing $2^{22}$ keys.
+
+Finally we used \textit{PapiEx} \cite{PapiExManual} to access performance
+counters on a Pentium 4 1.6 Ghz. This allows cycle counts (i.e. running times)
+to be determined for the sorting algorithms on a real processor.  The results
+presented from these tests are averaged over 1024 runs and use the system's
+random number generator to provide data. The data ranged in size from $2^{12}$
+to $2^{22}$ keys in size.  The processor we used had an 8-way associative 256KB
+level 2 cache with 64-byte cache lines. The separate data and instruction level
+1 caches are 4-way associative, 8KB in size and have 32-byte cache lines.  The
+Pentium 4 uses an unspecified form of dynamic branch prediction. A reference
+manual \cite{Intel248966-010} mentions a branch history register, so it is
+probable that the scheme is similar to two-level adaptive.
 
@@
 
-We begin by analysing the branch prediction behaviour of a number
-of common elementary sorting algorithms, each operating in
-$O(n^2)$ time. The main practical advantage of these
-algorithms is their simplicity, which generally allows them to
-operate faster than algorithms with $O(n \lg n)$ time complexity
-for small inputs. Indeed, as we will see later, optimized versions
-of $O(n \lg n)$ sorts use calls to elementary sorts for small inputs.
-Knuth \citeyear{KnuthVol3_98} provides a good exposition of these elementary
-sorting algorithms (as well as basic versions of the sorting algorithms
-we shall examine in later sections).
+We begin by analysing the branch prediction behaviour of a number of common
+elementary sorting algorithms, each operating in $O(n^2)$ time. The main
+practical advantage of these algorithms is their simplicity, which generally
+allows them to operate faster than algorithms with $O(n \lg n)$ time complexity
+for small inputs. Indeed, as we will see later, optimized versions of 
+$O(n \lg n)$ sorts use calls to elementary sorts for small inputs.  Knuth
+\citeyear{KnuthVol3_98} provides a good exposition of these elementary sorting
+algorithms (as well as basic versions of the sorting algorithms we shall examine
+in later sections).
 
@@
 
-Selection sort is a very simple quadratic time sorting algorithm which begins by sequentially searching
-for the smallest key in its input and moving it into position. The algorithm 
-continues by repeatedly searching for and then moving the smallest key in the unsorted
-section of its input into the appropriate position. The inner-loop of selection sort,
-which is run a total of $n - 1$ times on an input \texttt{a[0..n - 1]} is shown below
+Selection sort is a very simple quadratic time sorting algorithm which begins by
+sequentially searching for the smallest key in its input and moving it into
+position. The algorithm continues by repeatedly searching for and then moving
+the smallest key in the unsorted section of its input into the appropriate
+position. The inner-loop of selection sort, which is run a total of $n - 1$
+times on an input \texttt{a[0..n - 1]} is shown below
 
@@
 It is easy to see that selection sort performs approximately $n^2/2$ comparisons
-and exactly $n - 1$ exchanges. 
-On the $j^{th}$ iteration of the inner-loop the comparison branch is taken
-if \texttt{a[j]} is the minimum of \texttt{a[i..j]}, which occurs with
-probability $1 / (j - i + 1)$\footnote{Throughout this paper, for the purpose of analysis we make
-the usual assumption that the input to be sorted consists of a random permutation of $\lbrace1, .., n\rbrace$.}. 
-The expected number of times the branch is taken
-on the first iteration is thus given by $H_n - 1$ where 
-$H_n = \sum_{i = 1}^n 1/i$ is the $n^{th}$ harmonic number \cite{KnuthVol1_97}.
-Since $H_n \approx \ln n$, and $\sum_{i = 1}^n H_i = O(n \ln n)$ the comparison branch
-will be highly predictable. As expected, Figure \ref{elementary_sorts}(a) shows that number of mispredictions
-per key grows in a similar fashion to $\ln n$.
-
-If the array is initially in sorted order, then the value of \texttt{min}
-will never change in the inner-loop. Since this implies the comparison branch
-is never taken, it will be almost perfectly predictable. Similarly, 
-if the array is sorted in reverse order, the value of \texttt{min} must
-change at every element. That is, the comparison branch 
-is taken at every element, and the branch is therefore almost perfectly
-predictable.  From the point of view of branch prediction, frequent
-changes (preferably with no simple pattern) are the main indicator of
-poor performance, rather than the absolute number of times that the
-branch resolves in each direction. The worst case for selection sort
-would be an array which is partially sorted in reverse order; the value of
-\texttt{min} would change frequently, but not predictably so.
-Figure \ref{elementary_sorts}(a) shows the behaviour of selection
-sort's branches.
+and exactly $n - 1$ exchanges.  On the $j^{th}$ iteration of the inner-loop the
+comparison branch is taken if \texttt{a[j]} is the minimum of \texttt{a[i..j]},
+which occurs with probability $1 / (j - i + 1)$\footnote{Throughout this paper,
+for the purpose of analysis we make the usual assumption that the input to be
+sorted consists of a random permutation of $\lbrace1, .., n\rbrace$.}.  The
+expected number of times the branch is taken on the first iteration is thus
+given by $H_n - 1$ where $H_n = \sum_{i = 1}^n 1/i$ is the $n^{th}$ harmonic
+number \cite{KnuthVol1_97}.  Since $H_n \approx \ln n$, and 
+$\sum_{i = 1}^n H_i = O(n \ln n)$ the comparison branch will be highly
+predictable. As expected, Figure \ref{elementary_sorts}(a) shows that number of
+mispredictions per key grows in a similar fashion to $\ln n$.
+
+If the array is initially in sorted order, then the value of \texttt{min} will
+never change in the inner-loop. Since this implies the comparison branch is
+never taken, it will be almost perfectly predictable. Similarly, if the array is
+sorted in reverse order, the value of \texttt{min} must change at every element.
+That is, the comparison branch is taken at every element, and the branch is
+therefore almost perfectly predictable.  From the point of view of branch
+prediction, frequent changes (preferably with no simple pattern) are the main
+indicator of poor performance, rather than the absolute number of times that the
+branch resolves in each direction. The worst case for selection sort would be an
+array which is partially sorted in reverse order; the value of \texttt{min}
+would change frequently, but not predictably so.  Figure
+\ref{elementary_sorts}(a) shows the behaviour of selection sort's branches.
 
@@
 
-Insertion sort is another simple algorithm with quadratic
-worst case time. It works by iterating the inner-loop below
-a total of $n - 1$ times over its input
+Insertion sort is another simple algorithm with quadratic worst case time. It
+works by iterating the inner-loop below a total of $n - 1$ times over its input
 
@@
 \end{verbatim}
-On the $i^{th}$ iteration \texttt{a[0..i - 1]} is sorted, allowing
-\texttt{a[i]} to be placed in the correct position as shown in
-the inner-loop above. On average, insertion sort performs approximately
-$n^2/4$ comparisons and $n^2/4$ assignments in its inner-loop. Thus insertion sort
-performs around half as many branches as selection sort.
-
-A very pleasing property of insertion sort is that
-it generally causes just 
-a single branch misprediction per key, as Figure \ref{elementary_sorts}(a) shows. Typically
-the inner \texttt{while} loop exit causes this misprediction, when
-the correct location for the current item has been found.
-In fact, a static
-predictor would perform just as well as the two predictors shown, since it
-would mispredict exactly once per key.
+
+On the $i^{th}$ iteration \texttt{a[0..i - 1]} is sorted, allowing \texttt{a[i]}
+to be placed in the correct position as shown in the inner-loop above. On
+average, insertion sort performs approximately $n^2/4$ comparisons and $n^2/4$
+assignments in its inner-loop. Thus insertion sort performs around half as many
+branches as selection sort.
+
+A very pleasing property of insertion sort is that it generally causes just a
+single branch misprediction per key, as Figure \ref{elementary_sorts}(a) shows.
+Typically the inner \texttt{while} loop exit causes this misprediction, when the
+correct location for the current item has been found.  In fact, a static
+predictor would perform just as well as the two predictors shown, since it would
+mispredict exactly once per key.
 
@@
 \end{tabular}
-\caption{(a) Shows the branch mispredictions for selection and insertion sort using bimodal and two-level adaptive predictors both having 4096 table entries.
-It is noteworthy the simple bimodal predictor has essentially identical performance to the two-level adaptive predictors.
-(b) Shows the branch mispredictions for bubble and shaker sort using bimodal and two-level adaptive predictors, again both having 4096 table entries.
-For bubble sort the bimodal and two-level adaptive predictors have close to identical performance. However, for shaker sort the simple bimodal
-predictor results in fewer mispredictions than the two-level adaptive predictor. We also note that as a result of 
-the partial sorting performed by bubble and shaker sort they incur far more
-branch mispredictions per key than other $O(n^2)$ sorting algorithms.
-\texttt{sim-bpred} was used to simulate the branch predictors in both (a) and (b). The two-level adaptive predictors used 10-bit history registers.}
+\caption{(a) Shows the branch mispredictions for selection and insertion sort
+using bimodal and two-level adaptive predictors both having 4096 table entries.
+It is noteworthy the simple bimodal predictor has essentially identical
+performance to the two-level adaptive predictors. (b) Shows the branch
+mispredictions for bubble and shaker sort using bimodal and two-level adaptive
+predictors, again both having 4096 table entries. For bubble sort the bimodal
+and two-level adaptive predictors have close to identical performance. However,
+for shaker sort the simple bimodal predictor results in fewer mispredictions
+than the two-level adaptive predictor. We also note that as a result of the
+partial sorting performed by bubble and shaker sort they incur far more branch
+mispredictions per key than other $O(n^2)$ sorting algorithms.
+\texttt{sim-bpred} was used to simulate the branch predictors in both (a) and
+(b). The two-level adaptive predictors used 10-bit history registers.}
 \label{elementary_sorts}
@@
 Bubble sort is another simple sorting algorithm, although notoriously
-inefficient. It works by iterating the loop shown below at most $n - 1$
-times on an input \texttt{a[0..n - 1]}.
+inefficient. It works by iterating the loop shown below at most $n - 1$ times on
+an input \texttt{a[0..n - 1]}.
 
@@
 }
+
 \end{verbatim}
+
 In this code $i$ is the outer-loop iteration counter, and begins at 0 and counts
-towards $n - 1$. Bubble sort uses
-information yielded by the comparisons which selection
-sort simply discards. After a full iteration of the inner-loop shown larger
-elements have moved to the right, closer to their final location.
+towards $n - 1$. Bubble sort uses information yielded by the comparisons which
+selection sort simply discards. After a full iteration of the inner-loop shown
+larger elements have moved to the right, closer to their final location.
 
 Shaker sort is a variation of bubble sort in which there are two inner-loops
-which are alternated. One is the inner-loop shown above. The other
-inner-loop scans right-to-left moving small elements to the left in the
-same manner as moving larger elements to the right has just been described.
-Although shaker sort has the same worst-case time as bubble sort, it is generally
-more efficient in practice.
-
-Figure \ref{elementary_sorts}(b) shows the number of branch mispredictions per key for
-bubble and shaker sort. As the data set gets large, bubble sort causes nearly 10000
-branch mispredictions per key. Shaker sort executes a lot fewer
-branches, and so has fewer mispredictions, but the rate of
-misprediction is similar.
-
-To understand the high misprediction rate of bubble sort we define $Q^k_j$
-as the probability that $j^{th}$ inner-loop comparison branch is taken on the $k^{th}$ 
-outer-loop iteration, $1 \leq k < n$. It turns out that
+which are alternated. One is the inner-loop shown above. The other inner-loop
+scans right-to-left moving small elements to the left in the same manner as
+moving larger elements to the right has just been described.  Although shaker
+sort has the same worst-case time as bubble sort, it is generally more efficient
+in practice.
+
+Figure \ref{elementary_sorts}(b) shows the number of branch mispredictions per
+key for bubble and shaker sort. As the data set gets large, bubble sort causes
+nearly 10000 branch mispredictions per key. Shaker sort executes a lot fewer
+branches, and so has fewer mispredictions, but the rate of misprediction is
+similar.
+
+To understand the high misprediction rate of bubble sort we define $Q^k_l$ as
+the probability that $l^{th}$ inner-loop comparison branch is taken on the
+$k^{th}$ outer-loop iteration, $1 \leq k < n$. It turns out that
 
@@
 \label{bubble_equation}
-Q^k_j = 
+Q^k_l = 
 \left\{ 
 \begin{array}{ll}
-  \frac{j}{j + k} & \quad \mbox{if $j \leq n - k$}\\
+  \frac{l}{l + k} & \quad \mbox{if $l \leq n - k$}\\
   0 & \quad \mbox{otherwise}\\ 
@@
 \noindent
-This is a result of the fact that the $j^{th}$ comparison branch of the inner-loop
-is taken only if \texttt{a[j]} is not larger than all of \texttt{a[0..j - 1]}.
-For this
-to be the case, the $(j + 1)^{th}$ branch in the previous outer-loop iteration must have been taken,
-since otherwise the previous iteration of the outer-loop left the maximum of \texttt{a[0..j - 1]} in \texttt{a[j]}.
-The probability that the $(j + 1)^{th}$ branch of the
-previous outer-loop iteration is taken is $Q^{k - 1}_{j + 1}$. Given that the $(j + 1)^{th}$ branch of
-the previous outer-loop iteration was taken, the probability that \texttt{a[j]} is
-not larger than all of \texttt{a[0..j - 1]} is $1 - 1/(j + 1)$, thus,
+This is a result of the fact that the $l^{th}$ comparison branch of the
+inner-loop is taken only if \texttt{a[l]} is not larger than all of
+\texttt{a[0..l - 1]}.  For this to be the case, the $(l + 1)^{th}$ branch in the
+previous outer-loop iteration must have been taken, since otherwise the previous
+iteration of the outer-loop left the maximum of \texttt{a[0..l - 1]} in
+\texttt{a[l]}.  The probability that the $(l + 1)^{th}$ branch of the
+previous outer-loop iteration is taken is $Q^{k - 1}_{l + 1}$. Given that the
+$(l + 1)^{th}$ branch of the previous outer-loop iteration was taken, the
+probability that \texttt{a[j]} is not larger than all of \texttt{a[0..l - 1]} is
+$1 - 1/(l + 1)$, thus,
 
 \[
-Q^k_j = Q^{k-1}_{j+1}\left(1 - \frac{1}{j + 1} \right)
+Q^k_l = Q^{k-1}_{l+1}\left(1 - \frac{1}{l + 1} \right)
 \]
@@
 \noindent
-The bases $Q^{1}_j = 1 - 1 / (j + 1)$ for $1 \leq j < n$ and $Q^1_n = 0$ give the solution shown in Equation \ref{bubble_equation}.
-It is easy to see that in the first outer-loop iteration ($k = 1$) the comparison branch is very likely to be taken, whereas in the last
-outer-loop iteration ($k = n - 1$), the comparison branch is very unlikely to be taken. Moving between these two extremes causes the intermediate
-branches to be unpredictable. The remainder of this section examines this phenomenon in more detail.
-
-In order to analyse the behaviour of bubble sort when a bimodal predictor is used,
-it is useful to make use of the \textit{steady state predictability} function. This function
-generally gives a good estimate of the probability that a particular branch will be correctly predicted,
-given that we know the probability that the branch will be taken. For a bimodal predictor this function is
-given by
+The bases $Q^{1}_l = 1 - 1 / (l + 1)$ for $1 \leq l < n$ and $Q^1_n = 0$ give
+the solution shown in Equation \ref{bubble_equation}.  It is easy to see that in
+the first outer-loop iteration ($k = 1$) the comparison branch is very likely to
+be taken, whereas in the last outer-loop iteration ($k = n - 1$), the comparison
+branch is very unlikely to be taken. Moving between these two extremes causes
+the intermediate branches to be unpredictable. The remainder of this section
+examines this phenomenon in more detail.
+
+In order to analyse the behaviour of bubble sort when a bimodal predictor is
+used, it is useful to make use of the \textit{steady state predictability}
+function. This function generally gives a good estimate of the probability that
+a particular branch will be correctly predicted, given that we know the
+probability that the branch will be taken. For a bimodal predictor this function
+is given by
 
@@
 
-\noindent
-Where $p$ is the probability that the branch in question will be taken. Although the derivation of this
-function is simple, we relegate it to the appendix because we wish
-to focus mainly on our experimental results. $CP$ can be used to examine the predictability of the branches in bubble sort as the outer-loop iterates.
-Bubble sort executes $n - k$ comparison branches on its $k^{th}$ outer-loop iteration. Thus,
-using Equation \ref{bubble_equation} the average probability that the comparison branch is taken on the $k^{th}$ outer-loop iteration
-is
+\noindent Where $p$ is the probability that the branch in question will be
+taken. Although the derivation of this function is simple, we relegate it to the
+appendix because we wish to focus mainly on our experimental results. $CP$ can
+be used to examine the predictability of the branches in bubble sort as the
+outer-loop iterates.  Bubble sort executes $n - k$ comparison branches on its
+$k^{th}$ outer-loop iteration. Thus, using Equation \ref{bubble_equation} the
+average probability that the comparison branch is taken on the $k^{th}$
+outer-loop iteration is
 
 \begin{eqnarray*}
-Q^k_{avg} &=& \frac{1}{n - k} \sum_{j = 1}^{n - k} \frac{j}{j + k} \\
+Q^k_{avg} &=& \frac{1}{n - k} \sum_{l = 1}^{n - k} \frac{l}{l + k} \\
           &=& 1 + \frac{k}{n - k}(H_k - H_n)\\
@@
 \noindent
-Where $H_n$ denotes the $n^{th}$ harmonic number. An approximation to the predictability
-of the comparison branch of bubble sort on the $k^{th}$ iteration is then given by $CP(Q^k_{avg})$.
-Figure \ref{bubble_predictability}(a) plots this, showing how the predictability of the comparison branch varies
-with the number of outer-loop iterations. Figure \ref{bubble_predictability}(b)
-shows the measured number of correctly predicted branches as the outer-loop iterates, these results
-correspond approximately with what was obtained analytically in Figure \ref{bubble_predictability}(a).
+Where $H_n$ denotes the $n^{th}$ harmonic number. An approximation to the
+predictability of the comparison branch of bubble sort on the $k^{th}$ iteration
+is then given by $CP(Q^k_{avg})$.  Figure \ref{bubble_predictability}(a) plots
+this, showing how the predictability of the comparison branch varies with the
+number of outer-loop iterations. Figure \ref{bubble_predictability}(b) shows the
+measured number of correctly predicted branches as the outer-loop iterates,
+these results correspond approximately with what was obtained analytically in
+Figure \ref{bubble_predictability}(a).
 
@@
 \end{tabular}
-\caption{(a) Shows $CP(Q^k_{avg})$ as the sweep number, $k$, increases (i.e. as the outer-loop iterates). This gives a good approximation to the 
-predictability of the comparison branch of bubble sort. (b) Shows the number of correctly predicted branches in bubble sort, as well as the number of executed branches.
-Initially the number of correctly predicted branches is close to the number executed, however, the number of correctly predicted branches rapidly 
-decreases as the sweep number increases. Gradually, the branches become predictable again. These trends correspond to the 
-variation in predictability shown in (a). The results of (b)
-were obtained using our own software simulated bimodal predictor averaged over a large number of arrays with 8192 elements. Note also that in (b) 
-the version of bubble sort used does not terminate early if no swaps are performed for an entire execution of the inner-loop.  }
+\caption{(a) Shows $CP(Q^k_{avg})$ as the sweep number, $k$, increases (i.e. as
+the outer-loop iterates). This gives a good approximation to the predictability
+of the comparison branch of bubble sort. (b) Shows the number of correctly
+predicted branches in bubble sort, as well as the number of executed branches.
+Initially the number of correctly predicted branches is close to the number
+executed, however, the number of correctly predicted branches rapidly decreases
+as the sweep number increases. Gradually, the branches become predictable again.
+These trends correspond to the variation in predictability shown in (a). The
+results of (b) were obtained using our own software simulated bimodal predictor
+averaged over a large number of arrays with 8192 elements. Note also that in (b)
+the version of bubble sort used does not terminate early if no swaps are
+performed for an entire execution of the inner-loop.  }
 \label{bubble_predictability}
@@
 
-Initially the branches
-are quite predictable, however their predictability reduces quite rapidly as the outer-loop iterates.
-The incremental movement of large keys to the right in bubble sort causes this degredation in the
-predictability of its branches. As the data becomes close to being
-fully sorted the predictability of the branches improves again. The partial sorting bubble sort performs allows it to potentially converge to 
-fully sorted data in fewer outer-loop iterations than selection sort. We can
-detect that the data is sorted early if the branch shown in the inner-loop above
-is never taken for a whole iteration. However, this partial sorting also greatly increases
-the number of branch mispredictions bubble sort incurs, substantially slowing it over selection sort or insertion sort.
+Initially the branches are quite predictable, however their predictability
+reduces quite rapidly as the outer-loop iterates.  The incremental movement of
+large keys to the right in bubble sort causes this degredation in the
+predictability of its branches. As the data becomes close to being fully sorted
+the predictability of the branches improves again. The partial sorting bubble
+sort performs allows it to potentially converge to fully sorted data in fewer
+outer-loop iterations than selection sort. We can detect that the data is sorted
+early if the branch shown in the inner-loop above is never taken for a whole
+iteration. However, this partial sorting also greatly increases the number of
+branch mispredictions bubble sort incurs, substantially slowing it over
+selection sort or insertion sort.
 
@@
 
-Our results clearly demonstrate the importance of branch
-prediction for the elementary sorting algorithms. 
-Our shaker sort implementation has fewer cache misses than our selection sort implementation, as Figure \ref{shaker_plots}(a) shows. 
-It also has a lower instruction count than selection sort. However it is slower due to branch mispredictions, as Figure \ref{shaker_plots}(b) shows.
-Figure \ref{elementary_sorts} shows
-the contrast between the branch prediction performance of selection sort and bubble or shaker sort. Where
-selection sort causes on average fewer than 12 mispredictions per key for a set
-size of 65536 keys, bubble and shaker sort cause many thousands.
-Insertion sort behaves even better than selection sort, causing
-almost a single misprediction per key. Since insertion sort
-also has the fewest cache misses it has much to recommend it
-as an elementary sorting algorithm. It is bubble sort's
-partial sorting behaviour that wreaks havoc with branch prediction,
-adding further weight to Knuth's \citeyear{KnuthVol3_98} point of view that it
-has very little to recommend it, besides a catchy name and some interesting theoretical problems it
-poses.
-
-Finally our results show that over these elementary sorting algorithms two-level adaptive branch predictors are no
-better than simpler bimodal predictors. 
-This is noteworthy, because two-level
-predictors are similar to, or sigificantly more accurate than bimodal
-predictors for almost all other types of branches \cite{Uht+97}.
-In fact, a bimodal predictor out-performs a two-level adaptive predictor for shaker sort, as
-is shown in Figure \ref{elementary_sorts}(b).
+Our results clearly demonstrate the importance of branch prediction for the
+elementary sorting algorithms.  Our shaker sort implementation has fewer cache
+misses than our selection sort implementation, as Figure \ref{shaker_plots}(a)
+shows.  It also has a lower instruction count than selection sort. However it is
+slower due to branch mispredictions, as Figure \ref{shaker_plots}(b) shows.
+Figure \ref{elementary_sorts} shows the contrast between the branch prediction
+performance of selection sort and bubble or shaker sort. Where selection sort
+causes on average fewer than 12 mispredictions per key for a set size of 65536
+keys, bubble and shaker sort cause many thousands.  Insertion sort behaves even
+better than selection sort, causing almost a single misprediction per key. Since
+insertion sort also has the fewest cache misses it has much to recommend it as
+an elementary sorting algorithm. It is bubble sort's partial sorting behaviour
+that wreaks havoc with branch prediction, adding further weight to Knuth's
+\citeyear{KnuthVol3_98} point of view that it has very little to recommend it,
+besides a catchy name and some interesting theoretical problems it poses.
+
+Finally our results show that over these elementary sorting algorithms two-level
+adaptive branch predictors are no better than simpler bimodal predictors.  This
+is noteworthy, because two-level predictors are similar to, or sigificantly more
+accurate than bimodal predictors for almost all other types of branches
+\cite{Uht+97}.  In fact, a bimodal predictor out-performs a two-level adaptive
+predictor for shaker sort, as is shown in Figure \ref{elementary_sorts}(b).
 
@@
 \end{tabular}
-\caption{(a) Shows the level 1 cache misses of the elementary sorting algorithms. 
-These results were obtained by using \texttt{sim-cache} to simulate an 8KB direct mapped 
-level 1 cache with 32 byte cache lines. Note that there are no level 2 cache misses since the simulated 2nd
-level cache is 2MB in size.
-(b) Shows the number of cycles per key for the elementary sorting algorithms. 
-Despite bubble and shaker sort having better cache performance than selection sort, the vast number of branch mispredictions 
-(see Figure \ref{elementary_sorts}(b)) they cause make them less efficient than selection sort in practice.
-These results were measured using hardware performance counters on a Pentium 4.}
+\caption{(a) Shows the level 1 cache misses of the elementary sorting
+algorithms.  These results were obtained by using \texttt{sim-cache} to simulate
+an 8KB direct mapped level 1 cache with 32 byte cache lines. Note that there are
+no level 2 cache misses since the simulated 2nd level cache is 2MB in size.  (b)
+Shows the number of cycles per key for the elementary sorting algorithms.
+Despite bubble and shaker sort having better cache performance than selection
+sort, the vast number of branch mispredictions (see Figure
+\ref{elementary_sorts}(b)) they cause make them less efficient than selection
+sort in practice.  These results were measured using hardware performance
+counters on a Pentium 4.}
 \label{shaker_plots}
@@
 
-Shellsort \cite{Shell59} was the first in-place sorting algorithm
-with time complexity better than $O(n^2)$.
-Algorithms like selection and bubble sort use each comparison
-to resolve at most one inversion (an inversion is 
-a pair of keys in the input which appear in the wrong order) in the input, and
-it is well known that this restricts their average case running
-time to be in $O(n^2)$. Shellsort uses each comparison to
-resolve more than one inversion, giving better performance.
-
-Shellsort uses a sequence of decreasing integers $h_1, h_2, \ldots, h_k$
-with $h_k = 1$, called increments. 
-On the $i^{th}$ of its $k$ iterations, shellsort performs
-an insertion sort but treats \texttt{a[j]} and \texttt{a[j + $h_i$]} as though
-they were adjacent (rather than a normal insertion sort, which treats \texttt{a[j]} 
-and \texttt{a[j + 1]} as adjacent).
-
-The choice of increments influences shellsort's asymptotic behaviour.
-We used Gonnet's increments \cite{Gonnet+91} for our implementation, because
-we found them to be efficient empirically. Using
-Gonnet's increments $h_i = \lfloor \frac{5}{11} h_{i - 1} \rfloor$, with $h_1 = \lfloor \frac{5}{11}n \rfloor$ and the
-final increment being 1 as usual.
-
-We use a straightforward implementation of shellsort, except we perform the final iteration
-(i.e. the iteration in which we perform a standard insertion sort because the increment is 1) separately. 
-For this final insertion sort
-we extract $\min(a[0], \ldots, a[9])$ as a sentinel, removing the need
-for a bounds check in its inner-loop. This also simplifies
-the bounds checks on the outer-most loop of the other iterations.
+Shellsort \cite{Shell59} was the first in-place sorting algorithm with time
+complexity better than $O(n^2)$.  Algorithms like selection and bubble sort use
+each comparison to resolve at most one inversion (an inversion is a pair of keys
+in the input which appear in the wrong order) in the input, and it is well known
+that this restricts their average case running time to be in $O(n^2)$
+\cite{KnuthVol3_98}. Shellsort uses each comparison to resolve more than one
+inversion, giving better performance.
+
+Shellsort uses a sequence of decreasing integers $h_1, h_2, \ldots, h_k$ with
+$h_k = 1$, called increments.  On the $i^{th}$ of its $k$ iterations, shellsort
+performs an insertion sort but treats \texttt{a[j]} and \texttt{a[j + $h_i$]} as
+though they were adjacent (rather than a normal insertion sort, which treats
+\texttt{a[j]} and \texttt{a[j + 1]} as adjacent).
+
+The choice of increments influences shellsort's asymptotic behaviour.  We used
+Gonnet's increments \cite{Gonnet+91} for our implementation, because we found
+them to be efficient empirically. Using Gonnet's increments $h_i = \lfloor
+\frac{5}{11} h_{i - 1} \rfloor$, with $h_1 = \lfloor \frac{5}{11}n \rfloor$ and
+the final increment being 1 as usual.
+
+We use a straightforward implementation of shellsort, except we perform the
+final iteration (i.e. the iteration in which we perform a standard insertion
+sort because the increment is 1) separately.  For this final insertion sort we
+extract $\min(a[0], \ldots, a[9])$ as a sentinel, removing the need for a bounds
+check in its inner-loop. This also simplifies the bounds checks on the
+outer-most loop of the other iterations.
 
 \subsection{Branch Prediction Results}
+
+The behaviour of the comparison branches in shellsort is simple to analyse.
+There are $\lfloor \log_\frac{11}{5} n \rfloor$ passes, with each pass
+performing an insertion sort (using the current increment). As we saw in Section
+\ref{insertion_sort} insertion sort causes about one branch misprediction per
+key, thus we expect about $n \lfloor \log_\frac{11}{5}n \rfloor$ mispredictions
+in total, and about $\lfloor \log_\frac{11}{5}n \rfloor$ mispredictions per key.
+
+Figure \ref{shellsort_results_figs}(a) shows the branch mispredictions per key
+on average for our shellsort implementations, with bimodal and two-level
+adaptive predictors.  The scale on the x-axis is logarithmic, and the plots are
+nearly straight lines, as we would expect if the branch mispredictions per key
+are given by $\lfloor \log_\frac{11}{5} n \rfloor$. For example, the final
+measurement on the x-axis is for a set size of 4194304 and here we observe from
+the plot approximately 18 misses per key on average, and 
+$\lfloor \log_\frac{11}{5} 4194304 \rfloor = 19$.
+
+The comparison branch of the insertion sort inner-loop of shellsort are very
+unpredictable, unlike the nearly perfectly predictable branches of a standard
+insertion sort.  We measured a predictability of about 50\% on average for this
+branch using a simulated bimodal predictor.  A static predictor which predicts
+all branches as taken would be ideal for predicting the branches of the
+insertion sorts performed by shellsort.  We found that on average, using
+Gonnet's increments, a key moves 0.9744 places per iteration (note that a move
+is of $h$ positions, where $h$ is the current increment). Thus, although the
+insertion sort branch will still cause just a single misprediction per key, its
+outcome is not nearly as predictable as a standard insertion sort, where keys on
+average are moved a larger number of times per inner-loop iteration.
+
+One might expect shellsort's cache performance to be poor, since it repeatedly
+accesses distantly separated keys in its input while the increments are large.
+Each such access could potentially cause a cache miss. However, the fact that on
+average keys move just 0.9744 places per iteration implies that large numbers of
+references in distant parts of the input (which could lead to cache misses) are
+not generally necessary.  In fact, our shellsort implementation has better cache
+performance in both the level 1 and level 2 cache than for our basic heapsort
+implementation as Figure \ref{shellsort_results_figs}(b), and Figure
+\ref{shellsort_results_figs}(c) shows.  Indeed, our shellsort implementation
+performs significantly better than our basic heapsort implementation as Figure
+\ref{shellsort_results_figs}(d) shows (see Section \ref{heapsort} for further
+details on our heapsort implementations).
+
+\section{Heapsort}
+\label{heapsort}
+
+Another well known general purpose sorting algorithm is heapsort
+\cite{Williams64}. Heapsort's running time is $O(n \lg n)$.  Heapsort begins by
+constructing a \textit{heap}: a binary tree in which every level except possibly
+the deepest is entirely filled, with the deepest level filled from the left. In
+addition, a heap must also satisfy the \textit{heap property}: A key is
+associated with each node that is always at least as large as the keys
+associated with all its children nodes. The sequence of keys resulting from a
+pre-order traversal of a heap's nodes can be used to represent it unambiguously
+in an array.
 
@@
 (a) & (b) \\
+\includegraphics[width=0.5\textwidth]{plots/shell_heap_L2_misses.eps} & \includegraphics[width=0.5\textwidth]{plots/shell_heap_cycles.eps} \\
+(a) & (b) \\
 \end{tabular}
-\caption{(a) Shows the average branches misses per key for shellsort using a bimodal and two-level adaptive predictor, both having 4096 table entries.
-On this occasion, the two-level adaptive predictor out-performs the simpler bimodal predictor. \texttt{sim-bpred}
-was used to simulate these branch predictors.
-(b) Shows that the average number of cache misses per key of our basic heapsort implementation
-is higher than for our basic shellsort implementation. Shellsort incurs only a small number of cache misses
-because on average a key moves just 0.9744 places per iteration. These results use \texttt{sim-cache} to simulate
-an 8KB direct mapped cache with 32 byte cache lines.}
+\caption{(a) Shows the average branches misses per key for shellsort using a
+bimodal and two-level adaptive predictor, both having 4096 table entries.  On
+this occasion, the two-level adaptive predictor out-performs the simpler bimodal
+predictor. \texttt{sim-bpred} was used to simulate these branch predictors.  (b)
+and (c) show that our basic heapsort implementation has inferior cache
+performance at both levels of the cache compared to our shellsort
+implementation. Shellsort incurs only a small number of cache misses because on
+average a key moves just 0.9744 places per iteration. These results use
+\texttt{sim-cache} to simulate an 8KB direct mapped level 1 cache and a 2MB
+direct mapped level 2 cache both with 32 byte cache lines. Finally (d) shows
+that our shellsort implementation out-performs our heapsort implementation,
+these results were measured using Pentium 4 hardware performance counters.}
 \label{shellsort_results_figs}
@@
 
-The behaviour of the comparison branches in shellsort is
-simple to analyse. There are $\lfloor \log_\frac{11}{5} n \rfloor$ passes, with
-each pass performing an insertion sort (using the current
-increment). As we saw in Section \ref{insertion_sort} insertion
-sort causes about one branch misprediction per key, thus
-we expect about $n \lfloor \log_\frac{11}{5}n \rfloor$ mispredictions in total, and
-about $\lfloor \log_\frac{11}{5}n \rfloor$ mispredictions per key.
-
-Figure \ref{shellsort_results_figs}(a)
-shows the branch mispredictions per key on average for our shellsort
-implementations, with bimodal and two-level adaptive predictors.
-The scale on the x-axis is logarithmic, and the plots are nearly
-straight lines, as we would expect if the branch mispredictions per key
-are given by $\lfloor \log_\frac{11}{5} n \rfloor$. For example,
-the final measurement on the x-axis is for a set size of 4194304 and
-here we observe from the plot approximately 18 misses per key on average,
-and $\lfloor \log_\frac{11}{5} 4194304 \rfloor = 19$.
-
-The comparison branch of the insertion sort inner-loop of shellsort are very unpredictable, unlike the 
-nearly perfectly predictable branches of a standard insertion sort.
-We measured a predictability of about 50\% on average for this branch using a simulated bimodal
-predictor.
-A static predictor which predicts all branches as taken would be ideal for predicting the branches of the
-insertion sorts performed by shellsort.
-We found that on average, using Gonnet's increments, a key moves 
-0.9744 places per iteration (note that a move is of $h$ positions, 
-where $h$ is the current increment). Thus, although the insertion
-sort branch will still cause just a single misprediction per key, 
-its outcome is not nearly as predictable as a standard insertion
-sort, where keys on average are moved a larger number of times
-per inner-loop iteration.
-
-One might expect shellsort's cache performance to be poor, since it repeatedly accesses
-distantly separated keys in its input while the increments are large. Each such access
-could potentially cause a cache miss. However, the fact that on average 
-keys move just 0.9744 places per iteration
-implies that large numbers of references in distant parts
-of the input (which could lead to cache misses) are not generally necessary.
-In fact, the number of cache misses for our shellsort implementations, although significant,
-is smaller than for our basic heapsort implementation as Figure \ref{shellsort_results_figs}(b)
-shows. For our input sizes, which were up to $2^{22}$ keys, our shellsort implementations were on
-average about 50\% faster than the best of our heapsort implementations (see Section \ref{heapsort} for
-further details on our heapsort implementations).
-
-\section{Heapsort}
-\label{heapsort}
+An unordered array can be transformed into a heap in $O(n)$ time, using the
+approach of Floyd \citeyear{Floyd64}.  Given a node whose children are heaps but
+who may itself violate the heap property, an iteration of Floyd's approach swaps
+the node with the larger of its children, and then repeats the same procedure
+with the subtree rooted at that child, until the heap property is satisfied or a
+leaf is reached. We refer to this as ``sifting down'' a node.  To build the
+entire heap, we first sift-down the deepest nodes having children, followed by
+their parents, and so on until we reach the root.
+
+Once the heap is built, we gradually destroy it in order to sort. We iterate the
+code below $n - 1$ times.
+
+\begin{verbatim}
+    tmp = a[0];
+    a[0] = a[--n];
+    a[n] = tmp;    
+    sift_down(a, 0, n);
+\end{verbatim}
+
+\noindent On the $i^{th}$ iteration we swap the largest key, \texttt{a[0]}, with
+\texttt{a[n - i]}.  After each swap, the heap property is restored by sifting
+down the new root of the heap (an operation with worst case time $O(\lg n)$).
+After iteration $i$ the heap contains $n - i$ keys, and is found in
+\texttt{a[0..n - i - 1]}. Meanwhile \texttt{a[n - i..n - 1]} contains the $i$
+largest keys of the input from smallest to largest.
+
+\subsection{Heapsort Variations}
+
+For our base heapsort implementation, we used the approach described in the
+previous section, but removed a bounds check from the sift down procedure by
+introducing a sentinel at the cost of no extra instructions. The following
+optimization reduced the instruction count by about 10\% on average compared to
+an implementation of sift-down which uses a bounds check. We have not seen it
+described elsewhere. In a heap, exactly one node has a single left child, all
+other nodes have either two children or zero children. On every sift-down, we
+compare each node which has children with each of its children. As a result of
+the node with a single child, on every sift-down we must check that the node
+being compared to its children has a right child as well as a left child. To
+avoid this check while building the heap, we can just insert a maximal extra
+node at the end of the heap. When destroying the heap however, every second
+iteration requires the insertion of this extra node. To avoid this the code for
+destroying the heap shown in the previous section is changed to
+
+\begin{verbatim}
+    tmp = a[0];
+    a[0] = a[--n];
+    sift_down(a, 0, n);
+    a[n] = tmp;
+\end{verbatim}
+
+\noindent 
+By not modifying \texttt{a[n - 1]} until after sifting down, it acts as a
+sentinel for sift-down. This removes the need for the bounds check without
+introducing any extra instructions.
+
+Aside from this optimization, we also applied caching optimizations of LaMarca
+and Ladner \citeyear{LaMarca96b} to give several cache-optimized versions of the
+algorithm. Firstly, the heap can be padded so that siblings always reside in the
+same cache line. Another cache optimization is realised by abandoning Floyd's
+heap building method and simply repeatedly growing the heap one array element at
+a time, sifting each new key upwards as needed. This greatly improves the
+temporal locality of the algorithm when the data does not fit entirely within
+the cache.
+
+To increase the utilization of cache lines (and hence require fewer cache lines
+to be fetched) we can use a $k$-ary heap instead of the binary heap that has
+been assumed until now. When using a $k$-ary heap, every node having children
+(except perhaps one) has $k$ children. When sifting up or down a $k$-ary heap,
+the maximum (or minimum, depending on the heap property being used) child must
+be found, which requires $k - 1$ comparisons. Of course, as $k$ grows the sort
+degenerates to a selection sort. Note also that for a $k$-ary heap, the
+optimization described above, which avoids bounds checking while sifting down
+cannot be applied. However, it is still more efficient to add a sentinel on
+every second iteration of the heap destruction (since this introduces $O(n)$
+operations whereas the bounds check in the inner-loop introduces $O(n \lg n)$
+operations). For modest values of $k$ a performance improvement can be expected
+over a simple binary heap \cite{LaMarca96b}. We experimented with 4-heaps and
+8-heaps, combined with the aforementioned cache optimizations.
+
+\subsection{Branch Prediction Results}
+\label{heapsort_branch_results_text}
+
+Figure \ref{heapsort_results}(a) shows the average number of branches per key
+for our heapsort implementations. We note that the cache-optimized heapsort
+implementations execute fewer branches per key than our base heapsort
+implementation. Although these algorithms execute fewer branches per key on
+average, they cause a larger total number of branch mispredictions than our base
+heapsort implementation. Indeed, Figure \ref{heapsort_results}(b) shows that the
+base heapsort implementation incurs the fewest branch mispredictions of any of
+the algorithms.  This is despite the fact that it is executing more branches on
+average.  Figure \ref{heapsort_results}(b) also shows that the simpler bimodal
+predictors are generally no worse at predicting branches than the two-level
+adaptive predictors. With the exception of shellsort, this is in keeping with
+the trend we observed for the other sorting algorithms examined so far.
+
+Notwithstanding the base heapsort implementation having the lowest misprediction
+rate, Figure \ref{heapsort_results}(c) shows that the most efficient
+implementation on our hardware configuration to be the 8-heap variation of
+heapsort, followed by the 4-heap variation.  This suggests the improved cache
+performance of these algorithms outweighs their increased level of branch
+mispredictions shown in Figure \ref{heapsort_results}(b).
+
+When using a $k$-ary heap, finding the maximum (or minimum) child at a
+particular node causes the $i^{th}$ comparison branch to be taken with
+probability $1/(1 + i)$. The comparisons behave just like those in selection
+sort (see Section \ref{selection_sort}), and are therefore quite predictable.
+For our 8-heap we unrolled the loop to find the minimum child into 7 separate
+branches.  Figure \ref{heapsort_results}(d) shows the behaviour of these 7
+branches, for an unrolled loop.  As expected, the later branches are gradually
+taken less often on average, with a corresponding increase in their
+predictability.  Unrolling this loop not only removes the loop over-head, but is
+also experimentally observed to improve the average prediction accuracy by about
+2\% \cite{BiggarGregg05}.  The average steady state predicitability in the
+unrolled case, is given by
+
+\[
+\frac{1}{7}\sum^7_{i=1}CP\left(\frac{1}{1 + i}\right) \approx 0.724
+\]
+
+\noindent Where $CP$ is the steady state predictability function introduced in
+Section \ref{bubble_sort}.  Without unrolling, the comparison branch is taken on
+average with probability $\frac{1}{7}\sum^7_{i=1}\frac{1}{i + 1}$, and its
+predictability can be estimated as
+
+\[
+CP\left(\frac{1}{7}\sum^7_{i=1}\frac{1}{1 + i} \right) \approx 0.706
+\]
+
+\noindent As is observed experimentally, the unrolled loop results in about a
+2\% greater prediction accuracy.
 
@@
 \end{tabular}
-\caption{Branch prediction and performance characteristics of our heapsort implementations. (a) Shows
-the average number of branches per key measured using \texttt{sim-bpred}. (b) Shows the average branch mispredictions per key for tables of 4096
-predictors, again measured using \texttt{sim-bpred}. The two-level adaptive (which used a 10-bit history register) and bimodal predictors have close to identical performance, and to avoid
-cluttering the diagram only one instance of a two-level adaptive predictor is shown, the remainder of the predictors are bimodal. (c) Shows the cycle counts of the algorithms measured using
-Pentium 4 hardware performance counters. Note that the Pentium 4 has 32 byte cache lines in its first level cache, and 64 byte cache lines in its
-second level cache.
-Finally (d) shows the predictability of the unrolled branches associated with an 8-heap, measured with our own software simulated bimodal predictor.}
+\caption{Branch prediction and performance characteristics of our heapsort
+implementations. (a) Shows the average number of branches per key measured using
+\texttt{sim-bpred}. (b) Shows the average branch mispredictions per key for
+tables of 4096 predictors, again measured using \texttt{sim-bpred}. The
+two-level adaptive (which used a 10-bit history register) and bimodal predictors
+have close to identical performance, and to avoid cluttering the diagram only
+one instance of a two-level adaptive predictor is shown, the remainder of the
+predictors are bimodal. (c) Shows the cycle counts of the algorithms measured
+using Pentium 4 hardware performance counters. Note that the Pentium 4 has 32
+byte cache lines in its first level cache, and 64 byte cache lines in its second
+level cache.  Finally (d) shows the predictability of the unrolled branches
+associated with an 8-heap, measured with our own software simulated bimodal
+predictor.}
 \label{heapsort_results}
@@
 
-Another well known general purpose sorting algorithm is heapsort \cite{Williams64}. Heapsort's running time is $O(n \lg n)$.
-Heapsort begins by constructing a \textit{heap}: a binary tree in which every level except possibly the deepest is entirely filled, with the deepest level filled from the
-left. In addition, a heap must also satisfy the \textit{heap property}: A key is associated with each node that is always at least as large as the keys 
-associated with all its children nodes. The sequence of keys resulting from a pre-order traversal of a heap's nodes can be used to represent it unambiguously in an array.
-
-An unordered array can be transformed into a heap in $O(n)$ time, using the approach of Floyd \citeyear{Floyd64}.
-Given a node whose children are heaps but who may itself
-violate the heap property, an iteration of Floyd's approach swaps the node with the larger of its children, and then repeats the same procedure 
-with the subtree rooted at that child, until the heap property is satisfied or a leaf is reached. We refer to this as ``sifting down'' a node. 
-To build the entire heap, we first sift-down the deepest nodes having children, followed by their
-parents, and so on until we reach the root.
-
-Once the heap is built, we gradually destroy it in order to sort. We iterate the code below $n - 1$ times.
-
-\begin{verbatim}
-    tmp = a[0];
-    a[0] = a[--n];
-    a[n] = tmp;    
-    sift_down(a, 0, n);
-\end{verbatim}
-
-\noindent
-On the $i^{th}$ iteration we swap the largest key, 
-\texttt{a[0]}, with \texttt{a[n - i]}. 
-After each swap, the heap property is restored by sifting down the new root of the heap (an operation with worst case time $O(\lg n)$). 
-After iteration $i$ the heap contains $n - i$ keys, and is found in \texttt{a[0..n - i - 1]}. Meanwhile 
-\texttt{a[n - i..n - 1]} contains the $i$ largest keys of the input from smallest to largest.
-
-\subsection{Heapsort Variations}
-
-For our base heapsort implementation, we used the approach described in the previous section, but removed a bounds check from the sift down procedure
-by introducing a sentinel at the cost of no extra instructions. The following optimization reduced the instruction count by about 10\%
-on average compared to an implementation of sift-down which uses a bounds check. We have not seen it described elsewhere. In a heap, exactly one node has a single left child, all 
-other nodes have either two children
-or zero children. On every sift-down, we compare each node which has children with each of its children. As a result of the node with a single child, on 
-every sift-down we
-must check that the node being compared to its children has a right child as well as a left child. To avoid this check while building the heap, we can just 
-insert
-a maximal extra node at the end of the heap. When destroying the heap however, every second iteration requires the insertion of 
-this extra node. To avoid this the code for destroying the heap shown in the previous section is changed to
-
-\begin{verbatim}
-    tmp = a[0];
-    a[0] = a[--n];
-    sift_down(a, 0, n);
-    a[n] = tmp;
-\end{verbatim}
-
-\noindent
-By not modifying \texttt{a[n - 1]} until after sifting down,
-it acts as a sentinel for sift-down. This removes the need for the bounds check without introducing any extra instructions.
-
-Aside from this optimization, we also applied caching optimizations of LaMarca and Ladner \citeyear{LaMarca96b} to give several cache-optimized versions of the
-algorithm. Firstly, the heap can be padded so that siblings
-always reside in the same cache line. Another cache optimization is realised by abandoning Floyd's heap building method and simply repeatedly growing the
-heap one array element at a time, sifting each new key upwards as needed. This greatly improves the temporal locality of the algorithm when the data
-does not fit entirely within the cache.
-
-To increase the utilization of cache lines (and hence require fewer cache lines to be
-fetched) we can use a $k$-ary heap instead of the binary heap that has been assumed until now. When using a $k$-ary heap, every node having children (except perhaps one)
-has $k$ children. When sifting up or down a $k$-ary heap, the maximum (or minimum, depending on the heap property being used) child must be found, which
-requires $k - 1$ comparisons. Of course, as $k$ grows the sort degenerates to a selection sort. Note also that for a $k$-ary heap, the optimization described
-above, which avoids bounds checking while sifting down cannot be applied. However, it is still more efficient to add a sentinel on every second iteration of the heap 
-destruction (since this introduces
-$O(n)$ operations whereas the bounds check in the inner-loop introduces $O(n \lg n)$ operations). For modest values of $k$ a performance improvement
-can be expected over a simple binary heap \cite{LaMarca96b}. We experimented with 4-heaps and 8-heaps, combined with the aforementioned cache optimizations.
-
-\subsection{Branch Prediction Results}
-\label{heapsort_branch_results_text}
-
-Figure \ref{heapsort_results}(a) shows the average number of branches per key for our heapsort implementations. We note that the cache-optimized heapsort
-implementations execute fewer branches per key than our base heapsort implementation. Although these algorithms execute fewer branches
-per key on average, they cause a larger total number of branch mispredictions than our base heapsort implementation. Indeed, Figure
-\ref{heapsort_results}(b) shows that the base heapsort implementation incurs the fewest branch mispredictions of any of the algorithms. 
-This is despite the fact that it is executing more branches on average.
-Figure \ref{heapsort_results}(b) also shows that the simpler
-bimodal predictors are generally no worse at predicting branches than the two-level adaptive predictors. With the exception of shellsort, this is in 
-keeping with the trend we observed for 
-the other sorting algorithms examined so far.
-
-Notwithstanding the base heapsort implementation having the lowest misprediction rate,
-Figure \ref{heapsort_results}(c) shows that the most efficient implementation on our hardware configuration to be the 8-heap variation of heapsort,
-followed by the 4-heap variation.
-This suggests the improved cache performance of these algorithms outweighs their increased level of branch mispredictions shown in 
-Figure \ref{heapsort_results}(b).
-
-When using a $k$-ary heap, finding the maximum (or minimum) child at a particular node causes the $i^{th}$ comparison branch to be taken
-with probability $1/(1 + i)$. The comparisons behave just like those in selection sort (see Section \ref{selection_sort}), and are therefore quite
-predictable.  For our 8-heap we unrolled the
-loop to find the minimum child into 7 separate branches. 
-Figure \ref{heapsort_results}(d) shows the behaviour of these 7 branches, for an unrolled loop.
-As expected, the later branches are gradually taken less often on average, with a corresponding increase in their predictability.
-Unrolling this loop
-not only removes the loop over-head, but is also experimentally observed to improve the average prediction accuracy by about 2\%. 
-The average steady state predicitability in the unrolled case, is given by
-
-\[
-\frac{1}{7}\sum^7_{i=1}CP\left(\frac{1}{1 + i}\right) \approx 0.724
-\]
-
-\noindent
-Where $CP$ is the steady state predictability function introduced in Section \ref{bubble_sort}.
-Without unrolling, the comparison branch is taken on average with probability $\frac{1}{7}\sum^7_{i=1}\frac{1}{i + 1}$, and its predictability
-can be estimated as
-
-\[
-CP\left(\frac{1}{7}\sum^7_{i=1}\frac{1}{1 + i} \right) \approx 0.706
-\]
-
-\noindent
-As is observed experimentally, the unrolled loop results in about a 2\% greater prediction accuracy.
-
 \section{Mergesort}
@@
 
-Mergesort \cite{KnuthVol3_98} is a another $O(n \lg n)$ sorting algorithm, which works by 
-repeatedly merging pairs of sorted lists into a single sorted list of twice the length. These merges can
-be done very simply in linear time and space. Mergesort's
-input can be thought of as $n$ sorted lists of length one. To simplify the description, we 
-assume the number of keys, $n$, is a power of two. Mergesort makes $\lg n$
-passes over the input where the $i^{th}$ pass merges sorted lists of length $2^{i - 1}$ into 
-sorted lists of length $2^i$.
+Mergesort \cite{KnuthVol3_98} is a another $O(n \lg n)$ sorting algorithm, which
+works by repeatedly merging pairs of sorted lists into a single sorted list of
+twice the length. These merges can be done very simply in linear time and space.
+Mergesort's input can be thought of as $n$ sorted lists of length one. To
+simplify the description, we assume the number of keys, $n$, is a power of two.
+Mergesort makes $\lg n$ passes over the input where the $i^{th}$ pass merges
+sorted lists of length $2^{i - 1}$ into sorted lists of length $2^i$.
 
@@
 
-For our base
-mergesort implementation we apply several optimizations.
-Firstly, our mergesort alternates the merge
-operation between two arrays to avoid unnecessary copying.
-Secondly, a bitonic sort as described by Sedgewick \citeyear{Sedgewick02},
-removes instructions from the inner-loop. A bitonic sort
-sorts every second list in reverse, merging then works
-from the two ends of the array containing the elements to
-be merged, terminating when the indices cross. Since the indices
-cannot go outside the bounds of the array unless they cross first, 
-this eliminates the need for a bounds check in the inner-loop.
-An insertion sort is also initially used to create
-sorted lists of length four, rather than allowing mergesort to
-sort these lists.
-
-We also analysed the branch prediction behaviour in several
-cache conscious mergesort variations, using the work of LaMarca and Ladner \citeyear{LaMarca97} as a reference.
-Mergesort is not in-place\footnote{Worst case linear time in-place merging is possible but is generally inefficient in practice, see for example Katajainen 
-\textit{et al} \citeyear{Katajainen+96}.}
-and uses an auxiliary array 
-for
-its merge operation. With a cache of size $C$, the largest input we can sort in the cache therefore has size $C/2$. To avoid conflict misses, the
-arrays should be allocated so that the cache line index of the start of the source array is $C/2$ from the cache line index of the start of the destination array. 
-This can
-be done for both the level 1 and level 2 cache.
-
-Since each key is only used once per pass
-in mergesort, if the input is larger than the cache then we will have no temporal re-use of data at all. To address this we can \textit{tile}
-mergesort: cache sized blocks of the input are mergesorted, and then merged as usual. Combined with alignment this variation is referred to as
-tiled mergesort. Tiling mergesort in this way does not fully address the problem of its poor temporal re-use, since the merging of the cache sized blocks in tiled 
-mergesort will exhibit no temporal re-use and again have very poor
-cache performance. To address this, a multi-way merge of the cache sized blocks can be used. A multi-way merge combines a number of sorted lists into a 
-single sorted
-list while reading through each of them only once. A heap can be used to help implement the multi-way merge efficiently. 
-This multi-mergesort variation has better cache performance. We refer to this combination of alignment and tiling with multi-way merging
-simply as multi-mergesort.
+For our base mergesort implementation we apply several optimizations.  Firstly,
+our mergesort alternates the merge operation between two arrays to avoid
+unnecessary copying.  Secondly, a bitonic sort as described by Sedgewick
+\citeyear{Sedgewick02}, removes instructions from the inner-loop. A bitonic sort
+sorts every second list in reverse, merging then works from the two ends of the
+array containing the elements to be merged, terminating when the indices cross.
+Since the indices cannot go outside the bounds of the array unless they cross
+first, this eliminates the need for a bounds check in the inner-loop.  An
+insertion sort is also initially used to create sorted lists of length four,
+rather than allowing mergesort to sort these lists.
+
+We also analysed the branch prediction behaviour in several cache conscious
+mergesort variations using the work of LaMarca and Ladner \citeyear{LaMarca97}
+as a reference, the details of our implementations differ slightly however
+\cite{BiggarGregg05}.  Mergesort is not in-place\footnote{Worst case linear time
+in-place merging is possible but is generally inefficient in practice, see for
+example Katajainen \textit{et al} \citeyear{Katajainen+96}.} and uses an
+auxiliary array for its merge operation. With a cache of size $C$, the largest
+input we can sort in the cache therefore has size $C/2$. To avoid conflict
+misses, the arrays should be allocated so that the cache line index of the start
+of the source array is $C/2$ from the cache line index of the start of the
+destination array.  This can be done for both the level 1 and level 2 cache.
+
+Since each key is only used once per pass in mergesort, if the input is larger
+than the cache then we will have no temporal re-use of data at all. To address
+this we can \textit{tile} mergesort: cache sized blocks of the input are
+mergesorted, and then merged as usual. Combined with alignment this variation is
+referred to as tiled mergesort. Tiling mergesort in this way does not fully
+address the problem of its poor temporal re-use, since the merging of the cache
+sized blocks in tiled mergesort will exhibit no temporal re-use and again have
+very poor cache performance. To address this, a multi-way merge of the cache
+sized blocks can be used. A multi-way merge combines a number of sorted lists
+into a single sorted list while reading through each of them only once. A heap
+can be used to help implement the multi-way merge efficiently.  This
+multi-mergesort variation has better cache performance. We refer to this
+combination of alignment and tiling with multi-way merging simply as
+multi-mergesort.
 
@@
 
-Figure \ref{mergesort_branch_results}(a) shows the average number of branches per key executed by
-our mergesort implementations. Up to $2^{18}$ keys can be sorted in the simulated 2 MB level 2 cache
-(each key is 4 bytes, and mergesort is not in-place). As a result, when the number of keys being 
-sorted reaches $2^{19}$ there is a dramatic increase in the number of branches executed by multi-mergesort. This
-is because a multi-way merge begins to be used. Figure \ref{mergesort_branch_results}(b) 
-shows that there is a corresponding increase in the number of branch
-mispredictions caused by multi-mergesort as a result of the multi-way merge.
-
-Many of the additional comparison branches introduced by
-the multi-way merge are associated with a $k$-ary heap, and are therefore quite predictable, as we saw in
-Section \ref{heapsort_branch_results_text}. In addition, two-level adaptive predictors outperform bimodal predictors on multi-mergesort's branches.
-Figure \ref{mergesort_branch_results}(b) shows this. Both predictors have 4096 entry tables. Varying the table size
-of the bimodal predictor between 2048 and 8192 entries did not result in any reduction in the number of branch mispredictions.
-On the other hand, increasing the table size of the two-level adaptive predictor gave incremental
-improvements in branch predictability. This suggests that the two-level adaptive predictor is
-able to exploit correlations between branch outcomes which occur during the multi-way merge.
-
-Figure \ref{mergesort_branch_results}(c) shows the simulated level 2 cache misses per key for our mergesort
-implementations. It is noteworthy that although multi-mergesort has the best cache performance, its cache
-performance is only slightly better than tiled mergesort. However, as the set size increases the difference in cache
-performance will slowly become more pronounced, as the trend in Figure \ref{mergesort_branch_results}(c) suggests. 
-Figure \ref{mergesort_branch_results}(d) shows the cycles per key measured for our mergesort implementations
-running on actual hardware. These results show that our less cache conscious tiled mergesort implementation
-out-performs multi-mergesort. This is likely because the set sizes are not large enough to make the superior
-cache performance of multi-mergesort the dominating factor in the comparative performances. Instead, for the 
-data sizes we used, multi-mergesort's increased instruction count and branch mispredictions
-cause its performance to be inferior to tiled mergesort.
+Figure \ref{mergesort_branch_results}(a) shows the average number of branches
+per key executed by our mergesort implementations. Up to $2^{18}$ keys can be
+sorted in the simulated 2 MB level 2 cache (each key is 4 bytes, and mergesort
+is not in-place). As a result, when the number of keys being sorted reaches
+$2^{19}$ there is a dramatic increase in the number of branches executed by
+multi-mergesort. This is because a multi-way merge begins to be used. Figure
+\ref{mergesort_branch_results}(b) shows that there is a corresponding increase
+in the number of branch mispredictions caused by multi-mergesort as a result of
+the multi-way merge.
+
+Many of the additional comparison branches introduced by the multi-way merge are
+associated with a $k$-ary heap, and are therefore quite predictable, as we saw
+in Section \ref{heapsort_branch_results_text}. In addition, two-level adaptive
+predictors outperform bimodal predictors on multi-mergesort's branches.  Figure
+\ref{mergesort_branch_results}(b) shows this. Both predictors have 4096 entry
+tables. Varying the table size of the bimodal predictor between 2048 and 8192
+entries did not result in any reduction in the number of branch mispredictions.
+On the other hand, increasing the table size of the two-level adaptive predictor
+gave incremental improvements in branch predictability. This suggests that the
+two-level adaptive predictor is able to exploit correlations between branch
+outcomes which occur during the multi-way merge.
+
+Figure \ref{mergesort_branch_results}(c) shows the simulated level 2 cache
+misses per key for our mergesort implementations. It is noteworthy that although
+multi-mergesort has the best cache performance, its cache performance is only
+slightly better than tiled mergesort. However, as the set size increases the
+difference in cache performance will slowly become more pronounced, as the trend
+in Figure \ref{mergesort_branch_results}(c) suggests.  Figure
+\ref{mergesort_branch_results}(d) shows the cycles per key measured for our
+mergesort implementations running on actual hardware. These results show that
+our less cache conscious tiled mergesort implementation out-performs
+multi-mergesort. This is likely because the set sizes are not large enough to
+make the superior cache performance of multi-mergesort the dominating factor in
+the comparative performances. Instead, for the data sizes we used,
+multi-mergesort's increased instruction count and branch mispredictions cause
+its performance to be inferior to tiled mergesort.
 
@@
 \end{tabular}
-\caption{(a) Shows the branches per key executed by our mergesort variations. The sudden increase in the number of executed branches for multi-mergesort
-results from the use of the multi-way merge when the data no longer fits within the cache. (b) Shows the number of mispredictions per key, all predictors
-had 4096 table entries. Bimodal predictors were used for base and tiled mergesort. These results were generated using \texttt{sim-bpred}, the
-two-level adaptive predictor had a 10-bit history register.
-(c) Shows the level 2 cache misses per key, base mergesort has by far the worst cache behaviour, while multi-mergesort
-has slightly better cache performance than tiled mergesort. These results were generated by using \texttt{sim-cache} to simulate a
-direct mapped 2MB cache with 32-byte cache lines. Finally (d) shows the cycles per key of our algorithms measured using Pentium 4 hardware performance
-counters. Tiled mergesort performs the best out of the algorithms, despite its inferior cache performance compared to multi-mergesort.}
+\caption{(a) Shows the branches per key executed by our mergesort variations.
+The sudden increase in the number of executed branches for multi-mergesort
+results from the use of the multi-way merge when the data no longer fits within
+the cache. (b) Shows the number of mispredictions per key, all predictors had
+4096 table entries. Bimodal predictors were used for base and tiled mergesort.
+These results were generated using \texttt{sim-bpred}, the two-level adaptive
+predictor had a 10-bit history register.  (c) Shows the level 2 cache misses per
+key, base mergesort has by far the worst cache behaviour, while multi-mergesort
+has slightly better cache performance than tiled mergesort. These results were
+generated by using \texttt{sim-cache} to simulate a direct mapped 2MB cache with
+32-byte cache lines. Finally (d) shows the cycles per key of our algorithms
+measured using Pentium 4 hardware performance counters. Despite
+multi-mergesort's superior cache performance, its heightened instruction count
+and branch mispredictions mean that the less cache conscious tiled mergesort
+performs the best out of the algorithms.}
 \label{mergesort_branch_results}
@@
 Quicksort \cite{Hoare62} is an algorithm which selects a key from its input
-called the pivot and then partitions the other keys into two sets,
-one set containing keys at most equal to the pivot and another
-containing keys at least equal to the pivot. The keys of these sets should
-appear respectively to the left and right of pivot in sorted order.
-Therefore they can be sorted independently, each using quicksort.
+called the pivot and then partitions the other keys into two sets, one set
+containing keys at most equal to the pivot and another containing keys at least
+equal to the pivot. The keys of these sets should appear respectively to the
+left and right of pivot in sorted order.  Therefore they can be sorted
+independently, each using quicksort.
 
@@
 
-For our basic quicksort implementation, we make use of a number
-of improvements suggested by Sedgewick \citeyear{Sedgewick78}. Firstly, we
-use an explicit stack instead of recursing, to save space. Secondly, after
-partitioning, we always quicksort the set with the smaller number of elements first. 
-This reduces the worst case stack space to $O(\lg n)$ from $O(n)$.
-Thirdly, inputs of less than a certain \texttt{THRESHOLD} (typically \texttt{THRESHOLD} is 10) in size are not quicksorted
-at all. Instead, insertion sort is run as a post-pass over the output of quicksort.
-The smallest of the first \texttt{THRESHOLD} keys is used as a sentinel for insertion sort.
-Finally, we use median-of-3 pivoting, greatly reducing the chances of very unbalanced
-partitioning and also removing the need for bounds checks in the inner-loop, thus
-greatly reducing the instruction count.
+For our basic quicksort implementation, we make use of a number of improvements
+suggested by Sedgewick \citeyear{Sedgewick78}. Firstly, we use an explicit stack
+instead of recursing, to save space. Secondly, after partitioning, we always
+quicksort the set with the smaller number of elements first.  This reduces the
+worst case stack space to $O(\lg n)$ from $O(n)$.  Thirdly, inputs of less than
+a certain \texttt{THRESHOLD} (typically \texttt{THRESHOLD} is 10) in size are
+not quicksorted at all. Instead, insertion sort is run as a post-pass over the
+output of quicksort.  The smallest of the first \texttt{THRESHOLD} keys is used
+as a sentinel for insertion sort.  Finally, we use median-of-3 pivoting, greatly
+reducing the chances of very unbalanced partitioning and also removing the need
+for bounds checks in the inner-loop, thus greatly reducing the instruction
+count.
 
@@
 implementations. We also investigated the branch prediction behaviour of cache
-conscious quicksort implementations. LaMarca and Ladner \citeyear{LaMarca97} suggest that instead of running
-insertion sort as a post-pass over the quicksorted input, each time quicksort
-is presented with an input of size less than \texttt{THRESHOLD} it immediately
-insertion sorts it. Although this increases the instruction count, it is more
-than compensated for by the resultant drop in cache misses. We refer to this
-as memory-tuned quicksort.
-
-When quicksort's input fits within a particular level of the cache quicksort
-has good cache properties at that level. To combat a dramatic increase
-in cache misses when the input does not fit in the cache, LaMarca and Ladner also suggest multi-quicksort. 
-Multi-quicksort chooses a set of pivots and performs a multi-way partition which divides the input into
-cache sized containers. Each container is then quicksorted. We leave
-the precise details of performing a multi-way partition to elsewhere \cite{LaMarca96a,BiggarGregg05}, it is sufficient
-to note that we can decide which container a key should be placed in by using either a sequential
-or binary search.
+conscious quicksort implementations. LaMarca and Ladner \citeyear{LaMarca97}
+suggest that instead of running insertion sort as a post-pass over the
+quicksorted input, each time quicksort is presented with an input of size less
+than \texttt{THRESHOLD} it immediately insertion sorts it. Although this
+increases the instruction count, it is more than compensated for by the
+resultant drop in cache misses. We refer to this as memory-tuned quicksort.
+
+When quicksort's input fits within a particular level of the cache quicksort has
+good cache properties at that level. To combat a dramatic increase in cache
+misses when the input does not fit in the cache, LaMarca and Ladner also suggest
+multi-quicksort.  Multi-quicksort chooses a set of pivots and performs a
+multi-way partition which divides the input into cache sized containers. Each
+container is then quicksorted. We leave the precise details of performing a
+multi-way partition to elsewhere \cite{LaMarca96a,BiggarGregg05}, it is
+sufficient to note that we can decide which container a key should be placed in
+by using either a sequential or binary search.
 
@@
 \end{tabular}
-\caption{Overview of branch prediction behaviour in our quicksort implementations. Every figure shows the behaviour of the $i$ and $j$ branches when
-using a median-of-3 pivot. As described in Section \ref{quicksort_branch_results_text}, these branches are about 60\% biased and 64\% predictable when using
-the median-of-3. In (a) the \texttt{median} branch is the combined results of the branches which compute the median-of-3 (these branches are also executed for (b), (c) and (d)).
-Comparing (a) with (b), (c) and (d), we see that the \texttt{insertion} branch associated with its insertion sort is slightly less predictable than in 
-the other variations. This is due to it running as a post-pass. Finally, comparing (c) with (d) we see that the binary search branches of (c), 
-\texttt{binary left} and \texttt{binary right}, are very unpredictable compared to the \texttt{sequential} branch of (d).
-}
+\caption{Overview of branch prediction behaviour in our quicksort
+implementations. Every figure shows the behaviour of the $i$ and $j$ branches
+when using a median-of-3 pivot. As described in Section
+\ref{quicksort_branch_results_text}, these branches are about 60\% biased and
+64\% predictable when using the median-of-3. In (a) the \texttt{median} branch
+is the combined results of the branches which compute the median-of-3 (these
+branches are also executed for (b), (c) and (d)).  Comparing (a) with (b), (c)
+and (d), we see that the \texttt{insertion} branch associated with its insertion
+sort is slightly less predictable than in the other variations. This is due to
+it running as a post-pass. Finally, comparing (c) with (d) we see that the
+binary search branches of (c), \texttt{binary left} and \texttt{binary right},
+are very unpredictable compared to the \texttt{sequential} branch of (d).  }
 \label{branch_prediction_quicksort}
@@
 \end{verbatim}
-\caption{Quicksort's partition inner-loop. We refer to the inner while loops
-as the $i$ and $j$ loops. We refer to their associated branches as
-the $i$ and $j$ branches respectively.}
+\caption{Quicksort's partition inner-loop. We refer to the inner while loops as
+the $i$ and $j$ loops. We refer to their associated branches as the $i$ and $j$
+branches respectively.}
 \label{partition_code}
@@
 
-Figure \ref{branch_prediction_quicksort} shows the branch prediction results of our quicksort implementations.
-In all figures, the most important branches are the $i$ and $j$ branches (see Figure \ref{partition_code}), since these 
-are in quicksort's inner-loop, and are executed more often than any of the other branches (with the exception of the \texttt{sequential} branch
-of Figure \ref{branch_prediction_quicksort}(d)).
-
-The branch prediction behaviour of basic quicksort and memory-tuned quicksort are similar.
-As can be seen in Figure \ref{branch_prediction_quicksort}(b) 
-quicksort has a slightly lower branch misprediction rate on the \texttt{insertion} branch. This
-is because a basic quicksort implementation, which performs insertion sort as a post-pass,
-will generate branch mispredictions on keys chosen as pivots since they never require movement.
-In practice, the over-all reduction in branch misprediction rate from 35\% in basic quicksort
-to 34\% in memory-tuned quicksort is probably offset by the branch misprediction rate of the
-outer-loop exit in the many insertion sorts executed by memory-tuned quicksort. However,
-memory-tuned quicksort has better caching properties, and so from the point of view of both
-branch prediction and caching behaviour this optimization is to be recommended.
-
-Multi-quicksort results in a 19\% reduction in the number of iterations of the $i$ and $j$
-loops of quicksort's partition inner-loop, 
-though their misprediction rate remains the same at 37\%. Figure \ref{branch_prediction_quicksort}(c) shows branch behaviour
-for multi-quicksort when using binary search. The binary search branches are highly unpredictable,
-being incorrectly predicted 43\% of the time. If multi-quicksort uses sequential search, the
-number of executed branches dramatically increases, but these
-branches are correctly predicted 92\% of the time, as shown in Figure \ref{branch_prediction_quicksort}(d).
-
-The implementations of multi-quicksort have the best cache performance of the quicksort variations, as Figure \ref{quicksort_plots}(a) shows.
-However, the multi-quicksort implementations also introduce many extra instructions per key as Figure \ref{quicksort_plots}(b) shows. Naturally, the 
-sequentially searched version introduces more extra instructions than the binary searched version. However, as described in the previous paragraph,
-the branches introduced by binary search are highly unpredictable. As a result, multi-quicksort using binary search has more branch mispredictions than the
-sequential searched version (see Figure \ref{quicksort_plots}(c)), consequently, despite having identical cache performance, binary searched multi-quicksort
-performs worse than the sequentially searched multi-quicksort, as shown in Figure \ref{quicksort_plots}(d).
-Of course, on very large data sets (larger than the $2^{22}$ keys we experimented with), the
-reduced instruction count of binary searched multi-quicksort would eventually allow it to out-perform sequentially searched multi-quicksort.
-
-As Figure \ref{quicksort_plots}(d) shows, our base quicksort implementation performs the best of all. It seems the improved cache performance of
-multi-quicksort is not enough for it to out-perform base quicksort on the data set sizes we experimented with. Examining the trends of 
-Figure \ref{quicksort_plots}(a), it seems likely that multi-quicksort would eventually out-perform base quicksort due to its improved cache performance.
-The fact that base quicksort out-performs memory-tuned quicksort is very difficult to explain. In the simulations memory-tuned quicksort has the same
-cache performance and instruction count, as shown in Figures \ref{quicksort_plots}(a) and \ref{quicksort_plots}(b). Also in the simulations, 
-memory-tuned quicksort has slightly fewer branch mispredictions than base quicksort, as Figure \ref{quicksort_plots}(c) shows. Despite this, base 
-quicksort slightly out-performs memory-tuned quicksort on our non-simulated hardware test as Figure \ref{quicksort_plots}(d) shows.
+Figure \ref{branch_prediction_quicksort} shows the branch prediction results of
+our quicksort implementations.  In all figures, the most important branches are
+the $i$ and $j$ branches (see Figure \ref{partition_code}), since these are in
+quicksort's inner-loop, and are executed more often than any of the other
+branches (with the exception of the \texttt{sequential} branch of Figure
+\ref{branch_prediction_quicksort}(d)).
+
+The branch prediction behaviour of basic quicksort and memory-tuned quicksort
+are similar.  As can be seen in Figure \ref{branch_prediction_quicksort}(b)
+quicksort has a slightly lower branch misprediction rate on the
+\texttt{insertion} branch. This is because a basic quicksort implementation,
+which performs insertion sort as a post-pass, will generate branch
+mispredictions on keys chosen as pivots since they never require movement.  In
+practice, the over-all reduction in branch misprediction rate from 35\% in basic
+quicksort to 34\% in memory-tuned quicksort is probably offset by the branch
+misprediction rate of the outer-loop exit in the many insertion sorts executed
+by memory-tuned quicksort. However, memory-tuned quicksort has better caching
+properties, and so from the point of view of both branch prediction and caching
+behaviour this optimization is to be recommended.
+
+Multi-quicksort results in a 19\% reduction in the number of iterations of the
+$i$ and $j$ loops of quicksort's partition inner-loop, though their
+misprediction rate remains the same at 37\%. Figure
+\ref{branch_prediction_quicksort}(c) shows branch behaviour for multi-quicksort
+when using binary search. The binary search branches are highly unpredictable,
+being incorrectly predicted 43\% of the time. If multi-quicksort uses sequential
+search, the number of executed branches dramatically increases, but these
+branches are correctly predicted 92\% of the time, as shown in Figure
+\ref{branch_prediction_quicksort}(d).
+
+The implementations of multi-quicksort have the best cache performance of the
+quicksort variations, as Figure \ref{quicksort_plots}(a) shows.  However, the
+multi-quicksort implementations also introduce many extra instructions per key
+as is seen in Figure \ref{quicksort_plots}(b).  Naturally, the sequentially
+searched version introduces more extra instructions than the binary searched
+version. However, as described in the previous paragraph, the branches
+introduced by binary search are highly unpredictable. As a result,
+multi-quicksort using binary search has more branch mispredictions than the
+sequential searched version (see Figure \ref{quicksort_plots}(c)), consequently,
+despite having identical cache performance, binary searched multi-quicksort
+performs worse than the sequentially searched multi-quicksort, as shown in
+Figure \ref{quicksort_plots}(d). Of course, on very large data sets (larger than
+the $2^{22}$ keys we experimented with), the reduced instruction count of binary
+searched multi-quicksort would eventually allow it to out-perform sequentially
+searched multi-quicksort.
+
+As Figure \ref{quicksort_plots}(d) shows, our base quicksort implementation
+performs the best of all. It seems the improved cache performance of
+multi-quicksort is not enough for it to out-perform base quicksort on the data
+set sizes we experimented with. Examining the trends of Figure
+\ref{quicksort_plots}(a), it seems likely that multi-quicksort would eventually
+out-perform base quicksort due to its improved cache performance.  The fact that
+base quicksort out-performs memory-tuned quicksort is very difficult to explain.
+In the simulations memory-tuned quicksort has the same cache performance and
+instruction count, as shown in Figures \ref{quicksort_plots}(a) and
+\ref{quicksort_plots}(b). Also in the simulations, memory-tuned quicksort has
+slightly fewer branch mispredictions than base quicksort, as Figure
+\ref{quicksort_plots}(c) shows. Despite this, base quicksort slightly
+out-performs memory-tuned quicksort on our non-simulated hardware test as Figure
+\ref{quicksort_plots}(d) shows.
 
@@
 \end{tabular}
-\caption{(a) Shows the 2nd level cache misses for our quicksort implementations. As is expected, the multi-quicksort implementations have the best
-cache performance. These results were generated using \texttt{sim-cache} to simulate a 2MB direct-mapped cache with 32-byte cache lines.
-(b) Shows the instruction counts of our quicksort implementations, when the data sets no longer fit within the cache the multi-quicksort implementations
-show a large increase in instruction count, these results were also generated using SimpleScalar. (c) Shows the simulated branch mispredictions per key
-for our quicksort implementations. All predictors use 4096 table entries and are bimodal predictors except where indicated. Multi-quicksort using
-binary search is the only implementation which gives varying results from the use of a two-level adaptive predictor. These results were generated
-using \texttt{sim-bpred}. (d) Shows the cycles per key for our quicksort implementations measured using Pentium 4 hardware performance counters.
-}
+\caption{(a) Shows the 2nd level cache misses for our quicksort implementations.
+As is expected, the multi-quicksort implementations have the best cache
+performance. These results were generated using \texttt{sim-cache} to simulate a
+2MB direct-mapped cache with 32-byte cache lines.  (b) Shows the instruction
+counts of our quicksort implementations, when the data sets no longer fit within
+the cache the multi-quicksort implementations show a large increase in
+instruction count, these results were also generated using SimpleScalar. (c)
+Shows the simulated branch mispredictions per key for our quicksort
+implementations. All predictors use 4096 table entries and are bimodal
+predictors except where indicated. Multi-quicksort using binary search is the
+only implementation which gives varying results from the use of a two-level
+adaptive predictor. These results were generated using \texttt{sim-bpred}. (d)
+Shows the cycles per key for our quicksort implementations measured using
+Pentium 4 hardware performance counters.  }
 \label{quicksort_plots}
@@
 
-The choice of pivot element has a strong influence on the behaviour of the $i$ and 
-$j$ branches (see Figure \ref{partition_code}). Clearly, if the pivot is chosen as the median of the input then
-either branch is equally likely to be taken (i.e. the loop executes) as not taken (i.e.
-the loop exits). In this case, the branches are also on average approximately 50\% predictable. 
-However, on average over random input, with a fixed choice of pivot the $i$ and
-$j$ branches are about 66\% biased (that is, they are taken about 66\% of the time). It is
-straightforward to show why this is so. Assume that the data to be partitioned is a random permutation of $\lbrace 1, \ldots, n \rbrace$.
-If the chosen pivot has rank $q$ then the $i$ branch will be taken with probability $(q - 1)/n$.
-Moreover, the branch will be executed (but not necessarily taken) a total of $q$ times, since there are exactly $q - 1$ elements
-to the left of the pivot. The average bias is given by averaging over all 
-possible choices of pivot $q$, thus
+The choice of pivot element has a strong influence on the behaviour of the $i$
+and $j$ branches (see Figure \ref{partition_code}). Clearly, if the pivot is
+chosen as the median of the input then either branch is equally likely to be
+taken (i.e. the loop executes) as not taken (i.e.  the loop exits). In this
+case, the branches are also on average approximately 50\% predictable.  However,
+on average over random input, with a fixed choice of pivot the $i$ and $j$
+branches are about 66\% biased (that is, they are taken about 66\% of the time).
+It is straightforward to show why this is so. Assume that the data to be
+partitioned is a random permutation of $\lbrace 1, \ldots, n \rbrace$.  If the
+chosen pivot has rank $q$ then the $i$ branch will be taken with probability $(q
+- 1)/n$.  Moreover, the branch will be executed (but not necessarily taken) a
+total of $q$ times, since there are exactly $q - 1$ elements to the left of the
+pivot. The average bias is given by averaging over all possible choices of pivot
+$q$, thus
 
@@
 \noindent
-So the bias rapidly approaches 66\% as $n$ increases. With this bias, the $i$ and $j$ branches are about 71\% predictable. 
-The fact that predictability of these branches exceeds their biases
-may at first seem paradoxical. The reason the branches are more predictable than biased is because the bias 
-is an average. The average predictability of the $i$ and $j$ branches can be obtained very simply from
-Equation \ref{bias_equation} using the steady state predictability function, $CP$, introduced in Section \ref{bubble_sort}, thus
+So the bias rapidly approaches 66\% as $n$ increases. With this bias, the $i$
+and $j$ branches are about 71\% predictable.  The fact that predictability of
+these branches exceeds their biases may at first seem paradoxical. The reason
+the branches are more predictable than biased is because the bias is an average.
+The average predictability of the $i$ and $j$ branches can be obtained very
+simply from Equation \ref{bias_equation} using the steady state predictability
+function, $CP$, introduced in Section \ref{bubble_sort}, thus
 
@@
 \noindent 
-As $n \rightarrow \infty$ we have $P_{avg} = \int_0^1 CP(q)\,dq = 3/2 - \pi/4$.
-Thus we see $P_{avg}$ is about 0.71. This estimated average predictability is close to what is observed experimentally when using a fixed pivot
-over random data, as the predictability entry of the first row of Table \ref{median_table} shows.
-
-If the pivot is chosen as the median-of-3, then the $i$ and $j$ branches
-are about 60\% biased in the taken direction, and about 64\% predictable. It is especially
-noteworthy that although median-of-3 gives around a 14\% reduction in the total number of executed
-comparison branches, there is actually a 6\% \textit{increase} in the total
-number of branch mispredictions. As the pivot more closely approximates the median of the input,
-there is a continuing trend of a reduction in the total number of executed branches, as well as reducing bias and predictability 
-with an increasing total number of mispredictions, as Table \ref{median_table} shows.
+As $n \rightarrow \infty$ we have 
+$P_{avg} = \int_0^1 CP(q)\,dq = 3/2 - \pi/4$.  Thus we see $P_{avg}$ is about
+0.71. This estimated average predictability is close to what is observed
+experimentally when using a fixed pivot over random data, as the predictability
+entry of the first row of Table \ref{median_table} shows.
+
+If the pivot is chosen as the median-of-3, then the $i$ and $j$ branches are
+about 60\% biased in the taken direction, and about 64\% predictable. It is
+especially noteworthy that although median-of-3 gives around a 14\% reduction in
+the total number of executed comparison branches, there is actually a 6\%
+\textit{increase} in the total number of branch mispredictions. As the pivot
+more closely approximates the median of the input, there is a continuing trend
+of a reduction in the total number of executed branches, as well as reducing
+bias and predictability with an increasing total number of mispredictions, as
+Table \ref{median_table} shows.
 
@@
 \caption{The effect of choosing a pivot closer to the median on the prediction
-of the $i$ and $j$ branches (the results are the same for both). As a higher order median is used the bias of the $i$ and $j$
-branches is reduced, with a corresponding reduction in their predictability.
-The reductions in the number of executed branches are relative to a fixed choice of pivot (i.e. median-of-1). 
-The misprediction increases correspond to the total number
-of branches mispredicted with a higher order median compared to a fixed choice of pivot.
-These measurements were obtained 
-from our own software simulations of a bimodal branch predictor. Note that the small number of branches
-required to compute the median approximations are included in these results.}
+of the $i$ and $j$ branches (the results are the same for both). As a higher
+order median is used the bias of the $i$ and $j$ branches is reduced, with a
+corresponding reduction in their predictability.  The reductions in the number
+of executed branches are relative to a fixed choice of pivot (i.e. median-of-1).
+The misprediction increases correspond to the total number of branches
+mispredicted with a higher order median compared to a fixed choice of pivot.
+These measurements were obtained from our own software simulations of a bimodal
+branch predictor. Note that the small number of branches required to compute the
+median approximations are included in these results.}
 \label{median_table} 
@@
 
-The use of the median-of-3 or higher order median approximation greatly reduces the
-chances of the worst case behaviour of quicksort, and on average therefore results
-in fewer comparison branches being executed. From an information theoretic
-perspective, each comparison branch resolves more uncertainty, or yields
-more information, and is hence harder to predict. This observation applies also
-to the predictability of the branches used in the binary searched version of 
-multi-quicksort versus the sequentially searched version. There is a trade-off
-to be had between the penalty induced by unpredictable comparisons, versus the 
-increased instruction count caused by making comparisons more predictable.
-
-In work concurrent and independent of our own, Kaligosi and Sanders \citeyear{Kaligosi+06} 
-have investigated the effect of the choice of pivot on the performance of quicksort. They found that
-an artificially skewed pivot gives better performance than using a random pivot or the median-of-3,
-due to the reduction in branch mispredictions. This performance increase is despite the increased instruction count
-such an artificially skewed pivot causes, we describe this work in more detail in Section \ref{RelatedWork}.
+The use of the median-of-3 or higher order median approximation greatly reduces
+the chances of the worst case behaviour of quicksort, and on average therefore
+results in fewer comparison branches being executed. From an information
+theoretic perspective, each comparison branch resolves more uncertainty, or
+yields more information, and is hence harder to predict. This observation
+applies also to the predictability of the branches used in the binary searched
+version of multi-quicksort versus the sequentially searched version. There is a
+trade-off to be had between the penalty induced by unpredictable comparisons,
+versus the increased instruction count caused by making comparisons more
+predictable.
+
+In work concurrent to and independent of our own, Kaligosi and Sanders
+\citeyear{Kaligosi+06} have investigated the effect of the choice of pivot on
+the performance of quicksort. They found that an artificially skewed pivot gives
+better performance than using a random pivot or the median-of-3, due to the
+reduction in branch mispredictions. This performance increase is despite the
+increased instruction count such an artificially skewed pivot causes; we
+describe this work in more detail in Section \ref{RelatedWork}.
 
@@
 
-The purpose of this paper is to show the relevance of branch mispredictions to sorting. We now show how to develop a  
-radix sort \cite{Friend56} implementation for 32-bit integers that, in all our tests, is more efficient than all other sorting algorithms presented in this paper. 
-This is in part due to the tiny number of branch mispredictions incurred by radix sorting. It is also in part because radix sort 
-operates in linear worst case time. However, in spite of its better asymptotic performance a ``step'' for our radix sort is quite costly, requiring
-a key to be counted four times and copied four times, with the counting involving several layers of indirection. In contrast, a step for quicksort
-is as simple as a comparison and a possible exchange every few steps.
-
-The cache performance of radix sort is generally poor compared
-to algorithms like mergesort or quicksort. In radix sort, each key is used just once each time it is in the cache, when the data to be sorted
-is larger than the cache every key access will result in a cache miss. On the other hand, mergesort and quicksort have good temporal locality and use keys 
-repeatedly before they are ejected from the cache.
+The purpose of this paper is to show the relevance of branch mispredictions to
+sorting. We now show how to develop a  radix sort \cite{Friend56} implementation
+for 32-bit integers that, in all our tests, is more efficient than all other
+sorting algorithms presented in this paper.  This is in part due to the tiny
+number of branch mispredictions incurred by radix sorting. It is also in part
+because radix sort operates in linear worst case time. However, in spite of its
+better asymptotic performance a ``step'' for our radix sort is quite costly,
+requiring a key to be counted four times and copied four times, with the
+counting involving several layers of indirection. In contrast, a step for
+quicksort is as simple as a comparison and a possible exchange every few steps.
+
+The cache performance of radix sort is generally poor compared to algorithms
+like mergesort or quicksort. In radix sort, each key is used just once each time
+it is in the cache, when the data to be sorted is larger than the cache every
+key access will result in a cache miss. On the other hand, mergesort and
+quicksort have good temporal locality and use keys repeatedly before they are
+ejected from the cache.
 
@@
 
-Our basic radix sort implementation is a typical least significant digit radix sort. We operate in radix 256, and begin by counting the number of keys
-with a particular least significant byte by indexing a table of 256 entries with the least significant bytes of the keys. The entries of this table are then summed to 
-store the cumulative count of keys with a smaller least significant byte. Now each key is placed in an auxiliary array according to the value retrieved by 
-using its least significant byte to index into the table of cumulative counts. Before the next iteration, the auxiliary array is written back to the 
-original array and the sort can proceed on the next most significant byte. We use this implementation to compare against the simple caching improvements
-we now suggest.
-
-We can firstly remove the copying back and forth between the original and auxiliary array, by just alternating their roles on each step. Since we are
-sorting 32 bit integers, the number of iterations is even and the data will end up sorted in its original location. If the number of steps was odd, an
-additional copy would be required at the end. Since radix sort is out-of-place, we might consider aligning its two arrays as we did with mergesort (see
-Section \ref{mergesort_variations}). However, radix sort does not access its arrays in the same regular fashion that mergesort does and so is unlikely to
-benefit from such a change. Indeed, we implemented a double-aligned radix sort and did not observe any performance gains.
-Finally, we use an improvement suggested by LaMarca and Ladner \citeyear{LaMarca97}, where the counting 
-of the next most significant bytes is begun during their preceding iteration. We refer to this optimized version of radix sort as memory-tuned
-radix sort.
-
-In our basic radix sort implementation, the first array is accessed 3 times and the second array 4 times, in each of the four
-iterations over the four bytes in the 32-bit integers. This gives 20 accesses in total.
-In memory-tuned radix sort each key of the two arrays is accessed only nine times in total.
-At the start, it traverses the original array once to count the least significant bytes. It then iterates four times, positioning
-the keys based on the previously established counts. These four iterations cause
-a total of eight accesses, because each array is accessed once per iteration.
+Our basic radix sort implementation is a typical least significant digit radix
+sort. We operate in radix 256, and begin by counting the number of keys with a
+particular least significant byte by indexing a table of 256 entries with the
+least significant bytes of the keys. The entries of this table are then summed
+to store the cumulative count of keys with a smaller least significant byte. Now
+each key is placed in an auxiliary array according to the value retrieved by
+using its least significant byte to index into the table of cumulative counts.
+Before the next iteration, the auxiliary array is written back to the original
+array and the sort can proceed on the next most significant byte. We use this
+implementation to compare against the simple caching improvements we now
+suggest.
+
+We can firstly remove the copying back and forth between the original and
+auxiliary array, by just alternating their roles on each step. Since we are
+sorting 32 bit integers, the number of iterations is even and the data will end
+up sorted in its original location. If the number of steps was odd, an
+additional copy would be required at the end. Since radix sort is out-of-place,
+we might consider aligning its two arrays as we did with mergesort (see Section
+\ref{mergesort_variations}). However, radix sort does not access its arrays in
+the same regular fashion that mergesort does and so is unlikely to benefit from
+such a change. Indeed, we implemented a double-aligned radix sort and did not
+observe any performance gains.  Finally, we use an improvement suggested by
+LaMarca and Ladner \citeyear{LaMarca97}, where the counting of the next most
+significant bytes is begun during their preceding iteration. We refer to this
+optimized version of radix sort as memory-tuned radix sort.
+
+In our basic radix sort implementation, the first array is accessed 3 times and
+the second array twice, in each of the four iterations over the four bytes in
+the 32-bit integers. This gives 20 accesses in total.  In memory-tuned radix
+sort each key of the two arrays is accessed only nine times in total.  At the
+start, it traverses the original array once to count the least significant
+bytes. It then iterates four times, positioning the keys based on the previously
+established counts. These four iterations cause a total of eight accesses,
+because each array is accessed once per iteration.
 
@@
 \end{tabular}
-\caption{(a) Shows the cycles per key of our radix sort implementations alongside our base quicksort implementation, measured using Pentium 4
-hardware performance counters. (b) Shows the level 2 cache 
-misses incurred by our radix sort implementations and a multi-quicksort implementation for a 2MB direct mapped cache with 32-byte cache lines. These
-results were generated using \texttt{sim-cache}.
-(c) Shows the average branch mispredictions per key 
-incurred by our radix sort implementations while using bimodal predictors with 4096 table entries, the results for two-level adaptive predictors are extremely
+\caption{(a) Shows the cycles per key of our radix sort implementations
+alongside our base quicksort implementation, measured using Pentium 4 hardware
+performance counters. (b) Shows the level 2 cache misses incurred by our radix
+sort implementations and a multi-quicksort implementation for a 2MB direct
+mapped cache with 32-byte cache lines. These results were generated using
+\texttt{sim-cache}.  (c) Shows the average branch mispredictions per key
+incurred by our radix sort implementations while using bimodal predictors with
+4096 table entries, the results for two-level adaptive predictors are extremely
 similar. Finally, (d) shows the average number of branches per key executed by
-our  implementations. The results of (c) and (d) were generated using \texttt{sim-bpred}.}
+our  implementations. The results of (c) and (d) were generated using
+\texttt{sim-bpred}.}
 \label{radixsort_results_pictures}
@@
 
-Our radix sort implementation requires fewer processor cycles per key than even our base quicksort 
-implementation as Figure \ref{radixsort_results_pictures}(a) shows. The number of level 2 cache misses is only slightly higher than we measured for
-quicksort while the array fits inside the cache, presumably because radix sort is out-of-place.
-
-Figure \ref{radixsort_results_pictures}(b) shows the level 2 cache misses for our basic and memory-tuned implementations.
-We note that, when the array no longer fits inside the cache the memory-tuned implementation has approximately half the number of misses per key
-(1.25) than the basic version (2.5). When the data no longer fits inside the cache there should be about one miss for each key per pass per cache line.
-Each cache line can fit 8 keys, and thus we expect about 2 misses per key since there are 16 passes in total. The 2.5 misses per key which was measured
-likely exceeds this because of the irregular fashion of the array indexing in each pass. Similary, in memory-tuned radix sort we expect about 1.125
-misses per key, because there are 9 passes, again the measured 1.25 misses per key presumably exceeds this because
-of the irregular array indexing.
-
-The large advantage of radix sort is its almost complete lack of branch mispredictions however. Figure \ref{radixsort_results_pictures}(c) shows the tiny
-number of branch mispredictions incurred by radix sort. It also executes a similarly tiny number of branch instructions as 
-Figure \ref{radixsort_results_pictures}(d) shows.
+Our radix sort implementation requires fewer processor cycles per key than even
+our base quicksort implementation as Figure \ref{radixsort_results_pictures}(a)
+shows.
+
+Figure \ref{radixsort_results_pictures}(b) shows the level 2 cache misses for
+our basic and memory-tuned implementations.  The number of level 2 cache misses
+is only slightly higher than we measured for quicksort while the array fits
+inside the cache, presumably because radix sort is out-of-place.  We note that
+when the array no longer fits inside the cache the memory-tuned implementation
+has approximately half the number of misses per key (1.25) than the basic
+version (2.5). When the data no longer fits inside the cache there should be
+about one miss for each key per pass per cache line.  Each cache line can fit 8
+keys, and thus we expect the observed 2.5 misses per key since there are 20
+passes in total.  likely exceeds this because of the irregular fashion of the
+array indexing in each pass. Similary, in memory-tuned radix sort we expect
+slightly below 1.25 misses per key since there are 9 passes in total.
+
+The large advantage of radix sort is its almost complete lack of branch
+mispredictions however. Figure \ref{radixsort_results_pictures}(c) shows the
+tiny number of branch mispredictions incurred by radix sort. It also executes a
+similarly tiny number of branch instructions as Figure
+\ref{radixsort_results_pictures}(d) shows.
 
@@
 
-We present the first large-scale systematic study of branch prediction
-and sorting. However, researchers have been aware for many years that
-the branches in comparison-based sorts cause problems on pipelined
-architectures. As early as 1972 Knuth commented on the
-``ever-increasing number of `pipeline' or `number crunching' computers
-that have appeared in recent years'' whose ``efficiency deteriorates
-noticeably in the presence of conditional branch instructions unless
-the branch almost always goes the same way''. He also notes that ``radix
-sorting is usually more efficient than any other known method for
-internal sorting on such machines'' (the same comment appears in \cite{KnuthVol3_98}). Most likely, Knuth
-was referring to the IBM 7030 ``Stretch'' computer (released in 1961
-with a four-stage pipeline), Seymour Cray's Freon-cooled CDC 6600
-supercomputer (released 1964), or other early supercomputers which
-used pipelining in the execution of general-purpose instructions.
-
-A number of other authors have considered optimized implementations of
-sorting algorithms for more recent single-processor machines. Nyberg
-\textit{et al} \citeyear{Nyberg+94} present a quicksort-based algorithm for RISC
-architectures. They found that the main limit on
-performance was cache behaviour, and did not mention branch prediction
-even once. Bentley and McIlroy \citeyear{Bentley+93} implemented a highly-tuned quicksort,
-to be used in the \texttt{qsort} function of the standard C library. Again, they did not consider branch prediction,
-perhaps because the first desktop processors with dynamic branch
-predictors (such as the DEC Alpha 21064, MIPS R8000 and Intel Pentium)
-were only just appearing around that time.
-
-Agarwal \citeyear{Agarwal96} developed an optimized algorithm for sorting random records on
-an IBM RS/6000 RISC machine. His is the first
-published algorithm that we are aware of that deliberately attempts to
-improve performance by eliminating branch mispredictions. First, it
-uses the seven higher order bits of the key to divide the data into
-128 buckets. Each of these buckets is radix-sorted. However, before
-the radix sort is reaches the lowest order bits, the probability of
-any adjacent pair of keys being out of order is extremely low, because
-the problem states that the data is guaranteed to be random. In the
-final stage, the algorithm checks that large runs of keys are
-correctly ordered using subtraction and bitwise operations, and only
-sorts a section if some key is found to be out of order. Although
-Agarwal's techniques are interesting, they are very dependent on the
-randomness of the data; it is easy to construct commmon cases that
-would result in very poor performance. Nonetheless, the work is
-interesting because it shows the effectiveness of radix sorting
-techniques on pipelined architectures.
-
-A recent trend in optimizing algorithms for FFT, linear algebra, and
-digital signal processing has been to automatically generate and
-evaluate very large numbers of variants of the algorithm.  A similar
-approch is used by Li \textit{et al} \citeyear{Li+05} to automatically generate sorting
-algorithms that are tuned to the architecture of the target machine
-and the characteristics of the input data. They identify six sorting
-primitives, which can be combined to construct a sorting algorithm.  A
-large number of combinations are tested, using a genetic algorithms
-search to find the most efficient. The resulting implementations are,
-in many cases, significantly faster than commercial sorting libraries.
-Although they do not consider branch prediction explicitly, they find
-that the implementations that perform best usually use radix sort for
-a large part of the sorting process.
-
-Sanders and Winkel \citeyear{Sanders+04} investigate the use of \emph{predicated}
-instructions on Intel Itanium 2 processors. In
-addition to its normal inputs, a predicated instruction takes a
-predicate register input. If the value in the predicate register is
-true, the instruction takes effect; otherwise it does not. Predication
-is normally implemented by the instruction being executed regardless
-of whether the predicate is true. However, its result is only
-\textit{written back} to a register or memory if the predicate is
-true. Although predicated instructions use execution resources
-regardless of whether they are allowed to write back, they allow
-conditional branches to be eliminated.  Sanders and Winkel show how
-the partition step in quicksort can be rewritten to use predicated
-instructions on an Itanium 2, a highly instruction-level parallel
-machine. Itanium 2 provides large numbers of parallel functional
-units and registers which make such trade-offs worthwhile.
-
-In  
-preliminary experiments, we found that Sanders and Winkel's work is difficult 
-to reproduce on a Pentium 4 based machine. The only predicated instruction
-supported on the Pentium 4 is the conditional move instruction. We found that
-rewriting
-quicksort's loops so that they are suitable for predication
-reduces their efficiency and also increases the number of simultaneously
-live variables. Since the Pentium 4 has only eight general
-purpose registers the number of simultaneously live variables exceeded
-the number of available registers, reducing efficiency. In addition,
-we also discovered that on the Pentium 4 conditional moves are
-surprisingly expensive.
-
-Mudge \textit{et al} \citeyear{Mudge+96} examine the behaviour of the $i$ and $j$
-comparison branches in randomized quicksort, as an example to demonstrate a
-general method for estimating the limit on the predictability of
-branches in a program. They argue that the optimal
-predictor for these branches would 
-keep a running count of the total
-number of array elements examined so far that are
-greater than the pivot. If the majority of the elements examined so far
-are greater than the pivot, the next element is predicted as greater
-than the pivot, and vice versa. They estimate that this approach will give a
-prediction accuracy of 75\%, which is close to our
-result for these branches, using a bimodal predictor.
-This shows that a bimodal predictor is close to optimal
-for randomized quicksort's branches, since $P_{avg}$ of Section \ref{quicksort_branch_results_text}
-is about 0.71.
-
-Brodal \textit{et al} \citeyear{Brodal+05} examine the adaptiveness of quicksort.
-They find that although randomized quicksort is not adaptive (i.e. its asymptotic analysis does not 
-improve with respect to some measure of presortedness), its performance
-is better on data with a low number of inversions. They justify this by showing 
-that the expected number of element swaps performed by randomized quicksort falls as
-the number of inversions does. The number of branch mispredictions
-is roughly twice the number of element swaps, because two branch mispredictions
-tend to occur when the two while loops of the partition code exit (see Figure \ref{partition_code}).
-Furthermore they note that the branch mispredictions are the dominant part of the running time of
-randomized quicksort. This provides an empirical relation between the presortedness of the input, 
-randomized quicksort's performance, and the number of branch mispredictions.
-
-In closely related work Brodal and Moruz \citeyear{BrodalMoruz05} provide a lower bound on the
-number of branch mispredictions a comparison-based sorting algorithm must incur based on the
-number of comparisons it performs. They also show that a mergesort variation can achieve this lower
-bound based on the number of comparisons it performs, although they do not provide experimental
-performance characteristics. They also extend their results to adaptive sorting algorithms
-
-In recent work conducted independently of our own, Kaligosi and Sanders \citeyear{Kaligosi+06}
-investigate the choice of pivot on the performance of quicksort. They examine $\alpha$-skewed
-pivoting, that is, where the pivot has rank $\alpha n$ when sorting $n$ keys. They examine the performance
-of quicksort when the pivot is determined randomly, from the median-of-3, or is the true median
-(i.e. a 1/2-skewed pivot). They show that using a 1/10-skewed pivot gives a performance increase 
-over all the aforementioned pivot choices. This is despite the increase in instruction count that a 
-deliberately skewed pivot causes.
+We present the first large-scale systematic study of branch prediction and
+sorting. However, researchers have been aware for many years that the branches
+in comparison-based sorts cause problems on pipelined architectures. As early as
+1972 Knuth commented on the ``ever-increasing number of `pipeline' or `number
+crunching' computers that have appeared in recent years'' whose ``efficiency
+deteriorates noticeably in the presence of conditional branch instructions
+unless the branch almost always goes the same way''. He also notes that ``radix
+sorting is usually more efficient than any other known method for internal
+sorting on such machines'' (the same comment appears in \cite{KnuthVol3_98}).
+Most likely, Knuth was referring to the IBM 7030 ``Stretch'' computer (released
+in 1961 with a four-stage pipeline), Seymour Cray's Freon-cooled CDC 6600
+supercomputer (released 1964), or other early supercomputers which used
+pipelining in the execution of general-purpose instructions.
+
+A number of other authors have considered optimized implementations of sorting
+algorithms for more recent single-processor machines. Nyberg \textit{et al}
+\citeyear{Nyberg+94} present a quicksort-based algorithm for RISC architectures.
+They found that the main limit on performance was cache behaviour, and did not
+mention branch prediction even once. Bentley and McIlroy \citeyear{Bentley+93}
+implemented a highly-tuned quicksort, to be used in the \texttt{qsort} function
+of the standard C library. Again, they did not consider branch prediction,
+perhaps because the first desktop processors with dynamic branch predictors
+(such as the DEC Alpha 21064, MIPS R8000 and Intel Pentium) were only just
+appearing around that time.
+
+Agarwal \citeyear{Agarwal96} developed an optimized algorithm for sorting random
+records on an IBM RS/6000 RISC machine. His is the first published algorithm
+that we are aware of that deliberately attempts to improve performance by
+eliminating branch mispredictions. First, it uses the seven higher order bits of
+the key to divide the data into 128 buckets. Each of these buckets is
+radix-sorted. However, before the radix sort is reaches the lowest order bits,
+the probability of any adjacent pair of keys being out of order is extremely
+low, because the problem states that the data is guaranteed to be random. In the
+final stage, the algorithm checks that large runs of keys are correctly ordered
+using subtraction and bitwise operations, and only sorts a section if some key
+is found to be out of order. Although Agarwal's techniques are interesting, they
+are very dependent on the randomness of the data; it is easy to construct
+commmon cases that would result in very poor performance. Nonetheless, the work
+is interesting because it shows the effectiveness of radix sorting techniques on
+pipelined architectures.
+
+A recent trend in optimizing algorithms for FFT, linear algebra, and digital
+signal processing has been to automatically generate and evaluate very large
+numbers of variants of the algorithm.  A similar approch is used by Li
+\textit{et al} \citeyear{Li+05} to automatically generate sorting algorithms
+that are tuned to the architecture of the target machine and the characteristics
+of the input data. They identify six sorting primitives, which can be combined
+to construct a sorting algorithm.  A large number of combinations are tested,
+using a genetic algorithms search to find the most efficient. The resulting
+implementations are, in many cases, significantly faster than commercial sorting
+libraries.  Although they do not consider branch prediction explicitly, they
+find that the implementations that perform best usually use radix sort for a
+large part of the sorting process.
+
+Sanders and Winkel \citeyear{Sanders+04} investigate the use of
+\emph{predicated} instructions on Intel Itanium 2 processors. In addition to its
+normal inputs, a predicated instruction takes a predicate register input. If the
+value in the predicate register is true, the instruction takes effect; otherwise
+it does not. Predication is normally implemented by the instruction being
+executed regardless of whether the predicate is true. However, its result is
+only \textit{written back} to a register or memory if the predicate is true.
+Although predicated instructions use execution resources regardless of whether
+they are allowed to write back, they allow conditional branches to be
+eliminated.  Sanders and Winkel show how the partition step in quicksort can be
+rewritten to use predicated instructions on an Itanium 2, a highly
+instruction-level parallel machine. Itanium 2 provides large numbers of parallel
+functional units and registers which make such trade-offs worthwhile.
+
+\comment{ In preliminary experiments, we found that Sanders and Winkel's work is
+difficult to reproduce on a Pentium 4 based machine. The only predicated
+instruction supported on the Pentium 4 is the conditional move instruction. We
+found that rewriting quicksort's loops so that they are suitable for predication
+reduces their efficiency and also increases the number of simultaneously live
+variables.  Since the Pentium 4 has only eight general purpose registers the
+number of simultaneously live variables exceeded the number of available
+registers, reducing efficiency. In addition, we also discovered that on the
+Pentium 4 conditional moves are surprisingly expensive.  } Mudge \textit{et al}
+\citeyear{Mudge+96} examine the behaviour of the $i$ and $j$ comparison branches
+in randomized quicksort, as an example to demonstrate a general method for
+estimating the limit on the predictability of branches in a program. They argue
+that the optimal predictor for these branches would keep a running count of the
+total number of array elements examined so far that are greater than the pivot.
+If the majority of the elements examined so far are greater than the pivot, the
+next element is predicted as greater than the pivot, and vice versa. They
+estimate that this approach will give a prediction accuracy of 75\%, which is
+close to our result for these branches, using a bimodal predictor.  This shows
+that a bimodal predictor is close to optimal for randomized quicksort's
+branches, since $P_{avg}$ of Section \ref{quicksort_branch_results_text} is
+about 0.71.
+
+Brodal \textit{et al} \citeyear{Brodal+05} examine the adaptiveness of
+quicksort.  They find that although randomized quicksort is not adaptive (i.e.
+its asymptotic analysis does not improve with respect to some measure of
+presortedness), its performance is better on data with a low number of
+inversions. They justify this by showing that the expected number of element
+swaps performed by randomized quicksort falls as the number of inversions does.
+The number of branch mispredictions is roughly twice the number of element
+swaps, because two branch mispredictions tend to occur when the two while loops
+of the partition code exit (see Figure \ref{partition_code}).  Furthermore they
+note that the branch mispredictions are the dominant part of the running time of
+randomized quicksort. This provides an empirical relation between the
+presortedness of the input, randomized quicksort's performance, and the number
+of branch mispredictions.
+
+In closely related work Brodal and Moruz \citeyear{BrodalMoruz05} provide a
+lower bound on the number of branch mispredictions a comparison-based sorting
+algorithm must incur based on the number of comparisons it performs. They also
+show that a mergesort variation can achieve this lower bound based on the number
+of comparisons it performs, although they do not provide experimental
+performance characteristics. They also extend their results to adaptive sorting
+algorithms
+
+In recent work conducted independently of our own, Kaligosi and Sanders
+\citeyear{Kaligosi+06} investigate the choice of pivot on the performance of
+quicksort. They examine $\alpha$-skewed pivoting, that is, where the pivot has
+rank $\alpha n$ when sorting $n$ keys. They examine the performance of quicksort
+when the pivot is determined randomly, from the median-of-3, or is the true
+median (i.e. a 1/2-skewed pivot). They show that using a 1/10-skewed pivot gives
+a performance increase over all the aforementioned pivot choices. This is
+despite the increase in instruction count that a deliberately skewed pivot
+causes.
 
@@
 
-This paper has presented the first large scale study on the characteristics of branch instructions in 
-all of the most common sorting algorithms. We noted that except where the sorting algorithms involve reasonably complicated
-flow control (i.e. shellsort, multi-mergesort, multi-quicksort), bimodal predictors preform just as well as two-level adaptive predictors. Indeed we
-noted that for shaker sort a bimodal predictor out-performs a two-level adaptive predictor.
-
-Of the elementary sorting algorithms we noted that insertion sort,
-causing just a single misprediction per key, has the best branch prediction characteristics. On the other hand,
-bubble and shaker sort operate in a manner which causes surprisingly unpredictable branches.
-
-Sorting (and indeed, searching) algorithms have been designed to minimize the number of comparisons necessary 
-in the worst case. However, minimizing the number of comparisons makes each comparison less predictable. When a branch is
-mispredicted, there is often a large penalty on modern processors. 
-There is therefore a tension between reducing the number of executed comparisons and the penalty associated with 
-increasing the chance of branch mispredictions. This tension is especially evident in the choice of pivot for quicksort.
-
-Since cache optimizations are now recognized as very important, we 
-have also noted the affect that these cache optimizations have on the branch prediction behaviour of sorting algorithms. In particular we
-provided performance results for cache optimized versions of quicksort, mergesort, heapsort and radix sort.
-
-Our study provides concrete results of the branch prediction characteristics for a large number
-of sorting algorithms, accompanied with the relevant analysis.  These branch prediction results have been obtained for realistic dynamic predictors, using
-a variety of configurations of bimodal and two-level adaptive predictors.
+This paper has presented the first large scale study on the characteristics of
+branch instructions in all of the most common sorting algorithms. We noted that
+except where the sorting algorithms involve reasonably complicated flow control
+(i.e. shellsort, multi-mergesort, multi-quicksort), bimodal predictors preform
+just as well as two-level adaptive predictors. Indeed we noted that for shaker
+sort a bimodal predictor out-performs a two-level adaptive predictor.
+
+Of the elementary sorting algorithms we noted that insertion sort, causing just
+a single misprediction per key, has the best branch prediction characteristics.
+On the other hand, bubble and shaker sort operate in a manner which causes
+surprisingly unpredictable branches.
+
+Sorting (and indeed, searching) algorithms have been designed to minimize the
+number of comparisons necessary in the worst case. However, minimizing the
+number of comparisons makes each comparison less predictable. When a branch is
+mispredicted, there is often a large penalty on modern processors.  There is
+therefore a tension between reducing the number of executed comparisons and the
+penalty associated with increasing the chance of branch mispredictions. This
+tension is especially evident in the choice of pivot for quicksort.
+
+Since cache optimizations are now recognized as very important, we have also
+noted the affect that these cache optimizations have on the branch prediction
+behaviour of sorting algorithms. In particular we provided performance results
+for cache optimized versions of quicksort, mergesort, heapsort and radix sort.
+
+Our study provides concrete results of the branch prediction characteristics for
+a large number of sorting algorithms, accompanied with the relevant analysis.
+These branch prediction results have been obtained for realistic dynamic
+predictors, using a variety of configurations of bimodal and two-level adaptive
+predictors.
 
@@
 
-In Sections \ref{bubble_sort}, \ref{heapsort_branch_results_text} and \ref{quicksort_branch_results_text}
-we made use of the steady state predictability function $CP$. When a branch instruction is executed 
-repeatedly with a known probability $p$ of being taken, $CP(p)$ gives a good estimate of the probability
-that it will be correctly predicted.
-
-The bimodal predictor associated with a branch instruction can be thought
-of as a Markov chain. Note that we are thus assuming that successive branches
-are independent. If the branch has probability $p$ of being taken then 
-the transition matrix of the associated Markov chain is
+In Sections \ref{bubble_sort}, \ref{heapsort_branch_results_text} and
+\ref{quicksort_branch_results_text} we made use of the steady state
+predictability function $CP$. When a branch instruction is executed repeatedly
+with a known probability $p$ of being taken, $CP(p)$ gives a good estimate of
+the probability that it will be correctly predicted.
+
+The bimodal predictor associated with a branch instruction can be thought of as
+a Markov chain. Note that we are thus assuming that successive branches are
+independent. If the branch has probability $p$ of being taken then the
+transition matrix of the associated Markov chain is
 
@@
 \noindent 
-Here $P(ST)$, $P(T)$, $P(NT)$ and $P(SNT)$ denote the respective probabilities of the predictor states strongly taken, taken, not-taken
-and strongly not-taken described in Section \ref{branch_prediction}. A branch is correctly predicted with probability $p$ if we are 
-in the strongly taken or taken state. Similarly, a branch is correctly predicted with probability $1 - p$ if we are in the strongly not-taken
-or not-taken state. Thus,
+Here $P(ST)$, $P(T)$, $P(NT)$ and $P(SNT)$ denote the respective
+probabilities of the predictor states strongly taken, taken, not-taken and
+strongly not-taken described in Section \ref{branch_prediction}. A branch is
+correctly predicted with probability $p$ if we are in the strongly taken or
+taken state. Similarly, a branch is correctly predicted with probability $1 - p$
+if we are in the strongly not-taken or not-taken state. Thus,
 

### 2007-01-29 docs/branch_paper/paper.tex
Make pictures fit better

@@
 \category{E.5}{Data}{Files}[Sorting/Searching]
-\category{C.1.1}{Computer Systems Organization}{Processor Architectures, Single Data Stream Architectures}[Pipeline architectures]
+\category{C.1.1}{Computer Systems Organization}{Processor Architectures, Other
+Architecture Styles}[Pipeline processors]
 \terms{Algorithms, Experimentation, Measurement, Performance}
-\keywords{Sorting, Branch prediction}
+\keywords{Sorting, Branch Prediction, Pipeline Architectures, Caching}
 
@@
 \begin{tabular}{cc}
-\includegraphics[width=0.5\textwidth]{plots/selection_insertion_sort_branches.eps} & \includegraphics[width=0.5\textwidth]{plots/bubble_shaker_sort_branches.eps} \\
+\includegraphics[width=0.48\textwidth]{plots/selection_insertion_sort_branches.eps} & \includegraphics[width=0.48\textwidth]{plots/bubble_shaker_sort_branches.eps} \\
 (a) & (b) \\
@@
 \begin{tabular}{cc}
-\includegraphics[width=0.5\textwidth]{plots/bubble_predictability.eps} & \includegraphics[width=0.5\textwidth]{plots/bubble_sweep_misses.eps} \\
+\includegraphics[width=0.48\textwidth]{plots/bubble_predictability.eps} & \includegraphics[width=0.48\textwidth]{plots/bubble_sweep_misses.eps} \\
 (a) & (b) \\
@@
 \begin{tabular}{cc}
-\includegraphics[width=0.5\textwidth]{plots/elementary_cache1_misses.eps} & \includegraphics[width=0.5\textwidth]{plots/elementary_cycles.eps}\\
+\includegraphics[width=0.48\textwidth]{plots/elementary_cache1_misses.eps} & \includegraphics[width=0.48\textwidth]{plots/elementary_cycles.eps}\\
 (a) & (b) \\
@@
 \begin{tabular}{cc}
-\includegraphics[width=0.5\textwidth]{plots/shellsort_branch_misses.eps} & \includegraphics[width=0.5\textwidth]{plots/shell_heap_cache_misses.eps} \\
+\includegraphics[width=0.48\textwidth]{plots/shellsort_branch_misses.eps} & \includegraphics[width=0.48\textwidth]{plots/shell_heap_cache_misses.eps} \\
 (a) & (b) \\
-\includegraphics[width=0.5\textwidth]{plots/shell_heap_L2_misses.eps} & \includegraphics[width=0.5\textwidth]{plots/shell_heap_cycles.eps} \\
+\includegraphics[width=0.48\textwidth]{plots/shell_heap_L2_misses.eps} & \includegraphics[width=0.48\textwidth]{plots/shell_heap_cycles.eps} \\
 (a) & (b) \\
@@
 \begin{tabular}{cc}
-\includegraphics[width=0.5\textwidth]{plots/heapsort_branch_counts.eps} & \includegraphics[width=0.5\textwidth]{plots/heapsort_branch_misses.eps}\\
+\includegraphics[width=0.48\textwidth]{plots/heapsort_branch_counts.eps} & \includegraphics[width=0.48\textwidth]{plots/heapsort_branch_misses.eps}\\
 (a) & (b) \\
-\includegraphics[width=0.5\textwidth]{plots/heapsort_cycle_counts.eps} & \includegraphics[width=0.5\textwidth]{plots/heapsort_8heap_branches.eps} \\
+\includegraphics[width=0.48\textwidth]{plots/heapsort_cycle_counts.eps} & \includegraphics[width=0.48\textwidth]{plots/heapsort_8heap_branches.eps} \\
 (c) & (d) \\
@@
 \begin{tabular}{cc}
-\includegraphics[width=0.5\textwidth]{plots/mergesort_branches_per_key.eps} & \includegraphics[width=0.5\textwidth]{plots/mergesort_branch_misses.eps} \\
+\includegraphics[width=0.48\textwidth]{plots/mergesort_branches_per_key.eps} & \includegraphics[width=0.48\textwidth]{plots/mergesort_branch_misses.eps} \\
 (a) & (b) \\
-\includegraphics[width=0.5\textwidth]{plots/mergesort_level_2_misses.eps} & \includegraphics[width=0.5\textwidth]{plots/mergesort_cycle_counts.eps} \\
+\includegraphics[width=0.48\textwidth]{plots/mergesort_level_2_misses.eps} & \includegraphics[width=0.48\textwidth]{plots/mergesort_cycle_counts.eps} \\
 (c) & (d) \\
@@
 \begin{tabular}{cc}
-\includegraphics[width=0.5\textwidth]{plots/counter_base_quicksort__median_of_3__0.eps} & \includegraphics[width=0.5\textwidth]{plots/counter_memory_tuned_quicksort_0.eps} \\
+\includegraphics[width=0.48\textwidth]{plots/counter_base_quicksort__median_of_3__0.eps} & \includegraphics[width=0.48\textwidth]{plots/counter_memory_tuned_quicksort_0.eps} \\
 (a) Basic quicksort & (b) Memory-tuned quicksort \\ 
-\includegraphics[width=0.5\textwidth]{plots/counter_multi_quicksort__binary_search__0.eps} & \includegraphics[width=0.5\textwidth]{plots/counter_multi_quicksort__sequential_search__0.eps} \\
+\includegraphics[width=0.48\textwidth]{plots/counter_multi_quicksort__binary_search__0.eps} & \includegraphics[width=0.48\textwidth]{plots/counter_multi_quicksort__sequential_search__0.eps} \\
 (c) Multi-quicksort (binary search) & (d) Multi-quicksort (sequential search)
@@
 \begin{tabular}{cc}
-\includegraphics[width=0.5\textwidth]{plots/quicksort_cache_misses.eps} & \includegraphics[width=0.5\textwidth]{plots/quicksort_instruction_counts.eps} \\
+\includegraphics[width=0.48\textwidth]{plots/quicksort_cache_misses.eps} & \includegraphics[width=0.48\textwidth]{plots/quicksort_instruction_counts.eps} \\
 (a) & (b) \\ 
-\includegraphics[width=0.5\textwidth]{plots/quicksort_branch_misses.eps} & \includegraphics[width=0.5\textwidth]{plots/quicksort_cycles.eps} \\
+\includegraphics[width=0.48\textwidth]{plots/quicksort_branch_misses.eps} & \includegraphics[width=0.48\textwidth]{plots/quicksort_cycles.eps} \\
 (c) & (d) \\
@@
 \begin{tabular}{cc}
-\includegraphics[width=0.5\textwidth]{plots/radix_cycles.eps} & \includegraphics[width=0.5\textwidth]{plots/radix_versus_quick_level_2_misses.eps} \\
+\includegraphics[width=0.48\textwidth]{plots/radix_cycles.eps} & \includegraphics[width=0.48\textwidth]{plots/radix_versus_quick_level_2_misses.eps} \\
 (a) & (b) \\
-\includegraphics[width=0.5\textwidth]{plots/radix_branch_misses.eps} & \includegraphics[width=0.5\textwidth]{plots/radix_branches.eps} \\
+\includegraphics[width=0.48\textwidth]{plots/radix_branch_misses.eps} & \includegraphics[width=0.48\textwidth]{plots/radix_branches.eps} \\
 (c) & (d) \\

### 2007-02-14 docs/branch_paper/paper.tex
Committing after a long time. IM logs added Minor updates to other stuff

@@
 WILLIAMS\footnote[2]{Supported by the Irish Research Council for Science,
-Engineering and Technology (IRCSET) and IBM.}, and DAVID GREGG \\University of
-Dublin, Trinity College
+Engineering and Technology (IRCSET) and IBM.}, and DAVID GREGG \\Trinity College
+Dublin
 }
@@
 \terms{Algorithms, Experimentation, Measurement, Performance}
-\keywords{Sorting, Branch Prediction, Pipeline Architectures, Caching}
+\keywords{Sorting, Branch Prediction, Pipeline Architectures}
 
